---
title: "QC — rRNA pU sites after dropping replicates with BACS r < 0.3 (both reps)"
author: "XSun"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, warning=FALSE, message=FALSE}
library(data.table)
library(dplyr)
library(stringr)
library(ggplot2)
library(patchwork)
library(gridExtra)
library(scales)

knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center"
)
```

# Introduction

For each cell line, we computed the Pearson correlation of each treated
replicate's rRNA %pU against **both replicates** of the BACS dataset (PMID
39349603). Replicates with **r < 0.3 in both BACS rep1 and rep2** were
identified as problematic and removed before re-pooling.

The re-pooled results are stored in
`pooled_sites_chisq_probrep_dropped/pooled_rRNA.txt` within each cell line's
results directory (original results in `pooled_sites_chisq/` are untouched).

The filtering cutoff is **FDR < 0.01 & rate_diff > 0.05 &
n_samples >= max(n_kept - 1, 1)**.

```{r config}
FDR_CUT  <- 0.01
DIFF_CUT <- 0.05
CUTOFF_LABEL <- "FDR < 0.01 & rate_diff > 0.05"

RESULTS_DIRS <- c("/project2/xinhe/xsun/pumseq/pumseq_results",
                  "/project/xinhe/xsun/pumseq/pumseq_results")

DROP_INFO <- list(
  CL18486 = list(dropped = c("13-32-15", "13-32-16"),                   kept = c("12-187-5", "12-187-6")),
  CL18498 = list(dropped = c("13-32-54"),                                kept = c("12-187-85", "12-187-86", "13-32-53")),
  CL18501 = list(dropped = c("13-32-43", "13-32-44"),                   kept = c("12-187-75", "12-187-76")),
  CL18502 = list(dropped = c("13-32-45", "13-32-46"),                   kept = c("12-187-77", "12-187-78")),
  CL18504 = list(dropped = c("13-32-47", "13-32-48"),                   kept = c("12-187-79", "12-187-80")),
  CL18505 = list(dropped = c("13-32-49", "13-32-50"),                   kept = c("12-187-81", "12-187-82")),
  CL18510 = list(dropped = c("13-32-35", "13-32-36"),                   kept = c("12-187-115", "12-187-35")),
  CL18516 = list(dropped = c("13-32-37", "13-32-38"),                   kept = c("12-187-116", "12-187-36")),
  CL18517 = list(dropped = c("13-32-39", "13-32-40"),                   kept = c("12-187-117", "12-187-37")),
  CL18519 = list(dropped = c("13-32-17", "13-32-18"),                   kept = c("12-187-7", "12-187-8")),
  CL18523 = list(dropped = c("13-32-41", "13-32-42"),                   kept = c("12-187-118", "12-187-38")),
  CL18862 = list(dropped = c("13-32-51", "13-32-52"),                   kept = c("12-187-83", "12-187-84")),
  CL18909 = list(dropped = c("13-32-23", "13-32-24"),                   kept = c("12-187-13", "12-187-14")),
  CL18912 = list(dropped = c("13-32-55", "13-32-56"),                   kept = c("12-187-87", "12-187-88")),
  CL18913 = list(dropped = c("13-32-25", "13-32-26"),                   kept = c("12-187-15", "12-187-16")),
  CL19039 = list(dropped = c("13-32-3", "13-32-4"),                     kept = c("13-32-59", "13-32-60")),
  CL19114 = list(dropped = c("13-32-5", "13-32-6"),                     kept = c("13-32-61", "13-32-62")),
  CL19116 = list(dropped = c("13-32-19"),                                kept = c("12-187-10", "12-187-9")),
  CL19137 = list(dropped = c("13-32-1", "13-32-2"),                     kept = c("13-32-57", "13-32-58")),
  CL19140 = list(dropped = c("13-32-13", "13-32-14"),                   kept = c("13-32-69", "13-32-70")),
  CL19143 = list(dropped = c("13-32-21", "13-32-22"),                   kept = c("12-187-11", "12-187-12")),
  CL19152 = list(dropped = c("13-32-10", "13-32-9"),                    kept = c("13-32-65", "13-32-66")),
  CL19159 = list(dropped = c("12-187-20", "13-32-29", "13-32-30"),     kept = c("12-187-19")),
  CL19193 = list(dropped = c("13-32-11", "13-32-12"),                   kept = c("13-32-67", "13-32-68")),
  CL19200 = list(dropped = c("13-32-7", "13-32-8"),                     kept = c("13-32-63", "13-32-64")),
  CL19204 = list(dropped = c("13-32-27", "13-32-28"),                   kept = c("12-187-17", "12-187-18")),
  CL19239 = list(dropped = c("13-32-33", "13-32-34"),                   kept = c("12-187-23", "12-187-24")),
  CL19257 = list(dropped = c("13-32-31", "13-32-32"),                   kept = c("12-187-21", "12-187-22"))
)

DROP_CLS <- names(DROP_INFO)
```

## Replicates dropped and kept

```{r rep-table}
rep_df <- do.call(rbind, lapply(DROP_CLS, function(cl) {
  info <- DROP_INFO[[cl]]
  data.frame(
    `Cell line` = cl,
    `Kept replicates` = paste(info$kept, collapse = ", "),
    `Dropped replicates (r < 0.3)` = paste(info$dropped, collapse = ", "),
    `N kept` = length(info$kept),
    `N dropped` = length(info$dropped),
    check.names = FALSE
  )
}))

knitr::kable(rep_df, caption = "Replicates kept and dropped (BACS r < 0.3 (both reps) threshold)")
```

# Cross-cell-line comparison (pooled data)

For each cell line, the best available pooled data is used: if replicates
were dropped, the re-pooled result from `pooled_sites_chisq_probrep_dropped/`
is used; otherwise the original `pooled_sites_chisq/` result is used.

```{r load-all-pooled}
find_results_dir <- function(cl) {
  for (rd in RESULTS_DIRS) {
    if (dir.exists(file.path(rd, cl))) return(file.path(rd, cl))
  }
  NULL
}

all_pooled_cls <- c()
for (rd in RESULTS_DIRS) {
  cls <- list.dirs(rd, recursive = FALSE, full.names = FALSE)
  cls <- cls[grepl("^CL\\d+$", cls)]
  for (cl in cls) {
    has_orig <- file.exists(file.path(rd, cl, "pooled_sites_chisq", "pooled_rRNA.txt"))
    has_drop <- file.exists(file.path(rd, cl, "pooled_sites_chisq_probrep_dropped", "pooled_rRNA.txt"))
    if (has_orig || has_drop) all_pooled_cls <- c(all_pooled_cls, cl)
  }
}
all_pooled_cls <- sort(unique(all_pooled_cls))

load_best_pooled <- function(cl) {
  wdir <- find_results_dir(cl)
  is_dropped <- cl %in% DROP_CLS
  if (is_dropped) {
    f <- file.path(wdir, "pooled_sites_chisq_probrep_dropped", "pooled_rRNA.txt")
    n_kept <- length(DROP_INFO[[cl]]$kept)
    min_nsamp <- max(n_kept - 1, 1)
  } else {
    f <- file.path(wdir, "pooled_sites_chisq", "pooled_rRNA.txt")
    min_nsamp <- 2
  }
  if (!file.exists(f)) return(NULL)
  dt <- fread(f, header = TRUE, sep = "\t", colClasses = list(character = "ref"))
  dt[, site := paste(ref, pos, strand, sep = ":")]
  dt[, cell_line := cl]
  attr(dt, "min_nsamp") <- min_nsamp
  dt
}

best_pooled <- setNames(lapply(all_pooled_cls, load_best_pooled), all_pooled_cls)
best_pooled <- Filter(Negate(is.null), best_pooled)
ALL_POOLED_CLS <- names(best_pooled)

filter_best <- function(dt) {
  mn <- attr(dt, "min_nsamp")
  dt[fdr < FDR_CUT & rate_diff > DIFF_CUT & n_samples >= mn]
}

cat("Total cell lines:", length(ALL_POOLED_CLS), "\n")
cat("Using dropped-rep data for:", length(intersect(DROP_CLS, ALL_POOLED_CLS)), "cell lines\n")
```

## Number of pooled sites per cell line

```{r cross-site-counts, fig.width=10, fig.height=12}
cross_count_df <- data.frame(
  cell_line = ALL_POOLED_CLS,
  n_sites = sapply(ALL_POOLED_CLS, function(cl) nrow(filter_best(best_pooled[[cl]])))
)
cross_count_df$cell_line <- factor(cross_count_df$cell_line, levels = ALL_POOLED_CLS)
n_per_panel <- ceiling(nrow(cross_count_df) / 5)
cross_count_df$panel <- rep(1:5, each = n_per_panel, length.out = nrow(cross_count_df))

cross_count_plots <- lapply(1:5, function(p) {
  sub <- cross_count_df[cross_count_df$panel == p, ]
  ggplot(sub, aes(x = cell_line, y = n_sites, fill = cell_line)) +
    geom_col(width = 0.6, show.legend = FALSE) +
    geom_text(aes(label = comma(n_sites)), vjust = -0.3, size = 3.5) +
    scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15)),
                       limits = c(0, max(cross_count_df$n_sites) * 1.15)) +
    labs(x = NULL, y = "# sites") +
    theme_bw(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})

wrap_plots(cross_count_plots, ncol = 1) +
  plot_annotation(title = "Number of pooled rRNA pU sites per cell line",
                  subtitle = CUTOFF_LABEL)
```

## %pU Distribution — pooled data

```{r cross-dist, fig.width=14, fig.height=20}
cross_dist_plots <- lapply(ALL_POOLED_CLS, function(cl) {
  df <- as.data.frame(filter_best(best_pooled[[cl]])[, .(psi = ratio)])
  if (nrow(df) == 0) return(NULL)
  med_val <- median(df$psi); mean_val <- mean(df$psi)
  ggplot(df, aes(x = psi)) +
    geom_histogram(aes(y = after_stat(density)), binwidth = 0.02,
                   fill = "steelblue", color = "white", alpha = 0.7) +
    geom_density(color = "navy", linewidth = 0.7) +
    geom_vline(xintercept = med_val, color = "red", linetype = "dashed") +
    geom_vline(xintercept = mean_val, color = "darkorange", linetype = "dotted") +
    annotate("text", x = Inf, y = Inf,
             label = sprintf("n = %d\nmedian = %.3f\nmean = %.3f",
                             nrow(df), med_val, mean_val),
             hjust = 1.05, vjust = 1.05, size = 3, family = "mono") +
    scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
    labs(title = cl, x = "%pU (pooled)", y = "Density") +
    theme_bw(base_size = 10)
})
wrap_plots(Filter(Negate(is.null), cross_dist_plots), ncol = 5)
```

## Cross-cell-line Jaccard index

```{r cross-jaccard, fig.width=max(8, length(ALL_POOLED_CLS)*0.6), fig.height=max(7, length(ALL_POOLED_CLS)*0.5)}
jaccard <- function(a, b) {
  n_int <- length(intersect(a, b)); n_uni <- length(union(a, b))
  if (n_uni == 0) NA_real_ else n_int / n_uni
}

site_sets <- setNames(lapply(ALL_POOLED_CLS, function(cl) filter_best(best_pooled[[cl]])$site), ALL_POOLED_CLS)
n_cl <- length(ALL_POOLED_CLS)
j_mat <- matrix(NA_real_, n_cl, n_cl, dimnames = list(ALL_POOLED_CLS, ALL_POOLED_CLS))
n_int_mat <- matrix(NA_integer_, n_cl, n_cl, dimnames = list(ALL_POOLED_CLS, ALL_POOLED_CLS))
diag(j_mat) <- 1
for (i in 1:(n_cl-1)) for (j in (i+1):n_cl) {
  sa <- site_sets[[ALL_POOLED_CLS[i]]]; sb <- site_sets[[ALL_POOLED_CLS[j]]]
  j_mat[i,j] <- j_mat[j,i] <- jaccard(sa, sb)
  ni <- length(intersect(sa, sb)); n_int_mat[i,j] <- n_int_mat[j,i] <- ni
}

jac_df <- expand.grid(x = ALL_POOLED_CLS, y = ALL_POOLED_CLS, stringsAsFactors = FALSE) %>%
  rowwise() %>%
  mutate(j = j_mat[x, y], n_int = n_int_mat[x, y],
         label = if (x == y) "1.000" else if (is.na(j)) "NA"
                 else sprintf("%.3f\n(n=%d)", j, n_int)) %>%
  ungroup() %>%
  mutate(x = factor(x, levels = ALL_POOLED_CLS), y = factor(y, levels = rev(ALL_POOLED_CLS)))

ggplot(jac_df, aes(x = x, y = y, fill = j)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = label), size = 1.8, lineheight = 0.9) +
  scale_fill_gradient2(low = "#1a9641", mid = "#ffffbf", high = "#d7191c",
                       midpoint = 0.5, limits = c(0, 1), na.value = "grey90", name = "Jaccard") +
  labs(title = "Jaccard index — pooled rRNA sites across cell lines",
       subtitle = CUTOFF_LABEL, x = NULL, y = NULL) +
  coord_fixed() +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        axis.text.y = element_text(size = 7),
        panel.grid = element_blank(), panel.border = element_blank())
```

## Cross-cell-line Pearson's r of %pU

```{r cross-pearson, fig.width=max(8, length(ALL_POOLED_CLS)*0.6), fig.height=max(7, length(ALL_POOLED_CLS)*0.5)}
r_mat <- matrix(NA_real_, n_cl, n_cl, dimnames = list(ALL_POOLED_CLS, ALL_POOLED_CLS))
n_com_mat <- matrix(NA_integer_, n_cl, n_cl, dimnames = list(ALL_POOLED_CLS, ALL_POOLED_CLS))
diag(r_mat) <- 1
for (i in 1:(n_cl-1)) for (j in (i+1):n_cl) {
  d1 <- filter_best(best_pooled[[ALL_POOLED_CLS[i]]])[, .(site, ratio1 = ratio)]
  d2 <- filter_best(best_pooled[[ALL_POOLED_CLS[j]]])[, .(site, ratio2 = ratio)]
  m <- merge(d1, d2, by = "site")
  if (nrow(m) < 3) next
  r_mat[i,j] <- r_mat[j,i] <- cor(m$ratio1, m$ratio2, method = "pearson")
  n_com_mat[i,j] <- n_com_mat[j,i] <- nrow(m)
}

cor_df <- expand.grid(x = ALL_POOLED_CLS, y = ALL_POOLED_CLS, stringsAsFactors = FALSE) %>%
  rowwise() %>%
  mutate(r = r_mat[x, y], n_com = n_com_mat[x, y],
         label = if (x == y) "1.000" else if (is.na(r)) "NA"
                 else sprintf("%.3f\n(n=%d)", r, n_com)) %>%
  ungroup() %>%
  mutate(x = factor(x, levels = ALL_POOLED_CLS), y = factor(y, levels = rev(ALL_POOLED_CLS)))

ggplot(cor_df, aes(x = x, y = y, fill = r)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = label), size = 1.8, lineheight = 0.9) +
  scale_fill_gradient2(low = "#1a9641", mid = "#ffffbf", high = "#d7191c",
                       midpoint = 0.5, limits = c(0, 1), na.value = "grey90", name = "Pearson r") +
  labs(title = "Pearson's r of %pU — pooled rRNA sites across cell lines",
       subtitle = CUTOFF_LABEL, x = NULL, y = NULL) +
  coord_fixed() +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
        axis.text.y = element_text(size = 7),
        panel.grid = element_blank(), panel.border = element_blank())
```

---

# Comparing with published data (BACS)

```{r bacs-load}
path_BACS <- "/project2/xinhe/xsun/pumseq/data_others/pU_PMID39349603.xlsx"
rrna_BACS <- readxl::read_excel(path_BACS, sheet = 2)
colnames(rrna_BACS) <- rrna_BACS[2, ]
rrna_BACS <- rrna_BACS[-c(1:2), ]
rrna_BACS <- rrna_BACS[complete.cases(rrna_BACS$chr), ]
rrna_BACS_clean <- rrna_BACS %>%
  mutate(accession = str_extract(chr, "NR_[0-9]+"), pos_num = as.integer(pos))
n_bacs <- nrow(rrna_BACS_clean)
```

```{r bacs-overlap, fig.width=10, fig.height=15}
overlap_df <- do.call(rbind, lapply(ALL_POOLED_CLS, function(cl) {
  dt <- filter_best(best_pooled[[cl]])
  ours <- as.data.frame(dt) %>% mutate(accession = str_extract(ref, "NR_[0-9]+"))
  n_overlap <- nrow(inner_join(ours, rrna_BACS_clean, by = c("accession", "pos" = "pos_num")))
  data.frame(cell_line = cl,
             category = c("Our sites", "BACS sites", "Overlap"),
             count = c(nrow(ours), n_bacs, n_overlap))
}))
overlap_df$category <- factor(overlap_df$category, levels = c("Our sites", "BACS sites", "Overlap"))
overlap_df$cell_line <- factor(overlap_df$cell_line, levels = ALL_POOLED_CLS)

n_per_panel <- ceiling(length(ALL_POOLED_CLS) / 5)
cl_panels <- split(ALL_POOLED_CLS, rep(1:5, each = n_per_panel, length.out = length(ALL_POOLED_CLS)))

bacs_plots <- lapply(cl_panels, function(cls) {
  sub <- overlap_df[overlap_df$cell_line %in% cls, ]
  sub$cell_line <- factor(sub$cell_line, levels = cls)
  ggplot(sub, aes(x = cell_line, y = count, fill = category)) +
    geom_col(position = position_dodge(width = 0.7), width = 0.6) +
    geom_text(aes(label = count), position = position_dodge(width = 0.7),
              vjust = -0.3, size = 3) +
    scale_fill_manual(values = c("Our sites" = "#4E79A7", "BACS sites" = "#F28E2B",
                                  "Overlap" = "#59A14F")) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    labs(x = NULL, y = "# sites", fill = NULL) +
    theme_bw(base_size = 11) +
    theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
})
bacs_plots[[1]] <- bacs_plots[[1]] + theme(legend.position = "top")

wrap_plots(bacs_plots, ncol = 1) +
  plot_annotation(title = "Overlap of pooled rRNA pU sites with BACS",
                  subtitle = CUTOFF_LABEL)
```

### Before vs. after comparison

```{r bacs-before-after, fig.width=14, fig.height=6}
comp_df <- do.call(rbind, lapply(DROP_CLS, function(cl) {
  wdir <- find_results_dir(cl)
  n_kept <- length(DROP_INFO[[cl]]$kept)
  min_nsamp <- max(n_kept - 1, 1)

  f_drop <- file.path(wdir, "pooled_sites_chisq_probrep_dropped", "pooled_rRNA.txt")
  dt_drop <- fread(f_drop, header = TRUE, sep = "\t", colClasses = list(character = "ref"))
  filt_drop <- dt_drop[fdr < FDR_CUT & rate_diff > DIFF_CUT & n_samples >= min_nsamp]
  ours_drop <- as.data.frame(filt_drop) %>% mutate(accession = str_extract(ref, "NR_[0-9]+"))
  ov_drop <- nrow(inner_join(ours_drop, rrna_BACS_clean, by = c("accession", "pos" = "pos_num")))

  f_orig <- file.path(wdir, "pooled_sites_chisq", "pooled_rRNA.txt")
  ov_orig <- NA_integer_
  if (file.exists(f_orig)) {
    dt_orig <- fread(f_orig, header = TRUE, sep = "\t", colClasses = list(character = "ref"))
    filt_orig <- dt_orig[fdr < FDR_CUT & rate_diff > DIFF_CUT & n_samples >= 2]
    ours_orig <- as.data.frame(filt_orig) %>% mutate(accession = str_extract(ref, "NR_[0-9]+"))
    ov_orig <- nrow(inner_join(ours_orig, rrna_BACS_clean, by = c("accession", "pos" = "pos_num")))
  }

  rbind(
    data.frame(cell_line = cl, version = "Before", overlap = ov_orig,
               pct = round(ov_orig / n_bacs * 100, 1)),
    data.frame(cell_line = cl, version = "After", overlap = ov_drop,
               pct = round(ov_drop / n_bacs * 100, 1)))
}))
comp_df$version <- factor(comp_df$version, levels = c("Before", "After"))

ggplot(comp_df, aes(x = cell_line, y = pct, fill = version)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = sprintf("%d/%d\n(%.0f%%)", overlap, n_bacs, pct)),
            position = position_dodge(width = 0.7), vjust = -0.2, size = 2.8,
            lineheight = 0.85) +
  scale_fill_manual(values = c("Before" = "#E15759", "After" = "#59A14F")) +
  scale_y_continuous(limits = c(0, 110), breaks = seq(0, 100, 20),
                     labels = function(x) paste0(x, "%")) +
  labs(title = "BACS overlap: before vs. after dropping replicates (r < 0.3)",
       subtitle = CUTOFF_LABEL, x = NULL, y = "% BACS sites recovered", fill = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "top", axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r bacs-table}
summary_df <- do.call(rbind, lapply(DROP_CLS, function(cl) {
  wdir <- find_results_dir(cl)
  n_kept <- length(DROP_INFO[[cl]]$kept)
  min_nsamp <- max(n_kept - 1, 1)

  f_drop <- file.path(wdir, "pooled_sites_chisq_probrep_dropped", "pooled_rRNA.txt")
  dt_drop <- fread(f_drop, header = TRUE, sep = "\t", colClasses = list(character = "ref"))
  filt_drop <- dt_drop[fdr < FDR_CUT & rate_diff > DIFF_CUT & n_samples >= min_nsamp]
  ours_drop <- as.data.frame(filt_drop) %>% mutate(accession = str_extract(ref, "NR_[0-9]+"))
  n_drop <- nrow(ours_drop)
  ov_drop <- nrow(inner_join(ours_drop, rrna_BACS_clean, by = c("accession", "pos" = "pos_num")))

  f_orig <- file.path(wdir, "pooled_sites_chisq", "pooled_rRNA.txt")
  n_orig <- NA_integer_; ov_orig <- NA_integer_
  if (file.exists(f_orig)) {
    dt_orig <- fread(f_orig, header = TRUE, sep = "\t", colClasses = list(character = "ref"))
    filt_orig <- dt_orig[fdr < FDR_CUT & rate_diff > DIFF_CUT & n_samples >= 2]
    ours_orig <- as.data.frame(filt_orig) %>% mutate(accession = str_extract(ref, "NR_[0-9]+"))
    n_orig <- nrow(ours_orig)
    ov_orig <- nrow(inner_join(ours_orig, rrna_BACS_clean, by = c("accession", "pos" = "pos_num")))
  }

  data.frame(`Cell line` = cl, `N kept` = n_kept,
             `Sites (before)` = n_orig, `Overlap (before)` = ov_orig,
             `% BACS (before)` = round(ov_orig / n_bacs * 100, 1),
             `Sites (after)` = n_drop, `Overlap (after)` = ov_drop,
             `% BACS (after)` = round(ov_drop / n_bacs * 100, 1),
             check.names = FALSE)
}))

knitr::kable(summary_df, caption = "BACS overlap: before vs. after dropping replicates (r < 0.3)")
```
