---
title: "5-fold reproducibility filtering of joint-pooled pU site calling"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r setup, warning=FALSE, message=FALSE}
library(data.table)
library(ggplot2)

knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", fig.width = 9, fig.height = 5
)

ANNO <- "/project/xinhe/xsun/pumseq/pumseq_processed/joint/fold5_union_analysis/fold5_union_sites_annotated.txt.gz"
feat_levels <- c("5UTR", "CDS", "3UTR", "ncRNA", "Intron", "Intergenic")
N_FOLDS <- 5
```

# Strategy

Instead of calling pU sites once from all cell lines pooled, we partition
the 51 cell lines into **5 disjoint folds** (~20% each; seed = 2026), 
**joint-call** sites within each fold (pool that
fold's ~10 cell lines' reads, one chi-squared test per site,
**FDR < 0.01, rate_diff > 0.05, depth >= 20**), then combine across folds.

<!-- This turns the number of folds into a **reproducibility filter**: keeping -->
<!-- sites called in `>= k` of 5 folds. `k = 1` is the plain union (any fold); -->
<!-- higher `k` requires a site to reproduce across independent 20% subsets. The -->
<!-- motivation: the earlier discovery/replication analysis showed only ~40-60% -->
<!-- of sites called from a single 20% pool replicate, so a plain union carries -->
<!-- a lot of non-reproducible sites — requiring `>= k` folds should filter them. -->

```{r load}
sites <- fread(ANNO, colClasses = list(character = c("ref", "strand")))
sites[, feature := factor(feature, levels = feat_levels)]
cat("Total unique sites across the 5 folds:", nrow(sites), "\n")
```

# Reproducibility: how many folds called each site?

```{r n-folds}
nf <- sites[, .N, by = n_folds][order(n_folds)]
# knitr::kable(nf, col.names = c("n_folds", "n_sites"),
#              caption = "Number of sites called in exactly n folds")

ggplot(sites, aes(x = factor(n_folds))) +
  geom_bar(fill = "steelblue") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.4) +
  labs(x = "called in exactly n folds", y = "number of sites",
       title = "Site reproducibility across the 5 folds") +
  theme_bw(base_size = 13)
```

<!-- A large share of the union is **single-fold** sites — called by only one -->
<!-- 20% subset. These are the sites a plain union would keep but that don't -->
<!-- reproduce. -->

# Threshold curve: site count, feature mix, and BACS overlap vs. `>= k` folds

```{r threshold}
thr <- rbindlist(lapply(1:N_FOLDS, function(k) {
  sub <- sites[n_folds >= k]
  data.table(
    min_folds             = k,
    n_sites               = nrow(sub),
    pct_cds_utr           = round(100 * mean(sub$feature %in% c("CDS", "5UTR", "3UTR")), 1),
    pct_intron_intergenic = round(100 * mean(sub$feature %in% c("Intron", "Intergenic")), 1),
    n_bacs                = sum(sub$in_bacs)
  )
}))
knitr::kable(thr, caption = "Sites called in >= k of 5 folds: count, CDS/UTR %, intron+intergenic %, BACS overlap")
```

# Feature composition by threshold

```{r feature-comp, fig.height=5}
comp <- rbindlist(lapply(1:N_FOLDS, function(k) {
  sub <- sites[n_folds >= k]
  t <- as.data.table(prop.table(table(sub$feature)) * 100)
  setnames(t, c("feature", "pct")); t[, min_folds := k]; t
}))
comp[, feature := factor(feature, levels = feat_levels)]

ggplot(comp, aes(x = factor(min_folds), y = pct, fill = feature)) +
  geom_col() +
  labs(x = "site called in >= k folds", y = "percent of sites", fill = "feature",
       title = "Genomic feature composition vs. fold-reproducibility threshold") +
  theme_bw(base_size = 13)
```



<!-- # Findings -->

<!-- - **Reproducibility strongly enriches for validated (BACS) sites.** BACS -->
<!--   overlap rises from `r thr[min_folds==1, pct_bacs]`% (plain union) to -->
<!--   `r thr[min_folds==5, pct_bacs]`% (all 5 folds) — a -->
<!--   `r round(thr[min_folds==5, pct_bacs] / thr[min_folds==1, pct_bacs], 1)`× -->
<!--   enrichment. The single-fold-only sites are nearly devoid of BACS overlap -->
<!--   (`r thr[min_folds==1, n_bacs] - thr[min_folds==2, n_bacs]` BACS sites out -->
<!--   of `r sites[n_folds==1, .N]` single-fold sites), i.e. they are mostly -->
<!--   spurious — the fold filter removes them effectively. **A reproducibility -->
<!--   filter is well justified.** -->
<!-- - **It does NOT enrich for CDS/UTR — it slightly de-enriches.** CDS+UTR -->
<!--   fraction *drops* (`r thr[min_folds==1, pct_cds_utr]`% → -->
<!--   `r thr[min_folds==5, pct_cds_utr]`%) and intron+intergenic rises as more -->
<!--   folds are required. So the genomic feature distribution is *not* a good -->
<!--   "realness" discriminator here; BACS overlap is the trustworthy signal, and -->
<!--   it says reproducibility works even though the feature mix does not shift as -->
<!--   one might expect. -->

<!-- # Recommendation -->

<!-- - **Avoid the plain union (`>= 1`)** — `r thr[min_folds==1, pct_bacs]`% BACS -->
<!--   and dominated by non-reproducible single-fold sites. -->
<!-- - **`>= 2` folds** is the natural minimum (analogous to the existing -->
<!--   `n_called >= 2`): it roughly doubles BACS enrichment -->
<!--   (`r thr[min_folds==2, pct_bacs]`%) while keeping -->
<!--   `r format(thr[min_folds==2, n_sites], big.mark=",")` sites — the largest -->
<!--   single jump in quality. -->
<!-- - **`>= 5` folds** for maximum purity (`r thr[min_folds==5, pct_bacs]`% BACS) -->
<!--   at the cost of dropping to `r format(thr[min_folds==5, n_sites], big.mark=",")` -->
<!--   sites. -->

<!-- The `>= 2` to `>= 5` range is the precision-vs-sensitivity choice. Note this -->
<!-- used the stricter FDR < 0.01 calling filter, so even `>= 2` is fairly clean. -->
