---
title: "5-fold reproducibility filtering of joint-pooled pU site calling"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r setup, warning=FALSE, message=FALSE}
library(data.table)
library(ggplot2)

knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", fig.width = 9, fig.height = 5
)

ANNO <- "/project/xinhe/xsun/pumseq/pumseq_processed/joint/fold5_union_analysis/fold5_union_sites_annotated.txt.gz"
feat_levels <- c("5UTR", "CDS", "3UTR", "ncRNA", "Intron", "Intergenic")
N_FOLDS <- 5
```

# Strategy

Instead of calling pU sites once from all cell lines pooled, we partition
the 51 cell lines into **5 disjoint folds** (~20% each; seed = 2026), 
**joint-call** sites within each fold (pool that
fold's ~10 cell lines' reads, one chi-squared test per site,
**FDR < 0.01, rate_diff > 0.05, depth >= 20**), then combine across folds.

<!-- This turns the number of folds into a **reproducibility filter**: keeping -->
<!-- sites called in `>= k` of 5 folds. `k = 1` is the plain union (any fold); -->
<!-- higher `k` requires a site to reproduce across independent 20% subsets. The -->
<!-- motivation: the earlier discovery/replication analysis showed only ~40-60% -->
<!-- of sites called from a single 20% pool replicate, so a plain union carries -->
<!-- a lot of non-reproducible sites — requiring `>= k` folds should filter them. -->

```{r load}
sites <- fread(ANNO, colClasses = list(character = c("ref", "strand")))
sites[, feature := factor(feature, levels = feat_levels)]
cat("Total unique sites across the 5 folds:", nrow(sites), "\n")
```

# Reproducibility: how many folds called each site?

```{r n-folds}
nf <- sites[, .N, by = n_folds][order(n_folds)]
# knitr::kable(nf, col.names = c("n_folds", "n_sites"),
#              caption = "Number of sites called in exactly n folds")

ggplot(sites, aes(x = factor(n_folds))) +
  geom_bar(fill = "steelblue") +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.4) +
  labs(x = "called in exactly n folds", y = "number of sites",
       title = "Site reproducibility across the 5 folds") +
  theme_bw(base_size = 13)
```

<!-- A large share of the union is **single-fold** sites — called by only one -->
<!-- 20% subset. These are the sites a plain union would keep but that don't -->
<!-- reproduce. -->

# Threshold curve: site count, feature mix, and BACS overlap vs. `>= k` folds

```{r threshold}
thr <- rbindlist(lapply(1:N_FOLDS, function(k) {
  sub <- sites[n_folds >= k]
  data.table(
    min_folds             = k,
    n_sites               = nrow(sub),
    pct_cds_utr           = round(100 * mean(sub$feature %in% c("CDS", "5UTR", "3UTR")), 1),
    pct_intron_intergenic = round(100 * mean(sub$feature %in% c("Intron", "Intergenic")), 1),
    n_bacs                = sum(sub$in_bacs)
  )
}))
knitr::kable(thr, caption = "Sites called in >= k of 5 folds: count, CDS/UTR %, intron+intergenic %, BACS overlap")
```

# Feature composition by threshold

```{r feature-comp, fig.height=5}
comp <- rbindlist(lapply(1:N_FOLDS, function(k) {
  sub <- sites[n_folds >= k]
  t <- as.data.table(prop.table(table(sub$feature)) * 100)
  setnames(t, c("feature", "pct")); t[, min_folds := k]; t
}))
comp[, feature := factor(feature, levels = feat_levels)]

ggplot(comp, aes(x = factor(min_folds), y = pct, fill = feature)) +
  geom_col() +
  labs(x = "site called in >= k folds", y = "percent of sites", fill = "feature",
       title = "Genomic feature composition vs. fold-reproducibility threshold") +
  theme_bw(base_size = 13)
```




# Agreement with BACS (%pU vs. BACS Ψ level)

For the sites overlapping BACS, how well does our %pU (mean of the pooled
`ratio` across the folds that called each site) agree with the orthogonal
BACS Ψ level? A high Pearson correlation is a validation that our
quantification is accurate. We also check whether agreement changes with
the fold-reproducibility threshold.

```{r bacs-cor}
bcor <- rbindlist(lapply(1:N_FOLDS, function(k) {
  sub <- sites[n_folds >= k & in_bacs & !is.na(mean_ratio)]
  data.table(min_folds = k, n_bacs = nrow(sub),
             pearson_r = round(cor(sub$mean_ratio, sub$bacs_psi), 3))
}))
knitr::kable(bcor, caption = "%pU (ours) vs. BACS Ψ level: Pearson r, by >= k folds")
```

```{r bacs-scatter, fig.height=5}
sb <- sites[in_bacs & !is.na(mean_ratio)]
r_all <- round(cor(sb$mean_ratio, sb$bacs_psi), 3)
ggplot(sb, aes(x = bacs_psi, y = mean_ratio, color = factor(n_folds))) +
  geom_point(size = 1, alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "BACS Ψ level", y = "our %pU (mean fold ratio)", color = "n_folds",
       title = sprintf("Our %%pU vs. BACS Ψ level at overlapping sites (Pearson r = %.3f)", r_all)) +
  theme_bw(base_size = 13)
```

Our %pU correlates strongly with the BACS Ψ level (r ≈ `r r_all`), and the
correlation is essentially flat across fold thresholds — the sites that
overlap BACS are quantified accurately regardless of how many folds called
them. So the fold filter enriches for *which* sites are real (BACS overlap
fraction rises with `k`), while %pU *accuracy* at the real sites is already
high at every threshold.
