---
title: "Joint pooling vs union-of-per-cell-line site calling"
author: "XSun"
date: "2026-06-12"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.width = 7, fig.height = 7)
library(data.table)
library(ggplot2)
library(corrplot)
library(eulerr)
library(patchwork)
library(dplyr)
library(stringr)
```


```{r config}
output_root <- "/project2/xinhe/xsun/pumseq/pumseq_results/joint"
joint_run   <- "test_3CL_mRNA"
union_run   <- "test_3CL_mRNA_union"

results_root <- "/project2/xinhe/xsun/pumseq/pumseq_results"
cell_lines   <- c("CL18870", "CL18916", "CL19099")

joint_bed <- file.path(output_root, joint_run, "joint_mRNA.pseusite.bed")
union_bed <- file.path(output_root, union_run, "union_mRNA.pseusite.bed")

pooled_files <- setNames(
  file.path(results_root, cell_lines, "pooled_sites_chisq", "pooled_mRNA.txt"),
  cell_lines
)
```

# Introduction

We are trying to find a way to call sites jointly across samples. It’s much easier in QTL analysis to have a common set of sites/features across all samples, then we can associate with genotypes.

We tried two strategies for getting a common mRNA pU site list across cell lines.
Both end with the same two kinds of output, used for downstream analyses:

- A **global site list** (`*_mRNA.pseusite.bed` / `.txt.gz`) — the sites
  considered "real" pU sites for this cohort, one row of cross-cell-line
  pooled stats (C, T, depth, ratio, pval, fdr, rate_diff, ...) per site.
- A **global_sites annotation + matrices**
  (`*_mRNA.global_sites.txt.gz`, `.ratio_matrix.txt.gz`,
  `.depth_matrix.txt.gz`) — for every site in the global list, every cell
  line's own %pU (`C/(C+T)`) and read depth (`C+T`), computed directly from
  that cell line's raw mpileup counts with **no per-cell-line depth or
  significance filter** (a cell line is `NA` only if it has zero reads at
  that site). The annotation table has a fixed number of columns regardless
  of cohort size; the matrices gain one column per cell line added.

## Joint

1. Pool mpileup read counts (treated and untreated separately) across all
   cell lines, position by position.
2. Run a single chi-squared test per site on the pooled counts, with a
   genome-wide BH-FDR.
3. Sites passing `fdr < fdr_chisq & rate_diff > ratediff_chisq & depth >=
   out_depth` define the global site list (`joint_mRNA.pseusite.bed`).
4. For that global site list, compute each cell line's %pU/depth directly
   from its own raw mpileup counts -> `joint_mRNA.global_sites.*`.

## Union

1. For each cell line, take its own already-pooled chi-squared calls
   (`pooled_sites_chisq/pooled_mRNA.txt`) and filter by
   `fdr < fdr_chisq_celline & rate_diff > ratediff_chisq & depth >=
   out_depth & n_samples >= min_replicates` -> per-cell-line called sites.
2. Take the union of `(ref, pos, strand)` across all cell lines' called
   sites, recording which cell line(s) called each site (`called_in`).
3. Look up cross-cell-line pooled stats (C, T, depth, ratio, pval, fdr,
   rate_diff) for the union sites from an existing Joint run
   (`joint_mRNA.pseusite.txt.gz`) — `fdr` here is the Joint run's
   genome-wide value, **not recomputed**. The union itself, with no further
   filtering, **is** the global site list (`union_mRNA.pseusite.bed`).
4. Same as Joint step 4: per-cell-line %pU/depth computed directly from raw
   mpileup counts -> `union_mRNA.global_sites.*`.

The key difference is *which sites end up in the global list*: Joint
requires significance in the pooled (mega-sample) test, while Union requires
significance in at least one individual cell line. Union's list is therefore
typically a superset of Joint's, and can include sites that are significant
in one cell line but diluted away once pooled with the others.

# Comparing the sites between the two strategies

```{r load-bed}
read_bed_sites <- function(path) {
  bed <- fread(path, header = FALSE,
                col.names = c("ref", "start", "pos", "name", "score", "strand"))
  bed[, ref := as.character(ref)]
  unique(bed[, .(ref, pos, strand)])
}

joint_sites <- read_bed_sites(joint_bed)
union_sites <- read_bed_sites(union_bed)
setkey(joint_sites, ref, pos, strand)
setkey(union_sites, ref, pos, strand)

cat("Joint strategy final sites :", nrow(joint_sites), "\n")
cat("Union strategy final sites :", nrow(union_sites), "\n")
```


```{r site-overlap-venn, fig.width=5, fig.height=5}
site_id <- function(dt) dt[, paste(ref, pos, strand, sep = ":")]

fit <- euler(list(Joint = site_id(joint_sites), Union = site_id(union_sites)))
plot(fit, quantities = TRUE, main = "Overlap of global site lists",
     fills = list(fill = c("#4E79A7", "#F28E2B"), alpha = 0.6))
```


# Per-cell-line %pU lookup

For every site in a strategy's final list, look up that site's per-cell-line
conversion rate (`ratio` = C / (C+T), i.e. %pU as a fraction) from each cell
line's own `pooled_sites_chisq/pooled_mRNA.txt`. A site missing from a given
cell line's table (insufficient depth in that cell line) becomes `NA` for
that cell line.

```{r load-ratios}
load_ratio <- function(path) {
  dt <- fread(path, select = c("ref", "pos", "strand", "ratio"),
               colClasses = list(character = c("ref", "strand")))
  setkey(dt, ref, pos, strand)
  dt
}

ratios <- lapply(pooled_files, load_ratio)
```

```{r build-pu-matrix}
build_pu_matrix <- function(sites, ratios) {
  out <- copy(sites)
  for (cl in names(ratios)) {
    out <- merge(out, ratios[[cl]], by = c("ref", "pos", "strand"), all.x = TRUE)
    setnames(out, "ratio", cl)
  }
  out
}

joint_pu <- build_pu_matrix(joint_sites, ratios)
union_pu <- build_pu_matrix(union_sites, ratios)
```

```{r pu-scatter-helper}
cl_pairs <- combn(cell_lines, 2, simplify = FALSE)

plot_pu_pair <- function(pu, pair) {
  sub <- pu[, ..pair]
  sub <- sub[complete.cases(sub)]
  r   <- cor(sub[[1]], sub[[2]], method = "pearson")
  ggplot(pu, aes(x = .data[[pair[1]]], y = .data[[pair[2]]]))+
    geom_point(alpha = 0.4, size = 0.8)+
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50")+
    coord_equal(xlim = c(0, 1), ylim = c(0, 1))+
    labs(title = sprintf("%s vs %s (r = %.3f)", pair[1], pair[2], r),
         x = sprintf("%%pU: %s", pair[1]), y = sprintf("%%pU: %s", pair[2]))+
    theme_bw()
}
```

## Correlation of %pU across cell lines

### Joint strategy

```{r corr-joint-table}
pearson  <- cor(joint_pu[, ..cell_lines], use = "pairwise.complete.obs", method = "pearson")

#knitr::kable(pearson, digits = 3, caption = "Joint: Pearson correlation of %pU")
```

```{r corr-joint-plot, fig.width=4, fig.height=4}
corrplot(pearson, method = "color", type = "upper", addCoef.col = "black",
         tl.col = "black", title = "Joint: Pearson r (%pU)", mar = c(0,0,2,0),
         col = colorRampPalette(c("#1a9641", "#ffffbf", "#d7191c"))(200))
```

```{r scatter-joint, fig.width=10, fig.height=3.5}
joint_scatter <- lapply(cl_pairs, function(p) plot_pu_pair(joint_pu, p))
wrap_plots(joint_scatter, ncol = 3) +
  plot_annotation(title = sprintf("Joint strategy: %%pU across cell lines (n = %d sites)", nrow(joint_pu)))
```

### Union strategy

```{r corr-union-table}
pearson_u  <- cor(union_pu[, ..cell_lines], use = "pairwise.complete.obs", method = "pearson")

#knitr::kable(pearson_u, digits = 3, caption = "Union: Pearson correlation of %pU")
```

```{r corr-union-plot, fig.width=4, fig.height=4}
corrplot(pearson_u, method = "color", type = "upper", addCoef.col = "black",
         tl.col = "black", title = "Union: Pearson r (%pU)", mar = c(0,0,2,0),
         col = colorRampPalette(c("#1a9641", "#ffffbf", "#d7191c"))(200))
```

```{r scatter-union, fig.width=10, fig.height=3.5}
union_scatter <- lapply(cl_pairs, function(p) plot_pu_pair(union_pu, p))
wrap_plots(union_scatter, ncol = 3) +
  plot_annotation(title = sprintf("Union strategy: %%pU across cell lines (n = %d sites)", nrow(union_pu)))
```

# Comparing with published data

We benchmark the **global site lists** from both strategies against
two independent published pU datasets. For each global site, `ratio` is the
pooled %pU (C / (C+T)) across all three cell lines (for Union this value is
looked up from the Joint run, see [Strategy descriptions](#strategy-descriptions)).

- **BACS** (PMID 39349603) HeLa polyA+ RNA Supplementary Table 8.
- **PRAISE** (PMID 36997645) HEK293T mRNAs Supplementary Dataset 2, filtered to mRNA
  only

Both published datasets use genomic coordinates, matching our `ref` / `pos` /
`strand` columns. BACS mRNA sites with `feature == "ncRNA"` are excluded.
PRAISE sites are filtered to `NM_` / `XM_` (mRNA) accessions, excluding
`NR_` / `XR_` (ncRNA).

## Load published data

```{r pub-load}
path_BACS   <- "/project2/xinhe/xsun/pumseq/data_others/pU_PMID39349603.xlsx"
path_PRAISE <- "/project2/xinhe/xsun/pumseq/data_others/pU_PMID36997645.xlsx"

# BACS mRNA: Supplementary Table 8 (sheet 6) - HeLa polyA+ RNA
mrna_BACS <- readxl::read_excel(path_BACS, sheet = 6, skip = 2)
mrna_BACS_clean <- mrna_BACS %>%
  filter(feature != "ncRNA") %>%
  rename(psi_level = `Ψ level`) %>%
  mutate(chr_num   = sub("^chr", "", chr),
         pos_num   = as.integer(pos),
         psi_level = as.numeric(psi_level))

# PRAISE mRNA+ncRNA: Supplementary Dataset 2 (sheet 2) - HEK293T
# Filter to mRNA only: keep NM_ and XM_ accessions.
mrna_PRAISE <- readxl::read_excel(path_PRAISE, sheet = 2, skip = 2)
mrna_PRAISE_clean <- mrna_PRAISE %>%
  filter(str_starts(chr_name, "NM_") | str_starts(chr_name, "XM_")) %>%
  mutate(
    chr_num       = sub("^chr", "", str_extract(chr_site, "^chr[^_]+")),
    genomic_start = as.integer(str_extract(chr_site, "(?<=_)[0-9]+")),
    genomic_end   = as.integer(coalesce(
      str_extract(chr_site, "(?<=-)[0-9]+$"),
      str_extract(chr_site, "(?<=_)[0-9]+")
    ))
  ) %>%
  filter(!is.na(chr_num))

cat("BACS mRNA sites  :", nrow(mrna_BACS_clean), "\n")
cat("PRAISE mRNA sites:", nrow(mrna_PRAISE_clean), "\n")
```

```{r pub-load-ours}
load_global_sites <- function(path) {
  dt <- fread(path, select = c("ref", "pos", "strand", "ratio"),
               colClasses = list(character = c("ref", "strand")))
  as.data.frame(dt)
}

joint_global <- load_global_sites(file.path(output_root, joint_run, "joint_mRNA.global_sites.txt.gz"))
union_global <- load_global_sites(file.path(output_root, union_run, "union_mRNA.global_sites.txt.gz"))

```

## Overlap with BACS and PRAISE

For each strategy's global site list we report how many sites overlap BACS
and PRAISE (by exact `chr/pos/strand` for BACS, and by `chr` + genomic range
for PRAISE), what fraction of **our** sites that represents, and what fraction
of the **published** sites we recover.

```{r pub-overlap}
overlap_bacs <- function(dt) {
  dt %>%
    mutate(chr_num = ref) %>%
    inner_join(mrna_BACS_clean, by = c("chr_num", "pos" = "pos_num", "strand"))
}

overlap_praise <- function(dt) {
  dt %>%
    mutate(chr_num = ref) %>%
    inner_join(mrna_PRAISE_clean, by = "chr_num", relationship = "many-to-many") %>%
    filter(pos >= pmin(genomic_start, genomic_end) &
           pos <= pmax(genomic_start, genomic_end))
}

summarize_overlap <- function(dt, label) {
  ov_bacs   <- overlap_bacs(dt)
  ov_praise <- overlap_praise(dt)
  data.frame(
    Strategy           = label,
    n_global_sites     = nrow(dt),
    n_BACS_overlap     = nrow(ov_bacs),
    pct_ours_in_BACS   = round(nrow(ov_bacs) / nrow(dt) * 100, 1),
    pct_BACS_in_ours   = round(nrow(ov_bacs) / nrow(mrna_BACS_clean) * 100, 1),
    n_PRAISE_overlap   = nrow(ov_praise),
    pct_ours_in_PRAISE = round(nrow(ov_praise) / nrow(dt) * 100, 1),
    pct_PRAISE_in_ours = round(nrow(ov_praise) / nrow(mrna_PRAISE_clean) * 100, 1)
  )
}

overlap_summary <- rbind(
  summarize_overlap(joint_global, "Joint"),
  summarize_overlap(union_global, "Union")
)

knitr::kable(
  overlap_summary,
  col.names = c("Strategy", "# global sites",
                 "# overlap BACS", "% ours in BACS", "% BACS in ours",
                 "# overlap PRAISE", "% ours in PRAISE", "% PRAISE in ours"),
  caption = "Overlap between our global site lists and published mRNA pU data"
)
```

## %pU vs published %pU / deletion ratio

For the overlapping sites, we compare our pooled %pU (`ratio`) against the
BACS Ψ level and the PRAISE deletion ratios (replicate 1 and 2).

```{r pub-scatter-helper}
make_pub_scatter <- function(df, xcol, ycol, ttl, xlab, ylab) {
  if (nrow(df) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = "No overlapping sites",
                 size = 4, color = "grey50") +
        labs(title = ttl, x = xlab, y = ylab) +
        theme_bw(base_size = 11)
    )
  }
  r_val <- cor(df[[xcol]], df[[ycol]], method = "pearson")
  ggplot(df, aes(x = .data[[xcol]], y = .data[[ycol]])) +
    geom_point(alpha = 0.6, size = 1.5, color = "#2c7bb6") +
    geom_smooth(method = "lm", se = TRUE, color = "grey40", linetype = "dashed") +
    annotate("text", x = -Inf, y = Inf,
             label = sprintf("r = %.3f\nn = %d", r_val, nrow(df)),
             hjust = -0.1, vjust = 1.3, size = 3.5) +
    labs(title = ttl, x = xlab, y = ylab) +
    theme_bw(base_size = 11)
}
```

### Joint strategy

```{r pub-scatter-joint, fig.width=12, fig.height=4}
j_bacs   <- overlap_bacs(joint_global)
j_praise <- overlap_praise(joint_global)

p1 <- make_pub_scatter(j_bacs, "ratio", "psi_level",
                        "Joint vs BACS", "Our %pU (Joint global sites)", "BACS psi level")
p2 <- make_pub_scatter(j_praise, "ratio", "rep1_deletion_ratio",
                        "Joint vs PRAISE (rep1)", "Our %pU (Joint global sites)", "PRAISE deletion ratio (rep1)")
p3 <- make_pub_scatter(j_praise, "ratio", "rep2_deletion_ratio",
                        "Joint vs PRAISE (rep2)", "Our %pU (Joint global sites)", "PRAISE deletion ratio (rep2)")

wrap_plots(p1, p2, p3, ncol = 3)
```

### Union strategy

```{r pub-scatter-union, fig.width=12, fig.height=4}
u_bacs   <- overlap_bacs(union_global)
u_praise <- overlap_praise(union_global)

p1 <- make_pub_scatter(u_bacs, "ratio", "psi_level",
                        "Union vs BACS", "Our %pU (Union global sites)", "BACS psi level")
p2 <- make_pub_scatter(u_praise, "ratio", "rep1_deletion_ratio",
                        "Union vs PRAISE (rep1)", "Our %pU (Union global sites)", "PRAISE deletion ratio (rep1)")
p3 <- make_pub_scatter(u_praise, "ratio", "rep2_deletion_ratio",
                        "Union vs PRAISE (rep2)", "Our %pU (Union global sites)", "PRAISE deletion ratio (rep2)")

wrap_plots(p1, p2, p3, ncol = 3)
```


