---
title: "QC — Joint cross-cell-line mRNA pU sites (51 cell lines)"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r setup, warning=FALSE, message=FALSE}
library(data.table)
library(dplyr)
library(stringr)
library(ggplot2)
library(patchwork)
library(corrplot)
library(scales)

knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", fig.width = 10, fig.height = 8
)
```

# Introduction

Quality control of the **joint cross-cell-line mRNA pU site calling** across
all 51 cell lines (with problematic replicates dropped based on BACS r < 0.3
in both reps).

The joint strategy pools treated and untreated reads across all cell lines,
runs a single chi-squared test per uridine position, then applies genome-wide
BH-FDR. The resulting **global site list** defines a common feature set.

For each global site, per-cell-line %pU is computed as C/(C+T) from each cell
line's raw mpileup data → ratio matrix (sites × 51 cell lines).

```{r load-data}
OUT_DIR <- "/project2/xinhe/xsun/pumseq/pumseq_results/joint/all51_mRNA"

global_sites <- fread(
  file.path(OUT_DIR, "joint_mRNA.global_sites.txt.gz"),
  header = TRUE, sep = "\t", colClasses = list(character = c("ref", "strand")))

ratio_mat <- fread(
  file.path(OUT_DIR, "joint_mRNA.global_sites.ratio_matrix.txt.gz"),
  header = TRUE, sep = "\t", colClasses = list(character = c("ref", "strand")))

depth_mat <- fread(
  file.path(OUT_DIR, "joint_mRNA.global_sites.depth_matrix.txt.gz"),
  header = TRUE, sep = "\t", colClasses = list(character = c("ref", "strand")))

CL_NAMES <- setdiff(colnames(ratio_mat), c("ref", "pos", "strand"))

# Baseline filter: FDR < 0.01, rate_diff > 0.05, depth >= 20
baseline_mask <- global_sites$fdr < 0.01 & global_sites$rate_diff > 0.05 & global_sites$depth >= 20
baseline_rows <- which(baseline_mask)
global_sites <- global_sites[baseline_mask, ]
ratio_mat <- ratio_mat[baseline_mask, ]
depth_mat <- depth_mat[baseline_mask, ]

n_sites <- nrow(global_sites)
n_cls <- length(CL_NAMES)

cat("Global sites (FDR < 0.01, rate_diff > 0.05, depth >= 20):", n_sites, "\n")
cat("Cell lines:", n_cls, "\n")

cutoff_combos <- list(
  list(fdr = 0.01,   diff = 0.05, depth = 20,  label = "FDR<0.01, diff>0.05, depth>=20"),
  list(fdr = 0.001,  diff = 0.05, depth = 30,  label = "FDR<0.001, diff>0.05, depth>=30"),
  list(fdr = 0.0001, diff = 0.05, depth = 50,  label = "FDR<0.0001, diff>0.05, depth>=50"),
  list(fdr = 0.01,   diff = 0.10, depth = 20,  label = "FDR<0.01, diff>0.10, depth>=20"),
  list(fdr = 0.001,  diff = 0.10, depth = 30,  label = "FDR<0.001, diff>0.10, depth>=30"),
  list(fdr = 0.0001, diff = 0.20, depth = 50,  label = "FDR<0.0001, diff>0.20, depth>=50")
)
```

# Cutoff recommendation

```{r recommendation-table}
path_BACS_early <- "/project2/xinhe/xsun/pumseq/data_others/pU_PMID39349603.xlsx"
bacs_early <- readxl::read_excel(path_BACS_early, sheet = 6, skip = 2) %>%
  filter(feature != "ncRNA") %>%
  mutate(chr_num = sub("^chr", "", chr),
         pos_num = as.integer(pos),
         psi_level = as.numeric(`Ψ level`))
n_bacs_early <- nrow(bacs_early)

rec_table <- do.call(rbind, lapply(cutoff_combos, function(cc) {
  idx <- which(global_sites$fdr < cc$fdr &
               global_sites$rate_diff > cc$diff &
               global_sites$depth >= cc$depth)
  sub <- global_sites[idx, ]
  ov <- sub %>% mutate(chr_num = ref) %>%
    inner_join(bacs_early, by = c("chr_num", "pos" = "pos_num", "strand"))
  r_val <- if (nrow(ov) >= 3) round(cor(ov$ratio, ov$psi_level,
                                         method = "pearson", use = "complete.obs"), 3) else NA
  data.frame(
    Cutoff = cc$label,
    Sites = length(idx),
    BACS_overlap = nrow(ov),
    pct_ours_in_BACS = round(nrow(ov) / length(idx) * 100, 1),
    pct_BACS_in_ours = round(nrow(ov) / n_bacs_early * 100, 1),
    Pearson_r = r_val
  )
}))

knitr::kable(rec_table,
             col.names = c("Cutoff", "# sites", "# BACS overlap",
                           "% ours in BACS", "% BACS in ours", "Pearson r"),
             caption = "Cutoff comparison — recommended: FDR < 0.001, diff > 0.05, depth >= 30")
```

**Recommended cutoff: FDR < 0.001, rate_diff > 0.05, depth >= 30** (~10,600 sites)

- BACS overlap is stable (~400 sites) across the top three cutoffs; raising
  rate_diff > 0.10 halves overlap, suggesting many real sites have subtle
  modification levels
- Pearson r with BACS remains high (0.92)
- 10K+ sites provides good power for downstream QTL mapping

# Summary statistics

## Site counts at different thresholds

```{r site-summary}
cat("Global sites (FDR < 0.01 & rate_diff > 0.05 & depth >= 20):", n_sites, "\n")
```

The heatmap below shows how the number of sites changes across different
combinations of FDR, rate_diff, and depth cutoffs. All cutoffs are subsets
of the global site list (FDR < 0.01, rate_diff > 0.05, depth >= 20).

```{r site-cutoff-heatmap, fig.width=12, fig.height=10}
fdr_cuts   <- c(0.01, 0.001, 0.0001)
diff_cuts  <- c(0.05, 0.10, 0.2)
depth_cuts <- c(20, 30, 50)

grid <- expand.grid(fdr = fdr_cuts, diff = diff_cuts, depth = depth_cuts)
grid$n_sites <- mapply(function(f, d, dp) {
  sum(global_sites$fdr < f & global_sites$rate_diff > d & global_sites$depth >= dp)
}, grid$fdr, grid$diff, grid$depth)

grid$fdr_label   <- factor(sprintf("FDR < %s", formatC(grid$fdr, format = "g")),
                           levels = sprintf("FDR < %s", formatC(fdr_cuts, format = "g")))
grid$diff_label  <- factor(sprintf("rate_diff > %.2f", grid$diff),
                           levels = sprintf("rate_diff > %.2f", diff_cuts))
grid$depth_label <- factor(sprintf("depth >= %d", grid$depth),
                           levels = sprintf("depth >= %d", depth_cuts))

ggplot(grid, aes(x = fdr_label, y = diff_label, fill = n_sites)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = scales::comma(n_sites)), size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#deebf7", high = "#2171b5",
                      name = "# sites", labels = scales::comma) +
  facet_wrap(~ depth_label, ncol = 2) +
  labs(title = "Number of sites at different cutoff combinations",
       subtitle = sprintf("Subsets of %s global sites (FDR < 0.01)", scales::comma(n_sites)),
       x = "FDR cutoff", y = "Rate difference cutoff") +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1),
        strip.background = element_rect(fill = "#e8eaf6"),
        strip.text = element_text(face = "bold"))
```

## Distribution of pooled %pU {.tabset}

Distribution of the mega-pooled conversion ratio C/(C+T) at different cutoffs.
Red dashed = median; orange dotted = mean.

```{r pooled-pU-dist, results='asis', fig.width=8, fig.height=5}
for (cc in cutoff_combos) {
  idx <- which(global_sites$fdr < cc$fdr &
               global_sites$rate_diff > cc$diff &
               global_sites$depth >= cc$depth)
  n_sub <- length(idx)

  cat(sprintf("\n\n### %s\n\n", cc$label))
  cat(sprintf("**%s sites**\n\n", scales::comma(n_sub)))

  if (n_sub < 10) { cat("Too few sites.\n\n"); next }

  sub <- global_sites[idx, ]
  med_val <- median(sub$ratio)
  mean_val <- mean(sub$ratio)

  print(
    ggplot(sub, aes(x = ratio)) +
      geom_histogram(aes(y = after_stat(density)), binwidth = 0.01,
                     fill = "steelblue", color = "white", alpha = 0.7) +
      geom_density(color = "navy", linewidth = 0.7) +
      geom_vline(xintercept = med_val, color = "red", linetype = "dashed") +
      geom_vline(xintercept = mean_val, color = "darkorange", linetype = "dotted") +
      annotate("text", x = Inf, y = Inf,
               label = sprintf("n = %s\nmedian = %.3f\nmean = %.3f",
                               scales::comma(n_sub), med_val, mean_val),
               hjust = 1.05, vjust = 1.05, size = 4, family = "mono") +
      labs(title = sprintf("Pooled %%pU distribution (%s)", cc$label),
           x = "%pU (C/(C+T), mega-pooled)", y = "Density") +
      theme_bw(base_size = 12)
  )
  cat("\n\n")
}
```

## Per-cell-line coverage and %pU distribution at different cutoffs {.tabset}

For each cutoff combination, we subset the global site list and show:

- **Coverage**: % of subset sites with non-null %pU per cell line
- **%pU distribution**: histogram of per-cell-line %pU at subset sites

```{r coverage-setup}
ratio_only <- as.matrix(ratio_mat[, ..CL_NAMES])
depth_only <- as.matrix(depth_mat[, ..CL_NAMES])
```

```{r coverage-cutoffs, results='asis', fig.width=10, fig.height=14}
for (cc in cutoff_combos) {
  idx <- which(global_sites$fdr < cc$fdr &
               global_sites$rate_diff > cc$diff &
               global_sites$depth >= cc$depth)
  n_sub <- length(idx)

  cat(sprintf("\n\n### %s\n\n", cc$label))
  cat(sprintf("**%s sites**\n\n", scales::comma(n_sub)))

  if (n_sub == 0) { cat("No sites at this cutoff.\n\n"); next }

  sub_ratio <- ratio_only[idx, , drop = FALSE]

  # Coverage
  cov_df <- data.frame(
    cell_line = CL_NAMES,
    pct_covered = round(colSums(!is.na(sub_ratio)) / n_sub * 100, 1)
  )
  cov_df$cell_line <- factor(cov_df$cell_line, levels = CL_NAMES)
  n_per <- ceiling(nrow(cov_df) / 5)
  cov_df$panel <- rep(1:5, each = n_per, length.out = nrow(cov_df))

  cov_plots <- lapply(1:5, function(p) {
    sub <- cov_df[cov_df$panel == p, ]
    ggplot(sub, aes(x = cell_line, y = pct_covered, fill = cell_line)) +
      geom_col(width = 0.6, show.legend = FALSE) +
      geom_text(aes(label = sprintf("%.0f%%", pct_covered)), vjust = -0.3, size = 2.5) +
      scale_y_continuous(limits = c(0, 105)) +
      labs(x = NULL, y = "% covered") +
      theme_bw(base_size = 9) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))
  })

  print(
    wrap_plots(cov_plots, ncol = 1) +
      plot_annotation(title = sprintf("Coverage (%s)", cc$label),
                      subtitle = sprintf("%s sites", scales::comma(n_sub)))
  )
  cat("\n\n")

  # %pU distribution
  # dist_plots <- lapply(CL_NAMES, function(cl) {
  #   vals <- sub_ratio[, cl]
  #   vals <- vals[!is.na(vals)]
  #   if (length(vals) == 0) return(NULL)
  #   df <- data.frame(pU = vals)
  #   med_v <- median(vals)
  #   ggplot(df, aes(x = pU)) +
  #     geom_histogram(aes(y = after_stat(density)), binwidth = 0.02,
  #                    fill = "steelblue", color = "white", alpha = 0.7) +
  #     geom_density(color = "navy", linewidth = 0.7) +
  #     geom_vline(xintercept = med_v, color = "red", linetype = "dashed") +
  #     annotate("text", x = Inf, y = Inf,
  #              label = sprintf("n=%d\nmed=%.3f", length(vals), med_v),
  #              hjust = 1.05, vjust = 1.05, size = 2, family = "mono") +
  #     scale_x_continuous(limits = c(0, 1)) +
  #     labs(title = cl, x = "%pU", y = "Density") +
  #     theme_bw(base_size = 7)
  # })
  # dist_plots <- Filter(Negate(is.null), dist_plots)
  # 
  # print(wrap_plots(dist_plots, ncol = 5))
  # cat("\n\n")
}
```



# Correlation of %pU across cell lines {.tabset}

Pairwise Pearson correlation of per-cell-line %pU at different cutoffs. High
correlations indicate consistent modification patterns across cell lines.

```{r cor-cutoffs, results='asis', fig.width=12, fig.height=12}
for (cc in cutoff_combos) {
  idx <- which(global_sites$fdr < cc$fdr &
               global_sites$rate_diff > cc$diff &
               global_sites$depth >= cc$depth)
  n_sub <- length(idx)

  cat(sprintf("\n\n## %s\n\n", cc$label))
  cat(sprintf("**%s sites**\n\n", scales::comma(n_sub)))

  if (n_sub < 10) { cat("Too few sites.\n\n"); next }

  sub_ratio <- ratio_only[idx, , drop = FALSE]
  cor_mat <- cor(sub_ratio, use = "pairwise.complete.obs", method = "pearson")

  corrplot(cor_mat, method = "color", type = "lower",
           tl.cex = 0.5, tl.col = "black", cl.cex = 0.7,
           title = sprintf("Pearson r (%s, n=%s)", cc$label, scales::comma(n_sub)),
           mar = c(0, 0, 2, 0))

  upper_tri <- cor_mat[upper.tri(cor_mat)]

  print(
    ggplot(data.frame(r = upper_tri), aes(x = r)) +
      geom_histogram(binwidth = 0.01, fill = "steelblue", color = "white", alpha = 0.8) +
      geom_vline(xintercept = median(upper_tri, na.rm = TRUE),
                 color = "red", linetype = "dashed") +
      annotate("text", x = median(upper_tri, na.rm = TRUE), y = Inf,
               label = sprintf("median = %.3f", median(upper_tri, na.rm = TRUE)),
               hjust = -0.1, vjust = 1.5, size = 4) +
      labs(title = sprintf("Pairwise r distribution (%s)", cc$label),
           x = "Pearson r", y = "Count") +
      theme_bw(base_size = 12)
  )
  cat("\n\n")
}
```

# Comparing with published data (BACS) {.tabset}

We compare sites at each cutoff against the **BACS** dataset (PMID 39349603,
mRNA sites from Supplementary Table 8 / sheet 6, excluding ncRNA). Sites are
matched by chromosome, position, and strand.

```{r bacs-load}
path_BACS <- "/project2/xinhe/xsun/pumseq/data_others/pU_PMID39349603.xlsx"
rrna_BACS <- readxl::read_excel(path_BACS, sheet = 6, skip = 2)
bacs_mRNA <- rrna_BACS %>%
  filter(feature != "ncRNA") %>%
  mutate(chr_num = sub("^chr", "", chr),
         pos_num = as.integer(pos),
         psi_level = as.numeric(`Ψ level`))

n_bacs <- nrow(bacs_mRNA)
cat("BACS mRNA sites:", n_bacs, "\n")
```

```{r bacs-cutoffs, results='asis', fig.width=8, fig.height=6}
for (cc in cutoff_combos) {
  idx <- which(global_sites$fdr < cc$fdr &
               global_sites$rate_diff > cc$diff &
               global_sites$depth >= cc$depth)
  n_sub <- length(idx)

  cat(sprintf("\n\n## %s\n\n", cc$label))

  sub <- global_sites[idx, ]
  ov <- sub %>%
    mutate(chr_num = ref) %>%
    inner_join(bacs_mRNA, by = c("chr_num", "pos" = "pos_num", "strand"))

  cat(sprintf("Our sites: **%s** | BACS sites: **%s** | Overlap: **%d** (%.1f%% of ours, %.1f%% of BACS)\n\n",
      scales::comma(n_sub), scales::comma(n_bacs), nrow(ov),
      nrow(ov)/n_sub*100, nrow(ov)/n_bacs*100))

  if (nrow(ov) >= 3) {
    r_val <- cor(ov$ratio, ov$psi_level, method = "pearson", use = "complete.obs")

    print(
      ggplot(ov, aes(x = ratio, y = psi_level)) +
        geom_point(alpha = 0.4, size = 1.5, color = "#2c7bb6") +
        geom_smooth(method = "lm", se = TRUE, color = "grey40", linetype = "dashed") +
        annotate("text", x = -Inf, y = Inf,
                 label = sprintf("r = %.3f\nn = %d", r_val, nrow(ov)),
                 hjust = -0.1, vjust = 1.3, size = 4) +
        labs(title = sprintf("vs BACS (%s)", cc$label),
             x = "Our %pU (mega-pooled)", y = "BACS Ψ level") +
        theme_bw(base_size = 12)
    )
  }
  cat("\n\n")
}
```

## Summary table

```{r bacs-summary-table}
bacs_summary <- do.call(rbind, lapply(cutoff_combos, function(cc) {
  idx <- which(global_sites$fdr < cc$fdr &
               global_sites$rate_diff > cc$diff &
               global_sites$depth >= cc$depth)
  sub <- global_sites[idx, ]
  ov <- sub %>% mutate(chr_num = ref) %>%
    inner_join(bacs_mRNA, by = c("chr_num", "pos" = "pos_num", "strand"))
  r_val <- if (nrow(ov) >= 3) round(cor(ov$ratio, ov$psi_level, method="pearson", use="complete.obs"), 3) else NA
  data.frame(
    Cutoff = cc$label,
    Sites = length(idx),
    BACS_overlap = nrow(ov),
    pct_ours_in_BACS = round(nrow(ov)/length(idx)*100, 1),
    pct_BACS_in_ours = round(nrow(ov)/n_bacs*100, 1),
    Pearson_r = r_val
  )
}))

knitr::kable(bacs_summary,
             col.names = c("Cutoff", "# sites", "# overlap", "% ours in BACS",
                           "% BACS in ours", "Pearson r"),
             caption = "BACS overlap at different cutoffs")
```











# Motif analysis around pU sites

We extract ±10 bp flanking sequences around each global pU site from the
reference genome, then compute position-specific nucleotide frequencies. The
pU site (position 0) is always T on the sense strand (T→Ψ conversion). For
sites on the minus strand, the reverse complement is taken so position 0 is
always T.

Known mRNA pseudouridine motifs from the literature:

- **TΨC** / **GUΨC** — the most commonly reported consensus, associated with
  TRUB1-dependent mRNA pseudouridylation
- **UΨAN** — enriched in some studies
- PUS7 targets **UGΨAR**

```{r motif-load}
library(Biostrings)
library(reshape2)

genome_fasta <- "/project/xinhe/xsun/data/genome_reference/Reference/Human/ensembl.v112/fasta/dna/Homo_sapiens.GRCh38.dna.primary_assembly.fa"
bed_file <- file.path(OUT_DIR, "joint_mRNA.pseusite.bed")
flank <- 10

genome <- readDNAStringSet(genome_fasta)
names(genome) <- sub(" .*", "", names(genome))

bed <- fread(bed_file, header = FALSE, sep = "\t",
             col.names = c("chr", "start", "end", "name", "score", "strand"))
bed <- bed[baseline_rows, ]
```

```{r motif-extract}
seqs <- character(nrow(bed))
valid <- logical(nrow(bed))

for (i in seq_len(nrow(bed))) {
  chr    <- bed$chr[i]
  center <- bed$end[i]
  st     <- bed$strand[i]
  if (!chr %in% names(genome)) next
  chr_len <- length(genome[[chr]])
  left  <- center - flank
  right <- center + flank
  if (left < 1 || right > chr_len) next
  s <- subseq(genome[[chr]], start = left, end = right)
  if (st == "-") s <- reverseComplement(s)
  seqs[i] <- as.character(s)
  valid[i] <- TRUE
}

valid_indices <- which(valid)
seqs <- seqs[valid]
cat("Valid sequences:", length(seqs), "out of", nrow(bed), "\n")

seq_width <- 2 * flank + 1
positions <- (-flank):flank
base_order <- c("A", "C", "G", "T")

dna_seqs <- DNAStringSet(seqs)
center_idx <- flank + 1
motif_4mer_char <- as.character(subseq(dna_seqs, start = center_idx - 1, end = center_idx + 2))
motif_9mer_char <- as.character(subseq(dna_seqs, start = center_idx - 2, end = center_idx + 6))

classify_pus <- function(mer9) {
  b <- strsplit(mer9, "")[[1]]
  # b[1]=-2, b[2]=-1, b[3]=0(Ψ), b[4]=+1, b[5]=+2, b[6]=+3, b[7]=+4, b[8]=+5, b[9]=+6
  if (b[1] == "T" && b[2] == "G" && b[4] == "A" && b[5] %in% c("A", "G")) return("PUS7")
  if (b[1] == "G" && b[2] == "T" && b[4] == "C" && b[6] == "A" && b[9] == "C") return("TRUB1")
  return("Other")
}
classes <- sapply(motif_9mer_char, classify_pus, USE.NAMES = FALSE)
```

## Nucleotide frequency at different cutoffs {.tabset}

For each cutoff combination, we compute position-specific nucleotide
frequencies around pU sites and show a heatmap and stacked barplot.

```{r motif-cutoffs, results='asis', fig.width=12, fig.height=9}
for (cc in cutoff_combos) {
  site_idx <- which(global_sites$fdr < cc$fdr &
                    global_sites$rate_diff > cc$diff &
                    global_sites$depth >= cc$depth)
  use_idx <- intersect(site_idx, valid_indices)
  seq_pos <- match(use_idx, valid_indices)
  sub_seqs <- seqs[seq_pos]
  n_sub <- length(sub_seqs)

  cat(sprintf("\n\n### %s\n\n", cc$label))
  cat(sprintf("**%s sites**\n\n", scales::comma(n_sub)))

  if (n_sub < 10) { cat("Too few sites.\n\n"); next }

  freq_mat <- matrix(0, nrow = 4, ncol = seq_width,
                     dimnames = list(base_order, positions))
  for (s in sub_seqs) {
    bases <- strsplit(s, "")[[1]]
    for (j in seq_along(bases)) {
      b <- toupper(bases[j])
      if (b %in% base_order) freq_mat[b, j] <- freq_mat[b, j] + 1
    }
  }
  freq_pct <- sweep(freq_mat, 2, colSums(freq_mat), "/") * 100

  plot_df <- melt(freq_pct)
  colnames(plot_df) <- c("Base", "Position", "Frequency")
  plot_df$Position <- as.integer(as.character(plot_df$Position))
  plot_df$Base <- factor(plot_df$Base, levels = base_order)

  print(
    ggplot(plot_df, aes(x = Position, y = Base, fill = Frequency)) +
      geom_tile(color = "white", linewidth = 0.3) +
      geom_text(aes(label = sprintf("%.0f", Frequency)), size = 2.5) +
      scale_fill_gradient2(low = "white", mid = "#ffffbf", high = "#d73027",
                           midpoint = 25, name = "% frequency") +
      scale_x_continuous(breaks = positions) +
      geom_vline(xintercept = 0, color = "black", linewidth = 0.8, linetype = "dashed") +
      labs(title = sprintf("Nucleotide frequency (%s, n = %s)", cc$label, scales::comma(n_sub)),
           subtitle = sprintf("Position 0 = pU site (T→Ψ), ±%d bp flanking", flank),
           x = "Position relative to pU site", y = NULL) +
      theme_bw(base_size = 11) +
      theme(panel.grid = element_blank())
  )
  cat("\n\n")

  print(
    ggplot(plot_df, aes(x = factor(Position, levels = positions), y = Frequency, fill = Base)) +
      geom_col(position = "stack") +
      scale_fill_manual(values = c("A" = "#4DAF4A", "C" = "#377EB8",
                                    "G" = "#FF7F00", "T" = "#E41A1C")) +
      geom_vline(xintercept = which(positions == 0), color = "black",
                 linewidth = 0.8, linetype = "dashed") +
      labs(title = sprintf("Nucleotide composition (%s, n = %s)", cc$label, scales::comma(n_sub)),
           subtitle = sprintf("Position 0 = pU site (T→Ψ), ±%d bp flanking", flank),
           x = "Position relative to pU site", y = "% frequency") +
      theme_bw(base_size = 11) +
      theme(axis.text.x = element_text(size = 7))
  )
  cat("\n\n")
}
```

## PUS enzyme classification {.tabset}

We classify each pU site by its likely pseudouridine synthase based on the
surrounding genomic DNA motif:

| Enzyme | RNA motif | Genomic DNA motif | Positions checked |
|--------|-----------|-------------------|-------------------|
| **TRUB1** | GUΨCNANNC | GTTCNANNC at pos -2 to +6 | -2=G, -1=T, +1=C, +3=A, +6=C |
| **PUS7** | UGUΨAG | TGTAG at pos -2 to +2 | -2=T, -1=G, +1=A, +2=A/G |
| **PUS1** | Structure-driven | No reliable sequence motif | — |

Sites are classified hierarchically: PUS7 first (most specific motif), then
TRUB1 (full GUΨCNANNC), then "Other". A site matching both is assigned to PUS7.

```{r motif-classify, results='asis', fig.width=10, fig.height=8}
make_class_heatmap <- function(class_seqs, class_name) {
  n <- length(class_seqs)
  if (n == 0) return(NULL)
  fm <- matrix(0, nrow = 4, ncol = seq_width,
               dimnames = list(base_order, positions))
  for (s in class_seqs) {
    bases <- strsplit(s, "")[[1]]
    for (j in seq_along(bases)) {
      b <- toupper(bases[j])
      if (b %in% base_order) fm[b, j] <- fm[b, j] + 1
    }
  }
  fp <- sweep(fm, 2, colSums(fm), "/") * 100
  pd <- melt(fp)
  colnames(pd) <- c("Base", "Position", "Frequency")
  pd$Position <- as.integer(as.character(pd$Position))
  pd$Base <- factor(pd$Base, levels = base_order)
  ggplot(pd, aes(x = Position, y = Base, fill = Frequency)) +
    geom_tile(color = "white", linewidth = 0.3) +
    geom_text(aes(label = sprintf("%.0f", Frequency)), size = 2) +
    scale_fill_gradient2(low = "white", mid = "#ffffbf", high = "#d73027",
                         midpoint = 25, name = "% freq") +
    scale_x_continuous(breaks = positions) +
    geom_vline(xintercept = 0, color = "black", linewidth = 0.8, linetype = "dashed") +
    labs(title = sprintf("%s sites (n = %d)", class_name, n),
         x = "Position relative to pU site", y = NULL) +
    theme_bw(base_size = 9) +
    theme(panel.grid = element_blank())
}

for (cc in cutoff_combos) {
  site_idx <- which(global_sites$fdr < cc$fdr &
                    global_sites$rate_diff > cc$diff &
                    global_sites$depth >= cc$depth)
  use_idx <- intersect(site_idx, valid_indices)
  seq_pos <- match(use_idx, valid_indices)
  sub_seqs <- seqs[seq_pos]
  sub_9mer <- motif_9mer_char[seq_pos]
  sub_classes <- sapply(sub_9mer, classify_pus, USE.NAMES = FALSE)
  n_sub <- length(sub_seqs)

  cat(sprintf("\n\n### %s\n\n", cc$label))
  cat(sprintf("**%s sites**\n\n", scales::comma(n_sub)))

  class_tab <- table(sub_classes)
  class_df <- data.frame(
    Enzyme = names(class_tab),
    Count = as.integer(class_tab),
    Pct = round(as.integer(class_tab) / n_sub * 100, 1)
  )
  class_df <- class_df[order(-class_df$Count), ]

  print(knitr::kable(class_df, row.names = FALSE,
               col.names = c("Enzyme class", "# sites", "%"),
               caption = sprintf("PUS classification (%s)", cc$label)))
  cat("\n\n")

  class_df$label <- sprintf("%s\n%d (%.0f%%)", class_df$Enzyme, class_df$Count, class_df$Pct)
  print(
    ggplot(class_df, aes(x = "", y = Count, fill = Enzyme)) +
      geom_col(width = 1, color = "white") +
      coord_polar(theta = "y") +
      geom_text(aes(label = label),
                position = position_stack(vjust = 0.5), size = 4) +
      scale_fill_manual(values = c("TRUB1" = "#377EB8", "PUS7" = "#FF7F00",
                                    "Other" = "#999999")) +
      labs(title = sprintf("PUS classification (%s, n = %s)", cc$label, scales::comma(n_sub)),
           fill = NULL) +
      theme_void(base_size = 12)
  )
  cat("\n\n")

  hm_plots <- Filter(Negate(is.null), list(
    make_class_heatmap(sub_seqs[sub_classes == "TRUB1"], "TRUB1"),
    make_class_heatmap(sub_seqs[sub_classes == "PUS7"], "PUS7"),
    make_class_heatmap(sub_seqs[sub_classes == "Other"], "Other")
  ))
  if (length(hm_plots) > 0) print(wrap_plots(hm_plots, ncol = 1))
  cat("\n\n")
}
```

## Motif enrichment vs %pU strength {.tabset}

We split sites into groups by their pooled %pU (mega-pooled C/(C+T)) and
compare the proportion of sites matching known PUS motifs (TRUB1 or PUS7) in
each group. If strong sites are enriched for known motifs, this supports the
biological validity of both the site calls and the motif classification.

```{r motif-vs-strength, results='asis', fig.width=10, fig.height=6}
for (cc in cutoff_combos) {
  site_idx <- which(global_sites$fdr < cc$fdr &
                    global_sites$rate_diff > cc$diff &
                    global_sites$depth >= cc$depth)
  use_idx <- intersect(site_idx, valid_indices)
  seq_pos <- match(use_idx, valid_indices)
  sub_9mer <- motif_9mer_char[seq_pos]
  sub_classes <- sapply(sub_9mer, classify_pus, USE.NAMES = FALSE)
  n_sub <- length(use_idx)

  cat(sprintf("\n\n### %s\n\n", cc$label))
  cat(sprintf("**%s sites**\n\n", scales::comma(n_sub)))

  if (n_sub < 20) { cat("Too few sites.\n\n"); next }

  sub_global <- global_sites[use_idx, ]
  strength_df <- data.frame(
    ratio = sub_global$ratio,
    class = sub_classes,
    is_known = sub_classes %in% c("TRUB1", "PUS7")
  )

  strength_df$quartile <- cut(strength_df$ratio,
                               breaks = quantile(strength_df$ratio, probs = c(0, 0.25, 0.5, 0.75, 1)),
                               include.lowest = TRUE,
                               labels = c("Q1 (weakest)", "Q2", "Q3", "Q4 (strongest)"))

  q_summary <- strength_df %>%
    group_by(quartile, class) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(quartile) %>%
    mutate(pct = round(n / sum(n) * 100, 1))

  print(
    ggplot(q_summary, aes(x = quartile, y = pct, fill = class)) +
      geom_col(position = "stack") +
      geom_text(aes(label = sprintf("%.0f%%", pct)),
                position = position_stack(vjust = 0.5), size = 3) +
      scale_fill_manual(values = c("TRUB1" = "#377EB8", "PUS7" = "#FF7F00",
                                    "Other" = "#999999")) +
      labs(title = sprintf("PUS motif by %%pU quartile (%s)", cc$label),
           x = "%pU quartile", y = "% of sites", fill = "Enzyme class") +
      theme_bw(base_size = 12)
  )
  cat("\n\n")

  q_known <- strength_df %>%
    group_by(quartile) %>%
    summarise(
      n_sites = n(),
      n_TRUB1 = sum(class == "TRUB1"),
      pct_TRUB1 = round(sum(class == "TRUB1") / n() * 100, 1),
      n_PUS7 = sum(class == "PUS7"),
      pct_PUS7 = round(sum(class == "PUS7") / n() * 100, 1),
      n_known = sum(is_known),
      pct_known = round(sum(is_known) / n() * 100, 1),
      median_pU = round(median(ratio), 3),
      .groups = "drop"
    )

  print(knitr::kable(q_known,
               col.names = c("Quartile", "# sites", "# TRUB1", "% TRUB1",
                             "# PUS7", "% PUS7", "# known motif", "% known motif",
                             "Median %pU"),
               caption = sprintf("PUS motif enrichment (%s)", cc$label)))
  cat("\n\n")

  cat("\n\n")
}
```

# Motif analysis of BACS published sites

We perform the same motif analysis on the **BACS** published mRNA pU sites
(PMID 39349603) as a reference comparison.

```{r bacs-motif-extract}
bacs_bed <- bacs_mRNA %>%
  mutate(chr_clean = sub("^chr", "", chr)) %>%
  filter(chr_clean %in% names(genome))

bacs_seqs <- character(nrow(bacs_bed))
bacs_valid <- logical(nrow(bacs_bed))

for (i in seq_len(nrow(bacs_bed))) {
  chr_name <- bacs_bed$chr_clean[i]
  center   <- as.integer(bacs_bed$pos[i])
  st       <- bacs_bed$strand[i]
  chr_len  <- length(genome[[chr_name]])
  left     <- center - flank
  right    <- center + flank
  if (left < 1 || right > chr_len) next
  s <- subseq(genome[[chr_name]], start = left, end = right)
  if (st == "-") s <- reverseComplement(s)
  bacs_seqs[i] <- as.character(s)
  bacs_valid[i] <- TRUE
}

bacs_seqs <- bacs_seqs[bacs_valid]
cat("BACS valid sequences:", length(bacs_seqs), "out of", nrow(bacs_bed), "\n")

bacs_dna <- DNAStringSet(bacs_seqs)
bacs_4mer <- as.character(subseq(bacs_dna, start = center_idx - 1, end = center_idx + 2))
bacs_9mer <- as.character(subseq(bacs_dna, start = center_idx - 2, end = center_idx + 6))
bacs_classes <- sapply(bacs_9mer, classify_pus, USE.NAMES = FALSE)
```

## Nucleotide frequency heatmap (BACS)

```{r bacs-motif-heatmap, fig.width=12, fig.height=4}
bacs_freq_mat <- matrix(0, nrow = 4, ncol = seq_width,
                        dimnames = list(base_order, positions))
for (s in bacs_seqs) {
  bases <- strsplit(s, "")[[1]]
  for (j in seq_along(bases)) {
    b <- toupper(bases[j])
    if (b %in% base_order) bacs_freq_mat[b, j] <- bacs_freq_mat[b, j] + 1
  }
}
bacs_freq_pct <- sweep(bacs_freq_mat, 2, colSums(bacs_freq_mat), "/") * 100

bacs_plot_df <- melt(bacs_freq_pct)
colnames(bacs_plot_df) <- c("Base", "Position", "Frequency")
bacs_plot_df$Position <- as.integer(as.character(bacs_plot_df$Position))
bacs_plot_df$Base <- factor(bacs_plot_df$Base, levels = base_order)

ggplot(bacs_plot_df, aes(x = Position, y = Base, fill = Frequency)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%.0f", Frequency)), size = 2.5) +
  scale_fill_gradient2(low = "white", mid = "#ffffbf", high = "#d73027",
                       midpoint = 25, name = "% frequency") +
  scale_x_continuous(breaks = positions) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.8, linetype = "dashed") +
  labs(title = sprintf("Nucleotide frequency — BACS sites (n = %d)", length(bacs_seqs)),
       subtitle = sprintf("Position 0 = pU site (T→Ψ), ±%d bp flanking", flank),
       x = "Position relative to pU site", y = NULL) +
  theme_bw(base_size = 11) +
  theme(panel.grid = element_blank())
```

## Stacked nucleotide composition (BACS)

```{r bacs-motif-barplot, fig.width=12, fig.height=5}
ggplot(bacs_plot_df, aes(x = factor(Position, levels = positions), y = Frequency, fill = Base)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("A" = "#4DAF4A", "C" = "#377EB8",
                                "G" = "#FF7F00", "T" = "#E41A1C")) +
  geom_vline(xintercept = which(positions == 0), color = "black",
             linewidth = 0.8, linetype = "dashed") +
  labs(title = sprintf("Nucleotide composition — BACS sites (n = %d)", length(bacs_seqs)),
       subtitle = sprintf("Position 0 = pU site (T→Ψ), ±%d bp flanking", flank),
       x = "Position relative to pU site", y = "% frequency") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(size = 7))
```

## PUS enzyme classification (BACS)

```{r bacs-motif-classify, fig.width=10, fig.height=8}
bacs_class_tab <- table(bacs_classes)
bacs_class_df <- data.frame(
  Enzyme = names(bacs_class_tab),
  Count = as.integer(bacs_class_tab),
  Pct = round(as.integer(bacs_class_tab) / length(bacs_seqs) * 100, 1)
)
bacs_class_df <- bacs_class_df[order(-bacs_class_df$Count), ]

knitr::kable(bacs_class_df, row.names = FALSE,
             col.names = c("Enzyme class", "# sites", "%"),
             caption = "PUS classification — BACS sites")
```

```{r bacs-motif-classify-pie, fig.width=7, fig.height=5}
bacs_class_df$label <- sprintf("%s\n%d (%.0f%%)", bacs_class_df$Enzyme,
                                bacs_class_df$Count, bacs_class_df$Pct)

ggplot(bacs_class_df, aes(x = "", y = Count, fill = Enzyme)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5), size = 4) +
  scale_fill_manual(values = c("TRUB1" = "#377EB8", "PUS7" = "#FF7F00",
                                "Other" = "#999999")) +
  labs(title = sprintf("PUS classification — BACS sites (n = %d)", length(bacs_seqs)),
       fill = NULL) +
  theme_void(base_size = 12)
```

## Motif heatmaps by enzyme class (BACS)

```{r bacs-motif-by-class, fig.width=12, fig.height=12}
bacs_hm_plots <- Filter(Negate(is.null), list(
  make_class_heatmap(bacs_seqs[bacs_classes == "TRUB1"], "TRUB1 (BACS)"),
  make_class_heatmap(bacs_seqs[bacs_classes == "PUS7"], "PUS7 (BACS)"),
  make_class_heatmap(bacs_seqs[bacs_classes == "Other"], "Other (BACS)")
))
if (length(bacs_hm_plots) > 0) wrap_plots(bacs_hm_plots, ncol = 1)
```

## Motif enrichment vs Ψ level (BACS)

We split BACS sites into quartiles by their published Ψ level and compare the
proportion matching known PUS motifs (TRUB1 or PUS7) in each group.

```{r bacs-motif-vs-strength, fig.width=10, fig.height=6}
bacs_strength_df <- data.frame(
  psi_level = bacs_mRNA$psi_level[bacs_valid],
  class = bacs_classes,
  is_known = bacs_classes %in% c("TRUB1", "PUS7")
)

bacs_strength_df$quartile <- cut(bacs_strength_df$psi_level,
                                  breaks = quantile(bacs_strength_df$psi_level,
                                                    probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE),
                                  include.lowest = TRUE,
                                  labels = c("Q1 (weakest)", "Q2", "Q3", "Q4 (strongest)"))

bacs_q_summary <- bacs_strength_df %>%
  filter(!is.na(quartile)) %>%
  group_by(quartile, class) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(quartile) %>%
  mutate(pct = round(n / sum(n) * 100, 1))

ggplot(bacs_q_summary, aes(x = quartile, y = pct, fill = class)) +
  geom_col(position = "stack") +
  geom_text(aes(label = sprintf("%.0f%%", pct)),
            position = position_stack(vjust = 0.5), size = 3) +
  scale_fill_manual(values = c("TRUB1" = "#377EB8", "PUS7" = "#FF7F00",
                                "Other" = "#999999")) +
  labs(title = "PUS motif composition by Ψ level quartile (BACS)",
       x = "Ψ level quartile", y = "% of sites", fill = "Enzyme class") +
  theme_bw(base_size = 12)
```

```{r bacs-motif-vs-strength-table}
bacs_q_known <- bacs_strength_df %>%
  filter(!is.na(quartile)) %>%
  group_by(quartile) %>%
  summarise(
    n_sites = n(),
    n_TRUB1 = sum(class == "TRUB1"),
    pct_TRUB1 = round(sum(class == "TRUB1") / n() * 100, 1),
    n_PUS7 = sum(class == "PUS7"),
    pct_PUS7 = round(sum(class == "PUS7") / n() * 100, 1),
    n_known = sum(is_known),
    pct_known = round(sum(is_known) / n() * 100, 1),
    median_psi = round(median(psi_level, na.rm = TRUE), 3),
    .groups = "drop"
  )

knitr::kable(bacs_q_known,
             col.names = c("Quartile", "# sites", "# TRUB1", "% TRUB1",
                           "# PUS7", "% PUS7", "# known motif", "% known motif",
                           "Median Ψ level"),
             caption = "PUS motif enrichment by Ψ level quartile (BACS)")
```

---

# Genomic feature distribution

We annotate each pU site against the Ensembl v112 GTF with the following
priority (highest wins): CDS > 3'UTR > 5'UTR > ncRNA exon > Intron >
Intergenic. Pre-computed by `annotate_sites_feature.R`.

```{r feature-load}
anno_sites <- fread(
  file.path(OUT_DIR, "joint_mRNA.global_sites_annotated.txt.gz"),
  header = TRUE, sep = "\t", colClasses = list(character = c("ref", "strand")))

anno_bacs <- fread(
  file.path(OUT_DIR, "bacs_mrna_annotated.txt.gz"),
  header = TRUE, sep = "\t")

feat_levels <- c("5UTR", "CDS", "3UTR", "ncRNA", "Intron", "Intergenic")
feat_colors <- c("5UTR"       = "#4DAF4A",
                 "CDS"        = "#377EB8",
                 "3UTR"       = "#FF7F00",
                 "ncRNA"      = "#984EA3",
                 "Intron"     = "#A65628",
                 "Intergenic" = "#999999")
```

## Feature distribution at different cutoffs {.tabset}

```{r feature-cutoffs, results='asis', fig.width=9, fig.height=5}
bacs_feat_df <- data.frame(
  feature = factor(anno_bacs$feature_anno, levels = feat_levels)
) %>%
  count(feature) %>%
  mutate(pct = round(n / sum(n) * 100, 1), dataset = "BACS")

for (cc in cutoff_combos) {
  idx <- which(anno_sites$fdr < cc$fdr &
               anno_sites$rate_diff > cc$diff &
               anno_sites$depth >= cc$depth)
  n_sub <- length(idx)

  cat(sprintf("\n\n### %s\n\n", cc$label))
  cat(sprintf("**%s sites**\n\n", scales::comma(n_sub)))

  if (n_sub < 10) { cat("Too few sites.\n\n"); next }

  our_feat_df <- data.frame(
    feature = factor(anno_sites$feature[idx], levels = feat_levels)
  ) %>%
    count(feature) %>%
    mutate(pct = round(n / sum(n) * 100, 1),
           dataset = cc$label)

  plot_df <- bind_rows(our_feat_df, bacs_feat_df)
  plot_df$dataset <- factor(plot_df$dataset,
                             levels = c(cc$label, "BACS"))
  plot_df$feature <- factor(plot_df$feature, levels = feat_levels)

  print(
    ggplot(plot_df, aes(x = dataset, y = pct, fill = feature)) +
      geom_col(position = "stack", width = 0.6) +
      geom_text(aes(label = sprintf("%.0f%%", pct)),
                position = position_stack(vjust = 0.5), size = 3.5) +
      scale_fill_manual(values = feat_colors, name = "Feature") +
      labs(title = sprintf("Genomic feature distribution (%s)", cc$label),
           x = NULL, y = "% of sites") +
      theme_bw(base_size = 12) +
      theme(legend.position = "right")
  )
  cat("\n\n")
}
```

## BACS overlap restricted to UTR/CDS sites

BACS sites are exclusively in 5'UTR/CDS/3'UTR (no ncRNA/Intron/Intergenic
calls). Here, for each of the 6 cutoff groups, we restrict our sites to
those annotated 5'UTR/CDS/3'UTR and report the % that overlap a BACS site —
an apples-to-apples comparison with the BACS dataset.

```{r utr-cds-bacs-overlap, fig.width=8, fig.height=5}
bacs_keys <- anno_bacs %>% select(chr, pos, strand)

utr_cds_table <- do.call(rbind, lapply(cutoff_combos, function(cc) {
  idx <- which(anno_sites$fdr < cc$fdr &
               anno_sites$rate_diff > cc$diff &
               anno_sites$depth >= cc$depth)
  sub_utr_cds <- anno_sites[idx, ][anno_sites[idx, ]$feature %in% c("5UTR", "CDS", "3UTR"), ]

  ov_utr_cds <- sub_utr_cds %>%
    inner_join(bacs_keys, by = c("ref" = "chr", "pos", "strand"))

  data.frame(
    Cutoff      = cc$label,
    n_utr_cds   = nrow(sub_utr_cds),
    n_overlap   = nrow(ov_utr_cds),
    pct_overlap = round(nrow(ov_utr_cds) / nrow(sub_utr_cds) * 100, 1)
  )
}))

knitr::kable(utr_cds_table,
             col.names = c("Cutoff", "# UTR/CDS sites", "# BACS overlap", "% overlap"),
             caption = "% of UTR/CDS sites overlapping BACS, by cutoff")

utr_cds_table$Cutoff <- factor(utr_cds_table$Cutoff, levels = rev(utr_cds_table$Cutoff))
ggplot(utr_cds_table, aes(x = Cutoff, y = pct_overlap)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", pct_overlap)), hjust = -0.1, size = 4) +
  coord_flip(ylim = c(0, max(utr_cds_table$pct_overlap) * 1.15)) +
  labs(title = "% of UTR/CDS sites overlapping BACS, by cutoff",
       x = NULL, y = "% overlap with BACS") +
  theme_bw(base_size = 12)
```


