---
title: "QC — Discovery/replication validation of joint-pooled pU site calling"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r setup, warning=FALSE, message=FALSE}
library(data.table)
library(ggplot2)
library(patchwork)

knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", fig.width = 9, fig.height = 6
)

JOINT_DIR <- "/project/xinhe/xsun/pumseq/pumseq_processed/joint"
N_SPLITS <- 3
FDR_THRESH <- 0.01
RATEDIFF_THRESH <- 0.05
DEPTH_THRESH <- 20
```

# Introduction

We split the cell lines into a small **discovery** subset and a larger **replication** subset,
call sites by pooling reads within the discovery subset only (same
"joint" pooling + chi-squared approach used throughout this project),
then check — in the independent replication subset — (1) what fraction
of discovery-called sites are also significant, and (2) how well %pU
at those sites correlates between discovery and replication.

<!-- This also addresses two concerns raised before running it: (1) union -->
<!-- site calling has substantial per-cell-line noise in individual -->
<!-- `treated - input` values (see the QTL phenotype work), which is -->
<!-- informal evidence that not every "called" site cleanly replicates -->
<!-- across samples; (2) with only ~20% of cell lines in the discovery pool, -->
<!-- per-site read depth in that pool will be considerably lower than the -->
<!-- full 51-cell-line joint pool (which itself already showed many -->
<!-- low-depth sites) — so discovery-pool depth is reported explicitly -->
<!-- alongside the replication numbers, to separate "site wasn't real" from -->
<!-- "site wasn't callable with this few samples." -->

**Design**: 51 cell lines split into discovery (~20%, 10 cell lines)
and replication (~80%, 41 cell lines), independently for
`r N_SPLITS` random splits (different cell lines in the discovery pool
each time), to see how much the result depends on which specific cell
lines end up in discovery. Site calling in both pools uses the same
thresholds as the standard joint pipeline (FDR < `r FDR_THRESH`,
rate_diff > `r RATEDIFF_THRESH`, depth >= `r DEPTH_THRESH`).

```{r load-data}
load_split <- function(i) {
  disc_dir <- file.path(JOINT_DIR, paste0("discovery_split", i, "_mRNA"))
  repl_dir <- file.path(JOINT_DIR, paste0("replication_split", i, "_mRNA"))

  discovery_sites <- fread(file.path(disc_dir, "joint_mRNA.global_sites.txt.gz"))
  discovery_sites <- discovery_sites[fdr < FDR_THRESH]   # apply FDR cutoff to discovery calls too
  replication_full <- fread(file.path(repl_dir, "joint_mRNA.pseusite.txt.gz"))

  matched <- merge(
    discovery_sites, replication_full,
    by = c("ref", "pos", "strand"), suffixes = c(".disc", ".repl")
  )
  matched[, split := i]
  matched
}

all_splits <- rbindlist(lapply(seq_len(N_SPLITS), load_split))
#cat("Total discovery-called sites (pooled across splits):", nrow(all_splits), "\n")
```

# Discovery-pool depth

Read depth at discovery-called sites, using only the discovery pool's
own reads (10 cell lines) 

<!-- — the concern is that this may be -->
<!-- considerably shallower than the full-cohort joint pool. -->

```{r discovery-depth}
depth_summary <- all_splits[, .(
  n_sites = .N,
  median_depth = as.double(median(depth.disc)),
  q25_depth = as.double(quantile(depth.disc, 0.25)),
  q75_depth = as.double(quantile(depth.disc, 0.75))
), by = split]
knitr::kable(depth_summary, digits = 1, caption = "Discovery-pool read depth at discovery-called sites, by split")
```

```{r depth-hist, fig.height=4.5}
ggplot(all_splits, aes(x = depth.disc)) +
  geom_histogram(bins = 60, fill = "steelblue") +
  facet_wrap(~ split, ncol = 3, labeller = label_both) +
  labs(x = "discovery-pool depth at called sites", y = "count",
       title = "Discovery-pool depth distribution, by split") +
  theme_bw(base_size = 12)
```

# Replication rate

For each discovery-called site, checking whether it's *also* significant
under the same thresholds (FDR < `r FDR_THRESH`, rate_diff >
`r RATEDIFF_THRESH`, depth >= `r DEPTH_THRESH`) in the independent
replication pool.

```{r replication-rate}
all_splits[, replicated := (fdr.repl < FDR_THRESH) & (rate_diff.repl > RATEDIFF_THRESH) & (depth.repl >= DEPTH_THRESH)]
all_splits[, tested_in_repl := depth.repl >= DEPTH_THRESH]

repl_summary <- all_splits[, .(
  n_discovery_sites = .N,
  n_testable_in_replication = sum(tested_in_repl),
  n_replicated = sum(replicated, na.rm = TRUE),
  replication_rate = round(sum(replicated, na.rm = TRUE) / .N, 3)
), by = split]
knitr::kable(repl_summary, caption = "Replication rate by split")
```

Column definitions:

- `n_discovery_sites`: number of sites called in the discovery pool.
- `n_testable_in_replication`: of those, the number that have adequate
  depth (>= `r DEPTH_THRESH`) in the *replication* pool — i.e. sites
  with enough replication-pool reads to actually be evaluated for
  significance, whether or not they replicated.
- `n_replicated`: of the discovery sites, the number that are also
  significant in the replication pool (FDR < `r FDR_THRESH`, rate_diff >
  `r RATEDIFF_THRESH`, depth >= `r DEPTH_THRESH`).
- `replication_rate`: `n_replicated / n_discovery_sites`.

```{r replication-rate-plot, fig.height=4.5}
ggplot(repl_summary, aes(x = factor(split), y = replication_rate)) +
  geom_col(fill = "steelblue", width = 0.5) +
  geom_text(aes(label = scales::percent(replication_rate)), vjust = -0.5) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(x = "split", y = "replication rate",
       title = "Fraction of discovery-called sites significant in replication") +
  theme_bw(base_size = 13)
```

# %pU correlation between discovery and replication

Comparing `ratio` (treated %pU, from the chi-squared test's pooled
treated counts) at discovery-called sites, discovery pool vs.
replication pool.

```{r correlation-plot, fig.height=5}
cor_by_split <- all_splits[, .(
  pearson_r = round(cor(ratio.disc, ratio.repl, use = "complete.obs"), 3),
  n = .N
), by = split]
knitr::kable(cor_by_split, caption = "%pU correlation (discovery vs. replication), by split")

ggplot(all_splits, aes(x = ratio.disc, y = ratio.repl)) +
  geom_point(size = 0.6, alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, color = "firebrick", linetype = "dashed") +
  facet_wrap(~ split, ncol = 3, labeller = label_both) +
  labs(x = "%pU, discovery pool", y = "%pU, replication pool",
       title = "%pU at discovery-called sites: discovery vs. replication") +
  theme_bw(base_size = 12)
```

<!-- # Notes -->

<!-- - Discovery-pool depth (above) directly addresses the concern that -->
<!--   calling sites from only ~10 pooled cell lines may be depth-limited -->
<!--   compared to the full 51-cell-line joint pool. -->
<!-- - Replication rate and %pU correlation are reported per split (not -->
<!--   just pooled) to show how much the result depends on which specific -->
<!--   cell lines happen to land in the discovery pool. -->
<!-- - `n_testable_in_replication` turns out to be nearly identical to -->
<!--   `n_discovery_sites` (see the note under the replication-rate table): -->
<!--   the larger replication pool gives essentially every discovery site -->
<!--   adequate depth to be tested. So non-replication here reflects sites -->
<!--   that genuinely fail to reproduce as significant in independent -->
<!--   samples, not sites that were too shallow in replication to evaluate. -->
