---
title: "Quality control using sites called from mRNA"
author: "XSun"
date: "2026-06-06"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

# Introduction

We ran the PUMSEQ pipeline on four cell lines: **CL18870**, **CL18909**,
**CL18916**, and **CL19099**.  Each cell line has four treated replicates.

Pseudouridine (pU) sites are identified from mRNA (polyA+) reads using a
ratio/depth threshold strategy.  The pseudouridine ratio per site is defined as

$$\%\psi = \frac{C}{C + T}$$

where *C* counts cytosine calls (the signature of reverse-transcriptase stop at
a pU site) and *T* counts unmodified thymidine calls at that position.

We compare our results against two published datasets:

- **BACS** — PMID 39349603
- **PRAISE** — PMID 36997645

```{r setup, warning=FALSE, message=FALSE}
library(dplyr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(purrr)
library(patchwork)
library(tidyverse)
library(data.table)
library(scales)
library(knitr)
library(gridExtra)
library(DT)
library(grid)

knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center"
)

cutoffs <- expand.grid(min_depth = c(10, 20),
                       min_ratio = c(0.1, 0.2))
```

---

```{r setup-config}
RESULTS_DIR <- "/project2/xinhe/xsun/pumseq/pumseq_results"

CELL_LINES <- list(
  CL18870 = list(
    rep1 = "CL18870_12-187-59",
    rep2 = "CL18870_12-187-60",
    rep3 = "CL18870_12-187-97",
    rep4 = "CL18870_12-187-98"
  ),
  CL18909 = list(
    rep1 = "CL18909_12-187-13",
    rep2 = "CL18909_12-187-14",
    rep3 = "CL18909_13-32-23",
    rep4 = "CL18909_13-32-24"
  ),
  CL18916 = list(
    rep1 = "CL18916_12-187-61",
    rep2 = "CL18916_12-187-62",
    rep3 = "CL18916_12-187-101",
    rep4 = "CL18916_12-187-102"
  ),
  CL19099 = list(
    rep1 = "CL19099_12-187-65",
    rep2 = "CL19099_12-187-66",
    rep3 = "CL19099_12-187-105",
    rep4 = "CL19099_12-187-106"
  )
)
```

---

# Comparing pooled data among cell lines

After per-replicate QC, replicates for each cell line are **pooled**
(`pooled_sites/pooled_mRNA.txt`): the depth and C-count across all four
replicates are summed, and a pooled %pU ratio is computed.  Each site also
records how many individual replicates it was detected in (`n_samples`).

We explore the effect of three filtering dimensions simultaneously:

| Dimension | Values |
|-----------|--------|
| Minimum pooled depth | 10, 20 |
| Minimum pooled %pU ratio | 0.1, 0.2 |
| Minimum number of replicates | 1, 2, 3, 4 |

Stricter cutoffs reduce noise at the cost of fewer sites.  Requiring detection
in multiple replicates is a strong reproducibility filter.

```{r setup-pooled, message=FALSE, warning=FALSE}
CELL_LINES_V <- names(CELL_LINES)
DEPTH_CUTS   <- c(10, 20)
RATIO_CUTS   <- c(0.1, 0.2)
NSAMP_CUTS   <- 1:4

load_pooled_mRNA <- function(cl) {
  f <- file.path(RESULTS_DIR, cl, "pooled_sites", "pooled_mRNA.txt")
  dt <- fread(f, header = TRUE, sep = "\t",
              colClasses = list(character = "ref"))
  dt[, cell_line := cl]
  dt[, site := paste(ref, pos, strand, sep = ":")]
  dt
}

count_sites <- function(dt, min_depth, min_ratio, min_nsamp) {
  sum(dt$depth >= min_depth & dt$ratio >= min_ratio & dt$n_samples >= min_nsamp)
}

build_count_grid <- function(dt) {
  expand.grid(
    min_depth  = DEPTH_CUTS,
    min_ratio  = RATIO_CUTS,
    min_nsamp  = NSAMP_CUTS,
    stringsAsFactors = FALSE
  ) |>
    rowwise() |>
    mutate(n_sites = count_sites(dt, min_depth, min_ratio, min_nsamp)) |>
    ungroup() |>
    mutate(
      depth_label = paste0("depth>=", min_depth),
      ratio_label = paste0("ratio>=", min_ratio),
      nsamp_label = paste0("n_samples>=", min_nsamp)
    )
}

plot_count_heatmap <- function(grid, cl_title) {
  grid |>
    mutate(
      depth_label = factor(depth_label, levels = paste0("depth>=", DEPTH_CUTS)),
      ratio_label = factor(ratio_label, levels = paste0("ratio>=", rev(RATIO_CUTS))),
      nsamp_label = factor(nsamp_label,
                           levels = paste0("n_samples>=", NSAMP_CUTS))
    ) |>
    ggplot(aes(x = depth_label, y = ratio_label, fill = n_sites)) +
    geom_tile(color = "white", linewidth = 0.6) +
    geom_text(aes(label = comma(n_sites)), size = 4, fontface = "bold") +
    scale_fill_gradient(low = "#deebf7", high = "#2171b5",
                        name = "# sites", labels = comma) +
    facet_wrap(~nsamp_label, nrow = 1) +
    labs(title = cl_title,
         x = "Depth cutoff (pooled)", y = "Ratio cutoff (pooled %pU)") +
    theme_bw(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          strip.background = element_rect(fill = "#e8eaf6"),
          strip.text = element_text(face = "bold"))
}

dat_list <- lapply(CELL_LINES_V, load_pooled_mRNA)
names(dat_list) <- CELL_LINES_V

grid_list <- lapply(dat_list, build_count_grid)
names(grid_list) <- CELL_LINES_V

cl_pairs <- combn(CELL_LINES_V, 2, simplify = FALSE)
```

## Per cell-line site counts at different cutoffs

Each heatmap shows the number of mRNA sites surviving a combination of
**pooled depth** (x-axis), **pooled %pU ratio** (y-axis), and **minimum
number of replicates** detecting the site (panels, left to right).
Darker blue = more sites.

---

### CL18870

```{r mrna-cl18870-count-heatmap, fig.width=12, fig.height=3.5}
plot_count_heatmap(grid_list[["CL18870"]], "CL18870 – mRNA pU site counts")
```

---

### CL18909

```{r mrna-cl18909-count-heatmap, fig.width=12, fig.height=3.5}
plot_count_heatmap(grid_list[["CL18909"]], "CL18909 – mRNA pU site counts")
```

---

### CL18916

```{r mrna-cl18916-count-heatmap, fig.width=12, fig.height=3.5}
plot_count_heatmap(grid_list[["CL18916"]], "CL18916 – mRNA pU site counts")
```

---

### CL19099

```{r mrna-cl19099-count-heatmap, fig.width=12, fig.height=3.5}
plot_count_heatmap(grid_list[["CL19099"]], "CL19099 – mRNA pU site counts")
```

---

### %pU Distribution — pooled data {.tabset}

Distribution of the pooled %ψ ratio (C/(C+T)) across all four cell lines.
Each panel shows one cell line.  Red dashed = median; orange dotted = mean;
green long-dash = ratio threshold(s).

```{r mrna-pooled-dist-helper, include=FALSE}
plot_pooled_dist <- function(cl, thr = NULL, min_depth = 20) {
  dt  <- dat_list[[cl]]
  df  <- as.data.frame(dt[depth >= min_depth, .(psi = ratio)])
  df  <- df[!is.na(df$psi), , drop = FALSE]
  if (!is.null(thr)) df <- df[df$psi >= thr, , drop = FALSE]
  if (nrow(df) == 0) return(NULL)

  med_val  <- median(df$psi)
  mean_val <- mean(df$psi)
  n_01     <- sum(df$psi >= 0.1)
  n_02     <- sum(df$psi >= 0.2)

  ann_text <- if (is.null(thr)) {
    sprintf(
      "n = %d\nmedian = %.3f\nmean = %.3f\n>=0.1: %d (%.1f%%)\n>=0.2: %d (%.1f%%)",
      nrow(df), med_val, mean_val,
      n_01, 100 * n_01 / nrow(df),
      n_02, 100 * n_02 / nrow(df)
    )
  } else {
    sprintf("n = %d\nmedian = %.3f\nmean = %.3f", nrow(df), med_val, mean_val)
  }

  vlines <- if (is.null(thr)) c(0.1, 0.2) else if (thr == 0.1) 0.2 else NULL
  label  <- if (is.null(thr)) "all sites" else sprintf("%%pU >= %.1f", thr)

  p <- ggplot(df, aes(x = psi)) +
    geom_histogram(aes(y = after_stat(density)),
                   binwidth = 0.02, fill = "steelblue",
                   color = "white", linewidth = 0.2, alpha = 0.7) +
    geom_density(color = "navy", linewidth = 0.7) +
    geom_vline(xintercept = med_val,  color = "red",
               linetype = "dashed",  linewidth = 0.6) +
    geom_vline(xintercept = mean_val, color = "darkorange",
               linetype = "dotted",  linewidth = 0.6) +
    annotate("text", x = Inf, y = Inf,
             label = ann_text,
             hjust = 1.05, vjust = 1.05, size = 3, family = "mono") +
    scale_x_continuous(limits = c(if (is.null(thr)) 0 else thr, 1),
                       breaks = seq(0, 1, 0.1)) +
    labs(title    = cl,
         subtitle = label,
         x        = "%pU  (C / C+T)",
         y        = "Density") +
    theme_bw(base_size = 10) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  if (!is.null(vlines))
    p <- p + geom_vline(xintercept = vlines, color = "darkgreen",
                        linetype = "longdash", linewidth = 0.5)
  p
}
```

#### %pU Distribution — all sites

```{r mrna-pooled-dist-all, fig.width=14, fig.height=3.5}
plots_all <- lapply(CELL_LINES_V, plot_pooled_dist, thr = NULL)
do.call(grid.arrange, c(Filter(Negate(is.null), plots_all), list(ncol = 4)))
```

#### %pU Distribution — %pU >= 0.1

```{r mrna-pooled-dist-01, fig.width=14, fig.height=3.5}
plots_01 <- lapply(CELL_LINES_V, plot_pooled_dist, thr = 0.1)
do.call(grid.arrange, c(Filter(Negate(is.null), plots_01), list(ncol = 4)))
```

#### %pU Distribution — %pU >= 0.2

```{r mrna-pooled-dist-02, fig.width=14, fig.height=3.5}
plots_02 <- lapply(CELL_LINES_V, plot_pooled_dist, thr = 0.2)
do.call(grid.arrange, c(Filter(Negate(is.null), plots_02), list(ncol = 4)))
```

---

## Cross-cell-line comparison of site counts

We use the **Jaccard index** to quantify how much the detected site sets
overlap between every pair of cell lines.  For each combination of depth,
ratio, and n_samples cutoffs, sites passing all filters in each cell line are
extracted and the pairwise Jaccard is computed:

$$J(A, B) = \frac{|A \cap B|}{|A \cup B|}$$

A high Jaccard index means the two cell lines call largely the same mRNA
positions.  Each cell shows the Jaccard score and, in parentheses, the number
of sites in the intersection.  Rows = minimum n_samples, columns = depth ×
ratio cutoff.

```{r mrna-cross-cl-comparison, fig.width=14, fig.height=11}
jaccard_cl_rows <- do.call(rbind, lapply(cl_pairs, function(p) {
  do.call(rbind, lapply(DEPTH_CUTS, function(d) {
    do.call(rbind, lapply(RATIO_CUTS, function(r) {
      do.call(rbind, lapply(NSAMP_CUTS, function(n) {
        sites1  <- dat_list[[p[1]]][depth >= d & ratio >= r & n_samples >= n, site]
        sites2  <- dat_list[[p[2]]][depth >= d & ratio >= r & n_samples >= n, site]
        n_int   <- length(intersect(sites1, sites2))
        n_uni   <- length(union(sites1, sites2))
        j       <- if (n_uni == 0) NA_real_ else n_int / n_uni
        data.frame(cl1 = p[1], cl2 = p[2],
                   min_depth = d, min_ratio = r, min_nsamp = n,
                   jaccard = round(j, 3), n_intersect = n_int)
      }))
    }))
  }))
}))

jaccard_cl_sym <- jaccard_cl_rows |>
  bind_rows(jaccard_cl_rows |> rename(cl1 = cl2, cl2 = cl1)) |>
  bind_rows(
    expand.grid(cl1 = CELL_LINES_V, cl2 = CELL_LINES_V,
                min_depth = DEPTH_CUTS, min_ratio = RATIO_CUTS,
                min_nsamp = NSAMP_CUTS) |>
      filter(cl1 == cl2) |>
      mutate(jaccard = 1, n_intersect = NA_integer_)
  ) |>
  mutate(
    cl1         = factor(cl1, levels = CELL_LINES_V),
    cl2         = factor(cl2, levels = rev(CELL_LINES_V)),
    nsamp_label = factor(paste0("n_samples>=", min_nsamp),
                         levels = paste0("n_samples>=", NSAMP_CUTS)),
    dr_label    = factor(
      paste0("d>=", min_depth, " r>=", min_ratio),
      levels = apply(expand.grid(d = DEPTH_CUTS, r = RATIO_CUTS), 1,
                     function(x) paste0("d>=", x["d"], " r>=", x["r"]))
    ),
    cell_label  = case_when(
      cl1 == cl2     ~ "1.000",
      is.na(jaccard) ~ "NA",
      TRUE           ~ sprintf("%.3f\n(n=%d)", jaccard, n_intersect)
    )
  )

ggplot(jaccard_cl_sym, aes(x = cl1, y = cl2, fill = jaccard)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = cell_label), size = 2.8, fontface = "bold",
            lineheight = 0.9) +
  scale_fill_gradient2(
    low = "#1a9641", mid = "#ffffbf", high = "#d7191c",
    midpoint = 0.5, limits = c(0, 1), na.value = "grey90",
    name = "Jaccard"
  ) +
  facet_grid(nsamp_label ~ dr_label) +
  labs(
    title = "Jaccard index of site overlap between cell lines (mRNA pooled sites)",
    x = NULL, y = NULL
  ) +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1, size = 9),
    axis.text.y      = element_text(size = 9),
    strip.background = element_rect(fill = "#e8eaf6"),
    strip.text       = element_text(face = "bold", size = 9),
    panel.spacing    = unit(0.3, "lines"),
    legend.position  = "right"
  )
```

---

## Pearson's r of %pU between cell lines

To assess whether the same sites are similarly modified across cell lines, we
compute Pearson's r on the pooled %pU values at sites passing all three cutoff
filters in **both** cell lines of each pair.  High r across cell lines
suggests that the detected modification pattern is consistent — i.e., the
same mRNA sites are strongly modified regardless of cell line.

Sites that pass in one cell line but not the other are excluded from the
correlation (they contribute to the Jaccard-style overlap question, not the
quantitative agreement question).

```{r mrna-pearson-setup}
pearson_pair <- function(dt1, dt2, min_depth, min_ratio, min_nsamp) {
  pass1 <- dt1[depth >= min_depth & ratio >= min_ratio & n_samples >= min_nsamp,
               .(site, ratio1 = ratio)]
  pass2 <- dt2[depth >= min_depth & ratio >= min_ratio & n_samples >= min_nsamp,
               .(site, ratio2 = ratio)]
  merged <- merge(pass1, pass2, by = "site")
  if (nrow(merged) < 3) return(data.frame(r = NA_real_, n_common = nrow(merged)))
  data.frame(
    r        = cor(merged$ratio1, merged$ratio2, method = "pearson"),
    n_common = nrow(merged)
  )
}

cor_grid <- do.call(rbind, lapply(cl_pairs, function(p) {
  do.call(rbind, lapply(DEPTH_CUTS, function(d) {
    do.call(rbind, lapply(RATIO_CUTS, function(r) {
      do.call(rbind, lapply(NSAMP_CUTS, function(n) {
        res <- pearson_pair(
          dat_list[[p[1]]], dat_list[[p[2]]],
          min_depth = d, min_ratio = r, min_nsamp = n
        )
        data.frame(cl1 = p[1], cl2 = p[2],
                   min_depth = d, min_ratio = r, min_nsamp = n,
                   pearson_r = round(res$r, 3),
                   n_common  = res$n_common)
      }))
    }))
  }))
}))

cor_sym <- cor_grid |>
  bind_rows(
    cor_grid |> rename(cl1 = cl2, cl2 = cl1)
  ) |>
  bind_rows(
    expand.grid(cl1 = CELL_LINES_V, cl2 = CELL_LINES_V,
                min_depth = DEPTH_CUTS, min_ratio = RATIO_CUTS,
                min_nsamp = NSAMP_CUTS) |>
      filter(cl1 == cl2) |>
      mutate(pearson_r = 1, n_common = NA_integer_)
  ) |>
  mutate(
    cl1         = factor(cl1, levels = CELL_LINES_V),
    cl2         = factor(cl2, levels = rev(CELL_LINES_V)),
    nsamp_label = factor(paste0("n_samples>=", min_nsamp),
                         levels = paste0("n_samples>=", NSAMP_CUTS)),
    dr_label    = factor(
      paste0("d>=", min_depth, " r>=", min_ratio),
      levels = apply(
        expand.grid(d = DEPTH_CUTS, r = RATIO_CUTS), 1,
        function(x) paste0("d>=", x["d"], " r>=", x["r"])
      )
    )
  )
```

### Correlation matrices

The grid of heatmaps below shows Pearson's r between every pair of cell lines
under all combinations of cutoffs.  Rows = minimum n_samples; columns = depth
× ratio cutoff.  Diagonal cells equal 1 by definition.  Warmer (red) colours
indicate higher agreement; a grey cell means too few common sites to compute r.

```{r mrna-pearson-heatmap, fig.width=14, fig.height=11}
ggplot(cor_sym, aes(x = cl1, y = cl2, fill = pearson_r)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = ifelse(cl1 == cl2, "1", sprintf("%.3f", pearson_r))),
            size = 3, fontface = "bold") +
  scale_fill_gradient2(
    low = "#1a9641", mid = "#ffffbf", high = "#d7191c",
    midpoint = 0.5, limits = c(0, 1), na.value = "grey90",
    name = "Pearson r"
  ) +
  facet_grid(nsamp_label ~ dr_label) +
  labs(
    title = "Pearson's r of pooled %pU between cell lines (mRNA sites)",
    x = NULL, y = NULL
  ) +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1, size = 9),
    axis.text.y      = element_text(size = 9),
    strip.background = element_rect(fill = "#e8eaf6"),
    strip.text       = element_text(face = "bold", size = 9),
    panel.spacing    = unit(0.3, "lines"),
    legend.position  = "right"
  )
```

### Scatter plots (depth >= 20, ratio >= 0.1, n_samples >= 2)

Representative scatter plots at a moderate cutoff (depth >= 20, %pU >= 0.1,
detected in >= 2 replicates).  Each point is one mRNA site.

```{r mrna-pearson-scatter, fig.width=13, fig.height=8}
MIN_D <- 20; MIN_R <- 0.1; MIN_N <- 2

scatter_data <- lapply(CELL_LINES_V, function(cl) {
  dat_list[[cl]][depth >= MIN_D & ratio >= MIN_R & n_samples >= MIN_N,
                 .(site, ratio, cell_line = cl)]
})
names(scatter_data) <- CELL_LINES_V

scatter_plots <- lapply(cl_pairs, function(p) {
  d1 <- scatter_data[[p[1]]]
  d2 <- scatter_data[[p[2]]]
  m  <- merge(d1, d2, by = "site", suffixes = c("_1", "_2"))
  if (nrow(m) < 3) return(NULL)
  r_val <- cor(m$ratio_1, m$ratio_2, method = "pearson")
  ggplot(m, aes(x = ratio_1, y = ratio_2)) +
    geom_point(alpha = 0.55, size = 1.2, color = "#2c7bb6") +
    geom_abline(slope = 1, intercept = 0, color = "red",
                linetype = "dashed", linewidth = 0.6) +
    annotate("text", x = Inf, y = -Inf,
             label = sprintf("r = %.3f\nn = %d", r_val, nrow(m)),
             hjust = 1.1, vjust = -0.3, size = 3.5) +
    scale_x_continuous(labels = percent_format(accuracy = 1)) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      title = paste(p[1], "vs", p[2]),
      x     = paste0("%pU  ", p[1]),
      y     = paste0("%pU  ", p[2])
    ) +
    theme_bw(base_size = 10)
})
scatter_plots <- Filter(Negate(is.null), scatter_plots)

do.call(grid.arrange, c(scatter_plots, list(ncol = 3)))
```

---

# Comparing with published data

We benchmark our pooled mRNA calls against two independent published datasets.
Overlap with validated sites confirms that the pipeline is correctly identifying
true pseudouridine modifications in mRNA.

- **BACS** (PMID 39349603) detects pU via antibody-based capture; it reports
  absolute Ψ levels for HeLa polyA+ RNA (Supplementary Table 8).
- **PRAISE** (PMID 36997645) detects pU via the deletion signature at reverse-
  transcription termination; it reports deletion ratios for two replicates in
  HEK293T mRNAs and ncRNAs (Supplementary Dataset 2 — filtered to mRNA only).

Both published datasets use genomic coordinates, matching our pooled mRNA
output (chromosome + genomic position + strand). BACS mRNA sites with
`feature == "ncRNA"` are excluded. PRAISE sites with `NR_` or `XR_` accession
prefixes (ncRNA) are excluded; only `NM_` and `XM_` (mRNA) entries are kept.

For each cell line and each combination of depth / ratio cutoffs we report:
(a) how many of our pooled sites overlap with BACS sites, (b) how many overlap
with PRAISE sites, and (c) scatter plots comparing the quantitative %pU or
deletion ratio values at overlapping positions.

```{r mrna-pub-helpers}
make_scatter_publication <- function(df, x, y, col, ttl, xlab, ylab) {
  if (nrow(df) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = "No overlapping sites",
                 size = 4, color = "grey50") +
        labs(title = ttl, x = xlab, y = ylab) +
        theme_classic(base_size = 11) +
        theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))
    )
  }
  ggplot(df, aes(x = .data[[x]], y = .data[[y]],
                 color = .data[[col]], label = label)) +
    geom_point(size = 2.5, alpha = 0.8) +
    geom_text_repel(size = 2.5, max.overlaps = 20) +
    geom_smooth(method = "lm", se = TRUE, color = "grey40", linetype = "dashed") +
    scale_color_manual(values = c("CDS"  = "#4E79A7", "3UTR" = "#F28E2B",
                                  "5UTR" = "#59A14F", "mRNA" = "#888888")) +
    labs(title = ttl, x = xlab, y = ylab, color = "feature") +
    theme_classic(base_size = 11) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))
}

make_heatmap_publication <- function(data, fill_col, label_col, ttl, fill_label) {
  ggplot(data, aes(x = min_ratio, y = min_depth, fill = .data[[fill_col]])) +
    geom_tile(color = "white", linewidth = 1.2) +
    geom_text(aes(label = .data[[label_col]]), size = 6, fontface = "bold", color = "white") +
    scale_fill_gradient(low = "#d0e8f2", high = "#1a6fa8") +
    labs(title = ttl, x = "Min ratio cutoff", y = "Min depth cutoff", fill = fill_label) +
    theme_classic(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          axis.text  = element_text(size = 11))
}

combine_panels <- function(results, slot) {
  plots <- map(results, slot)
  wrap_plots(plots, ncol = 2) +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
}
```

## Published data overview

Site counts and the overlap between BACS and PRAISE are shown first to
establish the published reference landscape before comparing with our data.

```{r mrna-pub-load, warning=FALSE, message=FALSE}
path_BACS   <- "/project2/xinhe/xsun/pumseq/data_others/pU_PMID39349603.xlsx"
path_PRAISE <- "/project2/xinhe/xsun/pumseq/data_others/pU_PMID36997645.xlsx"

# BACS mRNA: Supplementary Table 8 (sheet 6) — HeLa polyA+ RNA
mrna_BACS <- readxl::read_excel(path_BACS, sheet = 6, skip = 2)

mrna_BACS_clean <- mrna_BACS %>%
  filter(feature != "ncRNA") %>%
  rename(psi_level = `Ψ level`) %>%
  mutate(
    chr_num   = sub("^chr", "", chr),
    pos_num   = as.integer(pos),
    psi_level = as.numeric(psi_level)
  )

# PRAISE mRNA+ncRNA: Supplementary Dataset 2 (sheet 2) — HEK293T mRNAs and ncRNAs
# Filter to mRNA only: keep NM_ and XM_ accessions (exclude NR_/XR_ ncRNA).
mrna_PRAISE <- readxl::read_excel(path_PRAISE, sheet = 2, skip = 2)

mrna_PRAISE_clean <- mrna_PRAISE %>%
  filter(str_starts(chr_name, "NM_") | str_starts(chr_name, "XM_")) %>%
  mutate(
    chr_num       = sub("^chr", "", str_extract(chr_site, "^chr[^_]+")),
    genomic_start = as.integer(str_extract(chr_site, "(?<=_)[0-9]+")),
    genomic_end   = as.integer(coalesce(
      str_extract(chr_site, "(?<=-)[0-9]+$"),
      str_extract(chr_site, "(?<=_)[0-9]+")
    ))
  ) %>%
  filter(!is.na(chr_num))

# BACS–PRAISE mRNA overlap (genomic coordinate join)
overlap <- mrna_BACS_clean %>%
  inner_join(mrna_PRAISE_clean, by = "chr_num",
             relationship = "many-to-many") %>%
  filter(pos_num >= pmin(genomic_start, genomic_end) &
         pos_num <= pmax(genomic_start, genomic_end))

DT::datatable(
  mrna_BACS_clean[, c("chr", "pos", "strand", "gene", "feature", "psi_level")],
  caption = htmltools::tags$caption(
    style = 'caption-side: left; text-align: left; color:black; font-size:150%;',
    'Sites from BACS (mRNA only)'),
  options = list(pageLength = 5))

DT::datatable(
  mrna_PRAISE_clean[, c("chr_name", "site", "chr_site", "gene_name",
                        "rep1_deletion_ratio", "rep2_deletion_ratio")],
  caption = htmltools::tags$caption(
    style = 'caption-side: left; text-align: left; color:black; font-size:150%;',
    'Sites from PRAISE (mRNA only)'),
  options = list(pageLength = 5))

cat("mRNA sites in BACS:    ", nrow(mrna_BACS_clean), "\n")
cat("mRNA sites in PRAISE:  ", nrow(mrna_PRAISE_clean), "\n")
cat("Overlapping sites:", nrow(overlap), "\n")
```

```{r mrna-pub-preload, include=FALSE}
run_cl_comparison <- function(cl) {
  pooled_path <- file.path(RESULTS_DIR, cl, "pooled_sites", "pooled_mRNA.txt")
  if (!file.exists(pooled_path)) {
    warning("File not found: ", pooled_path)
    return(NULL)
  }

  pooled_ours_mrna <- fread(pooled_path,
                             colClasses = list(character = "ref"))

  results <- lapply(seq_len(nrow(cutoffs)), function(i) {
    min_depth <- cutoffs$min_depth[i]
    min_ratio <- cutoffs$min_ratio[i]
    tag       <- paste0("depth>=", min_depth, "  ratio>=", min_ratio)

    ours_filtered <- pooled_ours_mrna %>%
      filter(depth >= min_depth, ratio >= min_ratio) %>%
      mutate(chr_num = as.character(ref))

    # Join with BACS by exact genomic coordinates (chr + pos + strand)
    overlap_bacs <- ours_filtered %>%
      inner_join(mrna_BACS_clean,
                 by = c("chr_num", "pos" = "pos_num", "strand")) %>%
      mutate(label = paste0(gene, " pos", pos))

    # Join with PRAISE by chr + genomic range
    overlap_praise <- ours_filtered %>%
      inner_join(mrna_PRAISE_clean, by = "chr_num",
                 relationship = "many-to-many") %>%
      filter(pos >= pmin(genomic_start, genomic_end) &
             pos <= pmax(genomic_start, genomic_end)) %>%
      mutate(
        deletion_rep1 = as.numeric(rep1_deletion_ratio),
        deletion_rep2 = as.numeric(rep2_deletion_ratio),
        feature       = "mRNA",
        label         = paste0(gene_name, " pos", pos)
      )

    list(
      tag      = tag,
      min_depth = min_depth,
      min_ratio = min_ratio,
      n_ours   = nrow(ours_filtered),
      n_bacs   = nrow(overlap_bacs),
      n_praise = nrow(overlap_praise),
      overlap_bacs   = overlap_bacs,
      overlap_praise = overlap_praise,

      p_bacs      = make_scatter_publication(
                      overlap_bacs, "ratio", "psi_level", "feature",
                      tag, "Our pU ratio", "BACS Ψ level"),
      p_praise_r1 = make_scatter_publication(
                      overlap_praise, "ratio", "deletion_rep1", "feature",
                      tag, "Our pU ratio", "PRAISE deletion ratio (rep1)"),
      p_praise_r2 = make_scatter_publication(
                      overlap_praise, "ratio", "deletion_rep2", "feature",
                      tag, "Our pU ratio", "PRAISE deletion ratio (rep2)")
    )
  })

  names(results) <- paste0("depth", cutoffs$min_depth, "_ratio", cutoffs$min_ratio)
  results
}

pub_data        <- lapply(CELL_LINES_V, run_cl_comparison)
names(pub_data) <- CELL_LINES_V
```

---

## CL18870 {.tabset}

```{r mrna-cl18870-pub-check, include=FALSE}
cl      <- "CL18870"
results <- pub_data[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled file not found for", cl)`

### Site counts

Tile colour indicates the number of sites: **Our sites** (left) are those
passing the given depth and ratio cutoff in our pooled data; **Overlap with
BACS/PRAISE** (middle, right) counts how many of those sites are also
present in the respective published dataset.

```{r mrna-cl18870-pub-heatmap, eval=ok, fig.width=12, fig.height=4}
heatmap_data <- map_dfr(results, ~ tibble(
  min_depth = .x$min_depth, min_ratio = .x$min_ratio,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(min_depth = factor(paste0("depth>=", min_depth)),
         min_ratio = factor(paste0("ratio>=", min_ratio)))

make_heatmap_publication(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, "  site counts across depth & ratio cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

### vs BACS

Scatter plots of our pooled %pU ratio (x-axis) vs. the BACS Ψ level (y-axis)
at overlapping sites, faceted by depth/ratio cutoff. Points are coloured by
mRNA feature (CDS / 3UTR / 5UTR).

```{r mrna-cl18870-pub-bacs, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS Ψ level"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

Our pooled %pU ratio vs. the PRAISE deletion ratio from replicate 1.

```{r mrna-cl18870-pub-praise-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r mrna-cl18870-pub-praise-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

---

## CL18909 {.tabset}

```{r mrna-cl18909-pub-check, include=FALSE}
cl      <- "CL18909"
results <- pub_data[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled file not found for", cl)`

### Site counts

```{r mrna-cl18909-pub-heatmap, eval=ok, fig.width=12, fig.height=4}
heatmap_data <- map_dfr(results, ~ tibble(
  min_depth = .x$min_depth, min_ratio = .x$min_ratio,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(min_depth = factor(paste0("depth>=", min_depth)),
         min_ratio = factor(paste0("ratio>=", min_ratio)))

make_heatmap_publication(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, "  site counts across depth & ratio cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

### vs BACS

```{r mrna-cl18909-pub-bacs, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS Ψ level"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

```{r mrna-cl18909-pub-praise-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r mrna-cl18909-pub-praise-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

---

## CL18916 {.tabset}

```{r mrna-cl18916-pub-check, include=FALSE}
cl      <- "CL18916"
results <- pub_data[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled file not found for", cl)`

### Site counts

```{r mrna-cl18916-pub-heatmap, eval=ok, fig.width=12, fig.height=4}
heatmap_data <- map_dfr(results, ~ tibble(
  min_depth = .x$min_depth, min_ratio = .x$min_ratio,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(min_depth = factor(paste0("depth>=", min_depth)),
         min_ratio = factor(paste0("ratio>=", min_ratio)))

make_heatmap_publication(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, "  site counts across depth & ratio cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

### vs BACS

```{r mrna-cl18916-pub-bacs, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS Ψ level"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

```{r mrna-cl18916-pub-praise-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r mrna-cl18916-pub-praise-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

---

## CL19099 {.tabset}

```{r mrna-cl19099-pub-check, include=FALSE}
cl      <- "CL19099"
results <- pub_data[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled file not found for", cl)`

### Site counts

```{r mrna-cl19099-pub-heatmap, eval=ok, fig.width=12, fig.height=4}
heatmap_data <- map_dfr(results, ~ tibble(
  min_depth = .x$min_depth, min_ratio = .x$min_ratio,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(min_depth = factor(paste0("depth>=", min_depth)),
         min_ratio = factor(paste0("ratio>=", min_ratio)))

make_heatmap_publication(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, "  site counts across depth & ratio cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

### vs BACS

```{r mrna-cl19099-pub-bacs, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS Ψ level"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

```{r mrna-cl19099-pub-praise-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r mrna-cl19099-pub-praise-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```
