---
title: "Quality control using sites called from mRNA — chi-squared strategy"
author: "XSun"
date: "2026-06-06"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

# Introduction

This report mirrors [qc_rRNA_chisq.html](qc_rRNA_chisq.html) but evaluates the sites called from mRNA.

For every uridine (T) position we build a 2x2 table comparing the treated and
(pooled) untreated samples:

|            | Conversion (C) | No conversion (T) |
|------------|----------------|-------------------|
| Treated    | a              | b                 |
| Untreated  | c              | d                 |

and test the null hypothesis that the conversion rate is the same in both
samples ($X^2 = \frac{n(ad-bc)^2}{(a+b)(c+d)(a+c)(b+d)}$, 1 d.f.). A site is
called pseudouridylated when:

1. the chi-squared Benjamini-Hochberg **FDR < cutoff**, and
2. the **conversion-rate difference** (treated − untreated) **> cutoff**.

<!-- This complements the original ratio/depth-based filter: instead of asking -->
<!-- "is the treated ratio high and the input ratio low?" it asks "is the -->
<!-- treated-vs-untreated difference at this position larger than expected by -->
<!-- chance, given the sequencing depth?" — a test that naturally accounts for -->
<!-- coverage when judging significance. -->

We ran this strategy on the same four cell lines: **CL18870**, **CL18909**,
**CL18916**, and **CL19099** (four treated replicates each).

We compare our results against two published datasets:

- **BACS** — PMID 39349603
- **PRAISE** — PMID 36997645

```{r setup, warning=FALSE, message=FALSE}
library(dplyr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(purrr)
library(patchwork)
library(tidyverse)
library(data.table)
library(scales)
library(knitr)
library(gridExtra)
library(DT)
library(grid)

knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center"
)
```

---

```{r setup-config}
RESULTS_DIR <- "/project2/xinhe/xsun/pumseq/pumseq_results"

CELL_LINES <- list(
  CL18870 = list(
    rep1 = "CL18870_12-187-59",
    rep2 = "CL18870_12-187-60",
    rep3 = "CL18870_12-187-97",
    rep4 = "CL18870_12-187-98"
  ),
  CL18909 = list(
    rep1 = "CL18909_12-187-13",
    rep2 = "CL18909_12-187-14",
    rep3 = "CL18909_13-32-23",
    rep4 = "CL18909_13-32-24"
  ),
  CL18916 = list(
    rep1 = "CL18916_12-187-61",
    rep2 = "CL18916_12-187-62",
    rep3 = "CL18916_12-187-101",
    rep4 = "CL18916_12-187-102"
  ),
  CL19099 = list(
    rep1 = "CL19099_12-187-65",
    rep2 = "CL19099_12-187-66",
    rep3 = "CL19099_12-187-105",
    rep4 = "CL19099_12-187-106"
  )
)

# Cutoff combinations applied to pooled chi-squared output
CUTOFF_COMBOS <- list(
  list(fdr = 0.001, diff = 0.05, label = "FDR < 0.001 & rate_diff > 0.05"),
  list(fdr = 0.001, diff = 0.10, label = "FDR < 0.001 & rate_diff > 0.10"),
  list(fdr = 0.01,  diff = 0.05, label = "FDR < 0.01 & rate_diff > 0.05"),
  list(fdr = 0.01,  diff = 0.10, label = "FDR < 0.01 & rate_diff > 0.10")
)
```

---

# Comparing pooled data among cell lines

After per-replicate chi-squared filtering, sites for each cell line are
**pooled** (`pooled_sites_chisq/pooled_mRNA.txt`): treated and untreated
counts are summed across the four replicates and the chi-squared test is
re-run on the pooled contingency table, yielding pooled `pval`, `fdr`, and
`rate_diff` statistics alongside the pooled %pU ratio. Each row also records
`n_samples`, the number of individual replicates in which the site was
detected.

We explore the effect of two filtering dimensions simultaneously, using the
same four FDR / rate-difference combinations introduced above together with a
minimum-replicate filter:

| Dimension | Values |
|-----------|--------|
| FDR / rate-difference combination | the four combinations above |
| Minimum number of replicates (`n_samples`) | 1, 2, 3, 4 |

Stricter combinations (lower FDR, larger rate difference) reduce noise at the
cost of fewer sites; requiring detection in multiple replicates is a strong
reproducibility filter.

```{r setup-pooled-chisq, message=FALSE, warning=FALSE}
CELL_LINES_V      <- names(CELL_LINES)
NSAMP_CUTS_POOLED <- 1:4
COMBO_LABELS      <- vapply(CUTOFF_COMBOS, `[[`, character(1), "label")

load_pooled_chisq_rRNA <- function(cl) {
  f <- file.path(RESULTS_DIR, cl, "pooled_sites_chisq", "pooled_mRNA.txt")
  dt <- fread(f, header = TRUE, sep = "\t",
              colClasses = list(character = "ref"))
  dt[, cell_line := cl]
  dt[, site := paste(ref, pos, strand, sep = ":")]
  dt
}

count_sites_chisq <- function(dt, combo, min_nsamp) {
  sum(dt$fdr < combo$fdr & dt$rate_diff > combo$diff & dt$n_samples >= min_nsamp)
}

build_count_grid_chisq <- function(dt) {
  do.call(rbind, lapply(seq_along(CUTOFF_COMBOS), function(i) {
    combo <- CUTOFF_COMBOS[[i]]
    data.frame(
      combo_label = combo$label,
      min_nsamp   = NSAMP_CUTS_POOLED,
      n_sites     = vapply(NSAMP_CUTS_POOLED,
                           function(n) count_sites_chisq(dt, combo, n),
                           numeric(1))
    )
  })) |>
    mutate(
      combo_label = factor(combo_label, levels = COMBO_LABELS),
      nsamp_label = factor(paste0("n_samples>=", min_nsamp),
                           levels = paste0("n_samples>=", NSAMP_CUTS_POOLED))
    )
}

plot_count_heatmap_chisq <- function(grid, cl_title) {
  ggplot(grid, aes(x = combo_label, y = nsamp_label, fill = n_sites)) +
    geom_tile(color = "white", linewidth = 0.6) +
    geom_text(aes(label = comma(n_sites)), size = 4, fontface = "bold") +
    scale_fill_gradient(low = "#deebf7", high = "#2171b5",
                        name = "# sites", labels = comma) +
    labs(title = cl_title,
         x = "FDR / rate-difference cutoff (pooled)",
         y = "Minimum number of replicates") +
    theme_bw(base_size = 11) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}

dat_list_pooled_chisq        <- lapply(CELL_LINES_V, load_pooled_chisq_rRNA)
names(dat_list_pooled_chisq) <- CELL_LINES_V

grid_list_pooled_chisq        <- lapply(dat_list_pooled_chisq, build_count_grid_chisq)
names(grid_list_pooled_chisq) <- CELL_LINES_V

cl_pairs_v_chisq <- combn(CELL_LINES_V, 2, simplify = FALSE)
```

## Per cell-line site counts at different cutoffs

Each heatmap shows the number of pooled mRNA sites surviving a combination of
**FDR / rate-difference cutoff** (x-axis, the four combinations introduced
above) and **minimum number of replicates** detecting the site (y-axis).
Darker blue = more sites.

---

### CL18870

```{r pooled-chisq-cl18870-count-heatmap, fig.width=8, fig.height=4}
plot_count_heatmap_chisq(grid_list_pooled_chisq[["CL18870"]], "CL18870 - pooled chi-squared site counts")
```

---

### CL18909

```{r pooled-chisq-cl18909-count-heatmap, fig.width=8, fig.height=4}
plot_count_heatmap_chisq(grid_list_pooled_chisq[["CL18909"]], "CL18909 - pooled chi-squared site counts")
```

---

### CL18916

```{r pooled-chisq-cl18916-count-heatmap, fig.width=8, fig.height=4}
plot_count_heatmap_chisq(grid_list_pooled_chisq[["CL18916"]], "CL18916 - pooled chi-squared site counts")
```

---

### CL19099

```{r pooled-chisq-cl19099-count-heatmap, fig.width=8, fig.height=4}
plot_count_heatmap_chisq(grid_list_pooled_chisq[["CL19099"]], "CL19099 - pooled chi-squared site counts")
```

---

### %pU Distribution — pooled data {.tabset}

Distribution of the pooled %ψ ratio (C/(C+T)) at sites passing each FDR /
rate-difference combination, shown for all four cell lines side by side
(mirroring the per-replicate distributions above). Red dashed = median;
orange dotted = mean.

```{r pooled-chisq-dist-helper, include=FALSE}
plot_pooled_dist_chisq <- function(cl, combo) {
  dt <- dat_list_pooled_chisq[[cl]]
  df <- as.data.frame(dt[fdr < combo$fdr & rate_diff > combo$diff, .(psi = ratio)])
  if (nrow(df) == 0) return(NULL)

  med_val  <- median(df$psi)
  mean_val <- mean(df$psi)

  ggplot(df, aes(x = psi)) +
    geom_histogram(aes(y = after_stat(density)),
                   binwidth = 0.02, fill = "steelblue",
                   color = "white", linewidth = 0.2, alpha = 0.7) +
    geom_density(color = "navy", linewidth = 0.7) +
    geom_vline(xintercept = med_val,  color = "red",
               linetype = "dashed",  linewidth = 0.6) +
    geom_vline(xintercept = mean_val, color = "darkorange",
               linetype = "dotted",  linewidth = 0.6) +
    annotate("text", x = Inf, y = Inf,
             label = sprintf("n = %d\nmedian = %.3f\nmean = %.3f",
                             nrow(df), med_val, mean_val),
             hjust = 1.05, vjust = 1.05, size = 3, family = "mono") +
    scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
    labs(title = cl, subtitle = combo$label,
         x = "%pU  (C / C+T, pooled)", y = "Density") +
    theme_bw(base_size = 10) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

#### %pU Distribution — FDR < 0.001 & rate_diff > 0.05

```{r pooled-chisq-dist1, fig.width=14, fig.height=3.5}
plots_pdist1 <- lapply(CELL_LINES_V, plot_pooled_dist_chisq, combo = CUTOFF_COMBOS[[1]])
do.call(grid.arrange, c(Filter(Negate(is.null), plots_pdist1), list(ncol = 4)))
```

#### %pU Distribution — FDR < 0.001 & rate_diff > 0.10

```{r pooled-chisq-dist2, fig.width=14, fig.height=3.5}
plots_pdist2 <- lapply(CELL_LINES_V, plot_pooled_dist_chisq, combo = CUTOFF_COMBOS[[2]])
do.call(grid.arrange, c(Filter(Negate(is.null), plots_pdist2), list(ncol = 4)))
```

#### %pU Distribution — FDR < 0.01 & rate_diff > 0.05

```{r pooled-chisq-dist3, fig.width=14, fig.height=3.5}
plots_pdist3 <- lapply(CELL_LINES_V, plot_pooled_dist_chisq, combo = CUTOFF_COMBOS[[3]])
do.call(grid.arrange, c(Filter(Negate(is.null), plots_pdist3), list(ncol = 4)))
```

#### %pU Distribution — FDR < 0.01 & rate_diff > 0.10

```{r pooled-chisq-dist4, fig.width=14, fig.height=3.5}
plots_pdist4 <- lapply(CELL_LINES_V, plot_pooled_dist_chisq, combo = CUTOFF_COMBOS[[4]])
do.call(grid.arrange, c(Filter(Negate(is.null), plots_pdist4), list(ncol = 4)))
```

---

## Cross-cell-line comparison of site counts

We use the **Jaccard index** to quantify how much the pooled, significance
-filtered site sets overlap between every pair of cell lines. For each
combination of FDR / rate-difference cutoff and minimum-replicate filter,
sites passing both filters in each cell line are extracted and the pairwise
Jaccard is computed:

$$J(A, B) = \frac{|A \cap B|}{|A \cup B|}$$

A high Jaccard index means the two cell lines call largely the same pooled
mRNA positions. Each cell shows the Jaccard score and, in parentheses, the
number of sites in the intersection. Rows = minimum n_samples; columns = FDR /
rate-difference combination.

```{r pooled-chisq-cross-cl-comparison, fig.width=14, fig.height=11}
jaccard_pooled_rows <- do.call(rbind, lapply(cl_pairs_v_chisq, function(p) {
  do.call(rbind, lapply(seq_along(CUTOFF_COMBOS), function(i) {
    combo <- CUTOFF_COMBOS[[i]]
    do.call(rbind, lapply(NSAMP_CUTS_POOLED, function(n) {
      dt1 <- dat_list_pooled_chisq[[p[1]]]
      dt2 <- dat_list_pooled_chisq[[p[2]]]
      sites1 <- dt1$site[dt1$fdr < combo$fdr & dt1$rate_diff > combo$diff & dt1$n_samples >= n]
      sites2 <- dt2$site[dt2$fdr < combo$fdr & dt2$rate_diff > combo$diff & dt2$n_samples >= n]
      n_int  <- length(intersect(sites1, sites2))
      n_uni  <- length(union(sites1, sites2))
      j      <- if (n_uni == 0) NA_real_ else n_int / n_uni
      data.frame(cl1 = p[1], cl2 = p[2],
                 combo_idx = i, combo_label = combo$label, min_nsamp = n,
                 jaccard = round(j, 3), n_intersect = n_int)
    }))
  }))
}))

jaccard_pooled_sym <- jaccard_pooled_rows |>
  bind_rows(jaccard_pooled_rows |> rename(cl1 = cl2, cl2 = cl1)) |>
  bind_rows(
    expand.grid(cl1 = CELL_LINES_V, cl2 = CELL_LINES_V,
                combo_idx = seq_along(CUTOFF_COMBOS),
                min_nsamp = NSAMP_CUTS_POOLED) |>
      filter(cl1 == cl2) |>
      mutate(combo_label = COMBO_LABELS[combo_idx],
             jaccard = 1, n_intersect = NA_integer_)
  ) |>
  mutate(
    cl1         = factor(cl1, levels = CELL_LINES_V),
    cl2         = factor(cl2, levels = rev(CELL_LINES_V)),
    nsamp_label = factor(paste0("n_samples>=", min_nsamp),
                         levels = paste0("n_samples>=", NSAMP_CUTS_POOLED)),
    combo_label = factor(combo_label, levels = COMBO_LABELS),
    cell_label  = case_when(
      cl1 == cl2     ~ "1.000",
      is.na(jaccard) ~ "NA",
      TRUE           ~ sprintf("%.3f\n(n=%d)", jaccard, n_intersect)
    )
  )

ggplot(jaccard_pooled_sym, aes(x = cl1, y = cl2, fill = jaccard)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = cell_label), size = 2.8, fontface = "bold",
            lineheight = 0.9) +
  scale_fill_gradient2(
    low = "#1a9641", mid = "#ffffbf", high = "#d7191c",
    midpoint = 0.5, limits = c(0, 1), na.value = "grey90",
    name = "Jaccard"
  ) +
  facet_grid(nsamp_label ~ combo_label) +
  labs(
    title = "Jaccard index of site overlap between cell lines (pooled chi-squared mRNA sites)",
    x = NULL, y = NULL
  ) +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1, size = 9),
    axis.text.y      = element_text(size = 9),
    strip.background = element_rect(fill = "#e8eaf6"),
    strip.text       = element_text(face = "bold", size = 9),
    panel.spacing    = unit(0.3, "lines"),
    legend.position  = "right"
  )
```

---

## Pearson's r of %pU between cell lines

To assess whether the same sites are similarly modified across cell lines, we
compute Pearson's r on the pooled %pU values at sites passing the FDR /
rate-difference filter and the minimum-replicate filter in **both** cell
lines of each pair. High r across cell lines suggests that the detected
modification pattern is consistent — i.e., the same mRNA sites are strongly
modified regardless of cell line.

Sites that pass in one cell line but not the other are excluded from the
correlation (they contribute to the Jaccard-style overlap question, not the
quantitative agreement question).

```{r pooled-chisq-pearson-setup}
pearson_pair_chisq <- function(dt1, dt2, combo, min_nsamp) {
  pass1 <- dt1[fdr < combo$fdr & rate_diff > combo$diff & n_samples >= min_nsamp,
               .(site, ratio1 = ratio)]
  pass2 <- dt2[fdr < combo$fdr & rate_diff > combo$diff & n_samples >= min_nsamp,
               .(site, ratio2 = ratio)]
  merged <- merge(pass1, pass2, by = "site")
  if (nrow(merged) < 3) return(data.frame(r = NA_real_, n_common = nrow(merged)))
  data.frame(
    r        = cor(merged$ratio1, merged$ratio2, method = "pearson"),
    n_common = nrow(merged)
  )
}

cor_pooled_grid <- do.call(rbind, lapply(cl_pairs_v_chisq, function(p) {
  do.call(rbind, lapply(seq_along(CUTOFF_COMBOS), function(i) {
    combo <- CUTOFF_COMBOS[[i]]
    do.call(rbind, lapply(NSAMP_CUTS_POOLED, function(n) {
      res <- pearson_pair_chisq(
        dat_list_pooled_chisq[[p[1]]], dat_list_pooled_chisq[[p[2]]],
        combo = combo, min_nsamp = n
      )
      data.frame(cl1 = p[1], cl2 = p[2],
                 combo_idx = i, combo_label = combo$label, min_nsamp = n,
                 pearson_r = round(res$r, 3), n_common = res$n_common)
    }))
  }))
}))

cor_pooled_sym <- cor_pooled_grid |>
  bind_rows(
    cor_pooled_grid |> rename(cl1 = cl2, cl2 = cl1)
  ) |>
  bind_rows(
    expand.grid(cl1 = CELL_LINES_V, cl2 = CELL_LINES_V,
                combo_idx = seq_along(CUTOFF_COMBOS),
                min_nsamp = NSAMP_CUTS_POOLED) |>
      filter(cl1 == cl2) |>
      mutate(combo_label = COMBO_LABELS[combo_idx],
             pearson_r = 1, n_common = NA_integer_)
  ) |>
  mutate(
    cl1         = factor(cl1, levels = CELL_LINES_V),
    cl2         = factor(cl2, levels = rev(CELL_LINES_V)),
    nsamp_label = factor(paste0("n_samples>=", min_nsamp),
                         levels = paste0("n_samples>=", NSAMP_CUTS_POOLED)),
    combo_label = factor(combo_label, levels = COMBO_LABELS)
  )
```

### Correlation matrices

The grid of heatmaps below shows Pearson's r between every pair of cell lines
under all combinations of cutoffs. Rows = minimum n_samples; columns = FDR /
rate-difference combination. Diagonal cells equal 1 by definition. Warmer
(red) colours indicate higher agreement; a grey cell means too few common
sites to compute r.

```{r pooled-chisq-pearson-heatmap, fig.width=14, fig.height=11}
ggplot(cor_pooled_sym, aes(x = cl1, y = cl2, fill = pearson_r)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = ifelse(cl1 == cl2, "1", sprintf("%.3f", pearson_r))),
            size = 3, fontface = "bold") +
  scale_fill_gradient2(
    low = "#1a9641", mid = "#ffffbf", high = "#d7191c",
    midpoint = 0.5, limits = c(0, 1), na.value = "grey90",
    name = "Pearson r"
  ) +
  facet_grid(nsamp_label ~ combo_label) +
  labs(
    title = "Pearson's r of pooled %pU between cell lines (chi-squared mRNA sites)",
    x = NULL, y = NULL
  ) +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1, size = 9),
    axis.text.y      = element_text(size = 9),
    strip.background = element_rect(fill = "#e8eaf6"),
    strip.text       = element_text(face = "bold", size = 9),
    panel.spacing    = unit(0.3, "lines"),
    legend.position  = "right"
  )
```

### Scatter plots (FDR < 0.001 & rate_diff > 0.05, n_samples >= 2)

Representative scatter plots at a moderately stringent cutoff (FDR < 0.001 &
rate_diff > 0.05, detected in >= 2 replicates). Each point is one pooled mRNA
site.

```{r pooled-chisq-pearson-scatter, fig.width=13, fig.height=8}
SCATTER_COMBO_POOLED <- CUTOFF_COMBOS[[1]]
MIN_N_POOLED         <- 2

scatter_data_chisq <- lapply(CELL_LINES_V, function(cl) {
  dt <- dat_list_pooled_chisq[[cl]]
  dt[fdr < SCATTER_COMBO_POOLED$fdr &
       rate_diff > SCATTER_COMBO_POOLED$diff &
       n_samples >= MIN_N_POOLED,
     .(site, ratio, cell_line = cl)]
})
names(scatter_data_chisq) <- CELL_LINES_V

scatter_plots_chisq <- lapply(cl_pairs_v_chisq, function(p) {
  d1 <- scatter_data_chisq[[p[1]]]
  d2 <- scatter_data_chisq[[p[2]]]
  m  <- merge(d1, d2, by = "site", suffixes = c("_1", "_2"))
  if (nrow(m) < 3) return(NULL)
  r_val <- cor(m$ratio_1, m$ratio_2, method = "pearson")
  ggplot(m, aes(x = ratio_1, y = ratio_2)) +
    geom_point(alpha = 0.55, size = 1.2, color = "#2c7bb6") +
    geom_abline(slope = 1, intercept = 0, color = "red",
                linetype = "dashed", linewidth = 0.6) +
    annotate("text", x = Inf, y = -Inf,
             label = sprintf("r = %.3f\nn = %d", r_val, nrow(m)),
             hjust = 1.1, vjust = -0.3, size = 3.5) +
    scale_x_continuous(labels = percent_format(accuracy = 1)) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      title = paste(p[1], "vs", p[2]),
      x     = paste0("%pU  ", p[1]),
      y     = paste0("%pU  ", p[2])
    ) +
    theme_bw(base_size = 10)
})
scatter_plots_chisq <- Filter(Negate(is.null), scatter_plots_chisq)

do.call(grid.arrange, c(scatter_plots_chisq, list(ncol = 3)))
```

---

# Comparing with published data

We benchmark our **pooled chi-squared** mRNA calls against two independent
published datasets. Overlap with validated sites confirms that the
chi-squared site-calling strategy is correctly identifying true
pseudouridine modifications in mRNA.

- **BACS** (PMID 39349603) detects pU via antibody-based capture; it reports
  absolute Ψ levels for HeLa polyA+ RNA (Supplementary Table 8).
- **PRAISE** (PMID 36997645) detects pU via the deletion signature at reverse-
  transcription termination; it reports deletion ratios for two replicates in
  HEK293T mRNAs and ncRNAs (Supplementary Dataset 2 — filtered to mRNA only).

Both published datasets use genomic coordinates, matching our pooled mRNA
output (chromosome + genomic position + strand). BACS mRNA sites with
`feature == "ncRNA"` are excluded. PRAISE sites with `NR_` or `XR_` accession
prefixes (ncRNA) are excluded; only `NM_` and `XM_` (mRNA) entries are kept.

For each cell line and each of the four FDR / rate-difference cutoff
combinations introduced above we report: (a) how many of our pooled,
significance-filtered sites overlap with BACS sites, (b) how many overlap
with PRAISE sites, and (c) scatter plots comparing the quantitative %pU or
deletion ratio values at overlapping positions.

```{r pub-chisq-helpers}
make_scatter_publication_chisq <- function(df, x, y, col, ttl, xlab, ylab) {
  if (nrow(df) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = "No overlapping sites",
                 size = 4, color = "grey50") +
        labs(title = ttl, x = xlab, y = ylab) +
        theme_classic(base_size = 11) +
        theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))
    )
  }
  ggplot(df, aes(x = .data[[x]], y = .data[[y]],
                 color = .data[[col]], label = label)) +
    geom_point(size = 2.5, alpha = 0.8) +
    geom_text_repel(size = 2.5, max.overlaps = 20) +
    geom_smooth(method = "lm", se = TRUE, color = "grey40", linetype = "dashed") +
    scale_color_manual(values = c("CDS"  = "#4E79A7", "3UTR" = "#F28E2B",
                                  "5UTR" = "#59A14F", "mRNA" = "#888888")) +
    labs(title = ttl, x = xlab, y = ylab, color = "feature") +
    theme_classic(base_size = 11) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))
}

make_heatmap_publication_chisq <- function(data, fill_col, label_col, ttl, fill_label) {
  ggplot(data, aes(x = diff_label, y = fdr_label, fill = .data[[fill_col]])) +
    geom_tile(color = "white", linewidth = 1.2) +
    geom_text(aes(label = .data[[label_col]]), size = 5, fontface = "bold", color = "white") +
    scale_fill_gradient(low = "#d0e8f2", high = "#1a6fa8") +
    facet_wrap(~ nsamp_label, ncol = 2) +
    labs(title = ttl, x = "Min rate-difference cutoff", y = "Max FDR cutoff", fill = fill_label) +
    theme_classic(base_size = 11) +
    theme(plot.title    = element_text(hjust = 0.5, face = "bold"),
          axis.text     = element_text(size = 9),
          strip.text    = element_text(size = 9, face = "bold"),
          strip.background = element_rect(fill = "#e8f0f7", color = NA))
}

make_pct_plot_chisq <- function(data, cl_title) {
  pal <- c(BACS = "#E41A1C", PRAISE = "#377EB8")

  combo_levels <- data %>%
    arrange(fdr_cut, diff_cut) %>%
    mutate(combo = paste0(as.character(fdr_label), "\n", as.character(diff_label))) %>%
    pull(combo) %>% unique()

  base <- data %>%
    mutate(combo = factor(paste0(as.character(fdr_label), "\n", as.character(diff_label)),
                          levels = combo_levels))

  long_ours <- base %>%
    select(nsamp_label, combo, BACS = pct_ours_in_bacs, PRAISE = pct_ours_in_praise) %>%
    pivot_longer(c(BACS, PRAISE), names_to = "dataset", values_to = "pct")

  long_pub <- base %>%
    select(nsamp_label, combo, BACS = pct_bacs_in_ours, PRAISE = pct_praise_in_ours) %>%
    pivot_longer(c(BACS, PRAISE), names_to = "dataset", values_to = "pct")

  bar_theme <- theme_bw(base_size = 10) +
    theme(legend.position  = "bottom",
          strip.text        = element_text(face = "bold", size = 9),
          axis.text.x       = element_text(size = 7.5, lineheight = 0.85))

  p1 <- ggplot(long_ours, aes(x = combo, y = pct, fill = dataset)) +
    geom_col(position = "dodge", width = 0.7) +
    facet_wrap(~ nsamp_label, ncol = 2) +
    scale_fill_manual(values = pal) +
    labs(subtitle = "% of our sites found in publication",
         x = NULL, y = "%", fill = "") +
    bar_theme

  p2 <- ggplot(long_pub, aes(x = combo, y = pct, fill = dataset)) +
    geom_col(position = "dodge", width = 0.7) +
    facet_wrap(~ nsamp_label, ncol = 2) +
    scale_fill_manual(values = pal) +
    labs(subtitle = "% of publication sites found in our data",
         x = "Cutoff combo", y = "%", fill = "") +
    bar_theme

  (p1 / p2) +
    plot_annotation(title = cl_title,
                    theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
}

combine_panels_chisq <- function(results, slot) {
  plots <- map(results, slot)
  wrap_plots(plots, ncol = 2) +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
}
```

## Published data overview

Site counts and the overlap between BACS and PRAISE are shown first to
establish the published reference landscape before comparing with our pooled
chi-squared calls.

```{r pub-chisq-load, warning=FALSE, message=FALSE}
path_BACS   <- "/project2/xinhe/xsun/pumseq/data_others/pU_PMID39349603.xlsx"
path_PRAISE <- "/project2/xinhe/xsun/pumseq/data_others/pU_PMID36997645.xlsx"

# BACS mRNA: Supplementary Table 8 (sheet 6) — HeLa polyA+ RNA
# Rows 1-2 are title/blank; row 3 is the header.
mrna_BACS <- readxl::read_excel(path_BACS, sheet = 6, skip = 2)

mrna_BACS_clean <- mrna_BACS %>%
  filter(feature != "ncRNA") %>%
  rename(psi_level = `Ψ level`) %>%
  mutate(
    chr_num   = sub("^chr", "", chr),
    pos_num   = as.integer(pos),
    psi_level = as.numeric(psi_level)
  )

# PRAISE mRNA+ncRNA: Supplementary Dataset 2 (sheet 2) — HEK293T mRNAs and ncRNAs
# Filter to mRNA only: keep NM_ and XM_ accessions (exclude NR_/XR_ ncRNA).
# chr_site carries genomic coordinates (e.g. "chr14_73219165-73219168").
mrna_PRAISE <- readxl::read_excel(path_PRAISE, sheet = 2, skip = 2)

mrna_PRAISE_clean <- mrna_PRAISE %>%
  filter(str_starts(chr_name, "NM_") | str_starts(chr_name, "XM_")) %>%
  mutate(
    chr_num       = sub("^chr", "", str_extract(chr_site, "^chr[^_]+")),
    genomic_start = as.integer(str_extract(chr_site, "(?<=_)[0-9]+")),
    genomic_end   = as.integer(coalesce(
      str_extract(chr_site, "(?<=-)[0-9]+$"),
      str_extract(chr_site, "(?<=_)[0-9]+")
    ))
  ) %>%
  filter(!is.na(chr_num))

# BACS–PRAISE mRNA overlap (genomic coordinate join)
overlap <- mrna_BACS_clean %>%
  inner_join(mrna_PRAISE_clean, by = "chr_num",
             relationship = "many-to-many") %>%
  filter(pos_num >= pmin(genomic_start, genomic_end) &
         pos_num <= pmax(genomic_start, genomic_end))

DT::datatable(
  mrna_BACS_clean[, c("chr", "pos", "strand", "gene", "feature", "psi_level")],
  caption = htmltools::tags$caption(
    style = 'caption-side: left; text-align: left; color:black; font-size:150%;',
    'Sites from BACS (mRNA only)'),
  options = list(pageLength = 5))

DT::datatable(
  mrna_PRAISE_clean[, c("chr_name", "site", "chr_site", "gene_name",
                        "rep1_deletion_ratio", "rep2_deletion_ratio")],
  caption = htmltools::tags$caption(
    style = 'caption-side: left; text-align: left; color:black; font-size:150%;',
    'Sites from PRAISE (mRNA only)'),
  options = list(pageLength = 5))

cat("mRNA sites in BACS:    ", nrow(mrna_BACS_clean), "\n")
cat("mRNA sites in PRAISE:  ", nrow(mrna_PRAISE_clean), "\n")
cat("Overlapping sites:", nrow(overlap), "\n")
```

```{r pub-chisq-preload, include=FALSE}
run_cl_comparison_chisq <- function(cl) {
  pooled_path <- file.path(RESULTS_DIR, cl, "pooled_sites_chisq", "pooled_mRNA.txt")
  if (!file.exists(pooled_path)) {
    warning("File not found: ", pooled_path)
    return(NULL)
  }

  pooled_ours_mrna_chisq <- fread(pooled_path,
                                   colClasses = list(character = "ref"))

  results_nested <- lapply(NSAMP_CUTS_POOLED, function(min_nsamp) {
    lapply(seq_along(CUTOFF_COMBOS), function(i) {
      combo <- CUTOFF_COMBOS[[i]]
      tag   <- combo$label

      ours_filtered <- pooled_ours_mrna_chisq %>%
        filter(fdr < combo$fdr, rate_diff > combo$diff, n_samples >= min_nsamp) %>%
        mutate(chr_num = as.character(ref))

      # Join with BACS by exact genomic coordinates (chr + pos + strand)
      overlap_bacs <- ours_filtered %>%
        inner_join(mrna_BACS_clean,
                   by = c("chr_num", "pos" = "pos_num", "strand")) %>%
        mutate(label = paste0(gene, " pos", pos))

      # Join with PRAISE by chr + genomic range (chr_site can span 1-4 bp)
      overlap_praise <- ours_filtered %>%
        inner_join(mrna_PRAISE_clean, by = "chr_num",
                   relationship = "many-to-many") %>%
        filter(pos >= pmin(genomic_start, genomic_end) &
               pos <= pmax(genomic_start, genomic_end)) %>%
        mutate(
          deletion_rep1 = as.numeric(rep1_deletion_ratio),
          deletion_rep2 = as.numeric(rep2_deletion_ratio),
          feature       = "mRNA",
          label         = paste0(gene_name, " pos", pos)
        )

      list(
        tag       = tag,
        fdr_cut   = combo$fdr,
        diff_cut  = combo$diff,
        min_nsamp = min_nsamp,
        n_ours    = nrow(ours_filtered),
        n_bacs    = nrow(overlap_bacs),
        n_praise  = nrow(overlap_praise),
        overlap_bacs   = overlap_bacs,
        overlap_praise = overlap_praise,

        p_bacs      = make_scatter_publication_chisq(
                        overlap_bacs, "ratio", "psi_level", "feature",
                        tag, "Our pU ratio (pooled, chi-squared)", "BACS Ψ level"),
        p_praise_r1 = make_scatter_publication_chisq(
                        overlap_praise, "ratio", "deletion_rep1", "feature",
                        tag, "Our pU ratio (pooled, chi-squared)", "PRAISE deletion ratio (rep1)"),
        p_praise_r2 = make_scatter_publication_chisq(
                        overlap_praise, "ratio", "deletion_rep2", "feature",
                        tag, "Our pU ratio (pooled, chi-squared)", "PRAISE deletion ratio (rep2)")
      )
    })
  })

  results <- unlist(results_nested, recursive = FALSE)
  names(results) <- paste0(
    "nsamp", rep(NSAMP_CUTS_POOLED, each = length(CUTOFF_COMBOS)),
    "_combo", rep(seq_along(CUTOFF_COMBOS), times = length(NSAMP_CUTS_POOLED))
  )
  results
}

pub_data_chisq        <- lapply(CELL_LINES_V, run_cl_comparison_chisq)
names(pub_data_chisq) <- CELL_LINES_V
```

---

## CL18870 {.tabset}

```{r pub-chisq-cl18870-check, include=FALSE}
cl      <- "CL18870"
results <- pub_data_chisq[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled chi-squared file not found for", cl)`

### Site counts

Tile colour indicates the number of sites. **Our sites** (left): sites passing
the given FDR / rate-difference and minimum-`n_samples` combination. **Overlap
with BACS / PRAISE** (middle, right): each tile shows the overlap count plus
two percentages — *X%* is the fraction of **our** sites found in the
publication, and *Y%* is the fraction of **publication** sites recovered by our
study. Facets correspond to the minimum-`n_samples` filter (1–4).

```{r pub-chisq-cl18870-heatmap, eval=ok, fig.width=14, fig.height=9}
n_bacs_total   <- nrow(mrna_BACS_clean)
n_praise_total <- nrow(mrna_PRAISE_clean)

heatmap_data <- map_dfr(results, ~ tibble(
  fdr_cut   = .x$fdr_cut, diff_cut  = .x$diff_cut, min_nsamp = .x$min_nsamp,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(
    fdr_label   = factor(paste0("FDR<", fdr_cut),
                         levels = paste0("FDR<", c(0.001, 0.01))),
    diff_label  = factor(paste0("rate_diff>", diff_cut),
                         levels = paste0("rate_diff>", c(0.05, 0.10))),
    nsamp_label = factor(paste0("n_samples>=", min_nsamp),
                         levels = paste0("n_samples>=", NSAMP_CUTS_POOLED)),
    pct_ours_in_bacs   = if_else(n_ours > 0, round(n_bacs   / n_ours * 100, 1), 0),
    pct_ours_in_praise = if_else(n_ours > 0, round(n_praise / n_ours * 100, 1), 0),
    pct_bacs_in_ours   = round(n_bacs   / n_bacs_total   * 100, 1),
    pct_praise_in_ours = round(n_praise / n_praise_total * 100, 1)
  )

make_heatmap_publication_chisq(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication_chisq(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication_chisq(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, " pooled chi-squared site counts across FDR, rate-diff & n_samples cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

```{r pub-chisq-cl18870-pct-lineplot, eval=ok, fig.width=10, fig.height=7}
make_pct_plot_chisq(heatmap_data,
  paste(cl, "- Publication overlap % by n_samples & cutoff"))
```

### vs BACS

Scatter plots of our pooled chi-squared %pU ratio (x-axis) vs. the BACS Ψ
level (y-axis) at overlapping sites, faceted by FDR / rate-difference cutoff
combination. Points are coloured by mRNA feature (CDS / 3UTR / 5UTR).

```{r pub-chisq-cl18870-bacs, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_bacs") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs BACS Ψ level"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

Our pooled chi-squared %pU ratio vs. the PRAISE deletion ratio from
replicate 1.

```{r pub-chisq-cl18870-praise-r1, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r pub-chisq-cl18870-praise-r2, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

---

## CL18909 {.tabset}

```{r pub-chisq-cl18909-check, include=FALSE}
cl      <- "CL18909"
results <- pub_data_chisq[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled chi-squared file not found for", cl)`

### Site counts

```{r pub-chisq-cl18909-heatmap, eval=ok, fig.width=14, fig.height=9}
n_bacs_total   <- nrow(mrna_BACS_clean)
n_praise_total <- nrow(mrna_PRAISE_clean)

heatmap_data <- map_dfr(results, ~ tibble(
  fdr_cut   = .x$fdr_cut, diff_cut  = .x$diff_cut, min_nsamp = .x$min_nsamp,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(
    fdr_label   = factor(paste0("FDR<", fdr_cut),
                         levels = paste0("FDR<", c(0.001, 0.01))),
    diff_label  = factor(paste0("rate_diff>", diff_cut),
                         levels = paste0("rate_diff>", c(0.05, 0.10))),
    nsamp_label = factor(paste0("n_samples>=", min_nsamp),
                         levels = paste0("n_samples>=", NSAMP_CUTS_POOLED)),
    pct_ours_in_bacs   = if_else(n_ours > 0, round(n_bacs   / n_ours * 100, 1), 0),
    pct_ours_in_praise = if_else(n_ours > 0, round(n_praise / n_ours * 100, 1), 0),
    pct_bacs_in_ours   = round(n_bacs   / n_bacs_total   * 100, 1),
    pct_praise_in_ours = round(n_praise / n_praise_total * 100, 1)
  )

make_heatmap_publication_chisq(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication_chisq(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication_chisq(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, " pooled chi-squared site counts across FDR, rate-diff & n_samples cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

```{r pub-chisq-cl18909-pct-lineplot, eval=ok, fig.width=10, fig.height=7}
make_pct_plot_chisq(heatmap_data,
  paste(cl, "- Publication overlap % by n_samples & cutoff"))
```

### vs BACS

```{r pub-chisq-cl18909-bacs, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_bacs") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs BACS Ψ level"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

```{r pub-chisq-cl18909-praise-r1, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r pub-chisq-cl18909-praise-r2, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

---

## CL18916 {.tabset}

```{r pub-chisq-cl18916-check, include=FALSE}
cl      <- "CL18916"
results <- pub_data_chisq[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled chi-squared file not found for", cl)`

### Site counts

```{r pub-chisq-cl18916-heatmap, eval=ok, fig.width=14, fig.height=9}
n_bacs_total   <- nrow(mrna_BACS_clean)
n_praise_total <- nrow(mrna_PRAISE_clean)

heatmap_data <- map_dfr(results, ~ tibble(
  fdr_cut   = .x$fdr_cut, diff_cut  = .x$diff_cut, min_nsamp = .x$min_nsamp,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(
    fdr_label   = factor(paste0("FDR<", fdr_cut),
                         levels = paste0("FDR<", c(0.001, 0.01))),
    diff_label  = factor(paste0("rate_diff>", diff_cut),
                         levels = paste0("rate_diff>", c(0.05, 0.10))),
    nsamp_label = factor(paste0("n_samples>=", min_nsamp),
                         levels = paste0("n_samples>=", NSAMP_CUTS_POOLED)),
    pct_ours_in_bacs   = if_else(n_ours > 0, round(n_bacs   / n_ours * 100, 1), 0),
    pct_ours_in_praise = if_else(n_ours > 0, round(n_praise / n_ours * 100, 1), 0),
    pct_bacs_in_ours   = round(n_bacs   / n_bacs_total   * 100, 1),
    pct_praise_in_ours = round(n_praise / n_praise_total * 100, 1)
  )

make_heatmap_publication_chisq(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication_chisq(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication_chisq(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, " pooled chi-squared site counts across FDR, rate-diff & n_samples cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

```{r pub-chisq-cl18916-pct-lineplot, eval=ok, fig.width=10, fig.height=7}
make_pct_plot_chisq(heatmap_data,
  paste(cl, "- Publication overlap % by n_samples & cutoff"))
```

### vs BACS

```{r pub-chisq-cl18916-bacs, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_bacs") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs BACS Ψ level"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

```{r pub-chisq-cl18916-praise-r1, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r pub-chisq-cl18916-praise-r2, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

---

## CL19099 {.tabset}

```{r pub-chisq-cl19099-check, include=FALSE}
cl      <- "CL19099"
results <- pub_data_chisq[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled chi-squared file not found for", cl)`

### Site counts

```{r pub-chisq-cl19099-heatmap, eval=ok, fig.width=14, fig.height=9}
n_bacs_total   <- nrow(mrna_BACS_clean)
n_praise_total <- nrow(mrna_PRAISE_clean)

heatmap_data <- map_dfr(results, ~ tibble(
  fdr_cut   = .x$fdr_cut, diff_cut  = .x$diff_cut, min_nsamp = .x$min_nsamp,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(
    fdr_label   = factor(paste0("FDR<", fdr_cut),
                         levels = paste0("FDR<", c(0.001, 0.01))),
    diff_label  = factor(paste0("rate_diff>", diff_cut),
                         levels = paste0("rate_diff>", c(0.05, 0.10))),
    nsamp_label = factor(paste0("n_samples>=", min_nsamp),
                         levels = paste0("n_samples>=", NSAMP_CUTS_POOLED)),
    pct_ours_in_bacs   = if_else(n_ours > 0, round(n_bacs   / n_ours * 100, 1), 0),
    pct_ours_in_praise = if_else(n_ours > 0, round(n_praise / n_ours * 100, 1), 0),
    pct_bacs_in_ours   = round(n_bacs   / n_bacs_total   * 100, 1),
    pct_praise_in_ours = round(n_praise / n_praise_total * 100, 1)
  )

make_heatmap_publication_chisq(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication_chisq(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication_chisq(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, " pooled chi-squared site counts across FDR, rate-diff & n_samples cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

```{r pub-chisq-cl19099-pct-lineplot, eval=ok, fig.width=10, fig.height=7}
make_pct_plot_chisq(heatmap_data,
  paste(cl, "- Publication overlap % by n_samples & cutoff"))
```

### vs BACS

```{r pub-chisq-cl19099-bacs, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_bacs") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs BACS Ψ level"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

```{r pub-chisq-cl19099-praise-r1, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r pub-chisq-cl19099-praise-r2, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```
