---
title: "Quality control using sites called from rRNA — chi-squared strategy"
author: "XSun"
date: "2026-06-06"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

# Introduction

This report mirrors [qc_rRNA.html](qc_rRNA.html) but evaluates an **alternative
site-calling strategy** based on a per-site contingency-table (chi-squared)
test rather than simple ratio/depth thresholds.

For every uridine (T) position we build a 2x2 table comparing the treated and
(pooled) untreated samples:

|            | Conversion (C) | No conversion (T) |
|------------|----------------|-------------------|
| Treated    | a              | b                 |
| Untreated  | c              | d                 |

and test the null hypothesis that the conversion rate is the same in both
samples ($X^2 = \frac{n(ad-bc)^2}{(a+b)(c+d)(a+c)(b+d)}$, 1 d.f.). A site is
called pseudouridylated when:

1. the chi-squared Benjamini-Hochberg **FDR < cutoff**, and
2. the **conversion-rate difference** (treated − untreated) **> cutoff**.

<!-- This complements the original ratio/depth-based filter: instead of asking -->
<!-- "is the treated ratio high and the input ratio low?" it asks "is the -->
<!-- treated-vs-untreated difference at this position larger than expected by -->
<!-- chance, given the sequencing depth?" — a test that naturally accounts for -->
<!-- coverage when judging significance. -->

We ran this strategy on the same four cell lines: **CL18870**, **CL18909**,
**CL18916**, and **CL19099** (four treated replicates each).

We compare our results against two published datasets:

- **BACS** — PMID 39349603
- **PRAISE** — PMID 36997645

```{r setup, warning=FALSE, message=FALSE}
library(dplyr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(purrr)
library(patchwork)
library(tidyverse)
library(data.table)
library(scales)
library(knitr)
library(gridExtra)
library(DT)
library(grid)

knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center"
)
```

---

# Examining replicates for each cell line

For each cell line we examine how consistently the **chi-squared site-calling
strategy** identifies pseudouridylated rRNA positions across its four treated
replicates, using the per-replicate filtered chi-squared output
(`out_chisq/filter.<sample>.rRNA.pseusite.txt.gz`), the same per-site table
that feeds into the pooling step described below. Each row of this table
already carries the chi-squared statistics (`pval`, `fdr`, `rate_diff`)
alongside the treated conversion ratio (`ratio`) for every candidate site.

Because "detection" under this strategy is a statistical call rather than a
simple ratio/depth threshold, we repeat each check at **four cutoff
combinations** spanning two stringency dimensions:

| Combination | Max FDR | Min rate difference |
|-------------|---------|----------------------|
| 1           | < 0.001 | > 0.05               |
| 2           | < 0.001 | > 0.10               |
| 3           | < 0.01  | > 0.05               |
| 4           | < 0.01  | > 0.10               |

For each combination we run four checks:

1. **Site overlap (Jaccard index)** — how many sites called significant are
   shared between each pair of replicates, expressed as a set-similarity score.
2. **%pU distribution** — histogram and density of the (treated) conversion
   ratio at sites called significant in each replicate, used to judge overall
   signal strength and consistency.
3. **Pairwise Pearson correlation** — quantitative agreement of %pU at sites
   called significant in **both** replicates of a pair.
4. **Pairwise Pearson heatmap** — compact summary of all pairwise correlations.

```{r config-replicate-chisq}
RESULTS_DIR <- "/project2/xinhe/xsun/pumseq/pumseq_results"

CELL_LINES <- list(
  CL18870 = list(
    rep1 = "CL18870_12-187-59",
    rep2 = "CL18870_12-187-60",
    rep3 = "CL18870_12-187-97",
    rep4 = "CL18870_12-187-98"
  ),
  CL18909 = list(
    rep1 = "CL18909_12-187-13",
    rep2 = "CL18909_12-187-14",
    rep3 = "CL18909_13-32-23",
    rep4 = "CL18909_13-32-24"
  ),
  CL18916 = list(
    rep1 = "CL18916_12-187-61",
    rep2 = "CL18916_12-187-62",
    rep3 = "CL18916_12-187-101",
    rep4 = "CL18916_12-187-102"
  ),
  CL19099 = list(
    rep1 = "CL19099_12-187-65",
    rep2 = "CL19099_12-187-66",
    rep3 = "CL19099_12-187-105",
    rep4 = "CL19099_12-187-106"
  )
)

# Cutoff combinations applied to each replicate's per-site chi-squared output
CUTOFF_COMBOS <- list(
  list(fdr = 0.001, diff = 0.05, label = "FDR < 0.001 & rate_diff > 0.05"),
  list(fdr = 0.001, diff = 0.10, label = "FDR < 0.001 & rate_diff > 0.10"),
  list(fdr = 0.01,  diff = 0.05, label = "FDR < 0.01 & rate_diff > 0.05"),
  list(fdr = 0.01,  diff = 0.10, label = "FDR < 0.01 & rate_diff > 0.10")
)
```

```{r helpers-replicate-chisq}
# ── Data loading ──────────────────────────────────────────────────────────────
read_rRNA_chisq_replicate <- function(cl, sample_id, chisq_dir) {
  f <- file.path(chisq_dir, sprintf("filter.%s.rRNA.pseusite.txt.gz", sample_id))
  if (!file.exists(f)) {
    warning("File not found: ", f)
    return(NULL)
  }
  dt <- fread(cmd = paste("zcat", f), header = TRUE, sep = "\t",
              colClasses = list(character = "ref"), data.table = FALSE)
  dt %>%
    mutate(site = paste(ref, pos, strand, sep = ":")) %>%
    select(site, ratio, fdr, rate_diff)
}

sites_passing <- function(dt, combo) dt$site[dt$fdr < combo$fdr & dt$rate_diff > combo$diff]

shared_passing_pair <- function(rep_tabs, p, combo) {
  d1 <- rep_tabs[[p[1]]]; d2 <- rep_tabs[[p[2]]]
  s1 <- sites_passing(d1, combo); s2 <- sites_passing(d2, combo)
  shared <- intersect(s1, s2)
  if (length(shared) < 5) return(NULL)
  merge(d1[d1$site %in% shared, c("site", "ratio")],
        d2[d2$site %in% shared, c("site", "ratio")],
        by = "site", suffixes = c("_1", "_2"))
}

# ── Distribution plots ────────────────────────────────────────────────────────
plot_distribution_chisq <- function(rep_tabs, rep_names, cl, combo) {
  lapply(rep_names, function(samp) {
    dt <- rep_tabs[[samp]]
    df <- dt[dt$fdr < combo$fdr & dt$rate_diff > combo$diff, , drop = FALSE]
    if (nrow(df) == 0) return(NULL)

    med_val  <- median(df$ratio)
    mean_val <- mean(df$ratio)

    ggplot(df, aes(x = ratio)) +
      geom_histogram(aes(y = after_stat(density)),
                     binwidth = 0.02, fill = "steelblue",
                     color = "white", linewidth = 0.2, alpha = 0.7) +
      geom_density(color = "navy", linewidth = 0.7) +
      geom_vline(xintercept = med_val,  color = "red",
                 linetype = "dashed",  linewidth = 0.6) +
      geom_vline(xintercept = mean_val, color = "darkorange",
                 linetype = "dotted",  linewidth = 0.6) +
      annotate("text", x = Inf, y = Inf,
               label = sprintf("n = %d\nmedian = %.3f\nmean = %.3f",
                               nrow(df), med_val, mean_val),
               hjust = 1.05, vjust = 1.05, size = 3, family = "mono") +
      scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
      labs(
        title    = samp,
        subtitle = combo$label,
        x        = "%pU  (C / C+T, treated)",
        y        = "Density"
      ) +
      theme_bw(base_size = 10) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  })
}

# ── Scatter plots ─────────────────────────────────────────────────────────────
make_scatter_plots_chisq <- function(rep_tabs, pairs, cl, combo) {
  lapply(pairs, function(p) {
    m <- shared_passing_pair(rep_tabs, p, combo)
    if (is.null(m)) return(NULL)

    r <- cor(m$ratio_1, m$ratio_2, method = "pearson")
    ggplot(m, aes(x = ratio_1, y = ratio_2)) +
      geom_point(alpha = 0.4, size = 0.8) +
      geom_abline(slope = 1, intercept = 0,
                  color = "red", linetype = "dashed") +
      annotate("text", x = Inf, y = -Inf,
               label = sprintf("r = %.3f\nn = %d", r, nrow(m)),
               hjust = 1.1, vjust = -0.3, size = 3.5) +
      labs(
        title = sprintf("%s vs %s\n[%s]", p[1], p[2], combo$label),
        x     = sprintf("%%pU  %s", p[1]),
        y     = sprintf("%%pU  %s", p[2])
      ) +
      theme_bw(base_size = 10)
  })
}

# ── Arrange + render a list of plots ─────────────────────────────────────────
render_plots_chisq <- function(plot_list, ncol = 3) {
  plot_list <- Filter(Negate(is.null), plot_list)
  if (length(plot_list) == 0) {
    message("No plots to render.")
    return(invisible(NULL))
  }
  grid.arrange(grobs = plot_list, ncol = min(ncol, length(plot_list)))
}

# ── Pearson correlation heatmap ───────────────────────────────────────────────
make_cor_heatmap_chisq <- function(rep_tabs, pairs, rep_names, cl, combo) {
  cor_df <- do.call(rbind, lapply(pairs, function(p) {
    m <- shared_passing_pair(rep_tabs, p, combo)
    r <- if (!is.null(m)) round(cor(m$ratio_1, m$ratio_2, method = "pearson"), 4) else NA_real_
    n <- if (!is.null(m)) nrow(m) else 0L
    data.frame(rep1 = p[1], rep2 = p[2], r = r, n = n, stringsAsFactors = FALSE)
  }))

  mat <- matrix(NA_real_, length(rep_names), length(rep_names),
                dimnames = list(rep_names, rep_names))
  diag(mat) <- 1
  n_mat <- matrix(NA_integer_, length(rep_names), length(rep_names),
                  dimnames = list(rep_names, rep_names))
  for (i in seq_len(nrow(cor_df))) {
    mat[cor_df$rep1[i], cor_df$rep2[i]] <- cor_df$r[i]
    mat[cor_df$rep2[i], cor_df$rep1[i]] <- cor_df$r[i]
    n_mat[cor_df$rep1[i], cor_df$rep2[i]] <- cor_df$n[i]
    n_mat[cor_df$rep2[i], cor_df$rep1[i]] <- cor_df$n[i]
  }

  short <- function(s) sub("^[^_]+_", "", s)
  lev   <- short(rep_names)

  plot_df <- expand.grid(x = rep_names, y = rep_names,
                         stringsAsFactors = FALSE) |>
    rowwise() |>
    mutate(
      r     = mat[x, y],
      n     = n_mat[x, y],
      label = case_when(
        x == y   ~ "1.000",
        is.na(r) ~ "NA",
        TRUE     ~ sprintf("%.3f\n(n=%d)", r, n)
      )
    ) |>
    ungroup() |>
    mutate(
      x_lab = factor(short(x), levels = lev),
      y_lab = factor(short(y), levels = rev(lev))
    )

  ggplot(plot_df, aes(x = x_lab, y = y_lab, fill = r)) +
    geom_tile(color = "white", linewidth = 0.8) +
    geom_text(aes(label = label), size = 3, lineheight = 0.9) +
    scale_fill_gradient2(
      low  = "#1a9641", mid = "#ffffbf", high = "#d7191c",
      midpoint = 0.5, limits = c(0, 1), na.value = "grey90",
      name = "Pearson r"
    ) +
    labs(title = cl, subtitle = combo$label, x = NULL, y = NULL) +
    coord_fixed() +
    theme_bw(base_size = 10) +
    theme(
      axis.text.x       = element_text(angle = 35, hjust = 1),
      panel.grid        = element_blank(),
      panel.border      = element_blank(),
      legend.key.height = unit(0.8, "cm")
    )
}

# ── Jaccard index helpers ─────────────────────────────────────────────────────
jaccard_index <- function(a, b) {
  n_int <- length(intersect(a, b))
  n_uni <- length(union(a, b))
  if (n_uni == 0) return(NA_real_)
  n_int / n_uni
}

make_jaccard_heatmap_chisq <- function(rep_tabs, pairs, rep_names, cl, combo) {
  site_sets <- lapply(rep_tabs, sites_passing, combo = combo)

  mat <- matrix(NA_real_, length(rep_names), length(rep_names),
                dimnames = list(rep_names, rep_names))
  diag(mat) <- 1
  n_int_mat <- matrix(NA_integer_, length(rep_names), length(rep_names),
                      dimnames = list(rep_names, rep_names))
  for (p in pairs) {
    sa <- site_sets[[p[1]]]; sb <- site_sets[[p[2]]]
    j   <- jaccard_index(sa, sb)
    n_i <- length(intersect(sa, sb))
    mat[p[1], p[2]]       <- j;   mat[p[2], p[1]]       <- j
    n_int_mat[p[1], p[2]] <- n_i; n_int_mat[p[2], p[1]] <- n_i
  }

  short <- function(s) sub("^[^_]+_", "", s)
  lev   <- short(rep_names)

  plot_df <- expand.grid(x = rep_names, y = rep_names,
                         stringsAsFactors = FALSE) |>
    rowwise() |>
    mutate(
      j     = mat[x, y],
      n_int = n_int_mat[x, y],
      label = case_when(
        x == y   ~ "1.000",
        is.na(j) ~ "NA",
        TRUE     ~ sprintf("%.3f\n(n=%d)", j, n_int)
      )
    ) |>
    ungroup() |>
    mutate(
      x_lab = factor(short(x), levels = lev),
      y_lab = factor(short(y), levels = rev(lev))
    )

  ggplot(plot_df, aes(x = x_lab, y = y_lab, fill = j)) +
    geom_tile(color = "white", linewidth = 0.8) +
    geom_text(aes(label = label), size = 3, lineheight = 0.9) +
    scale_fill_gradient2(
      low      = "#1a9641", mid = "#ffffbf", high = "#d7191c",
      midpoint = 0.5, limits = c(0, 1), na.value = "grey90",
      name     = "Jaccard"
    ) +
    labs(title = cl, subtitle = combo$label, x = NULL, y = NULL) +
    coord_fixed() +
    theme_bw(base_size = 10) +
    theme(
      axis.text.x       = element_text(angle = 35, hjust = 1),
      panel.grid        = element_blank(),
      panel.border      = element_blank(),
      legend.key.height = unit(0.8, "cm")
    )
}
```

```{r run-all-replicate-chisq, include=FALSE}
cl_data_chisq <- list()

for (cl in names(CELL_LINES)) {
  chisq_dir  <- file.path(RESULTS_DIR, cl, "out_chisq")
  reps       <- CELL_LINES[[cl]]
  sample_ids <- unlist(reps, use.names = FALSE)

  rep_tabs <- setNames(
    lapply(sample_ids, function(sid) read_rRNA_chisq_replicate(cl, sid, chisq_dir)),
    sample_ids
  )
  rep_tabs <- Filter(Negate(is.null), rep_tabs)

  if (length(rep_tabs) < 2) {
    cl_data_chisq[[cl]] <- list(ok = FALSE, reason = "Not enough replicates found.")
    next
  }

  rep_names <- names(rep_tabs)
  cl_data_chisq[[cl]] <- list(
    ok        = TRUE,
    rep_tabs  = rep_tabs,
    rep_names = rep_names,
    pairs     = combn(rep_names, 2, simplify = FALSE)
  )
}
```

---

## CL18870 {.tabset}

```{r rep-chisq-cl18870-check, include=FALSE}
cl  <- "CL18870"
dat <- cl_data_chisq[[cl]]
```

`r if (!dat$ok) paste("**Skipped:**", dat$reason)`

### Site overlap (Jaccard index)

The **Jaccard index** measures set similarity between the sets of sites called
significant in two replicates at a given cutoff combination:

$$J(A, B) = \frac{|A \cap B|}{|A \cup B|}$$

A score of 1 means every called site appears in both replicates; 0 means no
shared sites. The four panels correspond to the four cutoff combinations
above (stricter combinations are more conservative and typically show fewer,
but more reproducible, calls). The intersection count is shown in parentheses.

```{r rep-chisq-cl18870-jaccard, eval=dat$ok, fig.width=15, fig.height=12}
render_plots_chisq(
  lapply(CUTOFF_COMBOS, function(combo)
    make_jaccard_heatmap_chisq(dat$rep_tabs, dat$pairs, dat$rep_names, cl, combo)),
  ncol = 2
)
```

### %pU Distribution — FDR < 0.001 & rate_diff > 0.05

Each panel shows the %pU (treated conversion ratio) distribution for one
replicate, restricted to sites called significant under this cutoff
combination (histogram + density overlay). The **red dashed** line marks the
median; the **orange dotted** line marks the mean. Consistently shaped
distributions across replicates indicate stable, reproducible site calling.

```{r rep-chisq-cl18870-dist1, eval=dat$ok, fig.width=11, fig.height=9}
render_plots_chisq(plot_distribution_chisq(dat$rep_tabs, dat$rep_names, cl, CUTOFF_COMBOS[[1]]), ncol = 2)
```

### %pU Distribution — FDR < 0.001 & rate_diff > 0.10

```{r rep-chisq-cl18870-dist2, eval=dat$ok, fig.width=11, fig.height=9}
render_plots_chisq(plot_distribution_chisq(dat$rep_tabs, dat$rep_names, cl, CUTOFF_COMBOS[[2]]), ncol = 2)
```

### %pU Distribution — FDR < 0.01 & rate_diff > 0.05

```{r rep-chisq-cl18870-dist3, eval=dat$ok, fig.width=11, fig.height=9}
render_plots_chisq(plot_distribution_chisq(dat$rep_tabs, dat$rep_names, cl, CUTOFF_COMBOS[[3]]), ncol = 2)
```

### %pU Distribution — FDR < 0.01 & rate_diff > 0.10

```{r rep-chisq-cl18870-dist4, eval=dat$ok, fig.width=11, fig.height=9}
render_plots_chisq(plot_distribution_chisq(dat$rep_tabs, dat$rep_names, cl, CUTOFF_COMBOS[[4]]), ncol = 2)
```

### Pairwise correlation — FDR < 0.001 & rate_diff > 0.05

Scatter plots of %pU at sites called significant in **both** replicates of
each pair under this cutoff combination. The **red dashed** diagonal is the
identity line; Pearson's r and the number of shared sites are shown in the
lower-right corner.

```{r rep-chisq-cl18870-cor1, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots_chisq(make_scatter_plots_chisq(dat$rep_tabs, dat$pairs, cl, CUTOFF_COMBOS[[1]]))
```

### Pairwise correlation — FDR < 0.001 & rate_diff > 0.10

```{r rep-chisq-cl18870-cor2, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots_chisq(make_scatter_plots_chisq(dat$rep_tabs, dat$pairs, cl, CUTOFF_COMBOS[[2]]))
```

### Pairwise correlation — FDR < 0.01 & rate_diff > 0.05

```{r rep-chisq-cl18870-cor3, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots_chisq(make_scatter_plots_chisq(dat$rep_tabs, dat$pairs, cl, CUTOFF_COMBOS[[3]]))
```

### Pairwise correlation — FDR < 0.01 & rate_diff > 0.10

```{r rep-chisq-cl18870-cor4, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots_chisq(make_scatter_plots_chisq(dat$rep_tabs, dat$pairs, cl, CUTOFF_COMBOS[[4]]))
```

### Pairwise Pearson heatmap

Compact summary of all pairwise Pearson correlations of %pU at sites called
significant in both replicates of each pair. Each cell shows Pearson's r and
the number of shared sites used; the four panels correspond to the four
cutoff combinations above.

```{r rep-chisq-cl18870-heatmap, eval=dat$ok, fig.width=15, fig.height=12}
render_plots_chisq(
  lapply(CUTOFF_COMBOS, function(combo)
    make_cor_heatmap_chisq(dat$rep_tabs, dat$pairs, dat$rep_names, cl, combo)),
  ncol = 2
)
```

---

## CL18909 {.tabset}

```{r rep-chisq-cl18909-check, include=FALSE}
cl  <- "CL18909"
dat <- cl_data_chisq[[cl]]
```

`r if (!dat$ok) paste("**Skipped:**", dat$reason)`

### Site overlap (Jaccard index)

```{r rep-chisq-cl18909-jaccard, eval=dat$ok, fig.width=15, fig.height=12}
render_plots_chisq(
  lapply(CUTOFF_COMBOS, function(combo)
    make_jaccard_heatmap_chisq(dat$rep_tabs, dat$pairs, dat$rep_names, cl, combo)),
  ncol = 2
)
```

### %pU Distribution — FDR < 0.001 & rate_diff > 0.05

```{r rep-chisq-cl18909-dist1, eval=dat$ok, fig.width=11, fig.height=9}
render_plots_chisq(plot_distribution_chisq(dat$rep_tabs, dat$rep_names, cl, CUTOFF_COMBOS[[1]]), ncol = 2)
```

### %pU Distribution — FDR < 0.001 & rate_diff > 0.10

```{r rep-chisq-cl18909-dist2, eval=dat$ok, fig.width=11, fig.height=9}
render_plots_chisq(plot_distribution_chisq(dat$rep_tabs, dat$rep_names, cl, CUTOFF_COMBOS[[2]]), ncol = 2)
```

### %pU Distribution — FDR < 0.01 & rate_diff > 0.05

```{r rep-chisq-cl18909-dist3, eval=dat$ok, fig.width=11, fig.height=9}
render_plots_chisq(plot_distribution_chisq(dat$rep_tabs, dat$rep_names, cl, CUTOFF_COMBOS[[3]]), ncol = 2)
```

### %pU Distribution — FDR < 0.01 & rate_diff > 0.10

```{r rep-chisq-cl18909-dist4, eval=dat$ok, fig.width=11, fig.height=9}
render_plots_chisq(plot_distribution_chisq(dat$rep_tabs, dat$rep_names, cl, CUTOFF_COMBOS[[4]]), ncol = 2)
```

### Pairwise correlation — FDR < 0.001 & rate_diff > 0.05

```{r rep-chisq-cl18909-cor1, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots_chisq(make_scatter_plots_chisq(dat$rep_tabs, dat$pairs, cl, CUTOFF_COMBOS[[1]]))
```

### Pairwise correlation — FDR < 0.001 & rate_diff > 0.10

```{r rep-chisq-cl18909-cor2, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots_chisq(make_scatter_plots_chisq(dat$rep_tabs, dat$pairs, cl, CUTOFF_COMBOS[[2]]))
```

### Pairwise correlation — FDR < 0.01 & rate_diff > 0.05

```{r rep-chisq-cl18909-cor3, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots_chisq(make_scatter_plots_chisq(dat$rep_tabs, dat$pairs, cl, CUTOFF_COMBOS[[3]]))
```

### Pairwise correlation — FDR < 0.01 & rate_diff > 0.10

```{r rep-chisq-cl18909-cor4, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots_chisq(make_scatter_plots_chisq(dat$rep_tabs, dat$pairs, cl, CUTOFF_COMBOS[[4]]))
```

### Pairwise Pearson heatmap

```{r rep-chisq-cl18909-heatmap, eval=dat$ok, fig.width=15, fig.height=12}
render_plots_chisq(
  lapply(CUTOFF_COMBOS, function(combo)
    make_cor_heatmap_chisq(dat$rep_tabs, dat$pairs, dat$rep_names, cl, combo)),
  ncol = 2
)
```

---

## CL18916 {.tabset}

```{r rep-chisq-cl18916-check, include=FALSE}
cl  <- "CL18916"
dat <- cl_data_chisq[[cl]]
```

`r if (!dat$ok) paste("**Skipped:**", dat$reason)`

### Site overlap (Jaccard index)

```{r rep-chisq-cl18916-jaccard, eval=dat$ok, fig.width=15, fig.height=12}
render_plots_chisq(
  lapply(CUTOFF_COMBOS, function(combo)
    make_jaccard_heatmap_chisq(dat$rep_tabs, dat$pairs, dat$rep_names, cl, combo)),
  ncol = 2
)
```

### %pU Distribution — FDR < 0.001 & rate_diff > 0.05

```{r rep-chisq-cl18916-dist1, eval=dat$ok, fig.width=11, fig.height=9}
render_plots_chisq(plot_distribution_chisq(dat$rep_tabs, dat$rep_names, cl, CUTOFF_COMBOS[[1]]), ncol = 2)
```

### %pU Distribution — FDR < 0.001 & rate_diff > 0.10

```{r rep-chisq-cl18916-dist2, eval=dat$ok, fig.width=11, fig.height=9}
render_plots_chisq(plot_distribution_chisq(dat$rep_tabs, dat$rep_names, cl, CUTOFF_COMBOS[[2]]), ncol = 2)
```

### %pU Distribution — FDR < 0.01 & rate_diff > 0.05

```{r rep-chisq-cl18916-dist3, eval=dat$ok, fig.width=11, fig.height=9}
render_plots_chisq(plot_distribution_chisq(dat$rep_tabs, dat$rep_names, cl, CUTOFF_COMBOS[[3]]), ncol = 2)
```

### %pU Distribution — FDR < 0.01 & rate_diff > 0.10

```{r rep-chisq-cl18916-dist4, eval=dat$ok, fig.width=11, fig.height=9}
render_plots_chisq(plot_distribution_chisq(dat$rep_tabs, dat$rep_names, cl, CUTOFF_COMBOS[[4]]), ncol = 2)
```

### Pairwise correlation — FDR < 0.001 & rate_diff > 0.05

```{r rep-chisq-cl18916-cor1, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots_chisq(make_scatter_plots_chisq(dat$rep_tabs, dat$pairs, cl, CUTOFF_COMBOS[[1]]))
```

### Pairwise correlation — FDR < 0.001 & rate_diff > 0.10

```{r rep-chisq-cl18916-cor2, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots_chisq(make_scatter_plots_chisq(dat$rep_tabs, dat$pairs, cl, CUTOFF_COMBOS[[2]]))
```

### Pairwise correlation — FDR < 0.01 & rate_diff > 0.05

```{r rep-chisq-cl18916-cor3, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots_chisq(make_scatter_plots_chisq(dat$rep_tabs, dat$pairs, cl, CUTOFF_COMBOS[[3]]))
```

### Pairwise correlation — FDR < 0.01 & rate_diff > 0.10

```{r rep-chisq-cl18916-cor4, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots_chisq(make_scatter_plots_chisq(dat$rep_tabs, dat$pairs, cl, CUTOFF_COMBOS[[4]]))
```

### Pairwise Pearson heatmap

```{r rep-chisq-cl18916-heatmap, eval=dat$ok, fig.width=15, fig.height=12}
render_plots_chisq(
  lapply(CUTOFF_COMBOS, function(combo)
    make_cor_heatmap_chisq(dat$rep_tabs, dat$pairs, dat$rep_names, cl, combo)),
  ncol = 2
)
```

---

## CL19099 {.tabset}

```{r rep-chisq-cl19099-check, include=FALSE}
cl  <- "CL19099"
dat <- cl_data_chisq[[cl]]
```

`r if (!dat$ok) paste("**Skipped:**", dat$reason)`

### Site overlap (Jaccard index)

```{r rep-chisq-cl19099-jaccard, eval=dat$ok, fig.width=15, fig.height=12}
render_plots_chisq(
  lapply(CUTOFF_COMBOS, function(combo)
    make_jaccard_heatmap_chisq(dat$rep_tabs, dat$pairs, dat$rep_names, cl, combo)),
  ncol = 2
)
```

### %pU Distribution — FDR < 0.001 & rate_diff > 0.05

```{r rep-chisq-cl19099-dist1, eval=dat$ok, fig.width=11, fig.height=9}
render_plots_chisq(plot_distribution_chisq(dat$rep_tabs, dat$rep_names, cl, CUTOFF_COMBOS[[1]]), ncol = 2)
```

### %pU Distribution — FDR < 0.001 & rate_diff > 0.10

```{r rep-chisq-cl19099-dist2, eval=dat$ok, fig.width=11, fig.height=9}
render_plots_chisq(plot_distribution_chisq(dat$rep_tabs, dat$rep_names, cl, CUTOFF_COMBOS[[2]]), ncol = 2)
```

### %pU Distribution — FDR < 0.01 & rate_diff > 0.05

```{r rep-chisq-cl19099-dist3, eval=dat$ok, fig.width=11, fig.height=9}
render_plots_chisq(plot_distribution_chisq(dat$rep_tabs, dat$rep_names, cl, CUTOFF_COMBOS[[3]]), ncol = 2)
```

### %pU Distribution — FDR < 0.01 & rate_diff > 0.10

```{r rep-chisq-cl19099-dist4, eval=dat$ok, fig.width=11, fig.height=9}
render_plots_chisq(plot_distribution_chisq(dat$rep_tabs, dat$rep_names, cl, CUTOFF_COMBOS[[4]]), ncol = 2)
```

### Pairwise correlation — FDR < 0.001 & rate_diff > 0.05

```{r rep-chisq-cl19099-cor1, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots_chisq(make_scatter_plots_chisq(dat$rep_tabs, dat$pairs, cl, CUTOFF_COMBOS[[1]]))
```

### Pairwise correlation — FDR < 0.001 & rate_diff > 0.10

```{r rep-chisq-cl19099-cor2, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots_chisq(make_scatter_plots_chisq(dat$rep_tabs, dat$pairs, cl, CUTOFF_COMBOS[[2]]))
```

### Pairwise correlation — FDR < 0.01 & rate_diff > 0.05

```{r rep-chisq-cl19099-cor3, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots_chisq(make_scatter_plots_chisq(dat$rep_tabs, dat$pairs, cl, CUTOFF_COMBOS[[3]]))
```

### Pairwise correlation — FDR < 0.01 & rate_diff > 0.10

```{r rep-chisq-cl19099-cor4, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots_chisq(make_scatter_plots_chisq(dat$rep_tabs, dat$pairs, cl, CUTOFF_COMBOS[[4]]))
```

### Pairwise Pearson heatmap

```{r rep-chisq-cl19099-heatmap, eval=dat$ok, fig.width=15, fig.height=12}
render_plots_chisq(
  lapply(CUTOFF_COMBOS, function(combo)
    make_cor_heatmap_chisq(dat$rep_tabs, dat$pairs, dat$rep_names, cl, combo)),
  ncol = 2
)
```

---

# Comparing pooled data among cell lines

After per-replicate chi-squared filtering, sites for each cell line are
**pooled** (`pooled_sites_chisq/pooled_rRNA.txt`): treated and untreated
counts are summed across the four replicates and the chi-squared test is
re-run on the pooled contingency table, yielding pooled `pval`, `fdr`, and
`rate_diff` statistics alongside the pooled %pU ratio. Each row also records
`n_samples`, the number of individual replicates in which the site was
detected.

We explore the effect of two filtering dimensions simultaneously, using the
same four FDR / rate-difference combinations introduced above together with a
minimum-replicate filter:

| Dimension | Values |
|-----------|--------|
| FDR / rate-difference combination | the four combinations above |
| Minimum number of replicates (`n_samples`) | 1, 2, 3, 4 |

Stricter combinations (lower FDR, larger rate difference) reduce noise at the
cost of fewer sites; requiring detection in multiple replicates is a strong
reproducibility filter.

```{r setup-pooled-chisq, message=FALSE, warning=FALSE}
CELL_LINES_V      <- names(CELL_LINES)
NSAMP_CUTS_POOLED <- 1:4
COMBO_LABELS      <- vapply(CUTOFF_COMBOS, `[[`, character(1), "label")

load_pooled_chisq_rRNA <- function(cl) {
  f <- file.path(RESULTS_DIR, cl, "pooled_sites_chisq", "pooled_rRNA.txt")
  dt <- fread(f, header = TRUE, sep = "\t",
              colClasses = list(character = "ref"))
  dt[, cell_line := cl]
  dt[, site := paste(ref, pos, strand, sep = ":")]
  dt
}

count_sites_chisq <- function(dt, combo, min_nsamp) {
  sum(dt$fdr < combo$fdr & dt$rate_diff > combo$diff & dt$n_samples >= min_nsamp)
}

build_count_grid_chisq <- function(dt) {
  do.call(rbind, lapply(seq_along(CUTOFF_COMBOS), function(i) {
    combo <- CUTOFF_COMBOS[[i]]
    data.frame(
      combo_label = combo$label,
      min_nsamp   = NSAMP_CUTS_POOLED,
      n_sites     = vapply(NSAMP_CUTS_POOLED,
                           function(n) count_sites_chisq(dt, combo, n),
                           numeric(1))
    )
  })) |>
    mutate(
      combo_label = factor(combo_label, levels = COMBO_LABELS),
      nsamp_label = factor(paste0("n_samples>=", min_nsamp),
                           levels = paste0("n_samples>=", NSAMP_CUTS_POOLED))
    )
}

plot_count_heatmap_chisq <- function(grid, cl_title) {
  ggplot(grid, aes(x = combo_label, y = nsamp_label, fill = n_sites)) +
    geom_tile(color = "white", linewidth = 0.6) +
    geom_text(aes(label = comma(n_sites)), size = 4, fontface = "bold") +
    scale_fill_gradient(low = "#deebf7", high = "#2171b5",
                        name = "# sites", labels = comma) +
    labs(title = cl_title,
         x = "FDR / rate-difference cutoff (pooled)",
         y = "Minimum number of replicates") +
    theme_bw(base_size = 11) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}

dat_list_pooled_chisq        <- lapply(CELL_LINES_V, load_pooled_chisq_rRNA)
names(dat_list_pooled_chisq) <- CELL_LINES_V

grid_list_pooled_chisq        <- lapply(dat_list_pooled_chisq, build_count_grid_chisq)
names(grid_list_pooled_chisq) <- CELL_LINES_V

cl_pairs_v_chisq <- combn(CELL_LINES_V, 2, simplify = FALSE)
```

## Per cell-line site counts at different cutoffs

Each heatmap shows the number of pooled rRNA sites surviving a combination of
**FDR / rate-difference cutoff** (x-axis, the four combinations introduced
above) and **minimum number of replicates** detecting the site (y-axis).
Darker blue = more sites.

---

### CL18870

```{r pooled-chisq-cl18870-count-heatmap, fig.width=8, fig.height=4}
plot_count_heatmap_chisq(grid_list_pooled_chisq[["CL18870"]], "CL18870 - pooled chi-squared site counts")
```

---

### CL18909

```{r pooled-chisq-cl18909-count-heatmap, fig.width=8, fig.height=4}
plot_count_heatmap_chisq(grid_list_pooled_chisq[["CL18909"]], "CL18909 - pooled chi-squared site counts")
```

---

### CL18916

```{r pooled-chisq-cl18916-count-heatmap, fig.width=8, fig.height=4}
plot_count_heatmap_chisq(grid_list_pooled_chisq[["CL18916"]], "CL18916 - pooled chi-squared site counts")
```

---

### CL19099

```{r pooled-chisq-cl19099-count-heatmap, fig.width=8, fig.height=4}
plot_count_heatmap_chisq(grid_list_pooled_chisq[["CL19099"]], "CL19099 - pooled chi-squared site counts")
```

---

### %pU Distribution — pooled data {.tabset}

Distribution of the pooled %ψ ratio (C/(C+T)) at sites passing each FDR /
rate-difference combination, shown for all four cell lines side by side
(mirroring the per-replicate distributions above). Red dashed = median;
orange dotted = mean.

```{r pooled-chisq-dist-helper, include=FALSE}
plot_pooled_dist_chisq <- function(cl, combo) {
  dt <- dat_list_pooled_chisq[[cl]]
  df <- as.data.frame(dt[fdr < combo$fdr & rate_diff > combo$diff, .(psi = ratio)])
  if (nrow(df) == 0) return(NULL)

  med_val  <- median(df$psi)
  mean_val <- mean(df$psi)

  ggplot(df, aes(x = psi)) +
    geom_histogram(aes(y = after_stat(density)),
                   binwidth = 0.02, fill = "steelblue",
                   color = "white", linewidth = 0.2, alpha = 0.7) +
    geom_density(color = "navy", linewidth = 0.7) +
    geom_vline(xintercept = med_val,  color = "red",
               linetype = "dashed",  linewidth = 0.6) +
    geom_vline(xintercept = mean_val, color = "darkorange",
               linetype = "dotted",  linewidth = 0.6) +
    annotate("text", x = Inf, y = Inf,
             label = sprintf("n = %d\nmedian = %.3f\nmean = %.3f",
                             nrow(df), med_val, mean_val),
             hjust = 1.05, vjust = 1.05, size = 3, family = "mono") +
    scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
    labs(title = cl, subtitle = combo$label,
         x = "%pU  (C / C+T, pooled)", y = "Density") +
    theme_bw(base_size = 10) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

#### %pU Distribution — FDR < 0.001 & rate_diff > 0.05

```{r pooled-chisq-dist1, fig.width=14, fig.height=3.5}
plots_pdist1 <- lapply(CELL_LINES_V, plot_pooled_dist_chisq, combo = CUTOFF_COMBOS[[1]])
do.call(grid.arrange, c(Filter(Negate(is.null), plots_pdist1), list(ncol = 4)))
```

#### %pU Distribution — FDR < 0.001 & rate_diff > 0.10

```{r pooled-chisq-dist2, fig.width=14, fig.height=3.5}
plots_pdist2 <- lapply(CELL_LINES_V, plot_pooled_dist_chisq, combo = CUTOFF_COMBOS[[2]])
do.call(grid.arrange, c(Filter(Negate(is.null), plots_pdist2), list(ncol = 4)))
```

#### %pU Distribution — FDR < 0.01 & rate_diff > 0.05

```{r pooled-chisq-dist3, fig.width=14, fig.height=3.5}
plots_pdist3 <- lapply(CELL_LINES_V, plot_pooled_dist_chisq, combo = CUTOFF_COMBOS[[3]])
do.call(grid.arrange, c(Filter(Negate(is.null), plots_pdist3), list(ncol = 4)))
```

#### %pU Distribution — FDR < 0.01 & rate_diff > 0.10

```{r pooled-chisq-dist4, fig.width=14, fig.height=3.5}
plots_pdist4 <- lapply(CELL_LINES_V, plot_pooled_dist_chisq, combo = CUTOFF_COMBOS[[4]])
do.call(grid.arrange, c(Filter(Negate(is.null), plots_pdist4), list(ncol = 4)))
```

---

## Cross-cell-line comparison of site counts

We use the **Jaccard index** to quantify how much the pooled, significance
-filtered site sets overlap between every pair of cell lines. For each
combination of FDR / rate-difference cutoff and minimum-replicate filter,
sites passing both filters in each cell line are extracted and the pairwise
Jaccard is computed:

$$J(A, B) = \frac{|A \cap B|}{|A \cup B|}$$

A high Jaccard index means the two cell lines call largely the same pooled
rRNA positions, consistent with constitutively modified sites shared across
cell lines. Each cell shows the Jaccard score and, in parentheses, the number
of sites in the intersection. Rows = minimum n_samples; columns = FDR /
rate-difference combination.

```{r pooled-chisq-cross-cl-comparison, fig.width=14, fig.height=11}
jaccard_pooled_rows <- do.call(rbind, lapply(cl_pairs_v_chisq, function(p) {
  do.call(rbind, lapply(seq_along(CUTOFF_COMBOS), function(i) {
    combo <- CUTOFF_COMBOS[[i]]
    do.call(rbind, lapply(NSAMP_CUTS_POOLED, function(n) {
      dt1 <- dat_list_pooled_chisq[[p[1]]]
      dt2 <- dat_list_pooled_chisq[[p[2]]]
      sites1 <- dt1$site[dt1$fdr < combo$fdr & dt1$rate_diff > combo$diff & dt1$n_samples >= n]
      sites2 <- dt2$site[dt2$fdr < combo$fdr & dt2$rate_diff > combo$diff & dt2$n_samples >= n]
      n_int  <- length(intersect(sites1, sites2))
      n_uni  <- length(union(sites1, sites2))
      j      <- if (n_uni == 0) NA_real_ else n_int / n_uni
      data.frame(cl1 = p[1], cl2 = p[2],
                 combo_idx = i, combo_label = combo$label, min_nsamp = n,
                 jaccard = round(j, 3), n_intersect = n_int)
    }))
  }))
}))

jaccard_pooled_sym <- jaccard_pooled_rows |>
  bind_rows(jaccard_pooled_rows |> rename(cl1 = cl2, cl2 = cl1)) |>
  bind_rows(
    expand.grid(cl1 = CELL_LINES_V, cl2 = CELL_LINES_V,
                combo_idx = seq_along(CUTOFF_COMBOS),
                min_nsamp = NSAMP_CUTS_POOLED) |>
      filter(cl1 == cl2) |>
      mutate(combo_label = COMBO_LABELS[combo_idx],
             jaccard = 1, n_intersect = NA_integer_)
  ) |>
  mutate(
    cl1         = factor(cl1, levels = CELL_LINES_V),
    cl2         = factor(cl2, levels = rev(CELL_LINES_V)),
    nsamp_label = factor(paste0("n_samples>=", min_nsamp),
                         levels = paste0("n_samples>=", NSAMP_CUTS_POOLED)),
    combo_label = factor(combo_label, levels = COMBO_LABELS),
    cell_label  = case_when(
      cl1 == cl2     ~ "1.000",
      is.na(jaccard) ~ "NA",
      TRUE           ~ sprintf("%.3f\n(n=%d)", jaccard, n_intersect)
    )
  )

ggplot(jaccard_pooled_sym, aes(x = cl1, y = cl2, fill = jaccard)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = cell_label), size = 2.8, fontface = "bold",
            lineheight = 0.9) +
  scale_fill_gradient2(
    low = "#1a9641", mid = "#ffffbf", high = "#d7191c",
    midpoint = 0.5, limits = c(0, 1), na.value = "grey90",
    name = "Jaccard"
  ) +
  facet_grid(nsamp_label ~ combo_label) +
  labs(
    title = "Jaccard index of site overlap between cell lines (pooled chi-squared rRNA sites)",
    x = NULL, y = NULL
  ) +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1, size = 9),
    axis.text.y      = element_text(size = 9),
    strip.background = element_rect(fill = "#e8eaf6"),
    strip.text       = element_text(face = "bold", size = 9),
    panel.spacing    = unit(0.3, "lines"),
    legend.position  = "right"
  )
```

---

## Pearson's r of %pU between cell lines

To assess whether the same sites are similarly modified across cell lines, we
compute Pearson's r on the pooled %pU values at sites passing the FDR /
rate-difference filter and the minimum-replicate filter in **both** cell
lines of each pair. High r across cell lines suggests that the detected
modification pattern is consistent — i.e., the same rRNA sites are strongly
modified regardless of cell line — which is expected for constitutively
modified rRNA residues.

Sites that pass in one cell line but not the other are excluded from the
correlation (they contribute to the Jaccard-style overlap question, not the
quantitative agreement question).

```{r pooled-chisq-pearson-setup}
pearson_pair_chisq <- function(dt1, dt2, combo, min_nsamp) {
  pass1 <- dt1[fdr < combo$fdr & rate_diff > combo$diff & n_samples >= min_nsamp,
               .(site, ratio1 = ratio)]
  pass2 <- dt2[fdr < combo$fdr & rate_diff > combo$diff & n_samples >= min_nsamp,
               .(site, ratio2 = ratio)]
  merged <- merge(pass1, pass2, by = "site")
  if (nrow(merged) < 3) return(data.frame(r = NA_real_, n_common = nrow(merged)))
  data.frame(
    r        = cor(merged$ratio1, merged$ratio2, method = "pearson"),
    n_common = nrow(merged)
  )
}

cor_pooled_grid <- do.call(rbind, lapply(cl_pairs_v_chisq, function(p) {
  do.call(rbind, lapply(seq_along(CUTOFF_COMBOS), function(i) {
    combo <- CUTOFF_COMBOS[[i]]
    do.call(rbind, lapply(NSAMP_CUTS_POOLED, function(n) {
      res <- pearson_pair_chisq(
        dat_list_pooled_chisq[[p[1]]], dat_list_pooled_chisq[[p[2]]],
        combo = combo, min_nsamp = n
      )
      data.frame(cl1 = p[1], cl2 = p[2],
                 combo_idx = i, combo_label = combo$label, min_nsamp = n,
                 pearson_r = round(res$r, 3), n_common = res$n_common)
    }))
  }))
}))

cor_pooled_sym <- cor_pooled_grid |>
  bind_rows(
    cor_pooled_grid |> rename(cl1 = cl2, cl2 = cl1)
  ) |>
  bind_rows(
    expand.grid(cl1 = CELL_LINES_V, cl2 = CELL_LINES_V,
                combo_idx = seq_along(CUTOFF_COMBOS),
                min_nsamp = NSAMP_CUTS_POOLED) |>
      filter(cl1 == cl2) |>
      mutate(combo_label = COMBO_LABELS[combo_idx],
             pearson_r = 1, n_common = NA_integer_)
  ) |>
  mutate(
    cl1         = factor(cl1, levels = CELL_LINES_V),
    cl2         = factor(cl2, levels = rev(CELL_LINES_V)),
    nsamp_label = factor(paste0("n_samples>=", min_nsamp),
                         levels = paste0("n_samples>=", NSAMP_CUTS_POOLED)),
    combo_label = factor(combo_label, levels = COMBO_LABELS)
  )
```

### Correlation matrices

The grid of heatmaps below shows Pearson's r between every pair of cell lines
under all combinations of cutoffs. Rows = minimum n_samples; columns = FDR /
rate-difference combination. Diagonal cells equal 1 by definition. Warmer
(red) colours indicate higher agreement; a grey cell means too few common
sites to compute r.

```{r pooled-chisq-pearson-heatmap, fig.width=14, fig.height=11}
ggplot(cor_pooled_sym, aes(x = cl1, y = cl2, fill = pearson_r)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = ifelse(cl1 == cl2, "1", sprintf("%.3f", pearson_r))),
            size = 3, fontface = "bold") +
  scale_fill_gradient2(
    low = "#1a9641", mid = "#ffffbf", high = "#d7191c",
    midpoint = 0.5, limits = c(0, 1), na.value = "grey90",
    name = "Pearson r"
  ) +
  facet_grid(nsamp_label ~ combo_label) +
  labs(
    title = "Pearson's r of pooled %pU between cell lines (chi-squared rRNA sites)",
    x = NULL, y = NULL
  ) +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1, size = 9),
    axis.text.y      = element_text(size = 9),
    strip.background = element_rect(fill = "#e8eaf6"),
    strip.text       = element_text(face = "bold", size = 9),
    panel.spacing    = unit(0.3, "lines"),
    legend.position  = "right"
  )
```

### Scatter plots (FDR < 0.001 & rate_diff > 0.05, n_samples >= 2)

Representative scatter plots at a moderately stringent cutoff (FDR < 0.001 &
rate_diff > 0.05, detected in >= 2 replicates). Each point is one pooled rRNA
site.

```{r pooled-chisq-pearson-scatter, fig.width=13, fig.height=8}
SCATTER_COMBO_POOLED <- CUTOFF_COMBOS[[1]]
MIN_N_POOLED         <- 2

scatter_data_chisq <- lapply(CELL_LINES_V, function(cl) {
  dt <- dat_list_pooled_chisq[[cl]]
  dt[fdr < SCATTER_COMBO_POOLED$fdr &
       rate_diff > SCATTER_COMBO_POOLED$diff &
       n_samples >= MIN_N_POOLED,
     .(site, ratio, cell_line = cl)]
})
names(scatter_data_chisq) <- CELL_LINES_V

scatter_plots_chisq <- lapply(cl_pairs_v_chisq, function(p) {
  d1 <- scatter_data_chisq[[p[1]]]
  d2 <- scatter_data_chisq[[p[2]]]
  m  <- merge(d1, d2, by = "site", suffixes = c("_1", "_2"))
  if (nrow(m) < 3) return(NULL)
  r_val <- cor(m$ratio_1, m$ratio_2, method = "pearson")
  ggplot(m, aes(x = ratio_1, y = ratio_2)) +
    geom_point(alpha = 0.55, size = 1.2, color = "#2c7bb6") +
    geom_abline(slope = 1, intercept = 0, color = "red",
                linetype = "dashed", linewidth = 0.6) +
    annotate("text", x = Inf, y = -Inf,
             label = sprintf("r = %.3f\nn = %d", r_val, nrow(m)),
             hjust = 1.1, vjust = -0.3, size = 3.5) +
    scale_x_continuous(labels = percent_format(accuracy = 1)) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      title = paste(p[1], "vs", p[2]),
      x     = paste0("%pU  ", p[1]),
      y     = paste0("%pU  ", p[2])
    ) +
    theme_bw(base_size = 10)
})
scatter_plots_chisq <- Filter(Negate(is.null), scatter_plots_chisq)

do.call(grid.arrange, c(scatter_plots_chisq, list(ncol = 3)))
```

---

# Comparing with published data

We benchmark our **pooled chi-squared** rRNA calls against two independent
published datasets. Overlap with validated sites confirms that the
chi-squared site-calling strategy is correctly identifying true
pseudouridine modifications.

- **BACS** (PMID 39349603) detects pU via antibody-based capture; it reports
  absolute %pU levels for two replicates.
- **PRAISE** (PMID 36997645) detects pU via the deletion signature at reverse-
  transcription termination; it reports deletion ratios for two replicates.

For each cell line and each of the four FDR / rate-difference cutoff
combinations introduced above we report: (a) how many of our pooled,
significance-filtered sites overlap with BACS sites, (b) how many overlap
with PRAISE sites, and (c) scatter plots comparing the quantitative %pU or
deletion ratio values at overlapping positions.

```{r pub-chisq-helpers}
make_scatter_publication_chisq <- function(df, x, y, col, ttl, xlab, ylab) {
  if (nrow(df) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = "No overlapping sites",
                 size = 4, color = "grey50") +
        labs(title = ttl, x = xlab, y = ylab) +
        theme_classic(base_size = 11) +
        theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))
    )
  }
  ggplot(df, aes(x = .data[[x]], y = .data[[y]],
                 color = .data[[col]], label = label)) +
    geom_point(size = 2.5, alpha = 0.8) +
    geom_text_repel(size = 2.5, max.overlaps = 20) +
    geom_smooth(method = "lm", se = TRUE, color = "grey40", linetype = "dashed") +
    scale_color_manual(values = c("5.8S" = "#4E79A7", "18S" = "#F28E2B",
                                  "28S"  = "#59A14F", "5S"  = "#E15759")) +
    labs(title = ttl, x = xlab, y = ylab, color = "rRNA type") +
    theme_classic(base_size = 11) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))
}

make_heatmap_publication_chisq <- function(data, fill_col, label_col, ttl, fill_label) {
  ggplot(data, aes(x = diff_label, y = fdr_label, fill = .data[[fill_col]])) +
    geom_tile(color = "white", linewidth = 1.2) +
    geom_text(aes(label = .data[[label_col]]), size = 6, fontface = "bold", color = "white") +
    scale_fill_gradient(low = "#d0e8f2", high = "#1a6fa8") +
    labs(title = ttl, x = "Min rate-difference cutoff", y = "Max FDR cutoff", fill = fill_label) +
    theme_classic(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          axis.text  = element_text(size = 11))
}

combine_panels_chisq <- function(results, slot) {
  plots <- map(results, slot)
  wrap_plots(plots, ncol = 2) +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
}
```

## Published data overview

Site counts and the overlap between BACS and PRAISE are shown first to
establish the published reference landscape before comparing with our pooled
chi-squared calls.

```{r pub-chisq-load, warning=FALSE, message=FALSE}
path_BACS   <- "/project2/xinhe/xsun/pumseq/data_others/pU_PMID39349603.xlsx"
path_PRAISE <- "/project2/xinhe/xsun/pumseq/data_others/pU_PMID36997645.xlsx"

rrna_BACS <- readxl::read_excel(path_BACS, sheet = 2)
colnames(rrna_BACS) <- rrna_BACS[2, ]
rrna_BACS <- rrna_BACS[-c(1:2), ]
rrna_BACS <- rrna_BACS[complete.cases(rrna_BACS$chr), ]

rrna_PRAISE <- readxl::read_excel(path_PRAISE, sheet = 1)
colnames(rrna_PRAISE) <- rrna_PRAISE[2, ]
rrna_PRAISE <- rrna_PRAISE[-c(1:2), ]
rrna_PRAISE <- rrna_PRAISE %>% filter(!is.na(chr_name), !is.na(site))

rrna_BACS_clean <- rrna_BACS %>%
  mutate(
    rna_type = case_when(
      str_detect(chr, "RNA5_8S") ~ "5.8S",
      str_detect(chr, "RNA18S")  ~ "18S",
      str_detect(chr, "RNA28S")  ~ "28S",
      str_detect(chr, "RNA5SN")  ~ "5S",
      TRUE ~ NA_character_
    ),
    pos_num = as.integer(str_extract(pos, "^[0-9]+"))
  )

rrna_PRAISE_clean <- rrna_PRAISE %>%
  mutate(
    rna_type  = str_extract(chr_name, "(5\\.8S|18S|28S|5S)"),
    pos_start = as.integer(str_extract(site, "^[0-9]+")),
    pos_end   = as.integer(coalesce(
      str_extract(site, "(?<=-)[0-9]+$"),
      str_extract(site, "^[0-9]+")
    ))
  )

overlap <- rrna_BACS_clean %>%
  inner_join(rrna_PRAISE_clean, by = "rna_type") %>%
  filter(pos_num >= pos_start & pos_num <= pos_end)

DT::datatable(rrna_BACS, caption = htmltools::tags$caption(
  style = 'caption-side: left; text-align: left; color:black; font-size:150%;',
  'Sites from BACS'), options = list(pageLength = 5))

DT::datatable(rrna_PRAISE, caption = htmltools::tags$caption(
  style = 'caption-side: left; text-align: left; color:black; font-size:150%;',
  'Sites from PRAISE'), options = list(pageLength = 5))

cat("Sites in BACS:    ", nrow(rrna_BACS_clean), "\n")
cat("Sites in PRAISE:  ", nrow(rrna_PRAISE_clean), "\n")
cat("Overlapping sites:", nrow(overlap), "\n")

rrna_BACS_clean <- rrna_BACS_clean %>%
  mutate(accession = str_extract(chr, "NR_[0-9]+"),
         pos_num   = as.integer(pos)) %>%
  rename(`pU level_rep1` = `Ψ level_rep1`,
         `pU level_rep2` = `Ψ level_rep2`)

rrna_PRAISE_clean <- rrna_PRAISE_clean %>%
  mutate(accession = str_extract(chr_name, "NR_[0-9]+"))
```

```{r pub-chisq-preload, include=FALSE}
run_cl_comparison_chisq <- function(cl) {
  pooled_path <- file.path(RESULTS_DIR, cl, "pooled_sites_chisq", "pooled_rRNA.txt")
  if (!file.exists(pooled_path)) {
    warning("File not found: ", pooled_path)
    return(NULL)
  }

  pooled_ours_rrna_chisq <- fread(pooled_path)

  results <- lapply(seq_along(CUTOFF_COMBOS), function(i) {
    combo <- CUTOFF_COMBOS[[i]]
    tag   <- combo$label

    ours_filtered <- pooled_ours_rrna_chisq %>%
      filter(fdr < combo$fdr, rate_diff > combo$diff) %>%
      mutate(accession = str_extract(ref, "NR_[0-9]+"))

    overlap_bacs <- ours_filtered %>%
      inner_join(rrna_BACS_clean, by = c("accession", "pos" = "pos_num")) %>%
      mutate(
        psi_rep1 = as.numeric(`pU level_rep1`),
        psi_rep2 = as.numeric(`pU level_rep2`),
        label    = paste0(name, " pos", pos)
      )

    overlap_praise <- ours_filtered %>%
      inner_join(rrna_PRAISE_clean, by = "accession", relationship = "many-to-many") %>%
      filter(pos >= pos_start & pos <= pos_end) %>%
      mutate(
        deletion_rep1 = as.numeric(rep1_deletion_ratio),
        deletion_rep2 = as.numeric(rep2_deletion_ratio),
        label         = paste0(rna_type, " pos", pos)
      )

    list(
      tag      = tag,
      fdr_cut  = combo$fdr,
      diff_cut = combo$diff,
      n_ours   = nrow(ours_filtered),
      n_bacs   = nrow(overlap_bacs),
      n_praise = nrow(overlap_praise),
      overlap_bacs   = overlap_bacs,
      overlap_praise = overlap_praise,

      p_bacs_r1   = make_scatter_publication_chisq(overlap_bacs, "ratio", "psi_rep1", "name",
                                             tag, "Our pU ratio (pooled, chi-squared)", "BACS pU level (rep1)"),
      p_bacs_r2   = make_scatter_publication_chisq(overlap_bacs, "ratio", "psi_rep2", "name",
                                             tag, "Our pU ratio (pooled, chi-squared)", "BACS pU level (rep2)"),
      p_praise_r1 = make_scatter_publication_chisq(overlap_praise, "ratio", "deletion_rep1", "rna_type",
                                             tag, "Our pU ratio (pooled, chi-squared)", "PRAISE deletion ratio (rep1)"),
      p_praise_r2 = make_scatter_publication_chisq(overlap_praise, "ratio", "deletion_rep2", "rna_type",
                                             tag, "Our pU ratio (pooled, chi-squared)", "PRAISE deletion ratio (rep2)")
    )
  })

  names(results) <- paste0("combo", seq_along(CUTOFF_COMBOS))
  results
}

pub_data_chisq        <- lapply(CELL_LINES_V, run_cl_comparison_chisq)
names(pub_data_chisq) <- CELL_LINES_V
```

---

## CL18870 {.tabset}

```{r pub-chisq-cl18870-check, include=FALSE}
cl      <- "CL18870"
results <- pub_data_chisq[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled chi-squared file not found for", cl)`

### Site counts

Tile colour indicates the number of sites: **Our sites** (left) are those
passing the given FDR / rate-difference cutoff combination in our pooled
chi-squared data; **Overlap with BACS/PRAISE** (middle, right) counts how many
of those sites are also present in the respective published dataset. Higher
overlap validates the specificity of our chi-squared site calls.

```{r pub-chisq-cl18870-heatmap, eval=ok, fig.width=12, fig.height=4}
heatmap_data <- map_dfr(results, ~ tibble(
  fdr_cut = .x$fdr_cut, diff_cut = .x$diff_cut,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(fdr_label  = factor(paste0("FDR<", fdr_cut),
                             levels = paste0("FDR<", c(0.001, 0.01))),
         diff_label = factor(paste0("rate_diff>", diff_cut),
                             levels = paste0("rate_diff>", c(0.05, 0.10))))

make_heatmap_publication_chisq(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication_chisq(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication_chisq(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, " pooled chi-squared site counts across FDR & rate-diff cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

### vs BACS (rep1)

Scatter plots of our pooled chi-squared %pU ratio (x-axis) vs. the BACS %pU
level from replicate 1 (y-axis) at overlapping sites, faceted by FDR /
rate-difference cutoff combination. Points are coloured by rRNA subunit type.
A positive trend indicates quantitative agreement with the BACS dataset.

```{r pub-chisq-cl18870-bacs-r1, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_bacs_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs BACS pU level (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs BACS (rep2)

```{r pub-chisq-cl18870-bacs-r2, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_bacs_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs BACS pU level (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

Our pooled chi-squared %pU ratio vs. the PRAISE deletion ratio from
replicate 1. PRAISE uses a different biochemical readout, so the absolute
scale differs; we expect a monotonic positive relationship at high-confidence
sites.

```{r pub-chisq-cl18870-praise-r1, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r pub-chisq-cl18870-praise-r2, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

---

## CL18909 {.tabset}

```{r pub-chisq-cl18909-check, include=FALSE}
cl      <- "CL18909"
results <- pub_data_chisq[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled chi-squared file not found for", cl)`

### Site counts

```{r pub-chisq-cl18909-heatmap, eval=ok, fig.width=12, fig.height=4}
heatmap_data <- map_dfr(results, ~ tibble(
  fdr_cut = .x$fdr_cut, diff_cut = .x$diff_cut,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(fdr_label  = factor(paste0("FDR<", fdr_cut),
                             levels = paste0("FDR<", c(0.001, 0.01))),
         diff_label = factor(paste0("rate_diff>", diff_cut),
                             levels = paste0("rate_diff>", c(0.05, 0.10))))

make_heatmap_publication_chisq(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication_chisq(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication_chisq(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, " pooled chi-squared site counts across FDR & rate-diff cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

### vs BACS (rep1)

```{r pub-chisq-cl18909-bacs-r1, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_bacs_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs BACS pU level (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs BACS (rep2)

```{r pub-chisq-cl18909-bacs-r2, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_bacs_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs BACS pU level (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

```{r pub-chisq-cl18909-praise-r1, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r pub-chisq-cl18909-praise-r2, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

---

## CL18916 {.tabset}

```{r pub-chisq-cl18916-check, include=FALSE}
cl      <- "CL18916"
results <- pub_data_chisq[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled chi-squared file not found for", cl)`

### Site counts

```{r pub-chisq-cl18916-heatmap, eval=ok, fig.width=12, fig.height=4}
heatmap_data <- map_dfr(results, ~ tibble(
  fdr_cut = .x$fdr_cut, diff_cut = .x$diff_cut,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(fdr_label  = factor(paste0("FDR<", fdr_cut),
                             levels = paste0("FDR<", c(0.001, 0.01))),
         diff_label = factor(paste0("rate_diff>", diff_cut),
                             levels = paste0("rate_diff>", c(0.05, 0.10))))

make_heatmap_publication_chisq(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication_chisq(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication_chisq(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, " pooled chi-squared site counts across FDR & rate-diff cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

### vs BACS (rep1)

```{r pub-chisq-cl18916-bacs-r1, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_bacs_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs BACS pU level (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs BACS (rep2)

```{r pub-chisq-cl18916-bacs-r2, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_bacs_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs BACS pU level (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

```{r pub-chisq-cl18916-praise-r1, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r pub-chisq-cl18916-praise-r2, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

---

## CL19099 {.tabset}

```{r pub-chisq-cl19099-check, include=FALSE}
cl      <- "CL19099"
results <- pub_data_chisq[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled chi-squared file not found for", cl)`

### Site counts

```{r pub-chisq-cl19099-heatmap, eval=ok, fig.width=12, fig.height=4}
heatmap_data <- map_dfr(results, ~ tibble(
  fdr_cut = .x$fdr_cut, diff_cut = .x$diff_cut,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(fdr_label  = factor(paste0("FDR<", fdr_cut),
                             levels = paste0("FDR<", c(0.001, 0.01))),
         diff_label = factor(paste0("rate_diff>", diff_cut),
                             levels = paste0("rate_diff>", c(0.05, 0.10))))

make_heatmap_publication_chisq(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication_chisq(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication_chisq(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, " pooled chi-squared site counts across FDR & rate-diff cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

### vs BACS (rep1)

```{r pub-chisq-cl19099-bacs-r1, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_bacs_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs BACS pU level (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs BACS (rep2)

```{r pub-chisq-cl19099-bacs-r2, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_bacs_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs BACS pU level (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

```{r pub-chisq-cl19099-praise-r1, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r pub-chisq-cl19099-praise-r2, eval=ok, fig.width=12, fig.height=9}
combine_panels_chisq(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio (pooled, chi-squared) vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

