---
title: "Quality control using sites called from rRNA"
author: "XSun"
date: "2026-06-04"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

# Introduction

We ran the pipeline for several cell lines: CL18909, CL18870, CL18916, CL19099. 

We do quality control using the sites called from rRNA here. We compare the sites called from different cell lines and also, compare them with the published data: 

- Paper 1: PMID39349603 (BACS)
- Paper 2: PMID36997645 (PRAISE)

```{r warning=F, message=F}
library(dplyr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(ggvenn)
library(purrr)
library(patchwork)
library(tidyverse)
library(data.table)
library(scales)
library(knitr)
library(kableExtra)
library(gridExtra)
library(DT)
library(UpSetR)
library(grid)

knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center"
)

cutoffs <- expand.grid(min_depth = c(10, 20),
                       min_ratio = c(0.1, 0.2))


```

# Examing replicates for each cell line

We performed QC on rRNA pseudouridine sites (%pU = C / (C + T)) across
treated replicates for each cell line. For every cell line it shows:

1. **%pU distribution** per replicate (histogram + density)
2. **Pairwise Pearson correlation** — all sites, sites with %pU >= 0.1, and %pU >= 0.2
3. **Pairwise Pearson heatmap**  — all sites, sites with %pU >= 0.1, and %pU >= 0.2


```{r config-replicate}
RESULTS_DIR <- "/project2/xinhe/xsun/pumseq/pumseq_results"

MPILE_COLS <- c("ref", "start", "pos", "strand", "base",
                "AllDepth", "A", "C", "G", "T_count", "Gap")

CELL_LINES <- list(
  CL18870 = list(
    rep1 = "CL18870_12-187-59",
    rep2 = "CL18870_12-187-60",
    rep3 = "CL18870_12-187-97",
    rep4 = "CL18870_12-187-98"
  ),
  CL18909 = list(
    rep1 = "CL18909_12-187-13",
    rep2 = "CL18909_12-187-14",
    rep3 = "CL18909_13-32-23",
    rep4 = "CL18909_13-32-24"
  ),
  CL18916 = list(
    rep1 = "CL18916_12-187-61",
    rep2 = "CL18916_12-187-62",
    rep3 = "CL18916_12-187-101",
    rep4 = "CL18916_12-187-102"
  ),
  CL19099 = list(
    rep1 = "CL19099_12-187-65",
    rep2 = "CL19099_12-187-66",
    rep3 = "CL19099_12-187-105",
    rep4 = "CL19099_12-187-106"
  )
)

MIN_DEPTH <- 20
```


```{r helpers-replicate}
# ── Data loading ──────────────────────────────────────────────────────────────
read_rRNA_psi <- function(cl, sample_id, mpile_dir) {
  f <- file.path(mpile_dir,
                 sprintf("mpileup.%s.rRNA.pseusite.txt.gz", sample_id))
  if (!file.exists(f)) {
    warning("File not found: ", f)
    return(NULL)
  }
  dt <- fread(cmd = paste("zcat", f), header = FALSE, sep = "\t",
              col.names = MPILE_COLS, data.table = FALSE)
  dt <- dt %>%
    mutate(
      depth = C + T_count,
      psi   = ifelse(depth >= MIN_DEPTH, C / depth, NA_real_),
      site  = paste(ref, pos, sep = ":")
    ) %>%
    filter(!is.na(psi)) %>%
    select(site, psi)
  colnames(dt)[colnames(dt) == "psi"] <- sample_id
  dt
}

# ── Distribution plots ────────────────────────────────────────────────────────
plot_distribution <- function(psi_mat, rep_names, cl) {
  thresholds <- list(
    list(thr = NULL, label = "all sites",   tag = "all"),
    list(thr = 0.1,  label = "%\u03a8 \u2265 0.1", tag = "psi01"),
    list(thr = 0.2,  label = "%\u03a8 \u2265 0.2", tag = "psi02")
  )

  all_plots <- lapply(thresholds, function(th) {
    lapply(rep_names, function(samp) {
      df <- data.frame(psi = psi_mat[[samp]])
      df <- df[!is.na(df$psi), , drop = FALSE]

      # apply threshold filter
      if (!is.null(th$thr)) df <- df[df$psi >= th$thr, , drop = FALSE]

      if (nrow(df) == 0) return(NULL)

      med_val  <- median(df$psi)
      mean_val <- mean(df$psi)
      n_01     <- sum(df$psi >= 0.1)
      n_02     <- sum(df$psi >= 0.2)

      # annotation varies by threshold
      ann_text <- if (is.null(th$thr)) {
        sprintf(
          "n = %d\nmedian = %.3f\nmean = %.3f\n\u22650.1: %d (%.1f%%)\n\u22650.2: %d (%.1f%%)",
          nrow(df), med_val, mean_val,
          n_01, 100 * n_01 / nrow(df),
          n_02, 100 * n_02 / nrow(df)
        )
      } else {
        sprintf(
          "n = %d\nmedian = %.3f\nmean = %.3f",
          nrow(df), med_val, mean_val
        )
      }

      # threshold lines: only show the ones still relevant
      vlines <- if (is.null(th$thr)) {
        c(0.1, 0.2)
      } else if (th$thr == 0.1) {
        0.2   # only show 0.2 cutoff inside the >=0.1 plot
      } else {
        NULL  # nothing extra to mark inside >=0.2 plot
      }

      p <- ggplot(df, aes(x = psi)) +
        geom_histogram(aes(y = after_stat(density)),
                       binwidth = 0.02, fill = "steelblue",
                       color = "white", linewidth = 0.2, alpha = 0.7) +
        geom_density(color = "navy", linewidth = 0.7) +
        geom_vline(xintercept = med_val,  color = "red",
                   linetype = "dashed",  linewidth = 0.6) +
        geom_vline(xintercept = mean_val, color = "darkorange",
                   linetype = "dotted",  linewidth = 0.6) +
        annotate("text", x = Inf, y = Inf,
                 label = ann_text,
                 hjust = 1.05, vjust = 1.05, size = 3, family = "mono") +
        scale_x_continuous(limits = c(if (is.null(th$thr)) 0 else th$thr, 1),
                           breaks = seq(0, 1, 0.1)) +
        labs(
          title    = samp,
          subtitle = th$label,
          x        = "%\u03a8  (C / C+T)",
          y        = "Density"
        ) +
        theme_bw(base_size = 10) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))

      if (!is.null(vlines))
        p <- p + geom_vline(xintercept = vlines, color = "darkgreen",
                            linetype = "longdash", linewidth = 0.5)
      p
    })
  })

  # flatten and name: list of (threshold_group x replicates)
  # return as a named list of plot groups for flexible rendering
  names(all_plots) <- sapply(thresholds, `[[`, "label")
  lapply(all_plots, function(grp) Filter(Negate(is.null), grp))
}

# ── Scatter plots for one threshold ──────────────────────────────────────────
make_scatter_plots_replicate <- function(psi_mat, pairs, cl, thr = NULL) {
  label <- if (is.null(thr)) "all sites" else sprintf("%%\u03a8 \u2265 %.1f", thr)

  lapply(pairs, function(p) {
    df <- psi_mat[complete.cases(psi_mat[, p]), ]
    if (!is.null(thr))
      df <- df[df[[p[1]]] >= thr & df[[p[2]]] >= thr, ]
    if (nrow(df) < 5) return(NULL)

    r <- cor(df[[p[1]]], df[[p[2]]], method = "pearson")
    ggplot(df, aes(x = .data[[p[1]]], y = .data[[p[2]]])) +
      geom_point(alpha = 0.4, size = 0.8) +
      geom_abline(slope = 1, intercept = 0,
                  color = "red", linetype = "dashed") +
      annotate("text", x = Inf, y = -Inf,
               label = sprintf("r = %.3f\nn = %d", r, nrow(df)),
               hjust = 1.1, vjust = -0.3, size = 3.5) +
      labs(
        title = sprintf("%s vs %s\n[%s]", p[1], p[2], label),
        x     = sprintf("%%\u03a8  %s", p[1]),
        y     = sprintf("%%\u03a8  %s", p[2])
      ) +
      theme_bw(base_size = 10)
  })
}

# ── Arrange + render a list of plots inline ───────────────────────────────────
render_plots <- function(plot_list, ncol = 3,
                         width_per = 5, height_per = 4.5) {
  plot_list <- Filter(Negate(is.null), plot_list)
  if (length(plot_list) == 0) {
    message("No plots to render.")
    return(invisible(NULL))
  }
  n      <- length(plot_list)
  ncol_p <- min(ncol, n)
  nrow_p <- ceiling(n / ncol_p)
  grid.arrange(grobs = plot_list, ncol = ncol_p)
}


# ── Correlation heatmap for one threshold ─────────────────────────────────────
make_cor_heatmap <- function(psi_mat, pairs, rep_names, cl, thr = NULL) {
  label <- if (is.null(thr)) "all sites" else sprintf("%%\u03a8 \u2265 %.1f", thr)

  cor_df <- do.call(rbind, lapply(pairs, function(p) {
    df <- psi_mat[complete.cases(psi_mat[, p]), ]
    if (!is.null(thr)) df <- df[df[[p[1]]] >= thr & df[[p[2]]] >= thr, ]
    r <- if (nrow(df) >= 5)
           round(cor(df[[p[1]]], df[[p[2]]], method = "pearson"), 4)
         else NA_real_
    data.frame(rep1 = p[1], rep2 = p[2], r = r,
               n = nrow(df), stringsAsFactors = FALSE)
  }))

  # build a symmetric matrix (diagonal = 1)
  mat <- matrix(NA_real_, nrow = length(rep_names), ncol = length(rep_names),
                dimnames = list(rep_names, rep_names))
  diag(mat) <- 1
  for (i in seq_len(nrow(cor_df))) {
    mat[cor_df$rep1[i], cor_df$rep2[i]] <- cor_df$r[i]
    mat[cor_df$rep2[i], cor_df$rep1[i]] <- cor_df$r[i]
  }

  # n lookup (symmetric, diagonal = NA for display)
  n_mat <- matrix(NA_integer_, nrow = length(rep_names), ncol = length(rep_names),
                  dimnames = list(rep_names, rep_names))
  for (i in seq_len(nrow(cor_df))) {
    n_mat[cor_df$rep1[i], cor_df$rep2[i]] <- cor_df$n[i]
    n_mat[cor_df$rep2[i], cor_df$rep1[i]] <- cor_df$n[i]
  }

  # tidy for ggplot
  plot_df <- expand.grid(x = rep_names, y = rep_names,
                         stringsAsFactors = FALSE) %>%
    rowwise() %>%
    mutate(
      r     = mat[x, y],
      n     = n_mat[x, y],
      label = case_when(
        x == y ~ "1.000",
        is.na(r) ~ "NA",
        TRUE ~ sprintf("%.3f\n(n=%d)", r, n)
      )
    ) %>%
    ungroup()

  # shorten sample IDs for axis labels (keep last 2 dash-fields)
  short <- function(s) sub("^[^_]+_", "", s)   # drop "CLxxxxx_" prefix
  plot_df$x_lab <- short(plot_df$x)
  plot_df$y_lab <- short(plot_df$y)
  lev <- short(rep_names)
  plot_df$x_lab <- factor(plot_df$x_lab, levels = lev)
  plot_df$y_lab <- factor(plot_df$y_lab, levels = rev(lev))

  ggplot(plot_df, aes(x = x_lab, y = y_lab, fill = r)) +
    geom_tile(color = "white", linewidth = 0.8) +
    geom_text(aes(label = label), size = 3, lineheight = 0.9) +
    # scale_fill_gradientn(
    #   colors  = c("#4575B4", "#91BFDB", "#E0F3F8", "#FFFFBF",
    #               "#FEE090", "#FC8D59", "#D73027"),
    #   values  = scales::rescale(c(0, 0.5, 0.8, 0.9, 0.95, 0.99, 1)),
    #   limits  = c(0, 1),
    #   na.value = "grey90",
    #   name    = "Pearson r"
    # ) +
    scale_fill_gradient2(
      low  = "#1a9641",
      mid  = "#ffffbf",
      high = "#d7191c",
      midpoint = 0.5,
      limits   = c(0, 1),
      na.value = "grey90",
      name     = "Pearson r"
    ) +
    labs(
      title    = cl,
      subtitle = label,
      x        = NULL,
      y        = NULL
    ) +
    coord_fixed() +
    theme_bw(base_size = 10) +
    theme(
      axis.text.x      = element_text(angle = 35, hjust = 1),
      panel.grid       = element_blank(),
      panel.border     = element_blank(),
      legend.key.height = unit(0.8, "cm")
    )
}

# ── Three-panel heatmap (all / >=0.1 / >=0.2) ────────────────────────────────
make_cor_heatmaps <- function(psi_mat, pairs, rep_names, cl) {
  list(
    make_cor_heatmap(psi_mat, pairs, rep_names, cl, thr = NULL),
    make_cor_heatmap(psi_mat, pairs, rep_names, cl, thr = 0.1),
    make_cor_heatmap(psi_mat, pairs, rep_names, cl, thr = 0.2)
  )
}



```


```{r run-all-replicate, include=FALSE}
# Pre-load all cell line data so each section chunk can reference it cleanly.
cl_data <- list()

for (cl in names(CELL_LINES)) {
  mpile_dir <- file.path(RESULTS_DIR, cl, "mpile")
  reps      <- CELL_LINES[[cl]]

  rep_list <- lapply(names(reps), function(rep_name) {
    read_rRNA_psi(cl, reps[[rep_name]], mpile_dir)
  })
  names(rep_list) <- names(reps)
  rep_list <- Filter(Negate(is.null), rep_list)

  if (length(rep_list) < 2) {
    cl_data[[cl]] <- list(ok = FALSE, reason = "Not enough replicates found.")
    next
  }

  merged <- Reduce(function(a, b) inner_join(a, b, by = "site"), rep_list)

  if (nrow(merged) < 10) {
    cl_data[[cl]] <- list(ok = FALSE, reason = "Too few shared sites.")
    next
  }

  psi_mat   <- merged[, -1, drop = FALSE]
  rep_names <- colnames(psi_mat)
  pairs     <- combn(rep_names, 2, simplify = FALSE)

  cl_data[[cl]] <- list(
    ok        = TRUE,
    psi_mat   = psi_mat,
    rep_names = rep_names,
    pairs     = pairs,
    n_sites   = nrow(merged),
    site_sets = setNames(
      lapply(rep_list, function(df) df$site),
      sapply(rep_list, function(df) colnames(df)[2])  # grabs the sample ID column name
    )
  )
}
```



## CL18870 {.tabset}

```{r cl18870-check, include=FALSE}
cl  <- "CL18870"
dat <- cl_data[[cl]]
```

`r if (!dat$ok) paste("**Skipped:**", dat$reason)`

```{r cl18870-skip, eval=!dat$ok, echo=FALSE}
## nothing to render — message printed above via inline R
```

### %pU Distribution - All sites

```{r cl18870-dist-all, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["all sites"]],
             ncol = 2)
```

### %pU Distribution - %pU >= 0.1
```{r cl18870-dist-01, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["%\u03a8 \u2265 0.1"]],
             ncol = 2)
```

### %pU Distribution - %pU >= 0.2
```{r cl18870-dist-02, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["%\u03a8 \u2265 0.2"]],
             ncol = 2)
```

### Pairwise correlation scatter plot— all sites

Sites with depth >= `r MIN_DEPTH` shared across all replicates: **`r if(dat$ok) dat$n_sites`**

```{r cl18870-cor-all, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = NULL))
```

### Pairwise correlation scatter plot— %pU >= 0.1

```{r cl18870-cor-01, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = 0.1))
```

### Pairwise correlation scatter plot— %pU >= 0.2

```{r cl18870-cor-02, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = 0.2))
```


### Pairwise Pearson heatmap

```{r cl18870-heatmap1, eval=dat$ok, fig.width=13.5, fig.height=4.8}
render_plots(
  make_cor_heatmaps(dat$psi_mat, dat$pairs, dat$rep_names, cl),
  ncol = 3
)
```

---

## CL18909 {.tabset}

```{r cl18909-check, include=FALSE}
cl  <- "CL18909"
dat <- cl_data[[cl]]
```

`r if (!dat$ok) paste("**Skipped:**", dat$reason)`

### %pU Distribution

#### All sites
```{r cl18909-dist-all, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["all sites"]],
             ncol = 2)
```

### %pU Distribution - %pU >= 0.1
```{r cl18909-dist-01, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["%\u03a8 \u2265 0.1"]],
             ncol = 2)
```

### %pU Distribution - %pU >= 0.2
```{r cl18909-dist-02, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["%\u03a8 \u2265 0.2"]],
             ncol = 2)
```

### Pairwise correlation scatter plot— all sites

Sites with depth >= `r MIN_DEPTH` shared across all replicates: **`r if(dat$ok) dat$n_sites`**

```{r cl18909-cor-all, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = NULL))
```

### Pairwise correlation scatter plot— %pU >= 0.1

```{r cl18909-cor-01, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = 0.1))
```

### Pairwise correlation scatter plot— %pU >= 0.2

```{r cl18909-cor-02, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = 0.2))
```

### Pairwise Pearson heatmap

```{r cl18909-heatmap2, eval=dat$ok, fig.width=13.5, fig.height=4.8}
render_plots(
  make_cor_heatmaps(dat$psi_mat, dat$pairs, dat$rep_names, cl),
  ncol = 3
)
```

---

## CL18916 {.tabset}

```{r cl18916-check, include=FALSE}
cl  <- "CL18916"
dat <- cl_data[[cl]]
```

`r if (!dat$ok) paste("**Skipped:**", dat$reason)`

### %pU Distribution

#### All sites
```{r cl18916-dist-all, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["all sites"]],
             ncol = 2)
```

### %pU Distribution - %pU >= 0.1
```{r cl18916-dist-01, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["%\u03a8 \u2265 0.1"]],
             ncol = 2)
```

### %pU Distribution - %pU >= 0.2
```{r cl18916-dist-02, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["%\u03a8 \u2265 0.2"]],
             ncol = 2)
```

### Pairwise correlation scatter plot— all sites

Sites with depth >= `r MIN_DEPTH` shared across all replicates: **`r if(dat$ok) dat$n_sites`**

```{r cl18916-cor-all, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = NULL))
```

### Pairwise correlation scatter plot— %pU >= 0.1

```{r cl18916-cor-01, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = 0.1))
```

### Pairwise correlation scatter plot— %pU >= 0.2

```{r cl18916-cor-02, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = 0.2))
```

### Pairwise Pearson heatmap

```{r cl18916-heatmap1, eval=dat$ok, fig.width=13.5, fig.height=4.8}
render_plots(
  make_cor_heatmaps(dat$psi_mat, dat$pairs, dat$rep_names, cl),
  ncol = 3
)
```

---

## CL19099 {.tabset}

```{r cl19099-check, include=FALSE}
cl  <- "CL19099"
dat <- cl_data[[cl]]
```

`r if (!dat$ok) paste("**Skipped:**", dat$reason)`

### %pU Distribution

#### All sites
```{r cl19099-dist-all, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["all sites"]],
             ncol = 2)
```

### %pU Distribution - %pU >= 0.1
```{r cl19099-dist-01, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["%\u03a8 \u2265 0.1"]],
             ncol = 2)
```

### %pU Distribution - %pU >= 0.2
```{r cl19099-dist-02, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["%\u03a8 \u2265 0.2"]],
             ncol = 2)
```

### Pairwise correlation scatter plot— all sites

Sites with depth >= `r MIN_DEPTH` shared across all replicates: **`r if(dat$ok) dat$n_sites`**

```{r cl19099-cor-all, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = NULL))
```

### Pairwise correlation scatter plot— %pU >= 0.1

```{r cl19099-cor-01, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = 0.1))
```

### Pairwise correlation scatter plot— %pU >= 0.2

```{r cl19099-cor-02, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = 0.2))
```

### Pairwise Pearson heatmap

```{r cl19099-heatmap1, eval=dat$ok, fig.width=13.5, fig.height=4.8}
render_plots(
  make_cor_heatmaps(dat$psi_mat, dat$pairs, dat$rep_names, cl),
  ncol = 3
)
```

---






# Comparing pooled data among cell lines 

```{r setup-comparingcellline, message=FALSE, warning=FALSE}

RESULTS_DIR  <- "/project2/xinhe/xsun/pumseq/pumseq_results"
CELL_LINES   <- c("CL18870", "CL18909", "CL18916", "CL19099")
DEPTH_CUTS   <- c(10, 20)
RATIO_CUTS   <- c(0.1, 0.2)
NSAMP_CUTS   <- 1:4

## ── Helpers ──────────────────────────────────────────────────────────────────

load_pooled_rRNA <- function(cl) {
  f <- file.path(RESULTS_DIR, cl, "pooled_sites", "pooled_rRNA.txt")
  dt <- fread(f, header = TRUE, sep = "\t",
              colClasses = list(character = "ref"))
  dt[, cell_line := cl]
  dt[, site := paste(ref, pos, strand, sep = ":")]
  dt
}

## Count sites passing (pooled depth >= d) AND (ratio >= r) AND (n_samples >= n)
count_sites <- function(dt, min_depth, min_ratio, min_nsamp) {
  sum(dt$depth >= min_depth & dt$ratio >= min_ratio & dt$n_samples >= min_nsamp)
}

## Build the full site-count grid for one cell line
build_count_grid <- function(dt) {
  expand.grid(
    min_depth  = DEPTH_CUTS,
    min_ratio  = RATIO_CUTS,
    min_nsamp  = NSAMP_CUTS,
    stringsAsFactors = FALSE
  ) |>
    rowwise() |>
    mutate(n_sites = count_sites(dt, min_depth, min_ratio, min_nsamp)) |>
    ungroup() |>
    mutate(
      depth_label = paste0("depth>=", min_depth),
      ratio_label = paste0("ratio>=", min_ratio),
      nsamp_label = paste0("n_samples>=", min_nsamp)
    )
}

## Heatmap (tile plot) of site counts across depth × ratio cutoffs,
## faceted by min_n_samples.
plot_count_heatmap <- function(grid, cl_title) {
  grid |>
    mutate(
      depth_label = factor(depth_label, levels = paste0("depth>=", DEPTH_CUTS)),
      ratio_label = factor(ratio_label, levels = paste0("ratio>=", rev(RATIO_CUTS))),
      nsamp_label = factor(nsamp_label,
                           levels = paste0("n_samples>=", NSAMP_CUTS))
    ) |>
    ggplot(aes(x = depth_label, y = ratio_label, fill = n_sites)) +
    geom_tile(color = "white", linewidth = 0.6) +
    geom_text(aes(label = comma(n_sites)), size = 4, fontface = "bold") +
    scale_fill_gradient(low = "#deebf7", high = "#2171b5",
                        name = "## sites", labels = comma) +
    facet_wrap(~nsamp_label, nrow = 1) +
    labs(title = cl_title,
         x = "Depth cutoff (pooled)", y = "Ratio cutoff (pooled %pU)") +
    theme_bw(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          strip.background = element_rect(fill = "#e8eaf6"),
          strip.text = element_text(face = "bold"))
}

## ── Load all data ─────────────────────────────────────────────────────────────
dat_list <- lapply(CELL_LINES, load_pooled_rRNA)
names(dat_list) <- CELL_LINES

grid_list <- lapply(dat_list, build_count_grid)
names(grid_list) <- CELL_LINES
```

## Per cell-line site counts at different cutoffs

Each heatmap shows the number of rRNA sites passing a given combination of
**pooled depth** (x-axis), **pooled %pU ratio** (y-axis), and
**minimum number of replicates** (panels).

---

### CL18870

```{r cl18870-heatmap, fig.width=12, fig.height=3.5}
plot_count_heatmap(grid_list[["CL18870"]], "CL18870 – pU site counts")
```

---

### CL18909

```{r cl18909-heatmap, fig.width=12, fig.height=3.5}
plot_count_heatmap(grid_list[["CL18909"]], "CL18909 – pU site counts")
```

---

### CL18916

```{r cl18916-heatmap, fig.width=12, fig.height=3.5}
plot_count_heatmap(grid_list[["CL18916"]], "CL18916 – pU site counts")
```

---

### CL19099

```{r cl19099-heatmap, fig.width=12, fig.height=3.5}
plot_count_heatmap(grid_list[["CL19099"]], "CL19099 – pU site counts")
```

---

## Cross-cell-line comparison of site counts

The n_sample below means minimum number of replicates the site identified. 


```{r cross-cl-comparison, fig.width=13, fig.height=9}
## Combine grids
all_grids <- bind_rows(
  lapply(CELL_LINES, function(cl) {
    grid_list[[cl]] |> mutate(cell_line = cl)
  })
) |>
  mutate(
    cell_line   = factor(cell_line, levels = CELL_LINES),
    depth_label = factor(depth_label, levels = paste0("depth>=", DEPTH_CUTS)),
    ratio_label = factor(ratio_label, levels = paste0("ratio>=", rev(RATIO_CUTS))),
    nsamp_label = factor(nsamp_label, levels = paste0("n_samples>=", NSAMP_CUTS))
  )

## Grouped bar chart: x = cell line, fill = cell line, facet_grid(ratio × depth × n_samples)
ggplot(all_grids, aes(x = cell_line, y = n_sites, fill = cell_line)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = comma(n_sites)), vjust = -0.3, size = 2.8) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  facet_grid(ratio_label ~ depth_label + nsamp_label) +
  labs(
    title = "pU site counts across cell lines",
    x     = NULL,
    y     = "Number of sites"
  ) +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x    = element_text(angle = 40, hjust = 1, size = 8),
    strip.background = element_rect(fill = "#e8eaf6"),
    strip.text     = element_text(face = "bold", size = 8),
    panel.spacing  = unit(0.4, "lines")
  )
```



---

## Pearson's r of %pU between cell lines

For each cutoff combination, sites passing all three filters
(depth, ratio, n_samples) in **both** cell lines are used to compute
Pearson's r on the pooled %pU (`ratio`).
Below the correlation matrices, scatter plots are shown for
a representative cutoff (depth >= 20, ratio >= 0.1, n_samples >= 2).

```{r pearson-setup}
## For a given pair and cutoff, find common passing sites and return Pearson r
pearson_pair <- function(dt1, dt2, min_depth, min_ratio, min_nsamp) {
  pass1 <- dt1[depth >= min_depth & ratio >= min_ratio & n_samples >= min_nsamp,
               .(site, ratio1 = ratio)]
  pass2 <- dt2[depth >= min_depth & ratio >= min_ratio & n_samples >= min_nsamp,
               .(site, ratio2 = ratio)]
  merged <- merge(pass1, pass2, by = "site")
  if (nrow(merged) < 3) return(data.frame(r = NA_real_, n_common = nrow(merged)))
  data.frame(
    r        = cor(merged$ratio1, merged$ratio2, method = "pearson"),
    n_common = nrow(merged)
  )
}

## Build full correlation grid across all cell-line pairs and cutoff combos
cl_pairs <- combn(CELL_LINES, 2, simplify = FALSE)

cor_grid <- do.call(rbind, lapply(cl_pairs, function(p) {
  do.call(rbind, lapply(DEPTH_CUTS, function(d) {
    do.call(rbind, lapply(RATIO_CUTS, function(r) {
      do.call(rbind, lapply(NSAMP_CUTS, function(n) {
        res <- pearson_pair(
          dat_list[[p[1]]], dat_list[[p[2]]],
          min_depth = d, min_ratio = r, min_nsamp = n
        )
        data.frame(cl1 = p[1], cl2 = p[2],
                   min_depth = d, min_ratio = r, min_nsamp = n,
                   pearson_r = round(res$r, 3),
                   n_common  = res$n_common)
      }))
    }))
  }))
}))

## Add symmetric entries (cl2, cl1) for heatmap matrix
cor_sym <- cor_grid |>
  bind_rows(
    cor_grid |> rename(cl1 = cl2, cl2 = cl1)
  ) |>
  bind_rows(
    expand.grid(cl1 = CELL_LINES, cl2 = CELL_LINES,
                min_depth = DEPTH_CUTS, min_ratio = RATIO_CUTS,
                min_nsamp = NSAMP_CUTS) |>
      filter(cl1 == cl2) |>
      mutate(pearson_r = 1, n_common = NA_integer_)
  ) |>
  mutate(
    cl1         = factor(cl1, levels = CELL_LINES),
    cl2         = factor(cl2, levels = rev(CELL_LINES)),
    nsamp_label = factor(paste0("n_samples>=", min_nsamp),
                         levels = paste0("n_samples>=", NSAMP_CUTS)),
    dr_label    = factor(
      paste0("d>=", min_depth, " r>=", min_ratio),
      levels = apply(
        expand.grid(d = DEPTH_CUTS, r = RATIO_CUTS), 1,
        function(x) paste0("d>=", x["d"], " r>=", x["r"])
      )
    )
  )
```

### Correlation matrices

Facets: rows = min n_samples, columns = depth × ratio cutoff.
Values shown inside each cell are Pearson's r (diagonal = 1 by definition).

```{r pearson-heatmap, fig.width=14, fig.height=11}
ggplot(cor_sym, aes(x = cl1, y = cl2, fill = pearson_r)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = ifelse(cl1 == cl2, "1", sprintf("%.3f", pearson_r))),
            size = 3, fontface = "bold") +
  scale_fill_gradient2(
    low = "#1a9641", mid = "#ffffbf", high = "#d7191c",
    midpoint = 0.5, limits = c(0, 1), na.value = "grey90",
    name = "Pearson r"
  ) +
  facet_grid(nsamp_label ~ dr_label) +
  labs(
    title = "Pearson's r of pooled %pU between cell lines (rRNA sites)",
    x = NULL, y = NULL
  ) +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1, size = 9),
    axis.text.y      = element_text(size = 9),
    strip.background = element_rect(fill = "#e8eaf6"),
    strip.text       = element_text(face = "bold", size = 9),
    panel.spacing    = unit(0.3, "lines"),
    legend.position  = "right"
  )
```


### Scatter plots (depth >= 20, ratio >= 0.1, n_samples >= 2)

```{r pearson-scatter, fig.width=13, fig.height=8}
MIN_D <- 20; MIN_R <- 0.1; MIN_N <- 2

scatter_data <- lapply(CELL_LINES, function(cl) {
  dat_list[[cl]][depth >= MIN_D & ratio >= MIN_R & n_samples >= MIN_N,
                 .(site, ratio, cell_line = cl)]
})
names(scatter_data) <- CELL_LINES

scatter_plots <- lapply(cl_pairs, function(p) {
  d1 <- scatter_data[[p[1]]]
  d2 <- scatter_data[[p[2]]]
  m  <- merge(d1, d2, by = "site", suffixes = c("_1", "_2"))
  if (nrow(m) < 3) return(NULL)
  r_val <- cor(m$ratio_1, m$ratio_2, method = "pearson")
  ggplot(m, aes(x = ratio_1, y = ratio_2)) +
    geom_point(alpha = 0.55, size = 1.2, color = "#2c7bb6") +
    geom_abline(slope = 1, intercept = 0, color = "red",
                linetype = "dashed", linewidth = 0.6) +
    annotate("text", x = Inf, y = -Inf,
             label = sprintf("r = %.3f\nn = %d", r_val, nrow(m)),
             hjust = 1.1, vjust = -0.3, size = 3.5) +
    scale_x_continuous(labels = percent_format(accuracy = 1)) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      title = paste(p[1], "vs", p[2]),
      x     = paste0("%pU  ", p[1]),
      y     = paste0("%pU  ", p[2])
    ) +
    theme_bw(base_size = 10)
})
scatter_plots <- Filter(Negate(is.null), scatter_plots)

do.call(grid.arrange, c(scatter_plots, list(ncol = 3)))
```














# Comparing with published data

```{r pub-helpers}
make_scatter_publication <- function(df, x, y, col, ttl, xlab, ylab) {
  if (nrow(df) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = "No overlapping sites",
                 size = 4, color = "grey50") +
        labs(title = ttl, x = xlab, y = ylab) +
        theme_classic(base_size = 11) +
        theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))
    )
  }
  ggplot(df, aes(x = .data[[x]], y = .data[[y]],
                 color = .data[[col]], label = label)) +
    geom_point(size = 2.5, alpha = 0.8) +
    geom_text_repel(size = 2.5, max.overlaps = 20) +
    geom_smooth(method = "lm", se = TRUE, color = "grey40", linetype = "dashed") +
    scale_color_manual(values = c("5.8S" = "#4E79A7", "18S" = "#F28E2B",
                                  "28S"  = "#59A14F", "5S"  = "#E15759")) +
    labs(title = ttl, x = xlab, y = ylab, color = "rRNA type") +
    theme_classic(base_size = 11) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))
}

make_heatmap_publication <- function(data, fill_col, label_col, ttl, fill_label) {
  ggplot(data, aes(x = min_ratio, y = min_depth, fill = .data[[fill_col]])) +
    geom_tile(color = "white", linewidth = 1.2) +
    geom_text(aes(label = .data[[label_col]]), size = 6, fontface = "bold", color = "white") +
    scale_fill_gradient(low = "#d0e8f2", high = "#1a6fa8") +
    labs(title = ttl, x = "Min ratio cutoff", y = "Min depth cutoff", fill = fill_label) +
    theme_classic(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          axis.text  = element_text(size = 11))
}

# FIX: takes results as explicit argument to avoid global scoping issues
combine_panels <- function(results, slot) {
  plots <- map(results, slot)
  wrap_plots(plots, ncol = 2) +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
}
```

## Published data overview

We first check the site overlap in these two published data.

```{r pub-load, warning=FALSE, message=FALSE}
path_BACS   <- "/project2/xinhe/xsun/pumseq/data_others/pU_PMID39349603.xlsx"
path_PRAISE <- "/project2/xinhe/xsun/pumseq/data_others/pU_PMID36997645.xlsx"

rrna_BACS <- readxl::read_excel(path_BACS, sheet = 2)
colnames(rrna_BACS) <- rrna_BACS[2, ]
rrna_BACS <- rrna_BACS[-c(1:2), ]
rrna_BACS <- rrna_BACS[complete.cases(rrna_BACS$chr), ]

rrna_PRAISE <- readxl::read_excel(path_PRAISE, sheet = 1)
colnames(rrna_PRAISE) <- rrna_PRAISE[2, ]
rrna_PRAISE <- rrna_PRAISE[-c(1:2), ]
rrna_PRAISE <- rrna_PRAISE %>% filter(!is.na(chr_name), !is.na(site))

# --- Clean BACS ---
rrna_BACS_clean <- rrna_BACS %>%
  mutate(
    rna_type = case_when(
      str_detect(chr, "RNA5_8S") ~ "5.8S",
      str_detect(chr, "RNA18S")  ~ "18S",
      str_detect(chr, "RNA28S")  ~ "28S",
      str_detect(chr, "RNA5SN")  ~ "5S",
      TRUE ~ NA_character_
    ),
    pos_num = as.integer(str_extract(pos, "^[0-9]+"))
  )

# --- Clean PRAISE ---
rrna_PRAISE_clean <- rrna_PRAISE %>%
  mutate(
    rna_type  = str_extract(chr_name, "(5\\.8S|18S|28S|5S)"),
    pos_start = as.integer(str_extract(site, "^[0-9]+")),
    pos_end   = as.integer(coalesce(
      str_extract(site, "(?<=-)[0-9]+$"),
      str_extract(site, "^[0-9]+")
    ))
  )

# --- Overlap between the two published datasets ---
overlap <- rrna_BACS_clean %>%
  inner_join(rrna_PRAISE_clean, by = "rna_type") %>%
  filter(pos_num >= pos_start & pos_num <= pos_end)

DT::datatable(rrna_BACS, caption = htmltools::tags$caption(
  style = 'caption-side: left; text-align: left; color:black; font-size:150%;',
  'Sites from BACS'), options = list(pageLength = 5))

DT::datatable(rrna_PRAISE, caption = htmltools::tags$caption(
  style = 'caption-side: left; text-align: left; color:black; font-size:150%;',
  'Sites from PRAISE'), options = list(pageLength = 5))

cat("Sites in BACS:    ", nrow(rrna_BACS_clean), "\n")
cat("Sites in PRAISE:  ", nrow(rrna_PRAISE_clean), "\n")
cat("Overlapping sites:", nrow(overlap), "\n")

# --- Final cleaned versions used for all cell line comparisons ---
rrna_BACS_clean <- rrna_BACS_clean %>%
  mutate(accession = str_extract(chr, "NR_[0-9]+"),
         pos_num   = as.integer(pos)) %>%
  rename(`pU level_rep1` = `Ψ level_rep1`,
         `pU level_rep2` = `Ψ level_rep2`)

rrna_PRAISE_clean <- rrna_PRAISE_clean %>%
  mutate(accession = str_extract(chr_name, "NR_[0-9]+"))
```

```{r pub-preload, include=FALSE}
# FIX: pre-run all cell lines using lapply (not purrr::pmap) to avoid
# data.table child-environment scoping issues with ours_filtered.
run_cl_comparison <- function(cl) {
  pooled_path <- file.path(RESULTS_DIR, cl, "pooled_sites", "pooled_rRNA.txt")
  if (!file.exists(pooled_path)) {
    warning("File not found: ", pooled_path)
    return(NULL)
  }

  pooled_ours_rrna     <- fread(pooled_path)

  results <- lapply(seq_len(nrow(cutoffs)), function(i) {
    min_depth <- cutoffs$min_depth[i]
    min_ratio <- cutoffs$min_ratio[i]
    tag       <- paste0("depth>=", min_depth, "  ratio>=", min_ratio)

    ours_filtered <- pooled_ours_rrna %>%
      filter(depth >= min_depth, ratio >= min_ratio) %>%
      mutate(accession = str_extract(ref, "NR_[0-9]+"))

    # vs BACS
    overlap_bacs <- ours_filtered %>%
      inner_join(rrna_BACS_clean, by = c("accession", "pos" = "pos_num")) %>%
      mutate(
        psi_rep1 = as.numeric(`pU level_rep1`),
        psi_rep2 = as.numeric(`pU level_rep2`),
        label    = paste0(name, " pos", pos)
      )

    # vs PRAISE
    overlap_praise <- ours_filtered %>%
      inner_join(rrna_PRAISE_clean, by = "accession", relationship = "many-to-many") %>%
      filter(pos >= pos_start & pos <= pos_end) %>%
      mutate(
        deletion_rep1 = as.numeric(rep1_deletion_ratio),
        deletion_rep2 = as.numeric(rep2_deletion_ratio),
        label         = paste0(rna_type, " pos", pos)
      )

    list(
      tag            = tag,
      min_depth      = min_depth,
      min_ratio      = min_ratio,
      n_ours         = nrow(ours_filtered),
      n_bacs         = nrow(overlap_bacs),
      n_praise       = nrow(overlap_praise),
      overlap_bacs   = overlap_bacs,
      overlap_praise = overlap_praise,

      p_bacs_r1   = make_scatter_publication(overlap_bacs, "ratio", "psi_rep1", "name",
                                             tag, "Our pU ratio", "BACS pU level (rep1)"),
      p_bacs_r2   = make_scatter_publication(overlap_bacs, "ratio", "psi_rep2", "name",
                                             tag, "Our pU ratio", "BACS pU level (rep2)"),
      p_praise_r1 = make_scatter_publication(overlap_praise, "ratio", "deletion_rep1", "rna_type",
                                             tag, "Our pU ratio", "PRAISE deletion ratio (rep1)"),
      p_praise_r2 = make_scatter_publication(overlap_praise, "ratio", "deletion_rep2", "rna_type",
                                             tag, "Our pU ratio", "PRAISE deletion ratio (rep2)")
    )
  })

  names(results) <- paste0("depth", cutoffs$min_depth, "_ratio", cutoffs$min_ratio)
  results
}

pub_data <- lapply(c("CL18870", "CL18909", "CL18916", "CL19099"), run_cl_comparison)
names(pub_data) <- c("CL18870", "CL18909", "CL18916", "CL19099")
```

---

## CL18870 {.tabset}

```{r cl18870-pub-check, include=FALSE}
cl      <- "CL18870"
results <- pub_data[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled file not found for", cl)`

### Site counts

```{r cl18870-pub-heatmap, eval=ok, fig.width=12, fig.height=4}
heatmap_data <- map_dfr(results, ~ tibble(
  min_depth = .x$min_depth, min_ratio = .x$min_ratio,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(min_depth = factor(paste0("depth>=", min_depth)),
         min_ratio = factor(paste0("ratio>=", min_ratio)))

make_heatmap_publication(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, "  site counts across depth & ratio cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

### vs BACS (rep1)

```{r cl18870-pub-bacs-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS pU level (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs BACS (rep2)

```{r cl18870-pub-bacs-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS pU level (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

```{r cl18870-pub-praise-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r cl18870-pub-praise-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

---

## CL18909 {.tabset}

```{r cl18909-pub-check, include=FALSE}
cl      <- "CL18909"
results <- pub_data[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled file not found for", cl)`

### Site counts

```{r cl18909-pub-heatmap, eval=ok, fig.width=12, fig.height=4}
heatmap_data <- map_dfr(results, ~ tibble(
  min_depth = .x$min_depth, min_ratio = .x$min_ratio,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(min_depth = factor(paste0("depth>=", min_depth)),
         min_ratio = factor(paste0("ratio>=", min_ratio)))

make_heatmap_publication(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, "  site counts across depth & ratio cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

### vs BACS (rep1)

```{r cl18909-pub-bacs-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS pU level (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs BACS (rep2)

```{r cl18909-pub-bacs-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS pU level (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

```{r cl18909-pub-praise-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r cl18909-pub-praise-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

---

## CL18916 {.tabset}

```{r cl18916-pub-check, include=FALSE}
cl      <- "CL18916"
results <- pub_data[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled file not found for", cl)`

### Site counts

```{r cl18916-pub-heatmap, eval=ok, fig.width=12, fig.height=4}
heatmap_data <- map_dfr(results, ~ tibble(
  min_depth = .x$min_depth, min_ratio = .x$min_ratio,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(min_depth = factor(paste0("depth>=", min_depth)),
         min_ratio = factor(paste0("ratio>=", min_ratio)))

make_heatmap_publication(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, "  site counts across depth & ratio cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

### vs BACS (rep1)

```{r cl18916-pub-bacs-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS pU level (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs BACS (rep2)

```{r cl18916-pub-bacs-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS pU level (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

```{r cl18916-pub-praise-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r cl18916-pub-praise-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

---

## CL19099 {.tabset}

```{r cl19099-pub-check, include=FALSE}
cl      <- "CL19099"
results <- pub_data[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled file not found for", cl)`

### Site counts

```{r cl19099-pub-heatmap, eval=ok, fig.width=12, fig.height=4}
heatmap_data <- map_dfr(results, ~ tibble(
  min_depth = .x$min_depth, min_ratio = .x$min_ratio,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(min_depth = factor(paste0("depth>=", min_depth)),
         min_ratio = factor(paste0("ratio>=", min_ratio)))

make_heatmap_publication(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, "  site counts across depth & ratio cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

### vs BACS (rep1)

```{r cl19099-pub-bacs-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS pU level (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs BACS (rep2)

```{r cl19099-pub-bacs-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS pU level (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

```{r cl19099-pub-praise-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r cl19099-pub-praise-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```
