---
title: "QC — Per-replicate rRNA %pU vs BACS"
author: "XSun"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, warning=FALSE, message=FALSE}
library(data.table)
library(dplyr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(patchwork)

knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center"
)
```

# Introduction

For each cell line, we compare every treated replicate's %pU (conversion
ratio C/(C+T)) at rRNA sites against **both replicates** of the **BACS**
dataset (PMID 39349603). Sites are matched by rRNA accession (`NR_*`) and
position.

Each row shows scatter plots of one replicate vs BACS rep1 (left) and
rep2 (right), with Pearson r annotated. This helps identify replicates with
weak or noisy signal compared to the published reference.

```{r config}
RESULTS_DIRS <- c("/project2/xinhe/xsun/pumseq/pumseq_results",
                  "/project/xinhe/xsun/pumseq/pumseq_results")
```

```{r load-bacs}
path_BACS <- "/project2/xinhe/xsun/pumseq/data_others/pU_PMID39349603.xlsx"

rrna_BACS <- readxl::read_excel(path_BACS, sheet = 2)
colnames(rrna_BACS) <- rrna_BACS[2, ]
rrna_BACS <- rrna_BACS[-c(1:2), ]
rrna_BACS <- rrna_BACS[complete.cases(rrna_BACS$chr), ]

bacs <- rrna_BACS %>%
  mutate(
    accession = str_extract(chr, "NR_[0-9]+"),
    pos_num   = as.integer(pos),
    rna_type  = case_when(
      str_detect(chr, "RNA5_8S") ~ "5.8S",
      str_detect(chr, "RNA18S")  ~ "18S",
      str_detect(chr, "RNA28S")  ~ "28S",
      str_detect(chr, "RNA5SN")  ~ "5S",
      TRUE ~ NA_character_
    ),
    bacs_psi_r1 = as.numeric(`Ψ level_rep1`),
    bacs_psi_r2 = as.numeric(`Ψ level_rep2`)
  ) %>%
  select(accession, pos_num, rna_type, bacs_psi_r1, bacs_psi_r2)

cat("BACS rRNA sites:", nrow(bacs), "\n")
```

```{r discover-cls}
all_cls <- c()
for (rd in RESULTS_DIRS) {
  cls <- list.dirs(rd, recursive = FALSE, full.names = FALSE)
  cls <- cls[grepl("^CL\\d+$", cls)]
  for (cl in cls) {
    files <- list.files(file.path(rd, cl, "out_chisq"),
                        pattern = "filter\\..*\\.rRNA\\.pseusite\\.txt\\.gz$")
    if (length(files) > 0) all_cls <- c(all_cls, cl)
  }
}
all_cls <- sort(unique(all_cls))

find_rd <- function(cl) {
  for (rd in RESULTS_DIRS) {
    if (dir.exists(file.path(rd, cl))) return(rd)
  }
  NULL
}

cat("Cell lines:", length(all_cls), "\n")
```

```{r per-cl-scatter, results='asis', fig.width=12, fig.height=5}
make_bacs_scatter <- function(dt, short, ycol, ylab) {
  r_val <- cor(dt$ratio, dt[[ycol]], method = "pearson", use = "complete.obs")
  ggplot(dt, aes(x = ratio, y = .data[[ycol]], color = rna_type)) +
    geom_point(size = 2, alpha = 0.8) +
    geom_text_repel(aes(label = paste0(rna_type, " ", pos)),
                    size = 2, max.overlaps = 15, show.legend = FALSE) +
    geom_smooth(method = "lm", se = TRUE, color = "grey40",
                linetype = "dashed", inherit.aes = FALSE,
                aes(x = ratio, y = .data[[ycol]])) +
    annotate("text", x = Inf, y = -Inf,
             label = sprintf("r = %.3f\nn = %d", r_val, nrow(dt)),
             hjust = 1.1, vjust = -0.3, size = 3.5) +
    scale_color_manual(values = c("5.8S" = "#4E79A7", "18S" = "#F28E2B",
                                  "28S" = "#59A14F", "5S" = "#E15759")) +
    labs(title = paste(short, "vs", ylab), x = "Our %pU", y = ylab,
         color = "rRNA") +
    theme_bw(base_size = 9) +
    theme(legend.position = "none")
}

for (cl in all_cls) {
  rd <- find_rd(cl)
  files <- list.files(file.path(rd, cl, "out_chisq"),
                      pattern = "filter\\..*\\.rRNA\\.pseusite\\.txt\\.gz$",
                      full.names = TRUE)
  files <- files[!grepl("12-50-|12-47-", files)]
  files <- sort(files)
  if (length(files) == 0) next

  cat(sprintf("\n\n## %s\n\n", cl))

  plots <- list()
  for (f in files) {
    rep_id <- sub("\\.rRNA\\.pseusite\\.txt\\.gz$", "",
                  sub("^filter\\.", "", basename(f)))
    short <- sub("^CL\\d+_", "", rep_id)

    dt <- tryCatch(
      fread(cmd = paste("zcat", f), header = TRUE, sep = "\t",
            colClasses = list(character = "ref")),
      error = function(e) NULL
    )
    if (is.null(dt) || nrow(dt) == 0) next

    dt <- dt %>%
      mutate(accession = str_extract(ref, "NR_[0-9]+")) %>%
      inner_join(bacs, by = c("accession", "pos" = "pos_num"))

    if (nrow(dt) < 3) next

    plots[[length(plots)+1]] <- make_bacs_scatter(dt, short, "bacs_psi_r1", "BACS rep1")
    plots[[length(plots)+1]] <- make_bacs_scatter(dt, short, "bacs_psi_r2", "BACS rep2")
  }

  if (length(plots) == 0) { cat("No overlapping sites.\n\n"); next }

  plots[[1]] <- plots[[1]] + theme(legend.position = "bottom")

  n_reps <- length(plots) / 2
  print(wrap_plots(plots, ncol = 2 * min(n_reps, 2)) +
          plot_layout(guides = "collect") &
          theme(legend.position = "bottom"))

  cat("\n\n")
}
```

# Summary table

Pearson r of each replicate's %pU vs BACS Ψ level (rep1 and rep2) at
overlapping rRNA sites. Lower r indicates weaker agreement with the published
reference.

```{r summary-table}
summary_rows <- list()
for (cl in all_cls) {
  rd <- find_rd(cl)
  files <- list.files(file.path(rd, cl, "out_chisq"),
                      pattern = "filter\\..*\\.rRNA\\.pseusite\\.txt\\.gz$",
                      full.names = TRUE)
  files <- files[!grepl("12-50-|12-47-", files)]
  files <- sort(files)
  if (length(files) == 0) next

  rep_labels <- character(4)
  for (i in seq_along(files)) {
    if (i > 4) break
    short <- sub("^CL\\d+_", "",
                 sub("\\.rRNA\\.pseusite\\.txt\\.gz$", "",
                     sub("^filter\\.", "", basename(files[i]))))
    dt <- tryCatch(
      fread(cmd = paste("zcat", files[i]), header = TRUE, sep = "\t",
            colClasses = list(character = "ref")),
      error = function(e) NULL
    )
    r1 <- NA; r2 <- NA
    if (!is.null(dt) && nrow(dt) > 0) {
      dt <- dt %>%
        mutate(accession = str_extract(ref, "NR_[0-9]+")) %>%
        inner_join(bacs, by = c("accession", "pos" = "pos_num"))
      if (nrow(dt) >= 3) {
        r1 <- round(cor(dt$ratio, dt$bacs_psi_r1, method = "pearson", use = "complete.obs"), 3)
        r2 <- round(cor(dt$ratio, dt$bacs_psi_r2, method = "pearson", use = "complete.obs"), 3)
      }
    }
    rep_labels[i] <- sprintf("%s (r1=%.3f, r2=%.3f)", short, r1, r2)
  }
  for (i in seq_along(rep_labels)) {
    if (rep_labels[i] == "") rep_labels[i] <- ""
  }

  summary_rows[[length(summary_rows)+1]] <- data.frame(
    Cell_line = cl, rep1 = rep_labels[1], rep2 = rep_labels[2],
    rep3 = rep_labels[3], rep4 = rep_labels[4],
    stringsAsFactors = FALSE)
}

summary_df <- do.call(rbind, summary_rows)

knitr::kable(summary_df,
             col.names = c("Cell line", "rep1", "rep2", "rep3", "rep4"),
             caption = "Per-replicate Pearson r of %pU vs BACS (r1=rep1, r2=rep2)")
```

```{r summary-histogram, fig.width=12, fig.height=5}
all_r <- list()
for (cl in all_cls) {
  rd <- find_rd(cl)
  files <- list.files(file.path(rd, cl, "out_chisq"),
                      pattern = "filter\\..*\\.rRNA\\.pseusite\\.txt\\.gz$",
                      full.names = TRUE)
  files <- files[!grepl("12-50-|12-47-", files)]
  for (f in files) {
    short <- sub("^CL\\d+_", "",
                 sub("\\.rRNA\\.pseusite\\.txt\\.gz$", "",
                     sub("^filter\\.", "", basename(f))))
    dt <- tryCatch(
      fread(cmd = paste("zcat", f), header = TRUE, sep = "\t",
            colClasses = list(character = "ref")),
      error = function(e) NULL)
    if (is.null(dt) || nrow(dt) == 0) next
    dt <- dt %>%
      mutate(accession = str_extract(ref, "NR_[0-9]+")) %>%
      inner_join(bacs, by = c("accession", "pos" = "pos_num"))
    if (nrow(dt) < 3) next
    r1 <- cor(dt$ratio, dt$bacs_psi_r1, method = "pearson", use = "complete.obs")
    r2 <- cor(dt$ratio, dt$bacs_psi_r2, method = "pearson", use = "complete.obs")
    all_r[[length(all_r)+1]] <- data.frame(cl = cl, rep = short,
                                            r = c(r1, r2),
                                            bacs_rep = c("vs BACS rep1", "vs BACS rep2"))
  }
}
r_df <- do.call(rbind, all_r)

p1 <- ggplot(r_df[r_df$bacs_rep == "vs BACS rep1", ], aes(x = r)) +
  geom_histogram(binwidth = 0.02, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = median(r_df$r[r_df$bacs_rep == "vs BACS rep1"], na.rm = TRUE),
             color = "red", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = median(r_df$r[r_df$bacs_rep == "vs BACS rep1"], na.rm = TRUE),
           y = Inf, label = sprintf("median = %.3f",
           median(r_df$r[r_df$bacs_rep == "vs BACS rep1"], na.rm = TRUE)),
           hjust = -0.1, vjust = 1.5, size = 4) +
  labs(title = "vs BACS rep1", x = "Pearson r", y = "Count") +
  theme_bw(base_size = 12)

p2 <- ggplot(r_df[r_df$bacs_rep == "vs BACS rep2", ], aes(x = r)) +
  geom_histogram(binwidth = 0.02, fill = "#F28E2B", color = "white", alpha = 0.8) +
  geom_vline(xintercept = median(r_df$r[r_df$bacs_rep == "vs BACS rep2"], na.rm = TRUE),
             color = "red", linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = median(r_df$r[r_df$bacs_rep == "vs BACS rep2"], na.rm = TRUE),
           y = Inf, label = sprintf("median = %.3f",
           median(r_df$r[r_df$bacs_rep == "vs BACS rep2"], na.rm = TRUE)),
           hjust = -0.1, vjust = 1.5, size = 4) +
  labs(title = "vs BACS rep2", x = "Pearson r", y = "Count") +
  theme_bw(base_size = 12)

p1 + p2
```
