---
title: "Quality control using sites called from rRNA"
author: "XSun"
date: "2026-06-06"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

# Introduction

We ran the PUMSEQ pipeline on four cell lines: **CL18870**, **CL18909**,
**CL18916**, and **CL19099**.  Each cell line has four treated replicates.

Pseudouridine (pU) sites are identified from ribosomal RNA (rRNA) reads.  rRNA
is an ideal positive control because its pU sites are abundant, well-conserved,
and independently validated by multiple published methods.  A good sequencing
run should:

1. Show consistent %pU values across replicates of the same cell line.
2. Detect the same set of sites reproducibly across replicates (high overlap).
3. Agree with sites reported in published datasets.

The pseudouridine ratio per site is defined as

$$\%\psi = \frac{C}{C + T}$$

where *C* counts cytosine calls (the signature of reverse-transcriptase stop at
a pU site) and *T* counts unmodified thymidine calls at that position.

We compare our results against two published datasets:

- **BACS** — PMID 39349603: antibody-based capture sequencing.
- **PRAISE** — PMID 36997645: deletion-based direct sequencing.

```{r setup, warning=FALSE, message=FALSE}
library(dplyr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(ggvenn)
library(purrr)
library(patchwork)
library(tidyverse)
library(data.table)
library(scales)
library(knitr)
library(kableExtra)
library(gridExtra)
library(DT)
library(UpSetR)
library(grid)

knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center"
)

cutoffs <- expand.grid(min_depth = c(10, 20),
                       min_ratio = c(0.1, 0.2))
```

---

# Examining replicates for each cell line

For each cell line we run four quality checks across its treated replicates:

1. **Site overlap (Jaccard index)** — how many sites are detected in common
   between each pair of replicates, expressed as a set-similarity score.
2. **%pU distribution** — histogram and density of the modification ratio per
   replicate, used to judge overall signal strength and consistency.
3. **Pairwise Pearson correlation** — quantitative agreement of %pU values at
   shared sites.
4. **Pairwise Pearson heatmap** — compact summary of all pairwise correlations.

A minimum sequencing depth of **`r 20` reads** per site is required before a
site is considered detected.

```{r config-replicate}
RESULTS_DIR <- "/project2/xinhe/xsun/pumseq/pumseq_results"

MPILE_COLS <- c("ref", "start", "pos", "strand", "base",
                "AllDepth", "A", "C", "G", "T_count", "Gap")

CELL_LINES <- list(
  CL18870 = list(
    rep1 = "CL18870_12-187-59",
    rep2 = "CL18870_12-187-60",
    rep3 = "CL18870_12-187-97",
    rep4 = "CL18870_12-187-98"
  ),
  CL18909 = list(
    rep1 = "CL18909_12-187-13",
    rep2 = "CL18909_12-187-14",
    rep3 = "CL18909_13-32-23",
    rep4 = "CL18909_13-32-24"
  ),
  CL18916 = list(
    rep1 = "CL18916_12-187-61",
    rep2 = "CL18916_12-187-62",
    rep3 = "CL18916_12-187-101",
    rep4 = "CL18916_12-187-102"
  ),
  CL19099 = list(
    rep1 = "CL19099_12-187-65",
    rep2 = "CL19099_12-187-66",
    rep3 = "CL19099_12-187-105",
    rep4 = "CL19099_12-187-106"
  )
)

MIN_DEPTH <- 20
```

```{r helpers-replicate}
# ── Data loading ──────────────────────────────────────────────────────────────
read_rRNA_psi <- function(cl, sample_id, mpile_dir) {
  f <- file.path(mpile_dir,
                 sprintf("mpileup.%s.rRNA.pseusite.txt.gz", sample_id))
  if (!file.exists(f)) {
    warning("File not found: ", f)
    return(NULL)
  }
  dt <- fread(cmd = paste("zcat", f), header = FALSE, sep = "\t",
              col.names = MPILE_COLS, data.table = FALSE)
  dt <- dt %>%
    mutate(
      depth = C + T_count,
      psi   = ifelse(depth >= MIN_DEPTH, C / depth, NA_real_),
      site  = paste(ref, pos, sep = ":")
    ) %>%
    filter(!is.na(psi)) %>%
    select(site, psi)
  colnames(dt)[colnames(dt) == "psi"] <- sample_id
  dt
}

# ── Distribution plots ────────────────────────────────────────────────────────
plot_distribution <- function(psi_mat, rep_names, cl) {
  thresholds <- list(
    list(thr = NULL, label = "all sites",   tag = "all"),
    list(thr = 0.1,  label = "%pU >= 0.1", tag = "psi01"),
    list(thr = 0.2,  label = "%pU >= 0.2", tag = "psi02")
  )

  all_plots <- lapply(thresholds, function(th) {
    lapply(rep_names, function(samp) {
      df <- data.frame(psi = psi_mat[[samp]])
      df <- df[!is.na(df$psi), , drop = FALSE]
      if (!is.null(th$thr)) df <- df[df$psi >= th$thr, , drop = FALSE]
      if (nrow(df) == 0) return(NULL)

      med_val  <- median(df$psi)
      mean_val <- mean(df$psi)
      n_01     <- sum(df$psi >= 0.1)
      n_02     <- sum(df$psi >= 0.2)

      ann_text <- if (is.null(th$thr)) {
        sprintf(
          "n = %d\nmedian = %.3f\nmean = %.3f\n>=0.1: %d (%.1f%%)\n>=0.2: %d (%.1f%%)",
          nrow(df), med_val, mean_val,
          n_01, 100 * n_01 / nrow(df),
          n_02, 100 * n_02 / nrow(df)
        )
      } else {
        sprintf("n = %d\nmedian = %.3f\nmean = %.3f", nrow(df), med_val, mean_val)
      }

      vlines <- if (is.null(th$thr)) {
        c(0.1, 0.2)
      } else if (th$thr == 0.1) {
        0.2
      } else {
        NULL
      }

      p <- ggplot(df, aes(x = psi)) +
        geom_histogram(aes(y = after_stat(density)),
                       binwidth = 0.02, fill = "steelblue",
                       color = "white", linewidth = 0.2, alpha = 0.7) +
        geom_density(color = "navy", linewidth = 0.7) +
        geom_vline(xintercept = med_val,  color = "red",
                   linetype = "dashed",  linewidth = 0.6) +
        geom_vline(xintercept = mean_val, color = "darkorange",
                   linetype = "dotted",  linewidth = 0.6) +
        annotate("text", x = Inf, y = Inf,
                 label = ann_text,
                 hjust = 1.05, vjust = 1.05, size = 3, family = "mono") +
        scale_x_continuous(limits = c(if (is.null(th$thr)) 0 else th$thr, 1),
                           breaks = seq(0, 1, 0.1)) +
        labs(
          title    = samp,
          subtitle = th$label,
          x        = "%pU  (C / C+T)",
          y        = "Density"
        ) +
        theme_bw(base_size = 10) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))

      if (!is.null(vlines))
        p <- p + geom_vline(xintercept = vlines, color = "darkgreen",
                            linetype = "longdash", linewidth = 0.5)
      p
    })
  })

  names(all_plots) <- sapply(thresholds, `[[`, "label")
  lapply(all_plots, function(grp) Filter(Negate(is.null), grp))
}

# ── Scatter plots ─────────────────────────────────────────────────────────────
make_scatter_plots_replicate <- function(psi_mat, pairs, cl, thr = NULL) {
  label <- if (is.null(thr)) "all sites" else sprintf("%%pU >= %.1f", thr)

  lapply(pairs, function(p) {
    df <- psi_mat[complete.cases(psi_mat[, p]), ]
    if (!is.null(thr))
      df <- df[df[[p[1]]] >= thr & df[[p[2]]] >= thr, ]
    if (nrow(df) < 5) return(NULL)

    r <- cor(df[[p[1]]], df[[p[2]]], method = "pearson")
    ggplot(df, aes(x = .data[[p[1]]], y = .data[[p[2]]])) +
      geom_point(alpha = 0.4, size = 0.8) +
      geom_abline(slope = 1, intercept = 0,
                  color = "red", linetype = "dashed") +
      annotate("text", x = Inf, y = -Inf,
               label = sprintf("r = %.3f\nn = %d", r, nrow(df)),
               hjust = 1.1, vjust = -0.3, size = 3.5) +
      labs(
        title = sprintf("%s vs %s\n[%s]", p[1], p[2], label),
        x     = sprintf("%%pU  %s", p[1]),
        y     = sprintf("%%pU  %s", p[2])
      ) +
      theme_bw(base_size = 10)
  })
}

# ── Arrange + render a list of plots ─────────────────────────────────────────
render_plots <- function(plot_list, ncol = 3,
                         width_per = 5, height_per = 4.5) {
  plot_list <- Filter(Negate(is.null), plot_list)
  if (length(plot_list) == 0) {
    message("No plots to render.")
    return(invisible(NULL))
  }
  n      <- length(plot_list)
  ncol_p <- min(ncol, n)
  grid.arrange(grobs = plot_list, ncol = ncol_p)
}

# ── Pearson correlation heatmap ───────────────────────────────────────────────
make_cor_heatmap <- function(psi_mat, pairs, rep_names, cl, thr = NULL) {
  label <- if (is.null(thr)) "all sites" else sprintf("%%pU >= %.1f", thr)

  cor_df <- do.call(rbind, lapply(pairs, function(p) {
    df <- psi_mat[complete.cases(psi_mat[, p]), ]
    if (!is.null(thr)) df <- df[df[[p[1]]] >= thr & df[[p[2]]] >= thr, ]
    r <- if (nrow(df) >= 5)
           round(cor(df[[p[1]]], df[[p[2]]], method = "pearson"), 4)
         else NA_real_
    data.frame(rep1 = p[1], rep2 = p[2], r = r,
               n = nrow(df), stringsAsFactors = FALSE)
  }))

  mat <- matrix(NA_real_, nrow = length(rep_names), ncol = length(rep_names),
                dimnames = list(rep_names, rep_names))
  diag(mat) <- 1
  n_mat <- matrix(NA_integer_, nrow = length(rep_names), ncol = length(rep_names),
                  dimnames = list(rep_names, rep_names))
  for (i in seq_len(nrow(cor_df))) {
    mat[cor_df$rep1[i], cor_df$rep2[i]] <- cor_df$r[i]
    mat[cor_df$rep2[i], cor_df$rep1[i]] <- cor_df$r[i]
    n_mat[cor_df$rep1[i], cor_df$rep2[i]] <- cor_df$n[i]
    n_mat[cor_df$rep2[i], cor_df$rep1[i]] <- cor_df$n[i]
  }

  short <- function(s) sub("^[^_]+_", "", s)
  lev   <- short(rep_names)

  plot_df <- expand.grid(x = rep_names, y = rep_names,
                         stringsAsFactors = FALSE) %>%
    rowwise() %>%
    mutate(
      r     = mat[x, y],
      n     = n_mat[x, y],
      label = case_when(
        x == y ~ "1.000",
        is.na(r) ~ "NA",
        TRUE ~ sprintf("%.3f\n(n=%d)", r, n)
      )
    ) %>%
    ungroup() %>%
    mutate(
      x_lab = factor(short(x), levels = lev),
      y_lab = factor(short(y), levels = rev(lev))
    )

  ggplot(plot_df, aes(x = x_lab, y = y_lab, fill = r)) +
    geom_tile(color = "white", linewidth = 0.8) +
    geom_text(aes(label = label), size = 3, lineheight = 0.9) +
    scale_fill_gradient2(
      low  = "#1a9641", mid = "#ffffbf", high = "#d7191c",
      midpoint = 0.5, limits = c(0, 1), na.value = "grey90",
      name = "Pearson r"
    ) +
    labs(title = cl, subtitle = label, x = NULL, y = NULL) +
    coord_fixed() +
    theme_bw(base_size = 10) +
    theme(
      axis.text.x      = element_text(angle = 35, hjust = 1),
      panel.grid       = element_blank(),
      panel.border     = element_blank(),
      legend.key.height = unit(0.8, "cm")
    )
}

make_cor_heatmaps <- function(psi_mat, pairs, rep_names, cl) {
  list(
    make_cor_heatmap(psi_mat, pairs, rep_names, cl, thr = NULL),
    make_cor_heatmap(psi_mat, pairs, rep_names, cl, thr = 0.1),
    make_cor_heatmap(psi_mat, pairs, rep_names, cl, thr = 0.2)
  )
}

# ── Jaccard index helpers ─────────────────────────────────────────────────────
jaccard <- function(a, b) {
  n_int <- length(intersect(a, b))
  n_uni <- length(union(a, b))
  if (n_uni == 0) return(NA_real_)
  n_int / n_uni
}

make_jaccard_heatmap <- function(rep_psi, cl, thr = NULL) {
  label     <- if (is.null(thr)) "all sites" else sprintf("%%pU >= %.1f", thr)
  rep_names <- names(rep_psi)
  pairs     <- combn(rep_names, 2, simplify = FALSE)

  mat <- matrix(NA_real_, length(rep_names), length(rep_names),
                dimnames = list(rep_names, rep_names))
  diag(mat) <- 1
  n_int_mat <- matrix(NA_integer_, length(rep_names), length(rep_names),
                      dimnames = list(rep_names, rep_names))

  for (p in pairs) {
    sa <- rep_psi[[p[1]]]
    sb <- rep_psi[[p[2]]]
    if (!is.null(thr)) {
      sa <- sa[sa >= thr]
      sb <- sb[sb >= thr]
    }
    sites_a <- names(sa)
    sites_b <- names(sb)
    j   <- jaccard(sites_a, sites_b)
    n_i <- length(intersect(sites_a, sites_b))
    mat[p[1], p[2]]       <- j
    mat[p[2], p[1]]       <- j
    n_int_mat[p[1], p[2]] <- n_i
    n_int_mat[p[2], p[1]] <- n_i
  }

  short <- function(s) sub("^[^_]+_", "", s)
  lev   <- short(rep_names)

  plot_df <- expand.grid(x = rep_names, y = rep_names,
                         stringsAsFactors = FALSE) %>%
    rowwise() %>%
    mutate(
      j     = mat[x, y],
      n_int = n_int_mat[x, y],
      label = case_when(
        x == y   ~ "1.000",
        is.na(j) ~ "NA",
        TRUE     ~ sprintf("%.3f\n(n=%d)", j, n_int)
      )
    ) %>%
    ungroup() %>%
    mutate(
      x_lab = factor(short(x), levels = lev),
      y_lab = factor(short(y), levels = rev(lev))
    )

  ggplot(plot_df, aes(x = x_lab, y = y_lab, fill = j)) +
    geom_tile(color = "white", linewidth = 0.8) +
    geom_text(aes(label = label), size = 3, lineheight = 0.9) +
    scale_fill_gradient2(
      low      = "#1a9641", mid = "#ffffbf", high = "#d7191c",
      midpoint = 0.5, limits = c(0, 1), na.value = "grey90",
      name     = "Jaccard"
    ) +
    labs(title = cl, subtitle = label, x = NULL, y = NULL) +
    coord_fixed() +
    theme_bw(base_size = 10) +
    theme(
      axis.text.x       = element_text(angle = 35, hjust = 1),
      panel.grid        = element_blank(),
      panel.border      = element_blank(),
      legend.key.height = unit(0.8, "cm")
    )
}

make_jaccard_heatmaps <- function(rep_psi, cl) {
  list(
    make_jaccard_heatmap(rep_psi, cl, thr = NULL),
    make_jaccard_heatmap(rep_psi, cl, thr = 0.1),
    make_jaccard_heatmap(rep_psi, cl, thr = 0.2)
  )
}
```

```{r run-all-replicate, include=FALSE}
cl_data <- list()

for (cl in names(CELL_LINES)) {
  mpile_dir <- file.path(RESULTS_DIR, cl, "mpile")
  reps      <- CELL_LINES[[cl]]

  rep_list <- lapply(names(reps), function(rep_name) {
    read_rRNA_psi(cl, reps[[rep_name]], mpile_dir)
  })
  names(rep_list) <- names(reps)
  rep_list <- Filter(Negate(is.null), rep_list)

  if (length(rep_list) < 2) {
    cl_data[[cl]] <- list(ok = FALSE, reason = "Not enough replicates found.")
    next
  }

  merged <- Reduce(function(a, b) inner_join(a, b, by = "site"), rep_list)

  if (nrow(merged) < 10) {
    cl_data[[cl]] <- list(ok = FALSE, reason = "Too few shared sites.")
    next
  }

  psi_mat   <- merged[, -1, drop = FALSE]
  rep_names <- colnames(psi_mat)
  pairs     <- combn(rep_names, 2, simplify = FALSE)

  cl_data[[cl]] <- list(
    ok        = TRUE,
    psi_mat   = psi_mat,
    rep_names = rep_names,
    pairs     = pairs,
    n_sites   = nrow(merged),
    site_sets = setNames(
      lapply(rep_list, function(df) df$site),
      sapply(rep_list, function(df) colnames(df)[2])
    ),
    # Named numeric vector (site -> psi) per replicate, used for Jaccard
    rep_psi = setNames(
      lapply(rep_list, function(df) setNames(df[[2]], df$site)),
      sapply(rep_list, function(df) colnames(df)[2])
    )
  )
}
```

---

## CL18870 {.tabset}

```{r v2-cl18870-check, include=FALSE}
cl  <- "CL18870"
dat <- cl_data[[cl]]
```

`r if (!dat$ok) paste("**Skipped:**", dat$reason)`

### Site overlap (Jaccard index)

The **Jaccard index** measures set similarity between two replicates:

$$J(A, B) = \frac{|A \cap B|}{|A \cup B|}$$

A score of 1 means every detected site appears in both replicates; 0 means
no shared sites.  Three thresholds are shown side by side: all sites passing
the depth filter, then those additionally requiring %pU >= 0.1 or %pU >= 0.2.
The intersection count is shown in parentheses.

```{r v2-cl18870-jaccard, eval=dat$ok, fig.width=13.5, fig.height=4.8}
grid.arrange(grobs = make_jaccard_heatmaps(dat$rep_psi, cl), ncol = 3)
```

### %pU Distribution — all sites

Each panel shows the %pU distribution for one replicate (histogram + density
overlay).  The **red dashed** line marks the median; the **orange dotted**
line marks the mean.  **Green dashed** lines indicate the 0.1 and 0.2 ratio
cutoffs.  Consistently shaped distributions across replicates indicate
stable library preparation and sequencing quality.

```{r v2-cl18870-dist-all, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["all sites"]],
             ncol = 2)
```

### %pU Distribution — %pU >= 0.1

Sites restricted to %pU >= 0.1, enriching for confidently modified positions.

```{r v2-cl18870-dist-01, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["%pU >= 0.1"]],
             ncol = 2)
```

### %pU Distribution — %pU >= 0.2

Sites restricted to %pU >= 0.2, the most stringent detection threshold.

```{r v2-cl18870-dist-02, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["%pU >= 0.2"]],
             ncol = 2)
```

### Pairwise correlation — all sites

Scatter plots of %pU at sites detected (depth >= `r MIN_DEPTH`) in **both**
replicates of each pair.  The **red dashed** diagonal is the identity line.
Pearson's r and the number of shared sites are shown in the lower-right corner.

Sites shared across all replicates: **`r if(dat$ok) dat$n_sites`**

```{r v2-cl18870-cor-all, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = NULL))
```

### Pairwise correlation — %pU >= 0.1

```{r v2-cl18870-cor-01, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = 0.1))
```

### Pairwise correlation — %pU >= 0.2

```{r v2-cl18870-cor-02, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = 0.2))
```

### Pairwise Pearson heatmap

Compact summary of all pairwise Pearson correlations.  Each cell shows
Pearson's r and the number of shared sites used.  Three panels correspond to
the three %pU thresholds above.

```{r v2-cl18870-heatmap, eval=dat$ok, fig.width=13.5, fig.height=4.8}
render_plots(make_cor_heatmaps(dat$psi_mat, dat$pairs, dat$rep_names, cl), ncol = 3)
```

---

## CL18909 {.tabset}

```{r v2-cl18909-check, include=FALSE}
cl  <- "CL18909"
dat <- cl_data[[cl]]
```

`r if (!dat$ok) paste("**Skipped:**", dat$reason)`

### Site overlap (Jaccard index)

```{r v2-cl18909-jaccard, eval=dat$ok, fig.width=13.5, fig.height=4.8}
grid.arrange(grobs = make_jaccard_heatmaps(dat$rep_psi, cl), ncol = 3)
```

### %pU Distribution — all sites

```{r v2-cl18909-dist-all, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["all sites"]],
             ncol = 2)
```

### %pU Distribution — %pU >= 0.1

```{r v2-cl18909-dist-01, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["%pU >= 0.1"]],
             ncol = 2)
```

### %pU Distribution — %pU >= 0.2

```{r v2-cl18909-dist-02, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["%pU >= 0.2"]],
             ncol = 2)
```

### Pairwise correlation — all sites

Sites shared across all replicates: **`r if(dat$ok) dat$n_sites`**

```{r v2-cl18909-cor-all, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = NULL))
```

### Pairwise correlation — %pU >= 0.1

```{r v2-cl18909-cor-01, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = 0.1))
```

### Pairwise correlation — %pU >= 0.2

```{r v2-cl18909-cor-02, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = 0.2))
```

### Pairwise Pearson heatmap

```{r v2-cl18909-heatmap, eval=dat$ok, fig.width=13.5, fig.height=4.8}
render_plots(make_cor_heatmaps(dat$psi_mat, dat$pairs, dat$rep_names, cl), ncol = 3)
```

---

## CL18916 {.tabset}

```{r v2-cl18916-check, include=FALSE}
cl  <- "CL18916"
dat <- cl_data[[cl]]
```

`r if (!dat$ok) paste("**Skipped:**", dat$reason)`

### Site overlap (Jaccard index)

```{r v2-cl18916-jaccard, eval=dat$ok, fig.width=13.5, fig.height=4.8}
grid.arrange(grobs = make_jaccard_heatmaps(dat$rep_psi, cl), ncol = 3)
```

### %pU Distribution — all sites

```{r v2-cl18916-dist-all, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["all sites"]],
             ncol = 2)
```

### %pU Distribution — %pU >= 0.1

```{r v2-cl18916-dist-01, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["%pU >= 0.1"]],
             ncol = 2)
```

### %pU Distribution — %pU >= 0.2

```{r v2-cl18916-dist-02, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["%pU >= 0.2"]],
             ncol = 2)
```

### Pairwise correlation — all sites

Sites shared across all replicates: **`r if(dat$ok) dat$n_sites`**

```{r v2-cl18916-cor-all, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = NULL))
```

### Pairwise correlation — %pU >= 0.1

```{r v2-cl18916-cor-01, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = 0.1))
```

### Pairwise correlation — %pU >= 0.2

```{r v2-cl18916-cor-02, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = 0.2))
```

### Pairwise Pearson heatmap

```{r v2-cl18916-heatmap, eval=dat$ok, fig.width=13.5, fig.height=4.8}
render_plots(make_cor_heatmaps(dat$psi_mat, dat$pairs, dat$rep_names, cl), ncol = 3)
```

---

## CL19099 {.tabset}

```{r v2-cl19099-check, include=FALSE}
cl  <- "CL19099"
dat <- cl_data[[cl]]
```

`r if (!dat$ok) paste("**Skipped:**", dat$reason)`

### Site overlap (Jaccard index)

```{r v2-cl19099-jaccard, eval=dat$ok, fig.width=13.5, fig.height=4.8}
grid.arrange(grobs = make_jaccard_heatmaps(dat$rep_psi, cl), ncol = 3)
```

### %pU Distribution — all sites

```{r v2-cl19099-dist-all, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["all sites"]],
             ncol = 2)
```

### %pU Distribution — %pU >= 0.1

```{r v2-cl19099-dist-01, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["%pU >= 0.1"]],
             ncol = 2)
```

### %pU Distribution — %pU >= 0.2

```{r v2-cl19099-dist-02, eval=dat$ok, fig.width=11, fig.height=9}
render_plots(plot_distribution(dat$psi_mat, dat$rep_names, cl)[["%pU >= 0.2"]],
             ncol = 2)
```

### Pairwise correlation — all sites

Sites shared across all replicates: **`r if(dat$ok) dat$n_sites`**

```{r v2-cl19099-cor-all, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = NULL))
```

### Pairwise correlation — %pU >= 0.1

```{r v2-cl19099-cor-01, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = 0.1))
```

### Pairwise correlation — %pU >= 0.2

```{r v2-cl19099-cor-02, eval=dat$ok, fig.width=13.5, fig.height=4.5}
render_plots(make_scatter_plots_replicate(dat$psi_mat, dat$pairs, cl, thr = 0.2))
```

### Pairwise Pearson heatmap

```{r v2-cl19099-heatmap, eval=dat$ok, fig.width=13.5, fig.height=4.8}
render_plots(make_cor_heatmaps(dat$psi_mat, dat$pairs, dat$rep_names, cl), ncol = 3)
```

---

# Comparing pooled data among cell lines

After per-replicate QC, replicates for each cell line are **pooled**: the
depth and C-count across all four replicates are summed, and a pooled %pU ratio
is computed.  Each site also records how many individual replicates it was
detected in (`n_samples`).

We explore the effect of three filtering dimensions simultaneously:

| Dimension | Values |
|-----------|--------|
| Minimum pooled depth | 10, 20 |
| Minimum pooled %pU ratio | 0.1, 0.2 |
| Minimum number of replicates | 1, 2, 3, 4 |

Stricter cutoffs reduce noise at the cost of fewer sites.  Requiring detection
in multiple replicates is a strong reproducibility filter.

```{r setup-comparingcellline, message=FALSE, warning=FALSE}
RESULTS_DIR  <- "/project2/xinhe/xsun/pumseq/pumseq_results"
CELL_LINES_V <- c("CL18870", "CL18909", "CL18916", "CL19099")
DEPTH_CUTS   <- c(10, 20)
RATIO_CUTS   <- c(0.1, 0.2)
NSAMP_CUTS   <- 1:4

load_pooled_rRNA <- function(cl) {
  f <- file.path(RESULTS_DIR, cl, "pooled_sites", "pooled_rRNA.txt")
  dt <- fread(f, header = TRUE, sep = "\t",
              colClasses = list(character = "ref"))
  dt[, cell_line := cl]
  dt[, site := paste(ref, pos, strand, sep = ":")]
  dt
}

count_sites <- function(dt, min_depth, min_ratio, min_nsamp) {
  sum(dt$depth >= min_depth & dt$ratio >= min_ratio & dt$n_samples >= min_nsamp)
}

build_count_grid <- function(dt) {
  expand.grid(
    min_depth  = DEPTH_CUTS,
    min_ratio  = RATIO_CUTS,
    min_nsamp  = NSAMP_CUTS,
    stringsAsFactors = FALSE
  ) |>
    rowwise() |>
    mutate(n_sites = count_sites(dt, min_depth, min_ratio, min_nsamp)) |>
    ungroup() |>
    mutate(
      depth_label = paste0("depth>=", min_depth),
      ratio_label = paste0("ratio>=", min_ratio),
      nsamp_label = paste0("n_samples>=", min_nsamp)
    )
}

plot_count_heatmap <- function(grid, cl_title) {
  grid |>
    mutate(
      depth_label = factor(depth_label, levels = paste0("depth>=", DEPTH_CUTS)),
      ratio_label = factor(ratio_label, levels = paste0("ratio>=", rev(RATIO_CUTS))),
      nsamp_label = factor(nsamp_label,
                           levels = paste0("n_samples>=", NSAMP_CUTS))
    ) |>
    ggplot(aes(x = depth_label, y = ratio_label, fill = n_sites)) +
    geom_tile(color = "white", linewidth = 0.6) +
    geom_text(aes(label = comma(n_sites)), size = 4, fontface = "bold") +
    scale_fill_gradient(low = "#deebf7", high = "#2171b5",
                        name = "# sites", labels = comma) +
    facet_wrap(~nsamp_label, nrow = 1) +
    labs(title = cl_title,
         x = "Depth cutoff (pooled)", y = "Ratio cutoff (pooled %pU)") +
    theme_bw(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          strip.background = element_rect(fill = "#e8eaf6"),
          strip.text = element_text(face = "bold"))
}

dat_list   <- lapply(CELL_LINES_V, load_pooled_rRNA)
names(dat_list) <- CELL_LINES_V

grid_list  <- lapply(dat_list, build_count_grid)
names(grid_list) <- CELL_LINES_V
```

## Per cell-line site counts at different cutoffs

Each heatmap shows the number of rRNA sites surviving a combination of
**pooled depth** (x-axis), **pooled %pU ratio** (y-axis), and **minimum
number of replicates** detecting the site (panels, left to right).
Darker blue = more sites.

---

### CL18870

```{r v2-cl18870-count-heatmap, fig.width=12, fig.height=3.5}
plot_count_heatmap(grid_list[["CL18870"]], "CL18870 – pU site counts")
```

---

### CL18909

```{r v2-cl18909-count-heatmap, fig.width=12, fig.height=3.5}
plot_count_heatmap(grid_list[["CL18909"]], "CL18909 – pU site counts")
```

---

### CL18916

```{r v2-cl18916-count-heatmap, fig.width=12, fig.height=3.5}
plot_count_heatmap(grid_list[["CL18916"]], "CL18916 – pU site counts")
```

---

### CL19099

```{r v2-cl19099-count-heatmap, fig.width=12, fig.height=3.5}
plot_count_heatmap(grid_list[["CL19099"]], "CL19099 – pU site counts")
```

---

## Cross-cell-line comparison of site counts

We use the **Jaccard index** to quantify how much the detected site sets
overlap between every pair of cell lines.  For each combination of depth,
ratio, and n_samples cutoffs, sites passing all filters in each cell line are
extracted and the pairwise Jaccard is computed:

$$J(A, B) = \frac{|A \cap B|}{|A \cup B|}$$

A high Jaccard index means the two cell lines call largely the same rRNA
positions, consistent with constitutively modified sites shared across cell
lines.  Each cell shows the Jaccard score and, in parentheses, the number of
sites in the intersection.  The layout mirrors the Pearson heatmap below:
rows = minimum n_samples, columns = depth × ratio cutoff.

```{r v2-cross-cl-comparison, fig.width=14, fig.height=11}
cl_pairs_v <- combn(CELL_LINES_V, 2, simplify = FALSE)

jaccard_cl_rows <- do.call(rbind, lapply(cl_pairs_v, function(p) {
  do.call(rbind, lapply(DEPTH_CUTS, function(d) {
    do.call(rbind, lapply(RATIO_CUTS, function(r) {
      do.call(rbind, lapply(NSAMP_CUTS, function(n) {
        sites1  <- dat_list[[p[1]]][depth >= d & ratio >= r & n_samples >= n, site]
        sites2  <- dat_list[[p[2]]][depth >= d & ratio >= r & n_samples >= n, site]
        n_int   <- length(intersect(sites1, sites2))
        n_uni   <- length(union(sites1, sites2))
        j       <- if (n_uni == 0) NA_real_ else n_int / n_uni
        data.frame(cl1 = p[1], cl2 = p[2],
                   min_depth = d, min_ratio = r, min_nsamp = n,
                   jaccard = round(j, 3), n_intersect = n_int)
      }))
    }))
  }))
}))

jaccard_cl_sym <- jaccard_cl_rows |>
  bind_rows(jaccard_cl_rows |> rename(cl1 = cl2, cl2 = cl1)) |>
  bind_rows(
    expand.grid(cl1 = CELL_LINES_V, cl2 = CELL_LINES_V,
                min_depth = DEPTH_CUTS, min_ratio = RATIO_CUTS,
                min_nsamp = NSAMP_CUTS) |>
      filter(cl1 == cl2) |>
      mutate(jaccard = 1, n_intersect = NA_integer_)
  ) |>
  mutate(
    cl1         = factor(cl1, levels = CELL_LINES_V),
    cl2         = factor(cl2, levels = rev(CELL_LINES_V)),
    nsamp_label = factor(paste0("n_samples>=", min_nsamp),
                         levels = paste0("n_samples>=", NSAMP_CUTS)),
    dr_label    = factor(
      paste0("d>=", min_depth, " r>=", min_ratio),
      levels = apply(expand.grid(d = DEPTH_CUTS, r = RATIO_CUTS), 1,
                     function(x) paste0("d>=", x["d"], " r>=", x["r"]))
    ),
    cell_label  = case_when(
      cl1 == cl2   ~ "1.000",
      is.na(jaccard) ~ "NA",
      TRUE         ~ sprintf("%.3f\n(n=%d)", jaccard, n_intersect)
    )
  )

ggplot(jaccard_cl_sym, aes(x = cl1, y = cl2, fill = jaccard)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = cell_label), size = 2.8, fontface = "bold",
            lineheight = 0.9) +
  scale_fill_gradient2(
    low = "#1a9641", mid = "#ffffbf", high = "#d7191c",
    midpoint = 0.5, limits = c(0, 1), na.value = "grey90",
    name = "Jaccard"
  ) +
  facet_grid(nsamp_label ~ dr_label) +
  labs(
    title = "Jaccard index of site overlap between cell lines (rRNA pooled sites)",
    x = NULL, y = NULL
  ) +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1, size = 9),
    axis.text.y      = element_text(size = 9),
    strip.background = element_rect(fill = "#e8eaf6"),
    strip.text       = element_text(face = "bold", size = 9),
    panel.spacing    = unit(0.3, "lines"),
    legend.position  = "right"
  )
```

---

## Pearson's r of %pU between cell lines

To assess whether the same sites are similarly modified across cell lines, we
compute Pearson's r on the pooled %pU values at sites passing all three cutoff
filters in **both** cell lines of each pair.  High r across cell lines
suggests that the detected modification pattern is consistent — i.e., the
same rRNA sites are strongly modified regardless of cell line — which is
expected for constitutively modified rRNA residues.

Sites that pass in one cell line but not the other are excluded from the
correlation (they contribute to the Jaccard-style overlap question, not the
quantitative agreement question).

```{r v2-pearson-setup}
pearson_pair <- function(dt1, dt2, min_depth, min_ratio, min_nsamp) {
  pass1 <- dt1[depth >= min_depth & ratio >= min_ratio & n_samples >= min_nsamp,
               .(site, ratio1 = ratio)]
  pass2 <- dt2[depth >= min_depth & ratio >= min_ratio & n_samples >= min_nsamp,
               .(site, ratio2 = ratio)]
  merged <- merge(pass1, pass2, by = "site")
  if (nrow(merged) < 3) return(data.frame(r = NA_real_, n_common = nrow(merged)))
  data.frame(
    r        = cor(merged$ratio1, merged$ratio2, method = "pearson"),
    n_common = nrow(merged)
  )
}

cl_pairs <- combn(CELL_LINES_V, 2, simplify = FALSE)

cor_grid <- do.call(rbind, lapply(cl_pairs, function(p) {
  do.call(rbind, lapply(DEPTH_CUTS, function(d) {
    do.call(rbind, lapply(RATIO_CUTS, function(r) {
      do.call(rbind, lapply(NSAMP_CUTS, function(n) {
        res <- pearson_pair(
          dat_list[[p[1]]], dat_list[[p[2]]],
          min_depth = d, min_ratio = r, min_nsamp = n
        )
        data.frame(cl1 = p[1], cl2 = p[2],
                   min_depth = d, min_ratio = r, min_nsamp = n,
                   pearson_r = round(res$r, 3),
                   n_common  = res$n_common)
      }))
    }))
  }))
}))

cor_sym <- cor_grid |>
  bind_rows(
    cor_grid |> rename(cl1 = cl2, cl2 = cl1)
  ) |>
  bind_rows(
    expand.grid(cl1 = CELL_LINES_V, cl2 = CELL_LINES_V,
                min_depth = DEPTH_CUTS, min_ratio = RATIO_CUTS,
                min_nsamp = NSAMP_CUTS) |>
      filter(cl1 == cl2) |>
      mutate(pearson_r = 1, n_common = NA_integer_)
  ) |>
  mutate(
    cl1         = factor(cl1, levels = CELL_LINES_V),
    cl2         = factor(cl2, levels = rev(CELL_LINES_V)),
    nsamp_label = factor(paste0("n_samples>=", min_nsamp),
                         levels = paste0("n_samples>=", NSAMP_CUTS)),
    dr_label    = factor(
      paste0("d>=", min_depth, " r>=", min_ratio),
      levels = apply(
        expand.grid(d = DEPTH_CUTS, r = RATIO_CUTS), 1,
        function(x) paste0("d>=", x["d"], " r>=", x["r"])
      )
    )
  )
```

### Correlation matrices

The grid of heatmaps below shows Pearson's r between every pair of cell lines
under all combinations of cutoffs.  Rows = minimum n_samples; columns = depth
× ratio cutoff.  Diagonal cells equal 1 by definition.  Warmer (red) colours
indicate higher agreement; a grey cell means too few common sites to compute r.

```{r v2-pearson-heatmap, fig.width=14, fig.height=11}
ggplot(cor_sym, aes(x = cl1, y = cl2, fill = pearson_r)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = ifelse(cl1 == cl2, "1", sprintf("%.3f", pearson_r))),
            size = 3, fontface = "bold") +
  scale_fill_gradient2(
    low = "#1a9641", mid = "#ffffbf", high = "#d7191c",
    midpoint = 0.5, limits = c(0, 1), na.value = "grey90",
    name = "Pearson r"
  ) +
  facet_grid(nsamp_label ~ dr_label) +
  labs(
    title = "Pearson's r of pooled %pU between cell lines (rRNA sites)",
    x = NULL, y = NULL
  ) +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1, size = 9),
    axis.text.y      = element_text(size = 9),
    strip.background = element_rect(fill = "#e8eaf6"),
    strip.text       = element_text(face = "bold", size = 9),
    panel.spacing    = unit(0.3, "lines"),
    legend.position  = "right"
  )
```

### Scatter plots (depth >= 20, ratio >= 0.1, n_samples >= 2)

Representative scatter plots at a moderate cutoff (depth >= 20, %pU >= 0.1,
detected in >= 2 replicates).  Each point is one rRNA site.

```{r v2-pearson-scatter, fig.width=13, fig.height=8}
MIN_D <- 20; MIN_R <- 0.1; MIN_N <- 2

scatter_data <- lapply(CELL_LINES_V, function(cl) {
  dat_list[[cl]][depth >= MIN_D & ratio >= MIN_R & n_samples >= MIN_N,
                 .(site, ratio, cell_line = cl)]
})
names(scatter_data) <- CELL_LINES_V

scatter_plots <- lapply(cl_pairs, function(p) {
  d1 <- scatter_data[[p[1]]]
  d2 <- scatter_data[[p[2]]]
  m  <- merge(d1, d2, by = "site", suffixes = c("_1", "_2"))
  if (nrow(m) < 3) return(NULL)
  r_val <- cor(m$ratio_1, m$ratio_2, method = "pearson")
  ggplot(m, aes(x = ratio_1, y = ratio_2)) +
    geom_point(alpha = 0.55, size = 1.2, color = "#2c7bb6") +
    geom_abline(slope = 1, intercept = 0, color = "red",
                linetype = "dashed", linewidth = 0.6) +
    annotate("text", x = Inf, y = -Inf,
             label = sprintf("r = %.3f\nn = %d", r_val, nrow(m)),
             hjust = 1.1, vjust = -0.3, size = 3.5) +
    scale_x_continuous(labels = percent_format(accuracy = 1)) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      title = paste(p[1], "vs", p[2]),
      x     = paste0("%pU  ", p[1]),
      y     = paste0("%pU  ", p[2])
    ) +
    theme_bw(base_size = 10)
})
scatter_plots <- Filter(Negate(is.null), scatter_plots)

do.call(grid.arrange, c(scatter_plots, list(ncol = 3)))
```

---

# Comparing with published data

We benchmark our pooled rRNA calls against two independent published datasets.
Overlap with validated sites confirms that the pipeline is correctly identifying
true pseudouridine modifications.

- **BACS** (PMID 39349603) detects pU via antibody-based capture; it reports
  absolute %pU levels for two replicates.
- **PRAISE** (PMID 36997645) detects pU via the deletion signature at reverse-
  transcription termination; it reports deletion ratios for two replicates.

For each cell line and each combination of depth / ratio cutoffs we report:
(a) how many of our pooled sites overlap with BACS sites, (b) how many overlap
with PRAISE sites, and (c) scatter plots comparing the quantitative %pU or
deletion ratio values at overlapping positions.

```{r v2-pub-helpers}
make_scatter_publication <- function(df, x, y, col, ttl, xlab, ylab) {
  if (nrow(df) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = "No overlapping sites",
                 size = 4, color = "grey50") +
        labs(title = ttl, x = xlab, y = ylab) +
        theme_classic(base_size = 11) +
        theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))
    )
  }
  ggplot(df, aes(x = .data[[x]], y = .data[[y]],
                 color = .data[[col]], label = label)) +
    geom_point(size = 2.5, alpha = 0.8) +
    geom_text_repel(size = 2.5, max.overlaps = 20) +
    geom_smooth(method = "lm", se = TRUE, color = "grey40", linetype = "dashed") +
    scale_color_manual(values = c("5.8S" = "#4E79A7", "18S" = "#F28E2B",
                                  "28S"  = "#59A14F", "5S"  = "#E15759")) +
    labs(title = ttl, x = xlab, y = ylab, color = "rRNA type") +
    theme_classic(base_size = 11) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))
}

make_heatmap_publication <- function(data, fill_col, label_col, ttl, fill_label) {
  ggplot(data, aes(x = min_ratio, y = min_depth, fill = .data[[fill_col]])) +
    geom_tile(color = "white", linewidth = 1.2) +
    geom_text(aes(label = .data[[label_col]]), size = 6, fontface = "bold", color = "white") +
    scale_fill_gradient(low = "#d0e8f2", high = "#1a6fa8") +
    labs(title = ttl, x = "Min ratio cutoff", y = "Min depth cutoff", fill = fill_label) +
    theme_classic(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          axis.text  = element_text(size = 11))
}

combine_panels <- function(results, slot) {
  plots <- map(results, slot)
  wrap_plots(plots, ncol = 2) +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
}
```

## Published data overview

Site counts and the overlap between BACS and PRAISE are shown first to
establish the published reference landscape before comparing with our data.

```{r v2-pub-load, warning=FALSE, message=FALSE}
path_BACS   <- "/project2/xinhe/xsun/pumseq/data_others/pU_PMID39349603.xlsx"
path_PRAISE <- "/project2/xinhe/xsun/pumseq/data_others/pU_PMID36997645.xlsx"

rrna_BACS <- readxl::read_excel(path_BACS, sheet = 2)
colnames(rrna_BACS) <- rrna_BACS[2, ]
rrna_BACS <- rrna_BACS[-c(1:2), ]
rrna_BACS <- rrna_BACS[complete.cases(rrna_BACS$chr), ]

rrna_PRAISE <- readxl::read_excel(path_PRAISE, sheet = 1)
colnames(rrna_PRAISE) <- rrna_PRAISE[2, ]
rrna_PRAISE <- rrna_PRAISE[-c(1:2), ]
rrna_PRAISE <- rrna_PRAISE %>% filter(!is.na(chr_name), !is.na(site))

rrna_BACS_clean <- rrna_BACS %>%
  mutate(
    rna_type = case_when(
      str_detect(chr, "RNA5_8S") ~ "5.8S",
      str_detect(chr, "RNA18S")  ~ "18S",
      str_detect(chr, "RNA28S")  ~ "28S",
      str_detect(chr, "RNA5SN")  ~ "5S",
      TRUE ~ NA_character_
    ),
    pos_num = as.integer(str_extract(pos, "^[0-9]+"))
  )

rrna_PRAISE_clean <- rrna_PRAISE %>%
  mutate(
    rna_type  = str_extract(chr_name, "(5\\.8S|18S|28S|5S)"),
    pos_start = as.integer(str_extract(site, "^[0-9]+")),
    pos_end   = as.integer(coalesce(
      str_extract(site, "(?<=-)[0-9]+$"),
      str_extract(site, "^[0-9]+")
    ))
  )

overlap <- rrna_BACS_clean %>%
  inner_join(rrna_PRAISE_clean, by = "rna_type") %>%
  filter(pos_num >= pos_start & pos_num <= pos_end)

DT::datatable(rrna_BACS, caption = htmltools::tags$caption(
  style = 'caption-side: left; text-align: left; color:black; font-size:150%;',
  'Sites from BACS'), options = list(pageLength = 5))

DT::datatable(rrna_PRAISE, caption = htmltools::tags$caption(
  style = 'caption-side: left; text-align: left; color:black; font-size:150%;',
  'Sites from PRAISE'), options = list(pageLength = 5))

cat("Sites in BACS:    ", nrow(rrna_BACS_clean), "\n")
cat("Sites in PRAISE:  ", nrow(rrna_PRAISE_clean), "\n")
cat("Overlapping sites:", nrow(overlap), "\n")

rrna_BACS_clean <- rrna_BACS_clean %>%
  mutate(accession = str_extract(chr, "NR_[0-9]+"),
         pos_num   = as.integer(pos)) %>%
  rename(`pU level_rep1` = `Ψ level_rep1`,
         `pU level_rep2` = `Ψ level_rep2`)

rrna_PRAISE_clean <- rrna_PRAISE_clean %>%
  mutate(accession = str_extract(chr_name, "NR_[0-9]+"))
```

```{r v2-pub-preload, include=FALSE}
run_cl_comparison <- function(cl) {
  pooled_path <- file.path(RESULTS_DIR, cl, "pooled_sites", "pooled_rRNA.txt")
  if (!file.exists(pooled_path)) {
    warning("File not found: ", pooled_path)
    return(NULL)
  }

  pooled_ours_rrna <- fread(pooled_path)

  results <- lapply(seq_len(nrow(cutoffs)), function(i) {
    min_depth <- cutoffs$min_depth[i]
    min_ratio <- cutoffs$min_ratio[i]
    tag       <- paste0("depth>=", min_depth, "  ratio>=", min_ratio)

    ours_filtered <- pooled_ours_rrna %>%
      filter(depth >= min_depth, ratio >= min_ratio) %>%
      mutate(accession = str_extract(ref, "NR_[0-9]+"))

    overlap_bacs <- ours_filtered %>%
      inner_join(rrna_BACS_clean, by = c("accession", "pos" = "pos_num")) %>%
      mutate(
        psi_rep1 = as.numeric(`pU level_rep1`),
        psi_rep2 = as.numeric(`pU level_rep2`),
        label    = paste0(name, " pos", pos)
      )

    overlap_praise <- ours_filtered %>%
      inner_join(rrna_PRAISE_clean, by = "accession", relationship = "many-to-many") %>%
      filter(pos >= pos_start & pos <= pos_end) %>%
      mutate(
        deletion_rep1 = as.numeric(rep1_deletion_ratio),
        deletion_rep2 = as.numeric(rep2_deletion_ratio),
        label         = paste0(rna_type, " pos", pos)
      )

    list(
      tag            = tag,
      min_depth      = min_depth,
      min_ratio      = min_ratio,
      n_ours         = nrow(ours_filtered),
      n_bacs         = nrow(overlap_bacs),
      n_praise       = nrow(overlap_praise),
      overlap_bacs   = overlap_bacs,
      overlap_praise = overlap_praise,

      p_bacs_r1   = make_scatter_publication(overlap_bacs, "ratio", "psi_rep1", "name",
                                             tag, "Our pU ratio", "BACS pU level (rep1)"),
      p_bacs_r2   = make_scatter_publication(overlap_bacs, "ratio", "psi_rep2", "name",
                                             tag, "Our pU ratio", "BACS pU level (rep2)"),
      p_praise_r1 = make_scatter_publication(overlap_praise, "ratio", "deletion_rep1", "rna_type",
                                             tag, "Our pU ratio", "PRAISE deletion ratio (rep1)"),
      p_praise_r2 = make_scatter_publication(overlap_praise, "ratio", "deletion_rep2", "rna_type",
                                             tag, "Our pU ratio", "PRAISE deletion ratio (rep2)")
    )
  })

  names(results) <- paste0("depth", cutoffs$min_depth, "_ratio", cutoffs$min_ratio)
  results
}

pub_data <- lapply(c("CL18870", "CL18909", "CL18916", "CL19099"), run_cl_comparison)
names(pub_data) <- c("CL18870", "CL18909", "CL18916", "CL19099")
```

---

## CL18870 {.tabset}

```{r v2-cl18870-pub-check, include=FALSE}
cl      <- "CL18870"
results <- pub_data[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled file not found for", cl)`

### Site counts

Tile colour indicates the number of sites: **Our sites** (left) are those
passing the given depth and ratio cutoff in our pooled data; **Overlap with
BACS/PRAISE** (middle, right) counts how many of those sites are also
present in the respective published dataset.  Higher overlap validates the
specificity of our site calls.

```{r v2-cl18870-pub-heatmap, eval=ok, fig.width=12, fig.height=4}
heatmap_data <- map_dfr(results, ~ tibble(
  min_depth = .x$min_depth, min_ratio = .x$min_ratio,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(min_depth = factor(paste0("depth>=", min_depth)),
         min_ratio = factor(paste0("ratio>=", min_ratio)))

make_heatmap_publication(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, "  site counts across depth & ratio cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

### vs BACS (rep1)

Scatter plots of our pooled %pU ratio (x-axis) vs. the BACS %pU level from
replicate 1 (y-axis) at overlapping sites, faceted by depth/ratio cutoff.
Points are coloured by rRNA subunit type.  A positive trend indicates
quantitative agreement with the BACS dataset.

```{r v2-cl18870-pub-bacs-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS pU level (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs BACS (rep2)

```{r v2-cl18870-pub-bacs-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS pU level (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

Our pooled %pU ratio vs. the PRAISE deletion ratio from replicate 1.  PRAISE
uses a different biochemical readout, so the absolute scale differs; we expect
a monotonic positive relationship at high-confidence sites.

```{r v2-cl18870-pub-praise-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r v2-cl18870-pub-praise-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

---

## CL18909 {.tabset}

```{r v2-cl18909-pub-check, include=FALSE}
cl      <- "CL18909"
results <- pub_data[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled file not found for", cl)`

### Site counts

```{r v2-cl18909-pub-heatmap, eval=ok, fig.width=12, fig.height=4}
heatmap_data <- map_dfr(results, ~ tibble(
  min_depth = .x$min_depth, min_ratio = .x$min_ratio,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(min_depth = factor(paste0("depth>=", min_depth)),
         min_ratio = factor(paste0("ratio>=", min_ratio)))

make_heatmap_publication(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, "  site counts across depth & ratio cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

### vs BACS (rep1)

```{r v2-cl18909-pub-bacs-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS pU level (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs BACS (rep2)

```{r v2-cl18909-pub-bacs-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS pU level (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

```{r v2-cl18909-pub-praise-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r v2-cl18909-pub-praise-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

---

## CL18916 {.tabset}

```{r v2-cl18916-pub-check, include=FALSE}
cl      <- "CL18916"
results <- pub_data[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled file not found for", cl)`

### Site counts

```{r v2-cl18916-pub-heatmap, eval=ok, fig.width=12, fig.height=4}
heatmap_data <- map_dfr(results, ~ tibble(
  min_depth = .x$min_depth, min_ratio = .x$min_ratio,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(min_depth = factor(paste0("depth>=", min_depth)),
         min_ratio = factor(paste0("ratio>=", min_ratio)))

make_heatmap_publication(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, "  site counts across depth & ratio cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

### vs BACS (rep1)

```{r v2-cl18916-pub-bacs-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS pU level (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs BACS (rep2)

```{r v2-cl18916-pub-bacs-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS pU level (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

```{r v2-cl18916-pub-praise-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r v2-cl18916-pub-praise-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

---

## CL19099 {.tabset}

```{r v2-cl19099-pub-check, include=FALSE}
cl      <- "CL19099"
results <- pub_data[[cl]]
ok      <- !is.null(results)
```

`r if (!ok) paste("**Skipped:** pooled file not found for", cl)`

### Site counts

```{r v2-cl19099-pub-heatmap, eval=ok, fig.width=12, fig.height=4}
heatmap_data <- map_dfr(results, ~ tibble(
  min_depth = .x$min_depth, min_ratio = .x$min_ratio,
  n_ours = .x$n_ours, n_bacs = .x$n_bacs, n_praise = .x$n_praise
)) %>%
  mutate(min_depth = factor(paste0("depth>=", min_depth)),
         min_ratio = factor(paste0("ratio>=", min_ratio)))

make_heatmap_publication(heatmap_data, "n_ours",   "n_ours",   "Our sites passing cutoffs",  "# sites") +
make_heatmap_publication(heatmap_data, "n_bacs",   "n_bacs",   "Overlap with BACS",          "# overlapping sites") +
make_heatmap_publication(heatmap_data, "n_praise", "n_praise", "Overlap with PRAISE",        "# overlapping sites") +
plot_annotation(title = paste(cl, "  site counts across depth & ratio cutoffs"),
                theme = theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13)))
```

### vs BACS (rep1)

```{r v2-cl19099-pub-bacs-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS pU level (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs BACS (rep2)

```{r v2-cl19099-pub-bacs-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_bacs_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs BACS pU level (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep1)

```{r v2-cl19099-pub-praise-r1, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r1") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep1)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

### vs PRAISE (rep2)

```{r v2-cl19099-pub-praise-r2, eval=ok, fig.width=8, fig.height=9}
combine_panels(results, "p_praise_r2") +
  plot_annotation(title = paste(cl, "— Our pU ratio vs PRAISE deletion ratio (rep2)"),
                  theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```
