---
title: "Replicate site overlap — Jaccard index"
author: "XSun"
date: "2026-06-06"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

# Introduction

For each cell line we quantify how reproducibly the same sites are detected
across replicates by computing the **Jaccard index** between every pair of
replicates:

$$J(A,B) = \frac{|A \cap B|}{|A \cup B|}$$

A value of 1 means perfect overlap; 0 means no shared sites.

We report the index at three detection thresholds:

- **All sites**: any site passing the depth filter (≥ `r 20` reads)
- **%ψ ≥ 0.1**: sites also requiring a C/(C+T) ratio ≥ 0.1
- **%ψ ≥ 0.2**: sites also requiring a C/(C+T) ratio ≥ 0.2

```{r setup, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(data.table)
library(gridExtra)

knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center"
)
```

# Examining replicates for each cell line

## Site overlap within replicates (Jaccard index)

```{r config}
RESULTS_DIR <- "/project2/xinhe/xsun/pumseq/pumseq_results"

MPILE_COLS <- c("ref", "start", "pos", "strand", "base",
                "AllDepth", "A", "C", "G", "T_count", "Gap")

CELL_LINES <- list(
  CL18870 = list(
    rep1 = "CL18870_12-187-59",
    rep2 = "CL18870_12-187-60",
    rep3 = "CL18870_12-187-97",
    rep4 = "CL18870_12-187-98"
  ),
  CL18909 = list(
    rep1 = "CL18909_12-187-13",
    rep2 = "CL18909_12-187-14",
    rep3 = "CL18909_13-32-23",
    rep4 = "CL18909_13-32-24"
  ),
  CL18916 = list(
    rep1 = "CL18916_12-187-61",
    rep2 = "CL18916_12-187-62",
    rep3 = "CL18916_12-187-101",
    rep4 = "CL18916_12-187-102"
  ),
  CL19099 = list(
    rep1 = "CL19099_12-187-65",
    rep2 = "CL19099_12-187-66",
    rep3 = "CL19099_12-187-105",
    rep4 = "CL19099_12-187-106"
  )
)

MIN_DEPTH <- 20
```

```{r helpers}
read_rRNA_psi_full <- function(cl, sample_id, mpile_dir) {
  f <- file.path(mpile_dir,
                 sprintf("mpileup.%s.rRNA.pseusite.txt.gz", sample_id))
  if (!file.exists(f)) {
    warning("File not found: ", f)
    return(NULL)
  }
  dt <- fread(cmd = paste("zcat", f), header = FALSE, sep = "\t",
              col.names = MPILE_COLS, data.table = FALSE)
  dt %>%
    mutate(
      depth = C + T_count,
      psi   = ifelse(depth >= MIN_DEPTH, C / depth, NA_real_),
      site  = paste(ref, pos, sep = ":")
    ) %>%
    filter(!is.na(psi)) %>%
    select(site, psi)
}

jaccard <- function(a, b) {
  n_int <- length(intersect(a, b))
  n_uni <- length(union(a, b))
  if (n_uni == 0) return(NA_real_)
  n_int / n_uni
}

make_jaccard_heatmap <- function(rep_data, cl, thr = NULL) {
  label <- if (is.null(thr)) "all sites" else sprintf("%%Ψ ≥ %.1f", thr)

  rep_names <- names(rep_data)
  pairs     <- combn(rep_names, 2, simplify = FALSE)

  # Build symmetric Jaccard matrix
  mat <- matrix(NA_real_,
                nrow = length(rep_names), ncol = length(rep_names),
                dimnames = list(rep_names, rep_names))
  diag(mat) <- 1

  n_mat <- matrix(NA_integer_,
                  nrow = length(rep_names), ncol = length(rep_names),
                  dimnames = list(rep_names, rep_names))

  for (p in pairs) {
    sites_a <- rep_data[[p[1]]]
    sites_b <- rep_data[[p[2]]]

    if (!is.null(thr)) {
      sites_a <- sites_a[rep_data[[p[1]]] >= thr]   # named vector: names = site
      sites_b <- sites_b[rep_data[[p[2]]] >= thr]
    }

    j   <- jaccard(names(sites_a), names(sites_b))
    n_i <- length(intersect(names(sites_a), names(sites_b)))

    mat[p[1], p[2]] <- j
    mat[p[2], p[1]] <- j
    n_mat[p[1], p[2]] <- n_i
    n_mat[p[2], p[1]] <- n_i
  }

  # Tidy for ggplot
  short <- function(s) sub("^[^_]+_", "", s)
  lev   <- short(rep_names)

  plot_df <- expand.grid(x = rep_names, y = rep_names,
                         stringsAsFactors = FALSE) %>%
    rowwise() %>%
    mutate(
      j     = mat[x, y],
      n_int = n_mat[x, y],
      label = case_when(
        x == y  ~ "1.000",
        is.na(j) ~ "NA",
        TRUE    ~ sprintf("%.3f\n(n=%d)", j, n_int)
      )
    ) %>%
    ungroup() %>%
    mutate(
      x_lab = factor(short(x), levels = lev),
      y_lab = factor(short(y), levels = rev(lev))
    )

  ggplot(plot_df, aes(x = x_lab, y = y_lab, fill = j)) +
    geom_tile(color = "white", linewidth = 0.8) +
    geom_text(aes(label = label), size = 3, lineheight = 0.9) +
    scale_fill_gradient2(
      low      = "#1a9641",
      mid      = "#ffffbf",
      high     = "#d7191c",
      midpoint = 0.5,
      limits   = c(0, 1),
      na.value = "grey90",
      name     = "Jaccard"
    ) +
    labs(
      title    = cl,
      subtitle = label,
      x        = NULL,
      y        = NULL
    ) +
    coord_fixed() +
    theme_bw(base_size = 10) +
    theme(
      axis.text.x       = element_text(angle = 35, hjust = 1),
      panel.grid        = element_blank(),
      panel.border      = element_blank(),
      legend.key.height = unit(0.8, "cm")
    )
}

make_jaccard_heatmaps <- function(rep_data, cl) {
  list(
    make_jaccard_heatmap(rep_data, cl, thr = NULL),
    make_jaccard_heatmap(rep_data, cl, thr = 0.1),
    make_jaccard_heatmap(rep_data, cl, thr = 0.2)
  )
}
```

```{r load-data, include=FALSE}
# Load per-replicate site+psi as named vectors (names = site IDs)
cl_rep_data <- list()

for (cl in names(CELL_LINES)) {
  mpile_dir <- file.path(RESULTS_DIR, cl, "mpile")
  reps      <- CELL_LINES[[cl]]

  rep_list <- lapply(names(reps), function(rep_name) {
    df <- read_rRNA_psi_full(cl, reps[[rep_name]], mpile_dir)
    if (is.null(df)) return(NULL)
    v <- setNames(df$psi, df$site)   # named numeric vector: site -> psi
    v
  })
  names(rep_list) <- names(reps)
  rep_list <- Filter(Negate(is.null), rep_list)

  cl_rep_data[[cl]] <- list(
    ok       = length(rep_list) >= 2,
    rep_data = rep_list,
    reason   = if (length(rep_list) < 2) "Not enough replicates found." else NULL
  )
}
```

---

### CL18870 {.tabset}

```{r cl18870-jac-check, include=FALSE}
cl  <- "CL18870"
dat <- cl_rep_data[[cl]]
```

`r if (!dat$ok) paste("**Skipped:**", dat$reason)`

#### Jaccard heatmap

```{r cl18870-jaccard, eval=dat$ok, fig.width=13.5, fig.height=4.8}
gridExtra::grid.arrange(
  grobs = make_jaccard_heatmaps(dat$rep_data, cl),
  ncol  = 3
)
```

---

### CL18909 {.tabset}

```{r cl18909-jac-check, include=FALSE}
cl  <- "CL18909"
dat <- cl_rep_data[[cl]]
```

`r if (!dat$ok) paste("**Skipped:**", dat$reason)`

#### Jaccard heatmap

```{r cl18909-jaccard, eval=dat$ok, fig.width=13.5, fig.height=4.8}
gridExtra::grid.arrange(
  grobs = make_jaccard_heatmaps(dat$rep_data, cl),
  ncol  = 3
)
```

---

### CL18916 {.tabset}

```{r cl18916-jac-check, include=FALSE}
cl  <- "CL18916"
dat <- cl_rep_data[[cl]]
```

`r if (!dat$ok) paste("**Skipped:**", dat$reason)`

#### Jaccard heatmap

```{r cl18916-jaccard, eval=dat$ok, fig.width=13.5, fig.height=4.8}
gridExtra::grid.arrange(
  grobs = make_jaccard_heatmaps(dat$rep_data, cl),
  ncol  = 3
)
```

---

### CL19099 {.tabset}

```{r cl19099-jac-check, include=FALSE}
cl  <- "CL19099"
dat <- cl_rep_data[[cl]]
```

`r if (!dat$ok) paste("**Skipped:**", dat$reason)`

#### Jaccard heatmap

```{r cl19099-jaccard, eval=dat$ok, fig.width=13.5, fig.height=4.8}
gridExtra::grid.arrange(
  grobs = make_jaccard_heatmaps(dat$rep_data, cl),
  ncol  = 3
)
```

---

### Summary table — all cell lines

Pairwise Jaccard indices across all cell lines and thresholds.

```{r summary-table}
library(knitr)
library(kableExtra)

summary_rows <- list()

for (cl in names(CELL_LINES)) {
  dat <- cl_rep_data[[cl]]
  if (!dat$ok) next
  rep_names <- names(dat$rep_data)
  pairs     <- combn(rep_names, 2, simplify = FALSE)

  for (thr_val in list(NULL, 0.1, 0.2)) {
    thr_label <- if (is.null(thr_val)) "all" else as.character(thr_val)

    for (p in pairs) {
      sa <- dat$rep_data[[p[1]]]
      sb <- dat$rep_data[[p[2]]]
      if (!is.null(thr_val)) {
        sa <- sa[sa >= thr_val]
        sb <- sb[sb >= thr_val]
      }
      j   <- jaccard(names(sa), names(sb))
      n_a <- length(sa)
      n_b <- length(sb)
      n_i <- length(intersect(names(sa), names(sb)))
      n_u <- length(union(names(sa), names(sb)))

      summary_rows[[length(summary_rows) + 1]] <- data.frame(
        cell_line = cl,
        rep_A     = p[1],
        rep_B     = p[2],
        threshold = thr_label,
        n_A       = n_a,
        n_B       = n_b,
        n_intersect = n_i,
        n_union     = n_u,
        jaccard     = round(j, 4),
        stringsAsFactors = FALSE
      )
    }
  }
}

summary_df <- do.call(rbind, summary_rows)

summary_df %>%
  mutate(
    rep_A = sub("^[^_]+_", "", rep_A),
    rep_B = sub("^[^_]+_", "", rep_B)
  ) %>%
  kable(
    col.names = c("Cell line", "Rep A", "Rep B", "Threshold",
                  "n(A)", "n(B)", "n(A∩B)", "n(A∪B)", "Jaccard"),
    align     = "llllrrrrc"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  collapse_rows(columns = c(1, 4), valign = "top")
```
