---
title: "QC — rRNA pU sites (chi-squared strategy)"
author: "XSun"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, warning=FALSE, message=FALSE}
library(data.table)
library(dplyr)
library(stringr)
library(ggplot2)
library(patchwork)
library(gridExtra)
library(scales)

knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center"
)
```

# Introduction

Quality control of rRNA pseudouridine (pU) site calls using the **chi-squared
strategy**. All analyses use a single cutoff: **FDR < 0.01, rate_diff > 0.05,
n_samples >= 2**.

```{r config}
RESULTS_DIRS <- c("/project2/xinhe/xsun/pumseq/pumseq_results",
                  "/project/xinhe/xsun/pumseq/pumseq_results")

FDR_CUT  <- 0.01
DIFF_CUT <- 0.05
CUTOFF_LABEL <- "FDR < 0.01 & rate_diff > 0.05"
MIN_NSAMP <- 2
```

```{r discover-cell-lines}
discover_cell_lines <- function() {
  all_cls <- list()
  for (rd in RESULTS_DIRS) {
    cl_dirs <- sort(list.dirs(rd, recursive = FALSE, full.names = TRUE))
    for (d in cl_dirs) {
      cl <- basename(d)
      if (!grepl("^CL\\d+$", cl)) next
      rRNA_files <- list.files(file.path(d, "out_chisq"),
                               pattern = "filter\\..*\\.rRNA\\.pseusite\\.txt\\.gz",
                               full.names = TRUE)
      if (length(rRNA_files) == 0) next
      reps <- sub("\\.rRNA\\.pseusite\\.txt\\.gz$", "",
                  sub("^filter\\.", "", basename(rRNA_files)))
      has_pooled <- file.exists(file.path(d, "pooled_sites_chisq", "pooled_rRNA.txt"))
      qc_png <- file.path(d, "qc", "qc.pUFraction.rRNA.png")
      has_qc_png <- file.exists(qc_png)
      all_cls[[cl]] <- list(
        results_dir = rd, reps = sort(reps),
        has_pooled = has_pooled, has_qc_png = has_qc_png, qc_png = qc_png
      )
    }
  }
  all_cls
}

ALL_CLS <- discover_cell_lines()
ALL_CLS <- ALL_CLS[sort(names(ALL_CLS))]
CL_NAMES <- names(ALL_CLS)
POOLED_CLS <- CL_NAMES[sapply(ALL_CLS, `[[`, "has_pooled")]

cat("Cell lines with per-replicate rRNA data:", length(CL_NAMES), "\n")
cat("Cell lines with pooled rRNA data:", length(POOLED_CLS), "\n")
cat("All:", paste(CL_NAMES, collapse = ", "), "\n")
cat("Pooled:", paste(POOLED_CLS, collapse = ", "), "\n")
```

```{r helpers}
read_rep_rRNA <- function(cl, sample_id) {
  rd <- ALL_CLS[[cl]]$results_dir
  f <- file.path(rd, cl, "out_chisq",
                 sprintf("filter.%s.rRNA.pseusite.txt.gz", sample_id))
  if (!file.exists(f)) { warning("Missing: ", f); return(NULL) }
  dt <- fread(f, header = TRUE, sep = "\t",
              colClasses = list(character = "ref"), data.table = FALSE)
  dt %>% mutate(site = paste(ref, pos, strand, sep = ":"))
}

sites_pass <- function(dt) {
  dt$site[dt$fdr < FDR_CUT & dt$rate_diff > DIFF_CUT]
}

jaccard <- function(a, b) {
  n_int <- length(intersect(a, b))
  n_uni <- length(union(a, b))
  if (n_uni == 0) NA_real_ else n_int / n_uni
}

make_jaccard_heatmap <- function(rep_tabs, rep_names, title_text) {
  site_sets <- lapply(rep_tabs, sites_pass)
  n <- length(rep_names)
  mat <- matrix(NA_real_, n, n, dimnames = list(rep_names, rep_names))
  n_mat <- matrix(NA_integer_, n, n, dimnames = list(rep_names, rep_names))
  diag(mat) <- 1
  for (i in 1:(n-1)) for (j in (i+1):n) {
    sa <- site_sets[[rep_names[i]]]; sb <- site_sets[[rep_names[j]]]
    mat[i,j] <- mat[j,i] <- jaccard(sa, sb)
    ni <- length(intersect(sa, sb))
    n_mat[i,j] <- n_mat[j,i] <- ni
  }
  short <- function(s) sub("^[^_]+_", "", s)
  lev <- short(rep_names)
  plot_df <- expand.grid(x = rep_names, y = rep_names, stringsAsFactors = FALSE) %>%
    rowwise() %>%
    mutate(j = mat[x, y], n_int = n_mat[x, y],
           label = if (x == y) "1.000"
                   else if (is.na(j)) "NA"
                   else sprintf("%.3f\n(n=%d)", j, n_int)) %>%
    ungroup() %>%
    mutate(x_lab = factor(short(x), levels = lev),
           y_lab = factor(short(y), levels = rev(lev)))
  ggplot(plot_df, aes(x = x_lab, y = y_lab, fill = j)) +
    geom_tile(color = "white", linewidth = 0.8) +
    geom_text(aes(label = label), size = 3, lineheight = 0.9) +
    scale_fill_gradient2(low = "#1a9641", mid = "#ffffbf", high = "#d7191c",
                         midpoint = 0.5, limits = c(0, 1), na.value = "grey90",
                         name = "Jaccard") +
    labs(title = title_text, subtitle = CUTOFF_LABEL, x = NULL, y = NULL) +
    coord_fixed() +
    theme_bw(base_size = 10) +
    theme(axis.text.x = element_text(angle = 35, hjust = 1),
          panel.grid = element_blank(), panel.border = element_blank())
}

make_pU_dist <- function(rep_tabs, rep_names, title_prefix) {
  lapply(rep_names, function(samp) {
    dt <- rep_tabs[[samp]]
    df <- dt[dt$fdr < FDR_CUT & dt$rate_diff > DIFF_CUT, , drop = FALSE]
    if (nrow(df) == 0) return(NULL)
    med_val <- median(df$ratio); mean_val <- mean(df$ratio)
    ggplot(df, aes(x = ratio)) +
      geom_histogram(aes(y = after_stat(density)), binwidth = 0.02,
                     fill = "steelblue", color = "white", alpha = 0.7) +
      geom_density(color = "navy", linewidth = 0.7) +
      geom_vline(xintercept = med_val, color = "red", linetype = "dashed") +
      geom_vline(xintercept = mean_val, color = "darkorange", linetype = "dotted") +
      annotate("text", x = Inf, y = Inf,
               label = sprintf("n = %d\nmedian = %.3f\nmean = %.3f",
                               nrow(df), med_val, mean_val),
               hjust = 1.05, vjust = 1.05, size = 3, family = "mono") +
      scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
      labs(title = samp, x = "%pU (C/(C+T))", y = "Density") +
      theme_bw(base_size = 10)
  })
}

make_pearson_heatmap <- function(rep_tabs, rep_names, title_text) {
  n <- length(rep_names)
  mat <- matrix(NA_real_, n, n, dimnames = list(rep_names, rep_names))
  n_mat <- matrix(NA_integer_, n, n, dimnames = list(rep_names, rep_names))
  diag(mat) <- 1
  for (i in 1:(n-1)) for (j in (i+1):n) {
    d1 <- rep_tabs[[rep_names[i]]]; d2 <- rep_tabs[[rep_names[j]]]
    s1 <- sites_pass(d1); s2 <- sites_pass(d2)
    shared <- intersect(s1, s2)
    if (length(shared) < 3) next
    m <- merge(d1[d1$site %in% shared, c("site","ratio")],
               d2[d2$site %in% shared, c("site","ratio")],
               by = "site", suffixes = c("_1","_2"))
    r <- cor(m$ratio_1, m$ratio_2, method = "pearson")
    mat[i,j] <- mat[j,i] <- r
    n_mat[i,j] <- n_mat[j,i] <- nrow(m)
  }
  short <- function(s) sub("^[^_]+_", "", s)
  lev <- short(rep_names)
  plot_df <- expand.grid(x = rep_names, y = rep_names, stringsAsFactors = FALSE) %>%
    rowwise() %>%
    mutate(r = mat[x, y], n_common = n_mat[x, y],
           label = if (x == y) "1.000"
                   else if (is.na(r)) "NA"
                   else sprintf("%.3f\n(n=%d)", r, n_common)) %>%
    ungroup() %>%
    mutate(x_lab = factor(short(x), levels = lev),
           y_lab = factor(short(y), levels = rev(lev)))
  ggplot(plot_df, aes(x = x_lab, y = y_lab, fill = r)) +
    geom_tile(color = "white", linewidth = 0.8) +
    geom_text(aes(label = label), size = 3, lineheight = 0.9) +
    scale_fill_gradient2(low = "#1a9641", mid = "#ffffbf", high = "#d7191c",
                         midpoint = 0.5, limits = c(0, 1), na.value = "grey90",
                         name = "Pearson r") +
    labs(title = title_text, subtitle = CUTOFF_LABEL, x = NULL, y = NULL) +
    coord_fixed() +
    theme_bw(base_size = 10) +
    theme(axis.text.x = element_text(angle = 35, hjust = 1),
          panel.grid = element_blank(), panel.border = element_blank())
}
```

```{r load-replicate-data, include=FALSE}
cl_rep_data <- list()
for (cl in CL_NAMES) {
  sids <- ALL_CLS[[cl]]$reps
  tabs <- setNames(lapply(sids, function(s) read_rep_rRNA(cl, s)), sids)
  tabs <- Filter(Negate(is.null), tabs)
  cl_rep_data[[cl]] <- tabs
}
```

# Examining replicates for each cell line

For each cell line we show four QC panels:

**1. rRNA pU fraction** — Per-sample lollipop plot of the conversion rate
C/(C+T) at every uridine position along the human 45S rRNA (18S, 5.8S, 28S;
positions with depth >= 20 only).

- **Green** dots/lines: annotated pseudouridine (pU) sites.
- **Grey** dots/lines: unannotated T positions (background).
- **Top-left** (`T=X%`): mean conversion at unannotated positions (background noise).
- **Top-right** (`pU=X%`): mean conversion at annotated pU sites (signal).
- Untreated samples should show near-zero signal; treated samples should show
  high green (pU) and low grey (T).

**2. Jaccard index** — Set similarity of called sites between each pair of
replicates at FDR < 0.01 & rate_diff > 0.05:
$J(A, B) = |A \cap B| / |A \cup B|$. The intersection count is shown in
parentheses. Higher values indicate more reproducible site calling.

**3. %pU distribution** — Histogram + density of the treated conversion ratio
at sites passing FDR < 0.01 & rate_diff > 0.05 in each replicate. Red dashed
line = median; orange dotted line = mean.

**4. Pairwise Pearson heatmap** — Pearson correlation of %pU values at sites
called significant in **both** replicates of each pair. High r indicates
quantitative agreement.

```{r per-cl-replicate-sections, results='asis', fig.width=10, fig.height=8}
for (cl in CL_NAMES) {
  tabs <- cl_rep_data[[cl]]
  nms  <- names(tabs)
  info <- ALL_CLS[[cl]]

  cat("\n\n---\n\n")
  cat(sprintf("## %s {.tabset}\n\n", cl))

  if (info$has_qc_png) {
    cat("### rRNA pU fraction\n\n")
    cat(sprintf("![%s rRNA QC](%s){width=100%%}\n\n", cl, info$qc_png))
  }

  if (length(tabs) < 2) {
    cat("Not enough replicates found.\n\n")
    next
  }

  cat("### Jaccard index\n\n")
  print(make_jaccard_heatmap(tabs, nms, cl))
  cat("\n\n")

  cat("### %pU distribution\n\n")
  plots <- make_pU_dist(tabs, nms, cl)
  plots <- Filter(Negate(is.null), plots)
  if (length(plots) > 0) {
    ncol_dist <- min(2, length(plots))
    do.call(grid.arrange, c(plots, list(ncol = ncol_dist)))
  }
  cat("\n\n")

  cat("### Pearson heatmap\n\n")
  print(make_pearson_heatmap(tabs, nms, cl))
  cat("\n\n")
}
```

---

# Cross-cell-line comparison (pooled data)

All comparisons below use the **pooled** rRNA data
(`pooled_sites_chisq/pooled_rRNA.txt`) filtered at FDR < 0.01, rate_diff >
0.05, and n_samples >= 2. In the pooled data, treated counts (C, T, depth)
are summed across all replicates of a cell line and the chi-squared test is
re-run on the combined contingency table, yielding higher statistical power.
`n_samples` records how many individual replicates detected the site.

**Plots shown:**

1. **Site counts** — Bar chart of the number of pooled sites passing the
   cutoff for each cell line. Consistent counts across cell lines suggest
   stable site calling.
2. **%pU distribution** — Histogram + density of the pooled conversion ratio
   C/(C+T) at passing sites. Red dashed = median; orange dotted = mean.
3. **Jaccard index heatmap** — Pairwise set similarity of called sites
   between cell lines. High Jaccard indicates the same rRNA positions are
   detected across cell lines (expected for constitutive modifications).
4. **Pearson r heatmap** — Pairwise correlation of %pU at sites called in
   **both** cell lines. High r indicates quantitative agreement in
   modification levels.

Cell lines with pooled data: **`r paste(POOLED_CLS, collapse = ", ")`**

```{r load-pooled}
load_pooled <- function(cl) {
  rd <- ALL_CLS[[cl]]$results_dir
  f <- file.path(rd, cl, "pooled_sites_chisq", "pooled_rRNA.txt")
  dt <- fread(f, header = TRUE, sep = "\t",
              colClasses = list(character = "ref"))
  dt[, site := paste(ref, pos, strand, sep = ":")]
  dt[, cell_line := cl]
  dt
}

pooled_list <- setNames(lapply(POOLED_CLS, load_pooled), POOLED_CLS)

filter_pooled <- function(dt) {
  dt[fdr < FDR_CUT & rate_diff > DIFF_CUT & n_samples >= MIN_NSAMP]
}
```

## Number of pooled sites per cell line

```{r pooled-site-counts, fig.width=10, fig.height=12}
count_df <- data.frame(
  cell_line = POOLED_CLS,
  n_sites = sapply(POOLED_CLS, function(cl) nrow(filter_pooled(pooled_list[[cl]])))
)
count_df$cell_line <- factor(count_df$cell_line, levels = POOLED_CLS)

n_per_panel <- ceiling(nrow(count_df) / 5)
count_df$panel <- rep(1:5, each = n_per_panel, length.out = nrow(count_df))

count_plots <- lapply(1:5, function(p) {
  sub <- count_df[count_df$panel == p, ]
  ggplot(sub, aes(x = cell_line, y = n_sites, fill = cell_line)) +
    geom_col(width = 0.6, show.legend = FALSE) +
    geom_text(aes(label = comma(n_sites)), vjust = -0.3, size = 3.5) +
    scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15)),
                       limits = c(0, max(count_df$n_sites) * 1.15)) +
    labs(x = NULL, y = "# sites") +
    theme_bw(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})

wrap_plots(count_plots, ncol = 1) +
  plot_annotation(
    title = "Number of pooled rRNA pU sites per cell line",
    subtitle = paste(CUTOFF_LABEL, "& n_samples >=", MIN_NSAMP))
```

## %pU Distribution — pooled data

```{r pooled-dist, fig.width=14, fig.height=20}
dist_plots <- lapply(POOLED_CLS, function(cl) {
  df <- as.data.frame(filter_pooled(pooled_list[[cl]])[, .(psi = ratio)])
  if (nrow(df) == 0) return(NULL)
  med_val <- median(df$psi); mean_val <- mean(df$psi)
  ggplot(df, aes(x = psi)) +
    geom_histogram(aes(y = after_stat(density)), binwidth = 0.02,
                   fill = "steelblue", color = "white", alpha = 0.7) +
    geom_density(color = "navy", linewidth = 0.7) +
    geom_vline(xintercept = med_val, color = "red", linetype = "dashed") +
    geom_vline(xintercept = mean_val, color = "darkorange", linetype = "dotted") +
    annotate("text", x = Inf, y = Inf,
             label = sprintf("n = %d\nmedian = %.3f\nmean = %.3f",
                             nrow(df), med_val, mean_val),
             hjust = 1.05, vjust = 1.05, size = 3, family = "mono") +
    scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
    labs(title = cl, x = "%pU (pooled)", y = "Density") +
    theme_bw(base_size = 10)
})
dist_plots <- Filter(Negate(is.null), dist_plots)
wrap_plots(dist_plots, ncol = 5)
```

## Cross-cell-line Jaccard index

```{r pooled-jaccard, fig.width=max(6, length(POOLED_CLS)*1.2), fig.height=max(5, length(POOLED_CLS)*1.0)}
site_sets_pooled <- setNames(
  lapply(POOLED_CLS, function(cl) filter_pooled(pooled_list[[cl]])$site),
  POOLED_CLS
)

n_cl <- length(POOLED_CLS)
j_mat <- matrix(NA_real_, n_cl, n_cl, dimnames = list(POOLED_CLS, POOLED_CLS))
n_int_mat <- matrix(NA_integer_, n_cl, n_cl, dimnames = list(POOLED_CLS, POOLED_CLS))
diag(j_mat) <- 1
for (i in 1:(n_cl-1)) for (j in (i+1):n_cl) {
  sa <- site_sets_pooled[[POOLED_CLS[i]]]; sb <- site_sets_pooled[[POOLED_CLS[j]]]
  j_mat[i,j] <- j_mat[j,i] <- jaccard(sa, sb)
  ni <- length(intersect(sa, sb))
  n_int_mat[i,j] <- n_int_mat[j,i] <- ni
}

jac_df <- expand.grid(x = POOLED_CLS, y = POOLED_CLS, stringsAsFactors = FALSE) %>%
  rowwise() %>%
  mutate(j = j_mat[x, y], n_int = n_int_mat[x, y],
         label = if (x == y) "1.000"
                 else if (is.na(j)) "NA"
                 else sprintf("%.3f\n(n=%d)", j, n_int)) %>%
  ungroup() %>%
  mutate(x = factor(x, levels = POOLED_CLS),
         y = factor(y, levels = rev(POOLED_CLS)))

ggplot(jac_df, aes(x = x, y = y, fill = j)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = label), size = 3, lineheight = 0.9) +
  scale_fill_gradient2(low = "#1a9641", mid = "#ffffbf", high = "#d7191c",
                       midpoint = 0.5, limits = c(0, 1), na.value = "grey90",
                       name = "Jaccard") +
  labs(title = "Jaccard index — pooled rRNA sites across cell lines",
       subtitle = paste(CUTOFF_LABEL, "& n_samples >=", MIN_NSAMP),
       x = NULL, y = NULL) +
  coord_fixed() +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(), panel.border = element_blank())
```

## Cross-cell-line Pearson's r of %pU

```{r pooled-pearson, fig.width=max(6, length(POOLED_CLS)*1.2), fig.height=max(5, length(POOLED_CLS)*1.0)}
r_mat <- matrix(NA_real_, n_cl, n_cl, dimnames = list(POOLED_CLS, POOLED_CLS))
n_com_mat <- matrix(NA_integer_, n_cl, n_cl, dimnames = list(POOLED_CLS, POOLED_CLS))
diag(r_mat) <- 1
for (i in 1:(n_cl-1)) for (j in (i+1):n_cl) {
  d1 <- filter_pooled(pooled_list[[POOLED_CLS[i]]])[, .(site, ratio1 = ratio)]
  d2 <- filter_pooled(pooled_list[[POOLED_CLS[j]]])[, .(site, ratio2 = ratio)]
  m <- merge(d1, d2, by = "site")
  if (nrow(m) < 3) next
  r_mat[i,j] <- r_mat[j,i] <- cor(m$ratio1, m$ratio2, method = "pearson")
  n_com_mat[i,j] <- n_com_mat[j,i] <- nrow(m)
}

cor_df <- expand.grid(x = POOLED_CLS, y = POOLED_CLS, stringsAsFactors = FALSE) %>%
  rowwise() %>%
  mutate(r = r_mat[x, y], n_com = n_com_mat[x, y],
         label = if (x == y) "1.000"
                 else if (is.na(r)) "NA"
                 else sprintf("%.3f\n(n=%d)", r, n_com)) %>%
  ungroup() %>%
  mutate(x = factor(x, levels = POOLED_CLS),
         y = factor(y, levels = rev(POOLED_CLS)))

ggplot(cor_df, aes(x = x, y = y, fill = r)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = label), size = 3, lineheight = 0.9) +
  scale_fill_gradient2(low = "#1a9641", mid = "#ffffbf", high = "#d7191c",
                       midpoint = 0.5, limits = c(0, 1), na.value = "grey90",
                       name = "Pearson r") +
  labs(title = "Pearson's r of %pU — pooled rRNA sites across cell lines",
       subtitle = paste(CUTOFF_LABEL, "& n_samples >=", MIN_NSAMP),
       x = NULL, y = NULL) +
  coord_fixed() +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(), panel.border = element_blank())
```

---

# Comparing with published data (BACS)

We benchmark our pooled rRNA pU calls against the **BACS** dataset (PMID
39349603), which detects pseudouridine via antibody-based capture and reports
rRNA pU sites in HeLa polyA+ RNA (Supplementary Table, sheet 2). Sites are
matched by rRNA accession (`NR_*`) and position.

The grouped bar chart shows, for each cell line: the number of **our** pooled
sites passing the cutoff, the total number of **BACS** reference sites, and
the **overlap** count. Higher overlap validates the specificity of our calls.

```{r pub-bacs-overlap, fig.width=10, fig.height=15}
path_BACS <- "/project2/xinhe/xsun/pumseq/data_others/pU_PMID39349603.xlsx"

rrna_BACS <- readxl::read_excel(path_BACS, sheet = 2)
colnames(rrna_BACS) <- rrna_BACS[2, ]
rrna_BACS <- rrna_BACS[-c(1:2), ]
rrna_BACS <- rrna_BACS[complete.cases(rrna_BACS$chr), ]

rrna_BACS_clean <- rrna_BACS %>%
  mutate(accession = str_extract(chr, "NR_[0-9]+"),
         pos_num   = as.integer(pos))

n_bacs <- nrow(rrna_BACS_clean)

overlap_df <- do.call(rbind, lapply(POOLED_CLS, function(cl) {
  dt <- filter_pooled(pooled_list[[cl]])
  ours <- as.data.frame(dt) %>%
    mutate(accession = str_extract(ref, "NR_[0-9]+"))
  n_ours <- nrow(ours)
  n_overlap <- nrow(inner_join(ours, rrna_BACS_clean,
                               by = c("accession", "pos" = "pos_num")))
  data.frame(
    cell_line = cl,
    category  = rep(c("Our sites", "BACS sites", "Overlap"), 1),
    count     = c(n_ours, n_bacs, n_overlap)
  )
}))

overlap_df$category <- factor(overlap_df$category,
                              levels = c("Our sites", "BACS sites", "Overlap"))
overlap_df$cell_line <- factor(overlap_df$cell_line, levels = POOLED_CLS)

n_per_panel <- ceiling(length(POOLED_CLS) / 5)
cl_panels <- split(POOLED_CLS, rep(1:5, each = n_per_panel, length.out = length(POOLED_CLS)))

bacs_plots <- lapply(cl_panels, function(cls) {
  sub <- overlap_df[overlap_df$cell_line %in% cls, ]
  sub$cell_line <- factor(sub$cell_line, levels = cls)
  ggplot(sub, aes(x = cell_line, y = count, fill = category)) +
    geom_col(position = position_dodge(width = 0.7), width = 0.6) +
    geom_text(aes(label = count), position = position_dodge(width = 0.7),
              vjust = -0.3, size = 3) +
    scale_fill_manual(values = c("Our sites" = "#4E79A7",
                                  "BACS sites" = "#F28E2B",
                                  "Overlap" = "#59A14F")) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    labs(x = NULL, y = "# sites", fill = NULL) +
    theme_bw(base_size = 11) +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1))
})
bacs_plots[[1]] <- bacs_plots[[1]] + theme(legend.position = "top")

wrap_plots(bacs_plots, ncol = 1) +
  plot_annotation(
    title = "Overlap of pooled rRNA pU sites with BACS",
    subtitle = paste(CUTOFF_LABEL, "& n_samples >=", MIN_NSAMP))
```
