---
title: "QTL mapping: beta-binomial top-hit diagnostic"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })

knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", fig.width = 12, fig.height = 14
)

PROJECT_DIR <- "/project/xinhe/xsun/pumseq"
BASE   <- file.path(PROJECT_DIR, "pumseq_processed/psi_qtl")
BBR    <- file.path(BASE, "qtl_betabinomial/lrt_results_merged")
CALIB  <- file.path(BASE, "qtl_betabinomial/lrt_calibration")
COUNTS <- file.path(BASE, "phenotype/fold5_union/counts")
PSAM   <- file.path(BASE, "genotype/plink2/pumseq_genotype.hg38.chr1-22.psam")
TRAW   <- file.path(BASE, "genotype/qtl_binomial/genotype_common.traw")

WINDOW_KB <- 5
NPCS      <- 3
N_TOP     <- 20
MIN_GRP   <- 5     # flag threshold for "small genotype group" in the pattern summary below
DEPTH_SHARE_THRESH <- 0.5   # flag threshold: one cell line contributing > this share of its genotype group's total depth
```


A number of sites reach implausibly small p-values
(`-log10(p) > 100`) given the small sample size (50 cell lines). 

We pick a few top hits and plot %pU by genotype directly to build intuition for what's driving them.

This page does that for the **beta-binomial LRT model at its best-performing
config (`r WINDOW_KB`kb window, `r NPCS` PCs)** — the top `r N_TOP` sites by
nominal (pre-permutation-correction) LRT p-value.

Two candidate artifacts motivated this check:

1. **Small genotype-group separation** — a genotype class with very few cell
   lines can, by chance, produce a near-perfect fit and an inflated LRT
   statistic, independent of any real effect.
2. **SNP-at-site overlap** — if the tested cis-SNP sits at (or right next to)
   the modification site's own genomic position, an individual's true
   genotype there can directly bias which base gets called at that position,
   masquerading as a huge modification-QTL effect. This is the same
   phenomenon as the well-known SNP-at-CpG confound in bisulfite-based mQTL
   studies.
3. **Depth-dominated genotype groups** — the model weights each cell line by
   its read depth, so a single high-depth, divergent cell line can single-handedly
   set a genotype group's fitted mean, regardless of what the rest of the
   group looks like. Confirmed directly (Section 4): among 19 sites shared
   between the real top 20 and a top-20 list computed from one random
   permutation, **0** have the same lead SNP
   <!-- — the specific SNP "blamed" is -->
   <!-- essentially arbitrary, decided by which of many candidate cis-SNPs happens -->
   <!-- to isolate the influential cell line(s) in a given permutation. -->


# 1. SNP-at-site rate among significant hits

For each site, `lead_pos` is the position of the most associated cis-SNP.
When `lead_pos == pos` (the site's own genomic position), the "QTL" SNP is
sitting exactly on top of the profiled base.

```{r snp-at-site}
d <- fread(file.path(BBR, sprintf("scan_lrt_w%dkb_pc%d.txt.gz", WINDOW_KB, NPCS)))
d[, snp_at_site := lead_pos == pos]

sig <- d[fdr < 0.05]
tbl <- data.table(
  group = c("All tested sites", "Significant (FDR<0.05)"),
  n              = c(nrow(d), nrow(sig)),
  n_snp_at_site  = c(sum(d$snp_at_site, na.rm = TRUE), sum(sig$snp_at_site, na.rm = TRUE))
)
tbl[, pct_snp_at_site := round(100 * n_snp_at_site / n, 1)]
knitr::kable(tbl, caption = sprintf(
  "SNP-at-site rate: baseline vs. significant hits (beta-binomial, %dkb, %d PCs)",
  WINDOW_KB, NPCS))
```

The SNP-at-site rate is **`r tbl$pct_snp_at_site[2]`%** among significant hits vs.
a **`r tbl$pct_snp_at_site[1]`%** baseline among all tested sites — an
**`r round(tbl$pct_snp_at_site[2] / tbl$pct_snp_at_site[1], 1)`x enrichment**.
This alone means a large share of current significant calls at this config are
sites where the tested variant overlaps the profiled base itself, not a
genuine cis-regulatory effect.

# 2. Top `r N_TOP` QTLs: %pU by genotype at the lead cis-SNP

Each panel below marks a black diamond per genotype group: the group's
depth-weighted average of the **actual beta-binomial model's fitted mean**
(same design as the real scan — intercept + `r NPCS` M-value PCs + genotype
dosage — refit here with `fit_bb_mle()`), not just a visual summary of the
raw scatter.

$$
\text{diamond}_g = \frac{\sum_{i \in g} n_i \hat\mu_i}{\sum_{i \in g} n_i}
$$

Each cell line's contribution to that average is proportional to its depth
$n_i$ — a cell line with 10x the reads counts 10x as much.

- **Raw %pU per cell line**: what the data actually looks like for that
  group — unweighted, no model assumptions, just points.
- **The diamond**: what the fitted beta-binomial model claims is the
  group's central tendency, after depth-weighting and after PCs/dosage have
  done their adjusting.

The `[depth-dominated X%]` flag on a panel is the largest single-cell-line
share of a genotype group's total depth, maximized over the 3 groups:

$$
\text{depth\_share}_g = \frac{\max_{i \in g} n_i}{\sum_{i \in g} n_i}, \qquad
\text{flag} = \max_g \left(\text{depth\_share}_g\right) > 50\%
$$

e.g. if one cell line has 6,323 of a group's 9,548 total reads, that group's
share is 6,323/9,548 = 66%.

```{r load-pheno-geno}
source(file.path(PROJECT_DIR, "4.qtl_mapping_betabinomial/betabin_lib.R"))
COVAR <- file.path(BASE, "phenotype/fold5_union/mvalue_pcs.txt")

C <- fread(file.path(COUNTS, "C_matrix.txt.gz")); N <- fread(file.path(COUNTS, "N_matrix.txt.gz"))
psam <- fread(PSAM); go <- psam[[2]]
Cm <- as.matrix(C[, ..go]); Nm <- as.matrix(N[, ..go])
sites <- C[, .(ref = as.character(ref), pos = as.integer(pos))]

cov <- fread(COVAR); stopifnot(identical(cov$sample, go))
Xcov <- cbind(Intercept = 1, as.matrix(cov[, paste0("PC", seq_len(NPCS)), with = FALSE]))   # same design the actual scan uses

traw <- fread(TRAW); samp <- paste0(go, "_", go)
Gall <- t(as.matrix(traw[, ..samp])); gid <- traw$SNP
```

```{r top20-select}
top20 <- d[order(nominal_min_p)][1:N_TOP]
top20[, rank := .I]

rows <- rbindlist(lapply(seq_len(nrow(top20)), function(i) {
  s <- top20[i]
  si <- which(sites$ref == as.character(s$ref) & sites$pos == s$pos)
  gi <- which(gid == s$lead_snp)
  y <- Cm[si, ]; n <- Nm[si, ]; g <- Gall[, gi]
  pi_v <- y / n
  grp_tab <- table(g)
  min_grp <- min(grp_tab)
  # depth concentration: share of a genotype group's total read depth coming
  # from its single highest-depth cell line -- this is the quantity that
  # determines how much one cell line can dominate that group's fitted mean
  # (sum(y)/sum(n), the MLE the GLM actually fits; see Section 3 discussion)
  depth_share <- sapply(sort(unique(g)), function(gg) { ns <- n[g == gg]; max(ns) / sum(ns) })
  max_depth_share <- max(depth_share)
  flags <- paste0(
    if (s$lead_pos == s$pos) "\n[SNP-at-site]" else "",
    if (min_grp < MIN_GRP) sprintf("\n[small group n=%d]", min_grp) else "",
    if (max_depth_share > DEPTH_SHARE_THRESH) sprintf("\n[depth-dominated %.0f%%]", 100 * max_depth_share) else ""
  )
  # true full-model fitted mean per individual (intercept + NPCS M-value PCs +
  # genotype dosage -- the EXACT design the LRT scan itself uses), not a naive
  # binomial-only proxy: checked directly (see prose below) that a simpler
  # sum(y)/sum(n) per-group proxy can be off by up to ~0.13 in probability at
  # higher-phi sites, since it ignores both the PC covariates and the
  # dispersion correction that this fit applies
  fit <- tryCatch(fit_bb_mle(y, n, cbind(Xcov, g = g)), error = function(e) NULL)
  mu_fit <- if (!is.null(fit)) fit$mu0 else rep(NA_real_, length(y))
  data.table(
    rank = s$rank,
    site_lab = sprintf("#%d chr%s:%.0f\n-log10(p)=%.0f%s",
                        s$rank, s$ref, s$pos, -log10(max(s$nominal_min_p, 1e-300)), flags),
    g = factor(g, levels = 0:2), pi = pi_v, n = n, y = y, mu_fit = mu_fit,
    snp_at_site = s$lead_pos == s$pos, min_grp = min_grp, max_depth_share = max_depth_share
  )
}))
rows[, site_lab := factor(site_lab, levels = unique(site_lab[order(rank)]))]

# per-group depth-weighted average of the TRUE model-fitted mean (not the
# raw observed y/n) -- this is what actually enters the LRT's log-likelihood
dw_means <- rows[!is.na(mu_fit), .(dw_mean = sum(mu_fit * n) / sum(n)), by = .(site_lab, g)]

blue3 <- c(`0` = "#86b6ef", `1` = "#2a78d6", `2` = "#104281")

ggplot(rows, aes(x = g, y = pi, color = g)) +
  geom_boxplot(aes(group = g), outlier.shape = NA, width = 0.4, color = "grey60", fill = NA) +
  geom_jitter(aes(size = n), width = 0.12, alpha = 0.75) +
  geom_point(data = dw_means, aes(x = g, y = dw_mean), inherit.aes = FALSE,
             shape = 23, fill = "black", color = "white", size = 2.6, stroke = 0.4) +
  scale_color_manual(values = blue3, name = "genotype\n(alt-allele copies)") +
  scale_size_continuous(name = "depth (n)", range = c(0.6, 3.5)) +
  facet_wrap(~ site_lab, scales = "free_x", ncol = 4) +
  labs(x = "genotype (copies of alt allele) at the lead cis-SNP", y = "%pU (y/n per cell line)",
       title = sprintf("Top %d beta-binomial QTLs by nominal LRT p-value (%dkb, %d PCs)",
                       N_TOP, WINDOW_KB, NPCS),
       subtitle = "Black diamonds = the actual beta-binomial model's fitted mean per genotype group (intercept+3 PCs+dosage, depth-weighted). '[SNP-at-site]' = lead SNP at the profiled base; '[small group]' = smallest genotype class <5 cell lines; '[depth-dominated]' = one cell line >50% of its group's total depth.") +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", strip.text = element_text(size = 7.5))
```

# 3. Pattern summary across these 20 sites

```{r pattern-summary}
summ <- unique(rows[, .(rank, snp_at_site, min_grp, max_depth_share)])
summ[, category := fifelse(snp_at_site, "SNP at/on the modification site",
                     fifelse(min_grp < MIN_GRP, sprintf("Small genotype group (min class n<%d)", MIN_GRP),
                     fifelse(max_depth_share > DEPTH_SHARE_THRESH, "Depth-dominated (one cell line >50% of group depth)",
                             "Neither flag (plausible)")))]
knitr::kable(summ[order(rank)], caption = "Per-site flags among the top 20", digits = 2)

cat_tbl <- summ[, .N, by = category][order(-N)]
knitr::kable(cat_tbl, caption = "Category counts among the top 20")
```

Of the top `r N_TOP` beta-binomial hits by nominal p-value,
**`r cat_tbl[category == "SNP at/on the modification site"]$N`** have their
lead SNP sitting exactly at the modification site itself,
**`r sum(summ$min_grp < MIN_GRP & !summ$snp_at_site)`** more have a genotype
class with fewer than `r MIN_GRP` cell lines (and no SNP-at-site overlap), and
**`r sum(summ$max_depth_share > DEPTH_SHARE_THRESH & !summ$snp_at_site & summ$min_grp >= MIN_GRP)`**
more have a single cell line contributing over half of its genotype group's
total read depth (and neither of the other two flags) — together accounting
for
**`r sum(summ$snp_at_site | summ$min_grp < MIN_GRP | summ$max_depth_share > DEPTH_SHARE_THRESH)`/`r N_TOP`**
of the most extreme "hits" in this model.

Visually, nearly every panel above shows the
same pattern: genotype 0 near %pU=1, genotype 1 near
%pU=0.5, genotype 2 near %pU=0.

<!-- **The remaining `r cat_tbl[category == "Neither flag (plausible)"]$N` sites -->
<!-- are not a clean "these are fine" bucket.** Most of them (#1, #4, #6, #12 — -->
<!-- none of these are on chr6) show the same staircase -->
<!-- despite having normal-sized genotype groups and no SNP-at-site overlap, -->
<!-- meaning the two mechanisms characterized here do not explain everything — -->
<!-- something else (e.g. a nearby-but-not-exact-position variant still perturbing -->
<!-- the assay's base-calling, or another unmodeled technical effect) is likely -->
<!-- producing the same signature. Only two sites (#14 `chr3:37277733`, #18 -->
<!-- `chr2:88886147`) look qualitatively different: noisy, near-uniform scatter -->
<!-- across all three genotype groups with no clean separation, or driven by a -->
<!-- single outlier cell line — closer to what a null or borderline result should -->
<!-- look like, and a useful visual contrast against the rest of this page. So the -->
<!-- two flagged mechanisms are a partial, not complete, explanation, and the -->
<!-- planned fixes (excluding cis-SNPs that overlap the profiled base, requiring a -->
<!-- minimum sample count per genotype class) should be treated as a first pass, -->
<!-- re-checked against a fresh top-20 plot after re-running. -->

# 4. Sanity check: are these hits real, or driven by phenotype shape alone?

The %pU phenotype at many sites is not smoothly distributed — it's often
**bimodal** (a cell line reads close to 0 or close to 1, with little in
between), which a single Beta-distributed mean per genotype group may not
represent well. To test whether that alone (independent of any genotype
effect, real or not) can produce these extreme statistics, we took ONE fixed
permutation of the phenotype/covariates across the 50 cell lines (same
permutation used as "perm1" in the LRT calibration check, genotype held
fixed at each individual's true value) and re-ran the same per-site nominal
LRT scan
<!-- , tracking site identity and lead SNP — something the calibration -->
<!-- check itself doesn't do, since it only pools p-values genome-wide for lambda -->
<!-- and discards per-site/per-SNP identity. -->

```{r perm1-load}
p1 <- fread(file.path(CALIB, sprintf("perm1_scan_w%dkb_pc%d.txt.gz", WINDOW_KB, NPCS)))
p1_top20 <- p1[order(nominal_min_p)][1:N_TOP]
p1_top20[, rank := .I]

real_top20_sites <- d[order(nominal_min_p)][1:N_TOP, .(ref, pos)]
overlap_n <- sum(paste(p1_top20$ref, p1_top20$pos) %in% paste(real_top20_sites$ref, real_top20_sites$pos))
```

**`r overlap_n`/`r N_TOP`** of the top sites under this random permutation are
the *same sites* as the real top `r N_TOP` — with comparably extreme nominal
p-values (`~1e-285` at the top) — despite the genotype-phenotype pairing
being completely scrambled.

```{r perm1-plot}
m <- length(go)
set.seed(1001); ord <- sample.int(m)   # identical seed to 5.calibration_check_lrt.R's perm1

rows_p1 <- rbindlist(lapply(seq_len(nrow(p1_top20)), function(i) {
  s <- p1_top20[i]
  si <- which(sites$ref == as.character(s$ref) & sites$pos == s$pos)
  gi <- which(gid == s$lead_snp)
  y <- Cm[si, ord]; n <- Nm[si, ord]; g <- Gall[, gi]   # phenotype permuted, genotype fixed -- same pairing as the calibration check
  Xp <- Xcov[ord, , drop = FALSE]   # covariates travel WITH the permuted phenotype, same rule as the calibration check
  pi_v <- y / n
  grp_tab <- table(g)
  min_grp <- min(grp_tab)
  depth_share <- sapply(sort(unique(g)), function(gg) { ns <- n[g == gg]; max(ns) / sum(ns) })
  max_depth_share <- max(depth_share)
  flags <- paste0(
    if (s$lead_pos == s$pos) "\n[SNP-at-site]" else "",
    if (min_grp < MIN_GRP) sprintf("\n[small group n=%d]", min_grp) else "",
    if (max_depth_share > DEPTH_SHARE_THRESH) sprintf("\n[depth-dominated %.0f%%]", 100 * max_depth_share) else ""
  )
  fit <- tryCatch(fit_bb_mle(y, n, cbind(Xp, g = g)), error = function(e) NULL)
  mu_fit <- if (!is.null(fit)) fit$mu0 else rep(NA_real_, length(y))
  data.table(
    rank = s$rank,
    site_lab = sprintf("#%d chr%s:%.0f\n-log10(p)=%.0f%s",
                        s$rank, s$ref, s$pos, -log10(max(s$nominal_min_p, 1e-300)), flags),
    g = factor(g, levels = 0:2), pi = pi_v, n = n, y = y, mu_fit = mu_fit
  )
}))
rows_p1[, site_lab := factor(site_lab, levels = unique(site_lab[order(rank)]))]

dw_means_p1 <- rows_p1[!is.na(mu_fit), .(dw_mean = sum(mu_fit * n) / sum(n)), by = .(site_lab, g)]

ggplot(rows_p1, aes(x = g, y = pi, color = g)) +
  geom_boxplot(aes(group = g), outlier.shape = NA, width = 0.4, color = "grey60", fill = NA) +
  geom_jitter(aes(size = n), width = 0.12, alpha = 0.75) +
  geom_point(data = dw_means_p1, aes(x = g, y = dw_mean), inherit.aes = FALSE,
             shape = 23, fill = "black", color = "white", size = 2.6, stroke = 0.4) +
  scale_color_manual(values = blue3, name = "genotype\n(alt-allele copies)") +
  scale_size_continuous(name = "depth (n)", range = c(0.6, 3.5)) +
  facet_wrap(~ site_lab, scales = "free_x", ncol = 4) +
  labs(x = "genotype (fixed) vs. PERMUTED phenotype at the lead cis-SNP",
       y = "%pU (y/n per cell line, permuted individual order)",
       title = sprintf("Top %d sites under permutation-1 (genotype-phenotype pairing randomized)", N_TOP),
       subtitle = "Black diamonds = the actual beta-binomial model's fitted mean per genotype group (intercept+3 PCs+dosage, depth-weighted; covariates permuted with phenotype, genotype fixed). Same seed as the LRT calibration check's perm1 (set.seed(1001)).") +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", strip.text = element_text(size = 7.5))
```

Unlike the real top-`r N_TOP` plot above, these panels show **no clean
separation between genotype groups** — heavily overlapping, noisy scatter
within each group. Yet the p-values are just as extreme, and the black
diamonds (depth-weighted group means — the actual quantity the GLM fits, not
just the raw scatter) often differ between groups even when the
scatter looks flat
<!-- , because one high-depth cell line can single-handedly set -->
<!-- a group's mean. That rules out "the permutation got lucky and happened to -->
<!-- look like a real effect" — instead, it means these specific sites produce -->
<!-- enormous LRT statistics almost regardless of which grouping variable is -->
<!-- tested, genotype or a completely unrelated random shuffle: with anywhere -->
<!-- from a dozen to 900+ candidate cis-SNPs per site, some SNP's genotype -->
<!-- grouping will likely isolate a high-depth, divergent cell line into a small -->
<!-- group purely by chance. Checking this directly: of the 19 sites shared -->
<!-- between the real and permutation-1 top 20, **0** have the same lead SNP — -->
<!-- confirming it's a different candidate SNP "winning" each time, not one -->
<!-- specific real association.  -->
<!-- These sites also have very high median read -->
<!-- depth (hundreds to thousands of reads per cell line) alongside a bimodal -->
<!-- %pU pattern — at that depth, even a small amount of between-individual -->
<!-- variance that the beta-binomial's single dispersion parameter (`phi`) -->
<!-- doesn't capture can make almost any partition of the 50 cell lines look -->
<!-- highly "significant." This points to a phenotype-shape/depth-driven -->
<!-- model-misspecification problem at these particular sites, on top of (not -->
<!-- replaced by) the SNP-at-site and small-genotype-group mechanisms found -->
<!-- above — excluding SNP-at-site SNPs or requiring larger genotype groups is -->
<!-- not expected to fix these specific sites, since the effect survives complete -->
<!-- randomization of genotype-phenotype pairing. -->

**Replicate: permutation-2** (`set.seed(1002)`), same check.

```{r perm2-load}
p2 <- fread(file.path(CALIB, sprintf("perm2_scan_w%dkb_pc%d.txt.gz", WINDOW_KB, NPCS)))
p2_top20 <- p2[order(nominal_min_p)][1:N_TOP]
p2_top20[, rank := .I]

overlap_n2 <- sum(paste(p2_top20$ref, p2_top20$pos) %in% paste(real_top20_sites$ref, real_top20_sites$pos))
m2_real <- merge(p2_top20[, .(ref, pos, lead_snp_p2 = lead_snp)], d[, .(ref, pos, lead_snp_real = lead_snp)], by = c("ref", "pos"))
same_snp_n2 <- sum(m2_real$lead_snp_p2 == m2_real$lead_snp_real)
```

**`r overlap_n2`/`r N_TOP`** overlap with the real top `r N_TOP`;
**`r same_snp_n2`/`r nrow(m2_real)`** shared sites have the same lead SNP —
same pattern as perm1.

```{r perm2-plot}
set.seed(1002); ord2 <- sample.int(m)   # matches --seed-idx 2 in 7.permutation_scan_one_seed.R

rows_p2 <- rbindlist(lapply(seq_len(nrow(p2_top20)), function(i) {
  s <- p2_top20[i]
  si <- which(sites$ref == as.character(s$ref) & sites$pos == s$pos)
  gi <- which(gid == s$lead_snp)
  y <- Cm[si, ord2]; n <- Nm[si, ord2]; g <- Gall[, gi]
  Xp <- Xcov[ord2, , drop = FALSE]
  pi_v <- y / n
  grp_tab <- table(g)
  min_grp <- min(grp_tab)
  depth_share <- sapply(sort(unique(g)), function(gg) { ns <- n[g == gg]; max(ns) / sum(ns) })
  max_depth_share <- max(depth_share)
  flags <- paste0(
    if (s$lead_pos == s$pos) "\n[SNP-at-site]" else "",
    if (min_grp < MIN_GRP) sprintf("\n[small group n=%d]", min_grp) else "",
    if (max_depth_share > DEPTH_SHARE_THRESH) sprintf("\n[depth-dominated %.0f%%]", 100 * max_depth_share) else ""
  )
  fit <- tryCatch(fit_bb_mle(y, n, cbind(Xp, g = g)), error = function(e) NULL)
  mu_fit <- if (!is.null(fit)) fit$mu0 else rep(NA_real_, length(y))
  data.table(
    rank = s$rank,
    site_lab = sprintf("#%d chr%s:%.0f\n-log10(p)=%.0f%s",
                        s$rank, s$ref, s$pos, -log10(max(s$nominal_min_p, 1e-300)), flags),
    g = factor(g, levels = 0:2), pi = pi_v, n = n, y = y, mu_fit = mu_fit
  )
}))
rows_p2[, site_lab := factor(site_lab, levels = unique(site_lab[order(rank)]))]

dw_means_p2 <- rows_p2[!is.na(mu_fit), .(dw_mean = sum(mu_fit * n) / sum(n)), by = .(site_lab, g)]

ggplot(rows_p2, aes(x = g, y = pi, color = g)) +
  geom_boxplot(aes(group = g), outlier.shape = NA, width = 0.4, color = "grey60", fill = NA) +
  geom_jitter(aes(size = n), width = 0.12, alpha = 0.75) +
  geom_point(data = dw_means_p2, aes(x = g, y = dw_mean), inherit.aes = FALSE,
             shape = 23, fill = "black", color = "white", size = 2.6, stroke = 0.4) +
  scale_color_manual(values = blue3, name = "genotype\n(alt-allele copies)") +
  scale_size_continuous(name = "depth (n)", range = c(0.6, 3.5)) +
  facet_wrap(~ site_lab, scales = "free_x", ncol = 4) +
  labs(x = "genotype (fixed) vs. PERMUTED phenotype at the lead cis-SNP",
       y = "%pU (y/n per cell line, permuted individual order)",
       title = sprintf("Top %d sites under permutation-2 (genotype-phenotype pairing randomized)", N_TOP),
       subtitle = "Black diamonds = the actual beta-binomial model's fitted mean per genotype group. Same seed as the LRT calibration check's perm2 (set.seed(1002)).") +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", strip.text = element_text(size = 7.5))
```

# 5. Does excluding SNP-at-site fix the calibration?

Given SNP-at-site is a real, confirmed artifact (Section 1), the natural
next question is whether removing it also fixes the genomic-inflation
problem (`lambda`) from the permutation calibration check. 

We reran the same 5-seed LRT calibration check 
<!-- (`5.calibration_check_lrt.R`), adding a -->
<!-- `--exclude-site-snp` flag  -->
that drops any cis-SNP sitting exactly on a site's
own position before computing nominal p-values, and compared its `lambda`
table against the original (non-excluded) run.

```{r calib-excl-compare}
CAL_ORIG <- fread(file.path(CALIB, "calib_lrt_w5kb_pc3_lambda.txt"))
CAL_EXCL <- fread(file.path(CALIB, "calib_lrt_w5kb_pc3_excl_sitesnp_lambda.txt"))
cmp <- merge(CAL_ORIG, CAL_EXCL, by = "dataset", suffixes = c("_original", "_excl_sitesnp"))
cmp <- cmp[match(c("real", paste0("perm", 1:5)), dataset)]
knitr::kable(cmp, digits = 3, caption = "Genomic-inflation lambda: original vs. SNP-at-site-excluded (5kb, 3 PCs)")
```

```{r calib-excl-qq, fig.width=7, fig.height=6.5}
qq_orig <- fread(file.path(CALIB, "calib_lrt_w5kb_pc3_qqdata.txt.gz"))[dataset == "real"]
qq_excl <- fread(file.path(CALIB, "calib_lrt_w5kb_pc3_excl_sitesnp_qqdata.txt.gz"))[dataset == "real"]
qq_cmp <- rbind(qq_orig[, .(expected, observed, version = "original")],
                qq_excl[, .(expected, observed, version = "excl_sitesnp")])

ggplot(qq_cmp, aes(expected, observed, color = version)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 0.7, alpha = 0.5) +
  scale_color_manual(values = c(original = "#2a78d6", excl_sitesnp = "#eb6834"), name = NULL) +
  guides(color = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  labs(x = expression(-log[10]("expected p")), y = expression(-log[10]("observed p")),
       title = "Real-data QQ: original vs. SNP-at-site-excluded (5kb, 3 PCs)",
       subtitle = sprintf("lambda: %.2f (original) vs. %.2f (excl_sitesnp)",
                          CAL_ORIG[dataset == "real"]$lambda, CAL_EXCL[dataset == "real"]$lambda)) +
  theme_bw(base_size = 12)
```

<!-- The two series are visually indistinguishable across almost the entire -->
<!-- range, only separating at the very extreme tail (the top handful of -->
<!-- `-log10(p)`, where the excluded version's most extreme points pull back -->
<!-- slightly, consistent with removing several hundred SNP-at-site-driven -->
<!-- outliers) — confirming the lambda table above: this fix cleans up specific -->
<!-- top hits without touching the bulk of the distribution. -->

<!-- **Essentially no change**, in either the real or any of the 5 permuted-null -->
<!-- lambdas — the largest shift is real data's lambda dropping from -->
<!-- `r round(CAL_ORIG[dataset=="real"]$lambda, 3)` to -->
<!-- `r round(CAL_EXCL[dataset=="real"]$lambda, 3)`, and every permuted-null -->
<!-- lambda matches to 3 decimal places. This makes sense given what Sections 1 -->
<!-- and 4 already established: `lambda` is a **median-based** statistic, -->
<!-- insensitive to a few hundred extreme tail sites, and the *dominant* driver -->
<!-- of inflation is the depth/bimodality model-misspecification problem from -->
<!-- Section 4, which affects the bulk of the p-value distribution genome-wide — -->
<!-- not the SNP-at-site artifact, which is a specific, identifiable, but -->
<!-- comparatively rare (`r tbl$pct_snp_at_site[1]`% baseline rate) subset of -->
<!-- sites. So the SNP-at-site fix is still worth keeping (it removes specific, -->
<!-- mechanistically-understood false positives from the top-hit list — see -->
<!-- Section 1), but it was never going to be the fix for calibration; that -->
<!-- requires addressing the depth/bimodality issue directly. -->
