---
title: "Beta-binomial pU-QTL: the 26 sites that survive permutation"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 3
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 11)

BASE <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl"
DAT  <- file.path(BASE, "qtl_betabinomial/site26_page_data")
DIAG <- file.path(BASE, "qtl_betabinomial/diagnostic_figures")

pan   <- fread(file.path(DAT, "geno_panels.tsv.gz"))
labs  <- fread(file.path(DAT, "site_labels.tsv"))
qq    <- fread(file.path(DAT, "qq2.tsv.gz"))
lam   <- fread(file.path(DAT, "lambda2.tsv"))
cnt   <- fread(file.path(DAT, "counts.tsv"))
patho <- fread(file.path(DIAG, "pathological_sites.txt"))

pan[, geno := factor(geno, levels = c(0, 1, 2))]
COND_LV <- c("all sites", "minus the 26 sites", "minus all U/C-aliasing sites")
GRP <- c("Mechanism A: U/C variant on the pU base",
         "Mechanism A: same shape, no variant called",
         "Mechanism B: one cell line dominates",
         "Comparison: surviving significant hits",
         "Comparison: random non-significant sites")

# Panel label. Kept short: these figures carry up to 14 facets.
mk_lab <- function(g, with_p = TRUE) {
  L <- labs[group == g]
  setkey(L, sid)
  lab <- L[, sprintf("chr%s:%s%s\n%spooled %%pU=%.3f | %s reads",
                     chr, format(pos, big.mark = ",", trim = TRUE),
                     ifelse(snp_at_site == "", "", paste0("  (", snp_at_site, ")")),
                     ifelse(with_p & is.finite(nlp), sprintf("-log10p=%.0f | ", nlp), ""),
                     pooled_pu, format(total_depth, big.mark = ",", trim = TRUE))]
  data.table(sid = L$sid, lab = lab)
}
attach_lab <- function(g, with_p = TRUE) {
  d <- merge(pan[group == g], mk_lab(g, with_p), by = "sid")
  ord <- labs[group == g][order(-nlp)]$sid
  d[, lab := factor(lab, levels = unique(merge(data.table(sid = ord),
                     mk_lab(g, with_p), by = "sid", sort = FALSE)$lab))]
  d
}

# ---- figure 1 style: %pU against read depth (one dot per cell line) -------
plot_pu_depth <- function(dd, title, subtitle, ncol = 4) {
  ggplot(dd, aes(pu, pmax(n, 1))) +
    geom_vline(xintercept = c(0, 0.5, 1), colour = "grey88", linewidth = 0.4) +
    geom_point(colour = "#256abf", alpha = 0.7, size = 1.6) +
    scale_x_continuous(limits = c(-0.04, 1.04), breaks = c(0, 0.5, 1),
                       labels = c("0", "50", "100")) +
    scale_y_log10(labels = scales::comma) +
    facet_wrap(~ lab, ncol = ncol) +
    labs(x = "%pU in one cell line", y = "read depth (log scale)",
         title = title, subtitle = subtitle) +
    theme_bw(base_size = 11) +
    theme(strip.text = element_text(size = 7.5), panel.grid.minor = element_blank())
}

# ---- figure 2 style: %pU against genotype at the lead cis-SNP -------------
# A boxplot IS appropriate here: we are comparing three defined groups, which is
# what a box summarises well. Groups with fewer than 5 cell lines are outlined in
# red, because a "genotype effect" resting on 2 cell lines is not interpretable.
plot_pu_geno <- function(dd, title, subtitle, ncol = 4) {
  gn <- dd[, .(n_cl = .N, small = .N < 5), by = .(lab, geno)]
  dd2 <- merge(dd, gn, by = c("lab", "geno"))
  ggplot(dd2, aes(geno, pu)) +
    geom_boxplot(aes(colour = small), fill = "grey96", width = 0.6,
                 outlier.shape = NA, linewidth = 0.45) +
    geom_jitter(aes(size = pmax(n, 1)), width = 0.16, height = 0,
                colour = "#256abf", alpha = 0.65) +
    geom_text(data = gn, aes(geno, y = 1.09, label = paste0("n=", n_cl)),
              size = 2.4, colour = "grey30", inherit.aes = FALSE) +
    scale_colour_manual(values = c(`FALSE` = "grey45", `TRUE` = "#e34948"),
                        labels = c("5 or more cell lines", "fewer than 5"),
                        name = "genotype group size") +
    scale_size_continuous(trans = "log10", range = c(0.7, 3.6),
                          breaks = c(10, 100, 1000, 10000),
                          labels = scales::comma, name = "read depth (n)") +
    scale_y_continuous(limits = c(-0.04, 1.16), breaks = c(0, 0.5, 1),
                       labels = c("0", "50", "100")) +
    facet_wrap(~ lab, ncol = ncol) +
    labs(x = "copies of the ALT allele at the lead cis-SNP", y = "%pU",
         title = title, subtitle = subtitle) +
    theme_bw(base_size = 11) +
    theme(legend.position = "bottom", strip.text = element_text(size = 7.5),
          panel.grid.minor = element_blank())
}
```

# Summary

The beta-binomial cis scan produced top hits that **a phenotype permutation did
not displace** — 19 of the 20 strongest sites under permutation were the same
sites as the real top 20, none sharing a lead SNP. Permutation is supposed to
destroy all genotype–phenotype signal, so those sites were being selected for
some reason other than genotype.

`r nrow(patho)` sites are responsible. This page describes what they are and what
removing them does to calibration.

```{r headline}
knitr::kable(cnt, col.names = c("", "n"), caption = "Scale")
```

- **`r sum(labs$group == GRP[1] )` of the 26 carry a U↔C sequence variant on the
  pU base itself**, which makes the measured "%pU" partly a read-out of genotype
  instead of modification (§2.2).
- **`r sum(labs$group == GRP[2])` more have no variant called** at the base. Only
  one of those four has the heterozygote band a U↔C variant requires; the other
  three show a two-level split that aliasing does not explain, and are
  **unexplained** (§2.3).
- **`r sum(labs$group == GRP[3])` are a different problem**: sites with almost no
  modification where a single cell line supplies nearly all converted reads
  (§2.4).
- Removing the 26 collapses the QQ tail from
  `r round(lam[condition == "all sites" & dataset == "real"]$max_nlp, 0)` to
  `r round(lam[condition == "minus the 26 sites" & dataset == "real"]$max_nlp, 0)`;
  removing every U/C site instead is what fixes λ (§3).

A companion page, [the full pathology write-up](qtl_mapping_betabinomial_pathology_summary.html),
covers the statistical mechanism and the wider audit of the other significant hits.

# 1. How the 26 sites were found

The cis scan was rerun six times — once on the real data, once under each of five
phenotype permutations (`set.seed(1000 + k)`, phenotype and covariates reindexed
together, genotypes untouched) — this time **keeping the site identity attached to
every p-value**, which the original calibration script discarded.

Under permutation no genotype–phenotype relationship exists, so any site still
returning an extreme p-value there is broken on its own terms. Two criteria:

```{r criteria}
knitr::kable(data.table(
  criterion = c("nominal p < 1e-10 in at least one of the 5 permutations",
                "and real -log10(p) > 10"),
  `sites` = c("27", as.character(nrow(patho)))),
  caption = paste("Selection used the PERMUTED data only, so looking at the real",
                  "QQ plot afterwards (§3) is not circular."))
```

Of the 27 candidates, one was extreme in a single permutation by chance. The other
**`r nrow(patho)` were extreme in all six datasets** — checked individually, and
their permutation-to-permutation spread is tiny (median SD
`r sprintf("%.2f", median(patho$perm_sd))` on the `-log10 p` scale), so this is a
stable property of each site rather than a fluke.

```{r sec1-fig, fig.height=6.5}
P <- paste0("perm", 1:5)
lg <- melt(patho[, c("sid", "real", P), with = FALSE], id.vars = "sid",
           variable.name = "dataset", value.name = "nlp")
lg[, sid := factor(sid, levels = patho[order(real)]$sid)]
ggplot(lg, aes(nlp, sid)) +
  geom_line(aes(group = sid), colour = "grey70", linewidth = 0.4) +
  geom_point(data = lg[dataset != "real"], aes(colour = "the 5 permutations"),
             size = 1.7, alpha = 0.85) +
  geom_point(data = lg[dataset == "real"], aes(colour = "real data"), size = 2.6) +
  scale_colour_manual(values = c(`real data` = "#e34948",
                                 `the 5 permutations` = "#3987e5"), name = NULL) +
  labs(x = expression(-log[10]("p")~"of the site's best cis-SNP"), y = NULL,
       title = "At all 26 sites the permutations reach almost the same p-value as the real data",
       subtitle = "A working site would have its permutation points near 0.") +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

# 2. What the 26 sites are

## 2.1 Classification

```{r sec2-class}
knitr::kable(labs[is26 == TRUE, .(
  site = sprintf("chr%s:%s", chr, format(pos, big.mark = ",", trim = TRUE)),
  `-log10p` = nlp,
  `variant on the pU base` = ifelse(snp_at_site == "", "--", snp_at_site),
  `pooled %pU` = round(pooled_pu, 3),
  `top cell-line share` = round(top_share, 2),
  `lead cis-SNP` = lead_snp,
  `its distance` = sprintf("%+d bp", lead_dist),
  mechanism = c(`Mechanism A: U/C variant on the pU base` = "A: U/C variant on the base",
                `Mechanism A: same shape, no variant called` = "unexplained (see 2.3)",
                `Mechanism B: one cell line dominates` = "B: one cell line dominates")[group]
  )][order(-`-log10p`)],
  caption = paste("All 26. 'top cell-line share' is the largest fraction of the",
                  "site's converted reads coming from one cell line.",
                  "The four marked 'unexplained' have no variant called on the pU",
                  "base and, with one exception, lack the heterozygote band that",
                  "aliasing would produce -- see 2.3."))
```

## 2.2 Mechanism A: a U↔C variant on the pU base

### What the problem is

Pseudouridine is a **modified uridine** — chemically still a U, and the assay
detects it by reading the base at that position and scoring conversions. The
readout is therefore a statement about *what base is there*, and it cannot
distinguish two different reasons for seeing something other than a plain U:

1. the U was **modified** to pseudouridine — the biology we want to measure; or
2. the cell line **inherited a C** at that position instead of a U.

When a common sequence variant changes that exact base from U to C, the two are
indistinguishable. A cell line homozygous for the C allele reads as ~100%
"modified" at every read; a heterozygote reads ~50%, because half its transcripts
carry the C; a homozygous-reference cell line reads whatever the true modification
rate is. **The phenotype becomes a genotype assay.**

That gives three consequences, all visible below:

- %pU clusters into three discrete levels near 0%, 50% and 100% rather than
  varying continuously;
- those levels track the variant almost perfectly, so any SNP in linkage with it
  predicts the phenotype — a real association, but of the wrong thing; and
- the effect does not weaken with depth, because it is not a sampling artefact.

Written against the genome this appears as **C/T on the + strand and A/G on the −
strand** — the same U→C change on the transcript, seen from whichever strand the
transcript came from. Genome-wide,
**`r format(cnt[grepl("U/C-aliasing sites genome-wide", quantity)]$n, big.mark = ",")`
sites** carry such a variant.

### %pU across cell lines

```{r sec2a-depth, fig.height=8}
dA <- attach_lab(GRP[1])
plot_pu_depth(dA,
  "Mechanism A: %pU forms discrete bands, because it is reporting genotype",
  paste("All", uniqueN(dA$sid), "sites with a U/C variant called on the pU base.",
        "Each dot is one cell line; the panel header gives the variant.",
        "\nThe bands persist at every read depth, so they are not a coverage artefact."))
```

### %pU against genotype at the lead cis-SNP

If %pU is a genotype read-out, it should line up with genotype in steps rather
than as a gradual trend.

```{r sec2a-geno, fig.height=8.5}
plot_pu_geno(dA,
  "Mechanism A: %pU against copies of the ALT allele at the lead cis-SNP",
  paste("Dot size is read depth. At most sites the three genotype groups sit at",
        "separated %pU levels -- a step function rather than a dose-response.",
        "\nchr6:31,271,283 is the exception, showing a gradient instead.",
        "\nBoxes outlined in red rest on fewer than 5 cell lines, so their level is",
        "not interpretable."))
```

## 2.3 Four sites with no variant called — mostly unexplained

Four of the 26 have no variant called on the pU base. An earlier version of this
page assigned them to mechanism A on the strength of their %pU shape. **The plots
below do not support that**, so the counts are set out explicitly.

A codominant U↔C variant makes a specific prediction: three levels, including a
**heterozygote band near 50%**, because half of a heterozygote's transcripts carry
the C. That band is the signature, and it is testable.

```{r sec2b-bins}
bin_tab <- function(g) {
  ids <- labs[group == g][order(-nlp)]$sid
  rbindlist(lapply(ids, function(s) {
    pu <- pan[sid == s]$pu
    data.table(site = sprintf("chr%s:%s", sub(" .*", "", s),
                              format(as.integer(sub(".* ", "", s)), big.mark = ",")),
               `~0%` = sum(pu <= 0.08),
               `8-42%` = sum(pu > 0.08 & pu < 0.42),
               `~50% (42-58%)` = sum(pu >= 0.42 & pu <= 0.58),
               `58-92%` = sum(pu > 0.58 & pu < 0.92),
               `~100%` = sum(pu >= 0.92))
  }))
}
bt <- bin_tab(GRP[2])
knitr::kable(bt, caption = paste("Cell lines per %pU band, for the four sites with",
                                 "no variant call. The middle column is the",
                                 "heterozygote band a U/C variant requires."))
uc_ids <- labs[group == GRP[1]]$sid
n50_uc <- vapply(uc_ids, function(s) sum(pan[sid == s]$pu >= 0.42 & pan[sid == s]$pu <= 0.58), numeric(1))
n50_ns <- bt$`~50% (42-58%)`
knitr::kable(data.table(
  `site set` = c("the 14 confirmed U/C sites", "the 4 with no variant call"),
  `have a heterozygote band (>=5 cell lines at 42-58%)` =
    c(sprintf("%d of %d", sum(n50_uc >= 5), length(n50_uc)),
      sprintf("%d of %d", sum(n50_ns >= 5), length(n50_ns)))),
  caption = "The band is present at most confirmed sites and absent at most of these four.")
```

```{r sec2b, fig.height=3.4}
dB <- attach_lab(GRP[2])
plot_pu_depth(dB,
  "The four sites with no variant called on the pU base",
  paste("Only chr6:32,745,178 shows tight clusters, and they are at 0% and 100%",
        "with no 50% group. The other three are diffuse.",
        "\nThis is not the clean three-level pattern of 2.2."),
  ncol = 4)
```

```{r sec2b-geno, fig.height=3.8}
plot_pu_geno(dB,
  "The same four, against genotype at the lead cis-SNP",
  paste("No stepped pattern. Note also that two of the genotype-0 groups contain a",
        "single cell line (red), so no level can be read off them."), ncol = 4)
```

**What this leaves.** Only **chr14:22,551,976** has a genuine three-level structure
(`r n50_ns[1]` cell lines in the heterozygote band), and for that site the
uncalled-U/C explanation is reasonable. The other three show a **two-level split**
— a wall at 0% and a group at high %pU, with no heterozygote band — which is *not*
what a codominant U↔C variant produces. They remain **unexplained**: they behave
pathologically under permutation but neither mechanism on this page accounts for
them.

One further inconvenience for the aliasing story: the two of these four that sit
in the chr6 MHC (`r paste(sprintf("chr%s:%s", labs[group == GRP[2] & chr == "6"]$chr, format(labs[group == GRP[2] & chr == "6"]$pos, big.mark = ",")), collapse = " and ")`)
are among the three *without* the heterozygote band. The MHC is the most
polymorphic and hardest-to-map region of the genome, so an uncalled variant there
would be unsurprising — but it is the wrong site that looks like aliasing, so MHC
location is not evidence for it. For the record, **`r sum(labs[is26 == TRUE]$chr == "6")`
of the 26 are in the MHC** (chr6:28.5–33.4 Mb, ~0.16% of the genome), which is a
strong enrichment and worth following up on its own terms.

## 2.4 Mechanism B: one cell line supplies nearly all the converted reads

These sites have essentially **no modification** — pooled %pU between
`r sprintf("%.1f%%", 100*min(labs[group == GRP[3]]$pooled_pu))` and
`r sprintf("%.1f%%", 100*max(labs[group == GRP[3]]$pooled_pu))`. Most cell lines
report zero converted reads. Then one cell line reports a large %pU, and it alone
accounts for up to
`r sprintf("%.0f%%", 100*max(labs[group == GRP[3]]$top_share, na.rm = TRUE))` of
every converted read at the site.

```{r sec2c-depth, fig.height=5}
dC <- attach_lab(GRP[3])
plot_pu_depth(dC,
  "Mechanism B: a near-empty site plus one discordant cell line",
  paste("A wall of cell lines at exactly 0%, and one or two lone dots to the",
        "right. Nothing here supports a genotype effect."))
```

```{r sec2c-geno, fig.height=5.5}
plot_pu_geno(dC,
  "Mechanism B: %pU against genotype at the lead cis-SNP",
  paste("The apparent 'effect' is one or two cell lines lifting a single genotype",
        "group off the floor -- not a shift of the group as a whole."))
```

## 2.5 For comparison: what ordinary sites look like

Same axes, same builders. Two comparison sets: significant hits that are not
U/C-aliasing and not concentration-driven, and sites drawn at random from those
tested but never significant.

### Significant hits, spanning the significance range

```{r sec2d-depth, fig.height=4.4}
dD <- attach_lab(GRP[4])
plot_pu_depth(dD, "Ordinary significant hits: %pU against read depth",
  "Chosen to span the whole significance range, from the strongest to the weakest.",
  ncol = 3)
```

```{r sec2d-geno, fig.height=5}
plot_pu_geno(dD, "Ordinary significant hits: %pU against genotype",
  paste("Compare with 2.2: the genotype groups overlap and shift gradually",
        "rather than sitting at separated levels."), ncol = 3)
```

### Random sites, never significant

```{r sec2e-depth, fig.height=4.4}
dE <- attach_lab(GRP[5], with_p = FALSE)
plot_pu_depth(dE, "Random non-significant sites: %pU against read depth",
  paste("The typical site: low %pU and shallow coverage in most cell lines.",
        "Shallow coverage is the norm in this dataset, not a property of the",
        "pathological sites."), ncol = 3)
```

```{r sec2e-geno, fig.height=5}
plot_pu_geno(dE, "Random non-significant sites: %pU against genotype",
  "No relationship, which is what the bulk of the genome should look like.",
  ncol = 3)
```

# 3. Calibration after removing these sites

Two exclusions, each shown against the unfiltered scan. The 26 were selected from
the permuted data alone (§1), so the real panels are not circular.

```{r sec3-qq, fig.height=9}
qq[, condition := factor(condition, levels = COND_LV)]
qq[, dataset := factor(dataset, levels = c("real", paste0("perm", 1:5)))]
COLS <- c(real = "#e34948", perm1 = "#9ec5f4", perm2 = "#6da7ec",
          perm3 = "#3987e5", perm4 = "#256abf", perm5 = "#0d366b")
ggplot(qq, aes(expected, observed, colour = dataset)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45") +
  geom_point(size = 0.6, alpha = 0.75) +
  scale_colour_manual(values = COLS, name = NULL) +
  guides(colour = guide_legend(override.aes = list(size = 2.6, alpha = 1), nrow = 1)) +
  facet_wrap(~ condition, ncol = 1, scales = "free_y") +
  labs(x = expression(-log[10]("expected p")), y = expression(-log[10]("observed p")),
       title = "Beta-binomial LRT calibration, before and after each exclusion",
       subtitle = paste("Note the free y-axis. The permutation series (blues) should",
                        "hug the dashed line; where they rise above it the test is",
                        "producing signal from shuffled data.")) +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

```{r sec3-tab}
ORD <- c("real", paste0("perm", 1:5))
lw <- dcast(lam, dataset ~ condition, value.var = "lambda")
setcolorder(lw, c("dataset", COND_LV)); lw <- lw[order(factor(dataset, levels = ORD))]
knitr::kable(lw, digits = 3, caption = "lambda, the median-based inflation factor")

tw <- dcast(lam, dataset ~ condition, value.var = "max_nlp")
setcolorder(tw, c("dataset", COND_LV)); tw <- tw[order(factor(dataset, levels = ORD))]
knitr::kable(tw, digits = 1, caption = "Largest observed -log10(p): the QQ tail height")
```

The two tables move independently, and that is the main result here.

```{r sec3-compare}
g <- function(cond, col) {
  v <- lam[condition == cond & dataset == "real"][[col]]
  if (col == "lambda") round(v, 3) else round(v, 1)
}
pmax_of <- function(cond) round(max(lam[condition == cond & dataset != "real"]$max_nlp), 1)
knitr::kable(data.table(
  condition = COND_LV,
  `real lambda` = vapply(COND_LV, g, numeric(1), col = "lambda"),
  `real tail` = vapply(COND_LV, g, numeric(1), col = "max_nlp"),
  `worst permuted tail` = vapply(COND_LV, pmax_of, numeric(1)),
  `sites removed` = c(0L, nrow(patho),
                      cnt[grepl("U/C-aliasing sites genome-wide", quantity)]$n)),
  row.names = FALSE,
  caption = "With the phenotype shuffled the permuted tail should be near 0.")
```

**Removing the 26 fixes the tail.** The largest `-log10 p` falls from
`r g("all sites", "max_nlp")` to `r g("minus the 26 sites", "max_nlp")` in the
real data and from `r pmax_of("all sites")` to
`r pmax_of("minus the 26 sites")` in the permutations — so essentially all of the
manufactured signal lived in those 26 sites. λ barely moves
(`r g("all sites", "lambda")` → `r g("minus the 26 sites", "lambda")`), because λ
is computed from the *median* statistic and 26 sites out of ~11,000 cannot shift a
median.

**Removing all U/C sites fixes λ.** The opposite pattern: λ drops to
`r g("minus all U/C-aliasing sites", "lambda")`, a much larger change, while the
tail stays at `r g("minus all U/C-aliasing sites", "max_nlp")` — because the four
sites with no variant call (§2.3) are not in the U/C list and are among the worst
offenders.

Neither exclusion alone does both. Worth noting for interpretation: the ~1,100
U/C sites that are *not* among the 26 are not a calibration failure at all —
permutation removes their signal normally. They are a problem because the
phenotype is a genotype read-out, so a statistically sound association is being
reported as a modification QTL.

<!-- # Caveats -->

<!-- - **The four sites with no variant call** are assigned to mechanism A on the -->
<!--   strength of their %pU shape. Confirming that needs base-level sequence at those -->
<!--   positions, not more statistics. -->
<!-- - **A large-effect QTL and aliasing look alike.** If a variant largely abolished -->
<!--   modification, non-carriers would sit high, homozygous carriers near zero and -->
<!--   heterozygotes between — the same three bands. So banding is a reason to check -->
<!--   the base, not a verdict on its own. This matters for the ordinary hits in §2.5 -->
<!--   as well; the [companion page](qtl_mapping_betabinomial_pathology_summary.html) -->
<!--   quantifies how many are affected. -->
<!-- - **The exclusion has not been re-run.** These figures filter the existing -->
<!--   results. Rebuilding the phenotype matrix without the excluded sites would also -->
<!--   change the multiple-testing correction. -->
<!-- - **Assay direction unconfirmed.** The U/C reading assumes pseudouridine is -->
<!--   detected at the same base the variant changes; worth confirming with whoever ran -->
<!--   the protocol. -->

<!-- # Reproducing this page -->

<!-- ```{r prov} -->
<!-- knitr::kable(fread(file.path(DAT, "runinfo.tsv")), caption = "Parameters") -->
<!-- knitr::kable(data.table( -->
<!--   step = c("site-resolved calibration scan (6 datasets, site identity kept)", -->
<!--            "selection of the 26 sites", -->
<!--            "U/C-aliasing site list", -->
<!--            "classification of the 26", -->
<!--            "genotype panels + QQ for this page", -->
<!--            "this page"), -->
<!--   script = c("8.calibration_site_resolved.R", "9.qq_exclude_pathological.R", -->
<!--              "9.qq_exclude_pathological.R (+ .traw allele check)", -->
<!--              "10.summary_page_data.R", "11.site26_page_data.R", -->
<!--              "analysis/qtl_mapping_betabinomial_site26.Rmd"), -->
<!--   submitted_via = c("run_calibration_siteres.sbatch", "-- (light)", "-- (light)", -->
<!--                     "run_summary_page_data.sbatch", "run_site26_page_data.sbatch", -->
<!--                     "run_render_rmd.sbatch")), -->
<!--   caption = "Provenance. The heavy steps need a compute node.") -->
<!-- ``` -->
