---
title: "QTL mapping: beta-binomial model"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r setup, warning=FALSE, message=FALSE}
library(data.table)
library(ggplot2)

knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", fig.width = 10, fig.height = 6
)

PROJECT_DIR <- "/project/xinhe/xsun/pumseq"
BB  <- file.path(PROJECT_DIR, "pumseq_processed/psi_qtl/qtl_betabinomial")
BBR <- file.path(BB, "results")
BIN <- file.path(PROJECT_DIR, "pumseq_processed/psi_qtl/qtl_binomial/results")
WINDOWS_KB <- c(5, 20, 50, 100)
PC_COUNTS  <- c(3, 5, 7, 10, 15, 20)
```

# Introduction

This page tests the **beta-binomial**, which models that overdispersion directly:

$$
y_i \sim \text{BetaBinomial}(n_i, \mu_i, \phi), \qquad
\text{Var}(y_i) = n_i \mu_i (1-\mu_i)\,[\,1 + (n_i - 1)\phi\,],
$$

with the same mean model `logit(mu_i) = beta_0 + beta_g G_i + z_i'gamma`, the
same 5-fold union sites (12,416), the same 50 cell lines, and the same M-value
PC covariates. `phi` is the intra-class-correlation dispersion (`phi = 0`
recovers the binomial).

<!-- # 1. Is the data overdispersed? -->

<!-- For each site we fit the null model (no genotype) under both the binomial and -->
<!-- the beta-binomial and record the dispersion `phi` and a boundary-corrected LRT -->
<!-- for `H0: phi = 0`. -->

<!-- ```{r overdisp} -->
<!-- od <- fread(file.path(BB, "overdispersion_pc5.txt.gz")) -->
<!-- knitr::kable(data.table( -->
<!--   sites_fit          = nrow(od), -->
<!--   `median phi`       = round(median(od$phi), 4), -->
<!--   `mean phi`         = round(mean(od$phi), 3), -->
<!--   `q90 phi`          = round(quantile(od$phi, 0.9), 3), -->
<!--   `phi > 0.05 (%)`   = round(100 * mean(od$phi > 0.05), 1), -->
<!--   `overdispersed FDR<0.05 (%)` = round(100 * mean(od$fdr < 0.05), 1) -->
<!-- ), caption = "Per-site overdispersion (null model, 5 PCs)") -->
<!-- ``` -->

<!-- ```{r overdisp-fig, results='asis'} -->
<!-- cat(sprintf('<img src="%s" style="max-width:70%%;">', -->
<!--             knitr::image_uri(file.path(BB, "overdispersion_pc5.png")))) -->
<!-- ``` -->

<!-- The **median site has phi ~ 0** (binomial is fine for most), but the -->
<!-- distribution has a heavy right tail: **~22% of sites are significantly -->
<!-- overdispersed** (phi up to ~0.5). At exactly those sites the binomial p-values -->
<!-- are anti-conservative — which is what the beta-binomial corrects. -->

<!-- # 2. Model and significance -->

<!-- The beta-binomial **score test** for `beta_g` is the binomial score engine with -->
<!-- the residual and working weight each divided by `c_i = 1 + (n_i-1)phi` -->
<!-- (validated against `aod::betabin`'s LRT). Significance uses the **same procedure -->
<!-- as the binomial**: per-SNP score p, the FastQTL/tensorQTL **permutation -->
<!-- beta-approximation** within each site, then BH-FDR across sites. To keep the -->
<!-- permutation fast, `phi` is estimated once per site and held fixed across -->
<!-- permutations (each permutation then refits only the mean — a weighted binomial -->
<!-- GLM). Swept over the same 6 PC x 4 window grid. -->

# 1. Results: QTL sites across the full PC range

```{r load-bb}
read_bb <- function(w, k) {
  f <- file.path(BBR, sprintf("betabin_w%dkb_pc%d.txt.gz", w, k))
  if (!file.exists(f)) return(NULL)
  d <- fread(f); d[, `:=`(window_kb = w, n_pcs = k)]; d
}
grid <- CJ(w = WINDOWS_KB, k = PC_COUNTS)
bb <- rbindlist(mapply(read_bb, grid$w, grid$k, SIMPLIFY = FALSE), use.names = TRUE, fill = TRUE)
bb_sum <- bb[, .(n_tested = .N, qtl_FDR05 = sum(fdr < 0.05, na.rm = TRUE)),
             by = .(window_kb, n_pcs)][order(window_kb, n_pcs)]
bb_sum[, window_lab := factor(paste0(window_kb, "kb"), levels = paste0(WINDOWS_KB, "kb"))]

knitr::kable(dcast(bb_sum, window_kb ~ n_pcs, value.var = "n_tested"),
             caption = "Sites the beta-binomial could fit (rows=window kb, cols=PCs)")
```

The beta-binomial MLE jointly estimates `k + 2` parameters (PCs + intercept +
dispersion) on only 50 samples. An earlier version of this pipeline (relying
on `aod::betabin`'s optimizer) appeared to stop converging for most sites
once k >= 10 (fittable sites collapsing from thousands to tens). Replacing
that optimizer with a properly-seeded fit (see the [model
page](qtl_mapping_betabinomial_model.html)) resolved this — sites fittable at
k=20 rose to ~7,600 (close to the binomial's ~7,900 at the same k). The
beta-binomial is now reliably fittable across the **full swept range**,
k=3-20.

```{r bb-disc}
knitr::kable(dcast(bb_sum, window_kb ~ n_pcs, value.var = "qtl_FDR05"),
             caption = "Beta-binomial QTL sites (FDR<0.05)")
```

```{r bb-vs-pc, fig.height=5}
ggplot(bb_sum, aes(x = n_pcs, y = qtl_FDR05, color = window_lab)) +
  geom_line() + geom_point(size = 2) +
  scale_x_continuous(breaks = PC_COUNTS) +
  labs(x = "Number of covariate PCs", y = "QTL sites (FDR<0.05)", color = "cis-window",
       title = "Beta-binomial QTL sites vs. PCs") +
  theme_bw(base_size = 13)
```

# 2. Beta-binomial vs. binomial

At **k = 3** (both models fit ~11,000 sites — a fair comparison), the
beta-binomial finds **~1.3-1.4x more** QTL sites than the binomial:

```{r head-to-head}
read_bin <- function(w, k) fread(file.path(BIN, sprintf("scan_w%dkb_pc%d.txt.gz", w, k)))
h2h <- rbindlist(lapply(WINDOWS_KB, function(w) {
  db <- read_bb(w, 3); dn <- read_bin(w, 3)
  data.table(window = paste0(w, "kb"),
             binomial   = sum(dn$fdr < 0.05, na.rm = TRUE),
             betabinom  = sum(db$fdr < 0.05, na.rm = TRUE))
}))
h2h[, `fold change` := round(betabinom / binomial, 1)]
knitr::kable(h2h, caption = "QTL sites (FDR<0.05) at k=3: binomial vs. beta-binomial")
```

This is the **opposite** of removing false positives — the beta-binomial
*recovers* signal. The likely reason: at the overdispersed sites the binomial's
permutation null is itself inflated (both the observed and the permuted min-p
are too small), so the binomial is effectively **conservative** there and buries
real QTLs in overdispersion-driven noise. The beta-binomial divides that
variance out, de-inflates the null, and recovers those associations.

```{r phi-of-hits, fig.height=4.5}
hits <- read_bb(5, 3)
hits[, group := ifelse(fdr < 0.05, "QTL (FDR<0.05)", "not significant")]
ggplot(hits, aes(x = phi, fill = group)) +
  geom_histogram(bins = 50, position = "identity", alpha = 0.6) +
  scale_fill_manual(values = c("QTL (FDR<0.05)" = "firebrick", "not significant" = "grey60")) +
  labs(x = "site dispersion phi", y = "sites", fill = NULL,
       title = "Dispersion of beta-binomial QTL sites vs. the rest (k=3, 5kb)") +
  theme_bw(base_size = 12)
```

# 3. Calibration check: close to calibrated

Whether the extra discoveries are real power recovery or an artifact of how
`phi` is estimated hinges on the beta-binomial's own calibration. We ran the
same permutation calibration check used for the binomial (5 seeds, same
config: 5kb, 3 PCs) — with a full beta-binomial refit per site per
permutation (`phi` re-estimated fresh each time, not held fixed), the correct
test of whether the beta-binomial's own p-values are calibrated. Model
details, including two bugs found and fixed during this validation (a
quasi-score approximation, and an unreliable model-fitting optimizer): [model
page](qtl_mapping_betabinomial_model.html).

```{r bb-calib-lambda}
CAL <- file.path(BB, "calibration")
lam <- fread(file.path(CAL, "calib_w5kb_pc3_lambda.txt"))
knitr::kable(lam, digits = 2,
             caption = "Genomic-inflation lambda (5kb, 3 PCs): real vs. 5 permuted nulls (beta-binomial)")
```

```{r bb-calib-qq, results='asis'}
cat(sprintf('<img src="%s" style="max-width:70%%;">',
            knitr::image_uri(file.path(CAL, "calib_w5kb_pc3_qq.png"))))
```

```{r bb-calib-compare}
knitr::kable(data.table(
  model = c("binomial", "beta-binomial"),
  real_lambda = c(5.28, round(lam[dataset == "real"]$lambda, 2)),
  permuted_null_lambda_range = c("3.12 - 3.62", sprintf("%.2f - %.2f",
                                  min(lam[dataset != "real"]$lambda),
                                  max(lam[dataset != "real"]$lambda)))
), caption = "Nominal-p calibration: binomial vs. beta-binomial")
```

**The beta-binomial brings the permuted-null lambda down to ~1.13** (range
1.07-1.22 across 5 seeds) — close to the ideal 1, and a large improvement
over the binomial's ~3.4. This required fixing two bugs found during this
validation exercise (see the [model page](qtl_mapping_betabinomial_model.html)):
an earlier "quasi-score" approximation to the beta-binomial score test, and a
model-fitting optimizer (`aod::betabin`) that sometimes failed to converge to
the true MLE. With both fixed, the ~1.3-1.4x discovery increase over the
binomial is well-supported as real power recovery — the binomial's
overdispersion-driven miscalibration was burying signal at ~22% of sites
(those significantly overdispersed), and the beta-binomial recovers it while
also being much better calibrated.

**Practical takeaway**: the beta-binomial is the better model to report —
substantially closer to calibrated than the binomial, with a modest and
defensible increase in discoveries, not an inflated one.
