---
title: "How well can we estimate dispersion at each pU site?"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 2
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 10)
PSI <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl"
RC  <- file.path(PSI, "qtl_betabinomial/phi_recovery")
LP  <- file.path(PSI, "qtl_betabinomial/low_phi_validation")
```

Dispersion (φ) describes variation in pU rates beyond binomial read-sampling
noise, after accounting for the covariates. We tested how well it can be
estimated at individual sites using simulations based on our data.

**Small φ is difficult to estimate at many sites.** Estimates often approach
zero, especially when the data contain little information. Shrinkage reduces
this pile-up, but at small φ it often returns a value near the shared prior
centre rather than distinguishing individual sites.

# 1. Can the MLE recover a known φ?

For each site, we keep the real read depths and the expected pU rates fitted
from the real data using three phenotype PCs. We choose a known φ, simulate
counts from the beta-binomial model, and refit the model to those counts.

Each panel below uses one true φ for all sites. We repeat this at six values
from 0.005 to 0.4, with about 10,000 sites and typically 49 covered cell lines
per site. The generating means vary across sites and cell lines; the refit
estimates the means and φ from the simulated counts.

```{r prob, fig.height=7, fig.width=13}
RB <- fread(file.path(RC, "recovery_by_site.txt.gz"), sep = "\t")[is.finite(phi_hat)]
RS <- fread(file.path(RC, "recovery_summary.tsv"), sep = "\t")
SP <- fread(file.path(RC, "site_properties.txt.gz"), sep = "\t")
RB <- merge(RB, SP[, .(sid, med_depth)], by = "sid")
PT <- sort(unique(RB$phi_true))
RB[, panel := factor(sprintf("true phi = %g", phi_true),
                     levels = sprintf("true phi = %g", PT))]
# colour by READ DEPTH quintile -- the point is that the boundary spike and the
# hump on the truth are made of different sites
qb <- quantile(SP$med_depth, 0:5/5)
RB[, dq := cut(med_depth, qb, include.lowest = TRUE,
               labels = c("Lowest depth", "2", "3", "4", "Highest depth"))]
vl <- merge(unique(RB[, .(panel, phi_true)]), RS[, .(phi_true, mode = `mode phi_hat`)], by = "phi_true")
ggplot(RB, aes(phi_hat, fill = dq)) +
  geom_histogram(bins = 60, colour = NA) +
  geom_vline(xintercept = 1e-6, colour = "grey35", linetype = "dotted") +
  geom_vline(data = vl, aes(xintercept = phi_true), colour = "#c0392b", linewidth = 0.7) +
  facet_wrap(~ panel, nrow = 2, scales = "free_y") +
  scale_x_log10(breaks = 10^seq(-6, 0, 2), labels = c("1e-6", "1e-4", "1e-2", "1")) +
  scale_fill_viridis_d(option = "viridis", direction = -1,
                       name = "Read-depth quintile") +
  guides(fill = guide_legend(nrow = 1, reverse = TRUE)) +
  labs(x = expression(hat(varphi)*", axis log-spaced"), y = "sites",
       title = "Recovery of known dispersion, coloured by read depth",
       subtitle = "Red: true phi. Dotted grey: numerical floor (1e-6). Colours show the depth groups within each bar.") +
  theme_bw(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        plot.subtitle = element_text(size = 8))
```

At small true φ, the estimates include a group near the truth and a pile-up
at the numerical floor, 10⁻⁶. The pile-up becomes smaller as true φ increases.
A boundary estimate can arise from limited information; it does not by itself
show that the optimiser is broken or that the true dispersion is zero.

The histogram and accuracy table include successful fits only. Failed fits
are reported separately below.

```{r fitcounts}
all_recovery <- fread(file.path(RC, "recovery_by_site.txt.gz"))
knitr::kable(all_recovery[, .(simulated = .N,
  successful = sum(is.finite(phi_hat)), failed = sum(!is.finite(phi_hat))),
  by = .(`true phi` = phi_true)][order(`true phi`)],
  caption = "Fit success at each simulated dispersion.")
```

```{r probtab}
knitr::kable(RB[, { r <- phi_hat/phi_true[1]
  .(`within 2x of truth` = sprintf("%.1f%%", 100*mean(r > 0.5 & r < 2)),
    `>10x too low`       = sprintf("%.1f%%", 100*mean(r < 0.1)),
    `at the 1e-6 floor`  = sprintf("%.1f%%", 100*mean(phi_hat < 2e-6)),
    `median fold error`  = round(median(pmax(r, 1/r)), 1)) },
  by = .(`true phi` = phi_true)][order(`true phi`)],
  caption = "Per-site accuracy against known truth. Accuracy among successful fits. A reasonable population mode can coexist with poor estimates at individual sites.")
```

# 2. Read depth and mean pU both matter

Deeper sites generally have more accurate estimates. The table below compares
recovery across read-depth groups at true φ = 0.0126.

```{r accdepth}
knitr::kable(fread(file.path(RC, "accuracy_by_depth.tsv")),
  caption = "Recovery at true phi = 0.0126, grouped by read depth.")
```

Low mean pU is also associated with estimates near zero. To examine it alongside
depth, we group sites by their **real-data pooled pU percentage** (converted reads
divided by total reads). This is a summary of the observed rate, not the full
vector of fitted cell-line-specific means used to generate the simulations.

```{r mean_depth}
counts_dir <- file.path(PSI, "phenotype/fold5_union_auto_noUCcalled/counts")
C <- fread(file.path(counts_dir, "C_matrix.txt.gz"))
N <- fread(file.path(counts_dir, "N_matrix.txt.gz"))
stopifnot(identical(C[, 1:3], N[, 1:3]))
cls <- setdiff(names(C), c("ref", "pos", "strand"))
cm <- as.matrix(C[, ..cls]); nm <- as.matrix(N[, ..cls])
covered_depth <- nm; covered_depth[covered_depth == 0] <- NA_real_
props <- data.table(sid = paste(C$ref, C$pos),
  pooled_pu = rowSums(cm) / rowSums(nm),
  median_depth = apply(covered_depth, 1, median, na.rm = TRUE))
stopifnot(!anyDuplicated(props$sid))
md <- merge(all_recovery[phi_true == 0.0126 & is.finite(phi_hat)], props, by = "sid")
md[, pu_group := factor(ifelse(pooled_pu < 0.05, "Below 5%", "At least 5%"),
                        levels = c("Below 5%", "At least 5%"))]
md[, depth_group := cut(median_depth, c(0, 5, 15, Inf), include.lowest = TRUE,
                        labels = c("Up to 5", "More than 5 to 15", "More than 15"))]
knitr::kable(md[, .(sites = .N,
  `near-zero estimates` = sprintf("%.1f%%", 100 * mean(phi_hat < 2e-6))),
  by = .(`median reads per covered cell line` = depth_group,
         `pooled pU` = pu_group)][order(`median reads per covered cell line`, `pooled pU`)],
  caption = "Near-zero estimates at true phi = 0.0126, by depth and pooled pU. Successful fits only.")
```

These comparisons support a role for both depth and mean pU. They do not isolate
causal effects: sites differ in their depth distributions and fitted means.
A controlled simulation varying the mean while holding depth and φ fixed would
separate those effects more directly. These results do not yet establish a
universal depth cutoff for reporting per-site φ.

# 3. Does shrinkage improve estimation?

We compared ordinary maximum likelihood (MLE) with several ways to stabilise φ.
MAP estimation maximises the likelihood plus a log prior, pulling weakly
supported estimates toward a shared distribution.

| Approach | Purpose | Main observation |
|---|---|---|
| Multiple optimiser starts | Find a better likelihood optimum | Helps with local optima; does not remove the small-φ pile-up. |
| Select an optimum near the population mode | Move estimates away from the boundary | Does little to improve per-site tracking. |
| Predict a prior centre from site properties | Allow the centre to vary by site | The explored predictors explained little variation in estimated φ. |
| Fixed-centre Gaussian or Student-t prior | Shrink extreme estimates | Reduces boundary estimates, but can pull genuinely large φ downward. |
| Two-component prior | Allow low- and high-dispersion groups | Tracking at large φ is similar to MLE. |
| Moment-derived prior | Learn the prior from moment estimates | Gives the strongest tracking at larger φ among the approaches tested; tracking remains weak at small φ. |

For this comparison, **true φ varies between sites**. We draw site-specific
values from log-normal distributions centred at 0.0126, 0.05, 0.2, or 0.4,
restricted to [10⁻⁵, 0.9]. Counts again use the real depths and fitted means.

The moment-derived approach estimates the mean anew from each simulated dataset,
calculates a moment estimate of φ, and uses the collection of estimates to build
a Student-t prior on log₁₀φ. We compare three rules for the prior centre:

- **mom-median:** median of all finite moment estimates, including negatives.
- **mom-mode:** density mode on the log scale among positive estimates.
- **mom-q60:** 60th percentile of all finite moment estimates.

The prior width uses the spread of positive log-moment estimates. The prior is
rebuilt for each simulation setting before the MAP refits. The moment calculation
is closed form, but it uses means from a fitted model.

## How we compare methods

Being within a factor of two of the truth is useful, but insufficient: if true
values are tightly clustered, a constant guess can score well. We also report:

- **Spearman correlation:** whether estimates rank sites in the right order.
- **Log–log slope:** how strongly estimates change with truth; a slope near zero
  indicates little response, while a slope near one indicates similar scaling.

Neither measure alone establishes accurate estimates. The constant benchmark
uses the true median, which is available only in simulation; its correlation
is undefined.

```{r track}
TR <- fread(file.path(RC, "approach_tracking.tsv"), sep = "\t")
LV <- c("(constant guess)", "MLE", "gauss sd=0.83", "t df=2 sd=0.83",
        "t df=1 sd=0.4", "MIXTURE", "MAP: mom-median", "MAP: mom-mode", "MAP: mom-q60")
TR <- TR[est %chin% LV][, est := factor(est, levels = LV)]
TR[est == "(constant guess)", spearman := NA_real_]
knitr::kable(dcast(TR, est ~ truth_centre, value.var = "spearman")[order(est)],
  caption = "Spearman correlation by simulation setting. The constant benchmark has no rank variation.")
```

```{r trackslope}
knitr::kable(dcast(TR, est ~ truth_centre, value.var = "slope")[order(est)],
  caption = "Slope of log estimate against log truth. Assess alongside correlation and estimation error.")
```

```{r trackhit}
knitr::kable(dcast(TR, est ~ truth_centre, value.var = "within2x")[order(est)],
  caption = "Percentage within a factor of two of truth, compared with the constant benchmark. Methods are summarised over their own successful fits.")
```

At the smallest centre, shrinkage methods have correlations of about 0.40–0.42:
they improve on MLE, but still distinguish sites poorly. The moment-derived
mom-mode and mom-q60 rules perform better at larger φ, reaching correlations
around 0.77–0.78 in the setting centred at 0.4. Differences between nearby scores
should not be overinterpreted without repeated simulations.

```{r workedscatter, fig.height=6.4, fig.width=13}
sc <- fread(file.path(RC, "moment_prior_by_site.txt.gz"), sep = "\t")
sc <- sc[is.finite(phi_hat) & phi_hat > 0 & is.finite(phi_true) &
         est %chin% c("MLE", "MAP: mom-q60")]
# defined here rather than in an earlier chunk so this figure stands alone
EL <- c("MLE" = "MLE (no prior)", "MAP: mom-q60" = "MAP, moment-derived prior (q60)")
CT <- c(0.0126, 0.05, 0.2, 0.4)
sc[, rl := factor(EL[est], levels = EL)]
sc[, cl := factor(sprintf("true phi centred at %g", truth_centre),
                  levels = sprintf("true phi centred at %g", CT))]
sann <- sc[, .(sp = cor(log10(phi_true), log10(phi_hat), method = "spearman"),
               sl = coef(lm(log10(phi_hat) ~ log10(phi_true)))[2]), by = .(rl, cl)]
sann[, lab := sprintf("Spearman %.2f\nslope %.2f", sp, sl)]
ggplot(sc, aes(phi_true, phi_hat)) +
  geom_abline(slope = 1, intercept = 0, colour = "#c0392b", linewidth = 0.5) +
  geom_point(size = 0.25, alpha = 0.15, shape = 16, colour = "#1f4e9c") +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), se = FALSE,
              colour = "#f5a623", linewidth = 0.7) +
  geom_text(data = sann, aes(x = 1.5e-5, y = 1.3, label = lab), hjust = 0, vjust = 1,
            size = 2.7, colour = "grey15", lineheight = 0.95) +
  facet_grid(rl ~ cl, labeller = label_wrap_gen(width = 28)) +
  scale_x_log10() + scale_y_log10() +
  labs(x = expression(TRUE~varphi[i]~"(one point per site)"),
       y = expression(estimated~hat(varphi)[i]),
       title = "How closely does each estimate track its true dispersion?",
       subtitle = "Red: estimate equals truth. Orange: fitted trend. Top: MLE. Bottom: MAP with the moment-derived q60 prior.") +
  theme_bw(base_size = 10) +
  theme(panel.grid.minor = element_blank(), strip.text = element_text(size = 8),
        plot.subtitle = element_text(size = 7.5))
```

Shrinkage removes much of the boundary pile-up. At small φ, however, the
estimates respond only weakly to the true differences between sites. At larger
φ, they track those differences more closely. The choice of prior centre matters:
the median rule tends to set the centre below the simulated true median.

<!-- # 4. What we can report -->

<!-- The simulations support cautious interpretation of individual-site dispersion, -->
<!-- especially at low depth and low pU rates. Shrinkage can stabilise estimates, but -->
<!-- a stable estimate near the prior centre is not evidence that a site's φ is -->
<!-- precisely measured. -->

<!-- The current real-data analysis uses mom-q60 as its headline prior and retains -->
<!-- the other rules for comparison: -->

<!-- ```{r real_summary} -->
<!-- rp <- fread(file.path(RC, "real_phi_summary.tsv")) -->
<!-- knitr::kable(rp[, .(estimator, sites, mode, median)], -->
<!--   caption = "Real-data dispersion estimates. Modes are calculated on the log scale. These are summaries of estimates, not known population parameters.") -->
<!-- ``` -->

<!-- Dispersion estimation and QTL significance are separate questions. The production -->
<!-- QTL LRT uses unpenalized maximum-likelihood fits. The shrinkage results here do -->
<!-- not establish calibration for a shrinkage-based QTL test. -->
