---
title: "Beta-binomial pU-QTL: diagnosing the problematic sites (k = 3, ±5 kb, U/C sites removed)"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 3
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 11)

PSI  <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl"
PC   <- file.path(PSI, "qtl_betabinomial/prior_pi_cases")
DIAG <- file.path(PSI, "qtl_betabinomial/diagnostic_figures")

par   <- fread(file.path(PC, "case_params_patho.tsv"), sep = "\t")
bins  <- fread(file.path(PC, "prior_bins_patho.tsv"), sep = "\t")
obs   <- fread(file.path(PC, "observed_patho.tsv"), sep = "\t")
pgen  <- fread(file.path(PC, "pu_by_geno_perm_patho.tsv"), sep = "\t")
# sid contains a space, so the separator must be explicit
pl    <- fread(file.path(DIAG, "pathological_sites_w5kb_pc3_auto_noUCcalled.txt"), sep = "\t")

GRP <- c("problematic (significant in real AND permuted)", "matched control")
for (d in list(par, bins, obs)) d[, group := factor(group, levels = GRP)]
NPROB <- par[group == GRP[1], .N]
NCTRL <- par[group == GRP[2], .N]
PHI_CUT <- 0.99
ord <- par[order(group, -phi)]$sid
for (d in list(par, bins, obs)) d[, sidf := factor(sid, levels = ord)]
```

# 1. How many problematic sites, and what makes them problematic

**Definition.** A site is *problematic* if it returns an extreme p-value in the
**permuted** data as well as the real data. Permutation destroys any
genotype–phenotype relationship, so a site that stays extreme there is not
reporting association — it is reporting something about its own phenotype
distribution that the test mistakes for signal. Selection is made on the
permuted scans only, which keeps the look at the real data non-circular:

- minimum p < 10⁻¹⁰ in at least one of the 5 permutations, **and**
- real −log₁₀p > 10

**Configuration.** ±5 kb windows, k = 3 covariate PCs, chr1–22 only, phenotype
`fold5_union_auto_noUCcalled` — U/C-variant sites already excluded (1,118 of
11,856), φ-boundary sites deliberately **retained** so they can be examined here.

> **There are `r NPROB` problematic sites** out of 10,045 tested, on the
> phenotype **before** the φ filter is applied.

Which stage is meant matters, because the count differs by stage. On the
autosome-only phenotype at k = 3, ±5 kb:

| stage | problematic sites |
|---|---|
| all autosomal sites | 30 |
| minus U/C-variant sites (1,118) — **this page** | **`r NPROB`** |
| + minus the 33 φ-boundary sites | 2 |

All `r NPROB` of the sites on this page are in that 33-site φ list, so the filter
removes every one of them. The 2 that remain afterwards are *different* sites
(`chr2:89245076`, `chr2:88861447`), newly flagged once the covariate PCs are
recomputed on the reduced site set — the PCs derive from the phenotype, so
changing the site set changes every fit. See
[the covariate-count page](qtl_mapping_betabinomial_npcs_tradeoff.html).

`r NCTRL` well-behaved sites, matched on covered cell lines and pooled %pU, are
carried through every panel below for contrast.

```{r counts}
knitr::kable(pl[, .(
  site = paste0("chr", sub(" ", ":", sid)),
  `real -log10 p` = round(real, 1),
  `permuted, mean` = round(perm_mean, 1),
  `permuted, sd` = round(perm_sd, 2),
  `real minus worst permuted` = round(real_excess_over_perm_max, 2),
  `covered cell lines` = NA_integer_,
  `median depth` = median_depth_covered,
  `pooled %pU` = round(100 * pooled_pu, 1))][
    , `covered cell lines` := par$n_cl[match(pl$sid, par$sid)]][],
  caption = "The signature: the real and permuted p-values are essentially identical. For 6 of the 9 the real scan is no more extreme than the WORST permutation (negative last column), i.e. shuffling the phenotype cost nothing at all.")
```

The last column is the point. If these were genuine associations, permuting the
phenotype would destroy them and the real value would tower over the permuted
ones. Instead the real and permuted values agree to within a fraction of a log
unit, and for `r sum(pl$real_excess_over_perm_max < 0)` of the `r NPROB` sites
the real scan is *less* extreme than the worst permutation.

# 2. %pU by genotype at the lead cis-SNP

Real data plus the calibration's first two permutation orders. The permutation
moves the phenotype (and covariates) across cell lines while **genotype stays
fixed**, so any genotype–phenotype relationship is destroyed. The `%pU` values
themselves are unchanged — same multiset, different cell lines — verified: the
per-site mean `%pU` and cell-line count are identical across all three datasets.

```{r box, fig.height=8.4}
g <- pgen[!is.na(geno)]
g[, gf := factor(geno, levels = c(0, 1, 2))]
g[, dataset := factor(dataset, levels = c("real", "perm1", "perm2"))]
labg <- unique(pgen[, .(sid, group, lead_snp)])
labg[, panel := sprintf("chr%s   %s", sub(" ", ":", sid), lead_snp)]
g <- merge(g, labg[, .(sid, panel)], by = "sid")
g[, panel := factor(panel, levels = labg[order(match(sid, ord))]$panel)]

# x = dataset (3 clusters), genotype dodged WITHIN each cluster, so the 0/1/2
# trend can be compared across real vs permuted at a glance.
ggplot(g, aes(dataset, 100 * pu, fill = gf)) +
  geom_boxplot(width = 0.74, outlier.size = 0.6, colour = "grey30",
               position = position_dodge(width = 0.8, preserve = "single"),
               linewidth = 0.3) +
  scale_fill_manual(values = c(`0` = "#fde0c5", `1` = "#f0a86a", `2` = "#c8622d"),
                    name = "genotype (copies of the counted allele)") +
  facet_wrap(~ panel, ncol = 4, scales = "free_y") +
  labs(x = NULL, y = "%pU",
       title = "%pU by genotype, within real and two permutations",
       subtitle = paste("Each cluster is one dataset; the three boxes inside it are genotypes 0/1/2.",
                        "Genotype is FIXED and only the\nphenotype is shuffled, so a 0->1->2 trend inside the",
                        "'real' cluster that vanishes in the permuted clusters is real structure.",
                        "\nFree y-axes.")) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", strip.text = element_text(size = 7.4))
```

```{r lead-p}
lp <- dcast(pgen[, .(p = lead_p_this[1]), by = .(sid, group, dataset)],
            sid + group ~ dataset, value.var = "p")
setcolorder(lp, c("sid", "group", "real", "perm1", "perm2"))
knitr::kable(lp[order(match(sid, ord)), .(
  site = paste0("chr", sub(" ", ":", sid)), group = as.character(group),
  `real` = signif(real, 3), `perm 1` = signif(perm1, 3), `perm 2` = signif(perm2, 3))],
  caption = "LRT p at the SAME lead cis-SNP under each dataset. The controls collapse under permutation, as they should. Most problematic sites do not -- but note chr14:106639258 and chr6:32643315 DO collapse in perm 2, so it is not that any genotype vector works: it is that the MINIMUM over the site's cis-SNPs stays extreme, which is what the problematic-site selection uses.")
```

```{r shape}
knitr::kable(obs[, .(
  site = paste0("chr", sub(" ", ":", sid[1])),
  group = as.character(group[1]),
  `cell lines` = .N,
  `frac %pU = 0` = sprintf("%.2f", mean(pu == 0)),
  `frac %pU > 0.9` = sprintf("%.2f", mean(pu > 0.9)),
  `frac in 0.1-0.9` = sprintf("%.2f", mean(pu >= 0.1 & pu <= 0.9)),
  `median depth` = as.numeric(median(n))), by = sidf][
    order(match(sidf, ord))][, -1],
  caption = "The same thing numerically: how much of each site's %pU distribution sits at the extremes versus in the middle.")
```

# 3. Estimated φ

```{r phi-tab}
knitr::kable(par[order(match(sid, ord)), .(
  site = paste0("chr", sub(" ", ":", sid)),
  group = as.character(group),
  `cell lines` = n_cl,
  `phi (real)` = sprintf("%.6f", phi),
  `phi (perm 1)` = sprintf("%.6f", phi_perm1),
  `phi (perm 2)` = sprintf("%.6f", phi_perm2),
  `max |difference|` = signif(max_abs_phi_diff, 2),
  `alpha+beta` = signif(alpha_plus_beta, 3),
  `phi, intercept-only` = sprintf("%.6f", phi_intercept_only))],
  caption = "phi under the real data and under the calibration's first two permutation orders. The three columns are IDENTICAL -- see below.")
```

## φ is invariant under permutation, and that is the whole explanation

The permutation columns above are not approximately equal, they are **equal**:
the largest difference across all `r nrow(par)` sites and both permutations is
`r sprintf("%.1g", max(par$max_abs_phi_diff))`, i.e. optimiser noise.

This is forced, not coincidental. The calibration permutes the phenotype **and
the covariates** by the same index while leaving genotype fixed
(`y <- Cm[s, ord]; n <- Nm[s, ord]; Xcov_p <- Xcov[ord, ]`). The null
log-likelihood is a product over cell lines of terms in $(y_i, n_i, z_i)$, so
permuting all three together merely reorders the factors of a product:

$$\ell_0(\hat\beta_0, \hat\gamma, \hat\varphi) = \sum_i \log f(y_i \mid n_i, z_i)
\quad\text{is unchanged by any relabelling of } i$$

So the MLE is unchanged, $\hat\varphi$ is unchanged, and the set of 50
$(\mu_i, \alpha_i, \beta_i)$ is unchanged — only which cell line carries which
value moves. **The degenerate null fit is mathematically immune to permutation.**

### To be clear about what is and is not identical

**Identical:** the null fit — $\hat\varphi$, $\hat\ell_0$, and the 50 Betas.

**Not identical:** the LRT p-values. The *full* model contains genotype, which is
**not** permuted, so only the full-model term moves:

```{r perm-p}
knitr::kable(pl[, .(
  site = paste0("chr", sub(" ", ":", sid)),
  `real -log10 p` = round(real, 2),
  `perm 1` = round(perm1, 2), `perm 2` = round(perm2, 2),
  spread = round(pmax(real, perm1, perm2) - pmin(real, perm1, perm2), 2))],
  caption = "The p-values DO change under permutation -- just not meaningfully. A genuine QTL at -log10 p = 141 would collapse to about 2 when the phenotype is shuffled; these move by 1-5 out of 141.")
```

The mechanism follows. The statistic is
$\Lambda = 2(\hat\ell_{\text{full}} - \hat\ell_0)$, and permutation freezes
$\hat\ell_0$. It is free to change $\hat\ell_{\text{full}}$ — but the null is so
poor (for `chr3:37277733`, $\hat\ell_0 = -478$, with a prior that contributes no
regularisation at all) that **almost any** genotype vector buys a large
improvement over it. Which genotype pairs with which phenotype barely matters. So
$\Lambda$ is measuring how far a full model can climb above a broken null, not
association — and shuffling the phenotype does not repair the null.

```{r phi-fig, fig.height=4.2}
ggplot(par, aes(reorder(paste0("chr", sub(" ", ":", sid)), phi), phi, fill = group)) +
  geom_col(width = 0.68) +
  geom_hline(yintercept = PHI_CUT, linetype = "dashed", colour = "grey25") +
  annotate("text", x = 0.7, y = PHI_CUT, label = " phi = 0.99", hjust = 0, vjust = -0.6,
           size = 3.2, colour = "grey25") +
  scale_fill_manual(values = setNames(c("#e34948", "#3987e5"), GRP), name = NULL) +
  coord_flip() +
  labs(x = NULL, y = expression("fitted"~varphi),
       title = sprintf("All %d problematic sites have phi pinned to its upper boundary", NPROB),
       subtitle = sprintf("Every one sits at %.6f, the numerical cap in fit_bb_mle. Genome-wide only ~0.2%% of sites do.",
                          max(par$phi))) +
  theme_bw(base_size = 11) + theme(legend.position = "bottom")
```

**`r par[group == GRP[1] & phi > PHI_CUT, .N]` of `r NPROB` problematic sites have
φ > 0.99** — in fact every one is at 0.999999, the numerical cap. This is not a
threshold that was tuned to catch them; genome-wide only about 0.2% of sites
reach it.

Note the `intercept-only` column: at 3 of the sites φ is well below the boundary
when the covariates are dropped (0.51, 0.86, 0.92), so at those the PCs are part
of what drives the fit onto the boundary. At the other 6 it is at the cap either
way.

# 4. The fitted prior distribution of $p_i$

$p_i$ is cell line $i$'s true modification rate. It is latent — never observed —
but the fitted null model states its whole distribution, and that distribution is
what is plotted. It is obtained from the fit in three steps:

1. **The mean.** The fitted coefficients give
   $\mathrm{logit}(\mu_i) = \hat\beta_0 + z_i^\top\hat\gamma$ (no genotype term,
   since this is the null), so $\mu_i = \mathrm{logit}^{-1}(\cdot)$. Note
   $\mu_i = E[p_i]$.
2. **The shape parameters**, from that mean and the site's single fitted
   $\hat\varphi$:
   $$\alpha_i = \mu_i\frac{1-\hat\varphi}{\hat\varphi}, \qquad
     \beta_i = (1-\mu_i)\frac{1-\hat\varphi}{\hat\varphi}$$
3. **The distribution.** $p_i \sim \text{Beta}(\alpha_i, \beta_i)$, evaluated as
   exact binned mass (a density curve is unusable: at $\hat\varphi \to 1$ it
   diverges at 0 and 1):
   $$P(a < p_i \le b) = F_{\text{Beta}}(b;\alpha_i,\beta_i) - F_{\text{Beta}}(a;\alpha_i,\beta_i)$$

The same invariance applies here: refitting under permutations 1 and 2 leaves the
binned prior mass unchanged to
`r sprintf("%.1g", max(par$max_abs_prior_mass_diff))`, so the panels below are the
permuted priors as well as the real ones — there is nothing separate to plot.

**There are as many distributions as cell lines** — up to 50 per site, not one.
They share $\alpha_i+\beta_i = (1-\hat\varphi)/\hat\varphi$, but $\mu_i$ splits
that total differently for each, so each cell line gets its own Beta. The figure
therefore draws **all of them**: one thin line per cell line, with the heavy line
their average. The ticks below are the observed $y_i/n_i$, for reference.

```{r prior, fig.height=7.6}
# One Beta PER CELL LINE: alpha_i = mu_i*(1-phi)/phi, beta_i = (1-mu_i)*(1-phi)/phi.
# mu_i comes from observed.tsv (per cell line), phi from case_params (per site).
NB  <- 50
brk <- seq(0, 1, length.out = NB + 1)
mid <- (head(brk, -1) + tail(brk, -1)) / 2

cl_bins <- merge(obs[, .(sid, cell_line, mu)], par[, .(sid, group, phi)], by = "sid")
cl_bins <- cl_bins[, {
  ab <- (1 - phi) / phi
  .(bin_mid = mid, mass = diff(pbeta(brk, mu * ab, (1 - mu) * ab)))
}, by = .(sid, group, cell_line)]
avg <- cl_bins[, .(mass = mean(mass)), by = .(sid, group, bin_mid)]

labp <- par[, .(sid, panel = sprintf("chr%s   phi = %.6f   n = %d cell lines\nmu_i %.3f-%.3f",
                                     sub(" ", ":", sid), phi, n_cl, mu_min, mu_max))]
for (d in list(cl_bins, avg)) d[, panel := labp$panel[match(sid, labp$sid)]]
LV <- labp[order(match(sid, ord))]$panel
for (d in list(cl_bins, avg)) d[, panel := factor(panel, levels = LV)]
ob <- obs[, .(sid, pu)][, panel := factor(labp$panel[match(sid, labp$sid)], levels = LV)]

ggplot(cl_bins, aes(bin_mid, mass)) +
  geom_line(aes(group = cell_line), colour = "grey55", linewidth = 0.18, alpha = 0.5) +
  geom_line(data = avg, aes(colour = group), linewidth = 0.9) +
  geom_point(data = ob, aes(x = pu, y = -0.06), shape = 124, size = 2, colour = "grey15",
             alpha = 0.7) +
  scale_colour_manual(values = setNames(c("#e34948", "#3987e5"), GRP), name = NULL) +
  facet_wrap(~ panel, ncol = 4) +
  labs(x = expression(p[i]), y = "prior probability mass",
       title = "Fitted prior on the latent rate: one distribution per cell line",
       subtitle = paste("Thin grey lines: each cell line's own Beta (up to 50 per site).",
                        "Heavy line: their average.\nTicks: observed y_i/n_i.",
                        "The cell lines share alpha+beta but differ in mu_i, so the curves differ --",
                        "\nmarkedly so at chr15:48430687, where mu_i spans 0.000-0.476.")) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", strip.text = element_text(size = 7.4))
```

```{r prior-tab}
knitr::kable(par[order(match(sid, ord)), .(
  site = paste0("chr", sub(" ", ":", sid)),
  group = as.character(group),
  alpha = signif(alpha_median, 3), beta = signif(beta_median, 3),
  `mu_i range` = sprintf("%.3f - %.3f", mu_min, mu_max),
  `P(p<0.01)` = round(P_below_0.01, 3),
  `P(p>0.99)` = round(P_above_0.99, 3),
  `both tails` = round(P_below_0.01 + P_above_0.99, 3),
  `prior SD` = round(prior_sd_median, 3))],
  caption = "At the problematic sites the two tail masses sum to 1.000 -- the prior has NO interior mass -- and the prior SD is ~0.35-0.50 against a maximum of 0.5 attainable on (0,1).")
```

**The two tail masses sum to 1.000 at every problematic site.** The fitted prior
places no probability anywhere between 0.01 and 0.99: the model asserts that each
cell line's true pU rate is either essentially 0 or essentially 1, with no
intermediate stoichiometry possible.

<!-- # 5. What this says -->

<!-- Putting the three pieces together, the same story reads consistently from the -->
<!-- data side and the model side: -->

<!-- - **Data (§2)**: at these sites %pU is bimodal — cell lines sit at 0 or near 100 -->
<!--   with little in between — rather than varying smoothly with genotype. -->
<!-- - **Dispersion (§3)**: the fitter responds by driving φ to its upper boundary, -->
<!--   all `r NPROB` of `r NPROB` sites, at the numerical cap. -->
<!-- - **Prior (§4)**: at that boundary $\alpha+\beta \to 0$ and the Beta prior -->
<!--   degenerates to point masses at 0 and 1, which is the likelihood-maximising -->
<!--   description of a bimodal phenotype. -->

<!-- An LRT computed against a degenerate null is not testing what it appears to test, -->
<!-- which is why these sites stay extreme under permutation: the statistic is -->
<!-- responding to the *shape* of the phenotype distribution, which shuffling does not -->
<!-- change, rather than to any genotype–phenotype relationship. -->

<!-- This is why the scan excludes them, and why the durable fix is a **two-sided φ -->
<!-- guard inside `fit_bb_mle`** rather than a site exclusion list — see -->
<!-- [the covariate-count page](qtl_mapping_betabinomial_npcs_tradeoff.html) and -->
<!-- [the filtering strategy](qtl_mapping_betabinomial_strategy.html). -->

<!-- # Caveats -->

<!-- - **The controls are matched, not pristine.** They were chosen for similar -->
<!--   coverage and pooled %pU, and one of them (`chr15:48430687`, φ = 0.165, -->
<!--   α ≈ 0.037) is itself moderately dispersed with substantial mass near 0. It is a -->
<!--   fair contrast for coverage but not an example of an ideal fit. -->
<!-- - **`p_i` is latent.** These are the model's *prior* distributions, not estimates -->
<!--   of any cell line's rate. The posterior would add the observed counts: -->
<!--   $p_i \mid y_i \sim \text{Beta}(\alpha_i + y_i,\ \beta_i + n_i - y_i)$. -->
<!-- - **One lead SNP per site**, chosen as the minimum-LRT cis-SNP within ±5 kb at -->
<!--   MAF ≥ 0.05. At a site whose p-value is an artefact, "lead SNP" means the SNP -->
<!--   that best fits the artefact, not a candidate causal variant. -->
<!-- - **φ is on (0,1) here**, defined as $1/(\alpha+\beta+1)$ — matching `aod` and -->
<!--   VGAM's `rho`, not the `betareg` precision convention, which runs the opposite -->
<!--   way. -->

<!-- Related: [beta-binomial model](qtl_mapping_betabinomial_model.html) · -->
<!-- [covariate count and the φ boundary](qtl_mapping_betabinomial_npcs_tradeoff.html) · -->
<!-- [filtering strategy and calibration](qtl_mapping_betabinomial_strategy.html) -->
