---
title: "Beta-binomial pU-QTL: why permutation did not remove the top hits"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 3
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 11)

BASE   <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl"
DAT    <- file.path(BASE, "qtl_betabinomial/summary_page_data")
DIAG   <- file.path(BASE, "qtl_betabinomial/diagnostic_figures")
COUNTS <- file.path(BASE, "phenotype/fold5_union/counts")
PSAM   <- file.path(BASE, "genotype/plink2/pumseq_genotype.hg38.chr1-22.psam")

site26 <- fread(file.path(DAT, "site26_class.tsv"))
mf     <- fread(file.path(DAT, "misfit_by_class.tsv"))
pw     <- fread(file.path(DAT, "perm_winner.tsv"))
lev    <- fread(file.path(DAT, "leverage_demo.tsv"))
aud    <- fread(file.path(DAT, "sig_audit.tsv"))
c10    <- fread(file.path(DAT, "clean10_pearson.tsv"))
qq     <- fread(file.path(DAT, "qq_conditions.tsv.gz"))
lam    <- fread(file.path(DAT, "lambda_conditions.tsv"))
exc    <- fread(file.path(DAT, "exclusion_counts.tsv"))
patho  <- fread(file.path(DIAG, "pathological_sites.txt"))

C <- fread(file.path(COUNTS, "C_matrix.txt.gz")); N <- fread(file.path(COUNTS, "N_matrix.txt.gz"))
go <- fread(PSAM)[[2]]
Cm <- as.matrix(C[, ..go]); Nm <- as.matrix(N[, ..go]); sid_all <- paste(C$ref, C$pos)

CLS_COL <- c(`the 26 pathological sites`      = "#e34948",
             `clean significant hits`         = "#3987e5",
             `random non-significant sites`   = "#9aa4b1")
COND_LV <- c("all sites", "minus the 26 (curated list)",
             "minus U/C + concentration", "minus high-misfit sites",
             "minus U/C + concentration + misfit")
mf_all <- fread(file.path(DAT, "misfit_all_sites.tsv.gz"))
sens   <- fread(file.path(DAT, "misfit_threshold_sensitivity.tsv"))
runinf <- fread(file.path(DAT, "runinfo.tsv"))
MISFIT_CUT <- as.numeric(runinf[parameter == "misfit_cut"]$value)

# tail height per condition, used in the prose below
tailh <- qq[, .(mx = max(observed)), by = .(condition, dataset)]
th_of <- function(cond, ds = "real") round(tailh[condition == cond & dataset == ds]$mx, 1)
lam_of <- function(cond, ds = "real") round(lam[condition == cond & dataset == ds]$lambda, 3)
```

# Summary

The beta-binomial LRT scan produced a set of top hits that **permutation did not
displace**: 19 of the 20 strongest sites under a random permutation of the
phenotype were the same sites as the real top 20, and not one of them shared a
lead SNP with its real counterpart. A permutation is supposed to destroy all
genotype-phenotype signal, so this meant the sites were being selected for a
reason unrelated to genotype.

This page traces that to its cause. In brief:

1. **26 sites** produce extreme p-values even when the phenotype is shuffled.
   They were found by rerunning the whole cis scan six times — on the real data
   and under five phenotype permutations — while **keeping the site identity
   attached to every p-value**, which the original calibration script discarded.
   Of the ~11,000 tested sites, 27 reached a nominal p below 1e-10 in at least
   one permutation; 26 of those were also extreme in the real data
   (`-log10 p > 10`), and those 26 turn out to be extreme in all six datasets.
   The selection used the **permuted data only**, so nothing here is circular.
2. **14 of the 26 carry a genotyped U↔C variant on the pU base itself**, so the
   measured "%pU" is partly a read-out of genotype rather than of modification.
   A further **4** show the same three-level signature but have *no variant call
   at the base* — consistent with the same mechanism from a variant below our
   MAF 0.05 threshold, though that is inferred from the phenotype shape and not
   directly confirmed. The remaining **8** are a separate problem: sites with
   almost no modification where a *single* cell line supplies nearly all the
   converted reads.
3. Both make the phenotype impossible for a one-mean beta-binomial to fit. The
   LRT then rewards any covariate that absorbs part of that misfit — and
   **permutation preserves the misfit**, because it reorders the phenotype
   values without changing them.
4. Removing them collapses the QQ tail from `r th_of("all sites")` to
   `r th_of("minus the 26 (curated list)")`, and the *permuted* tail from ~285 to
   ~12. Notably **λ barely moves** (`r lam_of("all sites")` →
   `r lam_of("minus the 26 (curated list)")`), which is why the original
   calibration check looked acceptable while the tail was at 289. Three rules
   that generalize beyond this curated list — a U/C variant call, a
   single-cell-line cap, and a misfit cut — do better than the list itself when
   combined, but no one of them is sufficient alone (§4).
5. The remaining significant hits do **not** show this statistical signature:
   their p-values agree with an ordinary correlation test, and their permutation
   p-values sit at exactly the level chance predicts. One caveat survives that,
   though — **93 of the 311 survivors have %pU as banded as a U/C-aliasing
   site** (§2.4). Banding is produced both by aliasing *and* by a genuine
   large-effect QTL, so it cannot settle which, and for those sites the question
   is still open.

```{r headline}
knitr::kable(exc, col.names = c("", "n"),
             caption = "Scale of the problem, and what each filter removes")
```

# 1. How the 26 sites were found

## 1.1 The idea

Under permutation there is, by construction, no genotype-phenotype
relationship. So **any site that still returns an extreme p-value under
permutation is broken on its own terms** — no comparison with the real data is
needed to establish that.

This is important for what follows: the 26 sites were selected using the
**permuted data only**. Looking afterwards at the real QQ plot is therefore not
circular. Had they been picked using the real p-values, flattening the real tail
would have been guaranteed and would have proved nothing.

## 1.2 The scan

`8.calibration_site_resolved.R` repeats the genome-wide cis scan six times —
once on the real data and once under each of five phenotype permutations
(`set.seed(1000 + k)`; the phenotype *and* the covariates are reindexed, the
genotypes are not) — and, unlike the earlier calibration script, **keeps the
site identity attached to every p-value**. That is what makes it possible to
ask which sites the inflated tail comes from.

## 1.3 The two criteria

```{r criteria}
knitr::kable(data.table(
  criterion = c("minimum p < 1e-10 in at least one of the 5 permutations",
                "real -log10(p) > 10",
                "-> sites meeting both"),
  purpose = c("the site can fake an extreme p-value with no signal present",
              "drops sites that were extreme in a single permutation by chance",
              "extreme in every one of the 6 datasets (verified individually)"),
  n = c("27", "", paste0(nrow(site26)))),
  caption = "Selection criteria (9.qq_exclude_pathological.R)")
```

Exactly one of the 27 candidates was extreme in a single permutation only; the
other **26 are extreme in all six datasets**.

## 1.4 What the selected sites look like across the six datasets

Each row is one site. The red point is the real data; the blue points are the
five permutations. For a normal site the permutations would sit near zero,
far below the real point.

```{r sec1-fig, fig.height=7}
P <- paste0("perm", 1:5)
lg <- melt(patho[, c("sid", "real", P), with = FALSE], id.vars = "sid",
           variable.name = "dataset", value.name = "nlp")
lg[, is_real := dataset == "real"]
ordv <- patho[order(real)]$sid
lg[, sid := factor(sid, levels = ordv)]

ggplot(lg, aes(nlp, sid)) +
  geom_line(aes(group = sid), colour = "grey70", linewidth = 0.4) +
  geom_point(data = lg[!is_real & TRUE], aes(colour = "the 5 permutations"),
             size = 1.7, alpha = 0.85) +
  geom_point(data = lg[is_real == TRUE], aes(colour = "real data"), size = 2.6) +
  scale_colour_manual(values = c(`real data` = "#e34948",
                                 `the 5 permutations` = "#3987e5"), name = NULL) +
  labs(x = expression(-log[10]("p")~"of the site's best cis-SNP"), y = NULL,
       title = "The 26 sites: the permutations reach almost the same p-value as the real data",
       subtitle = "A working site would have its permutation points near 0. Here they land on top of the real point.") +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

```{r sec1-tab}
knitr::kable(patho[order(-real)][, .(
  site = sid,
  `real -log10p` = round(real, 1),
  `permutation mean` = round(perm_mean, 1),
  `permutation SD` = round(perm_sd, 2),
  `real - best permutation` = round(real_excess_over_perm_max, 1))],
  caption = paste("The 26 sites. The last column is how much of the real signal",
                  "survives after subtracting what permutation produces at the",
                  "same site -- for most of them, a few percent of the total."))
```

Two things stand out. The permutation SD is tiny (mostly < 1 unit on the
`-log10 p` scale), so the five permutations agree closely with each other: this
is a stable property of the site, not a fluke. And the real data exceeds the
best permutation by only **`r sprintf("%.0f%%", 100 * median(patho$real_excess_over_perm_max / patho$real))`**
of its total (median) — for five sites the real data does not exceed the
permutations at all.

# 2. What the 26 sites are

## 2.1 Classification

```{r sec2-class}
cnt <- site26[, .N, by = class][order(-N)]
knitr::kable(cnt, col.names = c("class", "n"),
             caption = "The 26 sites fall into two mechanisms")
knitr::kable(site26[, .(site = sid, `-log10p` = real_nlp,
                        `SNP at the pU base` = ifelse(snp_at_site == "", "--", snp_at_site),
                        `pooled %pU` = pooled_pu,
                        `frac at 0/50/100%` = spike3,
                        `top cell-line share` = top_share,
                        class)],
  caption = paste("Every one of the 26, classified.",
                  "'frac at 0/50/100%' is the fraction of cell lines whose %pU is within",
                  "0.08 of 0, 0.5 or 1 -- the signature of a genotype read-out.",
                  "'top cell-line share' is the largest share of the site's converted",
                  "reads coming from a single cell line."))
```

## 2.2 Mechanism A: a U↔C variant on the pU base (14 confirmed, 4 suspected)

Pseudouridine is a *modified uridine*: the assay reads out the base at that
position. If a common sequence variant changes that same base from U to C, then
an inherited C is indistinguishable from a modification event. The measured
"%pU" becomes a read-out of genotype.

This shows up as **C/T on the + strand and A/G on the − strand** — the same
U→C change on the transcript, written against whichever genome strand the
transcript came from. Genome-wide, **`r format(fread(file.path(DIAG, "uc_aliasing_sites.txt"))[, .N], big.mark = ",")`
sites** (10.0% of those tested) have such a variant.

The consequence is a phenotype with three discrete bands: homozygous
reference near 0%, heterozygous near 50%, homozygous alternate near 100%.

```{r sec2-uc, fig.height=6.5}
pu_long <- function(ids, labs) rbindlist(lapply(seq_along(ids), function(k) {
  i <- match(ids[k], sid_all); n <- Nm[i, ]; sel <- n > 0
  data.table(lab = labs[k], pu = Cm[i, sel] / n[sel], n = n[sel])
}))

# One dot per cell line: x = %pU, y = its read depth on a log axis. Both axes
# carry meaning, so the discrete %pU bands stay visible while depth is exact
# rather than binned -- and a band made only of shallow cell lines is obvious
# because it sits low.
plot_pu_depth <- function(dd, title, subtitle, ncol = 3) {
  ggplot(dd, aes(pu, pmax(n, 1))) +
    geom_vline(xintercept = c(0, 0.5, 1), colour = "grey88", linewidth = 0.4) +
    geom_point(colour = "#256abf", alpha = 0.7, size = 1.7) +
    scale_x_continuous(limits = c(-0.04, 1.04), breaks = c(0, 0.5, 1),
                       labels = c("0", "50", "100")) +
    scale_y_log10(labels = scales::comma) +
    facet_wrap(~ lab, ncol = ncol) +
    labs(x = "%pU in one cell line", y = "read depth of that cell line (log scale)",
         title = title, subtitle = subtitle) +
    theme_bw(base_size = 11) +
    theme(strip.text = element_text(size = 8), panel.grid.minor = element_blank())
}
uc_ex <- site26[class == "U/C SNP on the pU base"][order(-real_nlp)][1:6]
d <- pu_long(uc_ex$sid, sprintf("chr%s  (%s)\n-log10p=%.0f | %s reads",
                                sub(" .*", "", uc_ex$sid),
                                uc_ex$snp_at_site, uc_ex$real_nlp,
                                format(uc_ex$total_depth, big.mark = ",")))
plot_pu_depth(d,
  "Mechanism A: discrete %pU bands, because %pU is reporting genotype",
  paste("Each dot is one cell line. Bands at 0 / 50 / 100% are the three genotypes",
        "at the U/C variant, not three levels of modification, and they persist at",
        "every depth -- so they are not a low-coverage artefact.",
        "\nThe pattern is not uniformly clean: at the two strongest sites (chr6, chr2)",
        "only the 0% band is sharp and the rest is diffuse, so aliasing is plainly",
        "present but is not the whole story there."))
```

Beyond the 14 with an actual variant call, **four more show the same three-clump
shape but have no genotyped variant at the base**. The most likely explanation is
the same mechanism with a variant below our MAF 0.05 threshold or one that was
not called — but that is inferred from the phenotype shape and **not directly
confirmed**. The clearest is chr14:22551976: 11 cell lines at exactly 0.00, 26 in
a tight band near 0.40, 12 at ~1.00, with ~130 reads each. No enzymatic
modification rate behaves that way.

### Where these sites sit in the genome

```{r sec2-mhc}
s26 <- copy(site26)
s26[, chr := sub(" .*", "", sid)][, pos := as.integer(sub(".* ", "", sid))]
s26[, mhc := chr == "6" & pos > 28.5e6 & pos < 33.4e6]
knitr::kable(s26[, .(sites = .N,
                     `median -log10p` = round(median(real_nlp), 1),
                     classes = paste(sort(unique(sub(",.*", "", class))), collapse = "; ")),
                 by = .(region = ifelse(mhc, "chr6 MHC (28.5-33.4 Mb)", "rest of the genome"))],
  caption = paste("The MHC is 5 Mb, about 0.16% of the genome, but supplies",
                  sprintf("%d of the 26 (%.0f%%).", sum(s26$mhc), 100*mean(s26$mhc))))
```

**`r sum(s26$mhc)` of the 26 sit in the MHC** — every chr6 site among them, inside
a 5 Mb window that is roughly 0.16% of the genome. That is a ~`r sprintf("%.0f", (sum(s26$mhc)/nrow(s26)) / 0.0016)`-fold
enrichment, and it fits the mechanism: the MHC is the most polymorphic region in
the human genome and among the hardest to map, so it is exactly where you would
expect both uncalled U↔C variants and reference-mapping artefacts to concentrate.
It also strengthens the inference about the four sites with no variant call —
**two of those four are in the MHC**, where a variant missing from our call set is
unsurprising rather than a stretch.

These four matter out of proportion to their number. Because no variant is
called at the base, **no annotation-based filter can see them** — and they
include the second, third and fourth strongest offenders in the entire scan
(`-log10 p` of `r paste(site26[class == "3-clump shape, no genotyped SNP at the base"][order(-real_nlp)]$real_nlp[1:3], collapse = ", ")`).
Any filter that works only from variant calls will leave the QQ tail nearly
where it started; §4 shows this happening and what fixes it.

```{r sec2-nosnp, fig.height=3.2}
ns <- site26[class == "3-clump shape, no genotyped SNP at the base"][order(-real_nlp)]
d2 <- pu_long(ns$sid, sprintf("chr%s\n-log10p=%.0f | frac at 0/50/100%% = %.2f",
                              sub(" .*", "", ns$sid), ns$real_nlp, ns$spike3))
plot_pu_depth(d2,
  "The same signature, with no genotyped variant at the base",
  paste("Consistent with a U/C variant below MAF 0.05, or one that was not called.",
        "\nNote the cell lines at exactly 0% carry thousands of reads -- this is not missing data."),
  ncol = 4)
```

## 2.3 Mechanism B: one cell line supplies nearly all the converted reads (8 of 26)

These sites have essentially **no modification at all** — pooled %pU between
1% and 9%. Most cell lines have thousands of reads and a handful of converted
ones. Then one cell line reports a large %pU on modest depth.

```{r sec2-conc, fig.height=6.5}
cc <- site26[class == "one cell line dominates"][order(-real_nlp)]
d3 <- pu_long(cc$sid, sprintf("chr%s\n-log10p=%.0f | pooled %%pU=%.3f | top line=%.0f%%",
                              sub(" .*", "", cc$sid), cc$real_nlp,
                              cc$pooled_pu, 100 * cc$top_share))
plot_pu_depth(d3,
  "Mechanism B: a near-empty site plus one discordant cell line",
  paste("'top line' = share of the site's converted reads from a single cell line.",
        "\nThe deep cell lines all sit at 0%, and the one outlier to their right is",
        "usually shallower -- yet it carries the site's entire p-value (\u00a73.5)."))
```

The most extreme case, chr9:107330635, has **98% of its converted reads from
one cell line out of 50**. There is nothing to make a QTL out of.

One thing the figure invites but does not support: many of the cell lines sitting
at 0% have only a handful of reads, which looks like it might be the real
problem. It is not. Shallow cell lines are the norm throughout this dataset — at
a randomly chosen site the median cell line has **9** reads and **53%** of covered
cell lines have fewer than 10, whereas these eight sites are *better* covered than
average (median 40 reads, 25% below 10 reads). So low depth is background
conditions here, not the distinguishing feature. What singles these sites out is
the concentration of the converted reads, which is why the filter is written on
`top_share` rather than on depth.

## 2.4 For comparison: what ordinary sites look like

The panels above are only interpretable against a baseline. Same axes, same
builder, two comparison sets: significant hits that survive both filters, and
sites drawn at random from those tested but never significant.

```{r sec2-compare, fig.height=6.5}
set.seed(20260817)
surv_ids <- aud[class == "surviving" & !is.na(real)][order(real)]$sid
pick_s <- surv_ids[unique(round(seq(1, length(surv_ids), length.out = 6)))]
rand_ids <- sample(setdiff(mf_all$sid, aud$sid), 6)

lab_of <- function(ids, with_p) vapply(seq_along(ids), function(k) {
  i <- match(ids[k], sid_all)
  base <- sprintf("chr%s\npooled %%pU=%.3f | %s reads", sub(" .*", "", ids[k]),
                  sum(Cm[i, ]) / sum(Nm[i, ]), format(sum(Nm[i, ]), big.mark = ","))
  if (!with_p) return(base)
  sub("\n", sprintf("\n-log10p=%.1f | ", aud[sid == ids[k]]$real), base)
}, character(1))

ds <- pu_long(pick_s, lab_of(pick_s, TRUE))
dr <- pu_long(rand_ids, lab_of(rand_ids, FALSE))
```

### Significant hits that survive both filters

```{r sec2-surv, fig.height=4.6}
plot_pu_depth(ds,
  "Surviving significant hits, spanning the significance range",
  paste("Some (chr11, chr19, chr20) form the single gradual cloud one would hope for.",
        "\nOthers (chr5, chr15, chr16) look banded -- see the caveat below. The banded",
        "appearance alone does not separate a real large-effect QTL from an artefact."))
```

### Random sites, never significant

```{r sec2-rand, fig.height=4.6}
plot_pu_depth(dr,
  "Six sites drawn at random from those tested but never significant",
  paste("The typical site: low %pU, shallow coverage in most cell lines,",
        "no structure. Note how much shallower these are than the",
        "pathological panels -- low depth is the norm here, not the defect."))
```

### The banded pattern is not diagnostic on its own

The tempting conclusion — pathological sites have **gaps**, working sites have a
**continuum** — is only half right, and the comparison panels are what expose it.

```{r sec2-spike-overlap}
s3 <- function(x) sprintf("%.2f  (range %.2f-%.2f)", median(x, na.rm = TRUE),
                          min(x, na.rm = TRUE), max(x, na.rm = TRUE))
knitr::kable(data.table(
  `site set` = c("the 14 U/C-aliasing sites", "the 311 surviving significant hits"),
  `frac of cell lines at 0 / 50 / 100%` = c(s3(site26[uc_alias == TRUE]$spike3),
                                            s3(aud[class == "surviving"]$spike3)),
  `n with frac >= 0.80` = c(sum(site26[uc_alias == TRUE]$spike3 >= 0.80, na.rm = TRUE),
                            sum(aud[class == "surviving"]$spike3 >= 0.80, na.rm = TRUE))),
  caption = paste("The distributions overlap heavily:",
                  sum(aud[class == "surviving"]$spike3 >= 0.80, na.rm = TRUE),
                  "of the 311 survivors are as banded as the median U/C site."))
```

The reason is uncomfortable but simple. **A genuine large-effect pU-QTL produces
the same picture as aliasing.** If a variant largely abolishes modification, then
non-carriers sit high, homozygous carriers sit near 0%, heterozygotes sit between
— three bands, from real biology.

The clearest case is the strongest surviving hit, chr5:170083186:

```{r sec2-chr5}
knitr::kable(data.table(
  quantity = c("frac of cell lines at 0 / 50 / 100%",
               "|cor| of lead SNP with %pU",
               "beta-binomial LRT -log10p",
               "ordinary Pearson -log10p, same SNP",
               "permutation ceiling",
               "lead SNP distance from the pU base",
               "variant called on the pU base?"),
  value = c(sprintf("%.2f -- as banded as a U/C site", aud[sid == "5 170083186"]$spike3),
            sprintf("%.3f", c10[sid == "5 170083186"]$abs_cor),
            sprintf("%.1f", c10[sid == "5 170083186"]$lrt_nlp),
            sprintf("%.1f -- agrees", c10[sid == "5 170083186"]$pearson_nlp),
            sprintf("%.1f -- chance level", aud[sid == "5 170083186"]$perm_ceil),
            "202 bp", "no")),
  caption = "Statistically textbook; visually indistinguishable from aliasing.")
```

Every statistical check passes: permutation destroys the signal, an ordinary
correlation test agrees with the LRT, and the correlation is 0.97. Yet the %pU
pattern is banded and no variant is called at the base. Both readings fit —
either a real variant 202 bp away with a near-complete effect on modification, or
an uncalled U↔C variant at the base in linkage with it — **and the data on this
page cannot separate them.** Resolving it needs the base-level sequence, not more
statistics.

So the honest version of the rule: gaps in %pU are a **red flag requiring a
variant check at the base**, not a verdict. What §3 quantifies is why gaps break
the *test*; whether a given set of gaps is biology or artefact is a separate
question, and for a minority of the survivors it is still open.

# 3. Why permutation does not remove these sites

## 3.1 The LRT measures an improvement, not an association

The test fits the site twice — once without the SNP, once with it — and the
p-value comes from **the gap between the two fits**. It does not come from how
well the second fit works in absolute terms.

The model has only two knobs, a mean and a dispersion, and "the SNP has an
effect" can only be expressed as "the SNP shifts the mean". So the no-SNP fit
must describe all 50 cell lines with a single mean.

## 3.2 Null misfit: how much room the no-SNP fit leaves

### The idea, in one small example

Six cell lines, 100 reads each, and two toy sites. Fit each twice — once forced
to use a single rate for all six cell lines, once allowed three — then ask the
model how probable the observed data is:

```{r sec3-toy}
toy <- function(y) {
  sc  <- function(y, p) sum(dbinom(y, 100, p, log = TRUE))
  one <- sc(y, mean(y) / 100)
  three <- sum(sapply(split(y, rep(1:3, each = 2)),
                      function(g) sc(g, mean(g) / 100)))
  list(one = one, three = three)
}
A <- toy(c(96, 96, 49, 49, 3, 3)); B <- toy(c(52, 48, 51, 49, 50, 50))
knitr::kable(data.table(
  site = c("A: three clumps -- 96, 96, 49, 49, 3, 3",
           "B: one blob -- 52, 48, 51, 49, 50, 50"),
  `P(data), one rate`    = sprintf("%.0e", exp(c(A$one, B$one))),
  `P(data), three rates` = sprintf("%.0e", exp(c(A$three, B$three))),
  `misfit` = sprintf("%.1f", c(A$three - A$one, B$three - B$one))),
  caption = paste("A binomial version, for legibility. The one-rate model calls",
                  "site A's data essentially impossible; at site B, allowing",
                  "three rates buys nothing at all."))
```

**Misfit is that improvement.** At site A roughly 217 log-likelihood units of
improvement are available and the one-rate model collects none of them; at site B
there are none to collect. Nothing else is meant by "unexplained".

A beta distribution with one mean can be a single hump, a tilted hump, or a
U-shape with mass at both ends. It **cannot** place three separate spikes at 0,
0.5 and 1. So at the pathological sites the null fit is not merely imperfect,
it is structurally incapable, and the misfit is enormous.

### The definition used in this page

For cell line $i$, $y_i$ converted reads out of $n_i$ total. Beta-binomial with
mean $\mu_i$ and a single dispersion $\phi$ shared across cell lines:

$$\mu_i = \operatorname{logit}^{-1}\!\left(X_i^\top\beta\right), \qquad
  k = \frac{1-\phi}{\phi}, \qquad a_i = \mu_i k, \qquad b_i = (1-\mu_i)k$$

$$\ell(\beta,\phi) \;=\; \sum_i \left[\log\binom{n_i}{y_i}
  \;+\; \log B\!\left(y_i + a_i,\; n_i - y_i + b_i\right)
  \;-\; \log B\!\left(a_i,\, b_i\right)\right]$$

with $B$ the beta function. Each fit maximises this over $(\beta,\phi)$ using
`optim` with an analytic gradient, started from a binomial `glm.fit`; $\phi$ is
reparameterised as
$\phi = \varepsilon + (1-2\varepsilon)\operatorname{logit}^{-1}(\theta)$ with
$\varepsilon = 10^{-6}$, to keep it off the $\phi = 0$ boundary where
$k \to \infty$ degenerates the likelihood.

Two designs are fitted to the same $y, n$:

$$X_{\text{null}} = \left[\mathbf{1},\ \text{PC}_1,\ \text{PC}_2,\ \text{PC}_3\right]$$

$$X_{\text{ref}} = \left[X_{\text{null}},\ \mathbb{1}\{c_i = 2\},\ \mathbb{1}\{c_i = 3\}\right]$$

where $c_i \in \{1,2,3\}$ comes from 3-means clustering of the observed rates
$p_i = y_i/n_i$, relabelled by ascending cluster centre. Then

$$\textbf{misfit} \;=\; \hat\ell\!\left(X_{\text{ref}}\right) - \hat\ell\!\left(X_{\text{null}}\right)$$

Two properties make this the right yardstick.

**It is nested**, so misfit $\ge 0$ up to optimiser noise — and it has the same
form as the test statistic itself, which is
$2\!\left(\hat\ell_{\text{full}} - \hat\ell_{\text{null}}\right)$ with the SNP
appended in place of the indicators. The statistic and the misfit draw on the
same quantity, which is the whole argument of this section.

**It is genotype-free.** $X_{\text{ref}}$ is built from $p_i$ alone. Under a
permutation $\sigma$, all of $y$, $n$, $X_{\text{null}}$ and the cluster labels
are reindexed by that same $\sigma$, so every term in the sum is the same term in
a different order. Misfit is therefore *exactly* permutation-invariant — §3.3.

Two caveats, the second of them serious.

Three clusters is an arbitrary choice, so this is a **floor** on the unexplained
likelihood rather than all of it: a phenotype with four distinct levels would have
more than this measures.

More importantly, misfit is **undefined** wherever fewer than six cell lines are
covered, fewer than three distinct $p_i$ exist, or either fit fails to converge —
with 49 cell lines at exactly zero there is no third cluster to form. That is not
a rare edge case:

```{r sec3-na-rate}
knitr::kable(data.table(
  `site set` = c("all tested sites", "the 26 pathological sites",
                 "the 311 surviving significant hits"),
  n = c(nrow(mf_all), sum(mf_all$is26), sum(aud$class == "surviving")),
  `misfit undefined` = c(
    sprintf("%s (%.1f%%)", format(sum(is.na(mf_all$misfit)), big.mark = ","),
            100 * mean(is.na(mf_all$misfit))),
    sprintf("%d", sum(is.na(mf_all[is26 == TRUE]$misfit))),
    sprintf("%d (%.0f%%)",
            sum(is.na(mf_all[sid %chin% aud[class == "surviving"]$sid]$misfit)),
            100 * mean(is.na(mf_all[sid %chin% aud[class == "surviving"]$sid]$misfit))))),
  caption = "Where misfit cannot be computed at all")
```

**Misfit is uncomputable for over 40% of tested sites.** A filter blind to two
fifths of the data is a partial instrument, not a solution — and the blindness is
not random: it concentrates exactly where the phenotype is most degenerate, which
is where the artefacts live. Four of the five unmeasurable sites among the 26 are
concentration sites. This is the main reason misfit cannot replace the
concentration cap, and it should temper how §4.3's combined-filter numbers are
read.

```{r sec3-misfit, fig.height=4.6}
mfp <- mf[is.finite(misfit)]
mfp[, class := factor(class, levels = names(CLS_COL))]
knitr::kable(mfp[, .(sites = .N,
                     `median misfit` = round(median(misfit), 1),
                     `25th pct` = round(quantile(misfit, .25), 1),
                     `75th pct` = round(quantile(misfit, .75), 1),
                     `max` = round(max(misfit), 1)), by = class],
  caption = paste("Null misfit (log-likelihood units left unexplained by the",
                  "one-mean model), computed without using genotype."))
ggplot(mfp, aes(class, pmax(misfit, 0.05), fill = class)) +
  geom_boxplot(width = 0.5, alpha = 0.85, outlier.size = 0.9) +
  scale_y_log10() + scale_fill_manual(values = CLS_COL, guide = "none") +
  labs(x = NULL, y = "null misfit, log-likelihood units (log scale)",
       title = "The pathological sites leave orders of magnitude more log-likelihood unexplained",
       subtitle = paste("Computed from the phenotype values only -- no genotype enters,",
                        "so permutation cannot change it.",
                        sprintf("\nMisfit is undefined at %d of the 26 (too few distinct %%pU levels to fit a 3-group reference), so %d are shown.",
                                26 - nrow(mfp[class == "the 26 pathological sites"]),
                                nrow(mfp[class == "the 26 pathological sites"])))) +
  theme_bw(base_size = 12)
```

## 3.3 Permutation preserves the misfit exactly

Permutation reorders which cell line holds which %pU value. **The set of values
is unchanged.** Since misfit is a property of that set and not of its
alignment with genotype, permutation leaves it untouched — the whole pot of
unexplained log-likelihood is still on the table after shuffling.

That is the entire failure. The LRT depends on two things:

- how much unexplained log-likelihood is available, and
- how well genotype happens to line up with the phenotype.

Permutation destroys only the second. When the first is large enough, the
p-value barely moves.

## 3.4 The consequence: an association no ordinary test would report

If the p-value came from association, then an ordinary Pearson correlation test
on the same lead SNP should broadly agree with it. Pearson has no dispersion
parameter and no depth weighting, so **misfit cannot inflate it** — which makes
it a useful independent yardstick.

Below, for each site, the LRT `-log10 p` **under permutation 1** against the
Pearson `-log10 p` for the very same SNP and the same data.

```{r sec3-gap, fig.height=5}
pw[, class := factor(class, levels = names(CLS_COL))]
# NB: geom_abline() in a log-scaled panel is drawn in TRANSFORMED space, i.e. it
# renders y = 10^x rather than y = x. The identity line must be supplied as
# explicit data so that it curves correctly under the log transform.
# The line spans the x-axis data range only -- extending it to the y range would
# stretch the x-axis to ~290 and squash every point into a sliver.
xr   <- range(pw$pearson_nlp, na.rm = TRUE)
iden <- data.table(x = seq(max(xr[1], 0.15), xr[2], length.out = 200))
iden[, y := x]

ggplot(pw, aes(pearson_nlp, lrt_nlp, colour = class)) +
  geom_line(data = iden, aes(x, y), inherit.aes = FALSE,
            linetype = "dashed", colour = "grey45") +
  geom_point(size = 2, alpha = 0.8) +
  scale_colour_manual(values = CLS_COL, name = NULL) +
  scale_y_log10() +
  labs(x = expression(-log[10]("p")~"from an ordinary Pearson correlation test"),
       y = expression(-log[10]("p")~"from the beta-binomial LRT (log scale)"),
       title = "Under permutation, only the pathological sites disagree with an ordinary correlation test",
       subtitle = paste("Dashed line = the two tests give the same answer. Points far above it are",
                        "p-values the LRT manufactures and Pearson does not.",
                        "\nEvery point is permuted data, so the true association is zero throughout.")) +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")

knitr::kable(pw[, .(sites = .N,
                    `median LRT -log10p`     = round(median(lrt_nlp), 1),
                    `median Pearson -log10p` = round(median(pearson_nlp), 1),
                    `median gap`             = round(median(gap), 1),
                    `median |cor| of the LRT winner` = round(median(abs_cor), 2)),
                by = class],
  caption = paste("Under permutation 1: what the LRT reports versus what an",
                  "ordinary correlation test reports for the same SNP."))
```

Two worked examples make the size of the gap concrete.

```{r sec3-examples}
ex_ids <- c("19 46609391", "14 106360212")
ee <- pw[sid %chin% ex_ids]
knitr::kable(data.table(
  site = ee$sid,
  mechanism = ifelse(ee$sid == "19 46609391", "A: U/C variant on the base",
                                              "B: one cell line dominates"),
  `cis-SNPs tested` = ee$n_cis,
  `null misfit` = round(ee$misfit, 0),
  `|cor| of the LRT winner` = round(ee$abs_cor, 3),
  `best |cor| available at the site` = round(ee$best_abs_cor, 3),
  `Pearson p for that SNP` = signif(ee$pearson_p, 2),
  `beta-binomial LRT p` = sprintf("%.0e", 10^(-ee$lrt_nlp))),
  caption = "Both rows are permuted data, so the true association is zero by construction.")
```

At chr19:46609391 the winning SNP correlates with the permuted phenotype at
|cor| = `r sprintf("%.2f", pw[sid=="19 46609391"]$abs_cor)`, which an ordinary
test rates p = `r signif(pw[sid=="19 46609391"]$pearson_p, 2)` — nothing. At
chr14:106360212 it is |cor| = `r sprintf("%.3f", pw[sid=="14 106360212"]$abs_cor)`,
p = `r signif(pw[sid=="14 106360212"]$pearson_p, 2)` — a shapeless cloud. The
LRT converts both into p-values of astronomical size, because there is a large
pot of misfit for even a negligible correlation to draw on.

Note also that neither site's *best available* correlation is large. With a few
dozen cis-SNPs to choose from, one always lands slightly closer to the clumps
than the others by luck — which is why a different SNP wins in every
permutation, and why the top hits share sites but never lead SNPs.

## 3.5 For mechanism B, one cell line carries the whole p-value

For the concentration sites the point can be made directly: drop the single
most extreme cell line and rerun the site unchanged.

```{r sec3-lev}
knitr::kable(lev[order(-orders_lost, na.last = FALSE)][, .(
  site = sid,
  `top cell-line share` = top_share,
  `converted reads it holds` = format(dropped_conv, big.mark = ","),
  `its total reads` = format(dropped_depth, big.mark = ","),
  `-log10p with all cell lines` = round(nlp_all, 1),
  `-log10p after dropping it` = ifelse(is.na(nlp_drop1), "no fit",
                                       sprintf("%.1f", nlp_drop1)),
  `orders lost` = ifelse(is.na(orders_lost), "--", sprintf("%.1f", orders_lost)))],
  caption = paste("Under permutation 1: removing the single cell line that holds",
                  "the most converted reads. 'no fit' means the site could not be",
                  "fitted at all once that cell line was gone."))
```

Why Pearson misses this while the likelihood does not: a cell line with 6,000
reads at %pU 0.0005 and a cell line with 20 reads at 0.05 count **equally** in
a correlation, but the likelihood weights the first thousands of times more
heavily. A SNP can therefore split the deep near-zero cell lines from the rest
— worth an enormous amount of likelihood — while barely moving the correlation.
(For the record, the earlier guess that rare-allele SNPs isolating one or two
cell lines were responsible is **wrong**: with MAF ≥ 0.05 across 50 cell lines
the minor allele is always carried by at least 5, and none of the cis-SNPs at
these sites had fewer than 4 carriers.)

# 4. Calibration after removing these sites

## 4.1 A curated list is not a filter

Dropping the 26 sites by name would fix this page's QQ plot and nothing else:
the next dataset will have a different 26. So each exclusion below is stated as
a **rule computable for any site**, and the curated list is kept only as the
reference for what a perfect filter would achieve.

Three rules are available:

```{r sec4-rules}
knitr::kable(data.table(
  rule = c("U/C-aliasing", "concentration", "high misfit"),
  definition = c("a genotyped C/T (+ strand) or A/G (- strand) variant on the pU base",
                 sprintf("one cell line supplies >= %.0f%% of the site's converted reads",
                         100 * as.numeric(runinf[parameter == "conc_cut"]$value)),
                 sprintf("null misfit > %g log-likelihood units", MISFIT_CUT)),
  `needs genotype?` = c("yes, a variant call at the base", "no", "no"),
  `sites removed` = format(c(as.integer(runinf[parameter == "n_uc_alias"]$value),
                             as.integer(runinf[parameter == "n_conc_bad"]$value),
                             as.integer(runinf[parameter == "n_misfit_bad"]$value)),
                           big.mark = ",")),
  caption = "The three exclusion rules, each computable without knowing which sites are broken")
```

The misfit rule matters because **the first two rules miss four of the 26** —
the sites with the three-clump shape but no genotyped variant at the base
(§2.2). Those four include the second, third and fourth worst offenders in the
whole scan, so an annotation-only filter leaves the tail almost untouched.

The misfit threshold was chosen as the smallest round value that catches every
one of the 26 for which misfit is measurable, then reported here rather than
tuned against the QQ plot:

```{r sec4-sens}
knitr::kable(sens[, .(
  `misfit threshold` = threshold,
  `sites removed` = format(sites_removed, big.mark = ","),
  `% of tested` = sprintf("%.1f%%", pct_of_tested),
  `of the 26, caught` = sprintf("%d of %d", of_the_26_caught, of_the_26_measurable))],
  caption = sprintf("Threshold sensitivity. The value used is %g.", MISFIT_CUT))
```

## 4.2 QQ plots under each condition

`all sites` is the calibration as originally run. Recall that the 26 were chosen
using the permuted data only, so the real panels are not circular.

```{r sec4-qq, fig.height=12}
qq[, condition := factor(condition, levels = COND_LV)]
qq[, dataset := factor(dataset, levels = c("real", paste0("perm", 1:5)))]
COLS <- c(real = "#e34948", perm1 = "#9ec5f4", perm2 = "#6da7ec",
          perm3 = "#3987e5", perm4 = "#256abf", perm5 = "#0d366b")
lab <- lam[, .(t = paste(sprintf("%s=%.2f", dataset, lambda), collapse = "  ")),
           by = condition]
ggplot(qq, aes(expected, observed, colour = dataset)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45") +
  geom_point(size = 0.6, alpha = 0.75) +
  scale_colour_manual(values = COLS, name = NULL) +
  guides(colour = guide_legend(override.aes = list(size = 2.6, alpha = 1), nrow = 1)) +
  facet_wrap(~ condition, ncol = 2, scales = "free_y") +
  labs(x = expression(-log[10]("expected p")), y = expression(-log[10]("observed p")),
       title = sprintf("Beta-binomial LRT calibration under %d exclusion conditions",
                       length(COND_LV)),
       subtitle = paste("Note the free y-axis -- the tail height is what changes, not the bulk.",
                        "\nThe permutation series (blues) should hug the dashed line; where they do",
                        "not, the test is manufacturing signal from shuffled data.")) +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

## 4.3 What actually changed

```{r sec4-lam}
ORD <- c("real", paste0("perm", 1:5))
lw <- dcast(lam, dataset ~ condition, value.var = "lambda")
setcolorder(lw, c("dataset", COND_LV))
lw <- lw[order(factor(dataset, levels = ORD))]
knitr::kable(lw, digits = 3,
  caption = "lambda -- the median-based genomic inflation factor")

tw <- dcast(tailh, dataset ~ condition, value.var = "mx")
setcolorder(tw, c("dataset", COND_LV))
tw <- tw[order(factor(dataset, levels = ORD))]
knitr::kable(tw, digits = 1,
  caption = "Height of the QQ tail: the largest observed -log10(p)")
```

The two tables measure different things, and the key result of this section is
that **the rules fix different things, and no single one of them is enough.**

```{r sec4-compare}
knitr::kable(data.table(
  condition = COND_LV,
  `real lambda` = vapply(COND_LV, lam_of, numeric(1)),
  `real tail` = vapply(COND_LV, th_of, numeric(1)),
  `worst permuted tail` = vapply(COND_LV, function(c)
    round(max(tailh[condition == c & dataset != "real"]$mx), 1), numeric(1)),
  `p-values retained` = vapply(COND_LV, function(c)
    format(lam[condition == c & dataset == "real"]$n_pvalues, big.mark = ","),
    character(1))),
  row.names = FALSE,
  caption = paste("The permuted tail is the diagnostic that matters: with the",
                  "phenotype shuffled it should be near 0, and anything above",
                  "~3 is signal the test is manufacturing."))
```

Reading down that table:

**λ and the tail are almost independent.** Removing the 26 collapses the tail
from `r th_of("all sites")` to `r th_of("minus the 26 (curated list)")` — and
the permuted tail from ~285 to `r round(max(tailh[condition == "minus the 26 (curated list)" & tailh$dataset != "real"]$mx), 1)` —
while barely touching λ (`r lam_of("all sites")` →
`r lam_of("minus the 26 (curated list)")`). λ comes from the *median* test
statistic, and 26 sites out of ~11,000 cannot move a median. Conversely the U/C
plus concentration rules pull λ down to
`r lam_of("minus U/C + concentration")` — much the larger effect on λ — yet
leave the tail at `r th_of("minus U/C + concentration")`, essentially unchanged.
**λ was never the right diagnostic for this problem**, which is why the original
calibration check reported reassuring-looking numbers while the tail was at 289.

**Misfit alone is not sufficient either.** It cuts the tail to
`r th_of("minus high-misfit sites")`, better than the annotation rules but well
short of the curated list. The reason is specific and limits the rule generally:
misfit is **undefined at 5 of the 26**, and indeed at
`r sprintf("%.0f%%", 100 * mean(is.na(mf_all$misfit)))` of all tested sites
(§3.2), because a site with 49 cell lines at exactly zero has no three distinct
%pU levels for the reference model to use. Four of those five are concentration
sites, so misfit and the concentration cap are complementary by construction
rather than redundant — but it also means the misfit rule is silent on a large
fraction of the genome, and silent precisely where phenotypes are most
degenerate.

**The three rules together beat the curated list on these metrics.** Applying all three gives
λ = `r lam_of("minus U/C + concentration + misfit")`, a real tail of
`r th_of("minus U/C + concentration + misfit")`, and a permuted tail of
`r round(max(tailh[condition == "minus U/C + concentration + misfit" & tailh$dataset != "real"]$mx), 1)` —
better on every measure than naming the 26 by hand, and computable on data where
nobody has told you which sites are broken.

**What it costs.** The combined filter removes
`r format(as.integer(exc[grepl("all three filters combined", filter)]$n), big.mark = ",")`
of `r format(as.integer(exc[grepl("sites tested", filter)]$n), big.mark = ",")`
tested sites
(`r sprintf("%.0f%%", 100 * as.integer(exc[grepl("all three filters combined", filter)]$n) / as.integer(exc[grepl("sites tested", filter)]$n))`).
That is not a small price, and the misfit threshold is the loose part of it: the
26 sites where misfit is measurable range from
`r round(min(mf_all[is26 & is.finite(misfit)]$misfit), 1)` to
`r round(max(mf_all[is26 & is.finite(misfit)]$misfit), 1)`, while ordinary
significant hits reach `r round(max(mf[class == "clean significant hits" & is.finite(mf$misfit)]$misfit), 1)`,
so the two distributions overlap and a cut of `r MISFIT_CUT` necessarily removes
some legitimate sites — `r exc[grepl("survivors also flagged by misfit", filter)]$n`
of the `r exc[grepl("surviving the U/C", filter)]$n` hits that pass the other two
rules. Whether that trade is worth making is a judgement call, not something
these numbers settle.

A residual permuted excess remains even after all exclusions, so this is a large
improvement rather than a complete fix — see [Open questions](#open-questions).

# 5. Are the other significant hits affected?

## 5.1 Two independent checks

The concern is that misfit might explain a large share of the ~1,400
significant hits, not just 26 of them. Two checks, applied to the hits that are
*not* among the 26:

**Check 1 — a second opinion.** Compare the LRT against an ordinary Pearson
correlation test on the same lead SNP. Agreement means the signal is real;
a large gap means only the fancy machinery sees it.

**Check 2 — the site's own fake-signal ceiling.** For each site, take the best
`-log10 p` it reached in *any* of the five permutations. That is how large a
p-value the site can manufacture with no signal present. Compare the real
result against it.

## 5.2 Check 2, applied to every significant hit

```{r sec5-ceiling}
aud[, class := factor(class, levels = c("one of the 26", "U/C-aliasing (not in the 26)",
                                        "concentration-pathological", "surviving"))]
knitr::kable(aud[, .(hits = .N,
  `median real -log10p`   = round(median(real, na.rm = TRUE), 1),
  `median perm ceiling`   = round(median(perm_ceil, na.rm = TRUE), 1),
  `median excess`         = round(median(excess, na.rm = TRUE), 1),
  `excess < 2`            = sum(excess < 2, na.rm = TRUE),
  `excess > 5`            = sum(excess > 5, na.rm = TRUE)), by = class],
  caption = paste("For every q<0.05 hit: the real signal against the largest",
                  "p-value the same site reached under permutation."))
```

The permutation ceiling for the non-pathological hits sits at about **2.2**,
and that number is exactly what chance predicts: with roughly 40 cis-SNPs per
site across 5 permutations, the best of ~200 uniform p-values falls near
1/200, i.e. `-log10 p` ≈ 2.3. **These sites have no capacity to fake signal at
all.** The 26, by contrast, reach ~122.

```{r sec5-ceilfig, fig.height=4.6}
af <- aud[!is.na(perm_ceil)]
ggplot(af, aes(perm_ceil, fill = class)) +
  geom_histogram(bins = 60) +
  geom_vline(xintercept = 2.3, linetype = "dashed", colour = "grey30") +
  annotate("text", x = 2.3, y = Inf, label = "  chance expectation\n  for ~200 tests",
           hjust = 0, vjust = 1.4, size = 3, colour = "grey30") +
  scale_fill_manual(values = c(`one of the 26` = "#e34948",
                               `U/C-aliasing (not in the 26)` = "#f0a24f",
                               `concentration-pathological` = "#9aa4b1",
                               `surviving` = "#3987e5"), name = NULL) +
  scale_x_log10() +
  labs(x = "the site's permutation ceiling, -log10p (log scale)", y = "hits",
       title = "How large a p-value each significant site can fake under permutation") +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

## 5.3 Check 1, on hits spanning the significance range

Ten surviving hits, chosen to span the whole range from the strongest to the
weakest, each tested both ways on the **real** data.

```{r sec5-c10}
knitr::kable(c10[, .(site = sid, lead_SNP = lead,
  `LRT -log10p` = round(lrt_nlp, 1),
  `Pearson -log10p` = round(pearson_nlp, 1),
  gap = round(gap, 1),
  `|cor|` = round(abs_cor, 2),
  `perm ceiling` = round(perm_ceil, 1),
  `total depth` = format(round(total_depth), big.mark = ","),
  `pooled %pU` = round(pooled_pu, 3))],
  caption = "Ten surviving hits: the beta-binomial LRT against an ordinary correlation test")
```

The gap is small in both directions — at the strongest site the Pearson test is
*more* significant than the LRT — and the correlations are large and visible.
Against gaps of ~241 and ~34 at the two pathological examples, these behave
like ordinary associations tested by a slightly better-powered method, which is
what the beta-binomial is supposed to be.

## 5.4 What the checks did turn up

Check 2 screens every site cheaply, and it found a class the original
26-site criteria missed: hits with the **mechanism-B concentration signature**
that never happened to produce a p-value below 1e-10 under permutation.

```{r sec5-conc}
cn <- aud[class == "concentration-pathological"][order(-top_share)]
knitr::kable(cn[, .(site = sid,
  `top cell-line share` = round(top_share, 2),
  `pooled %pU` = round(pooled_pu, 3),
  `real -log10p` = round(real, 1),
  `perm ceiling` = round(perm_ceil, 1),
  excess = round(excess, 1),
  qvalue = signif(qvalue, 2))],
  caption = paste("Significant hits where one cell line supplies at least half",
                  "the converted reads. These are not misfit artifacts -- they are",
                  "single-observation artifacts, and they need a separate filter."))
```

The worst has a single cell line supplying **100%** of the site's converted
reads. At least one has a permutation ceiling *above* its real result (negative
excess) and still passes q < 0.05. Several cluster inside a narrow window on
chr14, which suggests a regional mapping artifact rather than independent
events.

## 5.5 Where the discovery count lands

```{r sec5-funnel}
knitr::kable(aud[, .(hits = .N,
  `median -log10p` = round(median(real, na.rm = TRUE), 1),
  `share of all hits` = sprintf("%.1f%%", 100 * .N / nrow(aud))), by = class],
  caption = "The q<0.05 hit list, partitioned")
```

Only **`r sum(aud$class == "one of the 26")`** of the significant hits are among
the 26 — the per-site permutation `perm_p` did filter out most of them, so it
offers partial protection, not none. The large category is
**U/C-aliasing sites that are *not* in the 26**
(`r sum(aud$class == "U/C-aliasing (not in the 26)")` hits), and these deserve a
careful distinction:

> Their permutation ceiling is ~2.3, i.e. chance level, so they are **not a
> calibration failure**. Statistically the associations are sound: SNPs in
> linkage with the site variant genuinely do predict the measured phenotype.
> The problem is **interpretive** — the phenotype is a genotype read-out rather
> than a modification rate, so a real association is being reported as a
> pU-QTL. Same exclusion, entirely different reason. This is also why `perm_p`
> cannot protect against them: there is nothing statistically wrong to detect.

# Open questions {#open-questions}

- **Residual permuted inflation.** After removing both the 26 and the full
  U/C-aliasing set, the fraction of permuted p-values below 0.05 remains ~0.058
  against an expected 0.050. Small, but a third effect that is not yet
  accounted for.
- **The exclusion has not been re-run.** The counts above partition the
  *existing* result list. Rebuilding the phenotype matrix without the excluded
  sites and rescanning would also change the multiple-testing correction, so
  the final number will not be exactly the surviving count here.
- **Assay direction.** The U/C interpretation rests on pseudouridine being read
  at the same base that the variant changes. Worth confirming with whoever ran
  the PUM-seq protocol.
- **The banded survivors (§2.4).** `r sum(aud[class == "surviving"]$spike3 >= 0.80, na.rm = TRUE)`
  of the 311 look like aliasing but pass every statistical check, chr5:170083186
  most conspicuously (|cor| 0.97, Pearson agrees, permutation at chance, lead SNP
  202 bp away, no variant called at the base). Distinguishing a real
  near-complete-effect QTL from an uncalled U↔C variant needs base-level sequence
  at those positions — reads piled at the pU base, or a deeper variant call set —
  not further statistics. Until then the surviving count should be read as an
  upper bound.
- **Misfit covers only ~59% of sites.** The rule is undefined wherever the
  phenotype is too degenerate to fit a three-level reference, which is
  disproportionately where artefacts are (§3.2). Any production filter needs
  something that works on those sites too.
- **Which filter to adopt.** All three rules in §4.1 are computable without
  knowing which sites are broken, but they are not interchangeable: the
  annotation-based ones cannot see the four sites with no variant call at the
  base, and misfit alone does not tell you *why* a site is broken. Using the
  U/C annotation for interpretation and misfit for calibration is probably the
  right split, but the threshold deserves a second look — it was set to catch
  all of the 26, which is a small calibration set.

# Reproducing this page

```{r prov}
knitr::kable(fread(file.path(DAT, "runinfo.tsv")),
             caption = "Parameters (10.summary_page_data.R)")
```

```{r prov2}
knitr::kable(data.table(
  step = c("site-resolved calibration scan (6 datasets, site identity retained)",
           "selection of the pathological sites",
           "U/C-aliasing site list",
           "all precomputed tables behind this page",
           "this page"),
  script = c("4.qtl_mapping_betabinomial/8.calibration_site_resolved.R",
             "4.qtl_mapping_betabinomial/9.qq_exclude_pathological.R",
             "4.qtl_mapping_betabinomial/9.qq_exclude_pathological.R (+ .traw allele check)",
             "4.qtl_mapping_betabinomial/10.summary_page_data.R",
             "analysis/qtl_mapping_betabinomial_pathology_summary.Rmd"),
  submitted_via = c("run_calibration_siteres.sbatch", "-- (light)", "-- (light)",
                    "run_summary_page_data.sbatch", "run_render_rmd.sbatch")),
  caption = "Provenance. The heavy steps require a compute node; a login node SIGKILLs them.")
```

Related pages: [top-hit diagnostic](qtl_mapping_betabinomial_diagnostic.html) ·
[why the same top hits](qtl_mapping_betabinomial_why_same_tophits.html) ·
[beta-binomial scan summary](qtl_mapping_summary_betabinomial.html)
