---
title: "Beta-binomial pU-QTL: the current filtering strategy and its calibration"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 3
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 11)

PSI <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl"
DAT <- file.path(PSI, "qtl_betabinomial/strategy_page_data_auto")

qq   <- fread(file.path(DAT, "qq_all_runs.tsv.gz"))
lam  <- fread(file.path(DAT, "lambda_all_runs.tsv"))
phid <- fread(file.path(DAT, "phi_distribution.tsv"))
phis <- fread(file.path(DAT, "phi_per_site.tsv.gz"))
iter <- fread(file.path(DAT, "phi_iteration.tsv"))
excl <- fread(file.path(DAT, "exclusion_steps.tsv"))
info <- fread(file.path(DAT, "runinfo.tsv"))
gv <- function(k) info[parameter == k]$value

RUN_LV <- c("1. all sites", "2. minus U/C-variant sites",
            "3. minus U/C + phi-boundary sites")
ORD <- c("real", paste0("perm", 1:5))
qq[,  run := factor(run, levels = RUN_LV)][, dataset := factor(dataset, levels = ORD)]
lam[, run := factor(run, levels = RUN_LV)][, dataset := factor(dataset, levels = ORD)]
COLS <- c(real = "#e34948", perm1 = "#9ec5f4", perm2 = "#6da7ec",
          perm3 = "#3987e5", perm4 = "#256abf", perm5 = "#0d366b")
lv <- function(r, d, col) lam[run == r & dataset == d][[col]]
FINAL <- RUN_LV[3]
```

Everything on this page is restricted to **chr1–22**, with the covariate PCs
recomputed on chr1–22 only. Non-autosomal sites never contributed a p-value in any
case — the genotype panel is `pumseq_genotype.hg38.chr1-22` — but they were
previously feeding the covariate PCs, and through them every null fit.

# Why filter at all: the fitted overdispersion

φ is the beta-binomial's overdispersion parameter, defined here — as in `aod` and
as VGAM's `rho` — by

$$\varphi = \frac{1}{a + b + 1}, \qquad
  \mathrm{Var}[y] = n\,\mu(1-\mu)\bigl[1 + (n-1)\varphi\bigr]$$

so it is the intracluster correlation and is bounded to (0, 1) by construction.
φ → 0 means `a+b → ∞` and the model reduces to a binomial; **φ → 1 means
`a+b → 0`, where the beta prior collapses onto point masses at 0 and 1** — the
model asserting that every cell line's true rate is exactly 0 or exactly 1. A fit
that lands there has not really been fitted.

```{r phi-dist, fig.height=4.2}
p <- copy(phis)[is.finite(phi)]
ggplot(p, aes(pmax(phi, 1e-6), fill = problematic)) +
  geom_histogram(bins = 60, position = "identity", alpha = 0.75, colour = NA) +
  geom_vline(xintercept = 0.99, linetype = "dashed", colour = "grey25") +
  annotate("text", x = 0.99, y = Inf, hjust = 1.05, vjust = 1.6, size = 3.2,
           label = "phi = 0.99 ") +
  scale_x_log10() + scale_y_sqrt() +
  scale_fill_manual(values = c(`FALSE` = "#3987e5", `TRUE` = "#e34948"),
                    labels = c("ordinary site", "was problematic under permutation"),
                    name = NULL) +
  labs(x = "fitted overdispersion phi (log scale)", y = "sites (sqrt scale)",
       title = "phi spans its whole range; the problematic sites sit at the very top of it",
       subtitle = paste("phi is broadly and multi-modally distributed from 1e-6 to 1 -- there is NO clean gap.",
                        "\nWhat is clean is the discrimination: every site still problematic after filter 1",
                        "has phi > 0.99, and only 0.4% of all sites do.")) +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

That right-hand pile-up is the motivation for the second filter below: **every site
that was still problematic after the U/C step sits in it.** Note the observed
maximum is 1 − 10⁻⁶, the numerical cap in `fit_bb_mle`, so those optimisations
were driving `a+b` to zero and were stopped by the guard rather than converging.

# The strategy in one table

Two filters are applied to the phenotype before mapping. Neither is a curated
list of bad sites: both are rules computable from a single site's own data, so
they apply unchanged to any future dataset.

```{r strategy}
knitr::kable(data.table(
  step = c("start", "1. U/C-variant sites", "2. phi-boundary fits", "final"),
  what = c("all called pU sites",
           "a U/C variant is CALLED on the pU base, so %pU partly reports genotype",
           sprintf("the fitted overdispersion phi lands on its boundary (phi > 0.99), i.e. the fit is degenerate. Applied ITERATIVELY -- %s rounds flagged sites (%s)",
                   gv("phi_rounds_that_flagged"), gv("phi_rounds_detail")),
           "used for all QTL mapping"),
  removed = c("--", format(as.integer(gv("excluded_UC")), big.mark = ","),
              gv("excluded_phi"), "--"),
  sites_left = c(format(as.integer(gv("sites_start")), big.mark = ","),
                 format(as.integer(gv("sites_start")) - as.integer(gv("excluded_UC")), big.mark = ","),
                 format(as.integer(gv("sites_final")), big.mark = ","),
                 format(as.integer(gv("sites_final")), big.mark = ","))),
  caption = sprintf("Phenotype: %s. Covariate PCs are recomputed on the retained sites at each step.", gv("final_phenotype")))
```

And the effect on calibration, which is the point of the whole exercise:

```{r headline}
knitr::kable(data.table(
  run = RUN_LV,
  `real lambda` = vapply(RUN_LV, lv, numeric(1), d = "real", col = "lambda"),
  `permutation lambda (range)` = vapply(RUN_LV, function(r)
    sprintf("%.3f - %.3f", min(lam[run == r & dataset != "real"]$lambda),
            max(lam[run == r & dataset != "real"]$lambda)), character(1)),
  `real tail` = round(vapply(RUN_LV, lv, numeric(1), d = "real", col = "max_nlp"), 1),
  `worst permuted tail` = round(vapply(RUN_LV, function(r)
    max(lam[run == r & dataset != "real"]$max_nlp), numeric(1)), 1),
  `problematic sites` = vapply(RUN_LV, function(r)
    lam[run == r][1]$n_problematic, integer(1))),
  row.names = FALSE, digits = 3,
  caption = "Under permutation there is no genotype-phenotype relationship, so those rows are the diagnostic: lambda should be 1 and the tail near 0.")
```

**`r lam[run == FINAL][1]$n_problematic` sites remain problematic** after both
filters, down from `r lam[run == RUN_LV[1]][1]$n_problematic`. Permutation λ sits
at `r sprintf("%.2f-%.2f", min(lam[run == FINAL & dataset != "real"]$lambda), max(lam[run == FINAL & dataset != "real"]$lambda))`,
which is what correct calibration looks like, and the real-data λ of
**`r sprintf("%.3f", lv(FINAL, "real", "lambda"))`** is the mild inflation
expected from genuine cis signal — not the
`r sprintf("%.3f", lv(RUN_LV[1], "real", "lambda"))` seen before any filtering,
which was artefact.

# 1. Calibration QQ plots

Each panel is one filtering stage. **Red is the real scan; the five blues are
independent phenotype permutations.** The permutation series carries the
information: shuffling the phenotype destroys any genotype–phenotype
relationship, so those points must lie on the dashed line. Where they rise above
it, the test is manufacturing signal out of nothing.

```{r qq-facets, fig.height=4.8}
ggplot(qq, aes(expected, observed, colour = dataset)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45") +
  geom_point(size = 0.55, alpha = 0.75) +
  scale_colour_manual(values = COLS, name = NULL) +
  guides(colour = guide_legend(override.aes = list(size = 2.6, alpha = 1), nrow = 1)) +
  facet_wrap(~ run, nrow = 1, scales = "free_y") +
  labs(x = expression(-log[10]("expected p")), y = expression(-log[10]("observed p")),
       title = "Beta-binomial LRT calibration at each filtering stage",
       subtitle = "Note the free y-axis -- the left panel reaches ~290, the right ~28. The blues are what matter.") +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

```{r lam-tab}
w <- dcast(lam, dataset ~ run, value.var = "lambda")
setcolorder(w, c("dataset", RUN_LV))
knitr::kable(w[order(factor(dataset, levels = ORD))], digits = 3,
             caption = "lambda by dataset and stage")
t <- dcast(lam, dataset ~ run, value.var = "max_nlp")
setcolorder(t, c("dataset", RUN_LV))
knitr::kable(t[order(factor(dataset, levels = ORD))], digits = 1,
             caption = "Largest observed -log10(p): the QQ tail height")
```

# 2. The phi-boundary filter

This step is not about the phenotype at all — it is about whether the model was
successfully fitted. The distribution shown at the top of this page is not neatly
bimodal: φ occupies its entire range with several modes, so 0.99 is a cut in the
extreme tail rather than a gap between two clean populations. What *is* clean is
the discrimination:

```{r phi-tab}
knitr::kable(phis[is.finite(phi), .(
  sites = .N,
  `median phi` = signif(median(phi), 3),
  `frac phi > 0.99` = sprintf("%.3f", mean(phi > 0.99)),
  `median refit failure rate` = round(median(refit_fail, na.rm = TRUE), 3)),
  by = .(was_problematic = problematic)],
  caption = paste("The MEDIAN ordinary site sits four orders of magnitude from the",
                  "boundary, but the distribution is broad, so this contrast is",
                  "between typical values rather than between separated clusters.",
                  "'refit failure rate' is the fraction of datasets simulated from a",
                  "site's own fit that then cannot be refitted -- a direct measure of",
                  "how unstable the fit is."))
```

Two things make this a fitter correctness issue rather than a tuned threshold:

**`betabin_lib.R` already rejects boundary fits for the other parameters** — it
guards against μ hitting 0 or 1 and against runaway coefficients
(`MAX_ABS_COEF`). φ was simply omitted from that set. The right long-term home
for this rule is inside `fit_bb_mle`, not a site exclusion list.

**Data simulated from these fits often cannot be refitted at all.** Simulate 30
datasets from a site's own fitted model and try to recover it: the median ordinary
site fails 0% of the time, the median boundary site 33%. If a model cannot be
recovered from data it generated itself, nothing computed from it is trustworthy.
(The ranges overlap — some ordinary sites fail often too — so this is a contrast
of typical values, not a clean separator.)

## Does the filter converge?

Removing sites changes the site set, so the covariate PCs must be recomputed —
which changes every site's fit, which can push new sites onto the boundary. The
filter therefore has to be iterated to a fixed point rather than applied once.

```{r phi-iter}
knitr::kable(iter, digits = 5,
  caption = paste("Each round drops the flagged sites, recomputes the PCs, refits",
                  "everything, and looks for new boundary fits."))
```

It converges after `r max(iter$round)` rounds at
**`r max(iter$n_dropped_total)` sites**. Applying the filter once would have
caught `r iter$n_new_flagged[1]` and missed
`r max(iter$n_dropped_total) - iter$n_new_flagged[1]`.

φ itself is stable for the sites that are kept (correlation 0.99987 between the
first and last round, with only 0.12% of sites moving by more than 0.01), which
is why convergence is quick — the PCs enter the model through μ, while φ is
driven mostly by a site's own scatter.

Related: [the 26 problematic sites](qtl_mapping_betabinomial_site26.html) ·
[calibration after the U/C filter](qtl_mapping_betabinomial_calibration_noUCcalled.html) ·
[full pathology write-up](qtl_mapping_betabinomial_pathology_summary.html)
