---
title: "The 9 problematic sites, revisited: large φ, all-or-nothing %pU, and inherited C"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 2
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
QB <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl/qtl_betabinomial"
DF <- file.path(QB, "diagnostic_figures")
T  <- fread(file.path(DF, "problematic_sites_stats.tsv"))
L  <- fread(file.path(DF, "problematic_sites_by_line.tsv"))
PB <- fread(file.path(DF, "phi_band_untreated.tsv"))
GM <- T[sid == "GENOME-WIDE MEDIAN"]; T <- T[sid != "GENOME-WIDE MEDIAN"]
lab <- function(s) paste0("chr", sub(" ", ":", s))
T[, site := lab(sid)][, panel := ifelse(region == "-", site, paste0(site, " (", region, ")"))]
LV <- T$panel                                   # one site order everywhere: old real -log10 p, highest first
L <- merge(L, T[, .(sid, panel)], by = "sid")[, panel := factor(panel, levels = LV)]
pc <- function(x) ifelse(is.na(x), "-", sprintf("%.0f%%", 100 * x))
f2 <- function(x) ifelse(is.na(x), "fit failed", formatC(x, format = "fg", digits = 2, flag = "#"))
BLUE <- "#2a78d6"                                # single series: one colour, no legend needed
thm <- theme_bw(base_size = 10) +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_line(colour = "grey92"),
        strip.background = element_rect(fill = "grey96", colour = NA), strip.text = element_text(size = 8.5),
        plot.subtitle = element_text(size = 8.5, colour = "grey30"))
n_deep  <- T[, sum(top1_share_reads >= 0.10)]
n_full  <- T[, sum(high_lines_both > 0 & high_lines_inherited / high_lines_both >= 0.9)]
n_part  <- T[, sum(high_lines_both > 0 & high_lines_inherited > 0 & high_lines_inherited / high_lines_both < 0.9)]
n_untst <- T[, sum(high_lines_both == 0)]
PB[, band := factor(band, levels = band)]
top <- PB[band == "> 0.5"]; low <- PB[band %in% c("< 1e-4", "1e-4 to 0.01")]
```

> **The question.** The φ-estimation work showed estimates collapsing to ~0 when
> the true φ is small. Are the *problematic sites* the same problem, or a
> different one — perhaps a few samples with much higher coverage than the rest?
> And what φ do they have?
>
> **Short answer.**
>
> - **A different problem.** Their φ is **large**: `r f2(min(T$phi_multistart))`–`r f2(max(T$phi_multistart))`
>   (multi-start MLE), against about 0.01 at a typical site (genome-wide median
>   MAP estimate `r signif(GM$phi_MAP, 2)`). They are well measured, not
>   information-starved.
> - **A few very deep samples exist at `r n_deep` of the 9 sites, but they are not
>   the cause:** refitting without the deepest 1 or 3 cell lines leaves φ unchanged.
> - **What all 9 share is an all-or-nothing pattern across cell lines** — most at
>   ~0% pU, a subset at 50–100%. **Inherited C** (C in untreated RNA, i.e. a
>   genomic variant) accounts for the high cell lines fully at `r n_full` sites and
>   partly at `r n_part`; `r n_untst` sites cannot be tested.
> - **With the corrected (multi-start) fitter none of them is a hit any more.**

# 1. The sites

A site is *problematic* if its minimum cis-SNP p-value was extreme in the
**permuted** data as well as the real data (min p < 10⁻¹⁰ in at least one of 5
permutations, and real −log₁₀p > 10). Permutation breaks any genotype–phenotype
link, so such a site is reporting something about its own phenotype, not
association. These 9 were found with the **old single-start fitter** at ±5 kb,
k = 3, on `fold5_union_auto_noUCcalled` (called U/C-variant sites removed). The
[earlier diagnosis page](qtl_mapping_betabinomial_problematic_diagnosis.html)
describes them with that fitter.

```{r sites}
knitr::kable(rbind(
  T[, .(site, region, `cell lines` = cell_lines, `%pU` = sprintf("%.1f%%", 100 * pooled_pU),
        `total reads` = format(round(total_reads), big.mark = ","), `median reads per line` = median_reads)],
  GM[, .(site = "*genome-wide median*", region = "", `cell lines` = cell_lines, `%pU` = sprintf("%.1f%%", 100 * pooled_pU),
         `total reads` = format(round(total_reads), big.mark = ","), `median reads per line` = median_reads)]),
  caption = "Region labels use approximate hg38 spans (IGH chr14:105.5-107 Mb, IGK chr2:88.8-90.4 Mb, MHC chr6:28.5-33.5 Mb), not yet checked against the gene annotation.")
```

They are **deep**: `r format(round(min(T$total_reads)), big.mark = ",")`–`r format(round(max(T$total_reads)), big.mark = ",")`
reads per site against a genome-wide median of `r format(round(GM$total_reads), big.mark = ",")`.
The low-φ collapse is a problem of shallow sites; these are the opposite.

# 2. Their φ is large, not collapsed

```{r phi}
knitr::kable(rbind(
  T[, .(site, `old single-start` = f2(phi_single_start), `multi-start MLE` = f2(phi_multistart),
        `moment estimate` = f2(phi_moment), `MAP (mom-q60 prior)` = f2(phi_MAP))],
  GM[, .(site = "*genome-wide median*", `old single-start` = f2(phi_single_start), `multi-start MLE` = f2(phi_multistart),
         `moment estimate` = f2(phi_moment), `MAP (mom-q60 prior)` = f2(phi_MAP))]),
  caption = "The old fitter pinned all 9 at the phi = 1 boundary -- a secondary optimum of a bimodal likelihood (see the optimiser page). The multi-start fit finds the real optimum, still large. The moment estimate is unbounded, hence 2.1 at one site.")
```

A φ of 0.5–0.9 means the fitted distribution of %pU across cell lines is
**U-shaped**: most mass near 0% and near 100%, little in between.

# 3. A few very deep cell lines — at some sites, but not the cause

```{r deep}
knitr::kable(rbind(
  T[, .(site, `max / median reads` = sprintf("%.0fx", max_over_median),
        `deepest line: share of reads` = pc(top1_share_reads), `3 deepest: share of reads` = pc(top3_share_reads),
        `phi, all lines` = f2(phi_multistart), `phi without deepest` = f2(phi_without_deepest),
        `phi without 3 deepest` = f2(phi_without_3_deepest))],
  GM[, .(site = "*genome-wide median*", `max / median reads` = sprintf("%.0fx", max_over_median),
         `deepest line: share of reads` = pc(top1_share_reads), `3 deepest: share of reads` = pc(top3_share_reads),
         `phi, all lines` = "", `phi without deepest` = "", `phi without 3 deepest` = "")]),
  caption = "Multi-start MLE, k = 3. Removing the deepest cell lines does not reduce phi at any site.")
```

At `r n_deep` sites — the four in IGK/IGH and chr17:36212383 — one cell line
carries 13–40% of all reads, far above the genome-wide 6%. But **dropping the
deepest line, or the three deepest, leaves φ where it was**. The figure shows why:
the deep lines are not special in their %pU.

```{r depthfig, fig.width=10, fig.height=7.2}
ggplot(L, aes(n, 100 * y / n)) +
  geom_point(colour = BLUE, size = 1.6, alpha = 0.75, shape = 16) +
  facet_wrap(~ panel, ncol = 3) +
  scale_x_log10(labels = scales::label_comma()) +
  scale_y_continuous(limits = c(0, 100), breaks = c(0, 50, 100)) +
  labs(x = "reads in this cell line (log scale)", y = "%pU in this cell line",
       title = "Each point is one cell line: its depth against its %pU",
       subtitle = "Deep lines (right) sit at 0% or at high %pU like the rest. What stands out is the gap: lines are near 0% or near 50-100%, rarely in between.") +
  thm
```

# 4. What all 9 share: all-or-nothing %pU

```{r pattern}
knitr::kable(rbind(
  T[, .(site, `cell lines with >=10 reads` = lines_ge10, `at ~0% (< 5%)` = pc(frac_lines_zero),
        `5-50%` = pc(frac_lines_mid), `above 50%` = pc(frac_lines_high))],
  GM[, .(site = "*genome-wide median*", `cell lines with >=10 reads` = lines_ge10, `at ~0% (< 5%)` = pc(frac_lines_zero),
         `5-50%` = pc(frac_lines_mid), `above 50%` = pc(frac_lines_high))]),
  caption = "Share of cell lines (with at least 10 reads) in each %pU band.")
```

At a typical site essentially no cell line is above 50% pU; here
`r pc(min(T$frac_lines_high))`–`r pc(max(T$frac_lines_high))` are, while the
in-between band is nearly empty. People differing by "none vs most" is what a
variant looks like, not a modification level.

# 5. The untreated controls: inherited C

Untreated input RNA has no chemical conversion, so a C at the pU base there is
**inherited** (genomic) C. This was validated on known controls: a known U/C
variant reads 0.51 C untreated, an ordinary modified site reads 0.00. If the
high-%pU cell lines also read C untreated, the "modification" is a variant that
the genotype calls missed.

```{r untreatedfig, fig.width=10, fig.height=7.4}
U <- L[n >= 10 & n_untreated >= 10]
R <- T[, .(panel = factor(panel, levels = LV), r = r_treated_untreated)]
ggplot(U, aes(y_untreated / n_untreated, y / n)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linewidth = 0.4) +
  geom_point(colour = BLUE, size = 1.6, alpha = 0.75, shape = 16) +
  geom_text(data = R, aes(x = 0.02, y = 0.97, label = ifelse(is.na(r), "r = -", sprintf("r = %.2f", r))),
            hjust = 0, vjust = 1, size = 3, colour = "grey15", inherit.aes = FALSE) +
  facet_wrap(~ panel, ncol = 3) +
  scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1), labels = scales::label_percent()) +
  scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1), labels = scales::label_percent()) +
  coord_equal() +
  labs(x = "C at the pU base in UNTREATED RNA (inherited)", y = "%pU in TREATED RNA",
       title = "Treated %pU against untreated C, per cell line (lines with >= 10 reads in both)",
       subtitle = "Grey line: y = x. Points on it mean the treated signal is entirely inherited C. r = correlation across cell lines.") +
  thm
```

```{r untreatedtab}
T[, verdict := fifelse(high_lines_both == 0, "not testable",
               fifelse(high_lines_inherited / high_lines_both >= 0.9, "inherited C",
               fifelse(high_lines_inherited > 0, "partly inherited C", "no inherited C")))]
knitr::kable(T[, .(site, `high cell lines (treated > 50%)` = high_lines_treated_only,
                   `...with untreated data` = high_lines_both,
                   `...carrying inherited C (untreated C >= 10%)` = high_lines_inherited,
                   `Spearman rho` = f2(rho_treated_untreated), `Pearson r` = f2(r_treated_untreated), verdict)],
  caption = "Cell lines with at least 10 reads. The count of high cell lines carrying inherited C is the main measure: with only a few high lines, a correlation is carried by one or two points.")
```

- **Inherited C at `r n_full` sites.** At chr6:29943223 all 26 high cell lines
  carry C untreated, and they sit on the diagonal; at chr3:37277733, 18 of 20; at
  chr6:32643315 both high lines.
- **Partly, at `r n_part`** (chr14:106627394, chr17:36212383, chr14:106639258):
  one high line carries inherited C, the other one or two do not. The
  correlations there (0.53–0.60) are carried by that one line.
- **chr2:89085177**: its one high line has no inherited C, but two other lines are
  ~50% C untreated (heterozygous-looking) while reading only ~35% treated.
- **Not testable, `r n_untst` sites** (chr2:90039467, chr17:53106304): their
  high-%pU cell lines have fewer than 10 untreated reads, so their correlations
  are computed on low cell lines only and say nothing.

# 6. With the corrected fitter they are no longer hits

```{r assoc}
knitr::kable(T[, .(site, `old fitter: real` = sprintf("%.1f", old_real_nlp),
                   `old fitter: worst permutation` = sprintf("%.1f", old_worst_perm_nlp),
                   `multi-start: real` = sprintf("%.1f", ms_real_nlp),
                   `multi-start: worst permutation` = sprintf("%.1f", ms_worst_perm_nlp))],
  caption = sprintf("Minimum p over cis-SNPs, as -log10 p. Multi-start permutation ceiling (best permuted site genome-wide): %.1f.",
                    T$perm_ceiling_multistart[1]))
```

Under the old fitter the real and permuted p-values were **equal** — the extreme
values had nothing to do with genotype; they came from the φ = 1 fit. With
multi-start, real −log₁₀p falls to `r sprintf("%.1f", min(T$ms_real_nlp))`–`r sprintf("%.1f", max(T$ms_real_nlp))`,
all below the permutation ceiling of `r sprintf("%.1f", T$perm_ceiling_multistart[1])`.

<!-- # 7. Beyond the 9: the high-φ tail -->

<!-- The same count, applied to every site: of all cell lines with high %pU -->
<!-- (treated > 50%, ≥ 10 reads treated and untreated), how many carry inherited C? -->

<!-- ```{r bandfig, fig.width=7.5, fig.height=3.8} -->
<!-- ggplot(PB, aes(band, high_lines_inherited_pct)) + -->
<!--   geom_col(fill = BLUE, width = 0.68) + -->
<!--   geom_text(aes(label = sprintf("%.0f%%", high_lines_inherited_pct)), vjust = -0.45, size = 3.2, colour = "grey15") + -->
<!--   scale_y_continuous(limits = c(0, 100), expand = expansion(mult = c(0, 0.04))) + -->
<!--   labs(x = expression("estimated " * varphi * " of the site (multi-start MLE)"), -->
<!--        y = "% of high-%pU cell lines", -->
<!--        title = "High-%pU cell lines that carry inherited C, by the site's estimated phi", -->
<!--        subtitle = "High = treated %pU > 50%; inherited = untreated C >= 10%; lines with >= 10 reads in both. Counts in the table below.") + -->
<!--   thm + theme(panel.grid.major.x = element_blank()) -->
<!-- ``` -->

<!-- ```{r bandtab} -->
<!-- knitr::kable(PB[, .(`estimated phi` = band, sites = format(sites, big.mark = ","), -->
<!--                     `sites with a high line` = sites_with_high_line, -->
<!--                     `all high lines inherited C` = sprintf("%.0f%%", all_high_inherited_pct), -->
<!--                     `some` = sprintf("%.0f%%", some_high_inherited_pct), -->
<!--                     `none` = sprintf("%.0f%%", none_high_inherited_pct), -->
<!--                     `high lines` = format(high_lines, big.mark = ","), -->
<!--                     `high lines with inherited C` = sprintf("%.0f%%", high_lines_inherited_pct))], -->
<!--   caption = "Most sites with small phi have no high-%pU cell line at all, which is why the counts are small in the upper rows.") -->
<!-- ``` -->

<!-- At sites with φ above 0.5, `r sprintf("%.0f%%", top$high_lines_inherited_pct)` of the -->
<!-- high-%pU cell lines carry inherited C, against about -->
<!-- `r sprintf("%.0f%%", mean(low$high_lines_inherited_pct))` at sites with φ below 0.01. So -->
<!-- most of the large-φ signal comes from inherited C rather than large differences -->
<!-- in modification between people. This strengthens the reading -->
<!-- that **genuine between-person φ is small**, and it means the untreated controls -->
<!-- should be used to filter aliasing sites before the final QTL scan. -->

<!-- # 8. Caveats -->

<!-- - **Region labels are approximate**: hand-entered hg38 spans, not yet checked -->
<!--   against the gene annotation. -->
<!-- - **Correlations mislead here.** With one to three high cell lines, Pearson r is -->
<!--   carried by those few points, so the page leads with the count of high lines -->
<!--   carrying inherited C. -->
<!-- - **"Inherited" uses a 10% untreated-C cut** in a line with ≥ 10 untreated reads; -->
<!--   a heterozygous carrier reads about 50%, so the cut is permissive. -->
<!-- - **Two sites (chr2:90039467, chr17:53106304) cannot be tested**: their high-%pU -->
<!--   cell lines lack untreated coverage. -->
<!-- - **Untreated depth limits section 7 too**: only cell lines with ≥ 10 reads in -->
<!--   both treated and untreated RNA count, which leaves out many shallow sites. -->
<!-- - At chr6:32643315 the refit without the three deepest lines failed; without the -->
<!--   deepest line alone, φ is unchanged (0.86). -->

<!-- # 9. Provenance -->

<!-- - All tables and figures: `4.qtl_mapping_betabinomial/42.problematic_sites_stats.R` -->
<!--   → `qtl_betabinomial/diagnostic_figures/problematic_sites_stats.tsv`, -->
<!--   `problematic_sites_by_line.tsv`, `phi_band_untreated.tsv`. -->
<!-- - Site list: `diagnostic_figures/pathological_sites_w5kb_pc3_auto_noUCcalled.txt` -->
<!--   (old single-start calibration, ±5 kb, k = 3). -->
<!-- - φ: single- and multi-start from `multistart_audit/multistart_by_site.txt.gz`; -->
<!--   moment and MAP from `phi_recovery/real_phi_estimates.txt.gz` -->
<!--   (`35.real_phi_moment_prior.R`). -->
<!-- - Untreated controls: `untreated_counts/` (`12.untreated_counts_matrix.R`). -->
<!-- - Multi-start p-values: `lrt_calibration/calib_siteres_w5kb_pc3_auto_noUCcalled_MS_pvals.txt.gz`. -->
