---
title: "Linear-regression QTL mapping: phenotype and genotype data"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 2
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages(library(data.table))
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
PR    <- "/project/xinhe/xsun/pumseq/pumseq_processed"
UNION <- file.path(PR, "union/all51_mRNA")
PHENO <- file.path(PR, "psi_qtl/phenotype/union_ncalled2/treated_minus_input_floor0/quantile")
GENO  <- file.path(PR, "psi_qtl/genotype")
fmt <- function(x) format(x, big.mark = ",", trim = TRUE, scientific = FALSE)

# sites: the union call set, then the two filters applied for QTL mapping
allsites <- fread(file.path(UNION, "union_mRNA.global_sites.txt.gz"), colClasses = list(character = c("ref", "strand")))
allsites[, n_called := lengths(strsplit(called_in, ","))]
final <- fread(file.path(PHENO, "sites_info.txt.gz"), colClasses = list(character = c("ref", "strand")))
pm    <- fread(file.path(PHENO, "phenotype_matrix.txt.gz"))
cls   <- setdiff(names(pm), "site_id")

# raw treated and input %pU at the final sites, to report the floor and the imputation
key <- function(d) paste(d$ref, d$pos, d$strand)
trt <- fread(file.path(UNION, "union_mRNA.global_sites.ratio_matrix.txt.gz"), colClasses = list(character = c("ref", "strand")))
inp <- fread(file.path(UNION, "union_mRNA.global_sites.ratio_matrix_unt.txt.gz"), colClasses = list(character = c("ref", "strand")))
i <- match(key(final), key(trt)); stopifnot(!anyNA(i), identical(key(trt), key(inp)))
D <- as.matrix(trt[i, ..cls]) - as.matrix(inp[i, ..cls])
pct_missing <- 100 * mean(is.na(D))
pct_floored <- 100 * mean(D[!is.na(D)] < 0)
pct_zero    <- 100 * mean(D[!is.na(D)] == 0)

# genotype count: variants in the PLINK2 file
pvar_n <- length(grep("^#", readLines(file.path(GENO, "plink2/pumseq_genotype.hg38.chr1-22.pvar")), invert = TRUE))
map <- fread(file.path(GENO, "sample_id_map.txt"))
```

This page describes the inputs to the linear-regression (tensorQTL) pU-QTL scan
summarised on the [linear QTL mapping page](qtl_mapping_summary_floor0_quantile.html).

```{r overview}
knitr::kable(data.table(
  item = c("cell lines", "pU sites", "phenotype", "covariates", "genotypes", "cis window", "test"),
  value = c(sprintf("%d (CL19039 has no genotype and is excluded)", length(cls)),
            fmt(nrow(final)),
            "quantile-normalised (treated - input) %pU, negative differences set to 0",
            "phenotype PCs, 3 to 30 swept",
            sprintf("%s autosomal variants (hg38), no allele-frequency filter", fmt(pvar_n)),
            "+/-5, 20, 50 and 100 kb",
            "linear regression of the phenotype on genotype dosage (tensorQTL)")),
  col.names = c("", ""), caption = "What the linear model sees.")
```

# 1. Sites

The sites come from **union site calling**. Each of the 51 sequenced cell lines
is called on its own (chi-squared test against pooled input, FDR < 0.01, rate
difference > 0.05, depth >= 20), and a site enters the list if any cell line
calls it. Details are on the
[union site-calling QC page](qc_union_all51_mRNA.html).

Two filters are then applied for QTL mapping:

```{r sitefunnel}
knitr::kable(data.table(
  step = c("union call set (any of 51 cell lines)",
           "called independently in >= 2 cell lines",
           "on chr1-22 (the genotype panel is autosomal)"),
  sites = fmt(c(nrow(allsites), allsites[n_called >= 2, .N], allsites[n_called >= 2 & ref %in% as.character(1:22), .N]))),
  caption = "Site counts at each step.")
```

The last count, `r fmt(nrow(final))`, is the number of sites in the phenotype matrix.

# 2. Phenotype

For each site and cell line:

1. **Treated %pU**: converted reads C / (C + T) in the treated library, with
   treated replicates pooled.
2. **Input %pU**: the same ratio in the cell line's pooled input (untreated)
   library. This is the background conversion rate.
3. **Treated minus input.** Negative differences are set to 0, since treatment
   should only add signal on top of background. This changes
   `r sprintf("%.1f%%", pct_floored)` of the measured values at these sites; with the
   `r sprintf("%.1f%%", pct_zero)` that are exactly 0 already, `r sprintf("%.1f%%", pct_floored + pct_zero)`
   of values end up at 0.
4. **Missing values** (no coverage in a cell line, `r sprintf("%.1f%%", pct_missing)` of
   entries) are filled with the site's median across cell lines.
5. **Quantile normalisation** across cell lines, so every cell line has the same
   distribution of values.

The result is a `r fmt(nrow(pm))` x `r length(cls)` matrix (sites x cell lines),
written as a BED file for tensorQTL.

# 3. Covariates

Principal components of the normalised phenotype matrix (cell lines as
observations, sites as features, centred). Covariate files with 3, 5, 7, 10,
15, 20, 25 and 30 PCs are produced; the scans sweep 3 to 30.

# 4. Genotypes

**Source.** Imputed genotypes for 60 Yoruba (YRI) lymphoblastoid cell lines,
downloaded from [Zenodo record 3870952](https://zenodo.org/records/3870952), on
hg19. They were lifted to hg38 with GATK `LiftoverVcf` (4.5.0, UCSC
hg19ToHg38 chain); variants that could not be lifted were dropped.

**Processing** (`2.qtl_mapping_linear/1.run_merge_genotype_vcf.sbatch`, then
`4.convert_genotype_plink2.sbatch`):

- chromosomes 1-22 merged into one VCF (chrX excluded);
- subset to the `r nrow(map)` sequenced cell lines that have genotypes, and
  renamed from `NA#####` to `CL#####` to match the phenotype;
- converted to PLINK2 format, keeping chr1-22 only, with every variant ID reset
  to `chrom:pos:ref:alt` (the VCF has missing and duplicated rsIDs) and half-calls
  set to missing.

**What the linear model uses:** all `r fmt(pvar_n)` variants in that PLINK2 file.
No minor-allele-frequency filter is set, which is tensorQTL's default. (The
binomial and beta-binomial models use a filtered SNP set; see their data pages.)

# 5. Files and scripts

```{r files}
knitr::kable(data.table(
  what = c("phenotype matrix (BED for tensorQTL)", "site metadata", "covariates", "genotypes (PLINK2)", "sample ID map"),
  path = c(file.path(PHENO, "phenotype_matrix.bed.gz"), file.path(PHENO, "sites_info.txt.gz"),
           file.path(PHENO, "pca/covariates_<N>PCs.txt"), file.path(GENO, "plink2/pumseq_genotype.hg38.chr1-22.{pgen,pvar,psam}"),
           file.path(GENO, "sample_id_map.txt"))),
  caption = "Scripts: 2.qtl_mapping_linear/ -- 1.run_merge_genotype_vcf.sbatch, 2.prepare_phenotype.R (metric treated_minus_input_floor0), 3.compute_phenotype_pcs.R, 4.convert_genotype_plink2.sbatch, 5.run_tensorqtl.sbatch.")
```
