---
title: "QTL mapping: treated - input, floor0 quantile normalization"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r setup, warning=FALSE, message=FALSE}
library(data.table)
library(dplyr)
library(ggplot2)
library(qvalue)
library(arrow)
library(patchwork)

knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", fig.width = 10, fig.height = 6
)

PROJECT_DIR <- "/project/xinhe/xsun/pumseq"
```

# Introduction

This page summarizes where cis-QTL mapping for the **treated − input**
pseudouridylation (%pU) phenotype currently stands: the data issue, how the phenotype is
now built and normalized, how the QTL scan is set up, and what the
results and calibration checks look like.

# 1. The data: treated - input has many negative values

The phenotype of interest is the differential %pU signal — %pU in the
treated sample minus %pU in the matched input (untreated) sample, at
each pseudouridylation site, for each of the 50 genotyped cell lines.
Under the **union** site-calling strategy, a site enters the analysis
if it clears a significance threshold (FDR < 0.01, rate-difference > 0.05, depth>=20) in
**at least one** cell line; every other cell line's raw %pU value at
that site is then reported as-is, with no per-cell-line filter.

Conceptually, `treated - input` should not be negative: input is the
background/baseline rate, and treatment should only add signal on top
of it. But because most cell lines don't necessarily show real signal
at a site that was only called significant in *some other* cell line,
the raw per-cell-line, per-site difference is often small, noisy, and
sometimes negative.

```{r load-raw-diff}
UNION_DIR <- file.path(PROJECT_DIR, "pumseq_processed/union/all51_mRNA")

ratio_treated <- fread(file.path(UNION_DIR, "union_mRNA.global_sites.ratio_matrix.txt.gz"),
                        colClasses = list(character = c("ref", "strand")))
ratio_input   <- fread(file.path(UNION_DIR, "union_mRNA.global_sites.ratio_matrix_unt.txt.gz"),
                        colClasses = list(character = c("ref", "strand")))

id_cols <- c("ref", "pos", "strand")
cl_cols <- setdiff(colnames(ratio_treated), id_cols)
stopifnot(identical(cl_cols, setdiff(colnames(ratio_input), id_cols)))

diff_mat <- as.matrix(ratio_treated[, ..cl_cols]) - as.matrix(ratio_input[, ..cl_cols])
diff_vec <- diff_mat[!is.na(diff_mat)]

diff_summary <- data.table(
  n_total       = length(diff_vec),
  n_negative    = sum(diff_vec < 0),
  `pct_<_0`  = round(100 * mean(diff_vec < 0), 1),
  `pct_<=_0`      = round(100 * mean(diff_vec <= 0), 1),
  median        = round(median(diff_vec), 4),
  min           = round(min(diff_vec), 3),
  max           = round(max(diff_vec), 3)
)
knitr::kable(diff_summary, caption = "per-site (treated - input) values across all union sites")
```

```{r raw-diff-hist, fig.height=4.5}
ggplot(data.table(diff = diff_vec), aes(x = diff)) +
  geom_histogram(bins = 100, fill = "steelblue", color = NA) +
  geom_vline(xintercept = 0, color = "firebrick", linetype = "dashed") +
  labs(x = "treated - input (%pU)", y = "count",
       title = "Raw (treated - input) values across all sites x cell lines",
       subtitle = sprintf("%.1f%% of %s values are negative; %.1f%% are <= 0",
                           diff_summary$pct_negative, format(diff_summary$n_total, big.mark=","),
                           diff_summary$pct_le_0)) +
  theme_bw(base_size = 13)
```

<!-- Over half of all (site, cell line) values sit at or below zero, and the -->
<!-- median is exactly zero — there is no meaningful biological -->
<!-- interpretation for a "negative" differential editing rate; it reflects -->
<!-- sampling noise around an absence of signal in that cell line at that -->
<!-- site, not real negative editing. -->

# 2. Flooring negative values to 0

Since a negative value isn't biologically meaningful, it's treated as
noise around "no detectable signal" rather than discarded outright.

If we require positivity in every cell line at a site, it keeps only ~700-800
of 5,928 sites (far too aggressive). Instead,
negative values are simply floored to 0, keeping every site:

```{r floor-illustration, fig.height=4.5}
diff_floored <- pmax(diff_vec, 0)

plot_dt <- rbind(
  data.table(value = diff_vec,     version = "raw (before flooring)"),
  data.table(value = diff_floored, version = "floored to 0")
)
plot_dt[, version := factor(version, levels = c("raw (before flooring)", "floored to 0"))]

ggplot(plot_dt, aes(x = value)) +
  geom_histogram(bins = 100, fill = "steelblue", color = NA) +
  geom_vline(xintercept = 0, color = "firebrick", linetype = "dashed") +
  facet_wrap(~ version, ncol = 1, scales = "free_y") +
  labs(x = "treated - input (%pU)", y = "count",
       title = "Effect of flooring negative values to 0") +
  theme_bw(base_size = 13)
```

The floored value (`pmax(treated - input, 0)`) is what feeds every
downstream step below.

# 3. Quantile normalization

After flooring, sites are restricted to the union set called in >= 2 cell lines and to autosomes (chr1-22), missing values are imputed
with the site-level median, and the resulting matrix is
quantile-normalized across cell lines — forcing every cell line to
share the same marginal distribution, so that differences in overall
signal level or spread between cell lines (technical, not biological)
don't drive spurious associations.

```{r quantile-norm-illustration, fig.height=8}
PHENO_DIR <- file.path(PROJECT_DIR, "pumseq_processed/psi_qtl/phenotype/union_ncalled2/treated_minus_input_floor0/quantile")

sites_info <- fread(file.path(PHENO_DIR, "sites_info.txt.gz"),
                     colClasses = list(character = c("ref", "strand")))
genotype_map <- fread(file.path(PROJECT_DIR, "pumseq_processed/psi_qtl/genotype/sample_id_map.txt"))
cl_genotyped <- intersect(cl_cols, genotype_map$CL_id)

setkey(ratio_treated, ref, pos, strand)
setkey(ratio_input, ref, pos, strand)
before_treated <- ratio_treated[sites_info[, .(ref, pos, strand)], on = c("ref", "pos", "strand")]
before_input   <- ratio_input[sites_info[, .(ref, pos, strand)], on = c("ref", "pos", "strand")]
before_mat <- pmax(as.matrix(before_treated[, ..cl_genotyped]) - as.matrix(before_input[, ..cl_genotyped]), 0)

after_mat <- as.matrix(fread(file.path(PHENO_DIR, "phenotype_matrix.txt.gz"))[, ..cl_genotyped])

box_dt <- rbind(
  melt(as.data.table(before_mat), measure.vars = cl_genotyped)[, version := "before quantile normalization"],
  melt(as.data.table(after_mat),  measure.vars = cl_genotyped)[, version := "after quantile normalization"]
)
box_dt[, version := factor(version, levels = c("before quantile normalization", "after quantile normalization"))]

ggplot(box_dt[!is.na(value)], aes(x = variable, y = value)) +
  geom_boxplot(outlier.size = 0.3, fill = "steelblue", alpha = 0.6) +
  facet_wrap(~ version, ncol = 1, scales = "free_y") +
  labs(x = "cell line", y = "phenotype value",
       title = "Per-cell-line distribution before vs. after quantile normalization") +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

After normalization, every cell line's marginal distribution is
identical by construction — the boxplots collapse to the same shape and
spread, so any remaining site-to-site variation is what QTL mapping
tests against genotype.

# 4. QTL mapping approach

cis-QTL mapping is done with [tensorQTL](https://github.com/broadinstitute/tensorqtl)
(tensorQTL reproduces the same beta-approximated permutation statistics, just faster). For each
phenotype site, both modes are run: `cis_nominal` (nominal association
p-value for every variant in the window) and `cis` (permutation pass,
one best-hit p-value per site, corrected via a beta-distribution
approximation for the number of variants tested).

Two things are swept to characterize the result:

- **Number of covariate PCs** (phenotype-derived PCA, used as
  covariates): 3, 5, 7, 10, 15, 20, 30.
- **cis-window size** (distance from the phenotype site within which
  variants are tested): **5kb, 20kb, 50kb, 100kb**.

That's a 7 x 4 grid, each cell a separate tensorQTL run, on the
floored, quantile-normalized phenotype above.

# 5. Results: eGenes by PC count and window size

```{r load-results}
RESULTS_ROOT <- file.path(PROJECT_DIR, "pumseq_processed/psi_qtl/qtl_results/union_ncalled2")
STRATEGY    <- "treated_minus_input_floor0"
PC_COUNTS   <- c(3, 5, 7, 10, 15, 20, 30)
WINDOWS_BP  <- c(5000, 20000, 50000, 100000)

compute_qval <- function(p) {
  q <- tryCatch(qvalue::qvalue(p)$qvalues, error = function(e) NULL)
  if (is.null(q)) q <- p.adjust(p, method = "BH")
  q
}

read_one <- function(pc, window) {
  window_dir <- paste0("window", format(window, scientific = FALSE), "bp/")
  f <- file.path(RESULTS_ROOT, STRATEGY, "quantile", window_dir, paste0("pcs", pc),
                 paste0("pumseq_", STRATEGY, ".cis_qtl.txt.gz"))
  if (!file.exists(f)) return(NULL)
  dt <- fread(f, header = TRUE, sep = "\t")
  dt[, `:=`(n_pcs = pc, window_bp = window)]
  dt
}

grid <- CJ(pc = PC_COUNTS, window = WINDOWS_BP)
loaded <- mapply(read_one, grid$pc, grid$window, SIMPLIFY = FALSE)
stopifnot(sum(sapply(loaded, is.null)) == 0)
cis_qtl <- rbindlist(loaded)
cis_qtl[, qval := compute_qval(pval_beta), by = .(n_pcs, window_bp)]

summary_tbl <- cis_qtl[, .(eGenes_FDR05 = sum(qval < 0.05, na.rm = TRUE)), by = .(window_bp, n_pcs)][order(window_bp, n_pcs)]
summary_tbl[, window_lab := factor(paste0(window_bp / 1000, "kb"), levels = paste0(WINDOWS_BP / 1000, "kb"))]
```

```{r egene-vs-pc, fig.height=5}
ggplot(summary_tbl, aes(x = n_pcs, y = eGenes_FDR05, color = window_lab)) +
  geom_line() +
  geom_point(size = 2) +
  scale_x_continuous(breaks = PC_COUNTS) +
  labs(x = "Number of covariate PCs", y = "eGenes (FDR<0.05)", color = "cis-window",
       title = "eGenes vs. number of covariate PCs, by cis-window size") +
  theme_bw(base_size = 13)
```

eGene counts fall off steadily as more covariate PCs are added at every
window size 
<!-- — with only 50 samples, higher PC counts spend degrees of -->
<!-- freedom faster than they remove noise, so fewer PCs (3-10) consistently -->
<!-- yield more eGenes across all four window sizes. -->

```{r egene-heatmap, fig.height=4.5}
ggplot(summary_tbl, aes(x = factor(n_pcs, levels = PC_COUNTS), y = window_lab, fill = eGenes_FDR05)) +
  geom_tile() +
  geom_text(aes(label = eGenes_FDR05), size = 3.8) +
  scale_fill_gradient(low = "white", high = "firebrick") +
  labs(x = "Number of covariate PCs", y = "cis-window size", fill = "eGenes\n(FDR<0.05)",
       title = "eGenes (FDR<0.05) by window size x PC count") +
  theme_bw(base_size = 13)
```

Both window size and PC count matter: smaller windows and fewer PCs
each independently yield more eGenes, with the largest counts in the
top-left of the heatmap (5-20kb, 3-7 PCs).

# 6. Permutation control

<!-- Because eGene counts are highest at the narrowest windows, it's -->
<!-- important to check whether that's real signal or an artifact of -->
<!-- scanning fewer variants.  -->

The check: break the true genotype-phenotype
relationship by permuting which individual's phenotype and covariates
get tested against which individual's genotype (5 independent random
permutations), then rerun the nominal association test on each permuted
copy exactly as on the real data, at PC=5. If the real
result's calibration is genuine, the permuted nulls should track the
null diagonal in a QQ plot regardless of window size

<!-- ; if the narrow-window -->
<!-- eGene counts were merely an artifact of fewer variants being tested, the -->
<!-- permuted nulls (which have exactly as few variants) would show it too. -->

```{r perm-load}
N_PCS_PERM <- 5
CHRS <- 1:22
N_PERM <- 5

read_nominal <- function(metric, prefix) {
  dir <- file.path(RESULTS_ROOT, metric, "quantile", paste0("pcs", N_PCS_PERM))
  rbindlist(lapply(CHRS, function(chr) {
    f <- file.path(dir, paste0(prefix, ".cis_qtl_pairs.", chr, ".parquet"))
    as.data.table(read_parquet(f, col_select = c("pval_nominal", "start_distance")))
  }))
}

real_dt <- read_nominal(STRATEGY, paste0("pumseq_", STRATEGY))
perm_dt <- lapply(seq_len(N_PERM), function(s) {
  metric <- paste0(STRATEGY, "_perm", s)
  read_nominal(metric, paste0("pumseq_", metric))
})

compute_lambda <- function(p) median(qchisq(1 - p, df = 1), na.rm = TRUE) / qchisq(0.5, df = 1)

lambda_by_window <- rbindlist(lapply(WINDOWS_BP, function(w) {
  real_p <- real_dt[abs(start_distance) <= w, pval_nominal]
  perm_p <- lapply(perm_dt, function(dt) dt[abs(start_distance) <= w, pval_nominal])
  data.table(
    window_kb = w / 1000,
    dataset   = c("real", paste0("permuted seed ", seq_len(N_PERM))),
    lambda_gc = c(compute_lambda(real_p), sapply(perm_p, compute_lambda))
  )
}))

knitr::kable(
  dcast(lambda_by_window, dataset ~ window_kb, value.var = "lambda_gc"),
  digits = 3, caption = "Genomic control lambda by cis-window size (kb): real vs. 5 permuted nulls, PC=5"
)
```

```{r perm-qq, fig.height=9}
thin_qq <- function(p, label, window_kb) {
  p <- sort(p)
  n <- length(p)
  if (n < 2) return(NULL)
  idx <- unique(as.integer(10^(seq(log10(1), log10(n), length.out = min(5000, n)))))
  data.table(observed = -log10(p[idx]), expected = -log10(ppoints(n)[idx]),
             dataset = label, window_kb = window_kb)
}

qq_dt <- rbindlist(lapply(WINDOWS_BP, function(w) {
  real_p <- real_dt[abs(start_distance) <= w, pval_nominal]
  rbindlist(c(
    list(thin_qq(real_p, "real", w / 1000)),
    lapply(seq_len(N_PERM), function(s) {
      p <- perm_dt[[s]][abs(start_distance) <= w, pval_nominal]
      thin_qq(p, paste0("permuted seed ", s), w / 1000)
    })
  ))
}))
qq_dt[, is_real := dataset == "real"]
qq_dt[, window_lab := factor(paste0(window_kb, "kb"), levels = paste0(WINDOWS_BP / 1000, "kb"))]

ggplot(qq_dt, aes(x = expected, y = observed, color = is_real, group = dataset)) +
  geom_point(size = 0.6, alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "black", linetype = "dashed") +
  facet_wrap(~ window_lab, ncol = 2, scales = "free") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "grey60"),
                      labels = c(`TRUE` = "real", `FALSE` = "permuted null"), name = NULL) +
  labs(x = expression(-log[10](expected~p)), y = expression(-log[10](observed~p)),
       title = "Real vs. 5 permuted nulls, by cis-window size (treated_minus_input_floor0, PC=5)") +
  theme_bw(base_size = 13)
```

<!-- The permuted nulls (grey) sit close to the diagonal at every window -->
<!-- size, while the real result (red) pulls progressively further above it -->
<!-- as the window narrows. Since the permuted nulls have exactly as few -->
<!-- variants per window as the real data and show no such inflation, this -->
<!-- rules out "too few variants" as the explanation — the effect is -->
<!-- specific to the real genotype-phenotype pairing. The most likely -->
<!-- explanation is a confound acting through local LD structure: nearby -->
<!-- variants are correlated with each other, so any residual correlation -->
<!-- between the phenotype and population structure/relatedness — which -->
<!-- current covariates (phenotype-derived PCs only, no genotype PCs) don't -->
<!-- correct for — shows up more strongly in a narrow window's tightly-linked -->
<!-- variant set than when diluted across a wider one. This means eGene -->
<!-- calls at the narrowest windows should be treated as provisional until -->
<!-- that confound is addressed, even though the calibration check itself -->
<!-- behaved as expected. -->

<!-- # Summary -->

<!-- - The union strategy's `treated - input` phenotype has substantial -->
<!--   negative mass (54.5% &le; 0) with no biological meaning; negative -->
<!--   values are floored to 0 rather than discarded or log-transformed. -->
<!-- - The floored phenotype is filtered (n_called &ge; 2, autosomes), -->
<!--   imputed, and quantile-normalized across cell lines before QTL mapping. -->
<!-- - cis-QTL mapping (tensorQTL) was swept across 7 PC counts x 4 window -->
<!--   sizes (5-100kb). Fewer PCs and narrower windows both increase eGene -->
<!--   counts. -->
<!-- - A label-permutation control confirms this narrow-window effect is a -->
<!--   real property of the genotype-phenotype relationship, not a -->
<!--   statistical artifact — but also flags a likely uncorrected confound -->
<!--   (population structure/relatedness via local LD) that should be -->
<!--   resolved before treating narrow-window eGene calls as final. -->
