---
title: "Beta-binomial pU-QTL: diagnosing the problematic sites (k = 3, ±5 kb, U/C sites removed)"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 3
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 11)

PSI  <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl"
PC   <- file.path(PSI, "qtl_betabinomial/prior_pi_cases")
DIAG <- file.path(PSI, "qtl_betabinomial/diagnostic_figures")

par   <- fread(file.path(PC, "case_params_patho.tsv"), sep = "\t")
bins  <- fread(file.path(PC, "prior_bins_patho.tsv"), sep = "\t")
obs   <- fread(file.path(PC, "observed_patho.tsv"), sep = "\t")
# sid contains a space, so the separator must be explicit
pl    <- fread(file.path(DIAG, "pathological_sites_w5kb_pc3_auto_noUCcalled.txt"), sep = "\t")

GRP <- c("problematic (significant in real AND permuted)", "matched control")
for (d in list(par, bins, obs)) d[, group := factor(group, levels = GRP)]
NPROB <- par[group == GRP[1], .N]
NCTRL <- par[group == GRP[2], .N]
PHI_CUT <- 0.99
ord <- par[order(group, -phi)]$sid
for (d in list(par, bins, obs)) d[, sidf := factor(sid, levels = ord)]
```

# 1. How many problematic sites, and what makes them problematic

**Definition.** A site is *problematic* if it returns an extreme p-value in the
**permuted** data as well as the real data. Permutation destroys any
genotype–phenotype relationship, so a site that stays extreme there is not
reporting association — it is reporting something about its own phenotype
distribution that the test mistakes for signal. Selection is made on the
permuted scans only, which keeps the look at the real data non-circular:

- minimum p < 10⁻¹⁰ in at least one of the 5 permutations, **and**
- real −log₁₀p > 10

**Configuration.** ±5 kb windows, k = 3 covariate PCs, chr1–22 only, phenotype
`fold5_union_auto_noUCcalled` — U/C-variant sites already excluded (1,118 of
11,856), φ-boundary sites deliberately **retained** so they can be examined here.

> **There are `r NPROB` problematic sites** out of 10,045 tested.

`r NCTRL` well-behaved sites, matched on covered cell lines and pooled %pU, are
carried through every panel below for contrast.

```{r counts}
knitr::kable(pl[, .(
  site = paste0("chr", sub(" ", ":", sid)),
  `real -log10 p` = round(real, 1),
  `permuted, mean` = round(perm_mean, 1),
  `permuted, sd` = round(perm_sd, 2),
  `real minus worst permuted` = round(real_excess_over_perm_max, 2),
  `covered cell lines` = NA_integer_,
  `median depth` = median_depth_covered,
  `pooled %pU` = round(100 * pooled_pu, 1))][
    , `covered cell lines` := par$n_cl[match(pl$sid, par$sid)]][],
  caption = "The signature: the real and permuted p-values are essentially identical. For 6 of the 9 the real scan is no more extreme than the WORST permutation (negative last column), i.e. shuffling the phenotype cost nothing at all.")
```

The last column is the point. If these were genuine associations, permuting the
phenotype would destroy them and the real value would tower over the permuted
ones. Instead the real and permuted values agree to within a fraction of a log
unit, and for `r sum(pl$real_excess_over_perm_max < 0)` of the `r NPROB` sites
the real scan is *less* extreme than the worst permutation.

# 2. %pU by genotype at the lead cis-SNP

```{r box, fig.height=7.6}
o <- obs[!is.na(geno)]
o[, gf := factor(geno, levels = c(0, 1, 2))]
lab <- unique(o[, .(sidf, group, lead_snp, lead_p)])
lab[, panel := sprintf("chr%s   %s\np = %.1e", sub(" ", ":", sidf), lead_snp, lead_p)]
o <- merge(o, lab[, .(sidf, panel)], by = "sidf")
o[, panel := factor(panel, levels = lab[order(match(sidf, ord))]$panel)]

ggplot(o, aes(gf, 100 * pu)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, colour = "grey35") +
  geom_jitter(aes(colour = log10(n)), width = 0.15, height = 0, size = 1.9) +
  scale_colour_viridis_c(option = "D", name = "log10 read depth") +
  facet_wrap(~ panel, ncol = 4, scales = "free_y") +
  labs(x = "genotype (copies of the counted allele)", y = "%pU",
       title = sprintf("%%pU by genotype at each site's lead cis-SNP (%d problematic + %d controls)",
                       NPROB, NCTRL),
       subtitle = paste("Dots are cell lines, coloured by read depth. Free y-axes.",
                        "At most problematic sites the values are pressed against 0 and 100",
                        "\nwith little in between -- the phenotype is bimodal rather than graded.")) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", strip.text = element_text(size = 7.6))
```

```{r shape}
knitr::kable(obs[, .(
  site = paste0("chr", sub(" ", ":", sid[1])),
  group = as.character(group[1]),
  `cell lines` = .N,
  `frac %pU = 0` = sprintf("%.2f", mean(pu == 0)),
  `frac %pU > 0.9` = sprintf("%.2f", mean(pu > 0.9)),
  `frac in 0.1-0.9` = sprintf("%.2f", mean(pu >= 0.1 & pu <= 0.9)),
  `median depth` = as.numeric(median(n))), by = sidf][
    order(match(sidf, ord))][, -1],
  caption = "The same thing numerically: how much of each site's %pU distribution sits at the extremes versus in the middle.")
```

# 3. Estimated φ

```{r phi-tab}
knitr::kable(par[order(match(sid, ord)), .(
  site = paste0("chr", sub(" ", ":", sid)),
  group = as.character(group),
  `cell lines` = n_cl,
  phi = sprintf("%.6f", phi),
  `alpha+beta = (1-phi)/phi` = signif(alpha_plus_beta, 3),
  `phi, intercept-only fit` = sprintf("%.6f", phi_intercept_only),
  `pooled %pU` = round(100 * pooled_pu, 1))],
  caption = "phi is the over-dispersion of the Beta prior, on (0,1). phi -> 1 means alpha+beta -> 0, where the prior collapses onto point masses at 0 and 1.")
```

```{r phi-fig, fig.height=4.2}
ggplot(par, aes(reorder(paste0("chr", sub(" ", ":", sid)), phi), phi, fill = group)) +
  geom_col(width = 0.68) +
  geom_hline(yintercept = PHI_CUT, linetype = "dashed", colour = "grey25") +
  annotate("text", x = 0.7, y = PHI_CUT, label = " phi = 0.99", hjust = 0, vjust = -0.6,
           size = 3.2, colour = "grey25") +
  scale_fill_manual(values = setNames(c("#e34948", "#3987e5"), GRP), name = NULL) +
  coord_flip() +
  labs(x = NULL, y = expression("fitted"~varphi),
       title = sprintf("All %d problematic sites have phi pinned to its upper boundary", NPROB),
       subtitle = sprintf("Every one sits at %.6f, the numerical cap in fit_bb_mle. Genome-wide only ~0.2%% of sites do.",
                          max(par$phi))) +
  theme_bw(base_size = 11) + theme(legend.position = "bottom")
```

**`r par[group == GRP[1] & phi > PHI_CUT, .N]` of `r NPROB` problematic sites have
φ > 0.99** — in fact every one is at 0.999999, the numerical cap. This is not a
threshold that was tuned to catch them; genome-wide only about 0.2% of sites
reach it.

Note the `intercept-only` column: at 3 of the sites φ is well below the boundary
when the covariates are dropped (0.51, 0.86, 0.92), so at those the PCs are part
of what drives the fit onto the boundary. At the other 6 it is at the cap either
way.

# 4. The fitted prior distribution of $p_i$

$p_i$ is cell line $i$'s true modification rate. It is latent — never observed —
but the fitted null model states its whole distribution, and that distribution is
what is plotted. It is obtained from the fit in three steps:

1. **The mean.** The fitted coefficients give
   $\mathrm{logit}(\mu_i) = \hat\beta_0 + z_i^\top\hat\gamma$ (no genotype term,
   since this is the null), so $\mu_i = \mathrm{logit}^{-1}(\cdot)$. Note
   $\mu_i = E[p_i]$.
2. **The shape parameters**, from that mean and the site's single fitted
   $\hat\varphi$:
   $$\alpha_i = \mu_i\frac{1-\hat\varphi}{\hat\varphi}, \qquad
     \beta_i = (1-\mu_i)\frac{1-\hat\varphi}{\hat\varphi}$$
3. **The distribution.** $p_i \sim \text{Beta}(\alpha_i, \beta_i)$, evaluated as
   exact binned mass (a density curve is unusable: at $\hat\varphi \to 1$ it
   diverges at 0 and 1):
   $$P(a < p_i \le b) = F_{\text{Beta}}(b;\alpha_i,\beta_i) - F_{\text{Beta}}(a;\alpha_i,\beta_i)$$

$\hat\varphi$ is one value per site and the covariates make $\mu_i$ differ only
slightly between cell lines (range printed on each panel), so this is effectively
one Beta per site. The ticks below each panel are the observed $y_i/n_i$, for
reference.

```{r prior, fig.height=7.6}
# One Beta prior per site, from that site's phi and median mu -- computed here
# straight from case_params, no simulation and no averaging over cell lines.
NB  <- 50
brk <- seq(0, 1, length.out = NB + 1)
pri <- par[, {
  ab <- (1 - phi) / phi
  a  <- mu_median * ab;  b <- (1 - mu_median) * ab
  .(bin_mid = (head(brk, -1) + tail(brk, -1)) / 2, mass = diff(pbeta(brk, a, b)))
}, by = .(sid, group)]

labp <- par[, .(sid, panel = sprintf("chr%s   phi = %.6f\nmu_i %.3f-%.3f   P(<.01)=%.2f  P(>.99)=%.2f",
                                     sub(" ", ":", sid), phi, mu_min, mu_max,
                                     P_below_0.01, P_above_0.99))]
pri <- merge(pri, labp, by = "sid")
pri[, panel := factor(panel, levels = labp[order(match(sid, ord))]$panel)]
ob  <- merge(obs[, .(sid, pu)], labp, by = "sid")
ob[, panel := factor(panel, levels = levels(pri$panel))]

ggplot(pri, aes(bin_mid, mass)) +
  geom_col(aes(fill = group), width = 1 / NB, colour = NA) +
  geom_point(data = ob, aes(x = pu, y = -0.06), shape = 124, size = 2.2,
             colour = "grey15", alpha = 0.7) +
  scale_fill_manual(values = setNames(c("#e34948", "#3987e5"), GRP), name = NULL) +
  facet_wrap(~ panel, ncol = 4) +
  labs(x = expression(p[i]), y = "prior probability mass",
       title = "Fitted prior on the latent rate, with the observed %pU beneath",
       subtitle = paste("Bars: the site's Beta prior (exact, sums to 1). Ticks below zero: the observed %pU,",
                        "one per cell line (these are y_i/n_i, so they carry sampling noise).",
                        "\nEvery problematic site's prior puts ALL its mass at 0 and 1. At chr3:37277733, chr6:29943223 and",
                        "\nchr17:36212383 well-covered cell lines nonetheless sit in the interior, where the prior gives ~zero mass.")) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", strip.text = element_text(size = 7.4))
```

```{r prior-tab}
knitr::kable(par[order(match(sid, ord)), .(
  site = paste0("chr", sub(" ", ":", sid)),
  group = as.character(group),
  alpha = signif(alpha_median, 3), beta = signif(beta_median, 3),
  `mu_i range` = sprintf("%.3f - %.3f", mu_min, mu_max),
  `P(p<0.01)` = round(P_below_0.01, 3),
  `P(p>0.99)` = round(P_above_0.99, 3),
  `both tails` = round(P_below_0.01 + P_above_0.99, 3),
  `prior SD` = round(prior_sd_median, 3))],
  caption = "At the problematic sites the two tail masses sum to 1.000 -- the prior has NO interior mass -- and the prior SD is ~0.35-0.50 against a maximum of 0.5 attainable on (0,1).")
```

**The two tail masses sum to 1.000 at every problematic site.** The fitted prior
places no probability anywhere between 0.01 and 0.99: the model asserts that each
cell line's true pU rate is either essentially 0 or essentially 1, with no
intermediate stoichiometry possible.

<!-- # 5. What this says -->

<!-- Putting the three pieces together, the same story reads consistently from the -->
<!-- data side and the model side: -->

<!-- - **Data (§2)**: at these sites %pU is bimodal — cell lines sit at 0 or near 100 -->
<!--   with little in between — rather than varying smoothly with genotype. -->
<!-- - **Dispersion (§3)**: the fitter responds by driving φ to its upper boundary, -->
<!--   all `r NPROB` of `r NPROB` sites, at the numerical cap. -->
<!-- - **Prior (§4)**: at that boundary $\alpha+\beta \to 0$ and the Beta prior -->
<!--   degenerates to point masses at 0 and 1, which is the likelihood-maximising -->
<!--   description of a bimodal phenotype. -->

<!-- An LRT computed against a degenerate null is not testing what it appears to test, -->
<!-- which is why these sites stay extreme under permutation: the statistic is -->
<!-- responding to the *shape* of the phenotype distribution, which shuffling does not -->
<!-- change, rather than to any genotype–phenotype relationship. -->

<!-- This is why the scan excludes them, and why the durable fix is a **two-sided φ -->
<!-- guard inside `fit_bb_mle`** rather than a site exclusion list — see -->
<!-- [the covariate-count page](qtl_mapping_betabinomial_npcs_tradeoff.html) and -->
<!-- [the filtering strategy](qtl_mapping_betabinomial_strategy.html). -->

<!-- # Caveats -->

<!-- - **The controls are matched, not pristine.** They were chosen for similar -->
<!--   coverage and pooled %pU, and one of them (`chr15:48430687`, φ = 0.165, -->
<!--   α ≈ 0.037) is itself moderately dispersed with substantial mass near 0. It is a -->
<!--   fair contrast for coverage but not an example of an ideal fit. -->
<!-- - **`p_i` is latent.** These are the model's *prior* distributions, not estimates -->
<!--   of any cell line's rate. The posterior would add the observed counts: -->
<!--   $p_i \mid y_i \sim \text{Beta}(\alpha_i + y_i,\ \beta_i + n_i - y_i)$. -->
<!-- - **One lead SNP per site**, chosen as the minimum-LRT cis-SNP within ±5 kb at -->
<!--   MAF ≥ 0.05. At a site whose p-value is an artefact, "lead SNP" means the SNP -->
<!--   that best fits the artefact, not a candidate causal variant. -->
<!-- - **φ is on (0,1) here**, defined as $1/(\alpha+\beta+1)$ — matching `aod` and -->
<!--   VGAM's `rho`, not the `betareg` precision convention, which runs the opposite -->
<!--   way. -->

<!-- Related: [beta-binomial model](qtl_mapping_betabinomial_model.html) · -->
<!-- [covariate count and the φ boundary](qtl_mapping_betabinomial_npcs_tradeoff.html) · -->
<!-- [filtering strategy and calibration](qtl_mapping_betabinomial_strategy.html) -->
