---
title: "Does a near-zero dispersion estimate create false pU-QTLs?"
author: "XSun"
date: "2026-09-29"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 2
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 10)
BASE <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl/qtl_betabinomial/null_simulation"
S <- fread(file.path(BASE, "null_sites.txt.gz"))
P <- fread(file.path(BASE, "null_pvals.txt.gz"))
SCEN <- c("phi_0.0126", "phi_spread", "phi_0.3")
SLAB <- c("Fixed phi = 0.0126", "Variable phi", "Fixed phi = 0.3")
S[, scenario_label := factor(SLAB[match(scenario, SCEN)], levels = SLAB)]
S[, collapsed := fifelse(is.finite(phi_hat) & phi_hat < phi_true / 10,
                         "Near-zero fit", "Other fit",
                         na = NA_character_)]
P <- merge(P, S[, .(scenario, sid, scenario_label, collapsed)],
           by = c("scenario", "sid"), all.x = TRUE)
P <- P[is.finite(p_bb) & p_bb >= 0 & p_bb <= 1]
P[, scenario_label := factor(scenario_label, levels = SLAB)]
P[, collapsed := factor(collapsed, levels = c("Near-zero fit", "Other fit"))]
site_min <- P[, .(min_p = min(p_bb), tests = .N),
              by = .(scenario, sid, scenario_label, collapsed)]
site_min[, hit := min_p < 0.05 / tests]
site_rate <- site_min[, .(sites = .N, hits = sum(hit), rate = mean(hit)),
                      by = .(scenario, scenario_label, collapsed)]
site_rate[, scenario_label := factor(scenario_label, levels = SLAB)]
site_rate[, collapsed := factor(collapsed, levels = c("Near-zero fit", "Other fit"))]
setorder(site_rate, scenario_label, collapsed)
fmt_n <- function(x) format(x, big.mark = ",", trim = TRUE, scientific = FALSE)
```

In the [dispersion-recovery analysis](qtl_mapping_betabinomial_phi_estimation.html),
we found that the estimated beta-binomial dispersion (phi) can be close to zero
at a site even when the true phi is positive. Xin's question is whether those
sites then produce **false-positive pU-QTLs**. We tested that directly by
simulating counts with **no genotype effect** and running the production QTL
test.

**Main result:** At true phi = 0.0126, the sites with a near-zero fitted phi
had *fewer* significant QTLs than the nominal 5% site-level benchmark:
**1.35%** (62 of 4,595 testable sites). The other sites were at **5.32%**
(265 of 4,982). A near-zero phi estimate therefore did **not** produce the
false-positive excess we feared in this simulation. The beta-binomial test
still showed some excess in the aggregate small-p tail, especially in the
large-phi scenario; that deserves separate investigation.

# How the null was made

For each eligible pU site, we retained its real per-cell-line read depths and
the mean pU probabilities fitted under the *null* model with an intercept and
three phenotype PCs. We simulated beta-binomial counts at a known phi, without
using genotype to generate the counts. We then tested every real cis-SNP within
5 kb using the production beta-binomial likelihood-ratio test: three
optimization starts, maximum likelihood, and no phi prior.

There were three settings: phi fixed at **0.0126**, phi varying across sites
with a log-normal distribution centred at **0.0126**, and phi fixed at **0.3**.
Each setting used 10,051 eligible sites and 320,967 attempted SNP tests.
We call a fit *near-zero* when its fitted null phi is less than one-tenth of
the known true phi. This includes small fitted values above the numerical
floor; it is a practical threshold, not a separate biological class.

The simulation was run by 43.null_simulation_fp.R through
run_null_simulation.sbatch. Its output tables are in the project's
pumseq_processed/psi_qtl/qtl_betabinomial/null_simulation/ directory.

# False positives per site

For each site with at least one valid beta-binomial test, we asked whether its
smallest cis-SNP p-value passed a within-site Bonferroni threshold of
0.05 / number of valid tests. Under a well-calibrated null, 5% is an
**upper bound** on this site-level rate, not an exact target, because cis-SNPs
are correlated.

```{r site-rates, fig.width=9, fig.height=4.8}
ggplot(site_rate, aes(scenario_label, 100 * rate, fill = collapsed)) +
  geom_hline(yintercept = 5, linetype = "dashed", colour = "grey40") +
  geom_col(position = position_dodge(width = 0.75), width = 0.67) +
  geom_text(aes(label = sprintf("%.2f%%\n(%s sites)", 100 * rate, fmt_n(sites))),
            position = position_dodge(width = 0.75), vjust = -0.15,
            size = 3.3, lineheight = 0.95) +
  scale_fill_manual(values = c("Near-zero fit" = "#276f9e",
                               "Other fit" = "#da8243"), name = NULL) +
  scale_y_continuous(limits = c(0, 7.4), breaks = 0:7,
                     expand = expansion(mult = c(0, 0))) +
  labs(x = NULL, y = "Sites with a significant cis-SNP (%)",
       title = "Near-zero phi fits do not show excess false-positive sites",
       subtitle = "Site-level threshold: best valid cis-SNP p < 0.05 / number of valid tests") +
  theme_bw(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```

```{r site-table}
tab <- copy(site_rate)
tab[, scenario := as.character(scenario_label)]
tab[, fit := as.character(collapsed)]
tab[, fraction := sprintf("%d / %s", hits, fmt_n(sites))]
tab[, percent := sprintf("%.2f%%", 100 * rate)]
knitr::kable(tab[, .(Scenario = scenario, "Null phi fit" = fit,
                       "Significant / testable sites" = fraction,
                       "Site-level false positives" = percent)],
             caption = "A site is testable here if it has at least one finite beta-binomial p-value.")
```

At the low-phi setting, near-zero fits occurred in 4,632 of 10,051 eligible
sites. A near-zero estimate is therefore common, but it is not where the
excess false positives are concentrated.

# The SNP-level p-value tail

The site-level result does not mean that every part of the beta-binomial null
is perfectly calibrated. A uniform p-value distribution would follow the
diagonal in the QQ plots. Points above the diagonal indicate more small p-values
than expected. These plots use every *valid* beta-binomial test; only plotting
points are thinned.

```{r qq, fig.width=12, fig.height=5.4}
qq_one <- function(d) {
  x <- sort(d$p_bb)
  n <- length(x)
  ix <- unique(as.integer(round(exp(seq(0, log(n), length.out = 700)))))
  data.table(expected = -log10((ix - 0.5) / n),
             observed = -log10(pmax(x[ix], .Machine$double.xmin)))
}
qq <- P[, qq_one(.SD), by = .(scenario_label, collapsed),
        .SDcols = "p_bb"]
ggplot(qq, aes(expected, observed, colour = collapsed)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey50",
              linetype = "dashed", linewidth = 0.4) +
  geom_line(linewidth = 0.75) +
  facet_wrap(~ scenario_label, nrow = 1) +
  scale_colour_manual(values = c("Near-zero fit" = "#276f9e",
                                 "Other fit" = "#da8243"), name = NULL) +
  coord_cartesian(xlim = c(0, 6), ylim = c(0, 9)) +
  labs(x = "Expected -log10(p)", y = "Observed -log10(p)",
       title = "Valid beta-binomial SNP tests under a true genotype null",
       subtitle = "Dashed line: uniform p-values; curves are grouped by the fitted null phi") +
  theme_bw(base_size = 10.5) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```

```{r tail-table}
lambda_bb <- function(p) median(qchisq(p, 1, lower.tail = FALSE)) / qchisq(0.5, 1)
summarize_tail <- function(d, groups) {
  d[, .(tests = .N, lambda = round(lambda_bb(p_bb), 3),
        tail_rate = 1e4 * mean(p_bb < 1e-4)), by = groups]
}
tail <- rbindlist(list(summarize_tail(P, "scenario")[, Group := "All valid tests"],
                       summarize_tail(P, c("scenario", "collapsed"))[
                         , Group := as.character(collapsed)][, collapsed := NULL]),
                  fill = TRUE)
tail[, Scenario := SLAB[match(scenario, SCEN)]]
tail <- tail[order(match(Scenario, SLAB),
                   match(Group, c("All valid tests", "Near-zero fit", "Other fit")))]
tail[, tail_rate := sprintf("%.2fx", tail_rate)]
knitr::kable(tail[, .(Scenario, Group, "SNP tests" = fmt_n(tests),
                       "Median inflation (lambda)" = lambda,
                       "Observed / expected p<1e-4" = tail_rate)],
             caption = "SNP-level tail rates are among valid beta-binomial tests. With 300,000 tests, a uniform null predicts about 30 p-values below 1e-4.")
```

At true phi = 0.0126, the near-zero group has **0.16 times** the expected
number of p-values below 1e-4, while the other group has **3.25 times**.
Across all valid tests at that phi, the ratio is **1.93 times**. At true phi =
0.3, the overall ratio rises to **6.51 times**. Thus the near-zero fits are
*not* the apparent source of the small-p excess, but the excess itself should
not be ignored.

# Fit failures and scope

```{r exclusions}
ex <- S[, .("Eligible sites" = .N,
             "Null-fit failures" = sum(!is.finite(phi_hat)),
             "Attempted SNP tests" = sum(n_tests),
             "Dropped SNP tests" = sum(dropped)), by = scenario]
ex[, Scenario := SLAB[match(scenario, SCEN)]]
ex[, "Dropped test fraction" := sprintf("%.1f%%",
                                       100 * get("Dropped SNP tests") / get("Attempted SNP tests"))]
knitr::kable(ex[, .(Scenario, "Eligible sites" = fmt_n(get("Eligible sites")),
                    "Null-fit failures" = fmt_n(get("Null-fit failures")),
                    "Attempted SNP tests" = fmt_n(get("Attempted SNP tests")),
                    "Dropped SNP tests" = fmt_n(get("Dropped SNP tests")),
                    "Dropped test fraction" = get("Dropped test fraction"))],
             caption = "Failure counts are included so the rates above are not mistaken for all attempted tests.")
```

The false-positive rates above condition on a valid p-value. Roughly 10% of
attempted tests were dropped in the low-phi settings, and 18% at phi = 0.3.
A failed null fit has no p-value and cannot be counted as a significant QTL.
This simulation uses one count realization per site and scenario, so a small
difference from 5% should not be overinterpreted. It tests genotype-free data
under the fitted beta-binomial mean structure and the observed coverage and
cis-SNP patterns. It does not address every form of model misspecification in
real pU counts.

The practical answer to Xin's question is that **phi collapse itself did not
generate excess false positives in these null simulations**. We still need to
understand the excess among non-collapsed fits, particularly when true phi is
large, before describing the full QTL test as calibrated under all conditions.
