---
title: "Beta-binomial pU-QTL: is the overdispersion really that small?"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 3
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 10)

PSI <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl"
LP  <- file.path(PSI, "qtl_betabinomial/low_phi_validation")
RC  <- file.path(PSI, "qtl_betabinomial/phi_recovery")
MLE_PEAK <- 0.0126; PHI_EPS <- 1e-6
```

# 1. The problem

The mode of the fitted overdispersion across sites is **φ ≈ 0.0126**, and that
number carries weight: it is the reference value the
[φ-selection heuristic](qtl_mapping_betabinomial_optimiser.html#7_Choosing_φ_when_it_is_not_identifiable)
aims at. It looks implausibly small. At our median depth of 8 reads it inflates
the variance of y/n by only about 10% over a binomial, and **79.2% of sites
cannot reject φ = 0 at all**. The natural suspicion is that 0.0126 describes our
*estimator* — a nearly flat likelihood with a hard boundary at zero — rather than
the dispersion in the data.

This page answers that objection in two steps. **§2 establishes that the
estimate is correct** — three checks with non-overlapping failure modes, one of
which uses no likelihood at all. **§3 then asks what the number means**: why so
small a value is the expected answer rather than a suspicious one, why
dispersion still has to be modelled when most sites cannot distinguish it from
zero, and what such a site looks like in the raw counts. §4 states what can and
cannot be concluded.

> **Every φ on this page is a 3-start estimate.** The fitter runs the optimiser
> from φ₀ ∈ {0.01, 0.1, 0.5} and keeps the highest-likelihood solution
> (`BB_PHI_STARTS=0.01,0.1,0.5`, `BB_PHI_SELECT=loglik`) — the plain MLE

# 2. Confirming the estimate is correct

Before asking what φ ≈ 0.013 means, we have to establish that it *is* the
number — that it describes the data rather than the machinery that produced it.
Three checks, deliberately chosen so their failure modes do not overlap: one
that tests the estimator against a known truth, one that avoids the likelihood
entirely, and one that bypasses estimation altogether and asks only what the raw
scatter permits.

## 2.1 Does the estimator recover a known φ?

The cleanest test is to generate data where the answer is known. Per site, take
the **real depths n_i** and the **real fitted mean vector μ_i** from the k = 3
null (μ_i varies across cell lines because logit μ_i = β₀ + z_iᵀγ), then for a
grid of **true** φ simulate

$$p_i \sim \mathrm{Beta}\!\left(\mu_i\tfrac{1-\varphi}{\varphi},\ (1-\mu_i)\tfrac{1-\varphi}{\varphi}\right), \qquad y_i \sim \mathrm{Binomial}(n_i, p_i)$$

and refit with the same multi-start fitter and the same covariates. The simulated
data therefore face exactly the estimation problem the real data pose.

```{r rectab}
RS <- fread(file.path(RC, "recovery_summary.tsv"), sep = "\t")
knitr::kable(RS, caption = "Recovery of a known phi at our real depths. The MODE is what matters, because it is the statistic the heuristic uses.")
```

**The mode is very nearly unbiased from 0.0126 upwards** — within 3% at 0.0126
itself and within 10% across the whole range up to 0.4. Only at a true φ of
0.005 does it collapse to the floor, which is below what our depth can resolve.

```{r recfig, fig.height=7, fig.width=13}
RB <- fread(file.path(RC, "recovery_by_site.txt.gz"), sep = "\t")[is.finite(phi_hat)]
PT <- sort(unique(RB$phi_true))
RB[, panel := factor(sprintf("true phi = %g", phi_true),
                     levels = sprintf("true phi = %g", PT))]
# two reference lines per panel: the phi used to SIMULATE, and the mode of what
# came back. Their agreement is the whole result, so nothing else is marked.
vl <- melt(merge(unique(RB[, .(panel, phi_true)]),
                 RS[, .(phi_true, mode = `mode phi_hat`)], by = "phi_true"),
           id.vars = c("panel", "phi_true"), measure.vars = c("phi_true", "mode"),
           variable.name = "what", value.name = "x")
vl[, what := factor(ifelse(what == "phi_true", "true phi (simulated)",
                           "mode of the refitted phi"),
                    levels = c("true phi (simulated)", "mode of the refitted phi"))]
ggplot(RB, aes(phi_hat)) +
  geom_histogram(bins = 60, fill = "#4a7fbf", colour = NA) +
  geom_vline(xintercept = PHI_EPS, colour = "grey35", linetype = "dotted") +
  geom_vline(data = vl, aes(xintercept = x, colour = what, linetype = what),
             linewidth = 0.7) +
  facet_wrap(~ panel, nrow = 2, scales = "free_y") +
  scale_x_log10(breaks = 10^seq(-6, 0, 2), labels = c("1e-6", "1e-4", "1e-2", "1")) +
  scale_colour_manual(values = c("true phi (simulated)" = "#c0392b",
                                 "mode of the refitted phi" = "#1f4e9c"), name = NULL) +
  scale_linetype_manual(values = c("true phi (simulated)" = "solid",
                                   "mode of the refitted phi" = "longdash"), name = NULL) +
  labs(x = expression(hat(varphi)*", axis log-spaced (ticks are "*hat(varphi)*" itself)"),
       y = "sites",
       title = "Simulated from a KNOWN phi, then refitted",
       subtitle = paste(
         "Red = the phi used to simulate. Blue dashed = the mode of the phi that came back. The closer the two, the better the recovery.",
         "Dotted grey = the PHI_EPS = 1e-6 numerical floor.",
         "Note the 'true phi = 0.0126' panel: ONE true value produces a bimodal estimate distribution with a pile at the floor.",
         sep = "\n")) +
  theme_bw(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        plot.subtitle = element_text(size = 8))
```

Two further results fall out of this, both of which correct earlier readings of
the real φ distribution.

**The low-end bimodality is an artefact of the estimator.** Look at the
`true phi = 0.0126` panel: a *single* true value produces both a pile at the
floor and a clean hump at the truth. So the floor pile and the ~46% of real
sites below 10⁻⁴ are not a distinct population of near-binomial sites — they are
what this estimator does to low-φ sites at 8 reads. An earlier version of the
optimiser page described these as two regimes; that reading was wrong.

**The median is badly biased; only the mode is usable.** At a true φ of 0.0126
the median estimate is 0.0023, five-fold low, because half the estimates fall
into the artefact pile. This is why the reference value must be a mode and not
an average — a point in favour of how the heuristic was constructed.

**But an unbiased mode does NOT mean accurate per-site estimates, and the gap is
large.** The mode is a property of the distribution; individual sites are
frequently wrong by orders of magnitude. Scored against the known truth:

```{r recacc}
RB2 <- fread(file.path(RC, "recovery_by_site.txt.gz"), sep = "\t")[is.finite(phi_hat)]
knitr::kable(RB2[, { r <- phi_hat/phi_true[1]
  .(sites = .N,
    `within 2x of truth`    = sprintf("%.1f%%", 100*mean(r > 0.5 & r < 2)),
    `within 10x`            = sprintf("%.1f%%", 100*mean(r > 0.1 & r < 10)),
    `more than 10x TOO LOW` = sprintf("%.1f%%", 100*mean(r < 0.1)),
    `at the 1e-6 floor`     = sprintf("%.1f%%", 100*mean(phi_hat < 2e-6)),
    `median fold error`     = round(median(pmax(r, 1/r)), 1)) },
  by = .(`true phi` = phi_true)][order(`true phi`)],
  caption = "Per-site accuracy against known truth. Accuracy is strongly phi-dependent: phi is only reliably estimated per site when it is LARGE, and the modal value is precisely where it is not.")
```

At the truth that matters, φ = 0.0126, only **29.0%** of sites land within 2× of
it, **48.2% are more than tenfold too low**, one site in seven collapses to the
floor, and the median fold error is **9×**. Accuracy climbs steadily with φ —
within-2× goes 18% → 29% → 52% → 62% → 71% → 79% as the truth rises from 0.005
to 0.4, with median fold error falling from 461× to 1.3×.

Since the real φ distribution is dominated by low-φ sites, the same must hold
there: **most real per-site φ estimates are wrong by a large factor.** This is
the same fact as "79.2% of sites cannot reject φ = 0", stated concretely.

It does not, however, undermine the QTL results. The LRT compares null against
full *at the same site* with φ free in both, so where the likelihood is flat both
fits slide together and the statistic is largely insensitive to where φ lands.
And that is measured rather than assumed: the permutation calibration gives
λ = 1.03–1.06 with zero problematic sites at k = 3 (§3.2). A badly-estimated
nuisance parameter inside a test verified on permuted data is tolerable; a
badly-estimated *reported quantity* is not. See §4.

Simulating a single true φ = 0.0126 reproduces the real data on every summary:

```{r recmatch}
knitr::kable(data.table(
  statistic = c("mode", "% at the 1e-6 floor", "% below 1e-4", "% identified (reject phi=0)"),
  `real data` = c("0.0126", "16.3%", "49.1%", "20.8%"),
  `simulated, truth = 0.0126` = c(
    sprintf("%.4f", RS[phi_true == 0.0126]$`mode phi_hat`),
    RS[phi_true == 0.0126]$`% at 1e-6 floor`,
    RS[phi_true == 0.0126]$`% below 1e-4`,
    RS[phi_true == 0.0126]$`% identified`)),
  caption = "A single true phi of 0.0126 reproduces the real distribution's mode, its boundary pile-up, its sub-1e-4 mass and its identification rate.")
```

## 2.2 An estimator with no likelihood in it

The recovery simulation still relies on `fit_bb_mle`. A fully independent check
is possible because φ enters the model through only one quantity, the design
effect:

$$\mathrm{Var}(y_i/n_i) = \frac{p(1-p)}{n_i}\bigl[1 + (n_i-1)\varphi\bigr]$$

So a Pearson χ² of the counts against their fitted mean recovers φ by moments.
With m covered cell lines and q fitted mean parameters,

$$S = \sum_i \frac{(y_i - n_i\hat\mu_i)^2}{n_i\hat\mu_i(1-\hat\mu_i)}, \qquad \mathbb{E}[S] \approx (m-q) + \varphi\sum_i (n_i-1), \qquad \hat\varphi_{\text{mom}} = \frac{S-(m-q)}{\sum_i (n_i-1)}$$

This is the classic overdispersion moment estimator (the construction behind
Tarone 1979). It is **closed form — no likelihood, no optimiser, no starting
value** — so it cannot suffer the bimodal-likelihood failure that made the
multi-start fix necessary in the first place.

```{r momtab}
MS <- fread(file.path(LP, "moment_summary.tsv"), sep = "\t")
knitr::kable(MS, caption = "Two moment variants and the MLE. NB the MLE row's mode is taken over ALL sites, which is the 2.4e-6 floor artefact; the 0.0126 peak comes from the subset where all three optimiser starts agree. The moment estimators need no such subsetting.")
```

**The moment estimators peak at 0.0146 and 0.0152, against the MLE's 0.0126** —
agreement within 16–20% between methods whose failure modes have nothing in
common. That is the direct answer to "is the peak an artefact of your fitter?".

It also independently vindicates the decision to build the peak on the
reproducible subset. Computed over *all* sites with no filtering, the moment
estimators land near 0.015 — close to the subset-based MLE peak of 0.0126, and
four orders of magnitude away from the all-sites MLE mode of 2.4×10⁻⁶.

```{r momfig, fig.height=4.6, fig.width=13}
MB <- fread(file.path(LP, "moment_by_site.txt.gz"), sep = "\t")
PL <- rbind(data.table(phi = MB$phi_mle,          est = "MLE, multi-start"),
            data.table(phi = MB$phi_mom_pooled,   est = "moments, pooled mean"),
            data.table(phi = MB$phi_mom_adjusted, est = "moments, covariate-adjusted"))
PL <- PL[is.finite(phi) & phi > 0]
PL[, est := factor(est, levels = c("MLE, multi-start", "moments, pooled mean",
                                   "moments, covariate-adjusted"))]
mode_log10 <- function(x) { d <- density(log10(x)); 10^d$x[which.max(d$y)] }
md <- PL[, .(mode = mode_log10(phi)), by = est]
ggplot(PL, aes(phi)) +
  geom_histogram(bins = 60, fill = "#4a7fbf", colour = NA) +
  geom_vline(xintercept = MLE_PEAK, colour = "#2c8a4a", linewidth = 0.7) +
  geom_vline(data = md, aes(xintercept = mode), colour = "#c0392b",
             linetype = "longdash", linewidth = 0.6) +
  geom_vline(xintercept = PHI_EPS, colour = "grey35", linetype = "dotted") +
  facet_wrap(~ est, nrow = 1, scales = "free_y") +
  scale_x_log10(breaks = 10^seq(-6, 0, 2), labels = c("1e-6", "1e-4", "1e-2", "1")) +
  labs(x = expression(hat(varphi)*", axis log-spaced (ticks are "*hat(varphi)*" itself)"),
       y = "sites",
       title = "Three estimators of phi, only the first of which uses a likelihood",
       subtitle = paste(
         "Green solid = 0.0126, the MLE peak on the reproducible subset. Red dashed = each panel's OWN mode. Dotted = the 1e-6 floor.",
         "Only positive estimates are shown: the moment estimators return phi <= 0 at 34% and 39% of sites respectively (see caveats below).",
         sep = "\n")) +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank(), plot.subtitle = element_text(size = 8))
```

**Caveats, stated plainly.** The moment estimator returns φ ≤ 0 at 34% (pooled)
and 39% (adjusted) of sites, meaning the observed scatter was at or below
binomial — the moment analogue of the MLE's pile-up at the floor. Those cannot
be a true variance, so the mode is taken over positive values only, and that
selection biases it **upward**. Adjusting for covariates was expected to shrink
the gap to the MLE and did not — 0.0152 against the pooled 0.0146 — so the
selection, not the mean model, is the likely cause of the residual 16–20%. The
moment estimator is also less efficient than the MLE, so its per-site values are
noisier.

## 2.3 Would a larger φ fit the observed scatter?

Independently of any estimator, we can ask which φ reproduces the spread of %pU
actually observed across cell lines.

```{r spread}
SP <- fread(file.path(LP, "spread_check.tsv"), sep = "\t")
knitr::kable(SP[, .(phi = label, `median predicted SD`, `median observed SD`,
                    `predicted / observed`, `% sites over-predicted`)],
  caption = "A well-chosen phi should sit near predicted/observed = 1 and split the sites about 50/50.")
```

**A φ of 0.45 predicts nearly double the spread we observe** and over-predicts at
90% of sites. So a substantially larger value is excluded by the raw scatter,
with no model fitting involved.

Note honestly that at the *median* site φ = 0 fits marginally better than 0.0126
(ratio 1.016 versus 1.051). That is the same fact as most sites being
unidentifiable, and it is not an argument for the binomial — see §3.2.

## 2.4 Does the peak-proximity rule estimate φ any better?

Everything above uses the plain MLE. The [optimiser page](qtl_mapping_betabinomial_optimiser.html)
proposes a different rule for unidentifiable sites: of the optima the three
starts found, return the one nearest the population mode. On real data we can
only check whether that changes *calibration* — it does not. We cannot check
whether it gets φ **right**, because the truth is unknown.

In simulation the truth is known, so accuracy is measurable. The same simulated
datasets were refit under both rules, so differences are attributable to the
rule rather than to simulation noise. Note the reference is fixed at 0.0126
while the truth varies — so the rule is tested both where its reference is
correct (φ = 0.0126) and where it is 32× too low (φ = 0.4).

(A third variant, applying the peak only to break near-ties in likelihood, was
also run and came out indistinguishable from the rule below at every true φ, so
it is not shown.)

```{r ruleacc}
RA <- fread(file.path(RC, "rule_accuracy.tsv"), sep = "\t")[rule != "peak_tiebreak"]
RA[, rule := factor(rule, levels = c("loglik", "peak"))]
setorder(RA, phi_true, rule)
knitr::kable(RA[, .(`true phi` = phi_true, rule,
                    `within 2x` = `within 2x of truth`, `within 10x`,
                    `at the 1e-6 floor`, `median fold error`,
                    `mean |log10 error|`, `loglik lost vs MLE` = `total loglik lost vs MLE`)],
  caption = "Accuracy against KNOWN truth. 'loglik' is the MLE used everywhere else on this page. 'loglik lost' is summed over ~10,000 sites, so it is negligible per site.")
```

```{r rulefig, fig.height=5.6, fig.width=13}
RR <- fread(file.path(RC, "rule_accuracy_by_site.txt.gz"), sep = "\t")[
  is.finite(phi_hat) & rule != "peak_tiebreak"]
RUL <- c("loglik" = "MLE (highest likelihood)",
         "peak" = "peak: nearest 0.0126")
RR[, rl := factor(RUL[rule], levels = RUL)]
PT2 <- sort(unique(RR$phi_true))
RR[, pcol := factor(sprintf("true phi = %g", phi_true),
                    levels = sprintf("true phi = %g", PT2))]
# per-panel: the true value, the recovered mode, and the % stuck at the floor --
# the floor is where the three rules actually differ, so label it directly
ann <- RR[, .(frac_floor = mean(phi_hat < 2e-6),
              mode = { d <- density(log10(phi_hat)); 10^d$x[which.max(d$y)] },
              phi_true = phi_true[1]), by = .(rl, pcol)]
ggplot(RR, aes(phi_hat)) +
  geom_histogram(bins = 55, fill = "#4a7fbf", colour = NA) +
  geom_vline(xintercept = PHI_EPS, colour = "grey35", linetype = "dotted") +
  geom_vline(data = ann, aes(xintercept = phi_true), colour = "#c0392b", linewidth = 0.6) +
  geom_vline(data = ann, aes(xintercept = mode), colour = "#1f4e9c",
             linetype = "longdash", linewidth = 0.55) +
  geom_text(data = ann, aes(x = 1e-6, y = Inf, label = sprintf("%.1f%% at floor", 100 * frac_floor)),
            hjust = -0.05, vjust = 1.6, size = 2.7, colour = "grey20") +
  facet_grid(rl ~ pcol, scales = "free_y", labeller = label_wrap_gen(width = 22)) +
  scale_x_log10(breaks = 10^c(-6, -3, 0), labels = c("1e-6", "1e-3", "1")) +
  labs(x = expression(hat(varphi)*", axis log-spaced (ticks are "*hat(varphi)*" itself)"),
       y = "sites",
       title = "The same simulated data, refitted under each selection rule",
       subtitle = paste(
         "Red = the true phi. Blue dashed = the recovered mode. Dotted grey = the 1e-6 floor.",
         "Compare DOWN a column: the right-hand hump is essentially unchanged between rules -- the rules do not improve typical accuracy.",
         "What changes is the LEFT pile at the floor, which the peak rule largely clears (e.g. 13.6% -> 1.1% at true phi = 0.0126).",
         "That happens in every column, including true phi = 0.4 where the 0.0126 reference is 32x too low: it is a boundary rescue, not shrinkage toward the truth.",
         sep = "\n")) +
  theme_bw(base_size = 10) +
  theme(panel.grid.minor = element_blank(), strip.text = element_text(size = 8),
        plot.subtitle = element_text(size = 8))
```

**The rule does not improve accuracy in any conventional sense.** Within-2× and
within-10× are unchanged to a decimal place at every true φ — 29.8% vs 29.9% at
the modal truth, 79.4% vs 79.4% at φ = 0.4. If you want to know a site's φ to
within a factor of two, the rule buys you nothing.

**What it does is rescue boundary collapses.** The fraction of estimates pinned
at the 10⁻⁶ floor falls from 13.6% to 1.1% at the modal truth (and 20.7% → 1.6%
at φ = 0.005), which shows up as a ~20% reduction in mean |log₁₀ error|. So it
converts *catastrophic* errors — off by four orders of magnitude — into merely
*large* ones. That is a real improvement, and it costs a negligible amount of
likelihood (a few log units summed over ten thousand sites).

**And it helps even where its reference is badly wrong**, which was not what I
expected: at φ = 0.4, where 0.0126 is 32× too low, mean error still improves
(0.424 → 0.398). The reason is that the rule is not shrinking estimates toward
the truth — it is pulling them off the boundary, and 0.0126 is closer to 0.4
than 10⁻⁶ is. It is a boundary rescue that happens to work whether or not the
reference value is right, which is a weaker and more honest description of it
than "choosing a more reasonable φ".

**What this does not change:** at the modal truth, 47.7% of estimates remain
more than tenfold too low under the rule, against 48.1% under the MLE. Per-site
φ is not made reportable by either rule — see §4.

# 3. Exploring what φ means

The estimate stands. The harder question is what a dispersion of 0.013 is
actually saying about pseudouridylation — why so small a number is the expected
answer rather than a suspicious one, why dispersion still has to be modelled
when most individual sites cannot distinguish it from zero, and what such a site
looks like in the raw counts.

## 3.1 What a small φ actually means

```{r interp}
knitr::kable(fread(file.path(LP, "interpretation.tsv"), sep = "\t"),
  caption = "At the median site: pooled %pU = 4.6%, median depth 8 reads. alpha = mu(1-phi)/phi; when alpha < 1 the Beta prior is U-shaped, i.e. the model asserts pU levels pile up at 0% and 100% across cell lines.")
```

φ = 0.0126 says the true pU level varies by about **±2.4 percentage points**
across cell lines around a mean of 4.6% — modest but real biological variation.
It is also the only value in that table keeping **α > 1**, i.e. a sensible
unimodal distribution of pU levels. From φ = 0.05 upward α falls below 1 and the
model asserts that pU levels pile up at 0% and 100%, which is not what these
sites look like.

```{r budget}
knitr::kable(fread(file.path(LP, "depth_budget.tsv"), sep = "\t"),
  caption = "The depth budget: why a small phi is expected, and why individual sites cannot see it.")
```

**This is the whole explanation for the 79.2%.** At 8 reads, one %pU measurement
carries **±7.4 pp** of pure sampling noise while the biological signal is
**2.4 pp** — noise exceeds signal **3.1-fold per site**, and resolving it would
need roughly **79 reads per cell line**. A single site therefore cannot reject
φ = 0, which is a statement about our depth, not about whether dispersion exists.

So the answer to "is φ too small?" is that φ is small **because at this depth the
cell-line-to-cell-line variability in %pU is mostly read sampling.** The genuine
biological component is real but modest, and the estimator finds it correctly.

## 3.2 Then why model dispersion at all?

Because per-site non-rejection is a power statement, and the aggregate is not
remotely ambiguous.

```{r aggtab}
knitr::kable(fread(file.path(LP, "aggregate_test.tsv"), sep = "\t"),
  caption = "Testing phi = 0 GLOBALLY. NB phi = 0 lies on the boundary of the parameter space, so the LRT null is the mixture 0.5*chi2_0 + 0.5*chi2_1 (Chernoff 1954; Self & Liang 1987) and the 95% cut is a loglik drop of 1.353, not the 1.92 of a chi2_1.")
```

```{r idtab}
knitr::kable(fread(file.path(LP, "identifiability.tsv"), sep = "\t"),
  caption = "Per-site identifiability at both critical values. Earlier passes of this project used 1.92 and so reported 81.4% unidentifiable; the boundary-correct figure is 79.2%.")
```

Twenty-one percent of sites reject φ = 0 against 5% expected, the summed LRT is
**296× its null expectation**, and 11.7% of sites reject individually with a
log-likelihood drop above 10. And the decisive practical point — the binomial
model *is* the φ = 0 model, and we ran it:

```{r binom}
bl <- fread("/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl/qtl_binomial/lrt_calibration/calib_lrt_w5kb_pc3_lambda.txt")
bb <- fread(file.path(PSI, "qtl_betabinomial/lrt_calibration/calib_siteres_w5kb_pc3_auto_noUCcalled_MS_lambda.txt"))
knitr::kable(data.table(
  model = c("binomial (phi = 0 by construction)", "beta-binomial, same window and k"),
  `real lambda` = c(round(bl[dataset == "real"]$lambda, 3), round(bb[dataset == "real"]$lambda, 3)),
  `permutation lambda` = c(
    sprintf("%.2f - %.2f", min(bl[dataset != "real"]$lambda), max(bl[dataset != "real"]$lambda)),
    sprintf("%.3f - %.3f", min(bb[dataset != "real"]$lambda), max(bb[dataset != "real"]$lambda)))),
  caption = "Permuted data contain no genotype-phenotype relationship by construction, so a permutation lambda of 3 is pure false-positive inflation.")
```

Setting φ = 0 roughly **triples the median test statistic on data with no
signal**. The dispersion costs nothing at the 79% of low-φ sites, where φ̂ goes
small and the model reduces to near-binomial, and prevents catastrophe at the
minority where it is large. That minority is real, not artefact: a single true
φ = 0.0126 produces only 1.3% of sites above 0.128, whereas the real data have
12.3% — a ten-fold excess, 1,290 sites against 135.

## 3.3 What a low-φ site looks like

Everything above argues from summary statistics. This section looks at
individual sites, all drawn from the **reproducible subset** — the sites where
all three optimiser starts agree on φ to within 5%, i.e. the middle panel of the
φ-distribution figure and the set the peak is measured on.

Three groups: five sites bracketing the peak (φ ∈ [0.010, 0.016]), three with
low φ even within that subset (φ < 0.003), and three with high φ (> 0.3), where
α < 1 and the Beta prior turns U-shaped.

> **The high-φ group is labelled "artefact-enriched" for a reason.** All three
> lie in IGH or MHC, which in LCLs are somatically rearranged or
> hyperpolymorphic — regions where φ is 6.8× more likely to be high than
> elsewhere in the genome. They show what a large φ looks like; they are not a
> fair sample of large-φ *biology*.
>
> **Genotype is deliberately not plotted.** φ is estimated from the null fit,
> which contains no genotype term, so it describes spread across cell lines and
> nothing about any SNP. An earlier version of this section plotted %pU against
> the lead cis-SNP; that was misleading twice over — the lead SNP is the
> *most-associated* of 11–56 tested with no significance filter (the low-φ sites'
> nominal p-values matched chance exactly, KS p = 0.22), and because the null
> cannot see genotype, a strong cis effect *causes* a high φ rather than the
> reverse. Sites above φ = 0.3 carry a p < 10⁻⁵ cis signal 12.2% of the time
> against 0.0% below φ = 10⁻⁴, so that axis measured the confound, not biology.

> **These are deliberately the deepest available sites, not a random sample.** At
> the median site (8 reads) the sampling noise on one %pU measurement is ±7.4 pp
> against a biological signal of 2.4 pp, so nothing is visible by eye at typical
> depth — which is §3.1's point, not a counterexample to it. The sites below carry
> 1,800–8,100 reads per cell line, so binomial noise is negligible and the
> **observed** scatter is essentially the biological scatter. That is what makes
> them a fair test of whether φ means what we claim.

```{r sitetab}
LS <- file.path(PSI, "qtl_betabinomial/lowphi_sites")
P  <- fread(file.path(LS, "site_table.tsv"), sep = "\t")
GRP <- c("near the peak (phi ~ 0.0126)", "low (phi < 0.003)",
         "high, for contrast (phi > 0.3)")
P[, group := factor(group, levels = GRP)]
setorder(P, group, -phi)
knitr::kable(P[, .(site = paste0("chr", sub(" ", ":", sid)), group,
                   phi = signif(phi, 4), `alpha (median)` = round(alpha_median, 2),
                   n_cl, `median depth` = median_depth,
                   `pooled %pU` = round(100 * pooled_pu, 1),
                   `observed SD (pp)` = round(100 * obs_sd_pu, 1),
                   `SD implied by phi (pp)` = round(100 * prior_sd_pi, 1),
                   `lead cis-SNP` = lead_snp)],
  caption = "All 11 sites are covered in all 50 cell lines. 'SD implied by phi' is sqrt(mu(1-mu)phi), the Beta spread around each cell line's own mu_i.")
```

**The observed scatter tracks what φ predicts across a 260-fold range of φ** —
1.4 pp observed against 1.0 pp implied at φ = 0.0023, about 4 pp against 3 pp at
the peak, and 40.7 pp against 38.5 pp at φ = 0.60. The observed value sits
consistently a little above the implied one at low φ because the implied figure
covers only the Beta dispersion around each cell line's μ_i, whereas the observed
scatter also contains the covariate-driven spread in μ_i itself; at high φ the
dispersion dominates and the two converge.

The important comparison is the near-peak group against the high-φ group. A
φ ≈ 0.013 site has a %pU spread of **4–8 pp**, not the **19–41 pp** of the
high-φ sites. So the peak value is describing a genuinely tight distribution,
which is what the objection needs to be tested against. Seen directly:

```{r strip, fig.height=5.2, fig.width=12}
OB <- fread(file.path(LS, "observed.tsv"), sep = "\t")
OB <- merge(OB, P[, .(sid, phi, pooled_pu, prior_sd_pi)], by = "sid")
# display labels: ascending phi, and the contrast group named for what it IS
relab <- function(g) factor(fifelse(grepl("^low", g), "low: phi < 0.003",
                            fifelse(grepl("^near", g), "near the peak: phi ~ 0.0126",
                                    "high phi > 0.3 (artefact-enriched: IGH/MHC)")),
                            levels = c("low: phi < 0.003", "near the peak: phi ~ 0.0126",
                                       "high phi > 0.3 (artefact-enriched: IGH/MHC)"))
OB[, grp := relab(group)]; P[, grp := relab(group)]
lev <- P[order(phi)][, sprintf("chr%s   phi = %.4f", sub(" ", ":", sid), phi)]
mklab <- function(d) factor(sprintf("chr%s   phi = %.4f", sub(" ", ":", d$sid), d$phi), levels = lev)
OB[, row := mklab(OB)]; P[, row := mklab(P)]
pal3 <- c("low: phi < 0.003" = "#2c8a4a",
          "near the peak: phi ~ 0.0126" = "#1f4e9c",
          "high phi > 0.3 (artefact-enriched: IGH/MHC)" = "#c0392b")
ggplot(OB, aes(100 * pu, row, colour = grp)) +
  geom_segment(data = P, inherit.aes = FALSE,
               aes(y = row, yend = row,
                   x = pmax(0, 100 * (pooled_pu - prior_sd_pi)),
                   xend = pmin(100, 100 * (pooled_pu + prior_sd_pi))),
               colour = "grey70", linewidth = 3.2, alpha = 0.55) +
  geom_point(size = 1.5, alpha = 0.8) +
  scale_colour_manual(values = pal3, name = NULL) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
  guides(colour = guide_legend(nrow = 3, override.aes = list(size = 3))) +
  labs(x = "%pU in one cell line", y = NULL,
       title = "What phi looks like in the raw data: %pU across the 50 cell lines",
       subtitle = paste(
         "One point per cell line, sites ordered by phi, SHARED x-axis. Grey bar = the fitted Beta's +/-1 SD around the site mean.",
         "Genotype is deliberately absent: phi describes spread across cell lines and has nothing to do with any SNP.",
         "Low phi = tight cluster. Near the peak = wider. High phi = mass pushed towards both 0% and 100%.",
         sep = "\n")) +
  theme_bw(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        axis.text.y = element_text(size = 8, family = "mono"),
        plot.subtitle = element_text(size = 8))
```

Eleven sites cannot establish a relationship, so the same question asked of
every site. For each one, compare the spread the fitted model predicts against
the spread actually observed — once using the fitted φ, and once forcing φ = 0
so that only binomial sampling noise remains.

```{r spreadtab}
SPF <- fread(file.path(LP, "phi_vs_spread_fit.tsv"), sep = "\t")
knitr::kable(SPF, caption = "Predicted vs observed SD of %pU, per site, using each site's OWN fitted phi and its full depth vector. 'BB pred / obs' near 1 means phi reproduces the observed scatter; 'binomial pred / obs' below 1 means sampling noise alone cannot.")
```

```{r spreadfig, fig.height=7.2, fig.width=12}
VS <- fread(file.path(LP, "phi_vs_spread.txt.gz"), sep = "\t")
# NB these are tertiles of READ DEPTH, not of phi. Depth and phi are nearly
# independent here (Spearman 0.05; high-phi fraction 12.9/9.5/14.3% across the
# three), so the columns are not a proxy for dispersion. Labels carry the actual
# read counts so "shallow" cannot be misread as "low phi".
VS[, dep := cut(median_depth, quantile(median_depth, c(0, 1/3, 2/3, 1)),
                include.lowest = TRUE,
                labels = c("median depth 1-5 reads", "6-14 reads", "15+ reads"))]
M <- melt(VS[, .(dep, obs_sd,
                 `predicted using the fitted phi` = pred_sd_bb,
                 `predicted with phi = 0 (sampling noise only)` = pred_sd_binom)],
          id.vars = c("dep", "obs_sd"), variable.name = "prediction", value.name = "pred")
ggplot(M, aes(pred, obs_sd, colour = prediction)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey30", linetype = "dashed", linewidth = 0.4) +
  geom_point(size = 0.35, alpha = 0.25, shape = 16) +
  facet_grid(prediction ~ dep, labeller = label_wrap_gen(width = 30)) +
  scale_x_log10() + scale_y_log10() +
  scale_colour_manual(values = c("predicted using the fitted phi" = "#1f4e9c",
                                 "predicted with phi = 0 (sampling noise only)" = "#d98c00"),
                      guide = "none") +
  labs(x = "predicted SD of %pU", y = "observed SD of %pU",
       title = "Does phi describe the observed spread? All 10,467 sites",
       subtitle = paste(
         "Dashed = perfect agreement. TOP row uses each site's fitted phi; BOTTOM row sets phi = 0, leaving only binomial sampling noise.",
         "Columns are tertiles of READ DEPTH, not of phi -- the two are nearly independent (Spearman 0.05).",
         "Top row sits on the diagonal everywhere. Bottom row falls below it, and the shortfall grows with depth (pred/obs 1.00 -> 1.02 -> 0.88):",
         "deep sites have little sampling noise, so whatever spread remains must be real dispersion. That gap is what phi measures.",
         sep = "\n")) +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank(), strip.text.y = element_text(size = 8),
        plot.subtitle = element_text(size = 8))
```

The top row tracks the diagonal in all three depth strata (BB pred/obs
0.96–1.05), so the fitted φ reproduces the observed scatter across the whole
dataset, not merely at hand-picked sites. The bottom row — sampling noise alone
— falls short, and does so most clearly at **15+ reads** (0.881), which is
exactly where it should: deep sites have little sampling noise, so whatever
spread remains has to be real dispersion. The correlation between φ and the
observed spread strengthens with depth for the same reason (Spearman 0.22 → 0.34
→ 0.54).

To be explicit, since the column labels could invite the wrong reading: those
are **read-depth** tertiles (1–5, 6–14, 15+ reads per cell line), and depth is
nearly unrelated to φ — the fraction of high-φ sites is 12.9%, 9.5% and 14.3%
across the three. The median φ does rise with depth (10⁻⁵ → 3×10⁻³), but that is
the floor artefact from §2.1 rather than a dispersion difference: shallow sites
collapse onto the 10⁻⁶ boundary far more often.

```{r priorfig, fig.height=8, fig.width=13}
BN <- fread(file.path(LS, "prior_bins.tsv"), sep = "\t")
BN <- merge(BN, P[, .(sid, phi)], by = "sid")
BN[, grp := relab(group)]
BN[, row := factor(sprintf("chr%s   phi = %.4f", sub(" ", ":", sid), phi), levels = lev)]
ggplot(BN, aes(100 * bin_mid, mass, colour = grp)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ row, ncol = 4, scales = "free_y") +
  scale_colour_manual(values = pal3, name = NULL) +
  guides(colour = guide_legend(nrow = 3)) +
  labs(x = expression("latent pU rate "*p[i]*" (%)"), y = "probability mass",
       title = "The fitted Beta prior on the latent rate p_i",
       subtitle = paste(
         "Exact probability mass from pbeta, averaged over the 50 per-cell-line Beta distributions -- no simulation.",
         "Near the peak (blue) and at low phi (green) the prior is unimodal and narrow. At high phi (red) alpha < 1 and it collapses towards the 0/1 boundaries.",
         sep = "\n")) +
  theme_bw(base_size = 10) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        strip.text = element_text(size = 8))
```

The prior panels make the α criterion from §3.1 concrete. At the peak and below,
α ≈ 5–23 and the prior on p_i is a narrow unimodal bump — a coherent description
of a site whose pU level varies modestly between cell lines. In the high-φ group
α ≈ 0.16–0.39 and the prior piles against 0 and 1, asserting that cell lines are
either unmodified or fully modified. **That is what a large φ commits you to**,
and it is the right description for only about 12% of sites — most of which, as
noted above, sit in the Ig/MHC artefact regions.

# 4. What can and cannot be claimed

**Supported.** φ ≈ 0.013 is a reliable estimate of the *modal* dispersion across
sites. Three methods with unrelated failure modes agree; the estimator recovers a
known φ to within 3% at this value; a substantially larger φ is excluded by the
observed scatter; and the magnitude is exactly what the depth budget predicts.

**Not supported, and should not be claimed.** φ is *not* estimable at an
individual site. Measured against known truth (§2.1), at the modal value only
29% of estimates land within 2× and 48% are more than tenfold too low, with a
median fold error of 9×. **Per-site φ is therefore not a reportable quantity**
and should never be interpreted, filtered on, or shown for an individual site —
the retired φ > 0.99 filter was, in hindsight, thresholding noise. The observed φ distribution is *not* the true distribution:
its low-end bimodality is an estimator artefact, so only the mode and the
high-φ tail can be read from it, and the median cannot.

The practical consequence is that per-site φ should probably be **shrunk toward
the population mode** rather than estimated freely — the modal value is well
determined in aggregate while being undetermined per site, which is precisely
the case for partial pooling. That would also make the φ-selection heuristic
unnecessary, since the heuristic is a crude discrete approximation to the same
idea. Full pooling would be wrong: the 12% of genuinely high-dispersion sites
must be able to pull away from the mode. **Not yet implemented.**

# 5. Provenance

- Recovery simulation: `28.phi_recovery_simulation.R` via `run_phi_recovery.sbatch`
  → `qtl_betabinomial/phi_recovery/`. True φ ∈ {0.005, 0.0126, 0.05, 0.1, 0.2, 0.4},
  seed 2026, real n_i and fitted μ_i, k = 3, multi-start MLE.
- Moment estimators, spread check, interpretation, aggregate test:
  `29.low_phi_validation.R` via `run_low_phi_validation.sbatch`
  → `qtl_betabinomial/low_phi_validation/`. Multi-start MLE for the `phi_mle`
  column; the moment estimators use no fitter at all beyond taking μ̂_i from it.
- The peak value 0.0126: mode of log₁₀φ over the 5,045 sites where all three
  starts agree on φ to within 5% (`25.multistart_agreement.R`).
- Selection-rule accuracy (§2.4): `32.phi_rule_accuracy.R` via
  `run_phi_rule_accuracy.sbatch`. Each site simulated once per true φ and refit
  under all three rules on the SAME counts, so the comparison is paired.
- Genome-wide predicted vs observed spread (§3.3): `31.phi_vs_spread.R` via
  `run_phi_vs_spread.sbatch`, using each site's own fitted φ and full depth vector.
- Per-site identifiability: `multistart_audit/identifiability.txt.gz`, profile
  log-likelihood with β re-optimised at fixed φ.
- Example sites: `30.lowphi_example_sites.R` via `run_lowphi_example_sites.sbatch`
  → `qtl_betabinomial/lowphi_sites/`. Lead cis-SNP by beta-binomial LRT within
  ±5 kb at MAF ≥ 0.05; genotypes from `genotype/qtl_binomial/by_chr/chr*.traw`.
- Related: [the optimiser failure and the φ heuristic](qtl_mapping_betabinomial_optimiser.html) ·
  [the model](qtl_mapping_betabinomial_model.html) ·
  [window × nPC calibration grid](qtl_mapping_betabinomial_grid.html)
