---
title: "Beta-binomial pU-QTL: a goodness-of-fit test that did not work (and what it revealed)"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 3
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 11)

PSI <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl"
G   <- file.path(PSI, "qtl_betabinomial/goodness_of_fit")

gof  <- fread(file.path(G, "site_goodness_of_fit_noUCcalled.txt.gz"))
bp1  <- fread(file.path(G, "calibration_bootstrap_p_noUCcalled.txt.gz"))$bootstrap_p_ks
bs   <- fread(file.path(G, "gof_bootstrap_sites.txt.gz"))
feat <- fread(file.path(G, "site_features_noUCcalled.txt.gz"))
crit <- fread(file.path(G, "filter_criteria_comparison.txt"))

gof[, class := factor(class, levels = c("still problematic after filtering", "rest"))]
bs[, grp := factor(grp, levels = c("problematic", "matched control"))]
ctrl <- bs[grp == "matched control"]
```

We test if Beta-Binomial is a good fit. If the data form discrete clusters of %pU levels, then Beta-Binomial may be a poor fit.

Two attempts were made and neither produced usable p-values. 


# 1. The literature does not offer a usable off-the-shelf test

```{r lit}
knitr::kable(data.table(
  test = c("Tarone (1979) Z", "Brooks et al. (1997)", "both of the above"),
  problem = c(
    "tests binomial vs beta-binomial, i.e. whether overdispersion exists at all -- the wrong direction. It would confirm the beta-binomial beats the binomial, which we already assume.",
    "is a genuine beta-binomial GOF test, but Garren, Smith & Piegorsch (Biometrics 2000; 56:947-950) showed it cannot distinguish the beta-binomial from 'some severely non-beta-binomial models' -- exactly our alternative.",
    "assume homogeneous data: one mu shared by all observations. Our model has covariates (intercept + 3 M-value PCs), so mu differs per cell line.")),
  caption = "Why the classical tests were not used")
```

So the test had to be built from a general principle instead: **randomized
quantile residuals** (Dunn & Smyth 1996). Ask the fitted model what percentile
each cell line's observation sits at; under a correctly specified model those
percentiles are uniform, so testing uniformity *is* the fit test. Randomization
inside each integer's probability interval is what makes the percentiles
continuous — without it a discrete CDF gives lumpy values that are not uniform
even under a true model.

Four quantities were computed per site:

```{r what}
knitr::kable(data.table(
  output = c("p_ks", "p_sw", "p_dip", "best_K"),
  question = c("does the model fit? (KS of percentiles vs uniform -- PRIMARY)",
               "same, shape-sensitive (Shapiro-Wilk on qnorm(u) -- secondary)",
               "is there more than one cluster? (Hartigan dip test, ignores the model)",
               "how many clusters? (beta-binomial mixture BIC over K = 1, 2, 3)"),
  note = c("chosen as primary because its null is fully specified -- nothing can absorb a misfit",
           "location- and scale-invariant, so BLIND to residuals that are normal but too wide, i.e. to dispersion misfit",
           "answers 'more than one', not 'how many'",
           "the interpretable number: K=3 is the 0/1/2 genotype signature")),
  caption = "The four per-site outputs")
```

# 2. Attempt 1 — nominal p-values. Conservative.

Run on `fold5_union_noUCcalled` (U/C-variant sites already excluded), 11,006 of
11,298 sites testable.

The calibration check is the whole ballgame: simulate data **from** each fitted
model — data that is beta-binomial by construction — and re-test. The p-values
must come out uniform.

```{r cal1, fig.height=4}
d1 <- data.table(p = bp1)
ggplot(d1, aes(p)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "#3987e5", colour = "white",
                 linewidth = 0.2) +
  geom_hline(yintercept = length(bp1) / 20, linetype = "dashed", colour = "#e34948") +
  annotate("text", x = 0.02, y = length(bp1) / 20, hjust = 0, vjust = -0.6,
           label = "uniform expectation", colour = "#e34948", size = 3.4) +
  labs(x = "nominal KS p-value on data simulated from the fitted model",
       y = "replicates",
       title = "Attempt 1 calibration: p-values pile up at the TOP, not uniform",
       subtitle = sprintf("n = %s;  frac(p<0.05) = %.3f against a target of 0.05  ->  the test is CONSERVATIVE",
                          format(length(bp1), big.mark = ","), mean(bp1 < 0.05))) +
  theme_bw(base_size = 12)
```

`frac(p<0.05) = `r sprintf("%.3f", mean(bp1 < 0.05))`` where 0.05 is required, and
a KS test of these against uniform gives p = `r sprintf("%.3g", suppressWarnings(ks.test(bp1, "punif")$p.value))`.

**Cause.** Randomized quantile residuals are exactly uniform when the parameters
are *known*. Ours are estimated from the same ~50 observations (4 coefficients
plus φ), so the fitted model sits closer to the data than the truth would, the
residuals are under-dispersed, and the KS p-value inflates. Dunn & Smyth's
"exactly normal" carries the caveat "apart from sampling variability in the
estimated parameters" — at n = 50 with 5 parameters that caveat dominates.

A conservative test still has one use: anything it *does* flag is real. On that
basis the problematic sites were enriched — but the measured rate is a lower
bound, so the enrichment cannot be quantified.

```{r a1-res}
knitr::kable(gof[, .(sites = .N,
  `median p_ks` = signif(median(p_ks, na.rm = TRUE), 3),
  `frac q_ks<0.05` = sprintf("%.1f%%", 100 * mean(q_ks < 0.05, na.rm = TRUE)),
  `frac q_sw<0.05` = sprintf("%.1f%%", 100 * mean(q_sw < 0.05, na.rm = TRUE)),
  `frac q_dip<0.05` = sprintf("%.1f%%", 100 * mean(q_dip < 0.05, na.rm = TRUE))),
  by = class], caption = "Attempt 1 results -- read as lower bounds, not rates")
```

# 3. Attempt 2 — parametric bootstrap. Anti-conservative, and it lost sites.

The fix for an estimation-induced bias is to put the estimation into the null:
for each site, simulate B = 500 datasets from its own fit, **refit each one**, and
compare the observed KS statistic to that site's own null. Refitting every
replicate should make the conservatism cancel.

Run on the 17 still-problematic sites plus 10 matched controls each (matched on
covered cell lines, total depth and pooled %pU). The controls are the calibration
check — being ordinary sites, their `p_boot` must be uniform.

```{r cal2, fig.height=4}
ggplot(ctrl, aes(p_boot)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "#3987e5", colour = "white",
                 linewidth = 0.2) +
  geom_hline(yintercept = nrow(ctrl) / 20, linetype = "dashed", colour = "#e34948") +
  annotate("text", x = 0.55, y = nrow(ctrl) / 20, hjust = 0, vjust = -0.6,
           label = "uniform expectation", colour = "#e34948", size = 3.4) +
  labs(x = "bootstrap p-value at MATCHED CONTROL sites", y = "sites",
       title = "Attempt 2 calibration: now the p-values pile up at the BOTTOM",
       subtitle = sprintf("n = %d;  frac(p<0.05) = %.3f against a target of 0.05  ->  the test is ANTI-CONSERVATIVE",
                          nrow(ctrl), mean(ctrl$p_boot < 0.05))) +
  theme_bw(base_size = 12)
```

```{r cal2-tab}
knitr::kable(data.table(
  quantile = c("5%", "10%", "25%", "50%", "75%", "90%", "95%"),
  observed = round(quantile(ctrl$p_boot, c(.05, .1, .25, .5, .75, .9, .95)), 3),
  `expected if uniform` = c(0.05, 0.10, 0.25, 0.50, 0.75, 0.90, 0.95)),
  row.names = FALSE,
  caption = "Control p_boot against the uniform it should follow -- skewed across the whole range, not just the tail")
```

So attempt 1 was conservative at 0.012 and attempt 2 is anti-conservative at
`r sprintf("%.3f", mean(ctrl$p_boot < 0.05))`. Neither is usable.

## Why the bootstrap overcorrected: refits fail, and selectively

```{r refit, fig.height=4.6}
ggplot(bs, aes(grp, n_sim_ok, fill = grp)) +
  geom_boxplot(width = 0.5, alpha = 0.85, outlier.size = 1) +
  geom_hline(yintercept = 500, linetype = "dashed", colour = "grey40") +
  annotate("text", x = 2.35, y = 500, hjust = 1, vjust = 1.6, size = 3.4,
           colour = "grey30", label = "all 500 fitted") +
  scale_fill_manual(values = c(problematic = "#e34948",
                               `matched control` = "#3987e5"), guide = "none") +
  labs(x = NULL, y = "replicates refitted (of 500)",
       title = "The bootstrap null is truncated, and only at the problematic sites",
       subtitle = "Replicates that fail to refit produce no KS statistic and silently drop out of the null.") +
  theme_bw(base_size = 12)
```

```{r refit-tab}
knitr::kable(bs[, .(sites = .N,
  `median refits ok (of 500)` = as.numeric(median(n_sim_ok)),
  `min` = min(n_sim_ok),
  `sites below 450/500` = sum(n_sim_ok < 450)), by = grp],
  caption = "Every problematic site loses a fifth to nearly half of its null replicates; controls lose almost none.")
```

If the replicates that fail are the ones that would have produced *large* KS
statistics, the null is cut off at the top and the observed statistic looks
falsely extreme. That is the mechanism, and it also explains a second problem:

**8 of the 17 problematic sites produced no result at all**, because more than
half their replicates failed, tripping the script's own guard. The comparison
therefore ran on 9 of 17, and the 8 excluded are plausibly the worst ones — so
any enrichment estimate from the survivors is biased by an unknown amount.

```{r dropped}
knitr::kable(bs[grp == "problematic"][order(p_boot), .(
  site = sid, ig_mhc, n_cl, pooled = round(pooled, 3),
  D_obs = round(D_obs, 3), p_boot = signif(p_boot, 3),
  q_boot = signif(q_boot, 3), p_nominal = signif(p_ks_nominal, 2), best_K)],
  caption = paste("The", nrow(bs[grp == "problematic"]), "problematic sites that completed.",
                  "After BH correction nothing reaches q < 0.05."))
```

# 4. What survives, and what does not

## Survives

**Three sites genuinely misfit.** They separate by a wide margin under every
version of the test, have the largest KS statistics in the set, and are the only
ones needing more than one mixture component:

```{r survive}
knitr::kable(gof[sid %chin% c("3 37277733", "14 22551976", "6 29943223"),
  .(site = sid, p_ks = signif(p_ks, 2), p_sw = signif(p_sw, 2),
    p_dip = signif(p_dip, 2), best_K, phi = signif(phi, 3))][order(p_ks)],
  caption = "The three sites where the beta-binomial demonstrably does not describe the data")
```

For these, the advisor's hypothesis is confirmed: the %pU values form discrete
clusters and the model cannot represent them.

**But clustering is not the general explanation.** Of the 17 problematic sites,
most need only one mixture component — no clustering at all:

```{r bestk}
knitr::kable(dcast(gof[!is.na(best_K), .N, by = .(class, best_K)],
                   class ~ best_K, value.var = "N", fill = 0),
  caption = "Mixture components preferred by BIC. K=1 means no clustering.")
```

`r sum(gof[class != "rest" & best_K == 1]$best_K == 1)` of the
`r nrow(gof[class != "rest"])` problematic sites have K = 1. Genome-wide only
`r sprintf("%.1f%%", 100*mean(gof$best_K >= 2, na.rm = TRUE))` of sites prefer
K ≥ 2, so multimodality is uncommon and is not what makes these sites misbehave.

## Does not survive

```{r wrong}
knitr::kable(gof[, .(sites = .N,
  `frac q_ks<0.05` = sprintf("%.1f%%", 100 * mean(q_ks < 0.05, na.rm = TRUE)),
  `median best_K` = median(best_K, na.rm = TRUE)), by = ig_mhc],
  caption = "Prediction that failed: GOF failure was expected to concentrate in the Ig/MHC loci. It does not.")
```

Two predictions made before running were wrong. **GOF failure does not
concentrate in immunoglobulin or MHC loci** — the failure rate is 0.2% inside and
outside alike, even though 13 of the 17 problematic sites sit in those regions.
And **clustering was expected to explain the problematic sites as a group**; it
explains three of them.

Any statement of the form "*X*% of sites fail the fit test" is also not supported,
in either direction, because neither attempt was calibrated.

# 5. What the failure led to

The refit-failure diagnostic above turned out to be the useful result. Read as a
question it is: *if the world worked the way this site's fitted model says, could
I recover that model from fresh data?* Ordinary sites: yes, every time.
Problematic sites: no, a fifth to nearly half the time.

That pointed at the fitted parameters being degenerate, and checking them found
it directly — **every problematic site had the overdispersion parameter φ pinned
to its boundary**:

```{r phi}
knitr::kable(feat[is.finite(phi), .(
  sites = .N,
  `median phi` = signif(median(phi), 3),
  `frac phi > 0.99` = sprintf("%.3f", mean(phi > 0.99)),
  `median refit failure rate` = round(median(refit_fail, na.rm = TRUE), 3)),
  by = .(was_problematic = problematic)],
  caption = "phi is bounded to (0,1); phi -> 1 collapses the beta prior, so the fit is degenerate.")
```

φ > 0.99 turned out to be a far better criterion than anything the GOF test
offered, and removing those 48 sites eliminated the remaining problematic sites
entirely. The comparison against the other candidates:

```{r crit}
knitr::kable(crit[order(-caught, sites_removed)], row.names = FALSE,
  caption = "Candidate filters. phi>0.99 is not in this table because it was found afterwards: it catches 17 of 17 for ~44 sites removed (0.4%).")
```

So the honest summary of this work: **the goodness-of-fit test failed, and its
diagnostics found the actual bug.** The problem at those sites was never that the
beta-binomial family is wrong for this data — it is that at those sites the model
was never successfully fitted, and the LRT was computed from a boundary solution.

# 6. What would be needed to make the test work

- **Treat failed refits as extreme rather than dropping them.** A replicate that
  cannot be fitted is evidence of poor fit, not missing data. This removes the
  truncation bias directly and would let all 17 sites report. Most promising fix,
  small change.
- **Or fix β and φ at the observed estimates** instead of refitting, which avoids
  failures but reintroduces the attempt-1 conservatism.
- **Or use many more replicates with a bias correction**, which needs a
  characterisation of the estimation effect at n = 50 that we do not have.

None of this is currently a priority: with the φ filter in place there are no
problematic sites left to explain, so a working GOF test would be confirmatory
rather than diagnostic.

# Provenance

```{r prov}
knitr::kable(fread(file.path(G, "runinfo_noUCcalled.txt")), caption = "Attempt 1 parameters")
knitr::kable(fread(file.path(G, "gof_bootstrap_runinfo.txt")), caption = "Attempt 2 parameters")
knitr::kable(data.table(
  step = c("attempt 1: nominal p-values", "attempt 2: parametric bootstrap",
           "candidate filter comparison", "phi filter, iterated", "this page"),
  script = c("14.betabin_goodness_of_fit.R", "16.gof_bootstrap_pvalues.R",
             "17.problematic_site_filter.R", "18.phi_filter_iteration.R",
             "analysis/qtl_mapping_betabinomial_gof_failed.Rmd")),
  caption = "Scripts")
```

Related: [the current strategy and its calibration](qtl_mapping_betabinomial_strategy.html) ·
[the 26 problematic sites](qtl_mapping_betabinomial_site26.html)
