---
title: "Beta-binomial pU-QTL: a goodness-of-fit test that did not work (and what it revealed)"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 3
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 11)

PSI <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl"
G   <- file.path(PSI, "qtl_betabinomial/goodness_of_fit")

gof  <- fread(file.path(G, "site_goodness_of_fit_noUCcalled.txt.gz"))
bp1  <- fread(file.path(G, "calibration_bootstrap_p_noUCcalled.txt.gz"))$bootstrap_p_ks
bs   <- fread(file.path(G, "gof_bootstrap_sites.txt.gz"))
feat <- fread(file.path(G, "site_features_noUCcalled.txt.gz"))
crit <- fread(file.path(G, "filter_criteria_comparison.txt"))

gof[, class := factor(class, levels = c("still problematic after filtering", "rest"))]
bs[, grp := factor(grp, levels = c("problematic", "matched control"))]
ctrl <- bs[grp == "matched control"]
```

We test if Beta-Binomial is a good fit. If the data form discrete clusters of %pU levels, then Beta-Binomial may be a poor fit.

Two attempts were made and neither produced usable p-values. 


# 1. The literature does not offer a usable off-the-shelf test

```{r lit}
knitr::kable(data.table(
  test = c("Tarone (1979) Z", "Brooks et al. (1997)", "both of the above"),
  problem = c(
    "tests binomial vs beta-binomial, i.e. whether overdispersion exists at all -- the wrong direction. It would confirm the beta-binomial beats the binomial, which we already assume.",
    "is a genuine beta-binomial GOF test, but Garren, Smith & Piegorsch (Biometrics 2000; 56:947-950) showed it cannot distinguish the beta-binomial from 'some severely non-beta-binomial models' -- exactly our alternative.",
    "assume homogeneous data: one mu shared by all observations. Our model has covariates (intercept + 3 M-value PCs), so mu differs per cell line.")),
  caption = "Why the classical tests were not used")
```

So the test had to be built from a general principle instead: **randomized
quantile residuals** (Dunn & Smyth 1996). Ask the fitted model what percentile
each cell line's observation sits at; under a correctly specified model those
percentiles are uniform, so testing uniformity *is* the fit test. (The
percentiles are the PIT values `u`; the residual proper is `qnorm(u)` -- see the
note in section 2.) Randomization inside each integer's probability interval is
what makes the percentiles continuous — without it a discrete CDF gives lumpy values that are not uniform
even under a true model.

Four quantities were computed per site:

```{r what}
knitr::kable(data.table(
  output = c("p_ks", "p_sw", "p_dip", "best_K"),
  question = c("does the model fit? (KS of percentiles vs uniform -- PRIMARY)",
               "same, shape-sensitive (Shapiro-Wilk on qnorm(u) -- secondary)",
               "is there more than one cluster? (Hartigan dip test, ignores the model)",
               "how many clusters? (beta-binomial mixture BIC over K = 1, 2, 3)"),
  note = c("chosen as primary because its null is fully specified -- nothing can absorb a misfit",
           "location- and scale-invariant, so BLIND to residuals that are normal but too wide, i.e. to dispersion misfit",
           "answers 'more than one', not 'how many'",
           "the interpretable number: K=3 is the 0/1/2 genotype signature")),
  caption = "The four per-site outputs")
```

# 2. Attempt 1 — nominal p-values. Conservative.

Attempt 1 is one script doing **two separate jobs**: Part A produces the actual
per-site result from the real data, and Part B audits whether those results can be
trusted. Part A involves no simulation at all.

## 2a. Method — Part A: the test itself (real data, every site)

For one site, with 50 cell lines each having a converted count `y` out of a total
`n`:

1. **Fit the null model** — `logit(mu_i) = intercept + PC1 + PC2 + PC3`, one shared
   phi. No SNP: the null is what the LRT's inflation traced to, and including a
   SNP would mean choosing one of the site's 10-900 cis-SNPs, the best of which
   would absorb the misfit we are looking for. So each cell line gets its own
   `mu_i` (4 parameters generate all 50) and the site gets one `phi`.
2. **The model now predicts a whole distribution per cell line**, e.g.
   `BetaBinomial(n = 180, mu = 0.07, phi)` for a cell line with 180 reads.
3. **Locate the observation in that prediction.** For an observed `y = 12`,
   `F(11) = 0.399` and `F(12) = 0.509`, so `y = 12` occupies the percentile
   *interval* [0.399, 0.509] -- the slice of probability the value 12 accounts for.
4. **Draw uniformly inside that interval** -> `u = 0.428`. This randomisation is
   what makes the percentiles continuous; without it every observation of 12
   returns exactly 0.509 and the percentiles could not be uniform even under a
   perfect model.
5. **Collect the 50 percentiles.** Under a correct model they are Uniform(0,1).
6. **KS-test them against Uniform(0,1)** -> one p-value for the site.
7. **Repeat 3-6 five times** with different jitter, take the median, so one
   unlucky draw cannot decide a site.

Run over `fold5_union_noUCcalled`: 11,006 of 11,298 sites testable (the other 292
have fewer than 10 covered cell lines or the fit failed). That is
**55,030 KS tests producing 11,006 p-values.**

## 2b. Result — Part A

```{r a1-dist, fig.height=4}
d0 <- gof[!is.na(p_ks)]
ggplot(d0, aes(p_ks)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "#3987e5", colour = "white",
                 linewidth = 0.2) +
  geom_hline(yintercept = nrow(d0) / 20, linetype = "dashed", colour = "#e34948") +
  annotate("text", x = 0.02, y = nrow(d0) / 20, hjust = 0, vjust = -0.6,
           label = "uniform expectation", colour = "#e34948", size = 3.4) +
  labs(x = "p_ks (real data, one value per site)", y = "sites",
       title = sprintf("Part A: the %s per-site p-values", format(nrow(d0), big.mark = ",")),
       subtitle = sprintf("median %.3f; %d sites reach q_ks < 0.05 after BH correction. Skewed high -- which section 2d explains.",
                          median(d0$p_ks), sum(d0$q_ks < 0.05, na.rm = TRUE))) +
  theme_bw(base_size = 12)
```

```{r a1-counts}
knitr::kable(data.table(
  quantity = c("sites in the phenotype", "testable (>=10 covered cell lines, fit converged)",
               "p_ks values produced", "KS tests run (5 jitter draws each)",
               "sites with q_ks < 0.05 (BH)", "sites with q_dip < 0.05 (BH)",
               "sites preferring K >= 2 components"),
  n = c("11,298", format(nrow(gof), big.mark = ","), format(nrow(gof), big.mark = ","),
        format(nrow(gof) * 5, big.mark = ","),
        sum(gof$q_ks < 0.05, na.rm = TRUE), sum(gof$q_dip < 0.05, na.rm = TRUE),
        sum(gof$best_K >= 2, na.rm = TRUE))),
  caption = "Part A output. The 21 is a FLOOR, not a rate -- see 2d.")
```

```{r a1-res}
knitr::kable(gof[, .(sites = .N,
  `median p_ks` = signif(median(p_ks, na.rm = TRUE), 3),
  `frac q_ks<0.05` = sprintf("%.1f%%", 100 * mean(q_ks < 0.05, na.rm = TRUE)),
  `frac q_sw<0.05` = sprintf("%.1f%%", 100 * mean(q_sw < 0.05, na.rm = TRUE)),
  `frac q_dip<0.05` = sprintf("%.1f%%", 100 * mean(q_dip < 0.05, na.rm = TRUE))),
  by = class], caption = "Part A by site class -- read as lower bounds, not rates")
```

## 2c. Method — Part B: the audit (simulated data)

Part A step 5 claims that a correct model gives uniform percentiles. But `mu` and
`phi` were estimated from the same 50 observations, which may break that. Part A
cannot tell you -- you need data where the model is true by construction:

```r
f    <- fit_bb_mle(y, n, X)                        # 1. fit the REAL data
ysim <- rbetabinom.ab(n, shape1 = a, shape2 = b)   # 2. simulate from its parameters
f2   <- fit_bb_mle(ysim, n, X)                     # 3. RE-FIT the simulated data
u    <- rqr_u(ysim, n, f2$mu0, f2$phi)             # 4. PERCENTILES under the re-fit
ks.test(u, "punif")$p.value                        # 5. are they uniform?
```

400 sites x 5 replicates = ~1,896 p-values, which **must** be uniform.

A note on naming, since it is a common trip-up. `u` holds the **percentiles**
(the probability-integral-transform values), which are Uniform(0,1) under a
correct model. The *residual* proper is `z = qnorm(u)`, which is N(0,1) under a
correct model — that is what Dunn & Smyth call the randomized quantile residual,
and it is what Shapiro-Wilk is given. The two carry the same information in
different coordinates:

| object | scale | null distribution | test used |
|---|---|---|---|
| `u` | (0, 1) | Uniform(0, 1) | KS (primary) |
| `z = qnorm(u)` | (-Inf, Inf) | N(0, 1) | Shapiro-Wilk (secondary) |

Note that the re-fit is already present at this stage — that matters for reading
attempt 2 below.

## 2d. Result — Part B: the audit fails

```{r cal1, fig.height=4}
d1 <- data.table(p = bp1)
ggplot(d1, aes(p)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "#3987e5", colour = "white",
                 linewidth = 0.2) +
  geom_hline(yintercept = length(bp1) / 20, linetype = "dashed", colour = "#e34948") +
  annotate("text", x = 0.02, y = length(bp1) / 20, hjust = 0, vjust = -0.6,
           label = "uniform expectation", colour = "#e34948", size = 3.4) +
  labs(x = "nominal KS p-value on data simulated from the fitted model",
       y = "replicates",
       title = "Attempt 1 calibration: p-values pile up at the TOP, not uniform",
       subtitle = sprintf("n = %s;  frac(p<0.05) = %.3f against a target of 0.05  ->  the test is CONSERVATIVE",
                          format(length(bp1), big.mark = ","), mean(bp1 < 0.05))) +
  theme_bw(base_size = 12)
```

`frac(p<0.05) = `r sprintf("%.3f", mean(bp1 < 0.05))`` where 0.05 is required, and
a KS test of these against uniform gives p = `r sprintf("%.3g", suppressWarnings(ks.test(bp1, "punif")$p.value))`.

**Cause.** Randomized quantile residuals are exactly uniform when the parameters
are *known*. Ours are estimated from the same ~50 observations (4 coefficients
plus φ), so the fitted model sits closer to the data than the truth would and the
residuals are under-dispersed. The failure is therefore in the **reference
distribution**: the asymptotic KS null does not apply to residuals built from
estimated parameters at this sample size. Dunn & Smyth's "exactly normal" carries
the caveat "apart from sampling variability in the estimated parameters" — at
n = 50 with 5 parameters that caveat dominates.

**Consequence for section 2b.** Every number in Part A's result is now a *floor*.
A conservative test still has one use — anything it *does* flag is real, so the
21 sites at q_ks < 0.05 are genuine misfits and the enrichment in the problematic
class is real in direction. But the rates cannot be quantified, and a site passing
the test cannot be called well-fitted. That is what makes attempt 1 unusable as a
per-site verdict, and it is what attempt 2 tried to repair.

# 3. Attempt 2 — parametric bootstrap. Anti-conservative, and it lost sites.

## 3a. Method

The simulate-and-refit machinery was already in place; what was wrong was the
**reference distribution**. So attempt 2 changes only that — the null is
*generated* rather than looked up:

1. **Fit the real data**, and compute `D_obs` — the KS **statistic**, not its
   p-value. The theoretical p-value is discarded entirely.
2. **500 times:** simulate from those parameters, **re-fit**, compute `D_sim`.
   `D` is averaged over 3 jitter draws each time.
3. **Ask how extreme `D_obs` is** among those 500:

$$p_{\text{boot}} = \frac{1 + \#\{D_{\text{sim}} \ge D_{\text{obs}}\}}{B + 1}$$

   The `1 +` counts the observed statistic as one of its own null draws, so the
   smallest reportable p-value is 1/501 rather than 0. This is self-calibrating by
   construction: anything that biases `D` -- estimating 5 parameters from 50
   points, coarse discreteness -- biases `D_obs` and every `D_sim` alike and
   cancels.

To be explicit about the difference, because it is easy to misread: **both
attempts fit the real data, simulate from it, and re-fit.** Attempt 1 used those
simulations only to *audit* a nominal p-value; attempt 2 uses them to *build* the
p-value.

**Scope:** 500 refits per site is ~500x the per-site cost of attempt 1, so this
ran on 172 sites, not 11,006 — the 17 still-problematic ones plus 10 matched
controls each, matched on covered cell lines, total depth and pooled %pU. (Total
runtime was therefore *lower* than attempt 1: 74 s against 12 min 20 s.) The
controls are the calibration check — being ordinary sites, their `p_boot` must be
uniform.

## 3b. Result

```{r cal2, fig.height=4}
ggplot(ctrl, aes(p_boot)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "#3987e5", colour = "white",
                 linewidth = 0.2) +
  geom_hline(yintercept = nrow(ctrl) / 20, linetype = "dashed", colour = "#e34948") +
  annotate("text", x = 0.55, y = nrow(ctrl) / 20, hjust = 0, vjust = -0.6,
           label = "uniform expectation", colour = "#e34948", size = 3.4) +
  labs(x = "bootstrap p-value at MATCHED CONTROL sites", y = "sites",
       title = "Attempt 2 calibration: now the p-values pile up at the BOTTOM",
       subtitle = sprintf("n = %d;  frac(p<0.05) = %.3f against a target of 0.05  ->  the test is ANTI-CONSERVATIVE",
                          nrow(ctrl), mean(ctrl$p_boot < 0.05))) +
  theme_bw(base_size = 12)
```

```{r cal2-tab}
knitr::kable(data.table(
  quantile = c("5%", "10%", "25%", "50%", "75%", "90%", "95%"),
  observed = round(quantile(ctrl$p_boot, c(.05, .1, .25, .5, .75, .9, .95)), 3),
  `expected if uniform` = c(0.05, 0.10, 0.25, 0.50, 0.75, 0.90, 0.95)),
  row.names = FALSE,
  caption = "Control p_boot against the uniform it should follow -- skewed across the whole range, not just the tail")
```

So attempt 1 was conservative at 0.012 and attempt 2 is anti-conservative at
`r sprintf("%.3f", mean(ctrl$p_boot < 0.05))`. Neither is usable.

Attempt 2 failing is the more surprising of the two: an empirical p-value built
from a site's own null *should* be calibrated whatever the estimation effect. The
reason it is not turns out to be a defect in how failures were handled, not in
the bootstrap principle.

## Why the bootstrap overcorrected: refits fail, and selectively

```{r refit, fig.height=4.6}
ggplot(bs, aes(grp, n_sim_ok, fill = grp)) +
  geom_boxplot(width = 0.5, alpha = 0.85, outlier.size = 1) +
  geom_hline(yintercept = 500, linetype = "dashed", colour = "grey40") +
  annotate("text", x = 2.35, y = 500, hjust = 1, vjust = 1.6, size = 3.4,
           colour = "grey30", label = "all 500 fitted") +
  scale_fill_manual(values = c(problematic = "#e34948",
                               `matched control` = "#3987e5"), guide = "none") +
  labs(x = NULL, y = "replicates refitted (of 500)",
       title = "The bootstrap null is truncated, and only at the problematic sites",
       subtitle = "Replicates that fail to refit produce no KS statistic and silently drop out of the null.") +
  theme_bw(base_size = 12)
```

```{r refit-tab}
knitr::kable(bs[, .(sites = .N,
  `median refits ok (of 500)` = as.numeric(median(n_sim_ok)),
  `min` = min(n_sim_ok),
  `sites below 450/500` = sum(n_sim_ok < 450)), by = grp],
  caption = "Every problematic site loses a fifth to nearly half of its null replicates; controls lose almost none.")
```

If the replicates that fail are the ones that would have produced *large* KS
statistics, the null is cut off at the top and the observed statistic looks
falsely extreme. That is the mechanism, and it also explains a second problem:

**8 of the 17 problematic sites produced no result at all**, because more than
half their replicates failed, tripping the script's own guard. The comparison
therefore ran on 9 of 17, and the 8 excluded are plausibly the worst ones — so
any enrichment estimate from the survivors is biased by an unknown amount.

```{r dropped}
knitr::kable(bs[grp == "problematic"][order(p_boot), .(
  site = sid, ig_mhc, n_cl, pooled = round(pooled, 3),
  D_obs = round(D_obs, 3), p_boot = signif(p_boot, 3),
  q_boot = signif(q_boot, 3), p_nominal = signif(p_ks_nominal, 2), best_K)],
  caption = paste("The", nrow(bs[grp == "problematic"]), "problematic sites that completed.",
                  "After BH correction nothing reaches q < 0.05."))
```

# 4. What survives, and what does not

## Survives

**Three sites genuinely misfit.** They separate by a wide margin under every
version of the test, have the largest KS statistics in the set, and are the only
ones needing more than one mixture component:

```{r survive}
knitr::kable(gof[sid %chin% c("3 37277733", "14 22551976", "6 29943223"),
  .(site = sid, p_ks = signif(p_ks, 2), p_sw = signif(p_sw, 2),
    p_dip = signif(p_dip, 2), best_K, phi = signif(phi, 3))][order(p_ks)],
  caption = "The three sites where the beta-binomial demonstrably does not describe the data")
```

For these, the advisor's hypothesis is confirmed: the %pU values form discrete
clusters and the model cannot represent them.

**But clustering is not the general explanation.** Of the 17 problematic sites,
most need only one mixture component — no clustering at all:

```{r bestk}
knitr::kable(dcast(gof[!is.na(best_K), .N, by = .(class, best_K)],
                   class ~ best_K, value.var = "N", fill = 0),
  caption = "Mixture components preferred by BIC. K=1 means no clustering.")
```

`r sum(gof[class != "rest" & best_K == 1]$best_K == 1)` of the
`r nrow(gof[class != "rest"])` problematic sites have K = 1. Genome-wide only
`r sprintf("%.1f%%", 100*mean(gof$best_K >= 2, na.rm = TRUE))` of sites prefer
K ≥ 2, so multimodality is uncommon and is not what makes these sites misbehave.

## Does not survive

```{r wrong}
knitr::kable(gof[, .(sites = .N,
  `frac q_ks<0.05` = sprintf("%.1f%%", 100 * mean(q_ks < 0.05, na.rm = TRUE)),
  `median best_K` = median(best_K, na.rm = TRUE)), by = ig_mhc],
  caption = "Prediction that failed: GOF failure was expected to concentrate in the Ig/MHC loci. It does not.")
```

Two predictions made before running were wrong. **GOF failure does not
concentrate in immunoglobulin or MHC loci** — the failure rate is 0.2% inside and
outside alike, even though 13 of the 17 problematic sites sit in those regions.
And **clustering was expected to explain the problematic sites as a group**; it
explains three of them.

Any statement of the form "*X*% of sites fail the fit test" is also not supported,
in either direction, because neither attempt was calibrated.

# 5. What the failure led to

The refit-failure diagnostic above turned out to be the useful result. Read as a
question it is: *if the world worked the way this site's fitted model says, could
I recover that model from fresh data?* Ordinary sites: yes, every time.
Problematic sites: no, a fifth to nearly half the time.

That pointed at the fitted parameters being degenerate, and checking them found
it directly — **every problematic site had the overdispersion parameter φ pinned
to its boundary**:

```{r phi}
knitr::kable(feat[is.finite(phi), .(
  sites = .N,
  `median phi` = signif(median(phi), 3),
  `frac phi > 0.99` = sprintf("%.3f", mean(phi > 0.99)),
  `median refit failure rate` = round(median(refit_fail, na.rm = TRUE), 3)),
  by = .(was_problematic = problematic)],
  caption = "phi is bounded to (0,1); phi -> 1 collapses the beta prior, so the fit is degenerate.")
```

φ > 0.99 turned out to be a far better criterion than anything the GOF test
offered, and removing those 48 sites eliminated the remaining problematic sites
entirely. The comparison against the other candidates:

```{r crit}
knitr::kable(crit[order(-caught, sites_removed)], row.names = FALSE,
  caption = "Candidate filters. phi>0.99 is not in this table because it was found afterwards: it catches 17 of 17 for ~44 sites removed (0.4%).")
```

So the honest summary of this work: **the goodness-of-fit test failed, and its
diagnostics found the actual bug.** The problem at those sites was never that the
beta-binomial family is wrong for this data — it is that at those sites the model
was never successfully fitted, and the LRT was computed from a boundary solution.

# 6. What would be needed to make the test work

- **Treat failed refits as extreme rather than dropping them.** A replicate that
  cannot be fitted is evidence of poor fit, not missing data. This removes the
  truncation bias directly and would let all 17 sites report. Most promising fix,
  small change.
- **Or fix β and φ at the observed estimates** instead of refitting, which avoids
  failures but reintroduces the attempt-1 conservatism.
- **Or use many more replicates with a bias correction**, which needs a
  characterisation of the estimation effect at n = 50 that we do not have.

None of this is currently a priority: with the φ filter in place there are no
problematic sites left to explain, so a working GOF test would be confirmatory
rather than diagnostic.

# Provenance

```{r prov}
knitr::kable(fread(file.path(G, "runinfo_noUCcalled.txt")), caption = "Attempt 1 parameters")
knitr::kable(fread(file.path(G, "gof_bootstrap_runinfo.txt")), caption = "Attempt 2 parameters")
knitr::kable(data.table(
  step = c("attempt 1: nominal p-values", "attempt 2: parametric bootstrap",
           "candidate filter comparison", "phi filter, iterated", "this page"),
  script = c("14.betabin_goodness_of_fit.R", "16.gof_bootstrap_pvalues.R",
             "17.problematic_site_filter.R", "18.phi_filter_iteration.R",
             "analysis/qtl_mapping_betabinomial_gof_failed.Rmd")),
  caption = "Scripts")
```

Related: [the current strategy and its calibration](qtl_mapping_betabinomial_strategy.html) ·
[the 26 problematic sites](qtl_mapping_betabinomial_site26.html)
