---
title: "Score test vs. LRT — validation, and a bug it caught"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r setup, warning=FALSE, message=FALSE}
library(data.table)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 7, fig.height = 5.5)
```

# Motivation

The advisor's question: *"I am not familiar with the statistical test used
here [the score test]. You can probably still use LRT, but then you'll need
to include a 'nuisance' parameter — the extent of overdispersion."*

Both the [binomial](qtl_mapping_binomial_model.html) and
[beta-binomial](qtl_mapping_betabinomial_model.html) pU-QTL models test
$H_0:\beta_g=0$ using a **Rao score test** rather than a **likelihood-ratio
test (LRT)**, for computational reasons explained below. This page:

1. explains the score test and how it relates to the LRT,
2. validates the binomial score test against the LRT on real data (it is
   **exact**, not an approximation),
3. shows that the beta-binomial's *first* score-test implementation was
   **not** exact — a real bug this validation exercise caught — and how it
   was fixed.

# Background: score test vs. LRT

Both test the same null hypothesis; they differ in what they require:

|  | **LRT** | **Score test** |
|---|---|---|
| fit the **full** model (with $\beta_g$) | yes | no |
| fit the **null** model (without $\beta_g$) | yes | yes |
| nuisance parameters (covariates; $\phi$ for BB) | re-estimated in *both* fits | estimated **once**, under the null |
| statistic | $2(\ell_{\text{full}}-\ell_{\text{null}})$ | $U^2/V$ (score $U$, information $V$, from the null fit) |
| asymptotic null | $\chi^2_1$ | $\chi^2_1$ |

They are **asymptotically equivalent** — same hypothesis, same reference
distribution. In both, $\phi$ (for the beta-binomial) or the covariates (for
either model) are **nuisance parameters**; the score test just estimates them
once rather than refitting them for every alternative.

## Why a score test at all: the permutation cost

The within-site cis-SNP correction (FastQTL/tensorQTL permutation
beta-approximation) needs the nominal test statistic for **every cis-SNP**,
recomputed for **~1000 phenotype permutations**, at **every one of 12,416
sites**. An LRT needs a full-model refit per (site, SNP, permutation):

$$
12{,}416 \text{ sites} \times \sim\!40 \text{ cis-SNPs} \times 1000 \text{ perms}
\times 2 \text{ fits} \times 25\text{ms} \;\approx\; 69{,}000 \text{ CPU-hours}
$$

A score test needs only **one** null fit per (site, permutation); every
cis-SNP is then a fast vectorized matrix computation — no per-SNP refit.
That is the entire reason for using it (the same reason large-scale genetic
association methods like SAIGE/regenie use score tests genome-wide).

```{r load-data}
BASE <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl"
source("/project/xinhe/xsun/pumseq/3.qtl_mapping_binomial/qtl_lib.R")
source("/project/xinhe/xsun/pumseq/4.qtl_mapping_betabinomial/betabin_lib.R")

C <- fread(file.path(BASE, "phenotype/fold5_union/counts/C_matrix.txt.gz"))
N <- fread(file.path(BASE, "phenotype/fold5_union/counts/N_matrix.txt.gz"))
go <- fread(file.path(BASE, "genotype/plink2/pumseq_genotype.hg38.chr1-22.psam"))[[2]]
Cm <- as.matrix(C[, ..go]); Nm <- as.matrix(N[, ..go])
sites <- C[, .(ref = as.character(ref), pos = as.integer(pos))]
cov <- fread(file.path(BASE, "phenotype/fold5_union/mvalue_pcs.txt"))
Xcov <- cbind(1, as.matrix(cov[, paste0("PC", 1:3), with = FALSE]))

# chr1 slice of the common-SNP dosage matrix (matches the +/-5kb scan window)
samp <- paste0(go, "_", go)
traw <- fread(file.path(BASE, "genotype/qtl_binomial/genotype_common.traw"), nrows = 400000)
Gall <- t(as.matrix(traw[, ..samp])); gpos <- traw$POS; gchr <- as.character(traw$CHR)
rm(traw); gc()
chr_idx <- split(seq_along(gpos), gchr)
for (ch in names(chr_idx)) { o <- order(gpos[chr_idx[[ch]]]); chr_idx[[ch]] <- chr_idx[[ch]][o] }

WINDOW <- 5000; MAF_MIN <- 0.05
cis_window <- function(s) {
  ch <- sites$ref[s]; if (is.null(chr_idx[[ch]])) return(NULL)
  cols <- chr_idx[[ch]]; pos <- gpos[cols]
  w <- cols[pos >= sites$pos[s] - WINDOW & pos <= sites$pos[s] + WINDOW]
  if (length(w) == 0) return(NULL)
  G <- mean_impute(Gall[, w, drop = FALSE]); keep <- maf_of(G) >= MAF_MIN
  if (!any(keep)) return(NULL)
  G[, keep, drop = FALSE]
}
set.seed(2026)
cand <- sample(which(sites$ref == "1" & sites$pos < max(gpos)), 150)
```

# 1. Binomial: the score test is exact

The binomial (logit link) is a **canonical one-parameter exponential
family** — its score/information formulas are the *exact* derivatives of the
likelihood, not an approximation:

$$
r_i = y_i - n_i\hat p_i, \qquad w_i = n_i\hat p_i(1-\hat p_i), \qquad U=G^\top r,\quad V = G^\top\text{diag}(w)G-(\cdots)
$$

```{r bin-validate}
res_bin <- rbindlist(lapply(cand, function(s) {
  y <- Cm[s, ]; n <- Nm[s, ]; if (sum(n >= 1) < 10) return(NULL)
  G <- cis_window(s); if (is.null(G)) return(NULL)
  nf <- fit_null(y, n, Xcov); if (!nf$converged) return(NULL)
  p_score <- score_pvals(nf, G); if (all(is.na(p_score))) return(NULL)
  lead <- which.min(p_score)
  ll_null <- binom_loglik(y, n, Xcov)
  ff <- suppressWarnings(glm.fit(cbind(Xcov, G[, lead]), cbind(y, n - y), family = binomial()))
  ll_full <- sum(dbinom(y, n, ff$fitted.values, log = TRUE))
  p_lrt <- pchisq(max(2 * (ll_full - ll_null), 0), 1, lower.tail = FALSE)
  data.table(site = s, p_score = p_score[lead], p_lrt = p_lrt)
}))
cat(sprintf("n=%d sites | Spearman rank cor = %.4f\n",
            nrow(res_bin), cor(res_bin$p_score, res_bin$p_lrt, method = "spearman")))
```

```{r bin-plot}
ggplot(res_bin, aes(-log10(p_lrt), -log10(p_score))) +
  geom_abline(slope = 1, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.6, color = "steelblue") +
  labs(x = expression(-log[10]("LRT p")), y = expression(-log[10]("score p")),
       title = "Binomial: score test vs. LRT (real sites, own lead SNP)",
       subtitle = sprintf("rank cor = %.3f — tight agreement, as expected for an exact score test",
                          cor(res_bin$p_score, res_bin$p_lrt, method = "spearman"))) +
  theme_bw(base_size = 12)
```

Tight, unbiased agreement — this is the expected behavior of a mathematically
exact score test, confirmed on real data.

# 2. Beta-binomial: the naive extension was *not* exact — a bug

## The quasi-score (wrong)

A first implementation simply divided the binomial's score/information by
the overdispersion factor $c_i=1+(n_i-1)\phi$ — matching the beta-binomial's
first two *moments*, but not its actual likelihood (a mixture distribution,
not an exponential family in $\mu$ given $\phi$):

```{r quasi-score-def}
# the WRONG formula (shown here only for comparison; removed from the codebase)
quasi_bb_rw <- function(y, n, mu, phi) {
  cc <- 1 + (n - 1) * phi
  list(r = (y - n * mu) / cc, w = n * mu * (1 - mu) / cc)
}
```

## The fix (exact)

The true score and observed information of the beta-binomial log-likelihood,
in closed form via digamma ($\psi$) / trigamma ($\psi'$) — see the [model
page](qtl_mapping_betabinomial_model.html#QTL_mapping:_the_beta-binomial_score_test)
for the derivation:

```{r exact-score-def}
exact_bb_rw   # already loaded from betabin_lib.R
```

## Side-by-side on real data

For each site, fit the beta-binomial null once (`aod::betabin`, giving the
exact joint MLE of $\beta_0,\gamma,\phi$), find the lead cis-SNP under the
**exact** score, then compute: the exact score p, the quasi-score p at the
*same* SNP, and the real LRT p (full model refit, $\phi$ freely
re-estimated) as ground truth.

```{r bb-validate}
res_bb <- rbindlist(lapply(cand, function(s) {
  y <- Cm[s, ]; n <- Nm[s, ]; if (sum(n >= 1) < 10) return(NULL)
  G <- cis_window(s); if (is.null(G)) return(NULL)
  nf <- fit_null_betabin(y, n, Xcov); if (is.null(nf)) return(NULL)
  p_exact <- score_pvals(nf, G); if (all(is.na(p_exact))) return(NULL)
  lead <- which.min(p_exact)

  qrw <- quasi_bb_rw(y, n, nf$mu0, nf$phi)
  WX <- qrw$w * Xcov; Ainv_q <- tryCatch(solve(crossprod(Xcov, WX)), error = function(e) NULL)
  if (is.null(Ainv_q)) return(NULL)
  g <- G[, lead]
  Uq <- sum(g * qrw$r); GtWXq <- crossprod(g, WX)
  Vq <- sum(g^2 * qrw$w) - as.numeric(GtWXq %*% Ainv_q %*% t(GtWXq))
  p_quasi <- if (is.finite(Vq) && Vq > 0) pchisq(Uq^2 / Vq, 1, lower.tail = FALSE) else NA

  nf_full <- fit_null_betabin(y, n, cbind(Xcov, g)); if (is.null(nf_full)) return(NULL)
  p_lrt <- pchisq(max(2 * (nf_full$logLik_bb - nf$logLik_bb), 0), 1, lower.tail = FALSE)

  data.table(site = s, phi = nf$phi, p_quasi = p_quasi, p_exact = p_exact[lead], p_lrt = p_lrt)
}))
cat(sprintf("n=%d sites\n", nrow(res_bb)))
cat(sprintf("QUASI-score  vs LRT: rank cor = %.3f\n",
            cor(res_bb$p_quasi, res_bb$p_lrt, method = "spearman", use = "complete.obs")))
cat(sprintf("EXACT-score  vs LRT: rank cor = %.3f\n",
            cor(res_bb$p_exact, res_bb$p_lrt, method = "spearman", use = "complete.obs")))
```

```{r bb-plot, fig.width=11, fig.height=5.5}
pd <- rbind(
  res_bb[, .(site, phi, p_test = p_quasi, p_lrt, method = "quasi-score (buggy)")],
  res_bb[, .(site, phi, p_test = p_exact, p_lrt, method = "exact score (fixed)")]
)
pd[, method := factor(method, levels = c("quasi-score (buggy)", "exact score (fixed)"))]
ggplot(pd, aes(-log10(p_lrt), -log10(p_test), color = phi)) +
  geom_abline(slope = 1, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.7) +
  scale_color_viridis_c(option = "C") +
  facet_wrap(~method) +
  labs(x = expression(-log[10]("LRT p")), y = expression(-log[10]("score p")),
       color = expression(phi),
       title = "Beta-binomial: quasi-score (before) vs. exact score (after), both vs. LRT",
       subtitle = "left: scattered off the diagonal, worse at high phi (color) -- right: tracks the diagonal") +
  theme_bw(base_size = 12)
```

**Left panel (quasi-score)**: scattered badly off the diagonal, worst at
high $\phi$ (yellow points) — some sites the quasi-score called essentially
certain ($-\log_{10}p > 10$) the LRT calls non-significant. **Right panel
(exact score)**: tracks the diagonal; residual scatter concentrates at high
$\phi$ points but is no longer one-sided.

## One concrete example

```{r worked-example}
worst <- res_bb[order(-abs(log10(p_quasi / p_lrt)))][1]
knitr::kable(worst[, .(site, phi = round(phi, 3),
                       quasi_score_p = formatC(p_quasi, format = "e", digits = 2),
                       exact_score_p = formatC(p_exact, format = "e", digits = 2),
                       LRT_p = formatC(p_lrt, format = "e", digits = 2))],
             caption = "The single largest quasi-score/LRT disagreement in this sample")
```

At this site the quasi-score and the LRT disagree on whether there is a QTL
at all; the exact score agrees with the LRT's conclusion.

## Is the residual exact-score/LRT gap a problem?

```{r phi-cor}
cat(sprintf("correlation(|log10(exact/lrt)|, phi) = %.3f\n",
            cor(abs(log10(res_bb$p_exact / res_bb$p_lrt)), res_bb$phi, use = "complete.obs")))
cat(sprintf("mean log10(exact/lrt) = %.3f (0 = unbiased)\n",
            mean(log10(res_bb$p_exact / res_bb$p_lrt), na.rm = TRUE)))
```

The remaining disagreement grows with $\phi$ and is **not** systematically
biased in one direction (mean log-ratio near 0) — consistent with the normal
finite-sample gap between a score test and an LRT when there is an extra
nuisance parameter ($\phi$) and effective sample size shrinks at high
overdispersion. This is expected statistical behavior, not a further
approximation error — unlike the quasi-score's one-sided divergence, which
was a genuine bug.

# 3. Details: the Rao score test

For a null hypothesis $H_0:\beta_g=0$ with nuisance parameters $\theta$
(covariates for the binomial; covariates + $\phi$ for the beta-binomial):

1. Fit $\hat\theta$ under the null (no $\beta_g$) — **once**.
2. **Score** $U(\beta_g)=\partial\ell/\partial\beta_g$, evaluated at
   $(\beta_g{=}0,\hat\theta)$ — how much the likelihood "wants" to move away
   from $\beta_g=0$.
3. **Information** $V$ = the curvature of $\ell$ at that point, with the
   nuisance parameters' own uncertainty partialled out (the
   $G^\top WX\, (X^\top WX)^{-1}\,X^\top WG$ correction term in
   `score_pvals()` — the standard nuisance-parameter adjustment for a score
   test with multiple nuisance parameters).
4. Statistic $U^2/V \sim \chi^2_1$ under $H_0$.

The binomial's $U,V$ are the *exact* derivatives of its likelihood (steps
2-3 have closed forms because the binomial is a canonical exponential
family). The beta-binomial's are *also* now exact — via digamma/trigamma —
after the fix above; the earlier quasi-score used a moment-matching
substitute for step 2-3 that is only valid when $\phi$ is small.

**Why this is fast**: because $\hat\theta$ is fit once, testing $M$ cis-SNPs
is $M$ evaluations of a closed-form formula (matrix algebra), not $M$ model
refits — this is what makes the 1000x-permutation cis-correction affordable
around a GLM.
