---
title: "Estimating phi per site: the problem, and six things we have tried"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 3
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 10)
PSI <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl"
RC  <- file.path(PSI, "qtl_betabinomial/phi_recovery")
LP  <- file.path(PSI, "qtl_betabinomial/low_phi_validation")
```

> This page records a problem with our per-site estimates of
> the overdispersion φ, and the six things we have tried against it.


# 1. The problem: estimates pile up at zero when φ is small

The test is simple: simulate data where we *know* φ, then refit with our usual
estimator and see what comes back.

**How the data were simulated.** We reuse the real design rather than an
idealised one. Fitting the null model to the real counts at every site gives, per
site, a fitted mean μ̂ᵢ for each cell line (which varies across lines through the
3 phenotype PCs). Those means and the **real read depths nᵢ** are held fixed as
templates; only φ is imposed. Then for a chosen true φ, at every site and every
cell line:

$$p_i \;\sim\; \mathrm{Beta}\!\left(\hat\mu_i\tfrac{1-\varphi}{\varphi},\; (1-\hat\mu_i)\tfrac{1-\varphi}{\varphi}\right), \qquad y_i \;\sim\; \mathrm{Binomial}(n_i,\, p_i)$$

The simulated counts therefore have our actual depth distribution, our actual
mean structure and our actual number of cell lines — so the estimation problem is
the one we really face, with the single difference that we know the answer.

**Scale of the simulation.**

```{r simscale}
RBx <- fread(file.path(RC, "recovery_by_site.txt.gz"), sep = "\t")
knitr::kable(data.table(
  quantity = c("true phi values imposed",
               "sites simulated per true phi",
               "cell lines per simulated site",
               "site-level datasets in total",
               "simulated count observations (y, n) in total",
               "random seed"),
  value = c(paste(sort(unique(RBx$phi_true)), collapse = ", "),
            format(uniqueN(RBx$sid), big.mark = ","),
            "median 49 (range 10-50)",
            format(nrow(RBx), big.mark = ","),
            "~3.08 million",
            "2026")),
  caption = "One dataset per site per true phi. Each 'site-level dataset' is ~49 paired counts (y_i, n_i), one per cell line, which is exactly what a real site provides. The histograms below pool the ~10,000 sites within each true-phi panel.")
```

**The design of one panel, step by step.** Take a single true φ, say 0.0126.
Give **every one of the 10,467 sites that same φ** — so within a panel there is
exactly one true value, and it is the red line. Simulate that site's ~49 paired
counts. Estimate φ̂ at that site, using only that site's own data, exactly as the
production pipeline does. Repeat independently for all 10,467 sites, and plot the
histogram of the resulting 10,467 estimates. Then repeat the whole procedure at
each of the six true φ values to get the six panels.

**What success would look like.** If the estimator were perfect, all 10,467
estimates would equal 0.0126 and the histogram would be a single narrow spike on
the red line. It cannot be that sharp in practice — each site has only ~49 cell
lines at ~8 reads, so some scatter is unavoidable and expected. A *well-behaved*
estimator would therefore give **one hump centred on the red line**, with its
width reflecting nothing but sampling noise.

What the figure actually shows at small φ is **a hump on the red line plus a
second spike jammed against the boundary**. The hump is fine and expected. The
second spike is the failure: those sites did not return a noisy version of
0.0126, they returned essentially zero.

A handful of fits fail outright at each value (9,491 to 10,076 succeed of 10,467)
and are excluded from the panel. Note the failure rate itself rises with φ, so
those exclusions are not random.

## 1.1 Which sites are estimated badly?

If the estimates are wrong at a large minority of sites, the natural question is
whether those sites share anything. They do, and it is overwhelmingly **read
depth**.

```{r accprop}
knitr::kable(fread(file.path(RC, "accuracy_by_property.tsv"), sep = "\t"),
  caption = "Sites at true phi = 0.0126, grouped by how accurate their estimate turned out. The accurate group has 9x more information (sum n(n-1)) and 3.7x lower predicted SE than the sites that collapsed to the floor. Cell-line count and pooled %pU barely differ between groups.")
```


```{r accdepth}
knitr::kable(fread(file.path(RC, "accuracy_by_depth.tsv"), sep = "\t"),
  caption = "Accuracy by read-depth quintile. This is the actionable result: at ~51 reads per cell line, only 3% of sites collapse and 68% land within 2x, whereas at ~3 reads only 7% are within 2x.")
```

**Three things follow.**

**Depth is the whole story, and it is actionable.** Sites in the top depth
quintile (~51 reads per cell line) collapse only 3% of the time and are within
2x of the truth 68% of the time. Sites in the bottom quintile (~3 reads) are
within 2x only 7% of the time. So a per-site φ *is* recoverable — for the deep
sites. The obvious consequence is that φ should be reported only above a depth
threshold, not genome-wide.

**Cell-line count and %pU are not the problem.** The four accuracy groups have
essentially the same number of covered cell lines (47-50) and similar
modification rates (4.0-5.4%). Only depth and the information summary
sum n(n-1) separate them.

**The boundary rate is not monotone in depth** — it peaks at ~10 reads (24%)
rather than at the lowest depth (12.8%). At very low depth the Pearson statistic
is so noisy that estimates scatter in *both* directions rather than settling on
the boundary, so the shallowest sites are wrong in both directions rather than
specifically collapsed.

```{r prob, fig.height=7, fig.width=13}
RB <- fread(file.path(RC, "recovery_by_site.txt.gz"), sep = "\t")[is.finite(phi_hat)]
RS <- fread(file.path(RC, "recovery_summary.tsv"), sep = "\t")
SP <- fread(file.path(RC, "site_properties.txt.gz"), sep = "\t")
RB <- merge(RB, SP[, .(sid, med_depth)], by = "sid")
PT <- sort(unique(RB$phi_true))
RB[, panel := factor(sprintf("true phi = %g", phi_true),
                     levels = sprintf("true phi = %g", PT))]
# colour by READ DEPTH quintile -- the point is that the boundary spike and the
# hump on the truth are made of different sites
qb <- quantile(SP$med_depth, 0:5/5)
RB[, dq := cut(med_depth, qb, include.lowest = TRUE,
               labels = c("1-3 reads (shallowest 20%)", "4-6", "7-11", "12-24",
                          "25+ reads (deepest 20%)"))]
vl <- merge(unique(RB[, .(panel, phi_true)]), RS[, .(phi_true, mode = `mode phi_hat`)], by = "phi_true")
ggplot(RB, aes(phi_hat, fill = dq)) +
  geom_histogram(bins = 60, colour = NA) +
  geom_vline(xintercept = 1e-6, colour = "grey35", linetype = "dotted") +
  geom_vline(data = vl, aes(xintercept = phi_true), colour = "#c0392b", linewidth = 0.7) +
  facet_wrap(~ panel, nrow = 2, scales = "free_y") +
  scale_x_log10(breaks = 10^seq(-6, 0, 2), labels = c("1e-6", "1e-4", "1e-2", "1")) +
  scale_fill_viridis_d(option = "viridis", direction = -1,
                       name = "median reads per cell line") +
  guides(fill = guide_legend(nrow = 1, reverse = TRUE)) +
  labs(x = expression(hat(varphi)*", axis log-spaced"), y = "sites",
       title = "Simulated from a KNOWN phi, then refitted -- coloured by read depth",
       subtitle = paste(
         "Red = the true phi used to simulate. Dotted grey = the 1e-6 numerical floor. Stacked bars, so colour shows COMPOSITION.",
         "A perfect estimator would give one narrow spike on the red line; a well-behaved one, a single hump centred there.",
         "The hump on the truth is built from the DEEP sites (yellow). Depth, not the value of phi, decides which a site lands in.",
         sep = "\n")) +
  theme_bw(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        plot.subtitle = element_text(size = 8))
```

At a true φ of 0.0126 — the value our real data suggest — **14% of sites return
essentially zero** and about half are more than tenfold too low, even though the
mode of the distribution sits correctly on the truth:

```{r probtab}
knitr::kable(RB[, { r <- phi_hat/phi_true[1]
  .(`within 2x of truth` = sprintf("%.1f%%", 100*mean(r > 0.5 & r < 2)),
    `>10x too low`       = sprintf("%.1f%%", 100*mean(r < 0.1)),
    `at the 1e-6 floor`  = sprintf("%.1f%%", 100*mean(phi_hat < 2e-6)),
    `median fold error`  = round(median(pmax(r, 1/r)), 1)) },
  by = .(`true phi` = phi_true)][order(`true phi`)],
  caption = "Per-site accuracy against known truth. The mode is nearly unbiased at every true phi >= 0.0126, but individual sites are frequently wrong by orders of magnitude, and worst where phi is smallest.")
```


# 2. What we tried

```{r tried}
knitr::kable(data.table(
  approach = c("multi-start optimiser",
               "peak-proximity rule",
               "trend-based prior centre (DESeq2/edgeR style)",
               "Gaussian prior on log phi",
               "Student-t prior on log phi",
               "two-component mixture prior",
               "prior derived from the MOMENT estimator"),
  idea = c("start the optimiser from 3 values of phi, keep the best fit",
           "of the optima found, take the one nearest the population mode",
           "predict each site's phi from its mean and depth, shrink toward that",
           "penalise phi values far from 0.0126",
           "same, but with a heavy tail so distant sites escape",
           "two plausible centres (0.019 and 0.46) instead of one",
           "estimate the prior from an optimiser-free estimator, then MAP"),
  outcome = c("FIXED A DIFFERENT BUG: 13 sites where the likelihood is bimodal and one fixed start found the worse optimum. Did NOT touch the boundary spike.",
              "NO. Per-site tracking unchanged; it only moves the pile off the boundary to ~1e-4. A relocation, not a repair.",
              "RULED OUT before building: phi is not predictable from site features (R-squared = 0.058), so there is no trend to shrink toward.",
              "NO. Loses to a CONSTANT GUESS at 3 of 4 true phi, and tracks WORSE than the MLE at large phi (Spearman 0.68 vs 0.74).",
              "NO. Same failure -- at 2 log-units out the tail has not saturated enough to help.",
              "MARGINAL. Roughly ties the MLE on tracking at large phi; and its second component is not supported by an optimiser-free estimator (see 3.2).",
              "BEST OF THE SIX, but not a solution: it is the only approach that beats a constant guess at all four true phi, yet per-site tracking at the modal phi is still only Spearman 0.41.")),
  caption = "Six attempts. Verdicts are by per-site tracking (see 3.1), not by the 'within 2x' rate, which is misleading -- see below.")
```

## 2.1 How we judge "worked": per-site tracking, not the hit rate

We first scored these by the fraction of sites within 2x of the truth. **That
metric is misleading, and it flattered every method.** If the true phi values are
tightly clustered, then simply returning a constant near the middle scores well
without using the data at all. So the table below adds two honest criteria:

- **slope** of log(estimate) on log(truth). 1 means the estimate tracks each
  site; 0 means it returns a constant regardless of the data.
- **Spearman** correlation between true and estimated phi, per site.
- a **constant-guess benchmark**: ignore the data entirely and return the true
  median at every site. Any method that cannot beat this is not measuring
  anything.

```{r track}
TR <- fread(file.path(RC, "approach_tracking.tsv"), sep = "\t")
LV <- c("(constant guess)", "MLE", "gauss sd=0.83", "t df=2 sd=0.83",
        "t df=1 sd=0.4", "MIXTURE", "MAP: mom-median")
TR <- TR[est %chin% LV][, est := factor(est, levels = LV)]
knitr::kable(dcast(TR, est ~ truth_centre, value.var = "spearman")[order(est)],
  caption = "SPEARMAN correlation between true and estimated phi, per site, by the centre of the true phi distribution. Higher is better; the constant guess is 0 by construction.")
```

```{r trackslope}
knitr::kable(dcast(TR, est ~ truth_centre, value.var = "slope")[order(est)],
  caption = "SLOPE of log(estimate) on log(truth). 1 = tracks each site; 0 = returns a constant; above 1 = spreads estimates more widely than the truth.")
```

```{r trackhit}
knitr::kable(dcast(TR, est ~ truth_centre, value.var = "within2x")[order(est)],
  caption = "The 'within 2x' rate, shown for comparison. Note the constant guess scores 68.7% at the smallest true phi -- which is why this metric cannot be used on its own.")
```

Three things follow, and they change the verdicts.

**The MLE spreads estimates too widely, not too narrowly.** Its slope is 1.55-1.85
everywhere, well above 1. The boundary collapse is the cause: sites pushed to
1e-6 create extreme low values that stretch the range.

**The fixed-centre priors do not work.** The Gaussian scores 61.6% within 2x at
the smallest true phi, which looked like a large gain over the MLE's 30.0% --
but a constant guess scores 68.7%, so it is worse than not using the data.
It loses to the constant at three of the four true phi values, and at large phi
it tracks worse than the plain MLE (Spearman 0.682 against 0.737). The
Student-t behaves the same way. The mixture is better but only ties the MLE on
tracking at large phi.

**The moment-derived prior is the best of the six, but it does not solve the
problem.** It has the best Spearman at every true phi (0.41 / 0.69 / 0.81 /
0.77), it is the only approach that beats the constant guess at all four, and its
slope rises toward 1 as information increases (0.36 -> 0.62 -> 0.82 -> 0.90),
which is the behaviour a shrinkage estimator should have. But **0.41 at the phi
our data actually suggest is weak per-site agreement**, and no approach we tried
does better there. Where information is plentiful (phi ~ 0.2-0.4) the tracking is
respectable; at the modal phi it is not.

<!-- ## 2.2 A caveat we had to correct ourselves on -->

<!-- The mixture prior's two components were fitted to the subset of sites where all -->
<!-- three optimiser starts agree on φ. But that subset is **defined by the estimator -->
<!-- under suspicion**, so it cannot be used to decide whether the estimator's -->
<!-- bimodality is real. Using an optimiser-free, covariate-adjusted estimator -->
<!-- instead, the second peak largely disappears: -->

<!-- ```{r bimod, fig.height=7, fig.width=12} -->
<!-- MA <- file.path(PSI, "qtl_betabinomial/multistart_audit") -->
<!-- ms <- fread(file.path(MA, "multistart_by_site.txt.gz"), sep = "\t") -->
<!-- ag <- fread(file.path(MA, "agreement_by_site.txt.gz"), sep = "\t") -->
<!-- mo <- fread(file.path(LP, "moment_by_site.txt.gz"), sep = "\t") -->
<!-- LV <- c("MLE, all sites\n(has the boundary spike)", -->
<!--         "MLE, 'reliable' subset\n(selected BY the MLE - circular)", -->
<!--         "moments, pooled mean\n(no optimiser, NO covariates)", -->
<!--         "moments, covariate-adjusted\n(no optimiser, WITH covariates)") -->
<!-- bd <- rbindlist(list( -->
<!--   data.table(phi = ms[is.finite(phi_3start)]$phi_3start,   src = LV[1]), -->
<!--   data.table(phi = ag[n_ok == 3L & n_opt_phi == 1L]$phi_1, src = LV[2]), -->
<!--   data.table(phi = mo$phi_mom_pooled,                      src = LV[3]), -->
<!--   data.table(phi = mo$phi_mom_adjusted,                    src = LV[4]))) -->
<!-- bd <- bd[is.finite(phi) & phi > 0][, src := factor(src, levels = LV)] -->
<!-- bann <- bd[, .(n = .N, hi = mean(phi > 0.128), -->
<!--                mode = { k <- density(log10(phi)); 10^k$x[which.max(k$y)] }), by = src] -->
<!-- bann[, lab := sprintf("n = %s\nmode = %.4f\n%.1f%% above 0.128", -->
<!--                       format(n, big.mark = ","), mode, 100 * hi)] -->
<!-- ggplot(bd, aes(phi)) + -->
<!--   geom_histogram(bins = 70, fill = "#4a7fbf", colour = NA) + -->
<!--   geom_vline(xintercept = 0.128, colour = "#c0392b", linetype = "dashed", linewidth = 0.5) + -->
<!--   geom_vline(xintercept = 1e-6, colour = "grey35", linetype = "dotted") + -->
<!--   geom_text(data = bann, aes(x = 1.3e-6, y = Inf, label = lab), hjust = 0, vjust = 1.2, -->
<!--             size = 2.7, colour = "grey15", lineheight = 0.95) + -->
<!--   facet_wrap(~ src, nrow = 2, scales = "free_y") + -->
<!--   scale_x_log10(breaks = 10^c(-6, -4, -2, 0), labels = c("1e-6", "1e-4", "1e-2", "1")) + -->
<!--   labs(x = expression(hat(varphi)*", axis log-spaced"), y = "sites", -->
<!--        title = "Is the phi distribution really two-humped? Four estimators, same data", -->
<!--        subtitle = paste( -->
<!--          "Red dashed = 0.128, where the Beta prior turns U-shaped. Dotted = the 1e-6 floor.", -->
<!--          "The two-humped shape is clear in the MLE-based views and in the unadjusted moment estimator,", -->
<!--          "but LARGELY DISAPPEARS in the covariate-adjusted moment estimator -- the fairest optimiser-free comparison.", -->
<!--          sep = "\n")) + -->
<!--   theme_bw(base_size = 10) + -->
<!--   theme(panel.grid.minor = element_blank(), strip.text = element_text(size = 8), -->
<!--         plot.subtitle = element_text(size = 7.5)) -->
<!-- ``` -->

# 3. The best of the six: a prior built from the moment estimator

Three passes, and the key point is that **nothing is assumed about where φ sits**:

1. compute the moment estimate at every site — closed form, no optimiser, no prior
2. derive the prior's centre and scale from those estimates by a fixed rule
3. refit each site by MAP with that prior

**How this simulation differs from the one in §1.** In §1 a single true φ was
imposed on every site, which is the right way to ask "does the estimator recover
a known value". It cannot be used to test a *prior*, because with one true value
the best possible prior is a spike on that value — it would score perfectly while
learning nothing. A prior describes how φ varies *between* sites, so the truth
has to vary. So here:

1. **invent a per-site truth**: `log10(φᵢ) ~ Normal(log10(centre), sd)`, clipped
   to [1e-5, 0.9]. Four centres are used — 0.0126, 0.05, 0.2, 0.4 — so the method
   is tested where our data suggest φ sits *and* well away from it.
2. **simulate counts** as in §1, reusing the real depths nᵢ and real fitted μ̂ᵢ.
3. **moment estimate** at every site from those simulated counts.
4. **derive the prior** from those estimates — median of all of them (negatives
   included) for the centre, MAD of log10 of the positives for the width.
5. **MAP refit** per site with that prior.
6. **score** against the per-site truth from step 1.

The prior is rebuilt from scratch inside every truth setting, so what is being
tested is the whole procedure rather than a prior that happened to suit one
answer. That is what makes the comparison fair: the derived centre had to find
the truth at four different values, and it did (truth 0.0126 → derived 0.0124;
0.05 → 0.0433; 0.2 → 0.140; 0.4 → 0.281). A fixed-centre prior would have been
right in one column and wrong in three.

```{r workedscatter, fig.height=6.4, fig.width=13}
sc <- fread(file.path(RC, "moment_prior_by_site.txt.gz"), sep = "\t")
sc <- sc[is.finite(phi_hat) & phi_hat > 0 & is.finite(phi_true) &
         est %chin% c("MLE", "MAP: mom-median")]
# defined here rather than in an earlier chunk so this figure stands alone
EL <- c("MLE" = "MLE (no prior)", "MAP: mom-median" = "MAP, prior from moment estimates")
CT <- c(0.0126, 0.05, 0.2, 0.4)
sc[, rl := factor(EL[est], levels = EL)]
sc[, cl := factor(sprintf("true phi centred at %g", truth_centre),
                  levels = sprintf("true phi centred at %g", CT))]
sann <- sc[, .(sp = cor(log10(phi_true), log10(phi_hat), method = "spearman"),
               sl = coef(lm(log10(phi_hat) ~ log10(phi_true)))[2]), by = .(rl, cl)]
sann[, lab := sprintf("Spearman %.2f\nslope %.2f", sp, sl)]
ggplot(sc, aes(phi_true, phi_hat)) +
  geom_abline(slope = 1, intercept = 0, colour = "#c0392b", linewidth = 0.5) +
  geom_point(size = 0.25, alpha = 0.15, shape = 16, colour = "#1f4e9c") +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), se = FALSE,
              colour = "#f5a623", linewidth = 0.7) +
  geom_text(data = sann, aes(x = 1.5e-5, y = 1.3, label = lab), hjust = 0, vjust = 1,
            size = 2.7, colour = "grey15", lineheight = 0.95) +
  facet_grid(rl ~ cl, labeller = label_wrap_gen(width = 28)) +
  scale_x_log10() + scale_y_log10() +
  labs(x = expression(TRUE~varphi[i]~"(one point per site)"),
       y = expression(estimated~hat(varphi)[i]),
       title = "Estimated against TRUE phi, site by site",
       subtitle = paste(
  "Red = y = x, perfect recovery. Orange = a smooth fit through the points. One point per site.",
  "This is the honest test: matching histograms can hide a wrong pairing, a scatter cannot.",
  "TOP (MLE): the vertical smear at the bottom is the boundary collapse -- sites whose truth varies over a decade all returned ~1e-6.",
  "BOTTOM (MAP): the collapse is gone, but at the LEFT-hand columns the cloud is nearly FLAT -- the estimate barely responds to the truth (slope 0.36).",
  "Only at large phi (right-hand columns) does the cloud approach the red line, which is where a per-site phi means something.",
  sep = "\n")) +
  theme_bw(base_size = 10) +
  theme(panel.grid.minor = element_blank(), strip.text = element_text(size = 8),
        plot.subtitle = element_text(size = 7.5))
```

The scatter is a stricter test than the histogram above it, and it shows two
things the histogram could not.

**The MLE's surviving estimates are unbiased.** In the top row the main cloud
sits squarely *on* the red line at every truth centre. The MLE's problem is not
that it estimates φ badly — it is that a detached blob of sites falls off the
bottom entirely, spanning the whole range of truth and all returning ~10⁻⁶. So
the MLE is really two estimators: an unbiased one, and a failure mode. That is a
more favourable reading of the MLE than the pooled accuracy figures give it.

**The MAP trades collapse for inflation, and the scatter makes the direction
visible.** In the bottom row the blob is gone, but at the low end the cloud sits
*above* the red line — roughly tenfold above it at the left edge of the
`0.0126` panel — before crossing it further right. That is shrinkage toward the
prior centre seen from the side: sites whose true φ lies below the centre are
pushed **up**, and sites above it are pulled **down**. Near-horizontal at small
φ (slope 0.36), approaching the line by φ = 0.4 (slope 0.90).

<!-- Worth noting which direction of error is safer. Overestimating φ widens the null -->
<!-- and is conservative for testing; the MLE's collapse to zero narrows it and is -->
<!-- anti-conservative. So although the MAP is biased at the low end, it is biased the -->
<!-- less dangerous way — which matters if these φ ever feed a test rather than being -->
<!-- reported. -->

<!-- ```{r workedtab} -->
<!-- M <- fread(file.path(RC, "moment_prior_accuracy.tsv"), sep = "\t") -->
<!-- knitr::kable(dcast(M[est %in% c("MLE", "MAP: mom-median", "MAP: mom-mode")], -->
<!--                    truth_centre ~ est, value.var = "within 2x")[ -->
<!--   , .(`true phi centre` = truth_centre, MLE, `MAP: mom-median`, `MAP: mom-mode`)], -->
<!--   caption = "Percent within 2x of truth. Unlike every fixed-centre prior, this beats the MLE at EVERY true phi, including 0.2 and 0.4.") -->
<!-- ``` -->

<!-- The boundary spike is eliminated (>10× too low falls from 47.6% to 0.1% at small -->
<!-- φ) **and** large-φ sites improve rather than degrade (80.4% → 89–92%). -->

<!-- **But it is not recovering per-site φ at the low end, and that must be said -->
<!-- plainly.** At the smallest true φ the slope of estimate on truth is **0.36** and -->
<!-- the Spearman correlation **0.41** — near the centre the method is largely -->
<!-- returning the prior rather than measuring the site. On the real data the effect -->
<!-- is visible directly: the estimated distribution collapses to a spike at the prior -->
<!-- centre, mode and median both 0.0048, and **57.8% of sites move by more than -->
<!-- five-fold** from their MLE value. Tracking is genuinely good where information -->
<!-- exists (slope 0.90, Spearman 0.77 at φ = 0.4), and that is where a reported -->
<!-- per-site φ means something. -->

<!-- So the defensible summary is: **the best available estimator of the population -->
<!-- distribution and of the well-informed sites, and a shrunk summary rather than a -->
<!-- measurement everywhere else. It is not a per-site measurement of phi at the low -->
<!-- end, and we do not think any of the six approaches achieves that.** -->

<!-- Applied to our real counts, the same calibrated rule puts the central φ at -->
<!-- **0.0052** — about 2.4× below the 0.0126 we had been quoting from the mode of the -->
<!-- MLE distribution. Worth noting these are different statistics (a median versus a -->
<!-- mode), but the direction is consistent. -->

