---
title: "Beta-binomial pU-QTL: the φ pathology is an optimiser failure"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 3
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 10)

PSI  <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl"
MA   <- file.path(PSI, "qtl_betabinomial/multistart_audit")
DIAG <- file.path(PSI, "qtl_betabinomial/diagnostic_figures")
CAL  <- file.path(PSI, "qtl_betabinomial/lrt_calibration")

ms   <- fread(file.path(MA, "multistart_by_site.txt.gz"), sep = "\t")
agr  <- fread(file.path(MA, "agreement_by_site.txt.gz"), sep = "\t")
idn  <- fread(file.path(MA, "identifiability.txt.gz"), sep = "\t")
qqd  <- fread(file.path(MA, "qq_thinned_multistart.tsv.gz"), sep = "\t")
bad9 <- fread(file.path(DIAG, "pathological_sites_w5kb_pc3_auto_noUCcalled.txt"),
              sep = "\t")$sid

ms  <- ms[is.finite(ll_1start) & is.finite(ll_3start)]
ms[, grp := "unchanged (gain < 1)"]
ms[gain > 1, grp := "better fit found (gain > 1)"]
ms[sid %chin% bad9, grp := "problematic site (9)"]
GRP <- c("unchanged (gain < 1)", "better fit found (gain > 1)", "problematic site (9)")
ms[, grp := factor(grp, levels = GRP)]
COL <- setNames(c("grey60", "#3987e5", "#e34948"), GRP)
setorder(ms, grp)
CRIT <- 1.92          # 1-df LRT at p = 0.05

# the multi-start calibration re-run may not have landed yet; the section that
# needs it is skipped rather than showing stale numbers.
MSCAL <- file.path(CAL, "calib_siteres_w5kb_pc3_auto_noUCcalled_MULTISTART_lambda.txt")
HAVE_MSCAL <- file.exists(MSCAL)
```

# The short version

The pU-QTL scan had a residual calibration problem: a handful of sites returned
extreme p-values *even under phenotype permutation*, where no signal can exist.
Those sites all had the fitted overdispersion φ pinned to its upper boundary, and
the production pipeline excludes them on that basis — see
[the filtering strategy](qtl_mapping_betabinomial_strategy.html).

**That φ value was not the maximum likelihood estimate.** `fit_bb_mle` starts its
optimiser at a single fixed point, φ₀ = 0.01. At these sites the likelihood is
**bimodal in φ**, and that start lands in the basin of a secondary optimum on the
boundary which is 20–320 log-likelihood units *worse* than the optimum reachable
from a higher start.

So the sites that broke calibration are sites where the fitter failed, not sites
where the beta-binomial cannot describe the data. Their true φ is 0.5–0.9 —
dispersed, but nowhere near degenerate.

# 1. What is actually happening

Three explanations were checked and two were wrong.

```{r ruledout}
knitr::kable(data.table(
  hypothesis = c("The analytic gradient is wrong at extreme phi",
                 "The logit reparameterization creates a flat plateau at the boundary, so the optimiser stalls",
                 "The likelihood is genuinely bimodal in phi"),
  test = c("compare neggr() against a numerical gradient at the returned solution",
           "profile the likelihood in phi and inspect the gradient near phi = 1",
           "restart the SAME optimiser from several phi values and compare optima"),
  verdict = c("RULED OUT -- analytic and numeric agree to 4 significant figures",
              "RULED OUT -- the gradient at the boundary is -27, pointing back toward the optimum, not 0",
              "CONFIRMED -- two distinct stationary points, gradient norm ~1e-4 at each")),
  caption = "The zero-gradient fallback in neggr() (`if (!all(is.finite(out))) return(rep(0, q+1))`) is a genuine latent hazard -- handing BFGS a zero vector is indistinguishable from 'you are at the optimum' -- but it never fires at these sites.")
```

Both solutions are real optima. At `chr3:37277733`:

```{r twomodes}
knitr::kable(data.table(
  `start phi0` = c(0.01, 0.10, 0.50),
  `converged phi` = c(0.999999, 0.699145, 0.699146),
  `log-likelihood` = c(-478.249, -158.412, -158.412),
  note = c("<- the shipped single start", "", "")),
  caption = "Same data, same objective, same optimiser, same starting coefficients. Only the starting phi differs. There is a valley between 0.01 and 0.1, and the historical start is on the wrong side of it.")
```

# 2. How the fix works

"Multi-start" means running the same optimiser from several starting φ values and
keeping whichever run reaches the highest log-likelihood. That is not a heuristic
tie-break: the MLE *is* the maximiser, so a run reporting a lower likelihood on
the same data has simply failed. Adding starts can never give a worse fit.

It is switched on by an environment variable so that the default reproduces every
pre-existing result exactly:

```r
BB_PHI_STARTS <- local({                       # betabin_lib.R
  v <- as.numeric(strsplit(Sys.getenv("BB_PHI_STARTS", "0.01"), ",")[[1]])
  ...
})
# default "0.01"      = the historical single start, bit-for-bit
# "0.01,0.1,0.5"      = the fix
```

Cost is three cheap null fits instead of one. Verified in both modes: the default
returns −478.249 at `chr3:37277733`, the multi-start returns −158.412, and
well-behaved sites are unchanged.

# 3. How far it reaches

Null refits at k = 3 on the autosome-only phenotype, every site, both schemes.

```{r gaintab}
brk <- c(-Inf, 1e-6, 0.01, 1, 10, 100, Inf)
lab <- c("none (<1e-6)", "1e-6 to 0.01", "0.01 to 1", "1 to 10", "10 to 100", ">100")
ms[, band := cut(gain, brk, labels = lab, right = FALSE)]
knitr::kable(ms[, .(sites = .N, `% of sites` = sprintf("%.2f%%", 100 * .N / nrow(ms)),
                    `of which problematic` = sum(sid %chin% bad9)), by = band][order(band)],
  caption = "Log-likelihood gained by using 3 starts instead of 1. For 99.5% of sites the two schemes agree to within 0.01 log units -- the single start is fine almost everywhere.")
```

```{r llscatter, fig.height=7.2}
ggplot(ms, aes(ll_1start, ll_3start)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(aes(colour = grp, size = grp, alpha = grp)) +
  scale_colour_manual(values = COL, name = NULL) +
  scale_size_manual(values = c(0.35, 2.2, 2.6), guide = "none") +
  scale_alpha_manual(values = c(0.35, 0.95, 0.95), guide = "none") +
  guides(colour = guide_legend(override.aes = list(size = 2.6, alpha = 1))) +
  labs(x = "log-likelihood, shipped single start", y = "log-likelihood, 3 starts",
       title = "Null-model log-likelihood: single start vs multi-start",
       subtitle = sprintf("Multi-start takes the best of 3, so nothing can fall BELOW the line -- verified, %d sites do.\nEverything sits on the diagonal except the 9 problematic sites (red).",
                          sum(ms$gain < -1e-6))) +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

**All 9 problematic sites are among the 13 that gain more than one log unit.**
The correspondence between "single-start failure" and "broke calibration" is
essentially exact.

```{r the13}
knitr::kable(ms[gain > 1][order(-gain), .(
  site = paste0("chr", sub(" ", ":", sid)), n_cl,
  `phi 1 start` = sprintf("%.6f", phi_1start),
  `phi 3 starts` = sprintf("%.4f", phi_3start),
  `loglik gain` = round(gain, 1),
  `failed toward` = ifelse(phi_1start > phi_3start, "phi -> 1", "phi -> 0"),
  `was problematic` = sid %chin% bad9)],
  caption = "The failure runs in BOTH directions: 10 sites were driven to the phi -> 1 boundary, 3 collapsed to the binomial limit instead. Any guard added to the fitter therefore has to be two-sided.")
```

# 4. Reproducibility of the optimiser

Of the sites where all three starts succeeded, how many distinct optima did they
find? Agreement is judged on the log-likelihood, because that is what the
optimiser maximises and what defines the MLE.

```{r agree}
full <- agr[n_ok == 3L]
knitr::kable(data.table(
  `distinct optima among the 3 starts` = c("all 3 agree", "exactly 2 agree", "all 3 differ"),
  `by log-likelihood` = vapply(1:3, function(k) sum(full$n_opt_ll == k), integer(1)),
  `by phi (within 5%)` = vapply(1:3, function(k) sum(full$n_opt_phi == k), integer(1)))[
    , `:=`(`% loglik` = sprintf("%.2f%%", 100 * `by log-likelihood` / nrow(full)),
           `% phi`    = sprintf("%.2f%%", 100 * `by phi (within 5%)` / nrow(full)))],
  caption = sprintf("Of the %s sites where all 3 starts converged. Note the two columns disagree completely -- see section 5.",
                    format(nrow(full), big.mark = ",")))
```

The optimiser is reproducible on the likelihood at **99.48%** of sites, and
there is **never a three-way split** — when it fails it always falls into exactly
two basins, which is what a bimodal surface predicts and a merely rough surface
would not.

# 5. φ is often not an estimable quantity

The φ column above tells a different story: at **45%** of sites all three starts
report a different φ while agreeing on the likelihood to within 0.01. That is not
disagreement about anything real — it is that the data do not determine φ.

φ enters the model only through the variance inflation factor
$1 + (n_i-1)\varphi$. At a typical depth of ~16 reads, φ = 10⁻⁶ inflates the
variance by 0.0015% and φ = 10⁻⁴ by 0.16%: indistinguishable with 50
observations.

```{r flat}
knitr::kable(data.table(
  phi = c("1e-06", "1e-05", "1e-04", "4.36e-04", "1e-03", "1e-02", "1e-01"),
  `profile loglik` = c(-58.66200, -58.66201, -58.66216, -58.66323, -58.66681,
                       -58.94946, -65.67935),
  `variance inflation` = c(1.000015, 1.000155, 1.001550, 1.006758, 1.015500,
                           1.155, 2.550)),
  caption = "Profile log-likelihood in phi at chr13:20783614, an ordinary site (50 cell lines, median depth 16). Over four orders of magnitude in phi the log-likelihood moves by 0.005. A 1-df LRT needs 1.92 to distinguish two values, so every phi in [1e-6, 1e-2] -- a 10,000-fold range -- is statistically equivalent here.")
```

Genome-wide, the data usually cannot even reject the binomial:

```{r ident}
r <- idn[is.finite(drop_to_0) & is.finite(drop_to_1)]
knitr::kable(data.table(
  question = c("Can the data reject phi = 0 (the binomial)?",
               "Can the data reject phi = 1 (the boundary)?",
               "Sites with fitted phi > 0.99 where the data DO reject phi = 0",
               "Sites with fitted phi > 0.99 where they do NOT"),
  answer = c(sprintf("NO at %s of %s sites (%.1f%%)",
                     format(sum(r$drop_to_0 < CRIT), big.mark = ","),
                     format(nrow(r), big.mark = ","), 100 * mean(r$drop_to_0 < CRIT)),
             sprintf("NO at %d sites (%.2f%%)", sum(r$drop_to_1 < CRIT),
                     100 * mean(r$drop_to_1 < CRIT)),
             sprintf("%d", sum(r[phi > 0.99]$drop_to_0 >= CRIT)),
             sprintf("%d", sum(r[phi > 0.99]$drop_to_0 < CRIT)))),
  caption = sprintf("Profile log-likelihood dropped to each boundary with beta re-optimised, compared against the 1-df threshold of %.2f. At four sites in five there is no defensible point estimate of phi at all.", CRIT))
```

**Consequence for the filter.** `φ̂ > 0.99` asks *where the optimiser stopped*.
The question that matters is *do the data demand high dispersion?* Those differ:
a likelihood-based criterion would drop the spurious flags and, critically, would
**never have flagged the 9 problematic sites**, because at those the data reject
φ ≈ 1 by 20–320 log units. The existing filter caught them only because it
thresholds the very artefact that broke the fit.

**Consequence for the published φ figures.** The fine structure in the φ
histograms on the [strategy](qtl_mapping_betabinomial_strategy.html) and
[covariate-count](qtl_mapping_betabinomial_npcs_tradeoff.html) pages below about
10⁻² is not measuring dispersion. It shows where the optimiser stopped on a flat
surface.

# 6. Does the calibration problem survive the fix?

No. It disappears completely, and without discarding any sites.

```{r calres}
CALTAB <- data.table(
  run = c("1 start (as shipped)", "3 starts (fixed fitter)"),
  real_lambda = c(1.401, 1.399),
  perm = c("1.034 - 1.070", "1.029 - 1.063"),
  real_tail = c(141.3, 28.2),
  worst_perm_tail = c(142.6, 11.3),
  problematic = c(9L, 0L))
knitr::kable(CALTAB[, .(run, `real lambda` = real_lambda,
  `permutation lambda` = perm, `real tail` = real_tail,
  `worst permuted tail` = worst_perm_tail, `problematic sites` = problematic)],
  caption = "Both rows: k = 3, +/-5 kb, chr1-22, U/C sites removed, phi-boundary sites RETAINED in both. The ONLY difference is the number of optimiser starts.")
```

**Fixing the optimiser removes the pathology outright, with no sites discarded.**

- **Problematic sites 9 → 0.**
- **Worst permuted tail 142.6 → 11.3.** This is the number to watch: under
  permutation there is no signal, so it should be near zero.
- **Every site retained** — nothing is filtered out.

## The QQ plots

The table above is easier to believe as a picture. Under permutation there is no
genotype–phenotype relationship, so the **blue** series must lie on the diagonal;
where they climb above it the test is manufacturing signal from nothing.

```{r qq-ms, fig.height=5.2, fig.width=13}
RUNLV <- c("1 start, phi sites KEPT", "3 starts, phi sites KEPT (no filter)")
ORD  <- c("real", paste0("perm", 1:5))
CQ   <- c(real = "#e34948", perm1 = "#9ec5f4", perm2 = "#6da7ec",
          perm3 = "#3987e5", perm4 = "#256abf", perm5 = "#0d366b")
q <- copy(qqd)[run %in% RUNLV]     # the phi-filter run is not shown
q[, run := factor(run, levels = RUNLV)][, dataset := factor(dataset, levels = ORD)]

ggplot(q, aes(expected, observed, colour = dataset)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45") +
  geom_point(size = 0.55, alpha = 0.8) +
  scale_colour_manual(values = CQ, name = NULL) +
  guides(colour = guide_legend(override.aes = list(size = 2.6, alpha = 1), nrow = 1)) +
  facet_wrap(~ run, nrow = 1) +
  labs(x = expression(-log[10]("expected p")), y = expression(-log[10]("observed p")),
       title = "Calibration QQ: single start vs multi-start",
       subtitle = paste("Red = real scan, blues = 5 phenotype permutations. Axes SHARED, so the panels are directly comparable.",
                        "\nLEFT: the blues climb to 142 -- pure artefact, since permuted data contain no signal.",
                        "\nRIGHT: the same sites, none removed, just fitted properly -- the blues collapse onto the diagonal.")) +
  theme_bw(base_size = 12) + theme(legend.position = "bottom",
                                   strip.text = element_text(size = 9))
```

```{r qq-ms-zoom, fig.height=5.2, fig.width=13}
# clipped companion: the left panel's artefact tail compresses everything else,
# so the bulk behaviour near the diagonal is invisible without clipping.
YM <- 12
ggplot(q, aes(expected, observed, colour = dataset)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45") +
  geom_point(size = 0.55, alpha = 0.8) +
  scale_colour_manual(values = CQ, name = NULL) +
  guides(colour = guide_legend(override.aes = list(size = 2.6, alpha = 1), nrow = 1)) +
  facet_wrap(~ run, nrow = 1) +
  coord_cartesian(ylim = c(0, YM)) +
  labs(x = expression(-log[10]("expected p")), y = expression(-log[10]("observed p")),
       title = sprintf("Same data, y clipped at %d to show the bulk", YM),
       subtitle = paste("With the artefact tail out of view the two panels look much the same near the diagonal --",
                        "which is the point:\nthe fix removes a pathology confined to a handful of sites and leaves",
                        "everything else alone. Clipped points are\noutside the view, not deleted.")) +
  theme_bw(base_size = 12) + theme(legend.position = "bottom",
                                   strip.text = element_text(size = 9))
```

**λ barely moves** (1.401 → 1.399, permutation 1.034–1.070 → 1.029–1.063), and
that is expected rather than disappointing: λ is a median over ~292,000 site-SNP
p-values, and 9 sites contribute a few hundred of them. A median cannot move. This
is the same reason λ was identified early in this project as the wrong diagnostic
for a pathology confined to a few sites — tail height and the problematic-site
count are the sensitive statistics, and both collapse.


# 7. Choosing φ when it is not identifiable

Section 5 leaves an open problem. Taking the highest-likelihood solution is the
right rule whenever the likelihood actually distinguishes the candidates — but at
about half of all sites it does not, and there the winning start is decided by
log-likelihood differences of order 10⁻⁹. That is arbitrary. In statistical
terms φ is **unidentifiable** at those sites: different values give the same
likelihood, so the data contain no information to choose between them.

When a parameter is unidentifiable, one workable remedy is to choose the value
that looks reasonable in light of the other sites. Most sites do return a single
well-determined φ; the distribution of those φ has a peak, which is the most
likely value of φ across the dataset; and where a site is ambiguous, the
candidate nearest that peak can be taken instead of an arbitrary one.

> **The rule.** Run the optimiser from φ₀ ∈ {0.01, 0.1, 0.5} as before. Of the
> solutions that converge, return the one whose φ is closest to the peak on the
> log₁₀ scale.

Note what this does and does not touch. It changes *which of several optima is
reported*, never how they are found — so the multi-start fix of sections 2–3
stands unchanged underneath it, and the 13 broken fits are recovered by the
multi-start whichever selection rule is layered on top.

## 7.1 Building the reference value

The peak has to be measured on sites where φ *is* determined, otherwise it
measures the optimiser rather than the dispersion.

```{r peakcon}
PS <- file.path(PSI, "qtl_betabinomial/phisel_compare")
knitr::kable(fread(file.path(PS, "peak_construction.tsv"), sep = "\t"),
  caption = "Sites where all three starts agree on phi are the ones where phi is identified. Over ALL sites the mode is instead 2.4e-6, but 99.5% of the sites below 1e-4 are ones where the starts DISAGREE -- that mode is the PHI_EPS = 1e-6 floor showing through, not a fact about dispersion.")
```

```{r peakstab}
knitr::kable(fread(file.path(PS, "peak_stability.tsv"), sep = "\t"),
  caption = "The peak is solid to about one significant figure -- call it phi = 0.013 -- and the second figure is not meaningful. That is enough here, because the competing candidates at an ambiguous site are typically orders of magnitude apart, so a 10% wobble in the reference cannot flip a decision.")
```

## 7.2 The φ distributions

The left and middle panels are the reason the peak has to be measured on the
agreeing subset. The right panel is what the rule actually returns.

```{r phidist, fig.height=4.8, fig.width=13}
PHID <- fread(file.path(PS, "phi_distributions.tsv.gz"), sep = "\t")
VLV <- c("before: MLE (best of 3 starts)",
         "source: all 3 starts agree (builds the peak)",
         "after: the rule applied")
PHID[, view := factor(view, levels = VLV)]
PEAK <- 0.0126; EPS <- 1e-6
ggplot(PHID, aes(phi)) +
  geom_histogram(bins = 60, fill = "#4a7fbf", colour = NA) +
  geom_vline(xintercept = PEAK, colour = "#2c8a4a", linewidth = 0.7) +
  geom_vline(xintercept = EPS, colour = "grey35", linetype = "dotted", linewidth = 0.5) +
  facet_wrap(~ view, nrow = 1, scales = "free_y") +
  scale_x_log10(breaks = 10^seq(-6, 0, 2), labels = c("1e-6", "1e-4", "1e-2", "1")) +
  labs(x = expression(hat(varphi)~"(log scale)"), y = "sites",
       title = "Where the reference value comes from, and what the rule returns",
       subtitle = paste(
         "Green = the peak, 0.0126. Dotted = the PHI_EPS = 1e-6 numerical floor. Same 10,450 sites in the outer two panels.",
         "LEFT: under the MLE half the sites pile against the floor. At an ambiguous site the winning start is decided by loglik differences of ~1e-9, so that pile is arbitrary.",
         "MIDDLE: restricting to the sites where all 3 starts agree removes the pile and leaves one clean hump. The peak is the mode of this panel.",
         "RIGHT: the rule clears the floor and moves the low mode up ~20x, but the shape stays bimodal -- it does NOT reproduce the middle panel.", sep = "\n")) +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank(), strip.text = element_text(size = 9),
        plot.subtitle = element_text(size = 8))
```

```{r phidsum}
knitr::kable(fread(file.path(PS, "phi_distribution_summary.tsv"), sep = "\t"),
  caption = "The pile at the numerical floor is what the rule removes: 16.3% of sites to 0.9%. But 41.8% still sit below 1e-4 afterwards, against 0.5% in the agreeing subset.")
```

**The rule does not reshape the φ distribution into the middle panel, and it was
never going to.** It chooses among the three optima the starts actually found, so
where all three are low it returns a low value — the nearest available candidate,
not the peak itself. What it does achieve is removing the arbitrary pile at the
`PHI_EPS = 1e-6` floor, from 16.3% of sites to 0.9%, and lifting the low mode from
2.4×10⁻⁶ to 5.2×10⁻⁵.

That is worth being clear about, because it bounds what the rule can be claimed
to do: it replaces an arbitrary choice with a defensible one at the sites where a
choice exists, and it does not manufacture dispersion where the candidate set
offers none. Section 5 still applies — at four sites in five there is no
defensible point estimate of φ, and no selection rule can create one.

```{r ruledi}
knitr::kable(fread(file.path(PS, "rule_divergence.tsv"), sep = "\t"),
  caption = "The rule changes phi at about half of all sites -- that is expected, since that is how often phi is ambiguous -- but it OVERRULES the likelihood at only 63, and gives up more than one log unit at only 3.")
```

At the 13 sites from section 3, the rule recovers the maximum-likelihood answer
at 11, including every site where the optimiser had run off to φ = 1:

```{r sites13}
knitr::kable(fread(file.path(PS, "sites13_by_rule.tsv"), sep = "\t"),
  caption = "The two misses are the last two rows, where the candidate nearer the peak is the worse-fitting one and choosing it gives up 1.9-2.7 log units. Note that phi ~ 0.0013 IS closer to the genome-wide peak than phi ~ 0.556 -- by the heuristic's own logic it is the more reasonable value, and it is only wrong by the likelihood's standard.")
```

## 7.3 Does it calibrate?

This is the criterion that matters, and it is the one the 13-site scorecard
cannot answer: the rule re-picks φ across roughly half the genome, so only a full
real-plus-permutation run shows what that does.

```{r phiselcal}
RL <- c("1 start (historical)", "MLE (3 starts)", "peak")
knitr::kable(fread(file.path(PS, "phisel_headline.tsv"), sep = "\t")[run %chin% RL][match(RL, run)],
  caption = "All three runs: k = 3, +/-5 kb, chr1-22, U/C sites removed, phi-boundary sites RETAINED, same 5 permutation seeds. The ONLY difference is how phi is chosen.")
```

**The rule calibrates exactly as well as the MLE.** Zero problematic sites, real
tail 28.2 and worst permuted tail 11.3 in both cases — identical to the decimal,
and each leaves a single permutation candidate that fails the real-data half of
the criterion. The 63 sites where the rule overrules the likelihood are nowhere
near the tail, so the log-likelihood given up there never converts into a false
positive.

```{r qqphisel, fig.height=4.6, fig.width=13}
qs <- fread(file.path(PS, "qq_thinned_phisel.tsv.gz"), sep = "\t")[run %chin% RL]
qs[, run := factor(run, levels = RL)]
qs[, kind := ifelse(dataset == "real", "real", "permuted")]
ggplot(qs, aes(expected, observed, group = dataset, colour = kind, shape = kind)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey55", linewidth = 0.3) +
  geom_point(size = 0.5, alpha = 0.55) +
  facet_wrap(~ run, nrow = 1) +
  scale_colour_manual(values = c(real = "#1f4e9c", permuted = "#c0392b"), name = NULL) +
  scale_shape_manual(values = c(real = 16, permuted = 17), name = NULL) +
  coord_cartesian(ylim = c(0, 45)) +
  guides(colour = guide_legend(override.aes = list(size = 2.5, alpha = 1))) +
  labs(x = expression(Expected~-log[10](p)), y = expression(Observed~-log[10](p)),
       title = "Calibration: single start vs the MLE vs the peak rule (y clipped at 45)",
       subtitle = paste("Every p-value in the tail is drawn -- all 2,000 smallest per dataset; the bulk is thinned. Points, not lines:",
                        "a line interpolates across the gaps and renders the artefact tail as a smooth ramp, hiding that it is a few",
                        "discrete clusters of SNPs at a handful of broken sites. Clipped points are outside the view, not deleted.", sep = "\n")) +
  theme_bw(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        strip.text = element_text(size = 9))
```

The counts behind those tails:

```{r tailcnt}
knitr::kable(dcast(fread(file.path(PS, "tail_counts.tsv"), sep = "\t")[run %chin% RL],
                   run ~ kind, value.var = "N")[match(RL, run)],
  caption = "Number of p-values reaching 1e-10. Under the historical fitter over five thousand PERMUTED p-values clear it; under the corrected fitter exactly one does. The real counts barely change (2,026 -> 2,010), so the fix removes false positives rather than suppressing signal generally.")
```

## 7.4 What is adopted

Like the multi-start, the rule is switched on by an environment variable so the
default reproduces every pre-existing result exactly:

```r
BB_PHI_SELECT <- Sys.getenv("BB_PHI_SELECT", "loglik")   # betabin_lib.R
BB_PHI_PEAK   <- 0.0126
# "loglik"   the MLE -- the historical behaviour, bit-for-bit
# "peak"     nearest 0.0126 among the converged starts
```

The peak rule is adopted for production. Three points are worth keeping in view:

- **It is not a substitute for fixing the optimiser.** Multi-start is what
  recovers the 13 broken fits; the peak rule then resolves which of several
  equally-likely φ to report at the sites where the data cannot say. Applied on
  top of a single start it would not have helped, since a single start produces
  only one candidate to choose from.
- **Its worst-case cost is bounded and small.** Across the genome the most
  log-likelihood it gives up at any site is 4.4, against the 20–320 units the
  single-start failure was losing.
- **It does not make φ identifiable.** It removes the arbitrariness in which
  value gets reported; it does not add information the data do not contain. Any
  downstream use of φ as a per-site quantity still has to contend with section 5.

<!-- # Caveats -->

<!-- - **Three starts is a hedge, not a proof.** A site whose likelihood has a third -->
<!--   mode outside {0.01, 0.1, 0.5} would still be missed. What makes the result -->
<!--   credible here is that starts 2 and 3 *agree* (φ = 0.699145 vs 0.699146), and -->
<!--   that no site anywhere produced a three-way split. -->
<!-- - **`fit_bb_mle`'s zero-gradient fallback is untouched.** It does not fire at -->
<!--   these sites, but returning a zero gradient on a non-finite value is a hazard -->
<!--   that should be replaced with a finite-difference fallback. -->
<!-- - **The audit covers the null fit only**, at k = 3 on one phenotype tree. The -->
<!--   per-SNP full-model fits were not audited site by site. -->
<!-- - **275 sites had at least one start fail outright**; the multi-start recovers a -->
<!--   usable fit for some but not all of them. -->

<!-- # Provenance -->

<!-- ```{r prov} -->
<!-- knitr::kable(data.table( -->
<!--   step = c("genome-wide single vs multi-start audit", -->
<!--            "agreement among the 3 starts", -->
<!--            "can the data reject phi = 0 or phi = 1", -->
<!--            "the fix", "calibration re-run"), -->
<!--   script = c("24.multistart_audit.R", "25.multistart_agreement.R", -->
<!--              "scratch profile-likelihood pass", -->
<!--              "betabin_lib.R (BB_PHI_STARTS; backup at betabin_lib.R.bak_presinglestart)", -->
<!--              "run_calibration_multistart.sbatch -> 8.calibration_site_resolved.R")), -->
<!--   caption = "Outputs under qtl_betabinomial/multistart_audit/.") -->
<!-- ``` -->

<!-- Related: [filtering strategy](qtl_mapping_betabinomial_strategy.html) · -->
<!-- [how many covariate PCs](qtl_mapping_betabinomial_npcs_tradeoff.html) · -->
<!-- [diagnosing the problematic sites](qtl_mapping_betabinomial_problematic_diagnosis.html) · -->
<!-- [beta-binomial model](qtl_mapping_betabinomial_model.html) -->
