---
title: "QTL mapping: beta-binomial top-hit diagnostic"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })

knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", fig.width = 12, fig.height = 14
)

PROJECT_DIR <- "/project/xinhe/xsun/pumseq"
BASE   <- file.path(PROJECT_DIR, "pumseq_processed/psi_qtl")
BBR    <- file.path(BASE, "qtl_betabinomial/lrt_results_merged")
COUNTS <- file.path(BASE, "phenotype/fold5_union/counts")
PSAM   <- file.path(BASE, "genotype/plink2/pumseq_genotype.hg38.chr1-22.psam")
TRAW   <- file.path(BASE, "genotype/qtl_binomial/genotype_common.traw")

WINDOW_KB <- 5
NPCS      <- 3
N_TOP     <- 20
MIN_GRP   <- 5   # flag threshold for "small genotype group" in the pattern summary below
```


A number of sites reach implausibly small p-values
(`-log10(p) > 100`) given the small sample size (50 cell lines). 

We pick a few top hits and plot %pU by genotype directly to build intuition for what's driving them.

This page does that for the **beta-binomial LRT model at its best-performing
config (`r WINDOW_KB`kb window, `r NPCS` PCs)** — the top `r N_TOP` sites by
nominal (pre-permutation-correction) LRT p-value.

Two candidate artifacts motivated this check:

1. **Small genotype-group separation** — a genotype class with very few cell
   lines can, by chance, produce a near-perfect fit and an inflated LRT
   statistic, independent of any real effect.
2. **SNP-at-site overlap** — if the tested cis-SNP sits at (or right next to)
   the modification site's own genomic position, an individual's true
   genotype there can directly bias which base gets called at that position,
   masquerading as a huge modification-QTL effect. This is the same
   phenomenon as the well-known SNP-at-CpG confound in bisulfite-based mQTL
   studies.

# 1. SNP-at-site rate among significant hits

For each site, `lead_pos` is the position of the most associated cis-SNP.
When `lead_pos == pos` (the site's own genomic position), the "QTL" SNP is
sitting exactly on top of the profiled base.

```{r snp-at-site}
d <- fread(file.path(BBR, sprintf("scan_lrt_w%dkb_pc%d.txt.gz", WINDOW_KB, NPCS)))
d[, snp_at_site := lead_pos == pos]

sig <- d[fdr < 0.05]
tbl <- data.table(
  group = c("All tested sites", "Significant (FDR<0.05)"),
  n              = c(nrow(d), nrow(sig)),
  n_snp_at_site  = c(sum(d$snp_at_site, na.rm = TRUE), sum(sig$snp_at_site, na.rm = TRUE))
)
tbl[, pct_snp_at_site := round(100 * n_snp_at_site / n, 1)]
knitr::kable(tbl, caption = sprintf(
  "SNP-at-site rate: baseline vs. significant hits (beta-binomial, %dkb, %d PCs)",
  WINDOW_KB, NPCS))
```

The SNP-at-site rate is **`r tbl$pct_snp_at_site[2]`%** among significant hits vs.
a **`r tbl$pct_snp_at_site[1]`%** baseline among all tested sites — an
**`r round(tbl$pct_snp_at_site[2] / tbl$pct_snp_at_site[1], 1)`x enrichment**.
This alone means a large share of current significant calls at this config are
sites where the tested variant overlaps the profiled base itself, not a
genuine cis-regulatory effect.

# 2. Top `r N_TOP` QTLs: %pU by genotype at the lead cis-SNP

```{r load-pheno-geno}
C <- fread(file.path(COUNTS, "C_matrix.txt.gz")); N <- fread(file.path(COUNTS, "N_matrix.txt.gz"))
psam <- fread(PSAM); go <- psam[[2]]
Cm <- as.matrix(C[, ..go]); Nm <- as.matrix(N[, ..go])
sites <- C[, .(ref = as.character(ref), pos = as.integer(pos))]

traw <- fread(TRAW); samp <- paste0(go, "_", go)
Gall <- t(as.matrix(traw[, ..samp])); gid <- traw$SNP
```

```{r top20-select}
top20 <- d[order(nominal_min_p)][1:N_TOP]
top20[, rank := .I]

rows <- rbindlist(lapply(seq_len(nrow(top20)), function(i) {
  s <- top20[i]
  si <- which(sites$ref == as.character(s$ref) & sites$pos == s$pos)
  gi <- which(gid == s$lead_snp)
  y <- Cm[si, ]; n <- Nm[si, ]; g <- Gall[, gi]
  pi_v <- y / n
  grp_tab <- table(g)
  min_grp <- min(grp_tab)
  flags <- paste0(
    if (s$lead_pos == s$pos) "\n[SNP-at-site]" else "",
    if (min_grp < MIN_GRP) sprintf("\n[small group n=%d]", min_grp) else ""
  )
  data.table(
    rank = s$rank,
    site_lab = sprintf("#%d chr%s:%.0f\n-log10(p)=%.0f%s",
                        s$rank, s$ref, s$pos, -log10(max(s$nominal_min_p, 1e-300)), flags),
    g = factor(g, levels = 0:2), pi = pi_v, n = n,
    snp_at_site = s$lead_pos == s$pos, min_grp = min_grp
  )
}))
rows[, site_lab := factor(site_lab, levels = unique(site_lab[order(rank)]))]

blue3 <- c(`0` = "#86b6ef", `1` = "#2a78d6", `2` = "#104281")

ggplot(rows, aes(x = g, y = pi, color = g)) +
  geom_boxplot(aes(group = g), outlier.shape = NA, width = 0.4, color = "grey60", fill = NA) +
  geom_jitter(aes(size = n), width = 0.12, alpha = 0.75) +
  scale_color_manual(values = blue3, name = "genotype\n(alt-allele copies)") +
  scale_size_continuous(name = "depth (n)", range = c(0.6, 3.5)) +
  facet_wrap(~ site_lab, scales = "free_x", ncol = 4) +
  labs(x = "genotype (copies of alt allele) at the lead cis-SNP", y = "%pU (y/n per cell line)",
       title = sprintf("Top %d beta-binomial QTLs by nominal LRT p-value (%dkb, %d PCs)",
                       N_TOP, WINDOW_KB, NPCS),
       subtitle = "Ranked by nominal (pre-permutation) p-value, most extreme first. '[SNP-at-site]' = lead SNP sits exactly at the profiled base; '[small group n=X]' = smallest genotype class has X < 5 cell lines.") +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", strip.text = element_text(size = 7.5))
```

# 3. Pattern summary across these 20 sites

```{r pattern-summary}
summ <- unique(rows[, .(rank, snp_at_site, min_grp)])
summ[, category := fifelse(snp_at_site, "SNP at/on the modification site",
                     fifelse(min_grp < MIN_GRP, sprintf("Small genotype group (min class n<%d)", MIN_GRP),
                             "Neither flag (plausible)"))]
knitr::kable(summ[order(rank)], caption = "Per-site flags among the top 20")

cat_tbl <- summ[, .N, by = category][order(-N)]
knitr::kable(cat_tbl, caption = "Category counts among the top 20")
```

Of the top `r N_TOP` beta-binomial hits by nominal p-value,
**`r cat_tbl[category == "SNP at/on the modification site"]$N`** have their
lead SNP sitting exactly at the modification site itself, and
**`r sum(summ$min_grp < MIN_GRP & !summ$snp_at_site)`** more have a genotype
class with fewer than `r MIN_GRP` cell lines (and no SNP-at-site overlap) —
together accounting for
**`r sum(summ$snp_at_site | summ$min_grp < MIN_GRP)`/`r N_TOP`** of the most
extreme "hits" in this model. 

Visually, nearly every panel above shows the
same pattern: genotype 0 near %pU=1, genotype 1 near
%pU=0.5, genotype 2 near %pU=0.

<!-- **The remaining `r cat_tbl[category == "Neither flag (plausible)"]$N` sites -->
<!-- are not a clean "these are fine" bucket.** Most of them (#1, #4, #6, #12 — -->
<!-- none of these are on chr6) show the same staircase -->
<!-- despite having normal-sized genotype groups and no SNP-at-site overlap, -->
<!-- meaning the two mechanisms characterized here do not explain everything — -->
<!-- something else (e.g. a nearby-but-not-exact-position variant still perturbing -->
<!-- the assay's base-calling, or another unmodeled technical effect) is likely -->
<!-- producing the same signature. Only two sites (#14 `chr3:37277733`, #18 -->
<!-- `chr2:88886147`) look qualitatively different: noisy, near-uniform scatter -->
<!-- across all three genotype groups with no clean separation, or driven by a -->
<!-- single outlier cell line — closer to what a null or borderline result should -->
<!-- look like, and a useful visual contrast against the rest of this page. So the -->
<!-- two flagged mechanisms are a partial, not complete, explanation, and the -->
<!-- planned fixes (excluding cis-SNPs that overlap the profiled base, requiring a -->
<!-- minimum sample count per genotype class) should be treated as a first pass, -->
<!-- re-checked against a fresh top-20 plot after re-running. -->
