---
title: "QTL mapping: beta-binomial model"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r setup, warning=FALSE, message=FALSE}
library(data.table)
library(ggplot2)

knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", fig.width = 10, fig.height = 6
)

PROJECT_DIR <- "/project/xinhe/xsun/pumseq"
BB  <- file.path(PROJECT_DIR, "pumseq_processed/psi_qtl/qtl_betabinomial")
BBR <- file.path(BB, "lrt_results_merged")
BIN <- file.path(PROJECT_DIR, "pumseq_processed/psi_qtl/qtl_binomial/lrt_results_merged")
WINDOWS_KB <- c(5, 20, 50, 100)
PC_COUNTS  <- c(0, 3, 5, 7, 10, 15, 20)
```

# Introduction

This page tests the **beta-binomial**, which models that overdispersion directly:

$$
y_i \sim \text{BetaBinomial}(n_i, \mu_i, \phi), \qquad
\text{Var}(y_i) = n_i \mu_i (1-\mu_i)\,[\,1 + (n_i - 1)\phi\,],
$$

with the same mean model `logit(mu_i) = beta_0 + beta_g G_i + z_i'gamma`, the
same 5-fold union sites (12,416), the same 50 cell lines, and the same M-value
PC covariates. `phi` is the intra-class-correlation dispersion (`phi = 0`
recovers the binomial).

**The genotype effect `beta_g` is tested with a likelihood-ratio test (LRT)**,
matching the binomial model's approach — see the [model
page](qtl_mapping_betabinomial_model.html). **The results below
use `NPERM=100`** (not the full 1000) — a preliminary pass, made affordable by
two optimizations to the beta-binomial fitter (dropping a constant
term from the optimizer's objective + supplying an analytic gradient, ~2.5-2.9x
speedup) plus running fewer permutations to first rank PC/window configs and
check calibration
<!-- (verified via Monte Carlo to stay unbiased with ~2x the -->
<!-- noise of 500 permutations — adequate for this purpose, not for final -->
<!-- per-site significance calls).  -->
Permutations will be increased once a final
config is chosen.

# 1. Results: QTL sites across the full PC range

```{r load-bb}
read_bb <- function(w, k) {
  f <- file.path(BBR, sprintf("scan_lrt_w%dkb_pc%d.txt.gz", w, k))
  if (!file.exists(f)) return(NULL)
  d <- fread(f); d[, `:=`(window_kb = w, n_pcs = k)]; d
}
grid <- CJ(w = WINDOWS_KB, k = PC_COUNTS)
bb <- rbindlist(mapply(read_bb, grid$w, grid$k, SIMPLIFY = FALSE), use.names = TRUE, fill = TRUE)
bb_sum <- bb[, .(n_tested = .N, qtl_FDR05 = sum(fdr < 0.05, na.rm = TRUE)),
             by = .(window_kb, n_pcs)][order(window_kb, n_pcs)]
bb_sum[, window_lab := factor(paste0(window_kb, "kb"), levels = paste0(WINDOWS_KB, "kb"))]

knitr::kable(dcast(bb_sum, window_kb ~ n_pcs, value.var = "n_tested"),
             caption = "Sites the beta-binomial could fit (rows=window kb, cols=PCs)")
```

<!-- The beta-binomial MLE jointly estimates `k + 2` parameters (PCs + intercept + -->
<!-- dispersion) on only 50 samples. An earlier version of this pipeline (relying -->
<!-- on `aod::betabin`'s optimizer) appeared to stop converging for most sites -->
<!-- once k >= 10 (fittable sites collapsing from thousands to tens). Replacing -->
<!-- that optimizer with a properly-seeded fit (see the [model -->
<!-- page](qtl_mapping_betabinomial_model.html)) resolved this — sites fittable at -->
<!-- k=20 rose to ~7,600 (close to the binomial's ~7,900 at the same k). The -->
<!-- beta-binomial is now reliably fittable across the **full swept range**, -->
<!-- k=3-20. -->

```{r bb-disc}
knitr::kable(dcast(bb_sum, window_kb ~ n_pcs, value.var = "qtl_FDR05"),
             caption = "Beta-binomial QTL sites (FDR<0.05)")
```

```{r bb-vs-pc, fig.height=5}
ggplot(bb_sum, aes(x = n_pcs, y = qtl_FDR05, color = window_lab)) +
  geom_line() + geom_point(size = 2) +
  scale_x_continuous(breaks = PC_COUNTS) +
  labs(x = "Number of covariate PCs", y = "QTL sites (FDR<0.05)", color = "cis-window",
       title = "Beta-binomial QTL sites vs. PCs") +
  theme_bw(base_size = 13)
```

Unlike the [binomial model](qtl_mapping_summary_binomial.html#4_Results:_QTL_sites_by_PC_count_and_window_size)
— where discovery counts rise monotonically as PCs are removed, peaking at
k=0 (intercept-only) — the beta-binomial's counts **peak at k=3** (1,360 at
5kb) and k=0 is slightly *lower* (1,323). k=0 also has noticeably fewer
fittable sites (9,885 vs. 11,112 at k=3, 5kb) — some sites' beta-binomial null
fit doesn't converge without any PC covariates. Both observations point the
same way: the beta-binomial's dispersion term is already absorbing whatever
confounding the binomial's inflation at k=0 was catching, so a few PCs still
help here even though they don't for the binomial. Calibration has only been
checked at k=3 (Section 3), not at k=0.

# 2. Beta-binomial vs. binomial

At **k = 3**, comparing both models' current LRT results:

```{r head-to-head}
read_bin <- function(w, k) {
  f <- file.path(BIN, sprintf("scan_lrt_w%dkb_pc%d.txt.gz", w, k))
  if (!file.exists(f)) return(NULL)
  fread(f)
}
h2h <- rbindlist(lapply(WINDOWS_KB, function(w) {
  db <- read_bb(w, 3); dn <- read_bin(w, 3)
  data.table(window = paste0(w, "kb"),
             binomial   = sum(dn$fdr < 0.05, na.rm = TRUE),
             betabinom  = sum(db$fdr < 0.05, na.rm = TRUE))
}))
h2h[, `fold change` := round(betabinom / binomial, 1)]
knitr::kable(h2h, caption = "QTL sites (FDR<0.05) at k=3: binomial LRT vs. beta-binomial LRT")
```

<!-- The beta-binomial finds substantially more sites than the binomial at every -->
<!-- window — a much larger gap than earlier (pre-LRT, score-test-based) estimates -->
<!-- of this comparison ever showed (~1.3-1.4x). Plausible in principle: the -->
<!-- beta-binomial's calibration is markedly better (lambda~2.47 vs. the -->
<!-- binomial's ~5.55 — see Section 3), so it may genuinely recover real signal -->
<!-- the binomial's model-misspecification noise was burying at overdispersed -->
<!-- sites, similar to the mechanism established earlier in this project. **This -->
<!-- specific magnitude has not yet been independently spot-checked** the way the -->
<!-- score-vs-LRT comparison was for the binomial model — worth verifying a sample -->
<!-- of beta-binomial-only hits before treating this fold-change as confirmed, -->
<!-- rather than just plausible. -->

<!-- Note: the current LRT scan output does not carry each site's fitted `phi` -->
<!-- (unlike the earlier score-test-based scan), so the dispersion-of-hits -->
<!-- breakdown from the previous version of this page isn't reproducible from the -->
<!-- current results without adding that column back to the scan and re-running. -->

# 3. Calibration check: the best-calibrated result in this project

We ran the same permutation calibration check used for the binomial (5 seeds,
same config: 5kb, 3 PCs), testing the LRT — a full beta-binomial refit per
site per permutation 
<!-- (`phi` re-estimated fresh each time). Model details: -->
<!-- [model page](qtl_mapping_betabinomial_model.html). -->

```{r bb-calib-lambda}
CAL <- file.path(BB, "lrt_calibration")
lam <- fread(file.path(CAL, "calib_lrt_w5kb_pc3_lambda.txt"))
knitr::kable(lam, digits = 2,
             caption = "Genomic-inflation lambda (5kb, 3 PCs): real vs. 5 permuted nulls (beta-binomial LRT)")
```

```{r bb-calib-qq, results='asis'}
cat(sprintf('<img src="%s" style="max-width:70%%;">',
            knitr::image_uri(file.path(CAL, "calib_lrt_w5kb_pc3_qq.png"))))
```

```{r bb-calib-compare}
knitr::kable(data.table(
  model = c("binomial LRT", "beta-binomial LRT"),
  real_lambda = c(5.55, round(lam[dataset == "real"]$lambda, 2)),
  permuted_null_lambda_range = c("2.93 - 3.44", sprintf("%.2f - %.2f",
                                  min(lam[dataset != "real"]$lambda),
                                  max(lam[dataset != "real"]$lambda)))
), caption = "Nominal-p calibration: binomial LRT vs. beta-binomial LRT")
```

<!-- **The beta-binomial LRT brings the permuted-null lambda down to ~1.08** (range -->
<!-- 0.96-1.19 across 5 seeds) — very close to the ideal 1, and the best-calibrated -->
<!-- test/model combination found across this entire project (binomial LRT sits at -->
<!-- ~3.28; both score tests were worse still). The residual gap between the -->
<!-- binomial's calibration and the beta-binomial's is consistent with real, -->
<!-- uncorrected overdispersion in the binomial model (~22% of sites are -->
<!-- significantly overdispersed by a separate test), not a fixable numerical -->
<!-- issue in either test — see the [model -->
<!-- page](qtl_mapping_betabinomial_model.html) for the full reasoning. -->

<!-- **Practical takeaway**: the beta-binomial LRT is the model/test combination to -->
<!-- report — closest to calibrated of everything tried, though the current -->
<!-- discovery numbers (Sections 1-2) are from a **preliminary NPERM=100 pass** and -->
<!-- the binomial-vs-beta-binomial fold-change is not yet independently verified -->
<!-- at this magnitude. Both should be revisited once permutations are increased -->
<!-- at the final chosen config. -->
