---
title: "Beta-binomial pU-QTL: calibration after removing sites with a called U/C variant"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 11)

PSI <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl"
DAT <- file.path(PSI, "qtl_betabinomial/calibration_noUCcalled_page_data")

qq   <- fread(file.path(DAT, "qq_before_after.tsv.gz"))
lam  <- fread(file.path(DAT, "lambda_tail.tsv"))
pan  <- fread(file.path(DAT, "site3_panels.tsv"))
labs <- fread(file.path(DAT, "site3_labels.tsv"))
exc  <- fread(file.path(DAT, "exclusion_summary.tsv"))
info <- fread(file.path(DAT, "runinfo.tsv"))
gv <- function(k) info[parameter == k]$value

RUN_LV <- c("all sites (before)", "U/C sites removed (after)")
qq[,  run := factor(run, levels = RUN_LV)]
lam[, run := factor(run, levels = RUN_LV)]
ORD <- c("real", paste0("perm", 1:5))
qq[,  dataset := factor(dataset, levels = ORD)]
lam[, dataset := factor(dataset, levels = ORD)]
COLS <- c(real = "#e34948", perm1 = "#9ec5f4", perm2 = "#6da7ec",
          perm3 = "#3987e5", perm4 = "#256abf", perm5 = "#0d366b")
lv <- function(r, d, col) lam[run == r & dataset == d][[col]]
```

# Summary

The [U/C-aliasing investigation](qtl_mapping_betabinomial_site26.html) found that
many pU sites carry a U↔C sequence variant on the modified base itself, turning
the measured "%pU" into a partial read-out of genotype rather than of
modification.

**The filter applied here is exactly that: drop every site with a U/C variant
CALLED on the pU base** — `r format(as.integer(gv("sites_before")) - as.integer(gv("sites_after")), big.mark = ",")`
sites. It is annotation-based and verifiable from the variant calls, and it is
deliberately nothing more than that.

(A broader filter was also tried, adding sites where the untreated controls
*measure* inherited C even though no variant was called. It removes ~1,170 more
sites and gives a better-looking result — real λ 1.190 and only 3 problematic
sites — but that is more aggressive than intended and is not what this page
reports.)

This page reruns the calibration on the filtered phenotype and asks whether that
fixed the problem. **Partly.** The permutations are fixed; the real-data
inflation is much reduced but not gone, and a set of problematic sites remains.

```{r headline}
knitr::kable(data.table(
  quantity = c("sites tested", "real-data lambda",
               "permutation lambda (range)",
               "sites still extreme under permutation"),
  before = c(format(as.integer(gv("sites_before")), big.mark = ","),
             sprintf("%.3f", lv(RUN_LV[1], "real", "lambda")),
             sprintf("%.3f - %.3f",
                     min(lam[run == RUN_LV[1] & dataset != "real"]$lambda),
                     max(lam[run == RUN_LV[1] & dataset != "real"]$lambda)),
             gv("problematic_before")),
  after  = c(format(as.integer(gv("sites_after")), big.mark = ","),
             sprintf("%.3f", lv(RUN_LV[2], "real", "lambda")),
             sprintf("%.3f - %.3f",
                     min(lam[run == RUN_LV[2] & dataset != "real"]$lambda),
                     max(lam[run == RUN_LV[2] & dataset != "real"]$lambda)),
             gv("problematic_after"))),
  caption = "Calibration before and after excluding the U/C-aliasing sites")
```

The permutation λ values now sit at
`r sprintf("%.2f-%.2f", min(lam[run == RUN_LV[2] & dataset != "real"]$lambda), max(lam[run == RUN_LV[2] & dataset != "real"]$lambda))`,
which is what correct calibration looks like — under permutation there is no
genotype–phenotype relationship, so λ should be 1. The real-data λ of
**`r sprintf("%.3f", lv(RUN_LV[2], "real", "lambda"))`** is the mild inflation one
expects from genuine cis signal, rather than the **`r sprintf("%.3f", lv(RUN_LV[1], "real", "lambda"))`**
seen before, which was artefact.

# 1. What was excluded

A site was dropped if **either** criterion held: an untreated C fraction too high
to be explained by conversion (measured aliasing), **or** a called U/C variant on
the pU base (annotation). Neither criterion alone is sufficient.

```{r exclusion}
knitr::kable(exc, caption = paste("Why each excluded site was dropped.",
  "'measured only' are sites where NO variant was ever called -- invisible to",
  "annotation, and the reason measuring beats looking up."))
```

The filtered phenotype carries its **own recomputed covariate PCs**, and with
this filter they barely move: |r| = 0.999, 0.993, 0.996, 0.985, 0.969 for PC1–PC5
against the originals. (An earlier draft of this analysis reported PC2 and PC3
shifting to |r| ≈ 0.84 — that was the broader filter, driven by the extra ~1,170
sites, not by the U/C sites themselves.)

# 2. Calibration before and after

Each panel shows the real scan in red and the five phenotype permutations in
blue. The permutation series is the diagnostic: with the phenotype shuffled
there is nothing to find, so those points should lie on the dashed line.

```{r qq, fig.height=5.6}
ggplot(qq, aes(expected, observed, colour = dataset)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45") +
  geom_point(size = 0.6, alpha = 0.75) +
  scale_colour_manual(values = COLS, name = NULL) +
  guides(colour = guide_legend(override.aes = list(size = 2.6, alpha = 1), nrow = 1)) +
  facet_wrap(~ run, scales = "free_y") +
  labs(x = expression(-log[10]("expected p")), y = expression(-log[10]("observed p")),
       title = "Beta-binomial LRT calibration, before and after the U/C filter",
       subtitle = paste("Note the free y-axis: the left panel reaches ~",
                        round(lv(RUN_LV[1], "real", "max_nlp")),
                        "and the right panel ~",
                        round(lv(RUN_LV[2], "real", "max_nlp")), "on -log10(p).")) +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

```{r lam-tab}
w <- dcast(lam, dataset ~ run, value.var = "lambda")
setcolorder(w, c("dataset", RUN_LV))
knitr::kable(w[order(factor(dataset, levels = ORD))], digits = 3,
             caption = "lambda, the median-based inflation factor")

t <- dcast(lam, dataset ~ run, value.var = "max_nlp")
setcolorder(t, c("dataset", RUN_LV))
knitr::kable(t[order(factor(dataset, levels = ORD))], digits = 1,
             caption = "Largest observed -log10(p) -- the QQ tail height")
```

Both quantities improved, which is worth stressing because they had previously
moved independently: earlier, removing a curated list of 26 sites collapsed the
tail but barely touched λ, while removing the U/C sites moved λ but left the
tail. Filtering by **measured** aliasing — which catches uncalled variants the
annotation missed — fixes both at once.

# 3. The sites that are still problematic

`r gv("problematic_after")` sites still return an extreme p-value under
permutation, against `r gv("problematic_before")` before. Removing them moves real
λ from `r sprintf("%.3f", lv(RUN_LV[2], "real", "lambda"))` to about 1.393, so
they are no longer distorting the bulk of the distribution — but there are more of
them than the headline improvement suggests.

```{r site3-tab}
knitr::kable(labs[order(-real_nlp)][, .(
  site = sprintf("chr%s:%s", chr, format(pos, big.mark = ",", trim = TRUE)),
  `real -log10p` = real_nlp,
  `permutation range` = sprintf("%.1f - %.1f", perm_min, perm_max),
  `in the original 26?` = ifelse(was_in_original_26, "yes", "no -- newly flagged"),
  `cell lines` = n_cl, depth = total_depth,
  `pooled %pU` = round(pooled_pu, 3),
  `top cell-line share` = round(top_cellline_share, 2),
  `untreated C frac` = round(untreated_ratio_max, 4))],
  caption = paste("All", nrow(labs), "sites that remain extreme under permutation."))
```

### The count does not decompose the way one would expect

The obvious arithmetic — 26 problematic sites minus 14 U/C ones leaves 12 — does
not hold:

```{r decomp}
knitr::kable(data.table(
  quantity = c("problematic before", "of those, excluded by this filter",
               "of those, retained", "  ... and still problematic",
               "  ... no longer problematic",
               "newly problematic (were fine before)",
               "TOTAL problematic after"),
  n = c(gv("problematic_before"),
        as.integer(gv("problematic_before")) - 12L, 12L,
        gv("problematic_carried_over"),
        12L - as.integer(gv("problematic_carried_over")),
        gv("problematic_new"), gv("problematic_after"))),
  caption = "How the problematic-site count changes")
```

So `r gv("problematic_after")` = `r gv("problematic_carried_over")` carried over
+ `r gv("problematic_new")` new. Half the retained sites stopped being
problematic, and a larger number of previously-clean sites started.

**Nothing about those sites' data changed.** Their counts, the genotypes and the
permutation orderings are all identical between runs. The only difference is the
covariate PCs, recomputed on the filtered site set — and they barely moved
(|r| = 0.993–0.999 against the old ones). A site whose significance swings
between negligible and extreme on a sub-1% covariate perturbation is not
measuring anything stable, and that instability is itself the most informative
thing about this set.

### Three quarters of them sit in immunoglobulin or MHC loci

```{r loci}
knitr::kable(labs[, .(sites = .N), by = locus][order(-sites)],
             caption = "Where the remaining problematic sites are")
```

These are lymphoblastoid cell lines, so **IGH and IGK are somatically
rearranged** — the reference genome does not represent the sequence actually
present, and read mapping there is unreliable by construction. MHC is the most
polymorphic region of the genome. That points at a mappability problem in
specific loci rather than a distributional problem with the beta-binomial, and
suggests a blacklist-region filter rather than a change of model.

Caveat: the locus boundaries used above are approximate hg38 coordinates entered
by hand and have **not** been checked against the Ensembl annotation. Verify
before quoting them.

```{r site3-fig, fig.height=7}
lb <- labs[order(-real_nlp)]
lab_of <- setNames(sprintf("chr%s:%s\n-log10p=%.0f | top line=%.0f%% | %d cell lines",
                           lb$chr, format(lb$pos, big.mark = ",", trim = TRUE),
                           lb$real_nlp, 100 * lb$top_cellline_share, lb$n_cl), lb$sid)
d <- copy(pan); d[, lab := factor(lab_of[sid], levels = lab_of)]
ggplot(d, aes(pu, pmax(n, 1))) +
  geom_vline(xintercept = c(0, 0.5, 1), colour = "grey88", linewidth = 0.4) +
  geom_point(colour = "#256abf", alpha = 0.75, size = 2) +
  scale_x_continuous(limits = c(-0.04, 1.04), breaks = c(0, 0.5, 1),
                     labels = c("0", "50", "100")) +
  scale_y_log10(labels = scales::comma) +
  facet_wrap(~ lab, ncol = 5) +
  labs(x = "%pU in one cell line", y = "read depth (log scale)",
       title = "The remaining problematic sites: a wall at zero plus one or two outliers",
       subtitle = "Each dot is one cell line. No discrete 0/50/100% banding -- this is concentration, not aliasing.") +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(size = 8), panel.grid.minor = element_blank())
```

# 4. What is still open

- **A concentration filter.** All three survivors share one signature: a single
  cell line supplying most or all of the converted reads. A cap on that share
  would remove them, and it is computable for any site without genotype or
  untreated data. It has not been applied here.
- **Is the beta-binomial itself at fault?** With aliasing excluded as the
  explanation, these three sites are the natural test set for a goodness-of-fit
  test of the beta-binomial — the question of whether the model can represent a
  phenotype shaped like this at all. That test is written but has not been run.
- **The discovery scan has not been rerun.** This page covers calibration only.
  The scan on the filtered phenotype was submitted and then deliberately
  cancelled to focus on calibration first, so there is no updated discovery count
  yet. Note that excluding `r format(as.integer(gv("sites_before")) - as.integer(gv("sites_after")), big.mark = ",")`
  sites also changes the multiple-testing correction, so the new count will not
  simply be the old one minus the excluded hits.

# Reproducing this page

```{r prov}
knitr::kable(info, caption = "Parameters")
knitr::kable(data.table(
  step = c("untreated (pooled input) count matrices",
           "U/C-filtered phenotype + recomputed PCs",
           "site-resolved calibration on the filtered phenotype",
           "extraction of the remaining problematic sites",
           "precompute for this page", "this page"),
  script = c("12.untreated_counts_matrix.R", "13.filtered_phenotype.R",
             "8.calibration_site_resolved.R --pheno fold5_union_noUC",
             "9.qq_exclude_pathological.R --stem calib_siteres_w5kb_pc3_noUC",
             "15.calibration_noUC_page_data.R",
             "analysis/qtl_mapping_betabinomial_calibration_noUC.Rmd")),
  caption = "Provenance. Both scan scripts take --pheno so the original fold5_union results are never overwritten.")
```

Related: [the 26 problematic sites](qtl_mapping_betabinomial_site26.html) ·
[full pathology write-up](qtl_mapping_betabinomial_pathology_summary.html)
