---
title: "Beta-binomial pU-QTL: calibration after removing the U/C-aliasing sites"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 11)

PSI <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl"
DAT <- file.path(PSI, "qtl_betabinomial/calibration_noUC_page_data")

qq   <- fread(file.path(DAT, "qq_before_after.tsv.gz"))
lam  <- fread(file.path(DAT, "lambda_tail.tsv"))
pan  <- fread(file.path(DAT, "site3_panels.tsv"))
labs <- fread(file.path(DAT, "site3_labels.tsv"))
exc  <- fread(file.path(DAT, "exclusion_summary.tsv"))
info <- fread(file.path(DAT, "runinfo.tsv"))
gv <- function(k) info[parameter == k]$value

RUN_LV <- c("all sites (before)", "U/C sites removed (after)")
qq[,  run := factor(run, levels = RUN_LV)]
lam[, run := factor(run, levels = RUN_LV)]
ORD <- c("real", paste0("perm", 1:5))
qq[,  dataset := factor(dataset, levels = ORD)]
lam[, dataset := factor(dataset, levels = ORD)]
COLS <- c(real = "#e34948", perm1 = "#9ec5f4", perm2 = "#6da7ec",
          perm3 = "#3987e5", perm4 = "#256abf", perm5 = "#0d366b")
lv <- function(r, d, col) lam[run == r & dataset == d][[col]]
```

# Summary

The [U/C-aliasing investigation](qtl_mapping_betabinomial_site26.html) found that
a large share of pU sites carry a U↔C sequence variant on the modified base
itself, which turns the measured "%pU" into a partial read-out of genotype. Those
sites were then identified by **direct measurement** rather than annotation: in
untreated RNA no chemical conversion occurs, so any C observed at a pU base is
inherited.

This page reruns the calibration on the filtered phenotype and asks whether that
fixed the problem.

**It largely did.**

```{r headline}
knitr::kable(data.table(
  quantity = c("sites tested", "real-data lambda",
               "permutation lambda (range)",
               "sites still extreme under permutation"),
  before = c(format(as.integer(gv("sites_before")), big.mark = ","),
             sprintf("%.3f", lv(RUN_LV[1], "real", "lambda")),
             sprintf("%.3f - %.3f",
                     min(lam[run == RUN_LV[1] & dataset != "real"]$lambda),
                     max(lam[run == RUN_LV[1] & dataset != "real"]$lambda)),
             gv("problematic_before")),
  after  = c(format(as.integer(gv("sites_after")), big.mark = ","),
             sprintf("%.3f", lv(RUN_LV[2], "real", "lambda")),
             sprintf("%.3f - %.3f",
                     min(lam[run == RUN_LV[2] & dataset != "real"]$lambda),
                     max(lam[run == RUN_LV[2] & dataset != "real"]$lambda)),
             gv("problematic_after"))),
  caption = "Calibration before and after excluding the U/C-aliasing sites")
```

The permutation λ values now sit at
`r sprintf("%.2f-%.2f", min(lam[run == RUN_LV[2] & dataset != "real"]$lambda), max(lam[run == RUN_LV[2] & dataset != "real"]$lambda))`,
which is what correct calibration looks like — under permutation there is no
genotype–phenotype relationship, so λ should be 1. The real-data λ of
**`r sprintf("%.3f", lv(RUN_LV[2], "real", "lambda"))`** is the mild inflation one
expects from genuine cis signal, rather than the **`r sprintf("%.3f", lv(RUN_LV[1], "real", "lambda"))`**
seen before, which was artefact.

# 1. What was excluded

A site was dropped if **either** criterion held: an untreated C fraction too high
to be explained by conversion (measured aliasing), **or** a called U/C variant on
the pU base (annotation). Neither criterion alone is sufficient.

```{r exclusion}
knitr::kable(exc, caption = paste("Why each excluded site was dropped.",
  "'measured only' are sites where NO variant was ever called -- invisible to",
  "annotation, and the reason measuring beats looking up."))
```

Note also that the filtered phenotype carries its **own recomputed covariate
PCs**. The original PCs were derived from all sites, ~18% of which are genotype
read-outs, so they partly encoded population structure rather than technical
variation. PC2 and PC3 moved substantially (|r| ≈ 0.84 between old and new), and
those are active covariates in this k=3 model.

# 2. Calibration before and after

Each panel shows the real scan in red and the five phenotype permutations in
blue. The permutation series is the diagnostic: with the phenotype shuffled
there is nothing to find, so those points should lie on the dashed line.

```{r qq, fig.height=5.6}
ggplot(qq, aes(expected, observed, colour = dataset)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45") +
  geom_point(size = 0.6, alpha = 0.75) +
  scale_colour_manual(values = COLS, name = NULL) +
  guides(colour = guide_legend(override.aes = list(size = 2.6, alpha = 1), nrow = 1)) +
  facet_wrap(~ run, scales = "free_y") +
  labs(x = expression(-log[10]("expected p")), y = expression(-log[10]("observed p")),
       title = "Beta-binomial LRT calibration, before and after the U/C filter",
       subtitle = paste("Note the free y-axis: the left panel reaches ~",
                        round(lv(RUN_LV[1], "real", "max_nlp")),
                        "and the right panel ~",
                        round(lv(RUN_LV[2], "real", "max_nlp")), "on -log10(p).")) +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

```{r lam-tab}
w <- dcast(lam, dataset ~ run, value.var = "lambda")
setcolorder(w, c("dataset", RUN_LV))
knitr::kable(w[order(factor(dataset, levels = ORD))], digits = 3,
             caption = "lambda, the median-based inflation factor")

t <- dcast(lam, dataset ~ run, value.var = "max_nlp")
setcolorder(t, c("dataset", RUN_LV))
knitr::kable(t[order(factor(dataset, levels = ORD))], digits = 1,
             caption = "Largest observed -log10(p) -- the QQ tail height")
```

Both quantities improved, which is worth stressing because they had previously
moved independently: earlier, removing a curated list of 26 sites collapsed the
tail but barely touched λ, while removing the U/C sites moved λ but left the
tail. Filtering by **measured** aliasing — which catches uncalled variants the
annotation missed — fixes both at once.

# 3. The sites that are still problematic

`r gv("problematic_after")` sites still return an extreme p-value under
permutation, down from `r gv("problematic_before")`. Removing even these moves λ
by less than 0.001, so they no longer distort the bulk of the distribution.

```{r site3-tab}
knitr::kable(labs[order(-real_nlp)][, .(
  site = sprintf("chr%s:%s", chr, format(pos, big.mark = ",", trim = TRUE)),
  `real -log10p` = real_nlp,
  `permutation range` = sprintf("%.1f - %.1f", perm_min, perm_max),
  `in the original 26?` = ifelse(was_in_original_26, "yes", "no -- newly flagged"),
  `cell lines` = n_cl, depth = total_depth,
  `pooled %pU` = round(pooled_pu, 3),
  `top cell-line share` = round(top_cellline_share, 2),
  `untreated C frac` = round(untreated_ratio_max, 4))],
  caption = paste("All", nrow(labs), "sites that remain extreme under permutation."))
```

Three things stand out.

**They are not aliasing.** Untreated C fractions are ~0.01, so the filter was
right to keep them — there is no inherited C at these bases.

**They are the single-cell-line concentration pattern.** One or two cell lines
carry nearly all the converted reads while the rest sit at zero. At
chr14:106,470,411 a **single** cell line supplies 100% of them, across only 15
covered cell lines.

**Two of the three were not problematic before.** They are newly flagged, and the
likely reason is that the covariate PCs changed when they were recomputed on the
filtered sites, altering the null fit. The problematic-site set is therefore not
a fixed property of the data — it moves with the model.

```{r site3-fig, fig.height=3.6}
lb <- labs[order(-real_nlp)]
lab_of <- setNames(sprintf("chr%s:%s\n-log10p=%.0f | top line=%.0f%% | %d cell lines",
                           lb$chr, format(lb$pos, big.mark = ",", trim = TRUE),
                           lb$real_nlp, 100 * lb$top_cellline_share, lb$n_cl), lb$sid)
d <- copy(pan); d[, lab := factor(lab_of[sid], levels = lab_of)]
ggplot(d, aes(pu, pmax(n, 1))) +
  geom_vline(xintercept = c(0, 0.5, 1), colour = "grey88", linewidth = 0.4) +
  geom_point(colour = "#256abf", alpha = 0.75, size = 2) +
  scale_x_continuous(limits = c(-0.04, 1.04), breaks = c(0, 0.5, 1),
                     labels = c("0", "50", "100")) +
  scale_y_log10(labels = scales::comma) +
  facet_wrap(~ lab, nrow = 1) +
  labs(x = "%pU in one cell line", y = "read depth (log scale)",
       title = "The remaining problematic sites: a wall at zero plus one or two outliers",
       subtitle = "Each dot is one cell line. No discrete 0/50/100% banding -- this is concentration, not aliasing.") +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(size = 8), panel.grid.minor = element_blank())
```

# 4. What is still open

- **A concentration filter.** All three survivors share one signature: a single
  cell line supplying most or all of the converted reads. A cap on that share
  would remove them, and it is computable for any site without genotype or
  untreated data. It has not been applied here.
- **Is the beta-binomial itself at fault?** With aliasing excluded as the
  explanation, these three sites are the natural test set for a goodness-of-fit
  test of the beta-binomial — the question of whether the model can represent a
  phenotype shaped like this at all. That test is written but has not been run.
- **The discovery scan has not been rerun.** This page covers calibration only.
  The scan on the filtered phenotype was submitted and then deliberately
  cancelled to focus on calibration first, so there is no updated discovery count
  yet. Note that excluding `r format(as.integer(gv("sites_before")) - as.integer(gv("sites_after")), big.mark = ",")`
  sites also changes the multiple-testing correction, so the new count will not
  simply be the old one minus the excluded hits.

# Reproducing this page

```{r prov}
knitr::kable(info, caption = "Parameters")
knitr::kable(data.table(
  step = c("untreated (pooled input) count matrices",
           "U/C-filtered phenotype + recomputed PCs",
           "site-resolved calibration on the filtered phenotype",
           "extraction of the remaining problematic sites",
           "precompute for this page", "this page"),
  script = c("12.untreated_counts_matrix.R", "13.filtered_phenotype.R",
             "8.calibration_site_resolved.R --pheno fold5_union_noUC",
             "9.qq_exclude_pathological.R --stem calib_siteres_w5kb_pc3_noUC",
             "15.calibration_noUC_page_data.R",
             "analysis/qtl_mapping_betabinomial_calibration_noUC.Rmd")),
  caption = "Provenance. Both scan scripts take --pheno so the original fold5_union results are never overwritten.")
```

Related: [the 26 problematic sites](qtl_mapping_betabinomial_site26.html) ·
[full pathology write-up](qtl_mapping_betabinomial_pathology_summary.html)
