---
title: "Beta-binomial pU-QTL: how many covariate PCs?"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 3
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 10)

PSI  <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl"
D    <- file.path(PSI, "qtl_betabinomial/phi_vs_npcs")
G    <- file.path(PSI, "qtl_betabinomial/goodness_of_fit")
FIG  <- file.path(PSI, "qtl_betabinomial/figures_npcs")
DIAG <- file.path(PSI, "qtl_betabinomial/diagnostic_figures")
SUF  <- "_auto_noUCcalled"            # the autosome-only sweep (chr1-22)

# sid contains a space ("3 37277733"), so the separator must be given explicitly
fit  <- fread(file.path(D, paste0("phi_by_npcs", SUF, ".txt.gz")), sep = "\t")
ov   <- fread(file.path(D, paste0("phi_set_overlap", SUF, ".txt")), sep = "\t")
stab <- fread(file.path(D, paste0("phi_stability_kept_sites", SUF, ".txt")), sep = "\t")
info <- fread(file.path(D, paste0("runinfo", SUF, ".txt")), sep = "\t")
gv   <- function(k) info[parameter == k]$value

KS   <- c(3, 5, 7, 10, 15, 20)         # the k values with a calibration run

# per-k iterated phi exclusion list. readLines, not fread: sid contains a space.
# k = 3 keeps the untagged name for back-compat inside 18.phi_filter_iteration.R.
list_for <- function(k) file.path(G, sprintf("phi_filter_dropped_sites%s%s.txt", SUF,
                                            if (k == 3) "" else sprintf("_pc%d", k)))
nexcl <- vapply(KS, function(k) {
  f <- list_for(k)
  if (!file.exists(f)) return(NA_integer_)
  length(setdiff(trimws(readLines(f)), c("", "sid")))
}, integer(1))
names(nexcl) <- KS
iter3 <- fread(file.path(G, paste0("phi_filter_iteration_history", SUF, ".txt")), sep = "\t")

# calibration: each k on its OWN phi-filtered phenotype
lam <- fread(file.path(FIG, "lambda_by_npcs_auto_filtered.tsv"), sep = "\t")
qqd <- fread(file.path(FIG, "qq_thinned_filtered.tsv.gz"), sep = "\t")
cal <- merge(lam[dataset != "real", .(perm_lo = min(lambda), perm_hi = max(lambda),
                                      perm_mid = median(lambda),
                                      worst_perm_tail = max(max_nlp)), by = npcs],
             lam[dataset == "real", .(npcs, real_lambda = lambda,
                                      real_tail = max_nlp, sites = n_sites,
                                      n_pvalues = n_pvalues)],
             by = "npcs")[order(npcs)]
cal[, phi_excluded := nexcl[as.character(npcs)]]

ORD  <- c("real", paste0("perm", 1:5))
COLS <- c(real = "#e34948", perm1 = "#9ec5f4", perm2 = "#6da7ec",
          perm3 = "#3987e5", perm4 = "#256abf", perm5 = "#0d366b")
NTOT    <- fit[npcs == 3, .N]        # sites with >=10 covered cell lines
PHI_CUT <- 0.99
PHI_EPS <- 1e-6                      # the numerical floor inside fit_bb_mle
```

Everything on this page is restricted to **chr1–22**, with the covariate PCs
recomputed on chr1–22 only. Phenotype **`r gv("phenotype")`**
(`r format(as.integer(gv("n_sites")), big.mark = ",")` sites ×
`r gv("n_cell_lines")` cell lines) — U/C-variant sites already excluded.

# 1. The problem: fits that land on the φ boundary

φ is the beta-binomial's overdispersion parameter. Here it is the intracluster
correlation,

$$\varphi = \frac{1}{a + b + 1}, \qquad
  \mathrm{Var}[y] = n\,\mu(1-\mu)\bigl[1 + (n-1)\varphi\bigr]$$

so it is bounded to (0, 1) by construction. **φ → 1 means `a + b → 0`, where the
beta prior collapses onto point masses at 0 and 1** — the model asserting that
every cell line's true rate is exactly 0 or exactly 1. A fit that lands there has
not really been fitted, and any LRT computed from it is computed from a boundary
solution.

This is what the pU-QTL calibration work traced its remaining miscalibration to:
every site still returning an extreme p-value *under permutation* had φ pinned
above 0.99, while only a fraction of a percent of all sites do — see
[the filtering strategy](qtl_mapping_betabinomial_strategy.html). The scan
therefore excludes those sites.

But the null model is

$$\mathrm{logit}(\mu_i) = \beta_0 + \beta_1 \mathrm{PC1} + \dots + \beta_k \mathrm{PC}k$$

so **k enters the fit directly.** More PCs give $\mu_i$ more freedom to track each
cell line, so scatter that φ currently absorbs could instead be explained by the
mean structure. That raises the question this page answers: how does the φ problem
behave across k, and which k should the scan use?

**Window size is not an input to the null fit.** φ comes from the null, which
contains no genotype; the window only decides which cis-SNPs are tested
afterwards. This is true *by construction* rather than by verification — the
window is not a parameter of the fit at all, so there is nothing to vary. (The
script that produces every φ on this page never reads a genotype file.)

That invariance covers φ **only**. Everything in section 3 is calibration, and
calibration *is* window-dependent: a wider window admits more cis-SNPs per site,
which changes the per-site minimum-p distribution, the tail heights and the
problematic-site counts. **All calibration on this page is at ±5 kb**, and no
other window has been run, so nothing here speaks to how λ behaves at 20, 50 or
100 kb.

# 2. φ estimated across many PC settings

Null fits only — no genotype, so this is cheap (all seven values of k in under a
minute).
`r format(as.integer(gv("n_sites")) - NTOT, big.mark = ",")` sites have fewer
than `r gv("min_cell_lines")` covered cell lines and are unfittable at any k;
they are excluded throughout, leaving **`r format(NTOT, big.mark = ",")`** sites,
constant across k.

```{r tab2}
knitr::kable(ov[order(npcs), .(
  `k (PCs)`      = npcs,
  `total sites`  = format(NTOT, big.mark = ","),
  `fittable`     = format(n_fitted, big.mark = ","),
  `unfittable`   = format(n_failed, big.mark = ","),
  `phi > 0.99`   = n_flagged,
  `UNUSABLE (unfittable + phi>0.99)` = format(n_unusable, big.mark = ","),
  `unusable %`   = sprintf("%.1f%%", 100 * n_unusable / NTOT))],
  caption = "'unfittable' means fit_bb_mle() returned NULL -- the fit was attempted and rejected -- NOT missing data; cell lines with zero depth are dropped before fitting. A site is equally lost to the scan whether it is unfittable or phi-degenerate, so the UNUSABLE column is the one to read.")
```

```{r tradeoff, fig.height=5.4}
m <- rbindlist(list(
  ov[, .(npcs, quantity = "sites with phi > 0.99\n(boundary-degenerate fit)", n = n_flagged)],
  ov[, .(npcs, quantity = "sites that cannot be fitted at all\n(fit_bb_mle returns NULL)", n = n_failed)]))
m[, quantity := factor(quantity, levels = unique(quantity))]

ggplot(m, aes(npcs, n)) +
  geom_line(colour = "grey55", linewidth = 0.6) +
  geom_point(colour = "#0d366b", size = 3) +
  geom_text(aes(label = sprintf("%s\n(%.1f%%)", format(n, big.mark = ","), 100 * n / NTOT)),
            vjust = -0.55, size = 3.0, lineheight = 0.9) +
  scale_x_continuous(breaks = sort(unique(m$npcs))) +
  scale_y_continuous(expand = expansion(mult = c(0.08, 0.34))) +
  facet_wrap(~ quantity, nrow = 1, scales = "free_y") +
  labs(x = "number of covariate PCs in the null model (k)", y = "sites",
       title = "More PCs remove the phi-boundary fits by destroying the fit entirely",
       subtitle = sprintf("Percentages are of the %s sites with >= %s covered cell lines.\nNOTE THE FREE y-AXES: the left panel tops out at %d, the right at %s.",
                          format(NTOT, big.mark = ","), gv("min_cell_lines"),
                          max(ov$n_flagged), format(max(ov$n_failed), big.mark = ","))) +
  theme_bw(base_size = 12) + theme(strip.text = element_text(size = 10.5))
```

The φ > 0.99 count falls to zero as k rises — but not because those sites became
well-behaved. They stop producing a fit at all. Unfittable sites go from
`r format(ov[npcs == 3]$n_failed, big.mark = ",")` at k = 3 to
`r format(ov[npcs == 20]$n_failed, big.mark = ",")` at k = 20, so the total
unusable count is a **U-shape with its minimum at
k = `r ov[which.min(n_unusable)]$npcs`** — the setting already in use. The
mechanism is sample size: there are `r gv("n_cell_lines")` cell lines, so k = 20
means estimating 22 parameters from `r gv("n_cell_lines")` observations.

**A zero in the φ > 0.99 column is therefore not good news, and it is not the
whole story either** — §4 shows the degeneracy reappearing at the opposite end of
φ's range.

# 3. Calibration after excluding the problematic sites

Each k here gets **its own** exclusion list: φ > 0.99 iterated to a fixed point at
that k, with the PCs recomputed each round, then the phenotype rebuilt and the PCs
recomputed again on what remains. So the comparison is not confounded by a list
derived at some other k.

```{r nexcl}
knitr::kable(data.table(`k (PCs)` = KS, `sites excluded by the phi filter at this k` = as.integer(nexcl)),
  caption = "The filter is a genuine intervention only at low k. At k >= 10 it removes nothing -- those sites have already been lost to fit failure rather than cured, so the k >= 10 panels below are NOT 'cleaned up' versions.")
```

**Read the permutation λ.** Shuffling the phenotype destroys any
genotype–phenotype relationship, so permutation λ must be 1 *whatever k is*.
Real-data λ cannot settle the question, because it cannot separate "more signal"
from "more inflation".

```{r cal-tab}
# join the fit-level counts (from the null-only sweep) onto the calibration
# results so one table carries the whole picture per k.
ct <- merge(cal, ov[, .(npcs, n_fitted, n_failed)], by = "npcs", all.x = TRUE)
# problematic sites AFTER that k's own phi filter, i.e. from the same filtered
# calibration run the lambda columns come from -- not the unfiltered tree.
ct[, problematic := vapply(npcs, function(k) {
  fp <- file.path(DIAG, sprintf("pathological_sites_w5kb_pc%d_auto_noUCcalled_noPhi_pc%d.txt", k, k))
  if (file.exists(fp)) max(0L, length(readLines(fp)) - 1L) else NA_integer_
}, integer(1))]

knitr::kable(ct[order(npcs), .(
  `k (PCs)`             = npcs,
  `total sites`         = format(NTOT, big.mark = ","),
  `fittable`            = format(n_fitted, big.mark = ","),
  `unfittable`          = format(n_failed, big.mark = ","),
  `phi excluded`        = phi_excluded,
  `sites tested`        = format(sites, big.mark = ","),
  `p-values`            = format(n_pvalues, big.mark = ","),
  `real lambda`         = round(real_lambda, 3),
  `permutation lambda`  = sprintf("%.3f - %.3f", perm_lo, perm_hi),
  `problematic (post-filter)` = problematic,
  `real tail`           = round(real_tail, 1),
  `worst permuted tail` = round(worst_perm_tail, 1))],
  caption = "One row per k. 'total sites' and 'fittable' are from the null-only fits (section 2); 'sites tested' is lower again because a site also needs a cis-SNP at MAF >= 0.05 within +/-5 kb. 'problematic (post-filter)' = sites still extreme under permutation AFTER that k's phi filter. NOTE this is not comparable across k as a measure of filter quality: at k = 3 the filter removes 33 sites and clears all 9 that were problematic beforehand, leaving 2 NEW ones created by recomputing the PCs; at k >= 10 it removes nothing, so the column is simply the unfiltered count. Real scan plus 5 phenotype permutations at each k, chr1-22.")
```

```{r cal-fig, fig.height=7.6}
g1 <- ggplot(cal, aes(npcs)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "#e34948") +
  annotate("text", x = max(cal$npcs), y = 1, label = "perfect calibration ",
           hjust = 1, vjust = -0.6, colour = "#e34948", size = 3.3) +
  geom_ribbon(aes(ymin = perm_lo, ymax = perm_hi), fill = "#3987e5", alpha = 0.25) +
  geom_line(aes(y = perm_mid), colour = "#0d366b", linewidth = 0.7) +
  geom_point(aes(y = perm_mid), colour = "#0d366b", size = 2.6) +
  geom_text(aes(y = perm_hi, label = sprintf("%.2f", perm_mid)), vjust = -0.9,
            size = 3.2, colour = "#0d366b") +
  scale_x_continuous(breaks = cal$npcs) +
  scale_y_continuous(expand = expansion(mult = c(0.06, 0.18))) +
  labs(x = NULL, y = expression(permutation~lambda),
       title = "Calibration degrades monotonically as PCs are added",
       subtitle = "Band = range over the 5 permutations, line = median. This must be 1.") +
  theme_bw(base_size = 12)

g2 <- ggplot(cal, aes(npcs, sites)) +
  geom_line(colour = "#7a4b8f", linewidth = 0.7) +
  geom_point(colour = "#7a4b8f", size = 2.6) +
  geom_text(aes(label = format(sites, big.mark = ",")), vjust = -1.0, size = 3.2,
            colour = "#7a4b8f") +
  scale_x_continuous(breaks = cal$npcs) +
  scale_y_continuous(expand = expansion(mult = c(0.08, 0.18))) +
  labs(x = "number of covariate PCs (k)", y = "sites tested",
       subtitle = "...and it costs sites: more PCs means more fits rejected outright.") +
  theme_bw(base_size = 12)

gridExtra::grid.arrange(g1, g2, ncol = 1, heights = c(1.3, 1))
```

Permutation λ rises from
**`r sprintf("%.3f-%.3f", cal[npcs==3]$perm_lo, cal[npcs==3]$perm_hi)`** at k = 3
to **`r sprintf("%.3f-%.3f", cal[npcs==20]$perm_lo, cal[npcs==20]$perm_hi)`** at
k = 20, monotonically and with no local optimum — **k = 5 is already worse than
k = 3**, so there is no headroom to buy. No k on this grid beats the setting in
use.

## The QQ plots

```{r qq, fig.height=8.6, fig.width=13}
q <- copy(qqd)[npcs %in% KS]
q[, dataset := factor(dataset, levels = ORD)]
sub <- cal[, .(npcs, lab = sprintf("k = %d PCs\nreal %.3f | perm %.3f-%.3f",
                                   npcs, real_lambda, perm_lo, perm_hi))]
q <- merge(q, sub, by = "npcs")
q[, panel := factor(lab, levels = sub[order(npcs)]$lab)]

ggplot(q, aes(expected, observed, colour = dataset)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45") +
  geom_point(size = 0.5, alpha = 0.8) +
  scale_colour_manual(values = COLS, name = NULL) +
  guides(colour = guide_legend(override.aes = list(size = 2.8, alpha = 1), nrow = 1)) +
  facet_wrap(~ panel, nrow = 2) +
  labs(x = expression(-log[10]("expected p")), y = expression(-log[10]("observed p")),
       title = "Calibration QQ after per-k phi filtering (autosomes only)",
       subtitle = paste("Red = real scan, blues = 5 phenotype permutations.",
                        "READ THE BLUES: with the phenotype shuffled there is no signal,",
                        "\nso they must lie on the dashed line. Axes shared across panels.")) +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

```{r qq-zoom, fig.height=8.6, fig.width=13}
YMAX <- 20
ggplot(q, aes(expected, observed, colour = dataset)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45") +
  geom_point(size = 0.5, alpha = 0.8) +
  scale_colour_manual(values = COLS, name = NULL) +
  guides(colour = guide_legend(override.aes = list(size = 2.8, alpha = 1), nrow = 1)) +
  facet_wrap(~ panel, nrow = 2) +
  coord_cartesian(ylim = c(0, YMAX)) +
  labs(x = expression(-log[10]("expected p")), y = expression(-log[10]("observed p")),
       title = sprintf("Same data, y clipped at %d to show the bulk", YMAX),
       subtitle = paste("At k = 3 the permutations stay near the diagonal through the bulk.",
                        "As k rises the WHOLE permutation curve lifts off it --",
                        "\ninflation with no signal to explain it, on every surviving site at once.",
                        "Clipped points are outside the view, not deleted.")) +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

**Why high k can look better and is not.** The real-data QQ tail height is
*erratic* across k — `r paste(sprintf("%.0f", cal$real_tail), collapse = " / ")`
for k = `r paste(cal$npcs, collapse = " / ")` — because it depends on whether a
given k happens to leave its worst sites fittable. At k = 15 every dataset,
including all five permutations, hits the −log₁₀p = 300 numerical clamp. So tail
height is **not** a usable criterion for choosing k. Permutation λ is smooth and
monotone over the same range, and it is the one to use.

A handful of catastrophic sites can be identified and excluded. Diffuse inflation
spread over every surviving site cannot be, and it corrupts every FDR threshold.

One bookkeeping note so the counts reconcile: the §2 sweep does a **single pass**
at each k, while the exclusion lists in this section are **iterated**. At k = 3
that is `r iter3[round == 0]$n_new_flagged` sites in round 0 and
`r max(iter3$n_dropped_total)` after `r sum(iter3$n_new_flagged > 0)` rounds that
flagged anything, because removing sites changes the PCs, which changes every
remaining fit.

# 4. Where the dispersion goes

Both ends of φ's range are degenerate: φ → 1 collapses the beta prior onto point
masses at 0 and 1, and φ → 0 reduces the model to a plain binomial. `fit_bb_mle`
floors φ at `PHI_EPS` = 1e-6 to keep the likelihood finite.

Watching the whole distribution, rather than just the upper tail, shows that
adding PCs does not remove the degeneracy — it **migrates to the other boundary**:

```{r phidist, fig.height=6.6}
d <- fit[ok == TRUE]
d[, kf := factor(npcs, levels = sort(unique(npcs)),
                 labels = paste0("k = ", sort(unique(npcs)), " PCs"))]
lab <- d[, .(txt = sprintf("at the 1e-6 floor: %.1f%%\nphi > 0.99: %.2f%%",
                           100 * mean(phi <= 1.01 * PHI_EPS), 100 * mean(phi > PHI_CUT))),
         by = kf]

ggplot(d, aes(pmax(phi, PHI_EPS))) +
  geom_histogram(bins = 60, fill = "#3987e5", colour = NA) +
  geom_vline(xintercept = PHI_CUT, linetype = "dashed", colour = "#e34948") +
  geom_vline(xintercept = PHI_EPS, linetype = "dashed", colour = "grey30") +
  geom_text(data = lab, aes(x = 3e-6, y = Inf, label = txt), hjust = 0, vjust = 1.25,
            size = 2.8, lineheight = 0.95, inherit.aes = FALSE) +
  scale_x_log10() + facet_wrap(~ kf, ncol = 3) +
  labs(x = "fitted phi (log scale); grey dashed = 1e-6 floor, red dashed = 0.99",
       y = "sites",
       title = "Adding PCs does not cure the degeneracy -- it migrates to the opposite boundary",
       subtitle = paste("At high k the typical site is pinned at the phi = 0 floor: the model has collapsed to a binomial,",
                        "\nbecause the PCs have absorbed the dispersion into the mean structure.",
                        "The y-axis is SHARED across panels.")) +
  theme_bw(base_size = 12)
```

```{r bothends}
knitr::kable(d[, .(
  `sites fitted`              = format(.N, big.mark = ","),
  `median phi`                = signif(median(phi), 3),
  `at the phi=0 floor (1e-6)` = sprintf("%.1f%%", 100 * mean(phi <= 1.01 * PHI_EPS)),
  `phi > 0.99`                = sprintf("%.2f%%", 100 * mean(phi > PHI_CUT))),
  by = .(`k (PCs)` = npcs)][order(`k (PCs)`)],
  caption = "Both boundaries matter, which is why 'phi > 0.99 = 0' at high k is not evidence of a healthier fit.")
```

At k = 20,
**`r sprintf("%.1f%%", 100 * mean(d[npcs == 20]$phi <= 1.01 * PHI_EPS))` of the
sites that do fit sit on the φ = 0 floor.** Combined with the
`r sprintf("%.1f%%", 100 * ov[npcs == 20]$n_failed / NTOT)` that fail outright,
almost nothing at k = 20 is being estimated properly.

This also changes what the eventual fix should look like: **the guard belongs
inside `fit_bb_mle`, and it must be two-sided**, covering φ → 0 as well as
φ → 1. `betabin_lib.R` already rejects boundary values for μ and for the
coefficients (`MAX_ABS_COEF`) — φ was simply omitted from that set, and an
exclusion list derived at one k is a stopgap for something that should be caught
at fit time at every k.

# 5. φ is not stable across k, even for the sites we keep

The exclusion list could be well-behaved while every *retained* site's φ moves,
which would matter for the LRT regardless of the filter. It does move:

```{r stab}
knitr::kable(stab[, .(pair, correlation = round(cor, 4),
                      `max |difference|` = round(max_abs_diff, 4),
                      `frac moving > 0.01` = sprintf("%.1f%%", 100 * frac_moving_gt_0.01))],
  caption = "phi among RETAINED sites, between pairs of k. Compare with the iterated filter at fixed k, where recomputing the SAME number of PCs on a changed site set leaves phi almost fixed (correlation 0.999+).")
```

Changing the *number* of PCs is a much larger perturbation than recomputing the
same number on a slightly different site set. **A scan at a different k is a
genuinely different model, not a small perturbation of this one** — which is why
the exclusion list has to be re-derived per k, as it was in §3, and why it should
not be carried across configs.

# Conclusion

**Keep k = 3.** Three independent lines of evidence agree:

- it minimises unusable sites (§2), with the total a U-shape whose minimum is k = 3;
- it is the best-calibrated point on the grid (§3), and calibration degrades
  monotonically upward with no local optimum, so k = 5 is already worse;
- at higher k the model is not merely worse-calibrated but structurally
  degenerate, with most fits collapsing to the binomial limit (§4).

The φ > 0.99 exclusion list is valid **at k = 3** and should be re-derived if k
ever changes. The durable fix is a two-sided φ guard inside `fit_bb_mle`.

# Caveats

- **Which rejection exit fires is not instrumented.** "Unfittable" means
  `fit_bb_mle` returned `NULL`; its four exits (no starting values,
  non-convergence, μ boundary guard, coefficient guard) are not distinguished.
  Overfitting makes the μ and coefficient guards likely at high k, but
  non-convergence would look identical here. Stated as inference, not measurement.
- **Only the k values shown were tested.** k = 3 is the best of
  {`r paste(sort(unique(fit$npcs)), collapse = ", ")`}, not a fitted optimum;
  nothing here explores k = 1, 2 or 4.
- **The φ filter is a no-op at k ≥ 10**, so those panels in §3 are not "cleaned"
  versions of the same analysis. The comparison across k is honest about
  calibration but is not a like-for-like test of the filter.
- **φ here is not a "common" φ.** φ = 1/(a+b+1) ∈ (0,1), matching `aod`'s φ and
  VGAM's `rho`, not a precision parameter on (0, ∞).

Related: [the current strategy and its calibration](qtl_mapping_betabinomial_strategy.html) ·
[the goodness-of-fit attempts](qtl_mapping_betabinomial_gof_failed.html)
