---
title: "Beta-binomial pU-QTL: the current filtering strategy and its calibration"
author: "XSun"
date: "`r Sys.Date()`"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_depth: 3
---

```{r setup, warning=FALSE, message=FALSE}
suppressMessages({ library(data.table); library(ggplot2) })
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 11)

PSI <- "/project/xinhe/xsun/pumseq/pumseq_processed/psi_qtl"
DAT <- file.path(PSI, "qtl_betabinomial/strategy_page_data")

qq   <- fread(file.path(DAT, "qq_all_runs.tsv.gz"))
lam  <- fread(file.path(DAT, "lambda_all_runs.tsv"))
phid <- fread(file.path(DAT, "phi_distribution.tsv"))
phis <- fread(file.path(DAT, "phi_per_site.tsv.gz"))
iter <- fread(file.path(DAT, "phi_iteration.tsv"))
excl <- fread(file.path(DAT, "exclusion_steps.tsv"))
info <- fread(file.path(DAT, "runinfo.tsv"))
gv <- function(k) info[parameter == k]$value

RUN_LV <- c("1. all sites", "2. minus U/C-variant sites",
            "3. minus U/C + phi-boundary sites")
ORD <- c("real", paste0("perm", 1:5))
qq[,  run := factor(run, levels = RUN_LV)][, dataset := factor(dataset, levels = ORD)]
lam[, run := factor(run, levels = RUN_LV)][, dataset := factor(dataset, levels = ORD)]
COLS <- c(real = "#e34948", perm1 = "#9ec5f4", perm2 = "#6da7ec",
          perm3 = "#3987e5", perm4 = "#256abf", perm5 = "#0d366b")
lv <- function(r, d, col) lam[run == r & dataset == d][[col]]
FINAL <- RUN_LV[3]
```

# The strategy in one table

Two filters are applied to the phenotype before mapping. Neither is a curated
list of bad sites: both are rules computable from a single site's own data, so
they apply unchanged to any future dataset.

```{r strategy}
knitr::kable(data.table(
  step = c("start", "1. U/C-variant sites", "2. phi-boundary fits", "final"),
  what = c("all called pU sites",
           "a U/C variant is CALLED on the pU base, so %pU partly reports genotype",
           "the fitted overdispersion phi lands on its boundary (phi > 0.99), i.e. the fit is degenerate",
           "used for all QTL mapping"),
  removed = c("--", format(as.integer(gv("excluded_UC")), big.mark = ","),
              gv("excluded_phi"), "--"),
  sites_left = c(format(as.integer(gv("sites_start")), big.mark = ","),
                 format(as.integer(gv("sites_start")) - as.integer(gv("excluded_UC")), big.mark = ","),
                 format(as.integer(gv("sites_final")), big.mark = ","),
                 format(as.integer(gv("sites_final")), big.mark = ","))),
  caption = sprintf("Phenotype: %s. Covariate PCs are recomputed on the retained sites at each step.", gv("final_phenotype")))
```

And the effect on calibration, which is the point of the whole exercise:

```{r headline}
knitr::kable(data.table(
  run = RUN_LV,
  `real lambda` = vapply(RUN_LV, lv, numeric(1), d = "real", col = "lambda"),
  `permutation lambda (range)` = vapply(RUN_LV, function(r)
    sprintf("%.3f - %.3f", min(lam[run == r & dataset != "real"]$lambda),
            max(lam[run == r & dataset != "real"]$lambda)), character(1)),
  `real tail` = round(vapply(RUN_LV, lv, numeric(1), d = "real", col = "max_nlp"), 1),
  `worst permuted tail` = round(vapply(RUN_LV, function(r)
    max(lam[run == r & dataset != "real"]$max_nlp), numeric(1)), 1),
  `problematic sites` = vapply(RUN_LV, function(r)
    lam[run == r][1]$n_problematic, integer(1))),
  row.names = FALSE, digits = 3,
  caption = "Under permutation there is no genotype-phenotype relationship, so those rows are the diagnostic: lambda should be 1 and the tail near 0.")
```

**`r lam[run == FINAL][1]$n_problematic` sites remain problematic** after both
filters, down from `r lam[run == RUN_LV[1]][1]$n_problematic`. Permutation λ sits
at `r sprintf("%.2f-%.2f", min(lam[run == FINAL & dataset != "real"]$lambda), max(lam[run == FINAL & dataset != "real"]$lambda))`,
which is what correct calibration looks like, and the real-data λ of
**`r sprintf("%.3f", lv(FINAL, "real", "lambda"))`** is the mild inflation
expected from genuine cis signal — not the
`r sprintf("%.3f", lv(RUN_LV[1], "real", "lambda"))` seen before any filtering,
which was artefact.

# 1. Calibration QQ plots

Each panel is one filtering stage. **Red is the real scan; the five blues are
independent phenotype permutations.** The permutation series carries the
information: shuffling the phenotype destroys any genotype–phenotype
relationship, so those points must lie on the dashed line. Where they rise above
it, the test is manufacturing signal out of nothing.

```{r qq-facets, fig.height=4.8}
ggplot(qq, aes(expected, observed, colour = dataset)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45") +
  geom_point(size = 0.55, alpha = 0.75) +
  scale_colour_manual(values = COLS, name = NULL) +
  guides(colour = guide_legend(override.aes = list(size = 2.6, alpha = 1), nrow = 1)) +
  facet_wrap(~ run, nrow = 1, scales = "free_y") +
  labs(x = expression(-log[10]("expected p")), y = expression(-log[10]("observed p")),
       title = "Beta-binomial LRT calibration at each filtering stage",
       subtitle = "Note the free y-axis -- the left panel reaches ~290, the right ~28. The blues are what matter.") +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

## The final calibration on its own

The same data as the right-hand panel above, at full size, since this is the
plot that describes the analysis actually being used.

```{r qq-final, fig.height=6}
ggplot(qq[run == FINAL], aes(expected, observed, colour = dataset)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(size = 1.1, alpha = 0.8) +
  scale_colour_manual(values = COLS, name = NULL,
                      labels = c("real data", paste("permutation", 1:5))) +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1), nrow = 1)) +
  labs(x = expression(-log[10]("expected p")), y = expression(-log[10]("observed p")),
       title = sprintf("Calibration after both filters (%s sites tested)",
                       format(lam[run == FINAL & dataset == "real"]$n_sites, big.mark = ",")),
       subtitle = paste0("real lambda = ", sprintf("%.3f", lv(FINAL, "real", "lambda")),
                         "; permutations ",
                         sprintf("%.3f-%.3f", min(lam[run == FINAL & dataset != "real"]$lambda),
                                 max(lam[run == FINAL & dataset != "real"]$lambda)),
                         ". The permutation series sit close to, though slightly above, the diagonal",
                         " -- consistent with lambda just over 1 -- and one reaches ~12 in the far tail.",
                         " The real series separates from all five.")) +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

The gap between the red series and the blue ones **is** the signal, and this is
the first time in this project the plot has looked like a working cis-QTL scan.

Two honest qualifications, both visible in the figure. The permutation series run
slightly *above* the diagonal rather than on it, which is what λ of 1.02–1.06
means — a small residual inflation that the filters did not remove. And one
permutation point still reaches ≈12 in the far tail: a single site can produce an
extreme p-value from shuffled data, it simply does not also do so in the real
data, which is why it is not counted as problematic.

```{r lam-tab}
w <- dcast(lam, dataset ~ run, value.var = "lambda")
setcolorder(w, c("dataset", RUN_LV))
knitr::kable(w[order(factor(dataset, levels = ORD))], digits = 3,
             caption = "lambda by dataset and stage")
t <- dcast(lam, dataset ~ run, value.var = "max_nlp")
setcolorder(t, c("dataset", RUN_LV))
knitr::kable(t[order(factor(dataset, levels = ORD))], digits = 1,
             caption = "Largest observed -log10(p): the QQ tail height")
```

# 2. Filter 1 — U/C-variant sites

Pseudouridine is a modified uridine and the assay reads the base at that
position: a modified U is called **C**, an unmodified U is called **T**. So if a
common sequence variant changes that same base from U to C, an inherited C is
indistinguishable from a modification, and the measured "%pU" becomes partly a
read-out of genotype.

This was established by measurement rather than inference. Each cell line has
untreated `input_pool` libraries where no chemical conversion occurs, so any C
seen there is inherited. At sites with a called U/C variant the untreated C
fraction is ~1.0; elsewhere the median is 0.0004.

**`r format(as.integer(gv("excluded_UC")), big.mark = ",")` sites are excluded**
on the basis of a called U/C variant on the pU base.

> A broader version of this filter also drops sites where the untreated controls
> *measure* inherited C although no variant was called (~1,170 more). It gives a
> better-looking result but removes more than intended, so it is not used here.
> See the [companion page](qtl_mapping_betabinomial_calibration_noUC.html).

# 3. Filter 2 — degenerate fits (phi on the boundary)

This step is not about the phenotype at all. φ is the beta-binomial's
overdispersion parameter, bounded to (0, 1). As φ → 1 the beta prior collapses
(`k = (1-φ)/φ → 0`) and the model degenerates, so a fit that lands there has not
really been fitted.

```{r phi-dist, fig.height=4.2}
p <- copy(phis)[is.finite(phi)]
ggplot(p, aes(pmax(phi, 1e-6), fill = problematic)) +
  geom_histogram(bins = 60, position = "identity", alpha = 0.75, colour = NA) +
  geom_vline(xintercept = 0.99, linetype = "dashed", colour = "grey25") +
  annotate("text", x = 0.99, y = Inf, hjust = 1.05, vjust = 1.6, size = 3.2,
           label = "phi = 0.99 ") +
  scale_x_log10() + scale_y_sqrt() +
  scale_fill_manual(values = c(`FALSE` = "#3987e5", `TRUE` = "#e34948"),
                    labels = c("ordinary site", "was problematic under permutation"),
                    name = NULL) +
  labs(x = "fitted overdispersion phi (log scale)", y = "sites (sqrt scale)",
       title = "phi spans its whole range; the problematic sites sit at the very top of it",
       subtitle = paste("phi is broadly and multi-modally distributed from 1e-6 to 1 -- there is NO clean gap.",
                        "\nWhat is clean is the discrimination: every site still problematic after filter 1",
                        "has phi > 0.99, and only 0.4% of all sites do.")) +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

The distribution is not neatly bimodal — φ occupies its entire range with
several modes, so 0.99 is a cut in the extreme tail rather than a gap between two
clean populations. What *is* clean is the discrimination:

```{r phi-tab}
knitr::kable(phis[is.finite(phi), .(
  sites = .N,
  `median phi` = signif(median(phi), 3),
  `frac phi > 0.99` = sprintf("%.3f", mean(phi > 0.99)),
  `median refit failure rate` = round(median(refit_fail, na.rm = TRUE), 3)),
  by = .(was_problematic = problematic)],
  caption = paste("The MEDIAN ordinary site sits four orders of magnitude from the",
                  "boundary, but the distribution is broad, so this contrast is",
                  "between typical values rather than between separated clusters.",
                  "'refit failure rate' is the fraction of datasets simulated from a",
                  "site's own fit that then cannot be refitted -- a direct measure of",
                  "how unstable the fit is."))
```

Two things make this a fitter correctness issue rather than a tuned threshold:

**`betabin_lib.R` already rejects boundary fits for the other parameters** — it
guards against μ hitting 0 or 1 and against runaway coefficients
(`MAX_ABS_COEF`). φ was simply omitted from that set. The right long-term home
for this rule is inside `fit_bb_mle`, not a site exclusion list.

**Data simulated from these fits often cannot be refitted at all.** Simulate 30
datasets from a site's own fitted model and try to recover it: ordinary sites
succeed every time, boundary sites fail 20–45% of the time. If a model cannot be
recovered from data it generated itself, nothing computed from it is trustworthy.

## Does the filter converge?

Removing sites changes the site set, so the covariate PCs must be recomputed —
which changes every site's fit, which can push new sites onto the boundary. The
filter therefore has to be iterated to a fixed point rather than applied once.

```{r phi-iter}
knitr::kable(iter, digits = 5,
  caption = paste("Each round drops the flagged sites, recomputes the PCs, refits",
                  "everything, and looks for new boundary fits."))
```

It converges after `r max(iter$round)` rounds at
**`r max(iter$n_dropped_total)` sites**. Applying the filter once would have
caught `r iter$n_new_flagged[1]` and missed
`r max(iter$n_dropped_total) - iter$n_new_flagged[1]`.

φ itself is stable for the sites that are kept (correlation 0.9986 between the
first and last round, with only 0.11% of sites moving by more than 0.01), which
is why convergence is quick — the PCs enter the model through μ, while φ is
driven mostly by a site's own scatter.

# 4. What each exclusion reason contributed

```{r excl}
knitr::kable(excl, caption = "Breakdown of the final exclusion manifest")
```

The two filters do not overlap: `r format(as.integer(gv("excluded_UC")), big.mark = ",")`
+ `r gv("excluded_phi")` = `r format(as.integer(gv("excluded_UC")) + as.integer(gv("excluded_phi")), big.mark = ",")`
exactly. They are catching genuinely different problems — a contaminated
phenotype versus a degenerate fit.

# Caveats and what is not done

- **The discovery scan has not been rerun on this phenotype.** This page is
  calibration only, so there is no updated hit list. Note that excluding
  `r format(as.integer(gv("excluded_UC")) + as.integer(gv("excluded_phi")), big.mark = ",")`
  sites also changes the multiple-testing correction, so the new count will not
  be the old one minus the excluded hits.
- **The φ guard lives in a filter script, not in the fitter.** Until it is moved
  into `fit_bb_mle`, any analysis that bypasses this phenotype can reproduce the
  problem.
- **"Sites tested" is smaller than "sites in the phenotype"** — `r format(as.integer(gv("sites_final")), big.mark = ",")`
  retained but `r format(lam[run == FINAL & dataset == "real"]$n_sites, big.mark = ",")`
  tested, the rest having no cis-SNP at MAF ≥ 0.05 within ±5 kb or failing to
  fit. The two numbers are not interchangeable.
- **A goodness-of-fit test was run separately** and found that only a minority of
  the previously-problematic sites showed genuine distributional misfit; most had
  no multimodality at all. Its own calibration check failed in both directions
  (conservative with nominal p-values, anti-conservative with a bootstrap), so
  those p-values should be read as indicative only.
- **The real-data λ of `r sprintf("%.3f", lv(FINAL, "real", "lambda"))`** is
  assumed to reflect genuine cis signal. That is an interpretation, not a
  measurement — a residual artefact affecting many sites weakly would look the
  same.

# Reproducing this page

```{r prov}
knitr::kable(info, caption = "Parameters")
knitr::kable(data.table(
  step = c("untreated (pooled input) count matrices",
           "U/C + phi filtered phenotype, PCs recomputed",
           "phi filter iterated to convergence",
           "site-resolved calibration (real + 5 permutations)",
           "extraction of remaining problematic sites",
           "precompute for this page", "this page"),
  script = c("12.untreated_counts_matrix.R",
             "13.filtered_phenotype.R --criterion called --exclude-list <phi sites>",
             "18.phi_filter_iteration.R",
             "8.calibration_site_resolved.R --pheno fold5_union_noUCcalled_noPhi",
             "9.qq_exclude_pathological.R --stem calib_siteres_w5kb_pc3_noUCcalled_noPhi",
             "19.strategy_page_data.R",
             "analysis/qtl_mapping_betabinomial_strategy.Rmd")),
  caption = "Provenance. Both scan scripts take --pheno / --stem so earlier results are never overwritten.")
```

Related: [the 26 problematic sites](qtl_mapping_betabinomial_site26.html) ·
[calibration after the U/C filter](qtl_mapping_betabinomial_calibration_noUCcalled.html) ·
[full pathology write-up](qtl_mapping_betabinomial_pathology_summary.html)
